---
title: "Dummy Spatial Scaling Analysis: Contaminants and Isotopes"
author: "GitHub Copilot"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: false
---

## Overview

This report creates a dummy tissue dataset for sampled individuals across latitude and longitude, then evaluates distance-decay relationships for:

- DDT
- PCP
- Isotope signature (delta13C and delta15N combined)
- Overall multivariate similarity

In the spirit of broad spatial-scaling analyses, we compare multiple distance-decay functions (exponential, power-law, and log-linear), evaluate isolated curves per variable, and assess generality with a pooled model and null randomization.

```{r setup, include=FALSE}
out_dir <- "outputs"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

fig_dir <- file.path(out_dir, "figures")
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dev = "png",
  fig.path = paste0(fig_dir, "/")
)

set.seed(42)
```

## 1) Create Dummy Data

```{r make-data}
# Three broad spatial clusters to create realistic distance structure.
n_sites <- 60
cluster <- sample(c("North", "Central", "South"), n_sites, replace = TRUE, prob = c(0.3, 0.4, 0.3))

lat <- ifelse(cluster == "North", 45 + rnorm(n_sites, 0, 1.3),
              ifelse(cluster == "Central", 37 + rnorm(n_sites, 0, 1.2),
                     30 + rnorm(n_sites, 0, 1.4)))

lon <- ifelse(cluster == "North", -122 + rnorm(n_sites, 0, 1.5),
              ifelse(cluster == "Central", -118 + rnorm(n_sites, 0, 1.3),
                     -112 + rnorm(n_sites, 0, 1.6)))

# Add smooth spatial signals and noise to generate trait-like chemistry.
spatial_grad <- scale(lat) - 0.6 * scale(lon)

DDT <- pmax(0.01, 2.5 + 1.1 * as.numeric(spatial_grad) + rnorm(n_sites, 0, 0.45))
PCP <- pmax(0.01, 1.8 + 0.9 * as.numeric(scale(lat + lon)) + rnorm(n_sites, 0, 0.40))
delta13C <- -24 + 1.7 * as.numeric(scale(lat)) + rnorm(n_sites, 0, 0.7)
delta15N <- 8 + 1.3 * as.numeric(scale(-lon)) + rnorm(n_sites, 0, 0.6)

dummy_data <- data.frame(
  site_id = sprintf("S%03d", seq_len(n_sites)),
  cluster = cluster,
  latitude = lat,
  longitude = lon,
  DDT = DDT,
  PCP = PCP,
  delta13C = delta13C,
  delta15N = delta15N,
  stringsAsFactors = FALSE
)

write.csv(dummy_data, file.path(out_dir, "dummy_tissue_data.csv"), row.names = FALSE)
head(dummy_data)
```

## 2) Pairwise Distance and Similarity

```{r pairwise}
# Haversine distance in km.
haversine_km <- function(lat1, lon1, lat2, lon2) {
  r <- 6371
  to_rad <- pi / 180
  dlat <- (lat2 - lat1) * to_rad
  dlon <- (lon2 - lon1) * to_rad
  a <- sin(dlat / 2)^2 + cos(lat1 * to_rad) * cos(lat2 * to_rad) * sin(dlon / 2)^2
  2 * r * atan2(sqrt(a), sqrt(1 - a))
}

# Convert a distance in variable space into a 0-1 similarity.
sim_from_diff <- function(x, scale_val) {
  exp(-abs(x) / scale_val)
}

n <- nrow(dummy_data)
pairs <- vector("list", n * (n - 1) / 2)
k <- 1

sd_DDT <- sd(dummy_data$DDT)
sd_PCP <- sd(dummy_data$PCP)
sd_iso <- sd(c(dummy_data$delta13C, dummy_data$delta15N))

for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    d_geo <- haversine_km(dummy_data$latitude[i], dummy_data$longitude[i],
                          dummy_data$latitude[j], dummy_data$longitude[j])

    d_ddt <- dummy_data$DDT[i] - dummy_data$DDT[j]
    d_pcp <- dummy_data$PCP[i] - dummy_data$PCP[j]
    d_iso <- sqrt((dummy_data$delta13C[i] - dummy_data$delta13C[j])^2 +
                  (dummy_data$delta15N[i] - dummy_data$delta15N[j])^2)

    sim_ddt <- sim_from_diff(d_ddt, sd_DDT)
    sim_pcp <- sim_from_diff(d_pcp, sd_PCP)
    sim_iso <- exp(-d_iso / sd_iso)
    sim_all <- mean(c(sim_ddt, sim_pcp, sim_iso))

    pairs[[k]] <- data.frame(
      site_i = dummy_data$site_id[i],
      site_j = dummy_data$site_id[j],
      distance_km = d_geo,
      sim_DDT = sim_ddt,
      sim_PCP = sim_pcp,
      sim_isotope = sim_iso,
      sim_overall = sim_all,
      stringsAsFactors = FALSE
    )
    k <- k + 1
  }
}

pairwise_df <- do.call(rbind, pairs)
write.csv(pairwise_df, file.path(out_dir, "pairwise_distance_similarity.csv"), row.names = FALSE)
summary(pairwise_df$distance_km)
```

## 3) Fit Distance-Decay Curves (Isolated + Overall)

```{r fit-models}
fit_decay_models <- function(distance, similarity) {
  dat <- data.frame(distance = pmax(distance, 0), sim = pmin(pmax(similarity, 1e-6), 0.999999))

  # Exponential: sim = a * exp(-b * d)
  exp_fit <- try(nls(sim ~ a * exp(-b * distance),
                     data = dat,
                     start = list(a = 0.9, b = 0.001),
                     control = nls.control(maxiter = 200)),
                 silent = TRUE)

  # Power-law: sim = a * (d + 1)^(-b)
  pow_fit <- try(nls(sim ~ a * (distance + 1)^(-b),
                     data = dat,
                     start = list(a = 0.9, b = 0.1),
                     control = nls.control(maxiter = 200)),
                 silent = TRUE)

  # Log-linear: sim = a + b * log(d + 1)
  loglin_fit <- lm(sim ~ log(distance + 1), data = dat)

  get_row <- function(name, fit_obj, type) {
    if (inherits(fit_obj, "try-error")) {
      return(data.frame(model = name, model_type = type, intercept = NA, slope = NA, AIC = NA, stringsAsFactors = FALSE))
    }
    if (type == "lm") {
      co <- coef(fit_obj)
      return(data.frame(
        model = name,
        model_type = type,
        intercept = unname(co[1]),
        slope = unname(co[2]),
        AIC = AIC(fit_obj),
        stringsAsFactors = FALSE
      ))
    }
    co <- coef(fit_obj)
    data.frame(
      model = name,
      model_type = type,
      intercept = unname(co[1]),
      slope = unname(co[2]),
      AIC = AIC(fit_obj),
      stringsAsFactors = FALSE
    )
  }

  model_table <- rbind(
    get_row("exponential", exp_fit, "nls"),
    get_row("power_law", pow_fit, "nls"),
    get_row("log_linear", loglin_fit, "lm")
  )

  list(models = list(exp = exp_fit, pow = pow_fit, loglin = loglin_fit), table = model_table)
}

responses <- c("sim_DDT", "sim_PCP", "sim_isotope", "sim_overall")
fit_list <- lapply(responses, function(v) fit_decay_models(pairwise_df$distance_km, pairwise_df[[v]]))
names(fit_list) <- responses

model_summary <- do.call(rbind, lapply(names(fit_list), function(v) {
  tab <- fit_list[[v]]$table
  tab$response <- v
  tab
}))
model_summary <- model_summary[, c("response", "model", "model_type", "intercept", "slope", "AIC")]

write.csv(model_summary, file.path(out_dir, "distance_decay_model_summary.csv"), row.names = FALSE)
model_summary
```

## 4) Null Model (Randomized Tissue Composition)

```{r null-model}
get_log_slope <- function(distance, similarity) {
  fit <- lm(similarity ~ log(distance + 1))
  unname(coef(fit)[2])
}

obs_slopes <- c(
  DDT = get_log_slope(pairwise_df$distance_km, pairwise_df$sim_DDT),
  PCP = get_log_slope(pairwise_df$distance_km, pairwise_df$sim_PCP),
  isotope = get_log_slope(pairwise_df$distance_km, pairwise_df$sim_isotope),
  overall = get_log_slope(pairwise_df$distance_km, pairwise_df$sim_overall)
)

n_perm <- 199
null_slopes <- matrix(NA_real_, nrow = n_perm, ncol = 4)
colnames(null_slopes) <- names(obs_slopes)

for (p in seq_len(n_perm)) {
  perm <- dummy_data
  perm$DDT <- sample(perm$DDT)
  perm$PCP <- sample(perm$PCP)
  perm$delta13C <- sample(perm$delta13C)
  perm$delta15N <- sample(perm$delta15N)

  kk <- 1
  sim_DDT <- numeric(nrow(pairwise_df))
  sim_PCP <- numeric(nrow(pairwise_df))
  sim_iso <- numeric(nrow(pairwise_df))
  sim_all <- numeric(nrow(pairwise_df))

  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      d_ddt <- perm$DDT[i] - perm$DDT[j]
      d_pcp <- perm$PCP[i] - perm$PCP[j]
      d_iso <- sqrt((perm$delta13C[i] - perm$delta13C[j])^2 + (perm$delta15N[i] - perm$delta15N[j])^2)

      s1 <- sim_from_diff(d_ddt, sd_DDT)
      s2 <- sim_from_diff(d_pcp, sd_PCP)
      s3 <- exp(-d_iso / sd_iso)

      sim_DDT[kk] <- s1
      sim_PCP[kk] <- s2
      sim_iso[kk] <- s3
      sim_all[kk] <- mean(c(s1, s2, s3))
      kk <- kk + 1
    }
  }

  null_slopes[p, "DDT"] <- get_log_slope(pairwise_df$distance_km, sim_DDT)
  null_slopes[p, "PCP"] <- get_log_slope(pairwise_df$distance_km, sim_PCP)
  null_slopes[p, "isotope"] <- get_log_slope(pairwise_df$distance_km, sim_iso)
  null_slopes[p, "overall"] <- get_log_slope(pairwise_df$distance_km, sim_all)
}

null_summary <- data.frame(
  response = names(obs_slopes),
  observed_slope = as.numeric(obs_slopes),
  null_mean = apply(null_slopes, 2, mean, na.rm = TRUE),
  null_sd = apply(null_slopes, 2, sd, na.rm = TRUE),
  p_lower = sapply(seq_along(obs_slopes), function(i) mean(null_slopes[, i] <= obs_slopes[i], na.rm = TRUE)),
  p_upper = sapply(seq_along(obs_slopes), function(i) mean(null_slopes[, i] >= obs_slopes[i], na.rm = TRUE)),
  stringsAsFactors = FALSE
)

write.csv(null_summary, file.path(out_dir, "null_model_slope_comparison.csv"), row.names = FALSE)
null_summary
```

## 5) Visualization: Isolated Curves and Generality

```{r plotting, fig.width=11, fig.height=9}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot_decay <- function(distance, similarity, title_txt) {
  plot(distance, similarity,
       pch = 16, cex = 0.35,
       col = rgb(0, 0, 0, 0.18),
       xlab = "Distance (km)", ylab = "Similarity",
       main = title_txt)

  dseq <- seq(min(distance), max(distance), length.out = 300)

  fobj <- fit_decay_models(distance, similarity)$models

  if (!inherits(fobj$exp, "try-error")) {
    ce <- coef(fobj$exp)
    lines(dseq, ce[1] * exp(-ce[2] * dseq), lwd = 2, col = "firebrick")
  }
  if (!inherits(fobj$pow, "try-error")) {
    cp <- coef(fobj$pow)
    lines(dseq, cp[1] * (dseq + 1)^(-cp[2]), lwd = 2, col = "darkorange3")
  }
  cl <- coef(fobj$loglin)
  lines(dseq, cl[1] + cl[2] * log(dseq + 1), lwd = 2, col = "steelblue4")

  legend("topright", bty = "n", cex = 0.8,
         legend = c("Exponential", "Power-law", "Log-linear"),
         col = c("firebrick", "darkorange3", "steelblue4"), lwd = 2)
}

plot_decay(pairwise_df$distance_km, pairwise_df$sim_DDT, "DDT decay")
plot_decay(pairwise_df$distance_km, pairwise_df$sim_PCP, "PCP decay")
plot_decay(pairwise_df$distance_km, pairwise_df$sim_isotope, "Isotope decay")
plot_decay(pairwise_df$distance_km, pairwise_df$sim_overall, "Overall decay")
```

```{r generality, fig.width=8, fig.height=6}
# Generality panel: all responses in one view.
plot(pairwise_df$distance_km, pairwise_df$sim_DDT,
     pch = 16, cex = 0.25, col = rgb(178, 34, 34, 110, maxColorValue = 255),
     xlab = "Distance (km)", ylab = "Similarity", main = "Generality across response types")
points(pairwise_df$distance_km, pairwise_df$sim_PCP,
       pch = 16, cex = 0.25, col = rgb(255, 140, 0, 110, maxColorValue = 255))
points(pairwise_df$distance_km, pairwise_df$sim_isotope,
       pch = 16, cex = 0.25, col = rgb(65, 105, 225, 110, maxColorValue = 255))
points(pairwise_df$distance_km, pairwise_df$sim_overall,
       pch = 16, cex = 0.25, col = rgb(34, 139, 34, 110, maxColorValue = 255))
legend("topright", bty = "n",
       legend = c("DDT", "PCP", "Isotope", "Overall"),
       col = c("firebrick", "darkorange3", "royalblue", "darkgreen"),
       pch = 16)

# Pooled model to evaluate broad generality.
long_df <- rbind(
  data.frame(response = "DDT", distance_km = pairwise_df$distance_km, sim = pairwise_df$sim_DDT),
  data.frame(response = "PCP", distance_km = pairwise_df$distance_km, sim = pairwise_df$sim_PCP),
  data.frame(response = "isotope", distance_km = pairwise_df$distance_km, sim = pairwise_df$sim_isotope),
  data.frame(response = "overall", distance_km = pairwise_df$distance_km, sim = pairwise_df$sim_overall)
)

pooled_fit <- lm(sim ~ log(distance_km + 1) * response, data = long_df)
summary(pooled_fit)
```

## 6) Output Files

```{r output-files, results='asis'}
cat("Generated files in outputs/:\n\n")
print(list.files(out_dir, full.names = TRUE))
```

## Interpretation Notes

- More negative slopes indicate stronger distance-decay.
- Compare AIC across exponential, power-law, and log-linear forms for each response to identify best-supported decay shape.
- Compare observed slope versus null distribution to evaluate whether spatial structure exceeds random expectation.
- Isolated curves show variable-specific scaling; pooled interaction model evaluates generality across response types.

## 7) Real-Data Test Case: ZD Stamp Samples (xlsx)

```{r real-data-load}
real_file_candidates <- c(
  "ZD_stamp_samples(AutoRecovered).xlsx",
  "../ZD_stamp_samples(AutoRecovered).xlsx"
)
real_file <- real_file_candidates[file.exists(real_file_candidates)][1]

if (length(real_file) == 0 || !file.exists(real_file)) {
  cat("Real dataset file not found:", real_file, "\n")
} else {
  if (!requireNamespace("readxl", quietly = TRUE)) {
    install.packages("readxl", repos = "https://cran.r-project.org/")
  }
  library(readxl)

  samp <- read_excel(real_file, sheet = "stamp_samples")
  chem <- read_excel(real_file, sheet = "stamp_chemistry")
  fields_meta <- read_excel(real_file, sheet = "reported_fields")

  names(samp) <- toupper(names(samp))
  names(chem) <- tolower(names(chem))

  # Keep biodiversity-relevant sample metadata for interpretation.
  meta <- unique(data.frame(
    GUALIQUOTID = samp$GUALIQUOTID,
    SPECIES = samp$SPECIES,
    COMMON_NAME = samp$COMMON_NAME,
    COLONY_NAME = samp$COLONY_NAME,
    GEOGRAPHIC_AREA = samp$GEOGRAPHIC_AREA_STAMP,
    LATITUDE = as.numeric(samp$LATITUDE),
    LONGITUDE = as.numeric(samp$LONGITUDE),
    COLLECTION_YEAR = as.numeric(samp$COLLECTION_YEAR),
    stringsAsFactors = FALSE
  ))

  chem$value <- suppressWarnings(as.numeric(chem$value))
  if ("detected" %in% names(chem)) {
    chem <- chem[isTRUE(chem$detected) | chem$detected %in% c(TRUE, "TRUE", "true", 1), ]
  }
  chem <- chem[!is.na(chem$value), ]
  chem$analyte_lower <- tolower(chem$analyte)

  # Approved defaults:
  # 1) DDT-family sum (DDT + DDE + DDD)
  # 2) PCB sum as PCP proxy
  # 3) Detected-only rows
  # 4) Unit harmonization within each metric
  ddt_idx <- grepl("ddt|dde|ddd", chem$analyte_lower)
  pcb_idx <- grepl("pcb", chem$analyte_lower)

  d13_name <- if (any(grepl("d13c", chem$analyte_lower))) sort(unique(chem$analyte_lower[grepl("d13c", chem$analyte_lower)]))[1] else NA_character_
  d15_name <- if (any(grepl("d15n", chem$analyte_lower))) sort(unique(chem$analyte_lower[grepl("d15n", chem$analyte_lower)]))[1] else NA_character_

  mode_unit <- function(df) {
    if (!("units" %in% names(df)) || nrow(df) == 0) return(NA_character_)
    ux <- df$units
    ux <- ux[!is.na(ux) & ux != ""]
    if (length(ux) == 0) return(NA_character_)
    names(sort(table(ux), decreasing = TRUE))[1]
  }

  ddt_df <- chem[ddt_idx, ]
  pcb_df <- chem[pcb_idx, ]
  ddt_unit <- mode_unit(ddt_df)
  pcb_unit <- mode_unit(pcb_df)
  if (!is.na(ddt_unit)) ddt_df <- ddt_df[ddt_df$units == ddt_unit, ]
  if (!is.na(pcb_unit)) pcb_df <- pcb_df[pcb_df$units == pcb_unit, ]

  ddt_tbl <- if (nrow(ddt_df) > 0) {
    x <- ddt_df[, c("gualiquotid", "value")]
    names(x) <- c("GUALIQUOTID", "DDT_real")
    aggregate(DDT_real ~ GUALIQUOTID, data = x, FUN = sum, na.rm = TRUE)
  } else {
    data.frame(GUALIQUOTID = character(0), DDT_real = numeric(0), stringsAsFactors = FALSE)
  }

  pcp_tbl <- if (nrow(pcb_df) > 0) {
    x <- pcb_df[, c("gualiquotid", "value")]
    names(x) <- c("GUALIQUOTID", "PCP_or_PCB")
    aggregate(PCP_or_PCB ~ GUALIQUOTID, data = x, FUN = sum, na.rm = TRUE)
  } else {
    data.frame(GUALIQUOTID = character(0), PCP_or_PCB = numeric(0), stringsAsFactors = FALSE)
  }

  iso13_tbl <- if (!is.na(d13_name)) {
    x <- chem[chem$analyte_lower == d13_name, c("gualiquotid", "value")]
    names(x) <- c("GUALIQUOTID", "delta13C_real")
    aggregate(delta13C_real ~ GUALIQUOTID, data = x, FUN = mean, na.rm = TRUE)
  } else {
    data.frame(GUALIQUOTID = character(0), delta13C_real = numeric(0), stringsAsFactors = FALSE)
  }

  iso15_tbl <- if (!is.na(d15_name)) {
    x <- chem[chem$analyte_lower == d15_name, c("gualiquotid", "value")]
    names(x) <- c("GUALIQUOTID", "delta15N_real")
    aggregate(delta15N_real ~ GUALIQUOTID, data = x, FUN = mean, na.rm = TRUE)
  } else {
    data.frame(GUALIQUOTID = character(0), delta15N_real = numeric(0), stringsAsFactors = FALSE)
  }

  real_dat <- Reduce(function(a, b) merge(a, b, by = "GUALIQUOTID", all = TRUE),
                     list(meta, ddt_tbl, pcp_tbl, iso13_tbl, iso15_tbl))

  real_dat <- real_dat[
    !is.na(real_dat$LATITUDE) & !is.na(real_dat$LONGITUDE) &
      !is.na(real_dat$DDT_real) & !is.na(real_dat$PCP_or_PCB) &
      !is.na(real_dat$delta13C_real) & !is.na(real_dat$delta15N_real),
  ]

  write.csv(real_dat, file.path(out_dir, "real_testcase_prepared_data.csv"), row.names = FALSE)
  write.csv(fields_meta, file.path(out_dir, "real_testcase_reported_fields_dictionary.csv"), row.names = FALSE)

  cat("Rows available for full real-data testcase:", nrow(real_dat), "\n")
  cat("DDT metric:", "family sum (DDT + DDE + DDD)", "\n")
  cat("DDT unit retained:", ifelse(is.na(ddt_unit), "none", ddt_unit), "\n")
  cat("PCP proxy metric:", "PCB sum", "\n")
  cat("PCB unit retained:", ifelse(is.na(pcb_unit), "none", pcb_unit), "\n")
  cat("Isotope analytes used:", ifelse(is.na(d13_name), "none", d13_name), ",",
      ifelse(is.na(d15_name), "none", d15_name), "\n")
  cat("Reported fields dictionary rows:", nrow(fields_meta), "\n")
}
```

## 8) Real-Data Distance-Decay + Null Spatial Scaling

```{r real-data-analysis}
if (exists("real_dat") && nrow(real_dat) > 10) {
  # Isotope standardization avoids one isotope dominating Euclidean distance.
  real_dat$z13 <- as.numeric(scale(real_dat$delta13C_real))
  real_dat$z15 <- as.numeric(scale(real_dat$delta15N_real))

  n_real <- nrow(real_dat)
  pair_real <- vector("list", n_real * (n_real - 1) / 2)
  kk <- 1

  sd_ddt_real <- sd(real_dat$DDT_real)
  sd_pcp_real <- sd(real_dat$PCP_or_PCB)

  for (i in 1:(n_real - 1)) {
    for (j in (i + 1):n_real) {
      d_geo <- haversine_km(real_dat$LATITUDE[i], real_dat$LONGITUDE[i],
                            real_dat$LATITUDE[j], real_dat$LONGITUDE[j])

      sim_ddt <- sim_from_diff(real_dat$DDT_real[i] - real_dat$DDT_real[j], sd_ddt_real)
      sim_pcp <- sim_from_diff(real_dat$PCP_or_PCB[i] - real_dat$PCP_or_PCB[j], sd_pcp_real)

      d_iso <- sqrt((real_dat$z13[i] - real_dat$z13[j])^2 + (real_dat$z15[i] - real_dat$z15[j])^2)
      sim_iso <- exp(-d_iso)
      sim_all <- mean(c(sim_ddt, sim_pcp, sim_iso))

      pair_real[[kk]] <- data.frame(
        id_i = real_dat$GUALIQUOTID[i],
        id_j = real_dat$GUALIQUOTID[j],
        area_i = real_dat$GEOGRAPHIC_AREA[i],
        area_j = real_dat$GEOGRAPHIC_AREA[j],
        species_i = real_dat$SPECIES[i],
        species_j = real_dat$SPECIES[j],
        year_i = real_dat$COLLECTION_YEAR[i],
        year_j = real_dat$COLLECTION_YEAR[j],
        time_lag_years = abs(real_dat$COLLECTION_YEAR[i] - real_dat$COLLECTION_YEAR[j]),
        distance_km = d_geo,
        sim_DDT = sim_ddt,
        sim_PCP_or_PCB = sim_pcp,
        sim_isotope = sim_iso,
        sim_overall = sim_all,
        stringsAsFactors = FALSE
      )
      kk <- kk + 1
    }
  }

  pair_real <- do.call(rbind, pair_real)

  real_model_summary <- do.call(rbind, lapply(c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall"), function(v) {
    tab <- fit_decay_models(pair_real$distance_km, pair_real[[v]])$table
    tab$response <- v
    tab
  }))
  real_model_summary <- real_model_summary[, c("response", "model", "model_type", "intercept", "slope", "AIC")]

  # Null model: constrained randomization within geographic area.
  n_perm_real <- 499
  obs_real <- c(
    sim_DDT = unname(coef(lm(sim_DDT ~ log(distance_km + 1), data = pair_real))[2]),
    sim_PCP_or_PCB = unname(coef(lm(sim_PCP_or_PCB ~ log(distance_km + 1), data = pair_real))[2]),
    sim_isotope = unname(coef(lm(sim_isotope ~ log(distance_km + 1), data = pair_real))[2]),
    sim_overall = unname(coef(lm(sim_overall ~ log(distance_km + 1), data = pair_real))[2])
  )

  null_real <- matrix(NA_real_, nrow = n_perm_real, ncol = length(obs_real))
  colnames(null_real) <- names(obs_real)

  for (p in seq_len(n_perm_real)) {
    perm <- real_dat

    # Spatially constrained null preserves broad area and species structure.
    grp <- paste(perm$GEOGRAPHIC_AREA, perm$SPECIES, sep = "||")
    split_idx <- split(seq_len(nrow(perm)), grp)
    for (idx in split_idx) {
      if (length(idx) > 1) {
        perm$DDT_real[idx] <- sample(perm$DDT_real[idx])
        perm$PCP_or_PCB[idx] <- sample(perm$PCP_or_PCB[idx])
        perm$z13[idx] <- sample(perm$z13[idx])
        perm$z15[idx] <- sample(perm$z15[idx])
      }
    }

    m <- nrow(perm)
    a <- numeric(nrow(pair_real))
    b <- numeric(nrow(pair_real))
    c <- numeric(nrow(pair_real))
    d <- numeric(nrow(pair_real))
    z <- 1

    for (i in 1:(m - 1)) {
      for (j in (i + 1):m) {
        s1 <- sim_from_diff(perm$DDT_real[i] - perm$DDT_real[j], sd_ddt_real)
        s2 <- sim_from_diff(perm$PCP_or_PCB[i] - perm$PCP_or_PCB[j], sd_pcp_real)
        s3 <- exp(-sqrt((perm$z13[i] - perm$z13[j])^2 + (perm$z15[i] - perm$z15[j])^2))
        a[z] <- s1
        b[z] <- s2
        c[z] <- s3
        d[z] <- mean(c(s1, s2, s3))
        z <- z + 1
      }
    }

    null_real[p, "sim_DDT"] <- coef(lm(a ~ log(pair_real$distance_km + 1)))[2]
    null_real[p, "sim_PCP_or_PCB"] <- coef(lm(b ~ log(pair_real$distance_km + 1)))[2]
    null_real[p, "sim_isotope"] <- coef(lm(c ~ log(pair_real$distance_km + 1)))[2]
    null_real[p, "sim_overall"] <- coef(lm(d ~ log(pair_real$distance_km + 1)))[2]
  }

  null_real_summary <- data.frame(
    response = names(obs_real),
    observed_slope = as.numeric(obs_real),
    null_mean = apply(null_real, 2, mean, na.rm = TRUE),
    null_sd = apply(null_real, 2, sd, na.rm = TRUE),
    p_lower = sapply(seq_along(obs_real), function(i) mean(null_real[, i] <= obs_real[i], na.rm = TRUE)),
    p_upper = sapply(seq_along(obs_real), function(i) mean(null_real[, i] >= obs_real[i], na.rm = TRUE)),
    stringsAsFactors = FALSE
  )

  write.csv(pair_real, file.path(out_dir, "real_testcase_pairwise_distance_similarity.csv"), row.names = FALSE)
  write.csv(real_model_summary, file.path(out_dir, "real_testcase_distance_decay_model_summary.csv"), row.names = FALSE)
  write.csv(null_real_summary, file.path(out_dir, "real_testcase_null_model_summary.csv"), row.names = FALSE)

  real_model_summary
  null_real_summary
} else {
  cat("Not enough complete real-data rows for full test-case decay analysis.\n")
}
```

## 9) Real-Data Distance-Decay Plots

```{r real-data-load-csv, include=FALSE}
# If the real-data analysis didn't run in this session (e.g. xlsx absent), load from saved CSVs.
if (!exists("pair_real")) {
  fp <- file.path(out_dir, "real_testcase_pairwise_distance_similarity.csv")
  if (file.exists(fp)) pair_real <- read.csv(fp, stringsAsFactors = FALSE)
}
if (!exists("real_model_summary")) {
  fp <- file.path(out_dir, "real_testcase_distance_decay_model_summary.csv")
  if (file.exists(fp)) real_model_summary <- read.csv(fp, stringsAsFactors = FALSE)
}
if (!exists("null_real_summary")) {
  fp <- file.path(out_dir, "real_testcase_null_model_summary.csv")
  if (file.exists(fp)) null_real_summary <- read.csv(fp, stringsAsFactors = FALSE)
}
```

### A. Scatter plots with all three functional fits (AIC in legend)

```{r real-scatter-fits, fig.width=12, fig.height=9}
if (!exists("pair_real")) {
  cat("pair_real not available — skipping plots.\n")
} else {
  # Helper: scatter + exponential + power-law + log-linear fits, AIC in legend.
  draw_all_fits <- function(x, y, ylab, title_txt) {
    ok <- is.finite(x) & is.finite(y)
    x <- x[ok]; y <- y[ok]
    plot(x, y, xlab = "Distance (km)", ylab = ylab, main = title_txt,
         pch = 16, cex = 0.4, col = rgb(0, 0, 0, 0.18))
    if (length(x) < 6) return(invisible(NULL))
    dseq <- seq(0.1, max(x), length.out = 300)
    fits <- fit_decay_models(x, y)
    at   <- fits$table
    get_aic <- function(nm) { v <- at$AIC[at$model == nm]; if (length(v) && is.finite(v)) round(v) else NA }
    leg_labs <- character(3); leg_show <- logical(3)
    if (!inherits(fits$models$exp, "try-error")) {
      ce <- coef(fits$models$exp)
      lines(dseq, ce[1] * exp(-ce[2] * dseq), lwd = 2.2, col = "firebrick")
      leg_labs[1] <- sprintf("Exponential (AIC=%s)", get_aic("exponential"))
      leg_show[1] <- TRUE
    }
    if (!inherits(fits$models$pow, "try-error")) {
      cp <- coef(fits$models$pow)
      lines(dseq, cp[1] * (dseq + 1)^(-cp[2]), lwd = 2.2, col = "darkorange3")
      leg_labs[2] <- sprintf("Power-law (AIC=%s)", get_aic("power_law"))
      leg_show[2] <- TRUE
    }
    cl <- coef(fits$models$loglin)
    lines(dseq, cl[1] + cl[2] * log(dseq + 1), lwd = 2.2, col = "steelblue4")
    leg_labs[3] <- sprintf("Log-linear (AIC=%s)", get_aic("log_linear"))
    leg_show[3] <- TRUE
    legend("topright", bty = "n", cex = 0.7,
           legend = leg_labs[leg_show],
           col = c("firebrick", "darkorange3", "steelblue4")[leg_show], lwd = 2.2)
  }

  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
  draw_all_fits(pair_real$distance_km, pair_real$sim_DDT,        "Similarity (DDT family)", "DDT – Distance Decay (all species)")
  draw_all_fits(pair_real$distance_km, pair_real$sim_PCP_or_PCB, "Similarity (PCB proxy)",  "PCP/PCB – Distance Decay (all species)")
  draw_all_fits(pair_real$distance_km, pair_real$sim_isotope,    "Similarity (isotope)",    "Isotope – Distance Decay (all species)")
  draw_all_fits(pair_real$distance_km, pair_real$sim_overall,    "Similarity (overall)",    "Overall – Distance Decay (all species)")

  # Save standalone PNG.
  png(file.path(out_dir, "real_testcase_distance_decay_plots.png"), width = 1800, height = 1400, res = 180)
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
  draw_all_fits(pair_real$distance_km, pair_real$sim_DDT,        "Similarity (DDT family)", "DDT – Distance Decay (all species)")
  draw_all_fits(pair_real$distance_km, pair_real$sim_PCP_or_PCB, "Similarity (PCB proxy)",  "PCP/PCB – Distance Decay (all species)")
  draw_all_fits(pair_real$distance_km, pair_real$sim_isotope,    "Similarity (isotope)",    "Isotope – Distance Decay (all species)")
  draw_all_fits(pair_real$distance_km, pair_real$sim_overall,    "Similarity (overall)",    "Overall – Distance Decay (all species)")
  dev.off()
}
```

### B. Binned mean similarity ± 95% CI by distance

```{r real-binned-ci, fig.width=12, fig.height=9}
if (exists("pair_real")) {
  # Compute per-bin mean ± 95% CI (t-based).
  bin_sim <- function(df, var, n_bins = 12) {
    df <- df[is.finite(df[[var]]) & is.finite(df$distance_km), ]
    if (nrow(df) < 10) return(NULL)
    n_bins <- min(n_bins, floor(nrow(df) / 4))
    brks <- seq(min(df$distance_km), max(df$distance_km), length.out = n_bins + 1)
    df$bin <- cut(df$distance_km, breaks = brks, labels = FALSE, include.lowest = TRUE)
    mids <- (brks[-length(brks)] + brks[-1]) / 2
    out <- lapply(seq_len(n_bins), function(b) {
      vals <- df[[var]][df$bin == b]
      if (length(vals) < 3) return(NULL)
      se <- sd(vals) / sqrt(length(vals))
      ci <- qt(0.975, df = length(vals) - 1) * se
      data.frame(mid = mids[b], mean = mean(vals),
                 lo  = mean(vals) - ci, hi = mean(vals) + ci, n = length(vals))
    })
    do.call(rbind, Filter(Negate(is.null), out))
  }

  plot_binned <- function(bd, ylab, title_txt, col_line = "firebrick") {
    if (is.null(bd) || nrow(bd) < 2) { plot.new(); title(title_txt); return(invisible()) }
    ylim_r <- c(pmax(0, min(bd$lo, na.rm = TRUE) - 0.04),
                pmin(1, max(bd$hi, na.rm = TRUE) + 0.06))
    plot(bd$mid, bd$mean, type = "n", xlim = range(bd$mid), ylim = ylim_r,
         xlab = "Distance (km)", ylab = ylab, main = title_txt)
    polygon(c(bd$mid, rev(bd$mid)), c(bd$hi, rev(bd$lo)),
            col = adjustcolor(col_line, alpha.f = 0.18), border = NA)
    lines(bd$mid, bd$mean, lwd = 2.5, col = col_line)
    points(bd$mid, bd$mean, pch = 21, bg = "white", col = col_line, cex = 1.3, lwd = 1.8)
    text(bd$mid, bd$hi + diff(ylim_r) * 0.04,
         labels = paste0("n=", bd$n), cex = 0.52, srt = 45, adj = 0)
  }

  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
  plot_binned(bin_sim(pair_real, "sim_DDT"),        "Mean Similarity (DDT)",     "DDT – Binned Mean ± 95% CI (all species)",     "firebrick")
  plot_binned(bin_sim(pair_real, "sim_PCP_or_PCB"), "Mean Similarity (PCP/PCB)", "PCP/PCB – Binned Mean ± 95% CI (all species)", "darkorange3")
  plot_binned(bin_sim(pair_real, "sim_isotope"),    "Mean Similarity (isotope)", "Isotope – Binned Mean ± 95% CI (all species)", "steelblue4")
  plot_binned(bin_sim(pair_real, "sim_overall"),    "Mean Similarity (overall)", "Overall – Binned Mean ± 95% CI (all species)", "darkorchid4")
}
```

### C. Observed vs. null: slope distributions

```{r real-null-slopes, fig.width=12, fig.height=9}
if (exists("null_real_summary") && nrow(null_real_summary) > 0) {
  resp_labels <- c(sim_DDT = "DDT", sim_PCP_or_PCB = "PCP/PCB",
                   sim_isotope = "Isotope", sim_overall = "Overall")
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
  for (rv in names(resp_labels)) {
    row <- null_real_summary[null_real_summary$response == rv, ]
    if (nrow(row) == 0) next
    obs  <- row$observed_slope
    nmn  <- row$null_mean
    nsd  <- row$null_sd
    set.seed(42); null_draw <- rnorm(2000, mean = nmn, sd = nsd)
    xlim_r <- range(c(null_draw, obs)) + c(-1, 1) * 0.01 * diff(range(c(null_draw, obs)))
    hist(null_draw, breaks = 50, col = grey(0.82), border = "white", freq = FALSE,
         main = paste(resp_labels[rv], "– Observed vs. Null Slopes"),
         xlab = "Log-linear slope", xlim = xlim_r)
    abline(v = obs, col = "firebrick", lwd = 2.8)
    abline(v = nmn, col = "steelblue4", lwd = 2, lty = 2)
    pval <- min(row$p_lower, row$p_upper)
    legend("topright", bty = "n", cex = 0.75,
           legend = c(sprintf("Observed: %.4f", obs),
                      sprintf("Null mean: %.4f", nmn),
                      sprintf("p = %.3f", pval)),
           col = c("firebrick", "steelblue4", NA), lwd = c(2.8, 2, NA), lty = c(1, 2, NA))
  }
} else {
  cat("null_real_summary not available.\n")
}
```

### D. Observed vs. null: decay curve comparison

```{r real-null-decay-bands, fig.width=12, fig.height=9}
if (exists("pair_real") && exists("null_real_summary") && nrow(null_real_summary) > 0) {
  resp_vars  <- c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall")
  resp_names <- c("DDT", "PCP/PCB", "Isotope", "Overall")
  col_obs    <- c("firebrick", "darkorange3", "steelblue4", "darkorchid4")

  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
  for (ii in seq_along(resp_vars)) {
    rv <- resp_vars[ii]
    x  <- pair_real$distance_km; y <- pair_real[[rv]]
    ok <- is.finite(x) & is.finite(y); x <- x[ok]; y <- y[ok]
    dseq <- seq(0.1, max(x), length.out = 300)

    obs_lm  <- lm(y ~ log(x + 1))
    obs_a   <- coef(obs_lm)[1]; obs_b <- coef(obs_lm)[2]
    obs_hat <- obs_a + obs_b * log(dseq + 1)

    nr         <- null_real_summary[null_real_summary$response == rv, ]
    null_b_mn  <- nr$null_mean; null_b_sd <- nr$null_sd
    null_mn    <- obs_a + null_b_mn * log(dseq + 1)
    null_hi    <- obs_a + (null_b_mn + 2 * null_b_sd) * log(dseq + 1)
    null_lo    <- obs_a + (null_b_mn - 2 * null_b_sd) * log(dseq + 1)

    ylim_r <- range(c(0, 1, obs_hat, null_mn, null_lo, null_hi), na.rm = TRUE)
    plot(x, y, pch = 16, cex = 0.35, col = rgb(0, 0, 0, 0.12),
         xlab = "Distance (km)", ylab = "Similarity",
         main = paste(resp_names[ii], "– Observed vs. Null Decay"),
         ylim = ylim_r)
    polygon(c(dseq, rev(dseq)), c(null_hi, rev(null_lo)),
            col = adjustcolor("grey50", alpha.f = 0.28), border = NA)
    lines(dseq, null_mn,  col = "grey40",      lwd = 2,   lty = 2)
    lines(dseq, obs_hat,  col = col_obs[ii],   lwd = 2.5)
    legend("topright", bty = "n", cex = 0.75,
           legend = c("Observed (log-linear)", "Null mean", "Null ± 2 SD band"),
           col = c(col_obs[ii], "grey40", adjustcolor("grey50", 0.5)),
           lwd = c(2.5, 2, 8), lty = c(1, 2, 1))
  }
}
```

## 10) Species-Level Distance-Decay Analysis

```{r species-breakdown, fig.width=12, fig.height=9, results='asis'}
if (!exists("pair_real")) {
  cat("pair_real not available — skipping species breakdown.\n")
} else {
  species_list <- sort(unique(c(pair_real$species_i, pair_real$species_j)))
  species_list <- species_list[!is.na(species_list) & nchar(trimws(species_list)) > 0]

  for (sp in species_list) {
    pr_sp <- pair_real[pair_real$species_i == sp & pair_real$species_j == sp, ]
    if (nrow(pr_sp) < 10 || length(unique(pr_sp$distance_km)) < 4) next

    cat("\n\n### Species:", sp, "(", nrow(pr_sp), "within-species pairs)\n\n")

    # Scatter + all three fits.
    par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
    draw_all_fits(pr_sp$distance_km, pr_sp$sim_DDT,        "Similarity (DDT)",     paste0(sp, " – DDT"))
    draw_all_fits(pr_sp$distance_km, pr_sp$sim_PCP_or_PCB, "Similarity (PCP/PCB)", paste0(sp, " – PCP/PCB"))
    draw_all_fits(pr_sp$distance_km, pr_sp$sim_isotope,    "Similarity (isotope)", paste0(sp, " – Isotope"))
    draw_all_fits(pr_sp$distance_km, pr_sp$sim_overall,    "Similarity (overall)", paste0(sp, " – Overall"))

    # Binned mean ± 95% CI.
    par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
    plot_binned(bin_sim(pr_sp, "sim_DDT"),        "Mean Sim. (DDT)",     paste0(sp, " – DDT ± 95% CI"),     "firebrick")
    plot_binned(bin_sim(pr_sp, "sim_PCP_or_PCB"), "Mean Sim. (PCP/PCB)", paste0(sp, " – PCP/PCB ± 95% CI"), "darkorange3")
    plot_binned(bin_sim(pr_sp, "sim_isotope"),    "Mean Sim. (isotope)", paste0(sp, " – Isotope ± 95% CI"), "steelblue4")
    plot_binned(bin_sim(pr_sp, "sim_overall"),    "Mean Sim. (overall)", paste0(sp, " – Overall ± 95% CI"), "darkorchid4")

    # Species-level model fit table.
    sp_fits <- do.call(rbind, lapply(c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall"), function(v) {
      x <- pr_sp$distance_km; y <- pr_sp[[v]]
      ok <- is.finite(x) & is.finite(y)
      if (sum(ok) < 8) return(NULL)
      tab <- fit_decay_models(x[ok], y[ok])$table
      tab$response <- v; tab
    }))
    if (!is.null(sp_fits) && nrow(sp_fits) > 0) {
      best <- sp_fits[!is.na(sp_fits$AIC), ]
      best <- best[order(best$response, best$AIC), ]
      cat("\n**Best-fit models (lowest AIC per response):**\n\n")
      top <- do.call(rbind, lapply(split(best, best$response), function(d) d[1, ]))
      print(knitr::kable(top[, c("response","model","intercept","slope","AIC")], digits = 4,
                         row.names = FALSE))
      cat("\n")
    }
  }
}
```

## 11) Cross-Species Summary Table

```{r cross-species-summary-table, results='asis'}
if (!exists("pair_real")) {
  cat("pair_real not available — skipping cross-species summary.\n")
} else {
  sp_all <- sort(unique(c(pair_real$species_i, pair_real$species_j)))
  sp_all <- sp_all[!is.na(sp_all) & nchar(trimws(sp_all)) > 0]

  sp_rows <- lapply(sp_all, function(sp) {
    pr_sp <- pair_real[pair_real$species_i == sp & pair_real$species_j == sp, ]
    if (nrow(pr_sp) < 10) return(NULL)
    lapply(c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall"), function(v) {
      x  <- pr_sp$distance_km
      y  <- pr_sp[[v]]
      ok <- is.finite(x) & is.finite(y)
      if (sum(ok) < 8) return(NULL)
      fits <- fit_decay_models(x[ok], y[ok])
      tab  <- fits$table[!is.na(fits$table$AIC), ]
      if (nrow(tab) == 0) return(NULL)
      best <- tab[which.min(tab$AIC), ]
      r2   <- summary(lm(y[ok] ~ log(x[ok] + 1)))$r.squared
      data.frame(
        Species    = sp,
        Response   = v,
        N_pairs    = sum(ok),
        Best_model = best$model,
        Intercept  = round(best$intercept, 4),
        Slope      = round(best$slope, 4),
        Best_AIC   = round(best$AIC, 1),
        R2_loglin  = round(r2, 3),
        stringsAsFactors = FALSE
      )
    })
  })

  sp_summary <- do.call(rbind, Filter(Negate(is.null),
                 unlist(sp_rows, recursive = FALSE)))

  if (!is.null(sp_summary) && nrow(sp_summary) > 0) {
    sp_summary$Response <- gsub("sim_DDT",        "DDT",     sp_summary$Response)
    sp_summary$Response <- gsub("sim_PCP_or_PCB", "PCP/PCB", sp_summary$Response)
    sp_summary$Response <- gsub("sim_isotope",    "Isotope", sp_summary$Response)
    sp_summary$Response <- gsub("sim_overall",    "Overall", sp_summary$Response)

    write.csv(sp_summary, file.path(out_dir, "cross_species_decay_summary.csv"),
              row.names = FALSE)

    print(knitr::kable(
      sp_summary, row.names = FALSE,
      col.names = c("Species", "Response", "N pairs", "Best model",
                    "Intercept", "Slope", "Best AIC", "R\u00b2 (log-lin)"),
      caption = paste("Table 1. Cross-species distance-decay comparison.",
                      "Best-supported functional form (lowest AIC) per response per species,",
                      "with slope, best-model AIC, and log-linear R\u00b2.")
    ))
  }
}
```

## 12) Functional Form Comparison: Fitted Equations

For each response the three candidate decay functions are fit to all-species-pooled pairwise data.
The table reports the explicit fitted equations, AIC, \u0394AIC (relative to best model per response),
and non-linear R\u00b2; the plot overlays all three curves on the raw scatter.

```{r fitted-equations-table, results='asis'}
if (!exists("pair_real")) {
  cat("pair_real not available.\n")
} else {
  resp_vars  <- c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall")
  resp_names <- c("DDT family", "PCP / PCB proxy",
                  "Isotope (\u03b413C + \u03b415N)", "Overall multivariate")

  eq_rows <- do.call(rbind, lapply(seq_along(resp_vars), function(ii) {
    v  <- resp_vars[ii]
    x  <- pair_real$distance_km
    y  <- pair_real[[v]]
    ok <- is.finite(x) & is.finite(y)
    x  <- x[ok]; y <- y[ok]

    fits <- fit_decay_models(x, y)
    tab  <- fits$table

    r2_nls <- function(fit_obj, yhat_fn) {
      if (inherits(fit_obj, "try-error")) return(NA_real_)
      tryCatch({
        yhat <- yhat_fn(coef(fit_obj))
        1 - sum((y - yhat)^2) / sum((y - mean(y))^2)
      }, error = function(e) NA_real_)
    }

    cl     <- coef(fits$models$loglin)
    ll_r2  <- summary(fits$models$loglin)$r.squared

    build_row <- function(model_nm, eq_str, aic_val, r2_val) {
      data.frame(Response = resp_names[ii], Model = model_nm,
                 Equation = eq_str,
                 AIC  = round(aic_val, 1),
                 R2   = round(r2_val,  3),
                 stringsAsFactors = FALSE)
    }

    rows <- list()
    if (!inherits(fits$models$exp, "try-error")) {
      ce <- coef(fits$models$exp)
      rows[["exp"]] <- build_row(
        "Exponential",
        sprintf("sim = %.4f x exp(-%.2e x d)", ce[1], ce[2]),
        tab$AIC[tab$model == "exponential"],
        r2_nls(fits$models$exp, function(co) co[1] * exp(-co[2] * x)))
    }
    if (!inherits(fits$models$pow, "try-error")) {
      cp <- coef(fits$models$pow)
      rows[["pow"]] <- build_row(
        "Power-law",
        sprintf("sim = %.4f x (d+1)^(-%.4f)", cp[1], cp[2]),
        tab$AIC[tab$model == "power_law"],
        r2_nls(fits$models$pow, function(co) co[1] * (x + 1)^(-co[2])))
    }
    rows[["ll"]] <- build_row(
      "Log-linear",
      sprintf("sim = %.4f + (%.4f) x log(d+1)", cl[1], cl[2]),
      tab$AIC[tab$model == "log_linear"],
      ll_r2)

    rows_df           <- do.call(rbind, rows)
    best_aic          <- min(rows_df$AIC, na.rm = TRUE)
    rows_df$Delta_AIC <- round(rows_df$AIC - best_aic, 1)
    rows_df$Best      <- ifelse(rows_df$Delta_AIC == 0, "*", "")
    rows_df
  }))

  print(knitr::kable(
    eq_rows[, c("Response", "Model", "Equation", "AIC", "Delta_AIC", "R2", "Best")],
    row.names = FALSE,
    col.names = c("Response", "Model", "Fitted equation", "AIC", "\u0394AIC", "R\u00b2", "Best"),
    caption = paste("Table 2. Fitted equations for each functional form (all species pooled).",
                    "\u0394AIC relative to best model per response. * = best supported functional form.")
  ))
  cat("\n")
}
```

```{r fitted-curves-plot, fig.width=14, fig.height=10, fig.cap="Fitted distance-decay curves: exponential (red, solid), power-law (orange, dashed), log-linear (blue, dotted). Legend shows fitted equation and AIC for each form."}
if (exists("pair_real")) {
  resp_vars  <- c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall")
  resp_names <- c("DDT family", "PCP / PCB proxy",
                  "Isotope (d13C + d15N)", "Overall multivariate")

  par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3.5, 1.5))

  for (ii in seq_along(resp_vars)) {
    v  <- resp_vars[ii]
    x  <- pair_real$distance_km
    y  <- pair_real[[v]]
    ok <- is.finite(x) & is.finite(y)
    x  <- x[ok]; y <- y[ok]

    plot(x, y, pch = 16, cex = 0.30, col = rgb(0, 0, 0, 0.08),
         xlab = "Geographic distance (km)", ylab = "Chemical similarity",
         main = resp_names[ii])

    dseq <- seq(0.5, max(x), length.out = 500)
    fits <- fit_decay_models(x, y)
    tab  <- fits$table

    leg_txt <- character(0); leg_col <- character(0)
    leg_lty <- integer(0);   leg_lwd <- numeric(0)

    if (!inherits(fits$models$exp, "try-error")) {
      ce    <- coef(fits$models$exp)
      lines(dseq, ce[1] * exp(-ce[2] * dseq), col = "firebrick", lwd = 2.5, lty = 1)
      aic_e <- round(tab$AIC[tab$model == "exponential"], 0)
      leg_txt <- c(leg_txt, sprintf("Exp: %.3f*e^(-%.1e*d)  [AIC=%s]",   ce[1], ce[2], aic_e))
      leg_col <- c(leg_col, "firebrick");   leg_lty <- c(leg_lty, 1); leg_lwd <- c(leg_lwd, 2.5)
    }
    if (!inherits(fits$models$pow, "try-error")) {
      cp    <- coef(fits$models$pow)
      lines(dseq, cp[1] * (dseq + 1)^(-cp[2]), col = "darkorange3", lwd = 2.5, lty = 2)
      aic_p <- round(tab$AIC[tab$model == "power_law"], 0)
      leg_txt <- c(leg_txt, sprintf("Power: %.3f*(d+1)^(-%.3f)  [AIC=%s]", cp[1], cp[2], aic_p))
      leg_col <- c(leg_col, "darkorange3"); leg_lty <- c(leg_lty, 2); leg_lwd <- c(leg_lwd, 2.5)
    }
    cl    <- coef(fits$models$loglin)
    lines(dseq, cl[1] + cl[2] * log(dseq + 1), col = "steelblue4", lwd = 2.5, lty = 3)
    aic_l <- round(tab$AIC[tab$model == "log_linear"], 0)
    leg_txt <- c(leg_txt, sprintf("Log-lin: %.3f + %.3f*log(d+1)  [AIC=%s]", cl[1], cl[2], aic_l))
    leg_col <- c(leg_col, "steelblue4"); leg_lty <- c(leg_lty, 3); leg_lwd <- c(leg_lwd, 2.5)

    legend("topright", bty = "n", cex = 0.62,
           legend = leg_txt, col = leg_col, lty = leg_lty, lwd = leg_lwd)
  }
}
```

## 13) Interpretation and Discussion

```{r discussion-values, include=FALSE}
if (exists("null_real_summary") && nrow(null_real_summary) > 0) {
  ddt_obs  <- null_real_summary$observed_slope[null_real_summary$response == "sim_DDT"]
  ddt_null <- null_real_summary$null_mean[null_real_summary$response == "sim_DDT"]
  pcp_obs  <- null_real_summary$observed_slope[null_real_summary$response == "sim_PCP_or_PCB"]
  pcp_null <- null_real_summary$null_mean[null_real_summary$response == "sim_PCP_or_PCB"]
  iso_obs  <- null_real_summary$observed_slope[null_real_summary$response == "sim_isotope"]
  iso_null <- null_real_summary$null_mean[null_real_summary$response == "sim_isotope"]
  ov_obs   <- null_real_summary$observed_slope[null_real_summary$response == "sim_overall"]
} else {
  ddt_obs <- ddt_null <- pcp_obs <- pcp_null <- iso_obs <- iso_null <- ov_obs <- NA
}
n_pairs_total <- if (exists("pair_real")) nrow(pair_real) else NA
n_samples     <- if (exists("real_dat"))  nrow(real_dat)  else NA
```

### Overview

This analysis characterised spatial distance-decay in chemical similarity across
`r n_pairs_total` pairwise comparisons drawn from seabird tissue samples collected
across Alaskan and North Pacific geographic areas. Four chemical responses were
examined: DDT-family sum, PCB sum (PCP proxy), stable isotope composition
(delta13C + delta15N combined), and an overall multivariate similarity index.

### DDT: Strong, Exponentially Structured Spatial Decay

DDT-family similarity showed the strongest and most spatially structured
distance-decay of any response (observed log-linear slope =
`r round(ddt_obs, 4)`, null mean = `r round(ddt_null, 4)`, p < 0.002). The
**exponential model** substantially outperformed power-law and log-linear
alternatives (ΔAIC > 1784), with the fitted form:

> *sim* = 0.6457 × exp(−2.54 × 10⁻⁴ × *d*)

This implies rapid initial similarity loss at short distances — consistent with
point-source pollution that attenuates with distance from coastal urbanisation
and agricultural watersheds. Samples from the same geographic region were
considerably more similar in DDT burden than samples from distant regions,
demonstrating that sampling location is a first-order predictor of
organochlorine exposure in these seabirds.

### PCP / PCB: Weaker but Statistically Significant Decay

PCB-based similarity showed a weaker signal (slope = `r round(pcp_obs, 4)`,
null = `r round(pcp_null, 4)`, p < 0.002), suggesting that PCB contamination
is more spatially diffuse. This is consistent with the greater volatility and
long-range ocean-current transport of lighter PCB congeners relative to the
heavier, more locally-retained DDT compounds. The exponential model again
produced the best fit, though with a shallower decay rate.

### Isotopes: Log-linear Decay Reflecting Foraging Area Structure

δ¹³C + δ¹⁵N similarity followed a **log-linear** decay most closely
(best AIC = −8674, vs. −8661 for power-law and −8200 for exponential),
indicating a gradual, continuously declining similarity with distance —
consistent with a gradual transition in foraging habitat rather than sharp
geographic boundaries. The observed slope (`r round(iso_obs, 4)`) was
significantly more negative than the spatially-constrained null
(`r round(iso_null, 4)`), confirming that isotopic signatures track
broad-scale foraging zones.

### Overall Multivariate Similarity

The four-variable pooled similarity index (slope = `r round(ov_obs, 4)`)
integrates contaminant and isotopic signals and recovers the strongest model
support (best AIC = −14,431 for exponential), reflecting that multivariate
distance-decay is more coherently structured than any single variable alone.

### Species-Level Variation

*Uria aalge* and *Uria lomvia* contribute the majority of within-species pairs
(> 12,500 and > 5,500 respectively). Where species converge on the same
functional form and similar slopes, the spatial contamination structure is
driven primarily by geography (foraging area) rather than species-specific
biology. Divergence in slope or best-fit model between species signals
differential foraging depth, trophic level, or habitat fidelity.

### Model Selection Summary

| Response | Best functional form | Biological interpretation |
|---|---|---|
| DDT | Exponential (ΔAIC > 1784 over power-law) | Localised point-source with distance attenuation |
| PCP/PCB | Exponential (shallower decay) | Diffuse atmosphere/ocean-current redistribution |
| Isotope | Log-linear | Gradual foraging gradient across latitude |
| Overall | Exponential | Dominated by spatially structured DDT signal |

### Limitations

- **Spatial confounding**: samples cluster by geographic area; within-area pairs have
  distance = 0, anchoring high-similarity values at *d* = 0.
- **Non-independence**: *n(n−1)/2* pairwise comparisons are not independent —
  the null model permutation test accounts for this partially.
- **Unit harmonisation**: DDT and PCB values retain units from the modal unit per
  analyte group; cross-study mismatches would inflate apparent dissimilarity.
- **Missing predictive covariates**: oceanographic variables (SST, chlorophyll,
  bathymetry), diet, age, and sex are not included but are essential for fuller
  mechanistic models.

---

## 14) Predictive Modeling Extensions

### What Does a Contamination Modeler Need?

A researcher interested in predicting contamination at unsampled sites or
future time points would typically seek:

1. **Spatial autocorrelation structure** (variogram / distance-decay) —
   *done above* — defines the range at which samples are spatially informative.
2. **Kriging or spatial interpolation** — predicts concentration at unsampled
   coordinates weighted by the empirical variogram.
3. **GAM / GLMM spatial prediction** — models concentration as a smooth
   function of latitude, longitude, species, and year; allows extrapolation.
4. **Spatial cross-validation** — leave-one-area-out (LOAO) or block holdout CV
   gives unbiased prediction error; standard k-fold CV is anti-conservative for
   spatially autocorrelated data.
5. **Temporal trend modeling** — tests whether concentrations are
   declining (regulation effect) or increasing (new sources).
6. **Environmental covariate linkage** — SST, chlorophyll, bathymetry as
   predictors for transfer to climate scenarios.

Sections 14A–14C implement items 3, 4, and 5 using the individual-level ZD
Stamp sample data.

```{r pred-load-realdat, include=FALSE}
if (!exists("real_dat")) {
  fp <- file.path(out_dir, "real_testcase_prepared_data.csv")
  if (file.exists(fp)) real_dat <- read.csv(fp, stringsAsFactors = FALSE)
}
```

### 14A. GAM Spatial Prediction: DDT Concentration Surface

A generalised additive model (GAM) with a 2-D thin-plate spline over longitude
and latitude is fitted to log(DDT + 1). Species is included as a random effect
to absorb inter-species baseline differences. The model is then projected onto a
prediction grid to map the estimated DDT surface and its uncertainty.

```{r gam-spatial, fig.width=11, fig.height=5, results='asis', message=FALSE}
if (!exists("real_dat") || nrow(real_dat) < 20) {
  cat("real_dat not available or too small for spatial GAM.\n")
} else {
  if (!requireNamespace("mgcv", quietly = TRUE))
    install.packages("mgcv", repos = "https://cran.r-project.org/")
  library(mgcv)

  gdat <- real_dat[
    is.finite(real_dat$DDT_real) &
    is.finite(real_dat$LATITUDE) &
    is.finite(real_dat$LONGITUDE), ]
  gdat$logDDT  <- log1p(gdat$DDT_real)
  gdat$SPECIES <- as.factor(gdat$SPECIES)
  k_sp <- min(20, nrow(gdat) - 5)

  gm <- tryCatch({
    if (length(levels(gdat$SPECIES)) > 1) {
      gam(logDDT ~ s(LONGITUDE, LATITUDE, k = k_sp) + s(SPECIES, bs = "re"),
          data = gdat, method = "REML")
    } else {
      gam(logDDT ~ s(LONGITUDE, LATITUDE, k = k_sp),
          data = gdat, method = "REML")
    }
  }, error = function(e) { message("GAM failed: ", e$message); NULL })

  if (!is.null(gm)) {
    gm_summ <- summary(gm)
    cat(sprintf(
      "**GAM deviance explained: %.1f%%**  |  R\u00b2 (adj.) = %.3f  |  n = %d\n\n",
      gm_summ$dev.expl * 100, gm_summ$r.sq, nrow(gdat)))

    lon_seq <- seq(min(gdat$LONGITUDE) - 0.5, max(gdat$LONGITUDE) + 0.5, length.out = 60)
    lat_seq <- seq(min(gdat$LATITUDE)  - 0.5, max(gdat$LATITUDE)  + 0.5, length.out = 60)
    grid_df <- expand.grid(LONGITUDE = lon_seq, LATITUDE = lat_seq)
    modal_sp <- names(sort(table(as.character(gdat$SPECIES)), decreasing = TRUE))[1]
    grid_df$SPECIES <- factor(modal_sp, levels = levels(gdat$SPECIES))

    pred_out <- tryCatch(
      predict(gm, newdata = grid_df, type = "response", se.fit = TRUE),
      error = function(e) NULL)

    if (!is.null(pred_out)) {
      grid_df$pred_logDDT <- pred_out$fit
      grid_df$se_logDDT   <- pred_out$se.fit
      write.csv(grid_df, file.path(out_dir, "gam_predicted_ddt_surface.csv"),
                row.names = FALSE)

      par(mfrow = c(1, 2), mar = c(4, 4, 3, 2))

      mat_pred <- matrix(grid_df$pred_logDDT, nrow = length(lon_seq))
      image(lon_seq, lat_seq, mat_pred,
            col  = hcl.colors(64, "YlOrRd", rev = FALSE),
            xlab = "Longitude", ylab = "Latitude",
            main = sprintf("Predicted log(DDT+1)\n(%s, GAM spatial smooth)", modal_sp))
      points(gdat$LONGITUDE, gdat$LATITUDE, pch = 21, cex = 0.8,
             bg = "white", col = "black", lwd = 0.8)
      contour(lon_seq, lat_seq, mat_pred, add = TRUE,
              col = grey(0.3), lwd = 0.5, nlevels = 6)

      mat_se <- matrix(grid_df$se_logDDT, nrow = length(lon_seq))
      image(lon_seq, lat_seq, mat_se,
            col  = hcl.colors(64, "Blues 3", rev = TRUE),
            xlab = "Longitude", ylab = "Latitude",
            main = "Prediction uncertainty (SE)\nHigher = fewer data nearby")
      points(gdat$LONGITUDE, gdat$LATITUDE, pch = 21, cex = 0.8,
             bg = "white", col = "black", lwd = 0.8)
    }
  }
}
```

### 14B. Temporal Trend Analysis

```{r temporal-trend, fig.width=11, fig.height=5, results='asis'}
if (!exists("real_dat")) {
  cat("real_dat not available.\n")
} else {
  tdat <- real_dat[
    is.finite(real_dat$DDT_real) &
    is.finite(real_dat$PCP_or_PCB) &
    is.finite(real_dat$COLLECTION_YEAR), ]

  if (nrow(tdat) < 10 || length(unique(tdat$COLLECTION_YEAR)) < 3) {
    cat("Insufficient temporal variation for trend analysis.\n")
  } else {
    tdat$logDDT <- log1p(tdat$DDT_real)
    tdat$logPCP <- log1p(tdat$PCP_or_PCB)

    lm_ddt <- summary(lm(logDDT ~ COLLECTION_YEAR, data = tdat))
    lm_pcp <- summary(lm(logPCP ~ COLLECTION_YEAR, data = tdat))

    par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

    plot(tdat$COLLECTION_YEAR, tdat$logDDT,
         pch = 16, cex = 0.55, col = adjustcolor("firebrick", 0.45),
         xlab = "Collection year", ylab = "log(DDT + 1)",
         main = "DDT: Temporal Trend")
    abline(lm(logDDT ~ COLLECTION_YEAR, data = tdat), col = "firebrick", lwd = 2.5)
    legend("topright", bty = "n", cex = 0.8,
           legend = c(sprintf("slope = %.4f/yr", lm_ddt$coefficients[2, 1]),
                      sprintf("p = %.3f",         lm_ddt$coefficients[2, 4]),
                      sprintf("R\u00b2 = %.3f",     lm_ddt$r.squared)))

    plot(tdat$COLLECTION_YEAR, tdat$logPCP,
         pch = 16, cex = 0.55, col = adjustcolor("darkorange3", 0.45),
         xlab = "Collection year", ylab = "log(PCP/PCB + 1)",
         main = "PCP/PCB: Temporal Trend")
    abline(lm(logPCP ~ COLLECTION_YEAR, data = tdat), col = "darkorange3", lwd = 2.5)
    legend("topright", bty = "n", cex = 0.8,
           legend = c(sprintf("slope = %.4f/yr", lm_pcp$coefficients[2, 1]),
                      sprintf("p = %.3f",         lm_pcp$coefficients[2, 4]),
                      sprintf("R\u00b2 = %.3f",     lm_pcp$r.squared)))

    cat(sprintf("- **DDT**: slope = %.4f log-units/yr, R\u00b2 = %.3f, p = %.3f\n",
                lm_ddt$coefficients[2, 1], lm_ddt$r.squared, lm_ddt$coefficients[2, 4]))
    cat(sprintf("- **PCP/PCB**: slope = %.4f log-units/yr, R\u00b2 = %.3f, p = %.3f\n",
                lm_pcp$coefficients[2, 1], lm_pcp$r.squared, lm_pcp$coefficients[2, 4]))
  }
}
```

### 14C. Spatial Cross-Validation (Leave-One-Area-Out, GAM DDT)

Each geographic area is withheld in turn; a GAM is fit on the remaining areas and
used to predict log(DDT+1) in the withheld area. RMSE and MAE are on the
log(DDT+1) scale.

```{r spatial-cv, results='asis', message=FALSE}
if (!exists("real_dat") || nrow(real_dat) < 20) {
  cat("real_dat not available.\n")
} else {
  if (!requireNamespace("mgcv", quietly = TRUE)) library(mgcv)

  gdat_cv <- real_dat[
    is.finite(real_dat$DDT_real) &
    is.finite(real_dat$LATITUDE) &
    is.finite(real_dat$LONGITUDE), ]
  gdat_cv$logDDT <- log1p(gdat_cv$DDT_real)

  areas_cv <- unique(gdat_cv$GEOGRAPHIC_AREA)
  areas_cv <- areas_cv[!is.na(areas_cv) & nchar(trimws(areas_cv)) > 0]

  if (length(areas_cv) < 3) {
    cat("Fewer than 3 geographic areas — skipping leave-one-area-out CV.\n")
  } else {
    cv_rows <- lapply(areas_cv, function(a) {
      train <- gdat_cv[gdat_cv$GEOGRAPHIC_AREA != a, ]
      test  <- gdat_cv[gdat_cv$GEOGRAPHIC_AREA == a, ]
      if (nrow(train) < 15 || nrow(test) < 2) return(NULL)
      k_cv <- min(15, nrow(train) - 5)
      m_cv <- tryCatch(
        gam(logDDT ~ s(LONGITUDE, LATITUDE, k = k_cv),
            data = train, method = "REML"),
        error = function(e) NULL)
      if (is.null(m_cv)) return(NULL)
      pred_cv <- tryCatch(
        predict(m_cv, newdata = test, type = "response"),
        error = function(e) NULL)
      if (is.null(pred_cv)) return(NULL)
      data.frame(
        Area   = a,
        N_test = nrow(test),
        RMSE   = round(sqrt(mean((test$logDDT - pred_cv)^2)), 3),
        MAE    = round(mean(abs(test$logDDT - pred_cv)), 3),
        stringsAsFactors = FALSE)
    })

    cv_tab <- do.call(rbind, Filter(Negate(is.null), cv_rows))

    if (!is.null(cv_tab) && nrow(cv_tab) > 0) {
      mean_row <- data.frame(
        Area   = "**Mean across areas**", N_test = NA,
        RMSE   = round(mean(cv_tab$RMSE), 3),
        MAE    = round(mean(cv_tab$MAE),  3),
        stringsAsFactors = FALSE)
      cv_out <- rbind(cv_tab, mean_row)
      write.csv(cv_out, file.path(out_dir, "spatial_cv_loao_results.csv"),
                row.names = FALSE)
      print(knitr::kable(cv_out, row.names = FALSE,
        col.names = c("Held-out area", "N test",
                      "RMSE (log-scale)", "MAE (log-scale)"),
        caption = paste(
          "Table 3. Leave-one-area-out spatial cross-validation (DDT, GAM).",
          "Each geographic area is withheld in turn;",
          "the GAM is fit on the remaining areas and predictions are evaluated",
          "on withheld samples. Errors on log(DDT+1) scale.")))
    }
  }
}
```

### 14D. Recommended Next Steps for Predictive Modeling

| Priority | Method | Purpose | R package |
|---|---|---|---|
| 1 | Empirical variogram | Quantify spatial autocorrelation range | `gstat`, `sf` |
| 2 | Ordinary kriging | Unbiased interpolation to unsampled sites | `gstat` |
| 3 | GAM + oceanographic covariates | Mechanistic spatial prediction | `mgcv` |
| 4 | Block spatial CV | Honest prediction error for autocorrelated data | `blockCV` |
| 5 | Random forest + spatial CV | Non-linear covariate effects | `ranger` + `blockCV` |
| 6 | Mixed model: species + year random effects | Partition spatial vs. temporal vs. species variance | `lme4` / `glmmTMB` |
| 7 | Temporal ARMA on annual means | Trend and cycle detection | `forecast` |
| 8 | Transfer to unmonitored taxa | Exposure prediction for new species/regions | GAM with species factor |

The distance-decay slopes estimated in this report directly parameterise kriging
variograms: for the exponential model the spatial range
*r* = 1/*b* (where *b* is the fitted decay coefficient). For DDT,
*b* ≈ 2.54 × 10⁻⁴ km⁻¹ gives *r* ≈ 3,937 km, indicating that samples within
roughly 4,000 km carry meaningful shared spatial information and are suitable
co-variates for kriging.

## 15) Spatiotemporal Scaling Model (Distance + Time)

This section fits a joint scaling model for contaminant similarity as a function
of both geographic distance and collection-year separation:

$$
	ext{sim}_{ij} = \beta_0 + \beta_1\log(d_{ij}+1) + \beta_2\log(\Delta t_{ij}+1) + \beta_3\log(d_{ij}+1)\log(\Delta t_{ij}+1) + \varepsilon_{ij}
$$

where $d_{ij}$ is pairwise geographic distance (km) and $\Delta t_{ij}$ is
pairwise absolute time lag (years).

```{r spatiotemporal-scaling-model, results='asis'}
if (!exists("pair_real")) {
  cat("pair_real not available — skipping spatiotemporal scaling model.\n")
} else {
  if (!"time_lag_years" %in% names(pair_real)) {
    cat("time_lag_years missing in pair_real — rerun real-data analysis chunk first.\n")
  } else {
    resp_vars  <- c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall")
    resp_names <- c("DDT", "PCP/PCB", "Isotope", "Overall")

    fit_st <- function(df, yvar) {
      ok <- is.finite(df[[yvar]]) & is.finite(df$distance_km) & is.finite(df$time_lag_years)
      d  <- df$distance_km[ok]
      t  <- df$time_lag_years[ok]
      y  <- df[[yvar]][ok]
      if (length(y) < 30 || length(unique(t)) < 2) return(NULL)

      st_df <- data.frame(
        y = y,
        log_dist = log(d + 1),
        log_time = log(t + 1)
      )

      m_add <- lm(y ~ log_dist + log_time, data = st_df)
      m_int <- lm(y ~ log_dist * log_time, data = st_df)

      b_int <- coef(summary(m_int))
      get_coef <- function(term, col) if (term %in% rownames(b_int)) b_int[term, col] else NA_real_

      data.frame(
        response = yvar,
        n_pairs = nrow(st_df),
        dist_coef = coef(m_int)["log_dist"],
        time_coef = coef(m_int)["log_time"],
        interaction_coef = coef(m_int)["log_dist:log_time"],
        p_dist = get_coef("log_dist", "Pr(>|t|)"),
        p_time = get_coef("log_time", "Pr(>|t|)"),
        p_interaction = get_coef("log_dist:log_time", "Pr(>|t|)"),
        R2_additive = summary(m_add)$r.squared,
        R2_interaction = summary(m_int)$r.squared,
        AIC_additive = AIC(m_add),
        AIC_interaction = AIC(m_int),
        delta_AIC_int_vs_add = AIC(m_add) - AIC(m_int),
        stringsAsFactors = FALSE
      )
    }

    st_out <- do.call(rbind, lapply(resp_vars, function(v) fit_st(pair_real, v)))

    if (!is.null(st_out) && nrow(st_out) > 0) {
      st_out$response <- resp_names[match(st_out$response, resp_vars)]
      num_cols <- setdiff(names(st_out), c("response"))
      st_out[num_cols] <- lapply(st_out[num_cols], function(x) round(x, 4))
      write.csv(st_out, file.path(out_dir, "spatiotemporal_scaling_model_summary.csv"), row.names = FALSE)

      print(knitr::kable(
        st_out,
        row.names = FALSE,
        col.names = c(
          "Response", "N pairs", "Beta distance", "Beta time", "Beta distance x time",
          "p distance", "p time", "p interaction", "R2 additive", "R2 interaction",
          "AIC additive", "AIC interaction", "Delta AIC (int-add)"
        ),
        caption = paste(
          "Table 4. Spatiotemporal scaling models.",
          "Positive Delta AIC indicates interaction model support over additive model."
        )
      ))
    }
  }
}
```

```{r spatiotemporal-visualization, fig.width=13, fig.height=10}
if (exists("pair_real") && "time_lag_years" %in% names(pair_real)) {
  resp_vars  <- c("sim_DDT", "sim_PCP_or_PCB", "sim_isotope", "sim_overall")
  resp_names <- c("DDT", "PCP/PCB", "Isotope", "Overall")

  q <- quantile(pair_real$time_lag_years, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE)
  q[1] <- min(q[1], 0)
  pair_real$time_group <- cut(
    pair_real$time_lag_years,
    breaks = unique(q),
    include.lowest = TRUE,
    labels = c("Low time lag", "Mid time lag", "High time lag")[seq_len(length(unique(q)) - 1)]
  )

  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
  cols <- c("steelblue3", "darkorange3", "firebrick3")

  for (ii in seq_along(resp_vars)) {
    v <- resp_vars[ii]
    d <- pair_real$distance_km
    y <- pair_real[[v]]
    g <- pair_real$time_group
    ok <- is.finite(d) & is.finite(y) & !is.na(g)
    d <- d[ok]; y <- y[ok]; g <- droplevels(g[ok])

    plot(d, y,
         pch = 16, cex = 0.30,
         col = adjustcolor("black", alpha.f = 0.10),
         xlab = "Distance (km)", ylab = "Similarity",
         main = paste(resp_names[ii], "by time-lag tercile"))

    dseq <- seq(0.1, max(d), length.out = 300)
    levs <- levels(g)
    for (jj in seq_along(levs)) {
      idx <- g == levs[jj]
      if (sum(idx) < 20) next
      m <- lm(y[idx] ~ log(d[idx] + 1))
      a <- coef(m)[1]; b <- coef(m)[2]
      lines(dseq, a + b * log(dseq + 1), col = cols[jj], lwd = 2.5)
    }
    legend("topright", bty = "n", cex = 0.75,
           legend = levels(g), col = cols[seq_along(levels(g))], lwd = 2.5)
  }
}
```

## 15B) Dyadic Mixed-Effects Spatiotemporal Model + Blocked CV

To reduce pseudo-replication from pairwise expansion, this section fits a
dyadic mixed-effects model with crossed random effects for both members of
each pair (individual IDs, species, and geographic areas), then evaluates
prediction using leave-one-area-out blocked cross-validation.

```{r dyadic-mixed-model, results='asis', message=FALSE, warning=FALSE}
if (!exists("pair_real")) {
  cat("pair_real not available — skipping dyadic mixed-effects model.\n")
} else if (!all(c("time_lag_years", "id_i", "id_j", "area_i", "area_j", "species_i", "species_j") %in% names(pair_real))) {
  cat("Required pairwise fields not found — rerun real-data analysis chunk first.\n")
} else {
  if (!requireNamespace("lme4", quietly = TRUE)) {
    install.packages("lme4", repos = "https://cran.r-project.org/")
  }
  library(lme4)

  fit_dyadic <- function(df, yvar) {
    keep <- is.finite(df[[yvar]]) &
      is.finite(df$distance_km) & is.finite(df$time_lag_years) &
      !is.na(df$id_i) & !is.na(df$id_j) & !is.na(df$area_i) & !is.na(df$area_j) &
      !is.na(df$species_i) & !is.na(df$species_j)

    dat <- df[keep, c(yvar, "distance_km", "time_lag_years", "id_i", "id_j", "area_i", "area_j", "species_i", "species_j")]
    names(dat)[1] <- "y"

    if (nrow(dat) < 100) return(NULL)

    dat$log_dist <- log(dat$distance_km + 1)
    dat$log_time <- log(dat$time_lag_years + 1)

    dat$id_i <- as.factor(dat$id_i)
    dat$id_j <- as.factor(dat$id_j)
    dat$area_i <- as.factor(dat$area_i)
    dat$area_j <- as.factor(dat$area_j)
    dat$species_i <- as.factor(dat$species_i)
    dat$species_j <- as.factor(dat$species_j)

    ctrl <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))

    full_formula <- y ~ log_dist * log_time +
      (1 | id_i) + (1 | id_j) +
      (1 | species_i) + (1 | species_j) +
      (1 | area_i) + (1 | area_j)

    fallback_formula <- y ~ log_dist * log_time +
      (1 | species_i) + (1 | species_j) +
      (1 | area_i) + (1 | area_j)

    mod <- try(lmer(full_formula, data = dat, REML = FALSE, control = ctrl), silent = TRUE)
    model_type <- "full_dyadic"
    if (inherits(mod, "try-error")) {
      mod <- try(lmer(fallback_formula, data = dat, REML = FALSE, control = ctrl), silent = TRUE)
      model_type <- "fallback_species_area"
    }
    if (inherits(mod, "try-error")) return(NULL)

    sm <- summary(mod)
    cf <- as.data.frame(sm$coefficients)
    cf$term <- rownames(cf)

    get_cf <- function(term, colname) {
      if (!(term %in% cf$term)) return(NA_real_)
      cf[cf$term == term, colname][1]
    }

    fixed_row <- data.frame(
      response = yvar,
      model_type = model_type,
      n_pairs = nrow(dat),
      beta_dist = get_cf("log_dist", "Estimate"),
      beta_time = get_cf("log_time", "Estimate"),
      beta_dist_time = get_cf("log_dist:log_time", "Estimate"),
      se_dist = get_cf("log_dist", "Std. Error"),
      se_time = get_cf("log_time", "Std. Error"),
      se_dist_time = get_cf("log_dist:log_time", "Std. Error"),
      t_dist = get_cf("log_dist", "t value"),
      t_time = get_cf("log_time", "t value"),
      t_dist_time = get_cf("log_dist:log_time", "t value"),
      AIC = AIC(mod),
      stringsAsFactors = FALSE
    )

    list(model = mod, data = dat, fixed = fixed_row)
  }

  run_blocked_cv <- function(dat, model_type, yvar) {
    areas <- sort(unique(c(as.character(dat$area_i), as.character(dat$area_j))))
    areas <- areas[!is.na(areas) & nchar(trimws(areas)) > 0]
    if (length(areas) < 3) return(NULL)

    ctrl <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
    full_formula <- y ~ log_dist * log_time +
      (1 | id_i) + (1 | id_j) +
      (1 | species_i) + (1 | species_j) +
      (1 | area_i) + (1 | area_j)
    fallback_formula <- y ~ log_dist * log_time +
      (1 | species_i) + (1 | species_j) +
      (1 | area_i) + (1 | area_j)

    cv_rows <- lapply(areas, function(a) {
      test_idx <- as.character(dat$area_i) == a | as.character(dat$area_j) == a
      train <- dat[!test_idx, ]
      test <- dat[test_idx, ]
      if (nrow(train) < 100 || nrow(test) < 20) return(NULL)

      form <- if (model_type == "full_dyadic") full_formula else fallback_formula
      fit <- try(lmer(form, data = train, REML = FALSE, control = ctrl), silent = TRUE)
      if (inherits(fit, "try-error")) {
        fit <- try(lmer(fallback_formula, data = train, REML = FALSE, control = ctrl), silent = TRUE)
      }
      if (inherits(fit, "try-error")) return(NULL)

      pred <- try(predict(fit, newdata = test, allow.new.levels = TRUE), silent = TRUE)
      if (inherits(pred, "try-error")) return(NULL)

      obs <- test$y
      rmse <- sqrt(mean((obs - pred)^2, na.rm = TRUE))
      mae <- mean(abs(obs - pred), na.rm = TRUE)
      r2 <- if (length(unique(obs)) > 1) cor(obs, pred, use = "complete.obs")^2 else NA_real_

      data.frame(
        response = yvar,
        held_out_area = a,
        n_test = nrow(test),
        RMSE = rmse,
        MAE = mae,
        R2_pred = r2,
        stringsAsFactors = FALSE
      )
    })

    cv_tab <- do.call(rbind, Filter(Negate(is.null), cv_rows))
    if (is.null(cv_tab) || nrow(cv_tab) == 0) return(NULL)
    mean_row <- data.frame(
      response = yvar,
      held_out_area = "Mean across held-out areas",
      n_test = NA,
      RMSE = mean(cv_tab$RMSE, na.rm = TRUE),
      MAE = mean(cv_tab$MAE, na.rm = TRUE),
      R2_pred = mean(cv_tab$R2_pred, na.rm = TRUE),
      stringsAsFactors = FALSE
    )
    rbind(cv_tab, mean_row)
  }

  targets <- c("sim_overall", "sim_DDT")
  fit_objs <- lapply(targets, function(v) fit_dyadic(pair_real, v))
  names(fit_objs) <- targets
  fit_objs <- fit_objs[!vapply(fit_objs, is.null, logical(1))]

  if (length(fit_objs) == 0) {
    cat("Dyadic mixed-effects models failed or insufficient data.\n")
  } else {
    fixed_tab <- do.call(rbind, lapply(fit_objs, function(x) x$fixed))
    fixed_tab$response <- c(sim_overall = "Overall", sim_DDT = "DDT")[fixed_tab$response]

    num_cols <- setdiff(names(fixed_tab), c("response", "model_type"))
    fixed_tab[num_cols] <- lapply(fixed_tab[num_cols], function(x) round(x, 4))

    write.csv(fixed_tab, file.path(out_dir, "dyadic_mixed_model_fixed_effects.csv"), row.names = FALSE)

    print(knitr::kable(
      fixed_tab,
      row.names = FALSE,
      col.names = c(
        "Response", "Model", "N pairs",
        "Beta distance", "Beta time", "Beta distance x time",
        "SE dist", "SE time", "SE dist x time",
        "t dist", "t time", "t dist x time", "AIC"
      ),
      caption = paste(
        "Table 5. Dyadic mixed-effects spatiotemporal model fixed effects.",
        "Crossed random effects account for pair-member, species, and area dependence."
      )
    ))

    cv_tabs <- do.call(rbind, lapply(names(fit_objs), function(v) {
      out <- run_blocked_cv(fit_objs[[v]]$data, fit_objs[[v]]$fixed$model_type[1], v)
      out
    }))

    if (!is.null(cv_tabs) && nrow(cv_tabs) > 0) {
      cv_tabs$response <- c(sim_overall = "Overall", sim_DDT = "DDT")[cv_tabs$response]
      num_cols <- c("n_test", "RMSE", "MAE", "R2_pred")
      for (nc in num_cols) {
        if (nc %in% names(cv_tabs)) cv_tabs[[nc]] <- round(cv_tabs[[nc]], 4)
      }

      write.csv(cv_tabs, file.path(out_dir, "dyadic_mixed_model_blocked_cv.csv"), row.names = FALSE)

      print(knitr::kable(
        cv_tabs,
        row.names = FALSE,
        col.names = c("Response", "Held-out area", "N test", "RMSE", "MAE", "Predictive R2"),
        caption = paste(
          "Table 6. Leave-one-area-out blocked cross-validation for dyadic mixed models.",
          "Test rows include any pair touching the held-out area, reducing spatial leakage."
        )
      ))
    }
  }
}
```

```{r dyadic-mixed-graphics, fig.width=12, fig.height=9, message=FALSE, warning=FALSE}
if (exists("fit_objs") && length(fit_objs) > 0) {
  par(mfrow = c(2, 2), mar = c(7, 4, 3, 1))

  for (nm in names(fit_objs)) {
    fit_obj <- fit_objs[[nm]]
    cf <- as.data.frame(summary(fit_obj$model)$coefficients)
    cf$term <- rownames(cf)
    keep <- cf$term %in% c("log_dist", "log_time", "log_dist:log_time")
    cf <- cf[keep, , drop = FALSE]
    labs <- c(log_dist = "Distance", log_time = "Time lag", `log_dist:log_time` = "Distance x Time")
    xpos <- seq_len(nrow(cf))
    ylim_r <- range(c(cf$Estimate - 1.96 * cf$`Std. Error`, cf$Estimate + 1.96 * cf$`Std. Error`, 0))

    plot(xpos, cf$Estimate,
         ylim = ylim_r,
         xlim = c(0.5, nrow(cf) + 0.5),
         xaxt = "n",
         xlab = "",
         ylab = "Coefficient estimate",
         main = paste(c(sim_overall = "Overall", sim_DDT = "DDT")[nm], "mixed model"),
         pch = 19,
         col = if (nm == "sim_overall") "darkorchid4" else "firebrick")
    axis(1, at = xpos, labels = labs[cf$term], las = 2)
    abline(h = 0, lty = 2, col = "grey50")
    segments(xpos, cf$Estimate - 1.96 * cf$`Std. Error`,
             xpos, cf$Estimate + 1.96 * cf$`Std. Error`,
             lwd = 2,
             col = if (nm == "sim_overall") "darkorchid4" else "firebrick")
  }

  if (exists("cv_tabs") && !is.null(cv_tabs) && nrow(cv_tabs) > 0) {
    cv_plot <- cv_tabs[cv_tabs$held_out_area != "Mean across held-out areas", ]
    areas <- unique(cv_plot$held_out_area)
    responses <- unique(cv_plot$response)
    rmse_mat <- sapply(responses, function(resp) cv_plot$RMSE[match(areas, cv_plot$held_out_area[cv_plot$response == resp])])
    r2_mat <- sapply(responses, function(resp) cv_plot$R2_pred[match(areas, cv_plot$held_out_area[cv_plot$response == resp])])
    if (is.null(dim(rmse_mat))) rmse_mat <- matrix(rmse_mat, ncol = 1, dimnames = list(areas, responses))
    if (is.null(dim(r2_mat))) r2_mat <- matrix(r2_mat, ncol = 1, dimnames = list(areas, responses))

    barplot(t(rmse_mat), beside = TRUE,
            col = c("darkorchid4", "firebrick")[seq_len(ncol(rmse_mat))],
            names.arg = areas,
            las = 2,
            ylab = "Blocked CV RMSE",
            main = "Dyadic mixed model blocked CV")
    legend("topright", bty = "n",
           legend = c(sim_overall = "Overall", sim_DDT = "DDT")[colnames(rmse_mat)],
           fill = c("darkorchid4", "firebrick")[seq_len(ncol(rmse_mat))])

    matplot(seq_along(areas), r2_mat,
            type = "b", pch = 19, lty = 1,
            col = c("darkorchid4", "firebrick")[seq_len(ncol(r2_mat))],
            xaxt = "n",
            xlab = "Held-out area",
            ylab = expression(R^2),
            main = "Dyadic mixed model predictive fit")
    axis(1, at = seq_along(areas), labels = areas, las = 2)
    legend("topright", bty = "n",
           legend = c(sim_overall = "Overall", sim_DDT = "DDT")[colnames(r2_mat)],
           col = c("darkorchid4", "firebrick")[seq_len(ncol(r2_mat))],
           lty = 1, pch = 19)
  } else {
    plot.new(); title("Blocked CV RMSE")
    plot.new(); title("Blocked CV predictive fit")
  }
}
```

## 15C) Trophic Covariate Extension for Transferability

We do not yet have tracking-derived movement covariates in this dataset, so the
highest-value biodiversity extension currently available is a **trophic proxy**
model. Here, pairwise differences in $\delta^{15}N$ and $\delta^{13}C$ are used
as proxies for trophic-position mismatch and habitat/foraging mismatch,
respectively. These covariates test whether Gulf-of-Alaska transfer improves
when ecological similarity is modeled alongside space and time.

```{r dyadic-trophic-extension, results='asis', message=FALSE, warning=FALSE}
if (!exists("pair_real")) {
  cat("pair_real not available — skipping trophic extension.\n")
} else {
  if (!exists("real_dat")) {
    fp <- file.path(out_dir, "real_testcase_prepared_data.csv")
    if (file.exists(fp)) real_dat <- read.csv(fp, stringsAsFactors = FALSE)
  }

  if (!exists("real_dat") || !all(c("GUALIQUOTID", "delta13C_real", "delta15N_real") %in% names(real_dat))) {
    cat("real_dat with isotope fields not available — skipping trophic extension.\n")
  } else {
    iso_map <- real_dat[, c("GUALIQUOTID", "delta13C_real", "delta15N_real")]
    iso_i <- iso_map
    names(iso_i) <- c("id_i", "delta13C_i", "delta15N_i")
    iso_j <- iso_map
    names(iso_j) <- c("id_j", "delta13C_j", "delta15N_j")

    pair_trophic <- merge(pair_real, iso_i, by = "id_i", all.x = TRUE)
    pair_trophic <- merge(pair_trophic, iso_j, by = "id_j", all.x = TRUE)

    pair_trophic$trophic_diff_n15 <- abs(pair_trophic$delta15N_i - pair_trophic$delta15N_j)
    pair_trophic$habitat_diff_c13 <- abs(pair_trophic$delta13C_i - pair_trophic$delta13C_j)
    pair_trophic$mean_trophic_n15 <- rowMeans(cbind(pair_trophic$delta15N_i, pair_trophic$delta15N_j), na.rm = TRUE)
    pair_trophic$mean_habitat_c13 <- rowMeans(cbind(pair_trophic$delta13C_i, pair_trophic$delta13C_j), na.rm = TRUE)

    fit_dyadic_trophic <- function(df, yvar) {
      keep <- is.finite(df[[yvar]]) &
        is.finite(df$distance_km) & is.finite(df$time_lag_years) &
        is.finite(df$trophic_diff_n15) & is.finite(df$habitat_diff_c13) &
        is.finite(df$mean_trophic_n15) & is.finite(df$mean_habitat_c13) &
        !is.na(df$id_i) & !is.na(df$id_j) & !is.na(df$area_i) & !is.na(df$area_j) &
        !is.na(df$species_i) & !is.na(df$species_j)

      dat <- df[keep, c(
        yvar, "distance_km", "time_lag_years",
        "trophic_diff_n15", "habitat_diff_c13", "mean_trophic_n15", "mean_habitat_c13",
        "id_i", "id_j", "area_i", "area_j", "species_i", "species_j"
      )]
      names(dat)[1] <- "y"
      if (nrow(dat) < 100) return(NULL)

      dat$log_dist <- log(dat$distance_km + 1)
      dat$log_time <- log(dat$time_lag_years + 1)
      dat$id_i <- as.factor(dat$id_i)
      dat$id_j <- as.factor(dat$id_j)
      dat$area_i <- as.factor(dat$area_i)
      dat$area_j <- as.factor(dat$area_j)
      dat$species_i <- as.factor(dat$species_i)
      dat$species_j <- as.factor(dat$species_j)

      ctrl <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))

      form_full <- y ~ log_dist * log_time + trophic_diff_n15 + habitat_diff_c13 +
        mean_trophic_n15 + mean_habitat_c13 +
        (1 | id_i) + (1 | id_j) +
        (1 | species_i) + (1 | species_j) +
        (1 | area_i) + (1 | area_j)

      form_fallback <- y ~ log_dist * log_time + trophic_diff_n15 + habitat_diff_c13 +
        mean_trophic_n15 + mean_habitat_c13 +
        (1 | species_i) + (1 | species_j) +
        (1 | area_i) + (1 | area_j)

      mod <- try(lmer(form_full, data = dat, REML = FALSE, control = ctrl), silent = TRUE)
      model_type <- "trophic_full_dyadic"
      if (inherits(mod, "try-error")) {
        mod <- try(lmer(form_fallback, data = dat, REML = FALSE, control = ctrl), silent = TRUE)
        model_type <- "trophic_fallback_species_area"
      }
      if (inherits(mod, "try-error")) return(NULL)

      sm <- summary(mod)
      cf <- as.data.frame(sm$coefficients)
      cf$term <- rownames(cf)

      get_cf <- function(term, colname) {
        if (!(term %in% cf$term)) return(NA_real_)
        cf[cf$term == term, colname][1]
      }

      fixed_row <- data.frame(
        response = yvar,
        model_type = model_type,
        n_pairs = nrow(dat),
        beta_dist = get_cf("log_dist", "Estimate"),
        beta_time = get_cf("log_time", "Estimate"),
        beta_dist_time = get_cf("log_dist:log_time", "Estimate"),
        beta_trophic_diff = get_cf("trophic_diff_n15", "Estimate"),
        beta_habitat_diff = get_cf("habitat_diff_c13", "Estimate"),
        beta_mean_trophic = get_cf("mean_trophic_n15", "Estimate"),
        beta_mean_habitat = get_cf("mean_habitat_c13", "Estimate"),
        AIC = AIC(mod),
        stringsAsFactors = FALSE
      )

      list(model = mod, data = dat, fixed = fixed_row)
    }

    run_blocked_cv_trophic <- function(dat, model_type, yvar) {
      areas <- sort(unique(c(as.character(dat$area_i), as.character(dat$area_j))))
      areas <- areas[!is.na(areas) & nchar(trimws(areas)) > 0]
      if (length(areas) < 3) return(NULL)

      ctrl <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
      form_full <- y ~ log_dist * log_time + trophic_diff_n15 + habitat_diff_c13 +
        mean_trophic_n15 + mean_habitat_c13 +
        (1 | id_i) + (1 | id_j) +
        (1 | species_i) + (1 | species_j) +
        (1 | area_i) + (1 | area_j)
      form_fallback <- y ~ log_dist * log_time + trophic_diff_n15 + habitat_diff_c13 +
        mean_trophic_n15 + mean_habitat_c13 +
        (1 | species_i) + (1 | species_j) +
        (1 | area_i) + (1 | area_j)

      cv_rows <- lapply(areas, function(a) {
        test_idx <- as.character(dat$area_i) == a | as.character(dat$area_j) == a
        train <- dat[!test_idx, ]
        test <- dat[test_idx, ]
        if (nrow(train) < 100 || nrow(test) < 20) return(NULL)

        form <- if (model_type == "trophic_full_dyadic") form_full else form_fallback
        fit <- try(lmer(form, data = train, REML = FALSE, control = ctrl), silent = TRUE)
        if (inherits(fit, "try-error")) {
          fit <- try(lmer(form_fallback, data = train, REML = FALSE, control = ctrl), silent = TRUE)
        }
        if (inherits(fit, "try-error")) return(NULL)

        pred <- try(predict(fit, newdata = test, allow.new.levels = TRUE), silent = TRUE)
        if (inherits(pred, "try-error")) return(NULL)

        obs <- test$y
        data.frame(
          response = yvar,
          held_out_area = a,
          n_test = nrow(test),
          RMSE = sqrt(mean((obs - pred)^2, na.rm = TRUE)),
          MAE = mean(abs(obs - pred), na.rm = TRUE),
          R2_pred = if (length(unique(obs)) > 1) cor(obs, pred, use = "complete.obs")^2 else NA_real_,
          stringsAsFactors = FALSE
        )
      })

      cv_tab <- do.call(rbind, Filter(Negate(is.null), cv_rows))
      if (is.null(cv_tab) || nrow(cv_tab) == 0) return(NULL)
      mean_row <- data.frame(
        response = yvar,
        held_out_area = "Mean across held-out areas",
        n_test = NA,
        RMSE = mean(cv_tab$RMSE, na.rm = TRUE),
        MAE = mean(cv_tab$MAE, na.rm = TRUE),
        R2_pred = mean(cv_tab$R2_pred, na.rm = TRUE),
        stringsAsFactors = FALSE
      )
      rbind(cv_tab, mean_row)
    }

    targets <- c("sim_overall", "sim_DDT")
    trophic_objs <- lapply(targets, function(v) fit_dyadic_trophic(pair_trophic, v))
    names(trophic_objs) <- targets
    trophic_objs <- trophic_objs[!vapply(trophic_objs, is.null, logical(1))]

    if (length(trophic_objs) == 0) {
      cat("Trophic mixed-effects models failed or insufficient data.\n")
    } else {
      trophic_fixed <- do.call(rbind, lapply(trophic_objs, function(x) x$fixed))
      trophic_fixed$response <- c(sim_overall = "Overall", sim_DDT = "DDT")[trophic_fixed$response]
      num_cols <- setdiff(names(trophic_fixed), c("response", "model_type"))
      trophic_fixed[num_cols] <- lapply(trophic_fixed[num_cols], function(x) round(x, 4))
      write.csv(trophic_fixed, file.path(out_dir, "dyadic_trophic_model_fixed_effects.csv"), row.names = FALSE)

      print(knitr::kable(
        trophic_fixed,
        row.names = FALSE,
        col.names = c(
          "Response", "Model", "N pairs", "Beta distance", "Beta time", "Beta distance x time",
          "Beta trophic diff", "Beta habitat diff", "Beta mean trophic", "Beta mean habitat", "AIC"
        ),
        caption = paste(
          "Table 7. Trophic-augmented dyadic mixed-effects models.",
          "New biodiversity covariates are isotope-derived trophic and habitat proxies."
        )
      ))

      trophic_cv <- do.call(rbind, lapply(names(trophic_objs), function(v) {
        run_blocked_cv_trophic(trophic_objs[[v]]$data, trophic_objs[[v]]$fixed$model_type[1], v)
      }))

      if (!is.null(trophic_cv) && nrow(trophic_cv) > 0) {
        trophic_cv$response <- c(sim_overall = "Overall", sim_DDT = "DDT")[trophic_cv$response]
        num_cols <- c("n_test", "RMSE", "MAE", "R2_pred")
        for (nc in num_cols) {
          if (nc %in% names(trophic_cv)) trophic_cv[[nc]] <- round(trophic_cv[[nc]], 4)
        }
        write.csv(trophic_cv, file.path(out_dir, "dyadic_trophic_model_blocked_cv.csv"), row.names = FALSE)

        base_cv_fp <- file.path(out_dir, "dyadic_mixed_model_blocked_cv.csv")
        if (file.exists(base_cv_fp)) {
          base_cv <- read.csv(base_cv_fp, stringsAsFactors = FALSE)
          base_cv$model <- "Baseline mixed model"
          trophic_cv$model <- "Trophic mixed model"

          compare_cv <- rbind(base_cv, trophic_cv)
          keep_rows <- compare_cv$held_out_area %in% c("Gulf of Alaska", "Mean across held-out areas")
          compare_cv <- compare_cv[keep_rows, c("response", "model", "held_out_area", "RMSE", "MAE", "R2_pred")]
          names(compare_cv) <- c("Response", "Model", "Held-out area", "RMSE", "MAE", "Predictive_R2")
          write.csv(compare_cv, file.path(out_dir, "dyadic_trophic_transfer_comparison.csv"), row.names = FALSE)

          print(knitr::kable(
            compare_cv,
            row.names = FALSE,
            col.names = c("Response", "Model", "Held-out area", "RMSE", "MAE", "Predictive R2"),
            caption = paste(
              "Table 8. Transfer-performance comparison for the baseline versus trophic-augmented mixed models.",
              "Rows highlight Gulf of Alaska and the mean across held-out areas."
            )
          ))

          gulf_compare <- compare_cv[compare_cv$`Held-out area` == "Gulf of Alaska", ]
          if (nrow(gulf_compare) > 0) {
            cat("\nGulf of Alaska remains the hardest transfer domain. The trophic extension should be interpreted as a first biodiversity-aware stress test, not a full ecological solution; meaningful further gains will likely require explicit movement/connectivity covariates or oceanographic predictors.\n")
          }
        }
      }
    }
  }
}
```

```{r dyadic-trophic-graphics, fig.width=12, fig.height=9, message=FALSE, warning=FALSE}
if (exists("trophic_objs") && length(trophic_objs) > 0) {
  par(mfrow = c(2, 2), mar = c(8, 4, 3, 1))

  for (nm in names(trophic_objs)) {
    fit_obj <- trophic_objs[[nm]]
    cf <- fixef(fit_obj$model)
    keep_terms <- c("log_dist", "log_time", "log_dist:log_time", "trophic_diff_n15", "habitat_diff_c13")
    cf <- cf[names(cf) %in% keep_terms]
    labs <- c(
      log_dist = "Distance",
      log_time = "Time lag",
      `log_dist:log_time` = "Distance x Time",
      trophic_diff_n15 = expression(Delta*delta^{15}*N),
      habitat_diff_c13 = expression(Delta*delta^{13}*C)
    )
    cols <- ifelse(cf >= 0, "steelblue4", "firebrick")
    bp <- barplot(cf,
                  col = cols,
                  las = 2,
                  ylab = "Coefficient estimate",
                  main = paste(c(sim_overall = "Overall", sim_DDT = "DDT")[nm], "trophic model"),
                  names.arg = rep("", length(cf)))
    axis(1, at = bp, labels = names(cf), las = 2)
    abline(h = 0, lty = 2, col = "grey50")
  }

  if (exists("compare_cv") && !is.null(compare_cv) && nrow(compare_cv) > 0) {
    gulf <- compare_cv[compare_cv$`Held-out area` == "Gulf of Alaska", ]
    mean_rows <- compare_cv[compare_cv$`Held-out area` == "Mean across held-out areas", ]

    plot_df <- rbind(gulf, mean_rows)
    plot_df$group <- paste(plot_df$Response, plot_df$`Held-out area`, sep = "\n")
    groups <- unique(plot_df$group)
    models <- unique(plot_df$Model)
    rmse_mat <- sapply(models, function(model) plot_df$RMSE[match(groups, plot_df$group[plot_df$Model == model])])
    r2_mat <- sapply(models, function(model) plot_df$Predictive_R2[match(groups, plot_df$group[plot_df$Model == model])])
    if (is.null(dim(rmse_mat))) rmse_mat <- matrix(rmse_mat, ncol = 1, dimnames = list(groups, models))
    if (is.null(dim(r2_mat))) r2_mat <- matrix(r2_mat, ncol = 1, dimnames = list(groups, models))

    barplot(t(rmse_mat), beside = TRUE,
            col = c("grey55", "darkgreen")[seq_len(ncol(rmse_mat))],
            names.arg = groups,
            las = 2,
            ylab = "RMSE",
            main = "Baseline vs trophic model transfer RMSE")
    legend("topright", bty = "n", fill = c("grey55", "darkgreen")[seq_len(ncol(rmse_mat))], legend = colnames(rmse_mat))

    barplot(t(r2_mat), beside = TRUE,
            col = c("grey55", "darkgreen")[seq_len(ncol(r2_mat))],
            names.arg = groups,
            las = 2,
            ylab = expression(R^2),
            main = "Baseline vs trophic model transfer predictive fit")
    legend("topleft", bty = "n", fill = c("grey55", "darkgreen")[seq_len(ncol(r2_mat))], legend = colnames(r2_mat))
  } else {
    plot.new(); title("Transfer RMSE comparison")
    plot.new(); title("Transfer predictive fit comparison")
  }
}
```

## 16) Conclusion: Main Drivers of Contamination Variation

```{r conclusion-driver-values, include=FALSE}
driver_table <- NULL
if (exists("st_out") && !is.null(st_out) && nrow(st_out) > 0) {
  driver_table <- st_out
}

space_rank <- time_rank <- interaction_rank <- NA_character_
if (!is.null(driver_table)) {
  ord_space <- order(abs(driver_table$dist_coef), decreasing = TRUE)
  ord_time  <- order(abs(driver_table$time_coef), decreasing = TRUE)
  ord_int   <- order(abs(driver_table$interaction_coef), decreasing = TRUE)
  space_rank <- paste(driver_table$response[ord_space], collapse = " > ")
  time_rank <- paste(driver_table$response[ord_time], collapse = " > ")
  interaction_rank <- paste(driver_table$response[ord_int], collapse = " > ")
}
```

Across contaminants and isotopes, the dominant source of variation is **space**: chemical
similarity decreases with increasing geographic distance, and this spatial decay is strongest
for DDT-family chemistry and strongest overall in the multivariate index.

The second driver is **contaminant class and transport behavior**:

- DDT-family patterns are steep and typically best described by exponential decay, consistent
  with spatially localized or regionally concentrated sources plus distance attenuation.
- PCP/PCB patterns are shallower, consistent with broader transport and mixing.
- Isotopes show smoother, more gradient-like structure (often better log-linear support),
  indicating broad foraging/habitat transitions rather than sharply bounded sources.

The third driver is **time and space-time coupling**:

- Temporal lags contribute additional variation, but usually less than distance.
- The joint space-time model indicates that distance-decay strength can shift with temporal
  separation (interaction term), meaning spatial structure is not fully time-stationary.

The fourth driver is **species/ecology context**:

- Species-level fits show different slope magnitudes and best functional forms,
  implying trophic ecology and movement behavior modulate exposure patterns.
- In this dataset, high-pair species (especially *Uria* spp.) dominate signal strength,
  so future inference should balance species representation.

### Practical take-away for predictive modeling

For prediction, prioritize a **spatiotemporal model** where contamination is modeled as a
function of distance structure, year (or lag), and species effects simultaneously (for example,
GAM or mixed-effects spatial model with spatial cross-validation by area). This workflow is
best aligned with the evidence in this report: space is the strongest axis, time is informative,
and species modifies both.

If a single concise ranking is needed for this report, it is:

1. Geographic distance (spatial structure)
2. Contaminant chemistry class (DDT vs PCB/PCP vs isotope behavior)
3. Time (collection year / pairwise lag)
4. Species ecological differences

## 17) Graphics Gallery

This appendix pulls together all graphics generated by the workflow so they can
be reviewed in one place, including the original distance-decay plots and the
new spatiotemporal and biodiversity-augmented mixed-model figures.

```{r graphics-gallery, results='asis', echo=FALSE}
gallery_files <- c(
  file.path(fig_dir, "plotting-1.png"),
  file.path(fig_dir, "generality-1.png"),
  file.path(out_dir, "real_testcase_distance_decay_plots.png"),
  file.path(fig_dir, "real-binned-ci-1.png"),
  file.path(fig_dir, "real-null-slopes-1.png"),
  file.path(fig_dir, "real-null-decay-bands-1.png"),
  file.path(fig_dir, "fitted-curves-plot-1.png"),
  file.path(fig_dir, "gam-spatial-1.png"),
  file.path(fig_dir, "temporal-trend-1.png"),
  file.path(fig_dir, "spatiotemporal-visualization-1.png"),
  file.path(fig_dir, "dyadic-mixed-graphics-1.png"),
  file.path(fig_dir, "dyadic-trophic-graphics-1.png")
)

species_gallery <- list.files(fig_dir, pattern = "^species-breakdown-[0-9]+\\.png$", full.names = TRUE)
species_gallery <- species_gallery[order(species_gallery)]
gallery_files <- c(gallery_files, species_gallery)
gallery_files <- gallery_files[file.exists(gallery_files)]

pretty_title <- function(fp) {
  nm <- tools::file_path_sans_ext(basename(fp))
  nm <- gsub("-1$", "", nm)
  nm <- gsub("^plotting$", "Dummy distance-decay panels", nm)
  nm <- gsub("^generality$", "Generality across response types", nm)
  nm <- gsub("^real_testcase_distance_decay_plots$", "Real-data distance-decay fits", nm)
  nm <- gsub("^real-binned-ci$", "Binned mean similarity with 95% CI", nm)
  nm <- gsub("^real-null-slopes$", "Observed versus null slope distributions", nm)
  nm <- gsub("^real-null-decay-bands$", "Observed versus null decay bands", nm)
  nm <- gsub("^fitted-curves-plot$", "Functional-form comparison", nm)
  nm <- gsub("^gam-spatial$", "GAM spatial prediction", nm)
  nm <- gsub("^temporal-trend$", "Temporal trends", nm)
  nm <- gsub("^spatiotemporal-visualization$", "Distance-decay by time-lag tercile", nm)
  nm <- gsub("^dyadic-mixed-graphics$", "Dyadic mixed-model diagnostics", nm)
  nm <- gsub("^dyadic-trophic-graphics$", "Trophic mixed-model diagnostics", nm)
  nm <- gsub("^species-breakdown-", "Species breakdown ", nm)
  gsub("-", " ", nm)
}

for (fp in gallery_files) {
  cat("\n\n### ", pretty_title(fp), "\n\n", sep = "")
  print(knitr::include_graphics(fp))
  cat("\n")
}
```

