---
title: "California Poppy (Eschscholzia californica) Species Distribution Model"
author: "GitHub Copilot"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: show
---

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c(
  "dplyr", "terra", "raster", "maxnet", "ggplot2", "BIEN",
  "knitr", "pROC", "viridis", "geodata"
)

to_install <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(to_install) > 0) {
  install.packages(to_install, repos = "https://cran.rstudio.com")
}

library(dplyr)
library(terra)
library(raster)
library(maxnet)
library(ggplot2)
library(BIEN)
library(knitr)
library(pROC)
library(viridis)
library(geodata)

set.seed(123)

harmonize_bioc_names <- function(x) {
  nm <- names(x)
  idx <- suppressWarnings(as.integer(sub("^bio0*([0-9]+)$", "\\1", nm, ignore.case = TRUE)))
  if (any(is.na(idx))) {
    idx <- suppressWarnings(as.integer(sub(".*_([0-9]+)$", "\\1", nm)))
  }
  if (all(!is.na(idx)) && length(idx) == terra::nlyr(x)) {
    names(x) <- paste0("bio", idx)
  }
  x
}

assign_nearest_centroid <- function(points_xy, centers_xy) {
  if (nrow(points_xy) == 0) {
    return(integer(0))
  }
  out <- integer(nrow(points_xy))
  for (i in seq_len(nrow(points_xy))) {
    d2 <- (centers_xy[, 1] - points_xy[i, 1])^2 + (centers_xy[, 2] - points_xy[i, 2])^2
    out[i] <- which.min(d2)
  }
  out
}

clamp_to_training_range <- function(x, ref_df, vars) {
  mins <- vapply(ref_df[, vars, drop = FALSE], min, numeric(1), na.rm = TRUE)
  maxs <- vapply(ref_df[, vars, drop = FALSE], max, numeric(1), na.rm = TRUE)
  out <- x[[vars]]
  for (v in vars) {
    out[[v]] <- terra::clamp(out[[v]], lower = mins[[v]], upper = maxs[[v]], values = TRUE)
  }
  out
}

novelty_fraction <- function(x, ref_df, vars) {
  mins <- vapply(ref_df[, vars, drop = FALSE], min, numeric(1), na.rm = TRUE)
  maxs <- vapply(ref_df[, vars, drop = FALSE], max, numeric(1), na.rm = TRUE)
  out_stack <- terra::rast(lapply(vars, function(v) {
    (x[[v]] < mins[[v]]) | (x[[v]] > maxs[[v]])
  }))
  terra::app(out_stack, sum, na.rm = TRUE) / length(vars)
}

safe_auc <- function(obs, pred) {
  if (length(unique(obs)) < 2) {
    return(NA_real_)
  }
  as.numeric(pROC::auc(pROC::roc(response = obs, predictor = pred, quiet = TRUE)))
}

calc_metrics <- function(obs, pred) {
  auc_val <- safe_auc(obs, pred)
  if (length(unique(obs)) < 2) {
    return(data.frame(
      auc = auc_val,
      sensitivity = NA_real_,
      specificity = NA_real_,
      tss = NA_real_,
      threshold = NA_real_
    ))
  }

  roc_obj <- pROC::roc(response = obs, predictor = pred, quiet = TRUE)
  youden <- pROC::coords(
    roc_obj,
    x = "best",
    best.method = "youden",
    ret = c("threshold", "sensitivity", "specificity")
  )
  sens <- as.numeric(youden["sensitivity"])
  spec <- as.numeric(youden["specificity"])

  data.frame(
    auc = auc_val,
    sensitivity = sens,
    specificity = spec,
    tss = sens + spec - 1,
    threshold = as.numeric(youden["threshold"])
  )
}

safe_sd <- function(x) {
  if (sum(!is.na(x)) <= 1) {
    return(NA_real_)
  }
  stats::sd(x, na.rm = TRUE)
}

make_accessible_area <- function(points_df, calibration_bbox, buffer_km = 150) {
  pres_pts_ll <- terra::vect(points_df, geom = c("longitude", "latitude"), crs = "EPSG:4326")
  pres_pts_ca <- terra::project(pres_pts_ll, "EPSG:3310")
  pres_hull <- terra::convHull(pres_pts_ca)
  M_poly_ca <- terra::buffer(pres_hull, width = buffer_km * 1000)
  bbox_poly <- terra::as.polygons(calibration_bbox, crs = "EPSG:4326")
  M_poly_ll <- terra::project(M_poly_ca, "EPSG:4326")
  M_poly_ll <- terra::intersect(M_poly_ll, bbox_poly)
  if (is.null(M_poly_ll) || nrow(M_poly_ll) == 0) {
    stop("Accessible area M could not be constructed.")
  }
  M_poly_ll
}

select_nonredundant_predictors <- function(env_df, candidate_vars, priority, cutoff = 0.7) {
  keep <- intersect(candidate_vars, names(env_df))
  if (length(keep) <= 1) {
    return(keep)
  }

  repeat {
    cor_mat <- abs(stats::cor(env_df[, keep, drop = FALSE], use = "pairwise.complete.obs"))
    diag(cor_mat) <- 0
    max_cor <- suppressWarnings(max(cor_mat, na.rm = TRUE))
    if (!is.finite(max_cor) || max_cor < cutoff || length(keep) <= 1) {
      break
    }
    pair_idx <- which(cor_mat == max_cor, arr.ind = TRUE)[1, ]
    pair <- unique(c(rownames(cor_mat)[pair_idx[1]], colnames(cor_mat)[pair_idx[2]]))
    drop_var <- pair[which.max(match(pair, priority))]
    keep <- setdiff(keep, drop_var)
  }

  keep
}

fit_maxnet_model <- function(df, vars, classes = "lqh", regmult = 1.5) {
  maxnet::maxnet(
    p = df$presence,
    data = df[, vars, drop = FALSE],
    f = maxnet::maxnet.formula(df$presence, df[, vars, drop = FALSE], classes = classes),
    regmult = regmult
  )
}
```

## 1. Ecological question and target of inference

The inferential target is relative climatic suitability for *Eschscholzia californica* across California under current and future climate scenarios. The output is interpreted as a correlative climatic suitability surface on the cloglog scale, not as occupancy probability, abundance, colonization probability, or demographic performance.

That distinction matters because California poppy occurrence reflects climate together with land use, disturbance, dispersal, soils, biotic interactions, and observation effort. The goal here is therefore not causal explanation, but defensible estimation of broad climatic association while keeping transfer risk and uncertainty visible.

## 2. Occurrence audit and explicit calibration region M

The reporting extent is California. Calibration is restricted to an explicit accessible area $M$: the convex hull of cleaned California records buffered by 150 km, then truncated to a California-centered calibration box. This is still an approximation, but it is more defensible than leaving extent implicit or drawing background from the full map.

```{r occurrence-and-m}
species_name <- "Eschscholzia californica"
report_bbox <- terra::ext(-125, -114, 32, 42)
calibration_bbox <- terra::ext(-125.5, -113.5, 31.5, 42.5)
main_buffer_km <- 150

occ_raw <- BIEN::BIEN_occurrence_species(species_name)

if (!all(c("latitude", "longitude") %in% names(occ_raw))) {
  stop("BIEN output does not include latitude/longitude columns.")
}

wc <- geodata::worldclim_global(var = "bio", res = 2.5, path = "worldclim_data")
wc_calibration <- harmonize_bioc_names(terra::crop(wc, calibration_bbox))
wc_report <- harmonize_bioc_names(terra::crop(wc, report_bbox))
template_report <- wc_report[[1]]

occ_basic <- occ_raw %>%
  filter(is.finite(latitude), is.finite(longitude)) %>%
  filter(abs(latitude) <= 90, abs(longitude) <= 180) %>%
  filter(!(abs(latitude) < 1e-8 & abs(longitude) < 1e-8)) %>%
  filter(
    longitude >= terra::xmin(calibration_bbox), longitude <= terra::xmax(calibration_bbox),
    latitude >= terra::ymin(calibration_bbox), latitude <= terra::ymax(calibration_bbox)
  ) %>%
  distinct(longitude, latitude, .keep_all = TRUE)

climate_extract <- terra::extract(
  template_report,
  as.matrix(occ_basic[, c("longitude", "latitude")])
)
climate_hit <- climate_extract[[ncol(climate_extract)]]
occ_on_grid <- occ_basic[!is.na(climate_hit), , drop = FALSE]

occ_cells <- terra::cellFromXY(template_report, as.matrix(occ_on_grid[, c("longitude", "latitude")]))
occ_cell_unique <- occ_on_grid[!duplicated(occ_cells) & !is.na(occ_cells), , drop = FALSE]

if (nrow(occ_cell_unique) < 30) {
  stop("Too few cleaned California occurrences remain for a stable SDM.")
}

occ_audit <- data.frame(
  step = c(
    "Raw BIEN records",
    "After basic coordinate and extent filters",
    "After removing points off the climate grid",
    "After one record per raster cell"
  ),
  n_records = c(
    nrow(occ_raw),
    nrow(occ_basic),
    nrow(occ_on_grid),
    nrow(occ_cell_unique)
  )
)

M_poly_ll <- make_accessible_area(
  occ_cell_unique[, c("longitude", "latitude")],
  calibration_bbox = calibration_bbox,
  buffer_km = main_buffer_km
)

M_path <- as.data.frame(terra::geom(M_poly_ll)[, c("x", "y", "part")])

knitr::kable(occ_audit, caption = "Occurrence filtering summary")

ggplot() +
  geom_path(data = M_path, aes(x = x, y = y, group = part), color = "black", linewidth = 0.5) +
  geom_point(data = occ_cell_unique, aes(x = longitude, y = latitude), color = "red3", alpha = 0.6, size = 1) +
  coord_quickmap(
    xlim = c(terra::xmin(report_bbox), terra::xmax(report_bbox)),
    ylim = c(terra::ymin(report_bbox), terra::ymax(report_bbox))
  ) +
  labs(
    title = "Cleaned occurrences and explicit calibration region M",
    subtitle = sprintf("M = occurrence hull buffered %s km and truncated to the calibration box", main_buffer_km),
    x = "Longitude",
    y = "Latitude"
  )
```

Filtering was limited to reproducible steps available with the installed toolchain: invalid coordinates removed, extent restricted, points off the climate grid removed, and duplicate records collapsed to one presence per raster cell. That does not solve all georeferencing or taxonomic issues, but it removes several common sources of spurious model signal.

## 3. Ecological predictor set and collinearity control

Rather than fitting all 19 BIO variables, the workflow starts from a short, ecologically motivated candidate set for a Mediterranean-climate annual and then thins it for redundancy within $M$ using a simple $|r| < 0.7$ rule.

```{r predictors}
wc_M <- terra::mask(terra::crop(wc_calibration, M_poly_ll), M_poly_ll)

candidate_predictors <- c("bio1", "bio6", "bio12", "bio15", "bio18", "bio4")
predictor_rationale <- data.frame(
  variable = candidate_predictors,
  rationale = c(
    "Annual thermal regime",
    "Cold limitation and frost exposure",
    "Overall moisture supply",
    "Mediterranean precipitation seasonality",
    "Warm-quarter moisture",
    "Temperature seasonality"
  ),
  stringsAsFactors = FALSE
)

env_values <- na.omit(as.data.frame(wc_M[[candidate_predictors]], xy = FALSE))
pred_names <- select_nonredundant_predictors(
  env_values,
  candidate_vars = candidate_predictors,
  priority = candidate_predictors,
  cutoff = 0.7
)

predictor_tbl <- predictor_rationale %>%
  mutate(selected = ifelse(variable %in% pred_names, "Kept", "Dropped as redundant"))

cor_tbl <- as.data.frame(as.table(round(stats::cor(env_values[, candidate_predictors, drop = FALSE], use = "pairwise.complete.obs"), 2)))
names(cor_tbl) <- c("var1", "var2", "correlation")

knitr::kable(predictor_tbl, caption = "Candidate predictors and retained set")
knitr::kable(cor_tbl, caption = "Candidate predictor correlations within M")

pred_names
```

## 4. Bias-aware background sampling

Background points are sampled within $M$ with probability proportional to a smoothed presence density surface. This is a practical bias-aware approximation of sampling effort and is preferable here to naive uniform background over the entire study region.

```{r background}
template_M <- wc_M[[1]]

pres_cells <- terra::cellFromXY(template_M, as.matrix(occ_cell_unique[, c("longitude", "latitude")]))
pres_cells <- unique(pres_cells[!is.na(pres_cells)])
pres_xy <- terra::xyFromCell(template_M, pres_cells)
pres_df <- data.frame(x = pres_xy[, 1], y = pres_xy[, 2])

bias_r <- template_M
terra::values(bias_r) <- 0
terra::values(bias_r)[pres_cells] <- 1
bias_r <- terra::focal(bias_r, w = matrix(1, 9, 9), fun = sum, na.policy = "omit")

bias_vals <- terra::values(bias_r)
bg_candidates <- setdiff(which(!is.na(bias_vals)), pres_cells)
bg_weights <- pmax(bias_vals[bg_candidates], 1e-6)
bg_weights <- bg_weights / sum(bg_weights)

bg_n <- min(10000, length(bg_candidates))
bg_cells <- sample(bg_candidates, size = bg_n, replace = FALSE, prob = bg_weights)
bg_xy <- terra::xyFromCell(template_M, bg_cells)
bg_df <- data.frame(x = bg_xy[, 1], y = bg_xy[, 2])

plot_df <- bind_rows(
  data.frame(x = pres_df$x, y = pres_df$y, type = "Presence"),
  data.frame(x = bg_df$x, y = bg_df$y, type = "Bias-aware background")
)

ggplot(plot_df, aes(x = x, y = y, color = type)) +
  geom_point(size = 0.25, alpha = 0.5) +
  geom_path(data = M_path, aes(x = x, y = y, group = part), inherit.aes = FALSE, color = "black") +
  coord_quickmap(
    xlim = c(terra::xmin(report_bbox), terra::xmax(report_bbox)),
    ylim = c(terra::ymin(report_bbox), terra::ymax(report_bbox))
  ) +
  scale_color_manual(values = c("Presence" = "red3", "Bias-aware background" = "gray50")) +
  labs(title = "Calibration data inside M", x = "Longitude", y = "Latitude")
```

```{r model-data}
pres_vals <- as.data.frame(terra::extract(wc_M[[pred_names]], as.matrix(pres_df)))
bg_vals <- as.data.frame(terra::extract(wc_M[[pred_names]], as.matrix(bg_df)))

if ("ID" %in% names(pres_vals)) {
  pres_vals <- pres_vals[, setdiff(names(pres_vals), "ID"), drop = FALSE]
}
if ("ID" %in% names(bg_vals)) {
  bg_vals <- bg_vals[, setdiff(names(bg_vals), "ID"), drop = FALSE]
}

model_input <- bind_rows(
  cbind(pres_df, presence = 1L, pres_vals),
  cbind(bg_df, presence = 0L, bg_vals)
) %>%
  filter(if_all(all_of(pred_names), ~ !is.na(.x)))

if (sum(model_input$presence == 1L) < 30) {
  stop("Insufficient cleaned presence records remain after predictor extraction.")
}

data.frame(
  class = c("Presence", "Background"),
  n = c(sum(model_input$presence == 1L), sum(model_input$presence == 0L))
) %>%
  knitr::kable(caption = "Presence-background sample sizes used for calibration")
```

## 5. Complexity control and spatial validation

The main model is intentionally constrained: linear, quadratic, and hinge features with `regmult = 1.5`. Validation is based on spatially structured folds rather than only random splits, because nearby records often make random validation look better than the model will perform under spatial transfer.

```{r validation}
coords_ll <- as.matrix(model_input[, c("x", "y")])
coords_m <- terra::project(coords_ll, from = "EPSG:4326", to = "EPSG:3310")

pres_rows <- which(model_input$presence == 1L)
bg_rows <- which(model_input$presence == 0L)
k_folds <- min(5, max(3, floor(length(pres_rows) / 20)))

if (k_folds < 3) {
  stop("Too few cleaned presences for spatial cross-validation.")
}

km <- kmeans(coords_m[pres_rows, , drop = FALSE], centers = k_folds, nstart = 25)
fold_id <- integer(nrow(model_input))
fold_id[pres_rows] <- km$cluster
fold_id[bg_rows] <- assign_nearest_centroid(coords_m[bg_rows, , drop = FALSE], km$centers)

cv_stats <- vector("list", k_folds)
cv_models <- vector("list", k_folds)
cv_thresholds <- rep(NA_real_, k_folds)

for (fold in seq_len(k_folds)) {
  test_idx <- which(fold_id == fold)
  train_idx <- which(fold_id != fold)

  train_df <- model_input[train_idx, , drop = FALSE]
  test_df <- model_input[test_idx, , drop = FALSE]

  mx_fold <- fit_maxnet_model(train_df, pred_names, classes = "lqh", regmult = 1.5)
  fold_pred <- predict(mx_fold, test_df[, pred_names, drop = FALSE], type = "cloglog")
  fold_metrics <- calc_metrics(test_df$presence, fold_pred)

  cv_models[[fold]] <- mx_fold
  cv_thresholds[fold] <- fold_metrics$threshold
  cv_stats[[fold]] <- cbind(
    data.frame(fold = fold, n_test = nrow(test_df), n_test_presence = sum(test_df$presence == 1L)),
    fold_metrics
  )
}

cv_tbl <- bind_rows(cv_stats)

spatial_summary <- data.frame(
  metric = c(
    "Spatial CV AUC mean",
    "Spatial CV AUC sd",
    "Spatial CV TSS mean",
    "Spatial CV TSS sd",
    "Spatial CV threshold median"
  ),
  value = c(
    mean(cv_tbl$auc, na.rm = TRUE),
    safe_sd(cv_tbl$auc),
    mean(cv_tbl$tss, na.rm = TRUE),
    safe_sd(cv_tbl$tss),
    median(cv_tbl$threshold, na.rm = TRUE)
  )
)

set.seed(2026)
rand_train_p <- sample(pres_rows, floor(0.8 * length(pres_rows)))
rand_train_b <- sample(bg_rows, floor(0.8 * length(bg_rows)))
rand_train <- c(rand_train_p, rand_train_b)
rand_test <- setdiff(seq_len(nrow(model_input)), rand_train)

rand_model <- fit_maxnet_model(model_input[rand_train, , drop = FALSE], pred_names, classes = "lqh", regmult = 1.5)
rand_pred <- predict(rand_model, model_input[rand_test, pred_names, drop = FALSE], type = "cloglog")
rand_metrics <- calc_metrics(model_input$presence[rand_test], rand_pred)

validation_compare <- data.frame(
  validation = c("Spatial CV mean", "Random 80/20 split"),
  auc = c(mean(cv_tbl$auc, na.rm = TRUE), rand_metrics$auc),
  tss = c(mean(cv_tbl$tss, na.rm = TRUE), rand_metrics$tss)
)

knitr::kable(cv_tbl, digits = 4, caption = "Fold-level diagnostics from spatial cross-validation")
knitr::kable(validation_compare, digits = 4, caption = "Validation sensitivity: spatial folds versus random split")
knitr::kable(spatial_summary, digits = 4, caption = "Primary validation summary")
```

Random-split performance is retained only as a sensitivity comparison. If it is substantially higher than the spatial result, that should be interpreted as validation inflation rather than evidence of strong transferability.

## 6. Final model, current suitability, and uncertainty

```{r final-model}
mx <- fit_maxnet_model(model_input, pred_names, classes = "lqh", regmult = 1.5)

thr <- median(cv_thresholds, na.rm = TRUE)
if (!is.finite(thr)) {
  thr <- 0.5
}

wc_report_selected <- wc_report[[pred_names]]
wc_report_clamped <- clamp_to_training_range(wc_report_selected, model_input, pred_names)
pred_ca <- raster::predict(raster::stack(wc_report_clamped), mx, type = "cloglog", progress = "text")
bin_map <- pred_ca > thr

pred_df <- as.data.frame(raster::rasterToPoints(pred_ca), stringsAsFactors = FALSE)
colnames(pred_df) <- c("x", "y", "suitability")

ggplot(pred_df, aes(x = x, y = y, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Suitability") +
  geom_point(data = pres_df, aes(x = x, y = y), inherit.aes = FALSE, color = "red3", alpha = 0.5, size = 0.7) +
  coord_quickmap() +
  labs(
    title = "Current relative climatic suitability",
    subtitle = "Projected over California with clamping to the calibration range"
  )
```

```{r uncertainty-and-novelty}
cv_pred_rasters <- lapply(cv_models, function(m) {
  raster::predict(raster::stack(wc_report_clamped), m, type = "cloglog", progress = "text")
})

cv_pred_stack <- raster::stack(cv_pred_rasters)
pred_sd <- raster::calc(cv_pred_stack, fun = sd, na.rm = TRUE)

sd_df <- as.data.frame(raster::rasterToPoints(pred_sd), stringsAsFactors = FALSE)
colnames(sd_df) <- c("x", "y", "sd")

ggplot(sd_df, aes(x = x, y = y, fill = sd)) +
  geom_raster() +
  scale_fill_viridis_c(name = "SD") +
  coord_quickmap() +
  labs(title = "Prediction uncertainty", subtitle = "SD across spatial CV models")

novelty_current <- novelty_fraction(wc_report_selected, model_input, pred_names)
nov_df <- as.data.frame(terra::as.points(novelty_current), geom = "xy", na.rm = TRUE)
colnames(nov_df) <- c("x", "y", "novelty")

ggplot(nov_df, aes(x = x, y = y, fill = novelty)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Novelty", limits = c(0, 1)) +
  coord_quickmap() +
  labs(title = "Current environmental novelty relative to calibration")

diag_tbl <- bind_rows(
  spatial_summary,
  data.frame(metric = "Current novelty fraction (>0)", value = mean(terra::values(novelty_current) > 0, na.rm = TRUE)),
  data.frame(metric = "Mean prediction SD (spatial CV)", value = mean(raster::values(pred_sd), na.rm = TRUE)),
  data.frame(metric = "Threshold used for binary map", value = thr)
)

knitr::kable(diag_tbl, digits = 4, caption = "Diagnostics summarizing evaluation and uncertainty")
```

## 7. Variable effects and interpretation

```{r interpretation}
med <- apply(model_input[, pred_names, drop = FALSE], 2, median, na.rm = TRUE)

response_df <- do.call(rbind, lapply(pred_names, function(v) {
  xseq <- seq(min(model_input[[v]], na.rm = TRUE), max(model_input[[v]], na.rm = TRUE), length.out = 120)
  nd <- as.data.frame(matrix(rep(med, each = length(xseq)), nrow = length(xseq)))
  names(nd) <- pred_names
  nd[[v]] <- xseq
  data.frame(variable = v, x = xseq, suitability = predict(mx, nd, type = "cloglog"))
}))

ggplot(response_df, aes(x = x, y = suitability)) +
  geom_line(color = "midnightblue") +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Partial response curves", x = "Predictor value", y = "Predicted suitability")

perm_imp_list <- lapply(pred_names, function(v) {
  fold_drops <- lapply(seq_len(k_folds), function(fold) {
    test_idx <- which(fold_id == fold)
    test_df <- model_input[test_idx, , drop = FALSE]
    if (nrow(test_df) == 0 || length(unique(test_df$presence)) < 2) {
      return(NA_real_)
    }

    base_pred <- predict(cv_models[[fold]], test_df[, pred_names, drop = FALSE], type = "cloglog")
    auc_base <- safe_auc(test_df$presence, base_pred)

    perm_df <- test_df[, pred_names, drop = FALSE]
    perm_df[[v]] <- sample(perm_df[[v]])
    perm_pred <- predict(cv_models[[fold]], perm_df, type = "cloglog")
    auc_perm <- safe_auc(test_df$presence, perm_pred)
    auc_base - auc_perm
  })
  data.frame(variable = v, importance = mean(unlist(fold_drops), na.rm = TRUE))
})

perm_imp <- bind_rows(perm_imp_list) %>% arrange(desc(importance))

knitr::kable(perm_imp, digits = 4, caption = "Permutation importance from spatial CV folds")

ggplot(perm_imp, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Permutation importance", x = "Predictor", y = "AUC drop")
```

Interpretation should remain conditional and correlative:

- Higher suitability indicates climates more similar to the cleaned occurrence-background contrast used here, not a causal effect of climate alone.
- Response curves and importance values depend on the chosen $M$, predictor set, bias model, and regularization settings.
- Areas with high novelty or high cross-validation variability are weakly supported and should not be treated as strong evidence of future occupancy.

## 8. Sensitivity analysis

At least one practical sensitivity comparison is useful because SDM conclusions can depend heavily on model settings. Here, the main model is compared with a more flexible alternative.

```{r sensitivity}
mx_complex <- fit_maxnet_model(model_input, pred_names, classes = "lqhp", regmult = 1.0)
pred_complex <- raster::predict(raster::stack(wc_report_clamped), mx_complex, type = "cloglog", progress = "text")

main_vals <- raster::values(pred_ca)
complex_vals <- raster::values(pred_complex)

sensitivity_tbl <- data.frame(
  comparison = c(
    "Map correlation: main vs more complex model",
    "Mean suitability: main model",
    "Mean suitability: more complex model",
    "Suitable fraction (> main threshold): main model",
    "Suitable fraction (> main threshold): more complex model"
  ),
  value = c(
    stats::cor(main_vals, complex_vals, use = "complete.obs"),
    mean(main_vals, na.rm = TRUE),
    mean(complex_vals, na.rm = TRUE),
    mean(main_vals > thr, na.rm = TRUE),
    mean(complex_vals > thr, na.rm = TRUE)
  )
)

knitr::kable(sensitivity_tbl, digits = 4, caption = "Sensitivity of current predictions to model complexity")
```

If that comparison changes broad spatial patterns substantially, interpretation should be conservative. If the maps remain similar, confidence is higher that the main results are not purely an artifact of one feature setting.

## 9. Future projections, novelty, and transfer risk

Future projections are treated as conditional transfer exercises rather than direct forecasts of realized occupancy. They assume that present-day climatic associations remain informative through time, while dispersal limits, habitat conversion, disturbance, and species interactions remain outside the fitted model.

```{r future-projections, message=TRUE, warning=FALSE}
ssp_levels <- c("126", "245", "370", "585")
future_time <- "2041-2060"
future_res <- 10

available_gcms <- get(".cmods", envir = asNamespace("geodata"))
set.seed(2026)
selected_gcms <- sample(available_gcms, 3)

future_pred_list <- list()
future_summary <- list()

for (gcm in selected_gcms) {
  for (ssp in ssp_levels) {
    message("Projecting: ", gcm, " | SSP", ssp, " | ", future_time)

    fut <- geodata::cmip6_world(
      model = gcm,
      ssp = ssp,
      time = future_time,
      var = "bio",
      res = future_res,
      path = "worldclim_data"
    )

    fut_ca <- harmonize_bioc_names(terra::crop(fut, report_bbox))
    missing_preds <- setdiff(pred_names, names(fut_ca))
    if (length(missing_preds) > 0) {
      stop(sprintf("Missing predictors for %s SSP%s: %s", gcm, ssp, paste(missing_preds, collapse = ", ")))
    }

    fut_selected <- fut_ca[[pred_names]]
    fut_novelty <- novelty_fraction(fut_selected, model_input, pred_names)
    fut_clamped <- clamp_to_training_range(fut_selected, model_input, pred_names)
    fut_pred <- raster::predict(raster::stack(fut_clamped), mx, type = "cloglog", progress = "text")

    key <- paste(gcm, paste0("SSP", ssp), sep = "__")
    future_pred_list[[key]] <- fut_pred

    future_summary[[key]] <- data.frame(
      gcm = gcm,
      scenario = paste0("RCP/SSP", ssp),
      period = future_time,
      mean_suitability = mean(raster::values(fut_pred), na.rm = TRUE),
      suitable_fraction = mean(raster::values(fut_pred > thr) == 1, na.rm = TRUE),
      novelty_fraction = mean(terra::values(fut_novelty) > 0, na.rm = TRUE),
      high_novelty_fraction = mean(terra::values(fut_novelty) > 0.2, na.rm = TRUE)
    )
  }
}

future_summary_df <- bind_rows(future_summary)

current_baseline <- data.frame(
  gcm = "Current climate",
  scenario = "Current",
  period = "1970-2000",
  mean_suitability = mean(raster::values(pred_ca), na.rm = TRUE),
  suitable_fraction = mean(raster::values(bin_map) == 1, na.rm = TRUE),
  novelty_fraction = mean(terra::values(novelty_current) > 0, na.rm = TRUE),
  high_novelty_fraction = mean(terra::values(novelty_current) > 0.2, na.rm = TRUE)
)

future_summary_w_baseline <- bind_rows(current_baseline, future_summary_df)

future_uncertainty <- future_summary_df %>%
  group_by(scenario) %>%
  summarize(
    mean_suitable_fraction = mean(suitable_fraction, na.rm = TRUE),
    sd_suitable_fraction = safe_sd(suitable_fraction),
    mean_novelty_fraction = mean(novelty_fraction, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(future_summary_w_baseline, digits = 4, caption = "Current and future projections with novelty diagnostics")
knitr::kable(future_uncertainty, digits = 4, caption = "Across-GCM summary by scenario")
```

```{r future-maps}
ensemble_by_ssp <- lapply(ssp_levels, function(ssp) {
  keys <- grep(paste0("__SSP", ssp, "$"), names(future_pred_list), value = TRUE)
  raster::calc(raster::stack(future_pred_list[keys]), fun = mean, na.rm = TRUE)
})
names(ensemble_by_ssp) <- paste0("RCP/SSP", ssp_levels)

ensemble_df <- bind_rows(lapply(names(ensemble_by_ssp), function(scn) {
  d <- as.data.frame(raster::rasterToPoints(ensemble_by_ssp[[scn]]), stringsAsFactors = FALSE)
  colnames(d) <- c("x", "y", "suitability")
  d$scenario <- scn
  d
}))

ggplot(ensemble_df, aes(x = x, y = y, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Suitability") +
  coord_quickmap() +
  facet_wrap(~ scenario, ncol = 2) +
  labs(
    title = sprintf("Future relative climatic suitability (%s)", future_time),
    subtitle = "Ensemble means across three GCMs; predictors clamped to the calibration range"
  )

ggplot(future_summary_df, aes(x = scenario, y = suitable_fraction, fill = gcm)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Suitable-area fraction by scenario and GCM",
    x = "Scenario",
    y = "Fraction of California predicted suitable"
  )
```

Future maps should therefore be read narrowly: they show where current climatic associations project into future climate space under a niche-stability assumption. They do not imply successful dispersal, persistence, or biotic release, and they are least trustworthy where novelty is high.

## 10. Limitations

- The occurrence audit is stronger than a raw BIEN pull but is not a full taxonomic or georeferencing review.
- The explicit $M$ is defensible, but still approximate; different dispersal assumptions would change background composition and model calibration.
- Climatic predictors are intentionally short and interpretable, which improves defensibility but omits soils, disturbance, habitat conversion, and biotic interactions.
- Novelty is summarized with a simple range-based diagnostic rather than a full multivariate analog analysis.

## 11. Save outputs

```{r save}
raster::writeRaster(pred_ca, filename = "california_poppy_suitability.tif", overwrite = TRUE)
raster::writeRaster(bin_map, filename = "california_poppy_suitability_binary.tif", overwrite = TRUE)
raster::writeRaster(pred_sd, filename = "california_poppy_prediction_uncertainty.tif", overwrite = TRUE)
raster::writeRaster(raster::raster(novelty_current), filename = "california_poppy_current_novelty.tif", overwrite = TRUE)

write.csv(diag_tbl, "sdm_diagnostics.csv", row.names = FALSE)
write.csv(perm_imp, "sdm_permutation_importance.csv", row.names = FALSE)
write.csv(future_summary_w_baseline, "sdm_future_projections_summary.csv", row.names = FALSE)
write.csv(sensitivity_tbl, "sdm_sensitivity_summary.csv", row.names = FALSE)

for (scn in names(ensemble_by_ssp)) {
  out_name <- paste0(
    "california_poppy_future_ensemble_",
    gsub("[^0-9]", "", scn),
    "_",
    gsub("-", "", future_time),
    ".tif"
  )
  raster::writeRaster(ensemble_by_ssp[[scn]], filename = out_name, overwrite = TRUE)
}

cat("Saved outputs:\n")
cat("- california_poppy_suitability.tif\n")
cat("- california_poppy_suitability_binary.tif\n")
cat("- california_poppy_prediction_uncertainty.tif\n")
cat("- california_poppy_current_novelty.tif\n")
cat("- sdm_diagnostics.csv\n")
cat("- sdm_permutation_importance.csv\n")
cat("- sdm_future_projections_summary.csv\n")
cat("- sdm_sensitivity_summary.csv\n")
cat("- california_poppy_future_ensemble_126_20412060.tif\n")
cat("- california_poppy_future_ensemble_245_20412060.tif\n")
cat("- california_poppy_future_ensemble_370_20412060.tif\n")
cat("- california_poppy_future_ensemble_585_20412060.tif\n")
```

## 12. Session info

```{r session-info}
sessionInfo()
```
