---
title: "California Poppy (Eschscholzia californica) Species Distribution Model"
author: "GitHub Copilot"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: show
---

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c(
  "dplyr", "terra", "raster", "maxnet", "ggplot2", "BIEN",
  "knitr", "pROC", "viridis", "geodata"
)

to_install <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(to_install) > 0) {
  install.packages(to_install, repos = "https://cran.rstudio.com")
}

library(dplyr)
library(terra)
library(raster)
library(maxnet)
library(ggplot2)
library(BIEN)
library(knitr)
library(pROC)
library(viridis)
library(geodata)

set.seed(123)

harmonize_bioc_names <- function(x) {
  nm <- names(x)
  idx <- suppressWarnings(as.integer(sub("^bio0*([0-9]+)$", "\\1", nm, ignore.case = TRUE)))
  if (any(is.na(idx))) {
    idx <- suppressWarnings(as.integer(sub(".*_([0-9]+)$", "\\1", nm)))
  }
  if (all(!is.na(idx)) && length(idx) == terra::nlyr(x)) {
    names(x) <- paste0("bio", idx)
  }
  x
}

assign_nearest_centroid <- function(points_xy, centers_xy) {
  if (nrow(points_xy) == 0) {
    return(integer(0))
  }
  out <- integer(nrow(points_xy))
  for (i in seq_len(nrow(points_xy))) {
    d2 <- (centers_xy[, 1] - points_xy[i, 1])^2 + (centers_xy[, 2] - points_xy[i, 2])^2
    out[i] <- which.min(d2)
  }
  out
}

clamp_to_training_range <- function(x, ref_df, vars) {
  mins <- vapply(ref_df[, vars, drop = FALSE], min, numeric(1), na.rm = TRUE)
  maxs <- vapply(ref_df[, vars, drop = FALSE], max, numeric(1), na.rm = TRUE)
  out <- x[[vars]]
  for (v in vars) {
    out[[v]] <- terra::clamp(out[[v]], lower = mins[[v]], upper = maxs[[v]], values = TRUE)
  }
  out
}

novelty_fraction <- function(x, ref_df, vars) {
  mins <- vapply(ref_df[, vars, drop = FALSE], min, numeric(1), na.rm = TRUE)
  maxs <- vapply(ref_df[, vars, drop = FALSE], max, numeric(1), na.rm = TRUE)
  out_stack <- terra::rast(lapply(vars, function(v) {
    (x[[v]] < mins[[v]]) | (x[[v]] > maxs[[v]])
  }))
  terra::app(out_stack, sum, na.rm = TRUE) / length(vars)
}

safe_auc <- function(obs, pred) {
  if (length(unique(obs)) < 2) {
    return(NA_real_)
  }
  as.numeric(pROC::auc(pROC::roc(response = obs, predictor = pred, quiet = TRUE)))
}
```

## 1. Modeling objective, target, and assumptions

This SDM is interpretation-first: estimate relative climatic suitability for *Eschscholzia californica* under current and future climates, then interpret ecological signal and transfer risks.

Prediction target: gridded climatic suitability (cloglog output) at WorldClim resolution, not true occupancy probability.

Key assumptions:

- occurrence records represent realized environmental associations after cleaning
- calibration is restricted to an explicit accessible area ($M$)
- background sampling approximates sampling effort (bias-aware)
- temporal transfer to future climates is conditional on niche stability
- extrapolation beyond calibration environments is flagged and constrained

## 2. Occurrence data from BIEN and initial cleaning

```{r occ-data}
species_name <- "Eschscholzia californica"

occ <- BIEN::BIEN_occurrence_species(species_name)

if (!all(c("latitude", "longitude") %in% names(occ))) {
  stop("BIEN output does not include latitude/longitude columns.")
}

occ <- occ %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  distinct(latitude, longitude, .keep_all = TRUE)

ca_bbox <- terra::ext(-125, -114, 32, 42)

occ_ca <- occ %>%
  filter(longitude >= -125, longitude <= -114, latitude >= 32, latitude <= 42)

if (nrow(occ_ca) < 30) {
  stop("Too few California occurrences after filtering; cannot build a stable SDM.")
}

nrow(occ_ca)
head(occ_ca)
```

## 3. Define accessible area (M) for calibration

```{r accessible-area}
# Accessible area M: convex hull of California occurrences + 150 km buffer,
# then clipped to California bounding extent.
pres_pts_ll <- terra::vect(occ_ca[, c("longitude", "latitude")], geom = c("longitude", "latitude"), crs = "EPSG:4326")
pres_pts_ca <- terra::project(pres_pts_ll, "EPSG:3310")
pres_hull <- terra::convHull(pres_pts_ca)
M_poly_ca <- terra::buffer(pres_hull, width = 150000)
M_poly_ll <- terra::project(M_poly_ca, "EPSG:4326")

ca_poly <- terra::as.polygons(ca_bbox, crs = "EPSG:4326")
M_poly_ll <- terra::intersect(M_poly_ll, ca_poly)

if (is.null(M_poly_ll) || nrow(M_poly_ll) == 0) {
  stop("Accessible area M could not be constructed.")
}

M_poly_ll
```

## 4. Climate predictors and calibration mask

```{r climate}
wc <- geodata::worldclim_global(var = "bio", res = 2.5, path = "worldclim_data")
wc_ca <- terra::crop(wc, ca_bbox)
wc_ca <- harmonize_bioc_names(wc_ca)

pred_names <- names(wc_ca)

wc_M <- terra::mask(terra::crop(wc_ca, M_poly_ll), M_poly_ll)

pred_names
```

## 5. Sampling bias handling and target-group-style background

```{r background}
template <- wc_M[[1]]

pres_xy_raw <- as.matrix(occ_ca[, c("longitude", "latitude")])
in_M <- !is.na(terra::extract(template, pres_xy_raw)[, 1])
pres_xy_raw <- pres_xy_raw[in_M, , drop = FALSE]

if (nrow(pres_xy_raw) < 30) {
  stop("Too few occurrences inside M after filtering to climate coverage.")
}

# One presence per raster cell reduces duplicate sampling effort signals.
pres_cells <- terra::cellFromXY(template, pres_xy_raw)
pres_cells <- unique(pres_cells[!is.na(pres_cells)])
pres_xy <- terra::xyFromCell(template, pres_cells)
pres_df <- data.frame(x = pres_xy[, 1], y = pres_xy[, 2])

# Bias-aware background: sample with probabilities proportional to smoothed
# presence density across M (target-group-style effort proxy).
bias_r <- template
terra::values(bias_r) <- 0
terra::values(bias_r)[pres_cells] <- 1
bias_r <- terra::focal(bias_r, w = matrix(1, 7, 7), fun = sum, na.policy = "omit")

bias_vals <- terra::values(bias_r)
valid_cells <- which(!is.na(bias_vals))
w <- pmax(bias_vals[valid_cells], 1e-6)
w <- w / sum(w)

bg_n <- 10000
bg_cells <- sample(valid_cells, size = bg_n, replace = TRUE, prob = w)
bg_xy <- terra::xyFromCell(template, bg_cells)
bg_df <- data.frame(x = bg_xy[, 1], y = bg_xy[, 2])

plot_df <- bind_rows(
  data.frame(x = pres_df$x, y = pres_df$y, type = "Presence"),
  data.frame(x = bg_df$x, y = bg_df$y, type = "Background")
)

ggplot(plot_df, aes(x, y, color = type)) +
  geom_point(size = 0.3, alpha = 0.5) +
  geom_path(
    data = as.data.frame(terra::geom(M_poly_ll)[, c("x", "y", "part")]),
    aes(x = x, y = y, group = part),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.4
  ) +
  coord_quickmap(xlim = c(-125, -114), ylim = c(32, 42)) +
  scale_color_manual(values = c("Presence" = "red3", "Background" = "gray50")) +
  labs(title = "Calibration data in M", x = "Longitude", y = "Latitude")
```

## 6. Presence-background matrix

```{r pb-data}
pres_vals_raw <- terra::extract(wc_M, as.matrix(pres_df))
bg_vals_raw <- terra::extract(wc_M, as.matrix(bg_df))

pres_vals <- as.data.frame(pres_vals_raw)[, intersect(pred_names, names(pres_vals_raw)), drop = FALSE]
bg_vals <- as.data.frame(bg_vals_raw)[, intersect(pred_names, names(bg_vals_raw)), drop = FALSE]

pres_dat <- cbind(pres_df, presence = 1, pres_vals)
bg_dat <- cbind(bg_df, presence = 0, bg_vals)

model_input <- bind_rows(pres_dat, bg_dat) %>%
  filter(if_all(all_of(pred_names), ~ !is.na(.x)))

if (sum(model_input$presence == 1) < 30) {
  stop("Insufficient cleaned presence records for modeling.")
}

head(model_input)
dim(model_input)
```

## 7. Spatially structured validation (clustered folds)

```{r spatial-cv}
coords_ll <- as.matrix(model_input[, c("x", "y")])
coords_m <- terra::project(coords_ll, from = "EPSG:4326", to = "EPSG:3310")

pres_rows <- which(model_input$presence == 1)
bg_rows <- which(model_input$presence == 0)

n_pres <- length(pres_rows)
k_folds <- min(5, max(2, floor(n_pres / 20)))

if (k_folds < 2) {
  stop("Too few presence points for spatially structured validation.")
}

km <- kmeans(coords_m[pres_rows, , drop = FALSE], centers = k_folds, nstart = 20)
fold_id <- integer(nrow(model_input))
fold_id[pres_rows] <- km$cluster
fold_id[bg_rows] <- assign_nearest_centroid(coords_m[bg_rows, , drop = FALSE], km$centers)

cv_stats <- vector("list", k_folds)
cv_models <- vector("list", k_folds)
cv_thresholds <- rep(NA_real_, k_folds)

for (fold in seq_len(k_folds)) {
  test_idx <- which(fold_id == fold)
  train_idx <- which(fold_id != fold)

  train_df <- model_input[train_idx, , drop = FALSE]
  test_df <- model_input[test_idx, , drop = FALSE]

  x_train <- train_df[, pred_names, drop = FALSE]
  p_train <- train_df$presence

  mx_fold <- maxnet::maxnet(
    p = p_train,
    data = x_train,
    f = maxnet.formula(p_train, x_train, classes = "lqhp")
  )

  x_test <- test_df[, pred_names, drop = FALSE]
  p_test <- test_df$presence
  score <- predict(mx_fold, x_test, type = "cloglog")
  auc_val <- safe_auc(p_test, score)

  if (length(unique(p_test)) >= 2) {
    roc_obj <- pROC::roc(response = p_test, predictor = score, quiet = TRUE)
    youden <- pROC::coords(
      roc_obj,
      x = "best",
      best.method = "youden",
      ret = c("threshold", "sensitivity", "specificity")
    )
    thr <- as.numeric(youden["threshold"])
    sens <- as.numeric(youden["sensitivity"])
    spec <- as.numeric(youden["specificity"])
    tss <- sens + spec - 1
  } else {
    thr <- NA_real_
    sens <- NA_real_
    spec <- NA_real_
    tss <- NA_real_
  }

  cv_models[[fold]] <- mx_fold
  cv_thresholds[fold] <- thr
  cv_stats[[fold]] <- data.frame(
    fold = fold,
    n_test = nrow(test_df),
    n_test_presence = sum(p_test == 1),
    auc = auc_val,
    sensitivity = sens,
    specificity = spec,
    tss = tss,
    threshold = thr
  )
}

cv_tbl <- bind_rows(cv_stats)

diag_tbl <- data.frame(
  metric = c("Spatial CV AUC (mean)", "Spatial CV AUC (sd)", "Spatial CV TSS (mean)", "Spatial CV TSS (sd)"),
  value = c(
    mean(cv_tbl$auc, na.rm = TRUE),
    sd(cv_tbl$auc, na.rm = TRUE),
    mean(cv_tbl$tss, na.rm = TRUE),
    sd(cv_tbl$tss, na.rm = TRUE)
  )
)

knitr::kable(cv_tbl, digits = 4, caption = "Spatially structured fold diagnostics")
knitr::kable(diag_tbl, digits = 4, caption = "Summary diagnostics (spatial CV)")
```

## 8. Final model fit and current prediction

```{r final-model}
x_all <- model_input[, pred_names, drop = FALSE]
p_all <- model_input$presence

mx <- maxnet::maxnet(
  p = p_all,
  data = x_all,
  f = maxnet.formula(p_all, x_all, classes = "lqhp")
)

wc_ca_clamped <- clamp_to_training_range(wc_ca, model_input, pred_names)
pred_ca <- raster::predict(raster::stack(wc_ca_clamped), mx, type = "cloglog", progress = "text")

thr <- median(cv_thresholds, na.rm = TRUE)
if (!is.finite(thr)) {
  thr <- 0.5
}
bin_map <- pred_ca > thr

pred_df <- as.data.frame(raster::rasterToPoints(pred_ca), stringsAsFactors = FALSE)
colnames(pred_df) <- c("x", "y", "suitability")

ggplot(pred_df, aes(x = x, y = y, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Suitability") +
  geom_point(data = pres_df, aes(x = x, y = y), color = "red", size = 0.9, alpha = 0.5, inherit.aes = FALSE) +
  coord_quickmap() +
  labs(
    title = "Current suitability prediction",
    subtitle = "Final model calibrated in M, projected with predictor clamping"
  )
```

## 9. Uncertainty and extrapolation diagnostics

```{r uncertainty-extrapolation}
# Spatial CV model spread as model-structure uncertainty proxy.
cv_pred_rasters <- lapply(cv_models, function(m) {
  raster::predict(raster::stack(wc_ca_clamped), m, type = "cloglog", progress = "text")
})

cv_pred_stack <- raster::stack(cv_pred_rasters)
pred_sd <- raster::calc(cv_pred_stack, fun = sd, na.rm = TRUE)

sd_df <- as.data.frame(raster::rasterToPoints(pred_sd), stringsAsFactors = FALSE)
colnames(sd_df) <- c("x", "y", "sd")

ggplot(sd_df, aes(x = x, y = y, fill = sd)) +
  geom_raster() +
  scale_fill_viridis_c(name = "SD") +
  coord_quickmap() +
  labs(title = "Prediction uncertainty (SD across spatial CV models)")

# Novelty map: fraction of predictors outside calibration range.
novelty_current <- novelty_fraction(wc_ca, model_input, pred_names)
nov_df <- as.data.frame(terra::as.points(novelty_current), geom = "xy", na.rm = TRUE)
colnames(nov_df) <- c("x", "y", "novelty")

ggplot(nov_df, aes(x = x, y = y, fill = novelty)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Novelty", limits = c(0, 1)) +
  coord_quickmap() +
  labs(title = "Current-climate environmental novelty relative to calibration")

novelty_warn_current <- mean(terra::values(novelty_current) > 0, na.rm = TRUE)
uncertainty_mean <- mean(raster::values(pred_sd), na.rm = TRUE)

diag_tbl <- bind_rows(
  diag_tbl,
  data.frame(metric = "Current novelty fraction (>0)", value = novelty_warn_current),
  data.frame(metric = "Mean prediction SD (spatial CV)", value = uncertainty_mean)
)

knitr::kable(diag_tbl, digits = 4, caption = "Diagnostics including uncertainty and extrapolation")
```

## 10. Variable effects, ecological interpretation, and limits

```{r interpretation}
med <- apply(model_input[, pred_names, drop = FALSE], 2, median, na.rm = TRUE)

vars_for_response <- intersect(c("bio1", "bio5", "bio6", "bio12"), pred_names)
if (length(vars_for_response) == 0) {
  vars_for_response <- head(pred_names, 4)
}

response_df <- do.call(rbind, lapply(vars_for_response, function(v) {
  xseq <- seq(min(model_input[[v]], na.rm = TRUE), max(model_input[[v]], na.rm = TRUE), length.out = 120)
  nd <- as.data.frame(matrix(rep(med, each = length(xseq)), nrow = length(xseq)))
  names(nd) <- pred_names
  nd[[v]] <- xseq
  yhat <- predict(mx, nd, type = "cloglog")
  data.frame(variable = v, x = xseq, suitability = yhat)
}))

ggplot(response_df, aes(x = x, y = suitability)) +
  geom_line(color = "midnightblue") +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Partial response curves", x = "Predictor value", y = "Predicted suitability")

# Permutation importance estimated from spatial CV folds.
perm_imp_list <- lapply(pred_names, function(v) {
  fold_drops <- lapply(seq_len(k_folds), function(fold) {
    test_idx <- which(fold_id == fold)
    test_df <- model_input[test_idx, , drop = FALSE]
    if (nrow(test_df) == 0 || length(unique(test_df$presence)) < 2) {
      return(NA_real_)
    }
    x_test <- test_df[, pred_names, drop = FALSE]
    y_test <- test_df$presence

    base_pred <- predict(cv_models[[fold]], x_test, type = "cloglog")
    auc_base <- safe_auc(y_test, base_pred)

    x_perm <- x_test
    x_perm[[v]] <- sample(x_perm[[v]])
    perm_pred <- predict(cv_models[[fold]], x_perm, type = "cloglog")
    auc_perm <- safe_auc(y_test, perm_pred)

    auc_base - auc_perm
  })
  data.frame(variable = v, importance = mean(unlist(fold_drops), na.rm = TRUE))
})

perm_imp <- bind_rows(perm_imp_list) %>% arrange(desc(importance))

knitr::kable(head(perm_imp, 10), digits = 4, caption = "Top permutation importance (spatial CV AUC drop)")

ggplot(perm_imp, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Permutation importance", x = "Climate variable", y = "AUC drop")
```

Interpretation guidance:

- response curves are conditional and correlational, not causal effects
- importance reflects predictive contribution in this calibration setup
- uncertainty and novelty maps should temper strong biogeographic claims
- forecast interpretation is strongest where novelty is low and uncertainty is low

## 11. Future range projections (all SSP pathways, 3 random GCMs)

```{r future-projections, message=TRUE, warning=FALSE}
ssp_levels <- c("126", "245", "370", "585")
future_time <- "2041-2060"
future_res <- 10

available_gcms <- get(".cmods", envir = asNamespace("geodata"))
set.seed(2026)
selected_gcms <- sample(available_gcms, 3)

selected_gcms

future_pred_list <- list()
future_novelty_list <- list()
future_summary <- list()

for (gcm in selected_gcms) {
  for (ssp in ssp_levels) {
    message("Projecting: ", gcm, " | SSP", ssp, " | ", future_time)

    fut <- geodata::cmip6_world(
      model = gcm,
      ssp = ssp,
      time = future_time,
      var = "bio",
      res = future_res,
      path = "worldclim_data"
    )

    fut_ca <- terra::crop(fut, ca_bbox)
    fut_ca <- harmonize_bioc_names(fut_ca)

    missing_preds <- setdiff(pred_names, names(fut_ca))
    if (length(missing_preds) > 0) {
      stop(sprintf(
        "Missing predictors for %s SSP%s: %s",
        gcm, ssp, paste(missing_preds, collapse = ", ")
      ))
    }

    fut_ca <- fut_ca[[pred_names]]
    fut_novelty <- novelty_fraction(fut_ca, model_input, pred_names)
    fut_ca_clamped <- clamp_to_training_range(fut_ca, model_input, pred_names)

    fut_pred <- raster::predict(
      raster::stack(fut_ca_clamped),
      mx,
      type = "cloglog",
      progress = "text"
    )

    key <- paste(gcm, paste0("SSP", ssp), sep = "__")
    future_pred_list[[key]] <- fut_pred
    future_novelty_list[[key]] <- fut_novelty

    suit_bin <- fut_pred > thr
    vals <- raster::values(fut_pred)
    bin_vals <- raster::values(suit_bin)
    nov_vals <- terra::values(fut_novelty)

    future_summary[[key]] <- data.frame(
      gcm = gcm,
      scenario = paste0("RCP/SSP", ssp),
      period = future_time,
      mean_suitability = mean(vals, na.rm = TRUE),
      suitable_fraction = mean(bin_vals == 1, na.rm = TRUE),
      novelty_fraction = mean(nov_vals > 0, na.rm = TRUE),
      high_novelty_fraction = mean(nov_vals > 0.2, na.rm = TRUE)
    )
  }
}

future_summary_df <- bind_rows(future_summary)

current_vals <- raster::values(pred_ca)
current_bin_vals <- raster::values(bin_map)
current_novelty_vals <- terra::values(novelty_current)

current_baseline <- data.frame(
  gcm = "Current climate",
  scenario = "Current",
  period = "1970-2000",
  mean_suitability = mean(current_vals, na.rm = TRUE),
  suitable_fraction = mean(current_bin_vals == 1, na.rm = TRUE),
  novelty_fraction = mean(current_novelty_vals > 0, na.rm = TRUE),
  high_novelty_fraction = mean(current_novelty_vals > 0.2, na.rm = TRUE)
)

future_summary_w_baseline <- bind_rows(current_baseline, future_summary_df)

knitr::kable(
  future_summary_w_baseline,
  digits = 4,
  caption = "Current and future suitability with novelty diagnostics"
)
```

## 12. Ensemble future maps and transferability signals

```{r future-ensemble-maps}
ensemble_by_ssp <- lapply(ssp_levels, function(ssp) {
  keys <- grep(paste0("__SSP", ssp, "$"), names(future_pred_list), value = TRUE)
  rs <- future_pred_list[keys]
  raster::calc(raster::stack(rs), fun = mean, na.rm = TRUE)
})
names(ensemble_by_ssp) <- paste0("RCP/SSP", ssp_levels)

ensemble_df <- bind_rows(lapply(names(ensemble_by_ssp), function(scn) {
  r <- ensemble_by_ssp[[scn]]
  d <- as.data.frame(raster::rasterToPoints(r), stringsAsFactors = FALSE)
  colnames(d) <- c("x", "y", "suitability")
  d$scenario <- scn
  d
}))

ggplot(ensemble_df, aes(x = x, y = y, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Suitability") +
  coord_quickmap() +
  facet_wrap(~ scenario, ncol = 2) +
  labs(
    title = sprintf("Future projected suitability (%s)", future_time),
    subtitle = "Ensemble mean across 3 random GCMs; predictors clamped to calibration range"
  )

future_uncertainty <- future_summary_df %>%
  group_by(scenario) %>%
  summarize(
    mean_suitable_fraction = mean(suitable_fraction, na.rm = TRUE),
    sd_suitable_fraction = sd(suitable_fraction, na.rm = TRUE),
    mean_novelty_fraction = mean(novelty_fraction, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(future_uncertainty, digits = 4, caption = "Across-GCM uncertainty and novelty by scenario")

ggplot(future_summary_df, aes(x = scenario, y = suitable_fraction, fill = gcm)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Suitable-area fraction by scenario and GCM",
    x = "Scenario",
    y = "Fraction of California predicted suitable"
  )
```

## 13. Save outputs

```{r save}
raster::writeRaster(pred_ca, filename = "california_poppy_suitability.tif", overwrite = TRUE)
raster::writeRaster(bin_map, filename = "california_poppy_suitability_binary.tif", overwrite = TRUE)
write.csv(diag_tbl, "sdm_diagnostics.csv", row.names = FALSE)
write.csv(perm_imp, "sdm_permutation_importance.csv", row.names = FALSE)
write.csv(future_summary_w_baseline, "sdm_future_projections_summary.csv", row.names = FALSE)

for (scn in names(ensemble_by_ssp)) {
  out_name <- paste0(
    "california_poppy_future_ensemble_",
    gsub("[^0-9]", "", scn),
    "_",
    gsub("-", "", future_time),
    ".tif"
  )
  raster::writeRaster(ensemble_by_ssp[[scn]], filename = out_name, overwrite = TRUE)
}

cat("Saved outputs:\n")
cat("- california_poppy_suitability.tif\n")
cat("- california_poppy_suitability_binary.tif\n")
cat("- sdm_diagnostics.csv\n")
cat("- sdm_permutation_importance.csv\n")
cat("- sdm_future_projections_summary.csv\n")
cat("- california_poppy_future_ensemble_126_20412060.tif\n")
cat("- california_poppy_future_ensemble_245_20412060.tif\n")
cat("- california_poppy_future_ensemble_370_20412060.tif\n")
cat("- california_poppy_future_ensemble_585_20412060.tif\n")
```

## 14. Session info

```{r session-info}
sessionInfo()
```

