wr---
title: "California Poppy (Eschscholzia californica) Species Distribution Model"
author: "GitHub Copilot"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: cerulean
    code_folding: show
---

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c(
  "dplyr", "terra", "raster", "maxnet", "ggplot2", "BIEN",
  "knitr", "pROC", "viridis", "geodata"
)

to_install <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(to_install) > 0) {
  install.packages(to_install, repos = "https://cran.rstudio.com")
}

library(dplyr)
library(terra)
library(raster)
library(maxnet)
library(ggplot2)
library(BIEN)
library(knitr)
library(pROC)
library(viridis)
library(geodata)

set.seed(123)
```

## 1. Occurrence data from BIEN

```{r occ-data}
species_name <- "Eschscholzia californica"

occ <- BIEN::BIEN_occurrence_species(species_name)

if (!all(c("latitude", "longitude") %in% names(occ))) {
  stop("BIEN output does not include latitude/longitude columns.")
}

occ <- occ %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  distinct(latitude, longitude, .keep_all = TRUE)

nrow(occ)
head(occ)
```

## 2. Define California extent and crop occurrences

```{r ca-extent}
ca_bbox <- terra::ext(-125, -114, 32, 42)

occ_ca <- occ %>%
  filter(longitude >= -125, longitude <= -114, latitude >= 32, latitude <= 42)

if (nrow(occ_ca) < 30) {
  stop("Too few California occurrences after filtering; cannot build a stable SDM.")
}

nrow(occ_ca)
```

## 3. WorldClim climate predictors (v2.1, 2.5 arc-min)

```{r climate}
wc <- geodata::worldclim_global(var = "bio", res = 2.5, path = "worldclim_data")
wc_ca <- terra::crop(wc, ca_bbox)

pred_names <- names(wc_ca)
pred_names
```

## 4. Background points (pseudo-absences)

```{r background}
bg_pts <- terra::spatSample(
  wc_ca[[1]], size = 10000, method = "random",
  na.rm = TRUE, xy = TRUE, values = FALSE
)
bg_df <- data.frame(x = bg_pts[, "x"], y = bg_pts[, "y"])

pres_df <- data.frame(x = occ_ca$longitude, y = occ_ca$latitude)

plot_df <- bind_rows(
  data.frame(x = pres_df$x, y = pres_df$y, type = "Presence"),
  data.frame(x = bg_df$x, y = bg_df$y, type = "Background")
)

ggplot(plot_df, aes(x, y, color = type)) +
  geom_point(size = 0.3, alpha = 0.5) +
  coord_quickmap(xlim = c(-125, -114), ylim = c(32, 42)) +
  scale_color_manual(values = c("Presence" = "red3", "Background" = "gray50")) +
  labs(title = "Sampling points for California poppy SDM", x = "Longitude", y = "Latitude")
```

## 5. Prepare presence-background data for maxnet

```{r pb-data}
pres_vals_raw <- terra::extract(wc_ca, as.matrix(pres_df))
bg_vals_raw <- terra::extract(wc_ca, as.matrix(bg_df))

pres_vals <- as.data.frame(pres_vals_raw)[, intersect(pred_names, names(pres_vals_raw)), drop = FALSE]
bg_vals <- as.data.frame(bg_vals_raw)[, intersect(pred_names, names(bg_vals_raw)), drop = FALSE]

pres_dat <- cbind(pres_df, presence = 1, pres_vals)
bg_dat <- cbind(bg_df, presence = 0, bg_vals)

model_input <- bind_rows(pres_dat, bg_dat) %>%
  filter(if_all(all_of(pred_names), ~ !is.na(.x)))

head(model_input)
dim(model_input)
```

## 6. Train/test split and maxnet fit

```{r maxnet-model}
pres_idx <- which(model_input$presence == 1)
bg_idx <- which(model_input$presence == 0)

train_pres <- sample(pres_idx, size = floor(0.8 * length(pres_idx)))
train_bg <- sample(bg_idx, size = floor(0.8 * length(bg_idx)))
train_idx <- c(train_pres, train_bg)
test_idx <- setdiff(seq_len(nrow(model_input)), train_idx)

train_df <- model_input[train_idx, , drop = FALSE]
test_df <- model_input[test_idx, , drop = FALSE]

x_train <- train_df[, pred_names, drop = FALSE]
p_train <- train_df$presence

mx <- maxnet::maxnet(
  p = p_train,
  data = x_train,
  f = maxnet.formula(p_train, x_train, classes = "lqhp")
)

mx
```

## 7. Standard SDM diagnostics on held-out test data

```{r diagnostics}
x_test <- test_df[, pred_names, drop = FALSE]
test_labels <- test_df$presence
test_score <- predict(mx, x_test, type = "cloglog")

roc_obj <- pROC::roc(response = test_labels, predictor = test_score, quiet = TRUE)
auc_val <- as.numeric(pROC::auc(roc_obj))

youden <- pROC::coords(roc_obj, x = "best", best.method = "youden", ret = c("threshold", "sensitivity", "specificity"))
thr <- as.numeric(youden["threshold"])

pred_class <- ifelse(test_score >= thr, 1, 0)
cm <- table(
  Truth = factor(test_labels, levels = c(0, 1)),
  Pred = factor(pred_class, levels = c(0, 1))
)

tn <- cm[1, 1]
fp <- cm[1, 2]
fn <- cm[2, 1]
tp <- cm[2, 2]

sensitivity <- tp / (tp + fn)
specificity <- tn / (tn + fp)
accuracy <- (tp + tn) / sum(cm)
tss <- sensitivity + specificity - 1

diag_tbl <- data.frame(
  metric = c("AUC", "Threshold (Youden)", "Sensitivity", "Specificity", "Accuracy", "TSS"),
  value = c(auc_val, thr, sensitivity, specificity, accuracy, tss)
)

knitr::kable(diag_tbl, digits = 4, caption = "Held-out test diagnostics")

cm
```

## 8. ROC and score distribution plots

```{r roc-plot}
roc_df <- data.frame(
  fpr = 1 - roc_obj$specificities,
  tpr = roc_obj$sensitivities
)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "dodgerblue3", linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "gray50") +
  coord_equal() +
  labs(
    title = "ROC curve on held-out test set",
    subtitle = paste("AUC =", round(auc_val, 4)),
    x = "False Positive Rate",
    y = "True Positive Rate"
  )

score_df <- data.frame(score = test_score, class = factor(test_labels, levels = c(0, 1), labels = c("Background", "Presence")))

ggplot(score_df, aes(x = score, fill = class)) +
  geom_histogram(position = "identity", alpha = 0.45, bins = 40) +
  geom_vline(xintercept = thr, linetype = 2, color = "black") +
  scale_fill_manual(values = c("Background" = "gray60", "Presence" = "firebrick3")) +
  labs(title = "Predicted suitability distributions", x = "cloglog score", y = "Count")
```

## 9. Predict full-region suitability and maps

```{r predict}
pred_ca <- raster::predict(raster::stack(wc_ca), mx, type = "cloglog", progress = "text")

pred_df <- as.data.frame(raster::rasterToPoints(pred_ca), stringsAsFactors = FALSE)
colnames(pred_df) <- c("x", "y", "suitability")

ggplot(pred_df, aes(x = x, y = y, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(name = "Suitability") +
  geom_point(data = pres_df, aes(x = x, y = y), color = "red", size = 0.9, alpha = 0.6, inherit.aes = FALSE) +
  coord_quickmap() +
  labs(title = "Predicted suitability for California poppy", subtitle = "Maxnet model, cloglog output")

bin_map <- pred_ca > thr
bin_df <- as.data.frame(raster::rasterToPoints(bin_map), stringsAsFactors = FALSE)
colnames(bin_df) <- c("x", "y", "suitability")

ggplot(bin_df, aes(x = x, y = y, fill = factor(suitability))) +
  geom_raster() +
  scale_fill_manual(values = c("lightgray", "darkgreen"), name = "Suitable") +
  coord_quickmap() +
  labs(title = sprintf("Binary suitability map > %.3f", thr))
```

## 10. Response curves and permutation importance

```{r importance}
med <- apply(model_input[, pred_names, drop = FALSE], 2, median, na.rm = TRUE)

vars_for_response <- intersect(c("bio1", "bio5", "bio6", "bio12"), pred_names)
if (length(vars_for_response) == 0) {
  vars_for_response <- head(pred_names, 4)
}

response_df <- do.call(rbind, lapply(vars_for_response, function(v) {
  xseq <- seq(min(model_input[[v]], na.rm = TRUE), max(model_input[[v]], na.rm = TRUE), length.out = 100)
  nd <- as.data.frame(matrix(rep(med, each = length(xseq)), nrow = length(xseq)))
  names(nd) <- pred_names
  nd[[v]] <- xseq
  yhat <- predict(mx, nd, type = "cloglog")
  data.frame(variable = v, x = xseq, suitability = yhat)
}))

ggplot(response_df, aes(x = x, y = suitability)) +
  geom_line(color = "midnightblue") +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Response curves", x = "Predictor value", y = "Predicted suitability")

baseline_auc <- auc_val
perm_imp <- lapply(pred_names, function(v) {
  x_perm <- x_test
  x_perm[[v]] <- sample(x_perm[[v]])
  p_perm <- predict(mx, x_perm, type = "cloglog")
  auc_perm <- as.numeric(pROC::auc(pROC::roc(test_labels, p_perm, quiet = TRUE)))
  data.frame(variable = v, importance = baseline_auc - auc_perm)
})
perm_imp <- bind_rows(perm_imp) %>% arrange(desc(importance))

knitr::kable(head(perm_imp, 10), digits = 4, caption = "Top permutation importance (AUC drop)")

ggplot(perm_imp, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Permutation importance", x = "Climate variable", y = "AUC drop after permutation")
```

## 11. Save outputs

```{r save}
raster::writeRaster(pred_ca, filename = "california_poppy_suitability.tif", overwrite = TRUE)
raster::writeRaster(bin_map, filename = "california_poppy_suitability_binary.tif", overwrite = TRUE)
write.csv(diag_tbl, "sdm_diagnostics.csv", row.names = FALSE)
write.csv(perm_imp, "sdm_permutation_importance.csv", row.names = FALSE)

cat("Saved outputs:\n")
cat("- california_poppy_suitability.tif\n")
cat("- california_poppy_suitability_binary.tif\n")
cat("- sdm_diagnostics.csv\n")
cat("- sdm_permutation_importance.csv\n")
```

## 12. Session info

```{r session-info}
sessionInfo()
```
