---
title: "GlobalBodySize: Science Summary"
subtitle: "Cross-Taxon Animal Body Mass Database — Phase 1 Tier-1 Intake"
author: "GlobalBodySize Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 6
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  cache      = FALSE
)

## Reproducibility: record session at time of render
session_render_time <- format(Sys.time(), "%Y-%m-%d %H:%M:%S UTC", tz = "UTC")
```

```{r packages}
## Required packages — install if absent
required_pkgs <- c("ggplot2", "dplyr", "scales", "knitr", "kableExtra",
                   "data.table", "tidyr", "patchwork")
missing_pkgs  <- required_pkgs[!required_pkgs %in% rownames(installed.packages())]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs, repos = "https://cran.r-project.org/",
                   quiet = TRUE)
}
invisible(lapply(required_pkgs, library, character.only = TRUE))

## ggplot theme consistent with publication-quality figures (Telford standard)
theme_globalsize <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      panel.grid.minor  = element_blank(),
      strip.background  = element_rect(fill = "grey92", colour = "grey60"),
      strip.text        = element_text(face = "bold"),
      plot.title        = element_text(face = "bold", size = base_size + 2),
      plot.subtitle     = element_text(size = base_size, colour = "grey40"),
      plot.caption      = element_text(size = base_size - 2, colour = "grey50",
                                       hjust = 0)
    )
}
theme_set(theme_globalsize())

## Colour palette: one colour per taxonomic group, colour-blind aware
GROUP_COLOURS <- c(
  "mammal"    = "#E69F00",
  "mammalia"  = "#E69F00",
  "bird"      = "#56B4E9",
  "aves"      = "#56B4E9",
  "fish"      = "#009E73",
  "actinopterygii" = "#009E73",
  "amphibian" = "#CC79A7",
  "amphibia"  = "#CC79A7",
  "reptile"   = "#D55E00",
  "reptilia"  = "#D55E00"
)
```

---

# 1. Project Overview

The **GlobalBodySize** project assembles a reproducible, provenance-rich, cross-taxonomic body mass database from authoritative curated sources. Body mass is the central ecological trait — it determines metabolic rate (Kleiber's law: $B \propto M^{3/4}$), home range, population density, generation time, extinction risk, trophic position, and biogeographic distribution (Peters 1983; Brown & Maurer 1989; West, Brown & Enquist 1997).

This document summarises the **Phase 1 Tier-1 intake**: six provider databases merged into a single flat file, covering mammals, birds, fish, and amphibians, with GBIF Backbone taxonomic reconciliation applied to all names.

**Data file used for this summary:** `data/compiled/tier1_reconciled.csv`

> **Honest uncertainty notice (Telford standard):** Statistics in this document are descriptive summaries of the current intake. The dataset is incomplete (reptiles critically underrepresented; amphibians sparse; AVONET birds not yet merged). No inferential claims should be drawn from provider-level comparisons without accounting for the different taxonomic scopes and mass methodologies of each source.

---

# 2. Data Loading and Basic Inventory

```{r load-data}
## Prefer reconciled file (includes GBIF taxonomy columns)
## Fall back to combined if reconciled not yet available
rec_file  <- "data/compiled/tier1_reconciled.csv"
base_file <- "data/compiled/tier1_combined.csv"

if (file.exists(rec_file)) {
  dat <- data.table::fread(rec_file, data.table = FALSE)
  using_reconciled <- TRUE
} else if (file.exists(base_file)) {
  dat <- data.table::fread(base_file, data.table = FALSE)
  using_reconciled <- FALSE
  warning("Using unreconciled tier1_combined.csv — reconciled file not found.")
} else {
  stop("Neither tier1_reconciled.csv nor tier1_combined.csv found. ",
       "Run scripts/merge_tier1.R first.")
}

## Coerce mass_g to numeric
dat$mass_g      <- suppressWarnings(as.numeric(dat$mass_g))
dat$log10_mass  <- log10(dat$mass_g)

## Normalise taxonomic group labels to title case for display
dat$group_display <- tools::toTitleCase(tolower(trimws(dat$input_taxonomic_group)))

## Flag mass type for display
dat$mass_type_display <- ifelse(
  grepl("LW_model", dat$mass_type, ignore.case = TRUE), "LW-modeled (FishBase)",
  tools::toTitleCase(tolower(trimws(dat$mass_type)))
)

cat(sprintf("Loaded: %d rows, %d columns\n", nrow(dat), ncol(dat)))
cat(sprintf("Using reconciled taxonomy: %s\n", using_reconciled))
cat(sprintf("Rendered: %s\n", session_render_time))
```

```{r provider-table}
## Provider-level summary
prov_summary <- dat |>
  dplyr::filter(!is.na(mass_g) & mass_g > 0) |>
  dplyr::group_by(source_display_name, group_display) |>
  dplyr::summarise(
    n_rows        = dplyr::n(),
    n_species     = dplyr::n_distinct(verbatim_taxon_name),
    mass_min_g    = min(mass_g, na.rm = TRUE),
    mass_max_g    = max(mass_g, na.rm = TRUE),
    mass_median_g = median(mass_g, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(group_display, dplyr::desc(n_rows))

prov_summary |>
  dplyr::mutate(
    mass_min_g    = round(mass_min_g, 2),
    mass_max_g    = scales::comma(round(mass_max_g, 0)),
    mass_median_g = round(mass_median_g, 1)
  ) |>
  knitr::kable(
    col.names = c("Source", "Group", "Rows", "Species",
                  "Min mass (g)", "Max mass (g)", "Median mass (g)"),
    caption   = "Table 1. Provider summary — all rows with positive mass_g."
  )
```

**Total rows in analysis: `r scales::comma(nrow(dat[!is.na(dat$mass_g) & dat$mass_g > 0, ]))`**

```{r species-per-group-bars, fig.height=5, fig.cap="Fig. 0a. Number of species per taxonomic group: total rows (left bar) vs. unique species after deduplication by GBIF accepted name (right bar, or verbatim name if reconciliation unavailable). The gap between the two bars reflects multi-source overlap (e.g., mammals appear in both PanTHERIA and EltonTraits)."}
## Build dedup name column (same logic used later in section 3b)
dedup_col_inv <- if ("accepted_name" %in% names(dat) &&
                      sum(!is.na(dat$accepted_name)) > nrow(dat) * 0.5)
                   "accepted_name" else "verbatim_taxon_name"

## Per-group counts: all rows vs. unique species
group_counts <- dat |>
  dplyr::filter(!is.na(mass_g), mass_g > 0) |>
  dplyr::group_by(group_display) |>
  dplyr::summarise(
    total_rows      = dplyr::n(),
    unique_species  = dplyr::n_distinct(.data[[dedup_col_inv]], na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(total_rows)) |>
  tidyr::pivot_longer(cols = c(total_rows, unique_species),
                      names_to = "count_type", values_to = "n") |>
  dplyr::mutate(
    count_label = dplyr::recode(count_type,
      total_rows     = "Total source rows",
      unique_species = paste0("Unique species\n(dedup by ", dedup_col_inv, ")")
    )
  )

ggplot(group_counts, aes(x = reorder(group_display, -n), y = n,
                          fill = count_label)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68,
           colour = "grey30", linewidth = 0.25, alpha = 0.88) +
  geom_text(aes(label = scales::comma(n)),
            position = position_dodge(width = 0.75),
            vjust = -0.4, size = 3.1, colour = "grey20") +
  scale_fill_manual(
    values = c("Total source rows"  = "#2c7bb6",
               "Unique species\n(dedup by accepted_name)" = "#abd9e9",
               "Unique species\n(dedup by verbatim_taxon_name)" = "#abd9e9"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12)),
                     name = "Count") +
  labs(
    title    = "Number of species per taxonomic group",
    subtitle = "Blue = total records from all sources; light = unique species after deduplication",
    caption  = paste0(
      "Note: mammals appear in multiple providers (PanTHERIA + EltonTraits + AnAge), ",
      "so total rows >> unique species.\n",
      "Deduplication key: ", dedup_col_inv, "."
    ),
    x = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r species-per-group-by-source, fig.height=5, fig.cap="Fig. 0b. Number of species (unique verbatim names) contributed by each source database, stacked by taxonomic group. Shows the relative contribution and overlap of each provider. PanTHERIA, EltonTraits mammals, and AnAge all overlap for many mammal species — the stacked total exceeds true unique mammal diversity."}
source_group_counts <- dat |>
  dplyr::filter(!is.na(mass_g), mass_g > 0) |>
  dplyr::mutate(
    source_short = dplyr::case_when(
      grepl("PanTHERIA",   source_display_name) ~ "PanTHERIA",
      grepl("EltonTraits.*bird|EltonTraits.*Bir", source_display_name) ~ "EltonTraits\n(birds)",
      grepl("EltonTraits.*mam|EltonTraits.*Mam", source_display_name) ~ "EltonTraits\n(mammals)",
      grepl("EltonTraits",  source_display_name) ~ "EltonTraits",
      grepl("AnAge",        source_display_name) ~ "AnAge",
      grepl("AmphiBIO",     source_display_name) ~ "AmphiBIO",
      grepl("FishBase",     source_display_name) ~ "FishBase",
      grepl("AVONET",       source_display_name) ~ "AVONET",
      TRUE                                        ~ source_display_name
    )
  ) |>
  dplyr::group_by(source_short, group_display) |>
  dplyr::summarise(
    n_species = dplyr::n_distinct(verbatim_taxon_name, na.rm = TRUE),
    .groups   = "drop"
  )

## Order sources by total descending
source_order <- source_group_counts |>
  dplyr::group_by(source_short) |>
  dplyr::summarise(tot = sum(n_species)) |>
  dplyr::arrange(dplyr::desc(tot)) |>
  dplyr::pull(source_short)

source_group_counts$source_short <- factor(source_group_counts$source_short,
                                            levels = source_order)

ggplot(source_group_counts, aes(x = source_short, y = n_species,
                                 fill = group_display)) +
  geom_col(colour = "grey30", linewidth = 0.25, alpha = 0.88, width = 0.72) +
  geom_text(aes(label = scales::comma(n_species)),
            position = position_stack(vjust = 0.5),
            size = 3, colour = "white", fontface = "bold") +
  scale_fill_manual(values = GROUP_COLOURS, name = "Taxonomic group") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.05)),
                     name = "Unique species (verbatim name)") +
  labs(
    title    = "Species contributed per source, broken down by taxonomic group",
    subtitle = "Stacked bars = group contribution within each source database",
    caption  = "Note: species shared across multiple sources are counted once per source — stacked totals reflect provider scope, not unique global diversity.",
    x        = NULL
  ) +
  theme(legend.position = "right",
        axis.text.x = element_text(size = 9))
```

```{r species-per-group-dedup-only, fig.height=4, fig.cap="Fig. 0c. Best-estimate unique species count per taxonomic group after deduplication. This is the most ecologically meaningful inventory statistic — it shows how many distinct species have at least one body mass record, independent of how many databases contributed data."}
## Unique species after dedup — the 'truth' for inventory purposes
unique_by_group <- dat |>
  dplyr::filter(!is.na(mass_g), mass_g > 0) |>
  dplyr::group_by(group_display) |>
  dplyr::summarise(
    unique_species = dplyr::n_distinct(.data[[dedup_col_inv]], na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(unique_species))

ggplot(unique_by_group,
       aes(x = reorder(group_display, unique_species), y = unique_species,
           fill = group_display)) +
  geom_col(colour = "grey30", linewidth = 0.3, alpha = 0.88, width = 0.65) +
  geom_text(aes(label = scales::comma(unique_species)),
            hjust = -0.15, size = 3.5, colour = "grey20") +
  scale_fill_manual(values = GROUP_COLOURS, guide = "none") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15)),
                     name = "Unique species") +
  coord_flip() +
  labs(
    title    = paste0("Unique species with body mass data — all sources combined"),
    subtitle = paste0("Deduplication key: ", dedup_col_inv,
                      " | Total unique species: ",
                      scales::comma(sum(unique_by_group$unique_species))),
    caption  = "Cross-source deduplication collapses multi-provider species to single counts.",
    x        = NULL
  )
```

---

# 3. Body Size Distributions by Taxonomic Group

Body mass distributions are examined on a log₁₀ scale. Empirical distributions within taxonomic groups typically approximate log-normal, reflecting multiplicative growth processes and diversification dynamics (Brown & Maurer 1989; Dial et al. 2008).

```{r dist-main, fig.height=8, fig.cap="Fig. 1. Log₁₀ body mass distributions by taxonomic group. Vertical dashed line = group median. Note: 3,443 FishBase rows carry LW-modeled mass (not directly measured) — shown in lighter shading."}
plot_dat <- dat |>
  dplyr::filter(!is.na(log10_mass) & is.finite(log10_mass)) |>
  dplyr::mutate(
    group_f = factor(group_display,
                     levels = c("Mammalia", "Aves", "Actinopterygii",
                                "Amphibia", "Reptilia", "Mammal", "Bird",
                                "Fish", "Amphibian", "Reptile")),
    is_modeled = grepl("LW_model|LW-model", mass_type, ignore.case = TRUE)
  )

## Compute group medians for reference lines
group_medians <- plot_dat |>
  dplyr::group_by(group_display) |>
  dplyr::summarise(med_log10 = median(log10_mass, na.rm = TRUE), .groups = "drop")

ggplot(plot_dat, aes(x = log10_mass, fill = is_modeled)) +
  geom_histogram(binwidth = 0.15, colour = "white", linewidth = 0.2, alpha = 0.85) +
  geom_vline(data = group_medians, aes(xintercept = med_log10),
             linetype = "dashed", colour = "grey20", linewidth = 0.7) +
  scale_fill_manual(
    values = c(`FALSE` = "#2c7bb6", `TRUE` = "#abd9e9"),
    labels = c(`FALSE` = "Directly measured / literature mean",
               `TRUE`  = "LW-modeled (FishBase only)"),
    name   = "Mass source"
  ) +
  scale_x_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = -2:8,
    labels = function(x) paste0("10^", x)
  ) +
  scale_y_continuous(name = "Number of species") +
  facet_wrap(~ group_display, scales = "free_y", ncol = 2) +
  labs(
    title    = "Body mass distributions across taxonomic groups",
    subtitle = "Dashed line = group median; log₁₀ scale",
    caption  = paste0(
      "Data: tier1_combined.csv | FishBase LW-modeled values in light blue | ",
      "n = ", scales::comma(nrow(plot_dat)), " species-source rows"
    )
  ) +
  theme(legend.position = "bottom")
```

```{r dist-violin, fig.height=5, fig.cap="Fig. 2. Violin + box plot of log₁₀ body mass by taxonomic group. Dot = group median. Width proportional to density."}
## Only groups with >20 records get a meaningful violin
groups_for_violin <- plot_dat |>
  dplyr::count(group_display) |>
  dplyr::filter(n >= 20) |>
  dplyr::pull(group_display)

violin_dat <- plot_dat |>
  dplyr::filter(group_display %in% groups_for_violin,
                !is_modeled)  # wet/direct only for fair comparison

group_medians_violin <- violin_dat |>
  dplyr::group_by(group_display) |>
  dplyr::summarise(med = median(log10_mass, na.rm = TRUE), .groups = "drop")

ggplot(violin_dat, aes(x = reorder(group_display, log10_mass, FUN = median),
                       y = log10_mass,
                       fill = group_display)) +
  geom_violin(trim = TRUE, alpha = 0.7, colour = "grey30", linewidth = 0.3) +
  geom_boxplot(width = 0.12, outlier.size = 0.5, outlier.alpha = 0.3,
               fill = "white", colour = "grey30") +
  geom_point(data = group_medians_violin,
             aes(x = group_display, y = med),
             inherit.aes = FALSE,
             size = 2.5, colour = "black", shape = 21, fill = "white") +
  scale_fill_manual(values = GROUP_COLOURS, guide = "none") +
  scale_y_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = -2:8,
    labels = function(x) paste0("10^", x)
  ) +
  labs(
    title    = "Body mass range by taxonomic group (directly measured / literature mean only)",
    subtitle = "LW-modeled FishBase mass excluded for comparability",
    caption  = "Hollow dot = median; box = IQR; violin = density",
    x        = NULL
  )
```

---

# 3b. Combined Body Size Distribution — All Datasets

```{r combined-all-histogram, fig.height=5, fig.cap="Fig. 3a. Frequency distribution of log₁₀ body mass across all datasets combined, stacked by taxonomic group. Each bar represents the total number of species-source records in that body size bin. The overall distribution shows the macroecological 'right skew of life' — most animal species are small, with a long tail of large-bodied forms."}
ggplot(plot_dat, aes(x = log10_mass, fill = group_display)) +
  geom_histogram(binwidth = 0.15, colour = "white", linewidth = 0.15, alpha = 0.88,
                 position = "stack") +
  geom_vline(xintercept = median(plot_dat$log10_mass, na.rm = TRUE),
             linetype = "dashed", colour = "grey20", linewidth = 0.8) +
  annotate("text",
           x     = median(plot_dat$log10_mass, na.rm = TRUE) + 0.07,
           y     = Inf,
           label = paste0("Overall median\n10^",
                          round(median(plot_dat$log10_mass, na.rm = TRUE), 2), " g"),
           hjust = 0, vjust = 1.4, size = 3.2, colour = "grey20") +
  scale_fill_manual(values = GROUP_COLOURS, name = "Taxonomic group") +
  scale_x_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = -2:9,
    labels = function(x) paste0("10^", x)
  ) +
  scale_y_continuous(name = "Number of records", labels = scales::comma) +
  labs(
    title    = "Body mass frequency distribution — all datasets combined",
    subtitle = paste0("All sources, all mass types | n = ",
                      scales::comma(nrow(plot_dat)), " species-source records"),
    caption  = "Dashed line = overall median. LW-modeled FishBase mass included (see mass_type column)."
  ) +
  theme(legend.position = "right")
```

```{r combined-all-density, fig.height=5, fig.cap="Fig. 3b. Density-normalised view of all datasets combined, with per-group overlaid curves. This shows the shape of the distribution independently of sample size differences between groups."}
ggplot(plot_dat, aes(x = log10_mass)) +
  geom_histogram(aes(y = after_stat(density), fill = group_display),
                 binwidth = 0.15, colour = "white", linewidth = 0.15,
                 alpha = 0.7, position = "stack") +
  geom_density(aes(colour = group_display), linewidth = 0.9, adjust = 1.2) +
  scale_fill_manual(values = GROUP_COLOURS, name = "Group (bars)") +
  scale_colour_manual(values = GROUP_COLOURS, name = "Group (density)") +
  scale_x_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = -2:9,
    labels = function(x) paste0("10^", x)
  ) +
  scale_y_continuous(name = "Density") +
  labs(
    title    = "Body mass density — all datasets combined, per-group curves overlaid",
    subtitle = "Stacked bars normalised to density; coloured curves = per-group kernel density estimates",
    caption  = "All mass types included."
  ) +
  theme(legend.position = "right")
```

```{r species-by-bodysize, fig.height=6, fig.cap="Fig. 3c. Number of unique species (deduplicated by GBIF accepted name, or verbatim name if reconciliation unavailable) as a function of body size bin (0.15 decade bins). This represents the body size frequency distribution of animal diversity independent of record duplication. Coloured by taxonomic group."}
## Deduplicate: one row per unique accepted species, taking median log10 mass
## Use accepted_name if available (from GBIF reconciliation), else verbatim_taxon_name
dedup_name_col <- if ("accepted_name" %in% names(dat) &&
                       sum(!is.na(dat$accepted_name)) > nrow(dat) * 0.5)
                    "accepted_name" else "verbatim_taxon_name"

species_dat <- dat |>
  dplyr::filter(!is.na(log10_mass), is.finite(log10_mass)) |>
  dplyr::group_by(.data[[dedup_name_col]], group_display) |>
  dplyr::summarise(
    log10_mass = median(log10_mass, na.rm = TRUE),
    .groups    = "drop"
  )

n_unique_spp <- nrow(species_dat)

## Bin into 0.15-decade bins
bin_width <- 0.15
species_binned <- species_dat |>
  dplyr::mutate(
    bin_floor = floor(log10_mass / bin_width) * bin_width,
    bin_mid   = bin_floor + bin_width / 2
  ) |>
  dplyr::count(bin_mid, group_display)

## Overall median across unique species
overall_spp_median <- median(species_dat$log10_mass, na.rm = TRUE)

ggplot(species_binned, aes(x = bin_mid, y = n, fill = group_display)) +
  geom_col(colour = "white", linewidth = 0.15, alpha = 0.88,
           position = "stack", width = bin_width * 0.92) +
  geom_vline(xintercept = overall_spp_median,
             linetype = "dashed", colour = "grey20", linewidth = 0.8) +
  annotate("text",
           x     = overall_spp_median + 0.08,
           y     = Inf,
           label = paste0("Median unique species\n10^",
                          round(overall_spp_median, 2), " g"),
           hjust = 0, vjust = 1.4, size = 3.2, colour = "grey20") +
  scale_fill_manual(values = GROUP_COLOURS, name = "Taxonomic group") +
  scale_x_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = seq(-2, 9, by = 1),
    labels = function(x) paste0("10^", x)
  ) +
  scale_y_continuous(name = "Number of unique species", labels = scales::comma) +
  labs(
    title    = "Number of species as a function of body size",
    subtitle = paste0(
      scales::comma(n_unique_spp), " unique species | deduplicated on ",
      ifelse(dedup_name_col == "accepted_name", "GBIF accepted name", "verbatim taxon name"),
      " | median mass per species used"
    ),
    caption  = paste0(
      "Bin width = 0.15 log₁₀ decades. Deduplication collapses multi-source species to one row.\n",
      "Stacked bars show group contribution to each body size bin."
    )
  ) +
  theme(legend.position = "right")
```

```{r species-by-bodysize-facet, fig.height=7, fig.cap="Fig. 3d. Per-group view of unique species as a function of body size. Same deduplication as Fig. 3c — each group's body size distribution shown on a free y-axis to reveal within-group structure independently of the large between-group richness differences."}
ggplot(species_binned, aes(x = bin_mid, y = n, fill = group_display)) +
  geom_col(colour = "white", linewidth = 0.15, alpha = 0.85, width = bin_width * 0.92) +
  geom_vline(
    data = species_dat |>
      dplyr::group_by(group_display) |>
      dplyr::summarise(med = median(log10_mass, na.rm = TRUE), .groups = "drop"),
    aes(xintercept = med), linetype = "dashed", colour = "grey20", linewidth = 0.6
  ) +
  scale_fill_manual(values = GROUP_COLOURS, guide = "none") +
  scale_x_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = seq(-2, 9, by = 2),
    labels = function(x) paste0("10^", x)
  ) +
  scale_y_continuous(name = "Number of unique species", labels = scales::comma) +
  facet_wrap(~ group_display, scales = "free_y", ncol = 2) +
  labs(
    title    = "Number of unique species by body size — per taxonomic group",
    subtitle = "Dashed line = group median | Free y-axis per panel",
    caption  = "Bin width = 0.15 log₁₀ decades."
  )
```

---

# 4. Cross-Provider Mass Comparison

```{r provider-mass-comparison, fig.height=6, fig.cap="Fig. 3. Body mass distribution per provider source. Useful for detecting systematic offsets between databases. Overlapping distributions indicate reasonable inter-source agreement; divergence warrants investigation."}
provider_plot_dat <- dat |>
  dplyr::filter(!is.na(log10_mass), is.finite(log10_mass)) |>
  dplyr::mutate(
    ## Shorten source labels for plot
    source_short = dplyr::case_when(
      grepl("PanTHERIA",   source_display_name) ~ "PanTHERIA\n(mammals)",
      grepl("EltonTraits", source_display_name) ~ "EltonTraits\n(birds+mammals)",
      grepl("AnAge",       source_display_name) ~ "AnAge\n(multi-taxon)",
      grepl("AmphiBIO",    source_display_name) ~ "AmphiBIO\n(amphibians)",
      grepl("FishBase",    source_display_name) ~ "FishBase\n(fish)",
      grepl("AVONET",      source_display_name) ~ "AVONET\n(birds)",
      TRUE                                       ~ source_display_name
    )
  )

ggplot(provider_plot_dat,
       aes(x = log10_mass, colour = source_short, fill = source_short)) +
  geom_density(alpha = 0.2, linewidth = 0.8) +
  scale_x_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = -2:8
  ) +
  scale_y_continuous(name = "Density") +
  scale_colour_brewer(palette = "Dark2", name = "Source") +
  scale_fill_brewer(palette = "Dark2", name = "Source") +
  labs(
    title    = "Body mass density by source database",
    subtitle = "Density curves allow inter-source comparison within overlapping taxonomic groups",
    caption  = "All mass types included. Use with caution: sources differ in taxonomic scope."
  ) +
  theme(legend.position = "right")
```

---

# 5. Mass Type and Data Quality Composition

```{r mass-type-breakdown, fig.height=4, fig.cap="Fig. 4. Composition of mass_type field. LW-modeled entries are FishBase only. All other sources contribute directly measured or literature-mean values. Analyses mixing mass types require explicit sensitivity testing."}
mt_summary <- dat |>
  dplyr::filter(!is.na(mass_g), mass_g > 0) |>
  dplyr::count(mass_type_display, group_display) |>
  dplyr::arrange(dplyr::desc(n))

ggplot(mt_summary, aes(x = reorder(mass_type_display, n), y = n, fill = group_display)) +
  geom_col(colour = "grey30", linewidth = 0.2) +
  scale_fill_manual(values = GROUP_COLOURS, name = "Group") +
  scale_y_continuous(labels = scales::comma, name = "Number of rows") +
  coord_flip() +
  labs(
    title    = "Mass type composition by taxonomic group",
    subtitle = "Critical: LW-modeled mass must not be mixed with wet mass without stratification",
    caption  = "mass_type = 'LW_modeled' entries are FishBase L-W equation estimates only",
    x        = NULL
  )
```

```{r confidence-breakdown}
## Confidence breakdown
conf_tbl <- dat |>
  dplyr::filter(!is.na(mass_g), mass_g > 0) |>
  dplyr::count(mass_confidence, source_display_name) |>
  tidyr::pivot_wider(names_from = mass_confidence, values_from = n, values_fill = 0)

conf_tbl |>
  knitr::kable(
    caption = "Table 2. Row counts by mass_confidence flag and source. 'high' = directly measured literature mean; 'medium' = LW-modeled or derived."
  )
```

---

# 6. Taxonomic Reconciliation Summary

```{r taxon-reconciliation}
if (using_reconciled && "match_type" %in% names(dat)) {
  ## Reconciliation match type breakdown
  match_summary <- dat |>
    dplyr::count(match_type, match_confidence) |>
    dplyr::arrange(dplyr::desc(n))

  match_summary |>
    knitr::kable(
      col.names = c("GBIF match type", "Match confidence", "Rows"),
      caption   = paste0(
        "Table 3. GBIF Backbone reconciliation results (version: ",
        if ("backbone_version" %in% names(dat)) dat$backbone_version[1] else "unknown",
        "). EXACT matches have high confidence that the verbatim name resolves unambiguously."
      )
    )
} else {
  cat("Reconciled data not available in this file. Run scripts/run_taxon_reconciliation.R.\n")
}
```

```{r synonym-rate}
if (using_reconciled && "matched_status" %in% names(dat)) {
  syn_rate <- mean(grepl("SYNONYM", dat$matched_status, ignore.case = TRUE), na.rm = TRUE)
  n_syn    <- sum(grepl("SYNONYM", dat$matched_status, ignore.case = TRUE), na.rm = TRUE)
  n_none   <- sum(dat$matched_status %in% c("NONE", "error"), na.rm = TRUE)

  cat(sprintf(
    "Synonym rate: %.1f%% (%d of %d rows carry a name that is a GBIF synonym)\n",
    100 * syn_rate, n_syn, nrow(dat)
  ))
  cat(sprintf("Unresolved names: %d rows (%.2f%%) — require manual review\n",
              n_none, 100 * n_none / nrow(dat)))
  cat(sprintf("\nNote: Synonym resolution does not change the analysis data — verbatim\n"))
  cat(sprintf("source names are preserved. accepted_name column carries the GBIF-preferred name.\n"))
}
```

```{r class-breakdown}
if (using_reconciled && "class" %in% names(dat)) {
  class_summary <- dat |>
    dplyr::filter(!is.na(class), nzchar(class), !is.na(mass_g), mass_g > 0) |>
    dplyr::count(class) |>
    dplyr::arrange(dplyr::desc(n))

  class_summary |>
    knitr::kable(
      col.names = c("GBIF-matched class", "Rows"),
      caption   = "Table 4. Row counts by GBIF-reconciled class. Reveals the actual taxonomic breadth of the dataset beyond the coarse input_taxonomic_group labels."
    )
}
```

---

# 7. Body Size Scaling Laws and Theoretical Context

This section contextualises the data against established macroecological theory, following the analytical framework of West, Brown & Enquist (1997) and Peters (1983).

## 7.1 The metabolic scaling prediction: $B \propto M^{3/4}$

The West-Brown-Enquist (WBE) theory derives the 3/4-power metabolic scaling law from the geometry of space-filling, impedance-minimizing fractal vascular networks. The prediction is:

$$B = B_0 \cdot M^{3/4}$$

where $B$ is whole-organism metabolic rate (watts), $M$ is body mass (kg), and $B_0$ is a normalization constant that varies by body temperature and taxonomic group. This dataset provides the $M$ axis for any cross-taxon test of this scaling law. Metabolic rate data must be joined from external sources (e.g., Savage et al. 2004 for mammals and birds; Clarke & Johnston 1999 for fish).

The scaling exponent $b = 3/4$ implies that per-unit-mass metabolic rate scales as $M^{-1/4}$ — larger animals are proportionally more metabolically efficient. This has direct consequences for the 7 other ecological pathways:

| Ecological variable | Predicted scaling | Exponent |
|---|---|---|
| Metabolic rate $B$ | $\propto M^{3/4}$ | 0.75 |
| Lifespan $T$ | $\propto M^{1/4}$ | 0.25 |
| Home range $A$ | $\propto M^{1.0}$ | 1.0 (mammals) |
| Population density $D$ | $\propto M^{-3/4}$ | -0.75 |
| Generation time $G$ | $\propto M^{1/4}$ | 0.25 |
| Heart rate | $\propto M^{-1/4}$ | -0.25 |
| Genome replication time | $\propto M^{1/4}$ | 0.25 |

*Exponents from Peters (1983) and West et al. (1997). All are UNVERIFIED estimates — confirm against primary literature before applying in specific taxonomic groups.*

## 7.2 Body size frequency distributions: the "right skew of life"

```{r log-normal-test, fig.height=5, fig.cap="Fig. 5. QQ-plot of log₁₀ body mass against theoretical normal distribution, by taxonomic group. Points on the line indicate log-normality. Systematic curvature indicates departures (e.g., right skew = upper tail heavier than log-normal; log-skew from body size constraints)."}
## QQ plot by group — Telford: diagnostics before interpretation
qq_dat <- dat |>
  dplyr::filter(!is.na(log10_mass), is.finite(log10_mass),
                !grepl("LW_model", mass_type, ignore.case = TRUE)) |>
  dplyr::group_by(group_display) |>
  dplyr::filter(dplyr::n() >= 50) |>  # only groups with ≥50 wet/direct rows
  dplyr::mutate(
    theoretical = {
      n <- dplyr::n()
      stats::qnorm(seq(0.5/n, 1 - 0.5/n, length.out = n))
    },
    sample_quantile = sort(log10_mass)
  ) |>
  dplyr::ungroup()

ggplot(qq_dat, aes(x = theoretical, y = sample_quantile, colour = group_display)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, colour = "black", linewidth = 0.7,
              linetype = "dashed") +
  scale_colour_manual(values = GROUP_COLOURS, guide = "none") +
  facet_wrap(~ group_display, scales = "free") +
  labs(
    title    = expression("QQ-plot: log"[10]*" body mass vs. theoretical normal"),
    subtitle = "LW-modeled FishBase mass excluded | Groups with ≥50 direct measurements shown",
    caption  = "Dashed line = perfect log-normality. Systematic departures indicate skew or multimodality.",
    x        = "Theoretical normal quantiles",
    y        = expression("log"[10]*" body mass (g) — sample quantiles")
  )
```

```{r summary-stats-by-group}
## Summary statistics — what enhanced-theory would want to see
scale_stats <- dat |>
  dplyr::filter(!is.na(log10_mass), is.finite(log10_mass)) |>
  dplyr::group_by(group_display) |>
  dplyr::summarise(
    n              = dplyr::n(),
    min_g          = min(mass_g, na.rm = TRUE),
    max_g          = max(mass_g, na.rm = TRUE),
    median_g       = median(mass_g, na.rm = TRUE),
    mean_log10     = mean(log10_mass, na.rm = TRUE),
    sd_log10       = sd(log10_mass, na.rm = TRUE),
    range_decades  = diff(range(log10_mass, na.rm = TRUE)),
    .groups        = "drop"
  )

scale_stats |>
  dplyr::mutate(
    min_g      = round(min_g, 2),
    max_g      = scales::comma(round(max_g, 0)),
    median_g   = scales::comma(round(median_g, 1)),
    mean_log10 = round(mean_log10, 2),
    sd_log10   = round(sd_log10, 2),
    range_decades = round(range_decades, 1)
  ) |>
  knitr::kable(
    col.names = c("Group", "n", "Min (g)", "Max (g)", "Median (g)",
                  "Mean log₁₀", "SD log₁₀", "Range (decades)"),
    caption   = "Table 5. Body mass summary statistics by taxonomic group. 'Range (decades)' = orders of magnitude spanned by group."
  )
```

## 7.3 Cross-group mass scaling context

```{r cross-group-scaling, fig.height=5, fig.cap="Fig. 6. Cumulative body size distribution (ECDF) on log₁₀ scale for each taxonomic group. Horizontal offset between curves reflects differences in modal body size. The width of each curve reflects the range of sizes within the group. For reference, dashed vertical lines mark iconic body sizes."}
ecdf_dat <- dat |>
  dplyr::filter(!is.na(log10_mass), is.finite(log10_mass),
                !grepl("LW_model", mass_type, ignore.case = TRUE)) |>
  dplyr::group_by(group_display) |>
  dplyr::filter(dplyr::n() >= 50) |>
  dplyr::ungroup()

## Reference masses (log10 g)
reference_masses <- data.frame(
  label = c("Hummingbird\n~3g", "House mouse\n~20g",
            "Mallard\n~1,200g", "African lion\n~180,000g",
            "Blue whale\n~150,000,000g"),
  log10_g = c(log10(3), log10(20), log10(1200),
               log10(180000), log10(1.5e8))
)

ggplot(ecdf_dat, aes(x = log10_mass, colour = group_display)) +
  stat_ecdf(linewidth = 1, pad = FALSE) +
  geom_vline(data = reference_masses, aes(xintercept = log10_g),
             linetype = "dotted", colour = "grey50", linewidth = 0.5) +
  geom_text(data = reference_masses, aes(x = log10_g, y = 0.08, label = label),
            inherit.aes = FALSE, size = 2.8, angle = 90, hjust = 0,
            vjust = -0.2, colour = "grey40") +
  scale_colour_manual(values = GROUP_COLOURS, name = "Taxonomic group") +
  scale_x_continuous(
    name   = expression(log[10]~"body mass (g)"),
    breaks = -2:9
  ) +
  scale_y_continuous(name = "Cumulative proportion", labels = scales::percent) +
  labs(
    title    = "Cumulative body mass distribution by taxonomic group",
    subtitle = "LW-modeled FishBase mass excluded | Groups with ≥50 directly measured rows",
    caption  = "Dotted lines = iconic reference body masses"
  )
```

---

# 8. Coverage and Gap Analysis

```{r gap-analysis}
## Known approximate species richness per group (UNVERIFIED — approximate for context only)
known_richness <- data.frame(
  group_display = c("Mammalia", "Aves", "Actinopterygii",
                    "Amphibia", "Reptilia"),
  approx_total_species = c(6600, 10900, 35000, 8500, 10900),
  richness_source = c("MDD v1.12", "BirdLife 2024", "FishBase 2024",
                      "ASW 2024", "Reptile Database 2024"),
  richness_verified = c(FALSE, FALSE, FALSE, FALSE, FALSE),
  stringsAsFactors = FALSE
)

coverage_dat <- scale_stats |>
  dplyr::left_join(known_richness, by = "group_display") |>
  dplyr::mutate(
    pct_covered = ifelse(!is.na(approx_total_species),
                         round(100 * n / approx_total_species, 1),
                         NA_real_)
  )

coverage_dat |>
  dplyr::select(group_display, n, approx_total_species, pct_covered, richness_source, richness_verified) |>
  dplyr::mutate(
    approx_total_species = scales::comma(approx_total_species),
    pct_covered = ifelse(is.na(pct_covered), "—",
                         paste0(pct_covered, "%"))
  ) |>
  knitr::kable(
    col.names = c("Group", "Rows in DB", "Known species (approx)",
                  "Est. coverage", "Richness source", "Count verified?"),
    caption   = paste0(
      "Table 6. Estimated taxonomic coverage. ",
      "CAUTION: 'Known species' counts are approximate and UNVERIFIED here — ",
      "confirm from authoritative checklists before publication. ",
      "Coverage % assumes one row per species (may overcount multi-source species)."
    )
  )
```

> **Coverage caveat:** Rows ≠ unique species. PanTHERIA, EltonTraits, and AnAge overlap substantially for mammals — many species appear in multiple providers. True unique species counts require deduplication on the GBIF accepted name, which should be done after synonym resolution. The coverage percentages above are upper bounds.

---

# 9. Darwin Core Compliance Status

Following the `BIODIVERSITY_INFORMATICS_AUDIT.md` recommendations:

```{r dwc-compliance}
## Check which DwC-critical fields are present and populated
dwc_critical <- data.frame(
  DwC_term        = c("measurementID", "basisOfRecord", "occurrenceID",
                       "kingdom", "phylum", "class", "order", "family",
                       "measurementDeterminedDate", "measurementAccuracy",
                       "measurementRemarks", "nameAccordingTo"),
  internal_field  = c("(missing)", "basis_of_record", "(missing)",
                       "kingdom", "phylum", "class", "order", "family",
                       "(missing)", "(missing)", "qa_note", "(missing)"),
  status          = c("ABSENT — required for MoF", "PRESENT", "ABSENT — required for specimen records",
                       "PRESENT (from reconciliation)", "PRESENT (from reconciliation)",
                       "PRESENT (from reconciliation)", "PRESENT (from reconciliation)",
                       "PRESENT (from reconciliation)", "ABSENT — add from source metadata",
                       "ABSENT — add SE where available", "PARTIAL (qa_note)", "ABSENT — add backbone version"),
  priority        = c("HIGH", "HIGH", "MEDIUM", "HIGH", "HIGH", "HIGH", "HIGH", "HIGH",
                       "MEDIUM", "MEDIUM", "LOW", "MEDIUM"),
  stringsAsFactors = FALSE
)

dwc_critical |>
  knitr::kable(
    caption = "Table 7. Darwin Core compliance status for MeasurementOrFact extension fields. ABSENT fields must be added before GBIF/OBIS submission."
  )
```

```{r basisofrecord-check}
## Check what basisOfRecord values are in use
if ("basis_of_record" %in% names(dat)) {
  bor_counts <- table(dat$basis_of_record, useNA = "always")
  bor_df <- as.data.frame(bor_counts)
  names(bor_df) <- c("basis_of_record", "n")
  knitr::kable(bor_df,
               caption = "Table 8. basisOfRecord values in use. Values must be from DwC controlled vocabulary.")
}
```

---

# 10. Known Limitations and Priority Next Steps

## Critical limitations (must address before any publication-grade analysis)

1. **LW-modeled mass must never be mixed with wet mass** without explicit stratification or sensitivity analysis. Always filter on `mass_type` before cross-taxon comparisons.

2. **Reptile coverage is essentially absent** (16 rows, all from AnAge). Do not include Reptilia in any scaling analysis without adding a dedicated reptile source (Meiri et al., Reptile Database).

3. **Amphibian mass is sparse** (591 of ~8,500 species). This coverage is sufficient for macroecological pattern analyses but not for species-level applications.

4. **AVONET bird data (11,009 rows) is not yet in tier1_combined.csv.** Add to `scripts/merge_tier1.R` PROVIDERS list and re-run to obtain the most complete bird mass dataset.

5. **Mammals contain duplicate species across PanTHERIA and EltonTraits.** Deduplication by GBIF `accepted_name` is needed before unique species counts or richness-based analyses.

6. **Darwin Core `measurementID` is not populated.** This blocks submission to GBIF MoF or OBIS.

7. **Zenodo discovery is incomplete.** Re-run of the 83-term discovery is in progress; `candidate_datasets.csv` currently reflects Dryad + Figshare only.

## Priority next steps

| Priority | Action | Blocker? |
|---|---|---|
| 1 | Add AVONET to tier1_combined.csv merge | No |
| 2 | Deduplicate species across providers by GBIF accepted_name | No |
| 3 | Run NEON small mammal intake | Requires neonUtilities install |
| 4 | Add VertNet museum specimens | Requires spocc API |
| 5 | Add Reptile Database / Meiri reptile traits | Requires download |
| 6 | Populate measurementID for DwC compliance | Design decision needed |
| 7 | Run renv::snapshot() to pin R environment | No |
| 8 | Submit to GBIF IPT or Zenodo for public archiving | After DwC compliance |

---

# 11. Session Information (Reproducibility)

```{r session-info}
cat("Rendered at:", session_render_time, "\n\n")
cat("Data file:", ifelse(using_reconciled, rec_file, base_file), "\n")
cat("GBIF Backbone version:", 
    if (using_reconciled && "backbone_version" %in% names(dat)) 
      dat$backbone_version[1] else "not available", "\n\n")
sessionInfo()
```

---

## References

- Brown JH, Maurer BA. 1989. Macroecology: The Division of Food and Space Among Species on Continents. *Science* 243(4895):1145–1150. DOI: 10.1126/science.243.4895.1145

- Clarke A, Johnston NM. 1999. Scaling of metabolic rate with body mass and temperature in teleost fish. *Journal of Animal Ecology* 68(5):893–905. DOI: 10.1046/j.1365-2656.1999.00337.x

- de Magalhães JP, Costa J. 2009. A database of vertebrate longevity records and their relation to other life-history traits. *Journal of Evolutionary Biology* 22(8):1770–1774. DOI: 10.1111/j.1420-9101.2009.01783.x

- Froese R, Pauly D (eds). FishBase. World Wide Web electronic publication. www.fishbase.org. Accessed 2026-05-10.

- Jones KE et al. 2009. PanTHERIA: a species-level database of life history, ecology, and geography of extant and recently extinct mammals. *Ecology* 90(9):2648. DOI: 10.1890/08-1494.1

- Oliveira BF et al. 2017. AmphiBIO, a global database for amphibian ecological traits. *Scientific Data* 4:160123. DOI: 10.1038/sdata.2016.123

- Peters RH. 1983. *The Ecological Implications of Body Size*. Cambridge University Press, Cambridge, UK. ISBN: 0-521-28886-5.

- Savage VM et al. 2004. The predominance of quarter-power scaling in biology. *Functional Ecology* 18(2):257–282. DOI: 10.1111/j.0269-8463.2004.00856.x

- Tobias JA et al. 2022. AVONET: morphological, ecological and geographical data for all birds. *Ecology Letters* 25(3):581–597. DOI: 10.1111/ele.13898

- West GB, Brown JH, Enquist BJ. 1997. A general model for the origin of allometric scaling laws in biology. *Science* 276(5309):122–126. DOI: 10.1126/science.276.5309.122

- Wilman H et al. 2014. EltonTraits 1.0: Species-level foraging attributes of the world's birds and mammals. *Ecology* 95(7):2027. DOI: 10.1890/13-1917.1
