---
title: "Plant Trait Harvest — Detailed Cross-Provider Report"
subtitle: "DataDryad · Zenodo · Scientific Data (AusTraits)"
author: "DryadPlantTraits Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE}
message(Sys.time(), " - loaded [setup]")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(knitr)
})
if (requireNamespace("kableExtra", quietly = TRUE)) library(kableExtra)
if (requireNamespace("scales",     quietly = TRUE)) library(scales)
if (requireNamespace("leaflet",    quietly = TRUE)) library(leaflet)
if (requireNamespace("DT",         quietly = TRUE)) library(DT)

# -- Project root detection
find_dryad_root <- function(start = getwd()) {
  path <- normalizePath(start, mustWork = FALSE)
  for (i in seq_len(12)) {
    if (file.exists(file.path(path, "R", "dryad_api.R"))) return(path)
    parent <- dirname(path)
    if (parent == path) break
    path <- parent
  }
  normalizePath(file.path(knitr::current_input(dir = TRUE), ".."),
                mustWork = FALSE)
}
proj_root <- find_dryad_root()

# -- kable helper
kable_styled <- function(df, caption = NULL) {
  df_fmt <- as.data.frame(df, stringsAsFactors = FALSE)
  num_cols <- vapply(df_fmt, is.numeric, logical(1))
  if (any(num_cols)) {
    df_fmt[num_cols] <- lapply(df_fmt[num_cols], function(x) {
      out <- format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
      out[is.na(x)] <- NA_character_
      out
    })
  }
  tbl <- knitr::kable(df_fmt, caption = caption, format = "html")
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    tryCatch(
      kableExtra::kable_styling(tbl,
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE),
      error = function(e) tbl
    )
  } else {
    tbl
  }
}

# -- Provider paths
paths <- list(
  dryad     = file.path(proj_root, "output/compiled_trait_observations_with_unit_inference.csv"),
  zenodo    = file.path(proj_root, "output/providers/zenodo/compiled_trait_observations_with_unit_inference.csv"),
  scidata   = file.path(proj_root, "output/providers/scientific_data/compiled_trait_observations_with_unit_inference.csv"),
  traithub  = file.path(proj_root, "output/providers/traithub/compiled_trait_observations.csv")
)
provider_labels <- c(dryad = "DataDryad", zenodo = "Zenodo", scidata = "Scientific Data", traithub = "TraitHub (tundra)")
provider_colors <- c(DataDryad = "#1f77b4", Zenodo = "#ff7f0e", `Scientific Data` = "#2ca02c", `TraitHub (tundra)` = "#9467bd")

for (.k in names(paths)) {
  if (!file.exists(paths[[.k]])) {
    stop("Missing provider file: ", provider_labels[[.k]],
         "\n  Expected: ", paths[[.k]])
  }
}

read_provider <- function(path, key) {
  d <- fread(path, data.table = TRUE)
  d[, provider := provider_labels[[key]]]
  d
}

dat_list <- lapply(names(paths), function(k) read_provider(paths[[k]], k))
names(dat_list) <- names(paths)

# -- Combined data table
all_dat <- rbindlist(dat_list, use.names = TRUE, fill = TRUE)

# -- Coordinate validity
all_dat[, lat_num := suppressWarnings(as.numeric(latitude))]
all_dat[, lon_num := suppressWarnings(as.numeric(longitude))]
all_dat[, has_coord := is.finite(lat_num) & is.finite(lon_num) &
                    lat_num >= -90 & lat_num <= 90 &
                    lon_num >= -180 & lon_num <= 180]

for (k in names(dat_list)) {
  dat_list[[k]][, lat_num := suppressWarnings(as.numeric(latitude))]
  dat_list[[k]][, lon_num := suppressWarnings(as.numeric(longitude))]
  dat_list[[k]][, has_coord := is.finite(lat_num) & is.finite(lon_num) &
                          lat_num >= -90 & lat_num <= 90 &
                          lon_num >= -180 & lon_num <= 180]
}
```

---

## 1 Executive Summary {#executive-summary}

```{r exec-summary}
message(Sys.time(), " - loaded [exec-summary]")

total_rows   <- nrow(all_dat)
total_traits <- uniqueN(all_dat[!is.na(trait_name) & nzchar(trait_name), trait_name])
total_taxa   <- uniqueN(all_dat[!is.na(scrubbed_species_binomial) & nzchar(scrubbed_species_binomial),
                                scrubbed_species_binomial])
total_coords <- sum(all_dat$has_coord, na.rm = TRUE)
```

<div class="alert alert-info" role="alert">
<h4 class="alert-heading">Key Takeaways</h4>
<ul>
  <li><strong>`r formatC(total_rows, format="d", big.mark=",")`</strong> total observation rows across <strong>3 providers</strong> (DataDryad, Zenodo, Scientific Data) as of `r Sys.Date()`.</li>
  <li><strong>`r formatC(total_traits, format="d", big.mark=",")`</strong> distinct trait names detected.</li>
  <li><strong>`r formatC(total_taxa, format="d", big.mark=",")`</strong> distinct taxa (scrubbed species binomials).</li>
  <li><strong>`r formatC(total_coords, format="d", big.mark=",")`</strong> rows with valid geographic coordinates (DataDryad + Zenodo only; Scientific Data does not carry coordinates in this extract).</li>
  <li><em>Note on Scientific Data / AusTraits unit confidence:</em> Rows from the Scientific Data provider show <code>inferred_unit_confidence = "none"</code> because the simple-schema mapper for this provider did not populate the <code>basis_type</code> or <code>value_type</code> metadata columns that the confidence scorer uses as inputs. AusTraits is a fully curated, peer-reviewed database (Falster et al. 2021, <em>Sci. Data</em>, doi:10.1038/s41597-021-01006-6). Its units were set by original data contributors and validated during publication. The "none" label reflects a <strong>pipeline labeling gap</strong>, not a unit quality failure.</li>
  <li><em>Cross-provider deduplication has not been applied.</em> The same measurement may appear in more than one provider if included in multiple source compilations.</li>
</ul>
</div>

---

## 2 Provider Overview {#provider-overview}

```{r provider-overview}
message(Sys.time(), " - loaded [provider-overview]")

provider_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  list(
    Provider             = provider_labels[[k]],
    `Observation Rows`   = nrow(d),
    `Distinct Traits`    = uniqueN(d[!is.na(trait_name) & nzchar(trait_name), trait_name]),
    `Distinct Taxa`      = uniqueN(d[!is.na(scrubbed_species_binomial) & nzchar(scrubbed_species_binomial),
                                     scrubbed_species_binomial]),
    `Source Datasets`    = uniqueN(d$source_title),
    `Rows with Coords`   = sum(d$has_coord, na.rm = TRUE),
    `Rows with QA Flags` = sum(nzchar(trimws(d$qa_flags)), na.rm = TRUE)
  )
}))

kable_styled(provider_summary, caption = "Provider-level harvest summary")
```

### 2.1 Observation rows by provider {#rows-by-provider}

```{r fig-provider-rows, fig.height=4}
prov_plot_df <- data.frame(
  Provider = provider_summary$Provider,
  Rows     = provider_summary$`Observation Rows`
)
ggplot(prov_plot_df, aes(x = reorder(Provider, -Rows), y = Rows / 1e3, fill = Provider)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = formatC(Rows, format = "d", big.mark = ",")),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = provider_colors) +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Observation rows by provider") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 2.2 Trait count by provider (stacked, top 20) {#trait-stacked}

```{r fig-trait-stacked, fig.height=5}
trait_by_prov <- all_dat[!is.na(trait_name) & nzchar(trait_name),
                         .(n = .N), by = .(trait_name, provider)]
top_traits_all <- all_dat[!is.na(trait_name) & nzchar(trait_name),
                           .(total = .N), by = trait_name][order(-total)]
top20_names <- head(top_traits_all$trait_name, 20L)
tbp20 <- trait_by_prov[trait_name %in% top20_names]
tbp20[, provider := factor(provider, levels = c("DataDryad", "Zenodo", "Scientific Data"))]
ggplot(tbp20, aes(x = reorder(trait_name, n, sum), y = n / 1e3, fill = provider)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(values = provider_colors, name = "Provider") +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Top 20 traits - stacked by provider") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

---

## 3 Source / Citation Breakdown {#source-citation}

```{r source-tables, results='asis'}
message(Sys.time(), " - loaded [source-citation]")

for (k in names(dat_list)) {
  d <- dat_list[[k]]
  src_tbl <- d[, .(
    Rows              = .N,
    `Distinct Traits` = uniqueN(trait_name),
    `Distinct Taxa`   = uniqueN(scrubbed_species_binomial),
    DOI               = dryad_dataset_doi[1L]
  ), by = source_title][order(-Rows)]
  setnames(src_tbl, "source_title", "Source Title")

  cat("\n### 3.", which(names(dat_list) == k), " ", provider_labels[[k]], "\n\n",
      sep = "")
  if (requireNamespace("DT", quietly = TRUE)) {
    print(DT::datatable(
      as.data.frame(src_tbl),
      filter = "top",
      extensions = "Scroller",
      options = list(scrollY = 300, scroller = TRUE, scrollX = TRUE, pageLength = 20),
      caption = paste(provider_labels[[k]], ": source datasets"),
      rownames = FALSE
    ))
  } else {
    print(kable_styled(head(src_tbl, 25L),
                       caption = paste(provider_labels[[k]], ": top 25 sources")))
  }
  cat("\n\n")
}
```

---

## 4 Trait Coverage - Detailed {#trait-coverage}

```{r trait-coverage-setup}
message(Sys.time(), " - loaded [trait-coverage]")
top_traits_tbl <- all_dat[!is.na(trait_name) & nzchar(trait_name),
  .(Rows = .N, `Distinct Taxa` = uniqueN(scrubbed_species_binomial)),
  by = trait_name][order(-Rows)]
```

### 4.1 Top 25 traits combined {#top25-traits}

```{r top25-traits}
kable_styled(
  head(top_traits_tbl, 25L)[, .(`Trait Name` = trait_name, Rows, `Distinct Taxa`)],
  caption = "Top 25 traits by row count (all providers combined)"
)
```

### 4.2 Top 20 traits - provider contribution {#top20-stacked}

```{r fig-top20-provider, fig.height=5}
ggplot(tbp20, aes(x = reorder(trait_name, n, sum), y = n / 1e3, fill = provider)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(values = provider_colors, name = "Provider") +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Top 20 traits - provider contribution") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

### 4.3 Provider-specific top 15 traits {#provider-top15}

```{r provider-top15, results='asis'}
for (k in names(dat_list)) {
  d <- dat_list[[k]]
  top15 <- d[!is.na(trait_name) & nzchar(trait_name),
    .(Rows = .N, `Distinct Taxa` = uniqueN(scrubbed_species_binomial)),
    by = trait_name][order(-Rows)]
  cat("\n#### ", provider_labels[[k]], "\n\n", sep = "")
  if (requireNamespace("DT", quietly = TRUE)) {
    print(DT::datatable(
      as.data.frame(head(top15, 15L))[, c("trait_name", "Rows", "Distinct Taxa")],
      extensions = "Scroller",
      options = list(scrollY = 280, scroller = TRUE, scrollX = TRUE, dom = "t"),
      colnames = c("Trait Name", "Rows", "Distinct Taxa"),
      rownames = FALSE
    ))
  } else {
    print(kable_styled(
      head(top15, 15L)[, .(`Trait Name` = trait_name, Rows, `Distinct Taxa`)],
      caption = paste(provider_labels[[k]], ": top 15 traits")
    ))
  }
  cat("\n\n")
}
```

### 4.4 Trait value summary statistics (top 20 numeric traits) {#trait-stats}

```{r trait-value-stats}
message(Sys.time(), " - loaded [trait-value-stats]")

all_dat[, val_num := suppressWarnings(as.numeric(trait_value))]

numeric_counts <- all_dat[is.finite(val_num) & !is.na(trait_name) & nzchar(trait_name),
  .(n_numeric = .N), by = trait_name][order(-n_numeric)]
top20_numeric <- head(numeric_counts[n_numeric >= 100], 20L)$trait_name

stats_list <- lapply(top20_numeric, function(tr) {
  vals <- all_dat[trait_name == tr & is.finite(val_num), val_num]
  list(
    `Trait Name` = tr,
    n            = length(vals),
    Mean         = round(mean(vals), 4),
    Median       = round(median(vals), 4),
    SD           = round(sd(vals), 4),
    Min          = round(min(vals), 4),
    P5           = round(quantile(vals, 0.05), 4),
    P25          = round(quantile(vals, 0.25), 4),
    P75          = round(quantile(vals, 0.75), 4),
    P95          = round(quantile(vals, 0.95), 4),
    Max          = round(max(vals), 4)
  )
})
stats_tbl <- rbindlist(stats_list)
kable_styled(stats_tbl,
  caption = "Summary statistics for top 20 numeric traits (rows with finite numeric trait_value only)")
```

### 4.5 Trait value boxplots (top 12 numeric traits) {#trait-boxplots}

```{r fig-trait-boxplots, fig.height=7, fig.width=10}
message(Sys.time(), " - loaded [trait-boxplots]")

top12_numeric <- head(top20_numeric, 12L)
box_dat <- all_dat[trait_name %in% top12_numeric & is.finite(val_num),
                   .(trait_name, val_num, provider)]

box_dat_clipped <- box_dat[, {
  p1  <- quantile(val_num, 0.01)
  p99 <- quantile(val_num, 0.99)
  .SD[val_num >= p1 & val_num <= p99]
}, by = trait_name]

box_dat_clipped[, provider := factor(provider,
  levels = c("DataDryad", "Zenodo", "Scientific Data"))]

ggplot(box_dat_clipped, aes(x = provider, y = val_num, fill = provider)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3) +
  facet_wrap(~ trait_name, scales = "free_y", ncol = 3) +
  scale_y_continuous(trans = "pseudo_log") +
  scale_fill_manual(values = provider_colors, name = "Provider") +
  labs(x = NULL, y = "Trait value (pseudo-log scale)",
       title = "Trait value distributions - top 12 numeric traits",
       caption = "Values clipped to 1st-99th percentile per trait for display. Pseudo-log y-axis.") +
  theme_classic(base_size = 10) +
  theme(plot.title   = element_text(face = "bold"),
        axis.text.x  = element_text(angle = 30, hjust = 1),
        legend.position = "bottom",
        strip.text   = element_text(size = 8))
```

---

## 5 Taxonomic Coverage {#taxonomic-coverage}

```{r taxonomic-setup}
message(Sys.time(), " - loaded [taxonomic-coverage]")
taxa_tbl <- all_dat[!is.na(scrubbed_species_binomial) & nzchar(scrubbed_species_binomial),
  .(Rows = .N, `Distinct Traits` = uniqueN(trait_name)),
  by = scrubbed_species_binomial][order(-Rows)]
```

### 5.1 Top 25 taxa by row count {#top25-taxa}

```{r top25-taxa}
kable_styled(
  head(taxa_tbl, 25L)[, .(
    `Species (scrubbed)` = scrubbed_species_binomial,
    Rows,
    `Distinct Traits`
  )],
  caption = "Top 25 taxa by row count (all providers)"
)
```

### 5.2 Distinct taxa per provider {#taxa-by-provider}

```{r fig-taxa-provider, fig.height=4}
taxa_prov <- data.table(
  Provider        = provider_summary$Provider,
  `Distinct Taxa` = provider_summary$`Distinct Taxa`
)
ggplot(taxa_prov, aes(x = reorder(Provider, -`Distinct Taxa`),
                      y = `Distinct Taxa` / 1e3, fill = Provider)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = formatC(`Distinct Taxa`, format = "d", big.mark = ",")),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = provider_colors) +
  labs(x = NULL, y = "Distinct taxa (thousands)",
       title = "Distinct taxa per provider") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 5.3 Country breakdown (top 20) {#country-breakdown}

```{r fig-country, fig.height=5}
country_tbl <- all_dat[,
  .(country = data.table::fifelse(
      is.na(country) | !nzchar(trimws(country)), "Unknown", country
    ))][, .(Rows = .N), by = country][order(-Rows)]
top20_ctry  <- head(country_tbl, 20L)
ggplot(top20_ctry, aes(x = reorder(country, Rows), y = Rows / 1e3)) +
  geom_col(fill = "#4e79a7") +
  coord_flip() +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Top 20 countries by row count") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

### 5.4 Date collected - year distribution {#date-histogram}

```{r fig-date-histogram, fig.height=4}
message(Sys.time(), " - loaded [date-histogram]")

if (requireNamespace("lubridate", quietly = TRUE)) {
  dates_raw <- all_dat[!is.na(date_collected) & nzchar(trimws(date_collected)), date_collected]
  parsed    <- lubridate::parse_date_time(
    dates_raw,
    orders = c("Ymd", "Y", "mdY", "dmy"),
    quiet  = TRUE
  )
  years <- lubridate::year(parsed)
  years <- years[is.finite(years) & years >= 1800 & years <= as.integer(format(Sys.Date(), "%Y"))]
  if (length(years) > 0) {
    year_df <- data.frame(year = years)
    print(
      ggplot(year_df, aes(x = year)) +
        geom_histogram(binwidth = 1, fill = "#59a14f", color = "white", linewidth = 0.2) +
        labs(x = "Year collected", y = "Records",
             title = "Records by year of collection") +
        theme_classic(base_size = 12) +
        theme(plot.title = element_text(face = "bold"))
    )
  } else {
    cat("> No parseable dates found in the data.\n")
  }
} else {
  cat("> **Note:** `lubridate` is not installed. Date histogram skipped.",
      "Install with `install.packages('lubridate')` and re-render.\n")
}
```

---

## 6 Unit Reconciliation Summary {#unit-reconciliation}

```{r unit-summary}
message(Sys.time(), " - loaded [unit-reconciliation]")

unit_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  iu  <- d$inferred_unit
  iuc <- trimws(as.character(d$inferred_unit_confidence))
  iuc[is.na(iuc) | !nzchar(iuc)] <- NA_character_
  list(
    Provider             = provider_labels[[k]],
    `Total Rows`         = nrow(d),
    `Inferred TRUE`      = sum(iu %in% c(TRUE, "TRUE",  "true",  1L), na.rm = TRUE),
    `Inferred FALSE`     = sum(iu %in% c(FALSE,"FALSE", "false", 0L), na.rm = TRUE),
    `Conf: high`         = sum(iuc == "high",   na.rm = TRUE),
    `Conf: medium`       = sum(iuc == "medium",  na.rm = TRUE),
    `Conf: low`          = sum(iuc == "low",     na.rm = TRUE),
    `Conf: none/missing` = sum(is.na(iuc))
  )
}))
kable_styled(unit_summary, caption = "6.1 Unit inference summary by provider")
```

### 6.2 inferred_unit_confidence breakdown {#confidence-breakdown}

```{r confidence-breakdown}
conf_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d   <- dat_list[[k]]
  lvl <- trimws(as.character(d$inferred_unit_confidence))
  lvl[is.na(lvl) | !nzchar(lvl)] <- "(missing)"
  tbl <- d[, .N, by = .(inferred_unit_confidence = lvl)][order(-N)]
  tbl[, provider := provider_labels[[k]]]
  tbl
}))
conf_wide <- dcast(conf_rows, inferred_unit_confidence ~ provider, value.var = "N", fill = 0L)
kable_styled(conf_wide, caption = "inferred_unit_confidence counts by provider (wide)")
```

### 6.3 basis_type breakdown {#basis-type}

```{r basis-type-breakdown}
bt_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d  <- dat_list[[k]]
  bt <- trimws(as.character(d$basis_type))
  bt[is.na(bt) | !nzchar(bt)] <- "(empty)"
  tbl <- d[, .N, by = .(basis_type = bt)][order(-N)]
  tbl[, provider := provider_labels[[k]]]
  tbl
}))
bt_wide <- dcast(bt_rows, basis_type ~ provider, value.var = "N", fill = 0L)
kable_styled(bt_wide, caption = "basis_type counts by provider (wide)")
```

### 6.4 Note on AusTraits / Scientific Data unit confidence {#austraits-note}

<div class="alert alert-warning" role="alert">
<strong>Pipeline labeling note - Scientific Data "none" unit confidence:</strong>
Scientific Data rows show <code>inferred_unit_confidence = "none"</code> because the simple-schema mapper used for this provider did not populate the <code>basis_type</code> or <code>value_type</code> metadata columns that the confidence scorer uses as inputs.
AusTraits (the dominant Scientific Data source) is a fully curated, peer-reviewed database (Falster et al. 2021, doi:10.1038/s41597-021-01006-6).
Its units were set by the original data contributors and validated during publication.
The <strong>"none" label reflects a pipeline labeling gap, not a unit quality failure.</strong>
</div>

---

## 7 Geographic Coverage - Interactive Map {#geographic-coverage}

```{r geo-coverage-table}
message(Sys.time(), " - loaded [geographic-coverage]")

geo_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  list(
    Provider           = provider_labels[[k]],
    `Total Rows`       = nrow(d),
    `Rows with Coords` = sum(d$has_coord, na.rm = TRUE),
    `Coverage (%)`     = round(100 * mean(d$has_coord, na.rm = TRUE), 1)
  )
}))
kable_styled(geo_summary, caption = "7.1 Geographic coordinate coverage by provider")
```

### 7.2 Interactive map (sampled) {#interactive-map}

```{r leaflet-map, fig.height=6}
message(Sys.time(), " - loaded [leaflet-map]")

if (requireNamespace("leaflet", quietly = TRUE)) {
  coord_dat <- all_dat[has_coord == TRUE,
    .(lat_num, lon_num, provider, scrubbed_species_binomial,
      trait_name, trait_value, unit, source_title, country)]

  total_georef <- nrow(coord_dat)

  if (total_georef == 0L) {
    cat(
      "> **Map note:** No valid coordinate rows are available after coordinate filtering. Displaying a base map with no points.\n\n"
    )
    leaflet::leaflet() |>
      leaflet::addTiles()
  } else {

    set.seed(42)
    samp_list <- lapply(unique(coord_dat$provider), function(prov) {
      sub <- coord_dat[provider == prov]
      if (nrow(sub) > 10000) sub <- sub[sample(.N, 10000)]
      sub
    })
    map_dat <- rbindlist(samp_list, fill = TRUE)
    plotted_n <- nrow(map_dat)

    if (plotted_n == 0L || !all(c("lat_num", "lon_num", "provider") %in% names(map_dat))) {
      cat(
        "> **Map note:** No mappable sampled points were produced after filtering/sampling. Displaying a base map with no points.",
        "Total georeferenced records:", formatC(total_georef, format = "d", big.mark = ","),
        "| Plotted points:", formatC(plotted_n, format = "d", big.mark = ","), "\n\n"
      )
      leaflet::leaflet() |>
        leaflet::addTiles()
    } else {

      esc <- function(x) htmltools::htmlEscape(ifelse(is.na(x), "", as.character(x)))
      map_dat[, popup_text := paste0(
        "<b>", esc(scrubbed_species_binomial), "</b><br>",
        ifelse(is.na(trait_name), "", paste0(esc(trait_name), ": ")),
        esc(trait_value), " ", esc(unit), "<br>",
        "Source: ", esc(source_title), "<br>",
        "Country: ", esc(country)
      )]

      map_providers <- unique(map_dat$provider)
      pal <- leaflet::colorFactor(
        palette = c("DataDryad" = "#1f77b4", "Zenodo" = "#ff7f0e",
                    "Scientific Data" = "#2ca02c"),
        domain  = c("DataDryad", "Zenodo", "Scientific Data"),
        na.color = "#999999"
      )

      map_out <- leaflet::leaflet(as.data.frame(map_dat)) |>
        leaflet::addProviderTiles("CartoDB.Positron",
          options = leaflet::providerTileOptions(noWrap = FALSE)) |>
        leaflet::addTiles(group = "OpenStreetMap") |>
        leaflet::addCircleMarkers(
          lng         = ~lon_num,
          lat         = ~lat_num,
          radius      = 5,
          color       = ~pal(provider),
          fillColor   = ~pal(provider),
          opacity     = 0.9,
          fillOpacity = 0.7,
          popup       = ~popup_text,
          group       = "Observations"
        ) |>
        leaflet::addLayersControl(
          baseGroups = c("CartoDB.Positron", "OpenStreetMap"),
          overlayGroups = "Observations",
          options = leaflet::layersControlOptions(collapsed = FALSE)
        ) |>
        leaflet::addLegend(
          position = "bottomright",
          pal      = pal,
          values   = ~provider,
          title    = "Provider"
        )

      cat("> **Map note:** Displaying a random sample of up to 10,000 records per provider for performance.",
        "Total georeferenced records:", formatC(total_georef, format = "d", big.mark = ","),
        "| Plotted points:", formatC(plotted_n, format = "d", big.mark = ","), "\n\n")

      map_out
    }
  }
} else {
  cat("> **Note:** `leaflet` is not installed. Interactive map skipped.",
      "Install with `install.packages('leaflet')` and re-render.\n")
}
```

### 7.3 Elevation distribution {#elevation}

```{r fig-elevation, fig.height=4}
elev_vals <- if ("elevation_m" %in% names(all_dat)) {
  suppressWarnings(as.numeric(all_dat$elevation_m))
} else {
  numeric(0)
}
elev_valid <- elev_vals[is.finite(elev_vals) & elev_vals >= -500 & elev_vals <= 6000]
if (length(elev_valid) > 0) {
  elev_df <- data.frame(elevation_m = elev_valid)
  ggplot(elev_df, aes(x = elevation_m)) +
    geom_histogram(bins = 60, fill = "#76b7b2", color = "white", linewidth = 0.2) +
    labs(x = "Elevation (m)", y = "Records",
         title = "Records by elevation",
         caption = "Filtered to -500 to 6000 m range.") +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(face = "bold"))
} else {
  cat("> No valid elevation values found in the range -500 to 6000 m.\n")
}
```

---

## 8 BIEN Database Mapping Status {#bien-mapping}

```{r bien-mapping-setup}
message(Sys.time(), " - loaded [bien-mapping]")

bien_trait_map <- data.table(
  trait_canonical = c(
    "leaf area", "specific leaf area", "leaf nitrogen content per leaf dry mass",
    "leaf phosphorus content per leaf dry mass", "stem specific density",
    "plant height vegetative", "leaf dry mass", "seed mass",
    "leaf carbon content per leaf dry mass", "maximum height vegetative",
    "wood density", "leaf thickness", "root depth",
    "stomatal conductance", "leaf water content per leaf dry mass",
    "leaf lifespan", "leaf chlorophyll content",
    "bark thickness", "vessel diameter", "conduit diameter lumen"
  ),
  bien_name = c(
    "leaf area", "specific leaf area", "leaf nitrogen content per leaf dry mass",
    "leaf phosphorus content per leaf dry mass", "stem specific density",
    "plant height vegetative", "leaf dry mass", "seed mass",
    "leaf carbon content per leaf dry mass", "maximum height vegetative",
    "wood density", "leaf thickness", "root depth",
    "stomatal conductance", "leaf water content per leaf dry mass",
    "leaf lifespan", "leaf chlorophyll content",
    "bark thickness", "vessel diameter", "conduit diameter lumen"
  )
)

all_dat[, trait_name_lower := trimws(tolower(trait_name))]
bien_set <- tolower(trimws(bien_trait_map$trait_canonical))
all_dat[, is_bien_mappable := trait_name_lower %in% bien_set]
```

### 8.1-8.2 BIEN-mappable rows per provider {#bien-table}

```{r bien-rows-table}
bien_prov <- all_dat[is_bien_mappable == TRUE,
  .(Rows = .N), by = .(bien_trait = trait_name_lower, provider)]
bien_wide <- dcast(bien_prov, bien_trait ~ provider, value.var = "Rows", fill = 0L)
bien_wide[, Total := rowSums(.SD, na.rm = TRUE),
          .SDcols = setdiff(names(bien_wide), "bien_trait")]
setorder(bien_wide, -Total)
setnames(bien_wide, "bien_trait", "BIEN Trait")
kable_styled(bien_wide, caption = "8.2 BIEN-mappable rows by trait and provider (exact canonical match)")
```

### 8.3 BIEN-mappable rows by provider {#bien-bar}

```{r fig-bien-bar, fig.height=4}
bien_summary <- all_dat[, .(
  Total         = .N,
  BIEN_mappable = sum(is_bien_mappable, na.rm = TRUE),
  Pct_mappable  = round(100 * mean(is_bien_mappable, na.rm = TRUE), 1)
), by = provider]

ggplot(bien_summary, aes(x = reorder(provider, -BIEN_mappable),
                         y = BIEN_mappable / 1e3, fill = provider)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = paste0(formatC(BIEN_mappable, format = "d", big.mark = ","),
                               "\n(", Pct_mappable, "%)")),
            vjust = -0.3, size = 3.2) +
  scale_fill_manual(values = provider_colors) +
  labs(x = NULL, y = "BIEN-mappable rows (thousands)",
       title = "8.3 BIEN-mappable rows by provider") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 8.4 Percentage of rows that are BIEN-mappable {#bien-pct}

```{r bien-pct-table}
kable_styled(
  bien_summary[, .(Provider = provider,
                   `Total Rows` = Total,
                   `BIEN-Mappable Rows` = BIEN_mappable,
                   `% Mappable` = Pct_mappable)],
  caption = "8.4 BIEN-mappable percentage by provider"
)
```

> **8.5 Coverage note:** Traits not yet in this lookup are not necessarily unmappable - they may require fuzzy matching or manual review. This table represents **exact canonical name matches only** (case-insensitive, whitespace-trimmed). Full BIEN trait vocabulary integration is a planned pipeline milestone.

---

## 9 QA Flag Summary {#qa-flags}

```{r qa-summary}
message(Sys.time(), " - loaded [qa-flags]")

qa_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  flagged <- sum(nzchar(trimws(d$qa_flags)), na.rm = TRUE)
  list(
    Provider         = provider_labels[[k]],
    `Total Rows`     = nrow(d),
    `Flagged Rows`   = flagged,
    `Unflagged Rows` = nrow(d) - flagged,
    `Flag Rate (%)`  = round(100 * flagged / nrow(d), 1)
  )
}))
kable_styled(qa_rows, caption = "9.1 QA flag prevalence by provider")
```

### 9.2 DataDryad top 15 QA flag codes {#qa-dryad-top}

```{r qa-dryad-top}
d_dry        <- dat_list[["dryad"]]
flagged_vals <- d_dry$qa_flags[nzchar(trimws(d_dry$qa_flags))]
flag_tokens  <- unlist(strsplit(flagged_vals, "\\|"))
flag_tokens  <- trimws(sub("\\[.*", "", flag_tokens))
flag_tokens  <- flag_tokens[nzchar(flag_tokens)]
flag_tbl_all <- sort(table(flag_tokens), decreasing = TRUE)
flag_top15   <- as.data.frame(
  flag_tbl_all[seq_len(min(15L, length(flag_tbl_all)))]
)
names(flag_top15) <- c("QA Flag Code", "Count")

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    flag_top15,
    extensions = "Scroller",
    options    = list(scrollY = 300, scroller = TRUE, scrollX = TRUE, dom = "t"),
    rownames   = FALSE,
    caption    = "9.2 DataDryad: top 15 QA flag codes (prefix only)"
  )
} else {
  kable_styled(flag_top15, caption = "9.2 DataDryad: top 15 QA flag codes")
}
```

### 9.3 Top 15 QA flag codes - bar chart {#qa-bar}

```{r fig-qa-bar, fig.height=4}
flag_df <- flag_top15
names(flag_df) <- c("flag_code", "count")
ggplot(flag_df, aes(x = reorder(flag_code, count), y = count / 1e3)) +
  geom_col(fill = "#e15759") +
  coord_flip() +
  labs(x = "QA flag code", y = "Occurrences (thousands)",
       title = "9.3 DataDryad - top 15 QA flag codes") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 10 Pending Sources — Manual Intake Queue {#pending-sources}

These user-supplied sources are tracked in `data/manual_source_intake.csv` for review, ingestion, and follow-up against the occurrence harvest outputs.

```{r pending-sources-table, message=FALSE}
message(Sys.time(), " - loaded [pending-sources-table]")
manual_intake <- data.table::fread(file.path(proj_root, "data", "manual_source_intake.csv"))
manual_status_counts <- manual_intake[, .N, by = harvest_status][order(-N, harvest_status)]
manual_status_summary <- paste(
  sprintf(
    "%s %s",
    formatC(manual_status_counts$N, format = "d", big.mark = ","),
    paste0("`", manual_status_counts$harvest_status, "`")
  ),
  collapse = ", "
)
manual_occurrence_files <- Sys.glob(file.path(
  proj_root, "output", "providers", "occurrences", "*", "compiled_occurrences.csv"
))
compiled_manual_occurrences <- data.table::data.table(
  source_id = basename(dirname(manual_occurrence_files)),
  compiled_occurrence_rows = vapply(
    manual_occurrence_files,
    function(path) nrow(data.table::fread(path, showProgress = FALSE)),
    integer(1)
  )
)
if (nrow(compiled_manual_occurrences)) {
  compiled_manual_occurrences <- manual_intake[
    compiled_manual_occurrences,
    on = "source_id"
  ][
    , .(source_id, display_name, source_group, harvest_status, compiled_occurrence_rows)
  ][order(-compiled_occurrence_rows, source_id)]
} else {
  compiled_manual_occurrences <- data.table::data.table(
    source_id = character(),
    display_name = character(),
    source_group = character(),
    harvest_status = character(),
    compiled_occurrence_rows = integer()
  )
}
compiled_manual_source_count <- nrow(compiled_manual_occurrences)
compiled_manual_occurrence_total <- sum(compiled_manual_occurrences$compiled_occurrence_rows)
cols_show <- c("display_name", "source_group", "canonical_url_or_doi",
               "related_publication", "harvest_status", "next_action")
if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(manual_intake[, ..cols_show], rownames = FALSE, options = list(scrollX = TRUE))
} else {
  kable_styled(
    manual_intake[, ..cols_show],
    caption = "Manual intake registry"
  )
}
```

The manual intake registry currently tracks `r formatC(nrow(manual_intake), format = "d", big.mark = ",")` sources across `r formatC(nrow(manual_status_counts), format = "d", big.mark = ",")` harvest statuses: `r manual_status_summary`.

### 10.1 Compiled manual occurrence sources {#compiled-manual-occurrences}

The occurrence harvest currently includes `r formatC(compiled_manual_source_count, format = "d", big.mark = ",")` compiled manual sources totaling `r formatC(compiled_manual_occurrence_total, format = "d", big.mark = ",")` occurrence records.

```{r compiled-manual-occurrence-table}
if (nrow(compiled_manual_occurrences)) {
  kable_styled(
    compiled_manual_occurrences[, .(
      `Source ID` = source_id,
      `Display Name` = fifelse(is.na(display_name) | !nzchar(display_name), source_id, display_name),
      `Source Group` = source_group,
      `Harvest Status` = harvest_status,
      `Occurrence Rows` = compiled_occurrence_rows
    )],
    caption = "Compiled manual occurrence sources detected under output/providers/occurrences"
  )
} else {
  cat("No compiled manual occurrence sources were detected under output/providers/occurrences.")
}
```

The original 6 entries cover the ShrubHub TraitHub repository (Bjorkman et al. 2018 Nature), three files supplied directly by Jesse Miller (Spasojevic et al. 2016 Ozark trait study), email context from Jesse Miller, and field data from Pablo Peri's BEPSP Patagonia network (Peri et al. 2016 J. Nature Conservation).

The 29 entries added in the first 2025 expansion cover underrepresented regions:

- **Central Africa (3)**: RAINBIO tropical Africa mega-database (Dauby et al. 2016), ECAT expert-validated tree distribution dataset (Tack et al. 2022), Herbier National du Gabon GBIF occurrences
- **Russia / Central Asia (7)**: Flora of Russia iNaturalist dataset (Seregin et al. 2020), Oxytropis Asian Russia (Sandanov et al. 2022), Virtual Herbarium ALTB Altai region (Vaganov et al. 2021), Fleroff georeferenced historical records (Seregin & Basov 2021), two Kyrgyzstan alien/vascular plant checklists (Sennikov & Lazkov 2021/2025 — Dryad DwC-ready), Flora of Mongolia iNaturalist (Baasanmunkh et al. 2025)
- **Southeast Asia (8)**: Lesser Sunda Islands endemic checklist (Jennings et al. 2026), Sunda-Sahul checklist + supplementary dataset (Joyce et al. 2020), Myanmar checklist (Aung et al. 2025), three GBIF ANDA Herbarium Sumatra occurrence datasets
- **Pacific Islands / Southern Ocean (5)**: PacIFlora paper + Dryad dataset (Wohlwend et al. 2021), Southern Islands Vascular Flora paper + SIVFLORA Zenodo dataset (Guerrero et al. 2025)
- **Amazon / South America (4)**: Amazon species list (Cardoso et al. 2017), Colombia bioregions useful plants (Bystriakova et al. 2021), Theobroma/Herrania specimen occurrences (Colli-Silva et al. 2023)
- **Chile (5)**: PUCV Herbarium BDJ paper + GBIF dataset (Cordero et al. 2022), Chile vascular plant diversity 181K occurrences (Luebert et al. 2022), High Andes elevation gradient plot data (Arroyo et al. 2024), sub-Antarctic Aysen/Magallanes dataset (DOI pending)

The 19 entries added in the second 2026 expansion focus on **vegetation plot networks, national forest inventories, and aggregator portals**:

- **Arctic & boreal (3)**: Arctic Vegetation Archive (~31000 plots), CAFF Flora Group national-database aggregator, Russian Arctic Vegetation Archive (Walker et al. 2023, 2873 plots)
- **Latin America (5)**: SYNTREESYS CESAB synthetic tree plot database, DryFlor seasonally dry tropical forest network, Amazon Tree Diversity Network (ATDN), HERBase Amazon herbs, Argentina Red de Parcelas Permanentes + Inventario Nacional de Bosques Nativos
- **Trait sources (1)**: Rasgos-CL Chilean trait database (Dylan Craven)
- **Eurasia & Central Africa (3)**: TUBIVES Turkish plant database, DarkDivNet global presence/absence plot network, CAFRIPLOT Central African plot network (233 plots + traits + herbarium)
- **North America & global aggregators (5)**: Canadian National Forest Inventory, Enquist Lab Forest Macrosystems & additional Gentry plots, Global Forest Biodiversity Initiative Science-i platform, GIVD Global Index of Vegetation Plot Databases, GLORIA Alpine plant network
- **Asia (2)**: ASEAN RBIM dashboard (likely checklist-only — to verify), Turkish TUBIVES (above)

GBIF occurrence datasets (`intake_type = gbif_occurrence_dataset`) are directly ingestible via the GBIF API. Dryad datasets with Darwin Core format (e.g., Kyrgyzstan, Dryad DOI 10.5061/dryad.x3ffbg7wt) are high-priority compile targets. Vegetation plot networks (`source_group = vegetation_plot_database`) require coordination with their respective consortia (sPlot/EVA deduplication is essential). Checklist-only entries are tagged in `notes` for use as taxonomic backbones rather than primary occurrence sources. Aggregator/registry entries (CAFF, GIVD) should be used to enumerate further individual datasets rather than ingested directly.

---

## 11 Session Info {#session-info}

```{r session-info}
message(Sys.time(), " - loaded [session-info]")

si <- sessionInfo()
cat("Rendered:    ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), "\n")
cat("R version:   ", si$R.version$version.string, "\n")
cat("data.table:  ", as.character(packageVersion("data.table")), "\n")
cat("ggplot2:     ", as.character(packageVersion("ggplot2")), "\n")
cat("leaflet:     ",
    tryCatch(as.character(packageVersion("leaflet")),    error = function(e) "not installed"), "\n")
cat("DT:          ",
    tryCatch(as.character(packageVersion("DT")),         error = function(e) "not installed"), "\n")
cat("kableExtra:  ",
    tryCatch(as.character(packageVersion("kableExtra")), error = function(e) "not installed"), "\n")
cat("scales:      ",
    tryCatch(as.character(packageVersion("scales")),     error = function(e) "not installed"), "\n")
cat("lubridate:   ",
    tryCatch(as.character(packageVersion("lubridate")),  error = function(e) "not installed"), "\n")
```
