---
title: "Cross-Provider Plant Trait Harvest Summary"
subtitle: "DataDryad | Zenodo | Scientific Data"
author: "DryadPlantTraits Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)

library(data.table)
library(ggplot2)

# Detect project root by walking up until R/dryad_api.R is found
find_dryad_root <- function(start = getwd()) {
  path <- normalizePath(start, mustWork = FALSE)
  for (i in seq_len(12)) {
    if (file.exists(file.path(path, "R", "dryad_api.R"))) return(path)
    parent <- dirname(path)
    if (parent == path) break
    path <- parent
  }
  # fall back to the parent of the reports/ directory (= DryadPlantTraits/)
  normalizePath(file.path(knitr::current_input(dir = TRUE), ".."),
                mustWork = FALSE)
}
proj_root <- find_dryad_root()

kable_styled <- function(df, caption = NULL) {
  df_fmt <- as.data.frame(df, stringsAsFactors = FALSE)
  num_cols <- vapply(df_fmt, is.numeric, logical(1))
  if (any(num_cols)) {
    df_fmt[num_cols] <- lapply(df_fmt[num_cols], function(x) {
      out <- format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
      out[is.na(x)] <- NA_character_
      out
    })
  }
  tbl <- knitr::kable(df_fmt, caption = caption, format = "html")
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    tryCatch(
      kableExtra::kable_styling(tbl,
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE),
      error = function(e) tbl
    )
  } else {
    tbl
  }
}

# -------------------------------------------------------------------------
# File paths
# -------------------------------------------------------------------------
paths <- list(
  dryad   = file.path(proj_root, "output/compiled_trait_observations_with_unit_inference.csv"),
  zenodo  = file.path(proj_root, "output/providers/zenodo/compiled_trait_observations_with_unit_inference.csv"),
  scidata = file.path(proj_root, "output/providers/scientific_data/compiled_trait_observations_with_unit_inference.csv")
)
provider_labels <- c(dryad = "DataDryad", zenodo = "Zenodo", scidata = "Scientific Data")

# Explicit file existence checks before loading
for (.k in names(paths)) {
  if (!file.exists(paths[[.k]])) {
    stop(
      "Missing provider file: ", provider_labels[[.k]],
      "\n  Expected: ", paths[[.k]]
    )
  }
}

read_provider <- function(path, key) {
  d <- fread(path, data.table = TRUE)
  d[, provider := provider_labels[[key]]]
  d
}

dat_list <- lapply(names(paths), function(k) read_provider(paths[[k]], k))
names(dat_list) <- names(paths)
```

---

## Key Takeaways

```{r key-takeaways}
total_rows   <- sum(sapply(dat_list, nrow))
total_traits <- length(unique(unlist(lapply(dat_list, function(d) unique(d$trait_name)))))
total_taxa   <- length(unique(unlist(lapply(dat_list, function(d) unique(d$scrubbed_species_binomial)))))

coord_count <- function(d) {
  lat <- suppressWarnings(as.numeric(d$latitude))
  lon <- suppressWarnings(as.numeric(d$longitude))
  sum(is.finite(lat) & is.finite(lon), na.rm = TRUE)
}
total_coords <- sum(sapply(dat_list, coord_count))

# Dynamic per-provider stats used in prose below
prov_rows     <- sapply(dat_list, nrow)
prov_datasets <- sapply(dat_list, function(d) data.table::uniqueN(d$source_title))

rows_rank <- names(sort(prov_rows, decreasing = TRUE))
dataset_rank <- names(sort(prov_datasets, decreasing = TRUE))
largest_volume_provider <- provider_labels[[rows_rank[1L]]]
largest_volume_rows <- prov_rows[[rows_rank[1L]]]
most_diverse_provider <- provider_labels[[dataset_rank[1L]]]
most_diverse_datasets <- prov_datasets[[dataset_rank[1L]]]
remaining_provider <- setdiff(names(provider_labels), c(rows_rank[1L], dataset_rank[1L]))[1L]
if (is.na(remaining_provider) || length(remaining_provider) == 0L) {
  remaining_provider <- rows_rank[2L]
}
remaining_provider_label <- provider_labels[[remaining_provider]]
remaining_provider_rows <- prov_rows[[remaining_provider]]
remaining_provider_datasets <- prov_datasets[[remaining_provider]]
```

> **`r formatC(total_rows, format="d", big.mark=",")`** observation rows harvested across **3 providers** (DataDryad, Zenodo, Scientific Data) as of `r Sys.Date()`.
>
> **`r total_traits`** distinct trait names detected; **`r formatC(total_taxa, format="d", big.mark=",")`** distinct taxa (scrubbed binomials); **`r formatC(total_coords, format="d", big.mark=",")`** rows with geographic coordinates.
>
> `r most_diverse_provider` is currently the most source-diverse contributor (`r formatC(most_diverse_datasets, format="d", big.mark=",")` source datasets), while `r largest_volume_provider` provides the largest observation volume (`r formatC(largest_volume_rows, format="d", big.mark=",")` rows). `r remaining_provider_label` adds a complementary set (`r formatC(remaining_provider_datasets, format="d", big.mark=",")` datasets, `r formatC(remaining_provider_rows, format="d", big.mark=",")` rows). All three provider files carry unit reconciliation columns (`inferred_unit`, `inferred_unit_confidence`); `basis_type` column is present in all files.
>
> _Note: observation row counts are summed across provider outputs independently -- no cross-provider deduplication has been applied; the same measurement may appear in more than one provider if it was included in multiple source compilations._

---

## 1. Provider Overview

```{r provider-overview}
provider_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  lat <- suppressWarnings(as.numeric(d$latitude))
  lon <- suppressWarnings(as.numeric(d$longitude))
  list(
    Provider             = provider_labels[[k]],
    `Observation Rows`   = nrow(d),
    `Distinct Traits`    = uniqueN(d$trait_name),
    `Distinct Taxa`      = uniqueN(d$scrubbed_species_binomial),
    `Source Datasets`    = uniqueN(d$source_title),
    `Rows with Coords`   = sum(is.finite(lat) & is.finite(lon), na.rm = TRUE),
    `Rows with QA Flags` = sum(nzchar(trimws(d$qa_flags)), na.rm = TRUE)
  )
}))

kable_styled(provider_summary, caption = "Provider-level harvest summary")
```

### Figure: Observation rows by provider

```{r fig-provider-rows}
prov_plot <- data.frame(
  Provider = provider_summary$Provider,
  Rows     = provider_summary$`Observation Rows`
)
ggplot(prov_plot, aes(x = reorder(Provider, -Rows), y = Rows / 1e3, fill = Provider)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = formatC(Rows, format = "d", big.mark = ",")),
            vjust = -0.4, size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Observation rows by provider") +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 2. Source / Citation Breakdown by Provider

### 2.1 DataDryad (top 25 sources by row count)

```{r dryad-sources}
d <- dat_list[["dryad"]]
src_dry <- d[, .(
  Rows              = .N,
  `Distinct Traits` = uniqueN(trait_name),
  `Source ID / DOI` = dryad_dataset_doi[1L],
  Authors           = substr(source_authors[1L], 1, 65)
), keyby = source_title][order(-Rows)]
kable_styled(head(src_dry, 25L)[, .(
  `Source Title`        = source_title,
  `Source ID / DOI`,
  `Authors (truncated)` = Authors,
  Rows,
  `Distinct Traits`
)], caption = "DataDryad: top 25 source datasets by row count")
```

### 2.2 Zenodo (all sources)

```{r zenodo-sources}
d <- dat_list[["zenodo"]]
src_zen <- d[, .(
  Rows              = .N,
  `Distinct Traits` = uniqueN(trait_name),
  `Source ID / DOI` = dryad_dataset_doi[1L],
  Authors           = substr(source_authors[1L], 1, 65)
), keyby = source_title][order(-Rows)]
kable_styled(src_zen[, .(
  `Source Title`        = source_title,
  `Source ID / DOI`,
  `Authors (truncated)` = Authors,
  Rows,
  `Distinct Traits`
)], caption = "Zenodo: all source datasets by row count")
```

### 2.3 Scientific Data (all sources)

```{r scidata-sources}
d <- dat_list[["scidata"]]
src_sci <- d[, .(
  Rows              = .N,
  `Distinct Traits` = uniqueN(trait_name),
  `Source ID / DOI` = dryad_dataset_doi[1L],
  Authors           = substr(source_authors[1L], 1, 65)
), keyby = source_title][order(-Rows)]
kable_styled(src_sci[, .(
  `Source Title`        = source_title,
  `Source ID / DOI`,
  `Authors (truncated)` = Authors,
  Rows,
  `Distinct Traits`
)], caption = "Scientific Data: all source datasets by row count")
```

---

## 3. Trait Coverage

### 3.1 Overall top 25 traits (all providers combined)

```{r top-traits-overall}
all_dat <- rbindlist(dat_list, use.names = TRUE, fill = TRUE)
top_traits <- all_dat[!is.na(trait_name) & nzchar(trait_name),
  .(Rows = .N, `Distinct Taxa` = uniqueN(scrubbed_species_binomial)),
  by = trait_name][order(-Rows)]
kable_styled(head(top_traits, 25L)[, .(
  `Trait Name`    = trait_name,
  Rows,
  `Distinct Taxa`
)], caption = "Top 25 traits by row count (all providers combined)")
```

### Figure: Top 20 traits across all providers

```{r fig-top-traits}
top20 <- head(top_traits, 20L)
ggplot(top20, aes(x = reorder(trait_name, Rows), y = Rows / 1e3, fill = Rows)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#a8d8ea", high = "#0a5275") +
  coord_flip() +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Top 20 plant traits -- all providers combined") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 3.2 Provider-specific top traits (top 15 each)

```{r top-traits-by-provider, results='asis'}
for (k in names(dat_list)) {
  d <- dat_list[[k]]
  top <- d[!is.na(trait_name) & nzchar(trait_name),
    .(Rows = .N, `Distinct Taxa` = uniqueN(scrubbed_species_binomial)),
    by = trait_name][order(-Rows)]
  cat("\n#### ", provider_labels[[k]], "\n\n")
  print(kable_styled(
    head(top, 15L)[, .(`Trait Name` = trait_name, Rows, `Distinct Taxa`)],
    caption = paste(provider_labels[[k]], ": top 15 traits")
  ))
  cat("\n")
}
```

---

## 4. Unit Reconciliation Summary

All three provider files carry the unit reconciliation columns added by the pipeline (`inferred_unit`, `inferred_unit_confidence`, `inference_evidence`, `basis_type`, `unit_conversion_factor`).

```{r unit-reconciliation}
unit_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  iu  <- d$inferred_unit
  iuc <- as.character(d$inferred_unit_confidence)
  iuc[is.na(iuc) | !nzchar(trimws(iuc))] <- NA_character_
  list(
    Provider               = provider_labels[[k]],
    `Total Rows`           = nrow(d),
    `Inferred TRUE`        = sum(iu %in% c(TRUE, "TRUE",  "true",  1L), na.rm = TRUE),
    `Inferred FALSE`       = sum(iu %in% c(FALSE,"FALSE", "false", 0L), na.rm = TRUE),
    `Conf: high`           = sum(iuc == "high",   na.rm = TRUE),
    `Conf: medium`         = sum(iuc == "medium",  na.rm = TRUE),
    `Conf: low`            = sum(iuc == "low",     na.rm = TRUE),
    `Conf: none/missing`   = sum(is.na(iuc))
  )
}))
kable_styled(unit_summary, caption = "Unit inference summary by provider")
```

### 4.1 inferred_unit_confidence breakdown (all providers)

```{r confidence-breakdown}
conf_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  lvl <- as.character(d$inferred_unit_confidence)
  lvl[is.na(lvl) | !nzchar(trimws(lvl))] <- "(missing)"
  lvl <- trimws(lvl)
  tbl <- d[, .N, by = .(inferred_unit_confidence = lvl)][order(-N)]
  tbl[, provider := provider_labels[[k]]]
  tbl
}))
conf_wide <- dcast(conf_rows, inferred_unit_confidence ~ provider, value.var = "N", fill = 0L)
kable_styled(conf_wide, caption = "inferred_unit_confidence counts by provider")
```

### 4.2 basis_type breakdown

```{r basis-type-breakdown}
bt_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  bt <- trimws(d$basis_type)
  bt[!nzchar(bt)] <- "(empty)"
  tbl <- d[, .N, by = .(basis_type = bt)][order(-N)]
  tbl[, provider := provider_labels[[k]]]
  tbl
}))
bt_wide <- dcast(bt_rows, basis_type ~ provider, value.var = "N", fill = 0L)
kable_styled(bt_wide, caption = "basis_type counts by provider")
```

---

## 5. Geographic Coverage

```{r geo-coverage}
geo_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  lat <- suppressWarnings(as.numeric(d$latitude))
  lon <- suppressWarnings(as.numeric(d$longitude))
  has_coord <- is.finite(lat) & is.finite(lon)
  list(
    Provider           = provider_labels[[k]],
    `Total Rows`       = nrow(d),
    `Rows with Coords` = sum(has_coord, na.rm = TRUE),
    `Coverage (%)`     = round(100 * mean(has_coord, na.rm = TRUE), 1)
  )
}))
kable_styled(geo_summary, caption = "Geographic coordinate coverage by provider")
```

---

## 6. QA Flag Summary

```{r qa-flags, results='asis'}
qa_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  flagged <- sum(nzchar(trimws(d$qa_flags)), na.rm = TRUE)
  list(
    Provider         = provider_labels[[k]],
    `Total Rows`     = nrow(d),
    `Flagged Rows`   = flagged,
    `Unflagged Rows` = nrow(d) - flagged,
    `Flag Rate (%)`  = round(100 * flagged / nrow(d), 1)
  )
}))
kable_styled(qa_rows, caption = "QA flag prevalence by provider")

# Top flag codes for DataDryad (most complex)
d_dry <- dat_list[["dryad"]]
flagged_vals <- d_dry$qa_flags[nzchar(trimws(d_dry$qa_flags))]
flag_tokens  <- unlist(strsplit(flagged_vals, "\\|"))
flag_tokens  <- trimws(sub("\\[.*", "", flag_tokens))
flag_tokens  <- flag_tokens[nzchar(flag_tokens)]
flag_tbl_all <- sort(table(flag_tokens), decreasing = TRUE)
flag_tbl     <- as.data.frame(flag_tbl_all[seq_len(min(15L, length(flag_tbl_all)))])
names(flag_tbl) <- c("QA Flag Code", "Count")
cat("\n\n**DataDryad: top 15 QA flag codes**\n\n")
print(kable_styled(flag_tbl, caption = "DataDryad: top 15 QA flag codes (prefix only)"))
```

---

## 7. Session Info

```{r session-info}
si <- sessionInfo()
cat("R version:", si$R.version$version.string, "\n")
cat("data.table:", as.character(packageVersion("data.table")), "\n")
cat("ggplot2:", as.character(packageVersion("ggplot2")), "\n")
cat("Rendered:", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), "\n")
```
