---
title: "Plant Trait Harvest — Detailed Cross-Provider Report"
subtitle: "DataDryad · Zenodo · Scientific Data (AusTraits)"
author: "DryadPlantTraits Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE}
message(Sys.time(), " - loaded [setup]")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(knitr)
})
if (requireNamespace("kableExtra", quietly = TRUE)) library(kableExtra)
if (requireNamespace("scales",     quietly = TRUE)) library(scales)
if (requireNamespace("leaflet",    quietly = TRUE)) library(leaflet)
if (requireNamespace("DT",         quietly = TRUE)) library(DT)

# -- Project root detection
find_dryad_root <- function(start = getwd()) {
  path <- normalizePath(start, mustWork = FALSE)
  for (i in seq_len(12)) {
    if (file.exists(file.path(path, "R", "dryad_api.R"))) return(path)
    parent <- dirname(path)
    if (parent == path) break
    path <- parent
  }
  normalizePath(file.path(knitr::current_input(dir = TRUE), ".."),
                mustWork = FALSE)
}
proj_root <- find_dryad_root()

# -- kable helper
kable_styled <- function(df, caption = NULL) {
  df_fmt <- as.data.frame(df, stringsAsFactors = FALSE)
  num_cols <- vapply(df_fmt, is.numeric, logical(1))
  if (any(num_cols)) {
    df_fmt[num_cols] <- lapply(df_fmt[num_cols], function(x) {
      out <- format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
      out[is.na(x)] <- NA_character_
      out
    })
  }
  tbl <- knitr::kable(df_fmt, caption = caption, format = "html")
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    tryCatch(
      kableExtra::kable_styling(tbl,
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE),
      error = function(e) tbl
    )
  } else {
    tbl
  }
}

# -- Provider paths
paths <- list(
  dryad     = file.path(proj_root, "output/compiled_trait_observations_with_unit_inference.csv"),
  zenodo    = file.path(proj_root, "output/providers/zenodo/compiled_trait_observations_with_unit_inference.csv"),
  scidata   = file.path(proj_root, "output/providers/scientific_data/compiled_trait_observations_with_unit_inference.csv"),
  traithub  = file.path(proj_root, "output/providers/traithub/compiled_trait_observations.csv")
)
provider_labels <- c(dryad = "DataDryad", zenodo = "Zenodo", scidata = "Scientific Data", traithub = "TraitHub (tundra)")
provider_colors <- c(DataDryad = "#1f77b4", Zenodo = "#ff7f0e", `Scientific Data` = "#2ca02c", `TraitHub (tundra)` = "#9467bd")

for (.k in names(paths)) {
  if (!file.exists(paths[[.k]])) {
    stop("Missing provider file: ", provider_labels[[.k]],
         "\n  Expected: ", paths[[.k]])
  }
}

read_provider <- function(path, key) {
  d <- fread(path, data.table = TRUE)
  d[, provider := provider_labels[[key]]]
  d
}

dat_list <- lapply(names(paths), function(k) read_provider(paths[[k]], k))
names(dat_list) <- names(paths)

# -- Combined data table
all_dat <- rbindlist(dat_list, use.names = TRUE, fill = TRUE)

# -- Coordinate validity
all_dat[, lat_num := suppressWarnings(as.numeric(latitude))]
all_dat[, lon_num := suppressWarnings(as.numeric(longitude))]
all_dat[, has_coord := is.finite(lat_num) & is.finite(lon_num) &
                    lat_num >= -90 & lat_num <= 90 &
                    lon_num >= -180 & lon_num <= 180]

for (k in names(dat_list)) {
  dat_list[[k]][, lat_num := suppressWarnings(as.numeric(latitude))]
  dat_list[[k]][, lon_num := suppressWarnings(as.numeric(longitude))]
  dat_list[[k]][, has_coord := is.finite(lat_num) & is.finite(lon_num) &
                          lat_num >= -90 & lat_num <= 90 &
                          lon_num >= -180 & lon_num <= 180]
}
```

---

## 1 Executive Summary {#executive-summary}

```{r exec-summary}
message(Sys.time(), " - loaded [exec-summary]")

total_rows   <- nrow(all_dat)
total_traits <- uniqueN(all_dat[!is.na(trait_name) & nzchar(trait_name), trait_name])
total_taxa   <- uniqueN(all_dat[!is.na(scrubbed_species_binomial) & nzchar(scrubbed_species_binomial),
                                scrubbed_species_binomial])
total_coords <- sum(all_dat$has_coord, na.rm = TRUE)
```

<div class="alert alert-info" role="alert">
<h4 class="alert-heading">Key Takeaways</h4>
<ul>
  <li><strong>`r formatC(total_rows, format="d", big.mark=",")`</strong> total observation rows across <strong>3 providers</strong> (DataDryad, Zenodo, Scientific Data) as of `r Sys.Date()`.</li>
  <li><strong>`r formatC(total_traits, format="d", big.mark=",")`</strong> distinct trait names detected.</li>
  <li><strong>`r formatC(total_taxa, format="d", big.mark=",")`</strong> distinct taxa (scrubbed species binomials).</li>
  <li><strong>`r formatC(total_coords, format="d", big.mark=",")`</strong> rows with valid geographic coordinates (DataDryad + Zenodo only; Scientific Data does not carry coordinates in this extract).</li>
  <li><em>Note on Scientific Data / AusTraits unit confidence:</em> Rows from the Scientific Data provider show <code>inferred_unit_confidence = "none"</code> because the simple-schema mapper for this provider did not populate the <code>basis_type</code> or <code>value_type</code> metadata columns that the confidence scorer uses as inputs. AusTraits is a fully curated, peer-reviewed database (Falster et al. 2021, <em>Sci. Data</em>, doi:10.1038/s41597-021-01006-6). Its units were set by original data contributors and validated during publication. The "none" label reflects a <strong>pipeline labeling gap</strong>, not a unit quality failure.</li>
  <li><em>Cross-provider deduplication has not been applied.</em> The same measurement may appear in more than one provider if included in multiple source compilations.</li>
</ul>
</div>

---

## 2 Provider Overview {#provider-overview}

```{r provider-overview}
message(Sys.time(), " - loaded [provider-overview]")

provider_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  list(
    Provider             = provider_labels[[k]],
    `Observation Rows`   = nrow(d),
    `Distinct Traits`    = uniqueN(d[!is.na(trait_name) & nzchar(trait_name), trait_name]),
    `Distinct Taxa`      = uniqueN(d[!is.na(scrubbed_species_binomial) & nzchar(scrubbed_species_binomial),
                                     scrubbed_species_binomial]),
    `Source Datasets`    = uniqueN(d$source_title),
    `Rows with Coords`   = sum(d$has_coord, na.rm = TRUE),
    `Rows with QA Flags` = sum(nzchar(trimws(d$qa_flags)), na.rm = TRUE)
  )
}))

kable_styled(provider_summary, caption = "Provider-level harvest summary")
```

### 2.1 Observation rows by provider {#rows-by-provider}

```{r fig-provider-rows, fig.height=4}
prov_plot_df <- data.frame(
  Provider = provider_summary$Provider,
  Rows     = provider_summary$`Observation Rows`
)
ggplot(prov_plot_df, aes(x = reorder(Provider, -Rows), y = Rows / 1e3, fill = Provider)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = formatC(Rows, format = "d", big.mark = ",")),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = provider_colors) +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Observation rows by provider") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 2.2 Trait count by provider (stacked, top 20) {#trait-stacked}

```{r fig-trait-stacked, fig.height=5}
trait_by_prov <- all_dat[!is.na(trait_name) & nzchar(trait_name),
                         .(n = .N), by = .(trait_name, provider)]
top_traits_all <- all_dat[!is.na(trait_name) & nzchar(trait_name),
                           .(total = .N), by = trait_name][order(-total)]
top20_names <- head(top_traits_all$trait_name, 20L)
tbp20 <- trait_by_prov[trait_name %in% top20_names]
tbp20[, provider := factor(provider, levels = c("DataDryad", "Zenodo", "Scientific Data"))]
ggplot(tbp20, aes(x = reorder(trait_name, n, sum), y = n / 1e3, fill = provider)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(values = provider_colors, name = "Provider") +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Top 20 traits - stacked by provider") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

---

## 3 Source / Citation Breakdown {#source-citation}

```{r source-tables, results='asis'}
message(Sys.time(), " - loaded [source-citation]")

for (k in names(dat_list)) {
  d <- dat_list[[k]]
  src_tbl <- d[, .(
    Rows              = .N,
    `Distinct Traits` = uniqueN(trait_name),
    `Distinct Taxa`   = uniqueN(scrubbed_species_binomial),
    DOI               = dryad_dataset_doi[1L]
  ), by = source_title][order(-Rows)]
  setnames(src_tbl, "source_title", "Source Title")

  cat("\n### 3.", which(names(dat_list) == k), " ", provider_labels[[k]], "\n\n",
      sep = "")
  if (requireNamespace("DT", quietly = TRUE)) {
    print(DT::datatable(
      as.data.frame(src_tbl),
      filter = "top",
      extensions = "Scroller",
      options = list(scrollY = 300, scroller = TRUE, scrollX = TRUE, pageLength = 20),
      caption = paste(provider_labels[[k]], ": source datasets"),
      rownames = FALSE
    ))
  } else {
    print(kable_styled(head(src_tbl, 25L),
                       caption = paste(provider_labels[[k]], ": top 25 sources")))
  }
  cat("\n\n")
}
```

---

## 4 Trait Coverage - Detailed {#trait-coverage}

```{r trait-coverage-setup}
message(Sys.time(), " - loaded [trait-coverage]")
top_traits_tbl <- all_dat[!is.na(trait_name) & nzchar(trait_name),
  .(Rows = .N, `Distinct Taxa` = uniqueN(scrubbed_species_binomial)),
  by = trait_name][order(-Rows)]
```

### 4.1 Top 25 traits combined {#top25-traits}

```{r top25-traits}
kable_styled(
  head(top_traits_tbl, 25L)[, .(`Trait Name` = trait_name, Rows, `Distinct Taxa`)],
  caption = "Top 25 traits by row count (all providers combined)"
)
```

### 4.2 Top 20 traits - provider contribution {#top20-stacked}

```{r fig-top20-provider, fig.height=5}
ggplot(tbp20, aes(x = reorder(trait_name, n, sum), y = n / 1e3, fill = provider)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(values = provider_colors, name = "Provider") +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Top 20 traits - provider contribution") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

### 4.3 Provider-specific top 15 traits {#provider-top15}

```{r provider-top15, results='asis'}
for (k in names(dat_list)) {
  d <- dat_list[[k]]
  top15 <- d[!is.na(trait_name) & nzchar(trait_name),
    .(Rows = .N, `Distinct Taxa` = uniqueN(scrubbed_species_binomial)),
    by = trait_name][order(-Rows)]
  cat("\n#### ", provider_labels[[k]], "\n\n", sep = "")
  if (requireNamespace("DT", quietly = TRUE)) {
    print(DT::datatable(
      as.data.frame(head(top15, 15L))[, c("trait_name", "Rows", "Distinct Taxa")],
      extensions = "Scroller",
      options = list(scrollY = 280, scroller = TRUE, scrollX = TRUE, dom = "t"),
      colnames = c("Trait Name", "Rows", "Distinct Taxa"),
      rownames = FALSE
    ))
  } else {
    print(kable_styled(
      head(top15, 15L)[, .(`Trait Name` = trait_name, Rows, `Distinct Taxa`)],
      caption = paste(provider_labels[[k]], ": top 15 traits")
    ))
  }
  cat("\n\n")
}
```

### 4.4 Trait value summary statistics (top 20 numeric traits) {#trait-stats}

```{r trait-value-stats}
message(Sys.time(), " - loaded [trait-value-stats]")

all_dat[, val_num := suppressWarnings(as.numeric(trait_value))]

numeric_counts <- all_dat[is.finite(val_num) & !is.na(trait_name) & nzchar(trait_name),
  .(n_numeric = .N), by = trait_name][order(-n_numeric)]
top20_numeric <- head(numeric_counts[n_numeric >= 100], 20L)$trait_name

stats_list <- lapply(top20_numeric, function(tr) {
  vals <- all_dat[trait_name == tr & is.finite(val_num), val_num]
  list(
    `Trait Name` = tr,
    n            = length(vals),
    Mean         = round(mean(vals), 4),
    Median       = round(median(vals), 4),
    SD           = round(sd(vals), 4),
    Min          = round(min(vals), 4),
    P5           = round(quantile(vals, 0.05), 4),
    P25          = round(quantile(vals, 0.25), 4),
    P75          = round(quantile(vals, 0.75), 4),
    P95          = round(quantile(vals, 0.95), 4),
    Max          = round(max(vals), 4)
  )
})
stats_tbl <- rbindlist(stats_list)
kable_styled(stats_tbl,
  caption = "Summary statistics for top 20 numeric traits (rows with finite numeric trait_value only)")
```

### 4.5 Trait value boxplots (top 12 numeric traits) {#trait-boxplots}

```{r fig-trait-boxplots, fig.height=7, fig.width=10}
message(Sys.time(), " - loaded [trait-boxplots]")

top12_numeric <- head(top20_numeric, 12L)
box_dat <- all_dat[trait_name %in% top12_numeric & is.finite(val_num),
                   .(trait_name, val_num, provider)]

box_dat_clipped <- box_dat[, {
  p1  <- quantile(val_num, 0.01)
  p99 <- quantile(val_num, 0.99)
  .SD[val_num >= p1 & val_num <= p99]
}, by = trait_name]

box_dat_clipped[, provider := factor(provider,
  levels = c("DataDryad", "Zenodo", "Scientific Data"))]

ggplot(box_dat_clipped, aes(x = provider, y = val_num, fill = provider)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3) +
  facet_wrap(~ trait_name, scales = "free_y", ncol = 3) +
  scale_y_continuous(trans = "pseudo_log") +
  scale_fill_manual(values = provider_colors, name = "Provider") +
  labs(x = NULL, y = "Trait value (pseudo-log scale)",
       title = "Trait value distributions - top 12 numeric traits",
       caption = "Values clipped to 1st-99th percentile per trait for display. Pseudo-log y-axis.") +
  theme_classic(base_size = 10) +
  theme(plot.title   = element_text(face = "bold"),
        axis.text.x  = element_text(angle = 30, hjust = 1),
        legend.position = "bottom",
        strip.text   = element_text(size = 8))
```

---

## 5 Taxonomic Coverage {#taxonomic-coverage}

```{r taxonomic-setup}
message(Sys.time(), " - loaded [taxonomic-coverage]")
taxa_tbl <- all_dat[!is.na(scrubbed_species_binomial) & nzchar(scrubbed_species_binomial),
  .(Rows = .N, `Distinct Traits` = uniqueN(trait_name)),
  by = scrubbed_species_binomial][order(-Rows)]
```

### 5.1 Top 25 taxa by row count {#top25-taxa}

```{r top25-taxa}
kable_styled(
  head(taxa_tbl, 25L)[, .(
    `Species (scrubbed)` = scrubbed_species_binomial,
    Rows,
    `Distinct Traits`
  )],
  caption = "Top 25 taxa by row count (all providers)"
)
```

### 5.2 Distinct taxa per provider {#taxa-by-provider}

```{r fig-taxa-provider, fig.height=4}
taxa_prov <- data.table(
  Provider        = provider_summary$Provider,
  `Distinct Taxa` = provider_summary$`Distinct Taxa`
)
ggplot(taxa_prov, aes(x = reorder(Provider, -`Distinct Taxa`),
                      y = `Distinct Taxa` / 1e3, fill = Provider)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = formatC(`Distinct Taxa`, format = "d", big.mark = ",")),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = provider_colors) +
  labs(x = NULL, y = "Distinct taxa (thousands)",
       title = "Distinct taxa per provider") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 5.3 Country breakdown (top 20) {#country-breakdown}

```{r fig-country, fig.height=5}
country_tbl <- all_dat[,
  .(country = data.table::fifelse(
      is.na(country) | !nzchar(trimws(country)), "Unknown", country
    ))][, .(Rows = .N), by = country][order(-Rows)]
top20_ctry  <- head(country_tbl, 20L)
ggplot(top20_ctry, aes(x = reorder(country, Rows), y = Rows / 1e3)) +
  geom_col(fill = "#4e79a7") +
  coord_flip() +
  labs(x = NULL, y = "Observations (thousands)",
       title = "Top 20 countries by row count") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

### 5.4 Date collected - year distribution {#date-histogram}

```{r fig-date-histogram, fig.height=4}
message(Sys.time(), " - loaded [date-histogram]")

if (requireNamespace("lubridate", quietly = TRUE)) {
  dates_raw <- all_dat[!is.na(date_collected) & nzchar(trimws(date_collected)), date_collected]
  parsed    <- lubridate::parse_date_time(
    dates_raw,
    orders = c("Ymd", "Y", "mdY", "dmy"),
    quiet  = TRUE
  )
  years <- lubridate::year(parsed)
  years <- years[is.finite(years) & years >= 1800 & years <= as.integer(format(Sys.Date(), "%Y"))]
  if (length(years) > 0) {
    year_df <- data.frame(year = years)
    print(
      ggplot(year_df, aes(x = year)) +
        geom_histogram(binwidth = 1, fill = "#59a14f", color = "white", linewidth = 0.2) +
        labs(x = "Year collected", y = "Records",
             title = "Records by year of collection") +
        theme_classic(base_size = 12) +
        theme(plot.title = element_text(face = "bold"))
    )
  } else {
    cat("> No parseable dates found in the data.\n")
  }
} else {
  cat("> **Note:** `lubridate` is not installed. Date histogram skipped.",
      "Install with `install.packages('lubridate')` and re-render.\n")
}
```

---

## 6 Unit Reconciliation Summary {#unit-reconciliation}

```{r unit-summary}
message(Sys.time(), " - loaded [unit-reconciliation]")

unit_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  iu  <- d$inferred_unit
  iuc <- trimws(as.character(d$inferred_unit_confidence))
  iuc[is.na(iuc) | !nzchar(iuc)] <- NA_character_
  list(
    Provider             = provider_labels[[k]],
    `Total Rows`         = nrow(d),
    `Inferred TRUE`      = sum(iu %in% c(TRUE, "TRUE",  "true",  1L), na.rm = TRUE),
    `Inferred FALSE`     = sum(iu %in% c(FALSE,"FALSE", "false", 0L), na.rm = TRUE),
    `Conf: high`         = sum(iuc == "high",   na.rm = TRUE),
    `Conf: medium`       = sum(iuc == "medium",  na.rm = TRUE),
    `Conf: low`          = sum(iuc == "low",     na.rm = TRUE),
    `Conf: none/missing` = sum(is.na(iuc))
  )
}))
kable_styled(unit_summary, caption = "6.1 Unit inference summary by provider")
```

### 6.2 inferred_unit_confidence breakdown {#confidence-breakdown}

```{r confidence-breakdown}
conf_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d   <- dat_list[[k]]
  lvl <- trimws(as.character(d$inferred_unit_confidence))
  lvl[is.na(lvl) | !nzchar(lvl)] <- "(missing)"
  tbl <- d[, .N, by = .(inferred_unit_confidence = lvl)][order(-N)]
  tbl[, provider := provider_labels[[k]]]
  tbl
}))
conf_wide <- dcast(conf_rows, inferred_unit_confidence ~ provider, value.var = "N", fill = 0L)
kable_styled(conf_wide, caption = "inferred_unit_confidence counts by provider (wide)")
```

### 6.3 basis_type breakdown {#basis-type}

```{r basis-type-breakdown}
bt_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d  <- dat_list[[k]]
  bt <- trimws(as.character(d$basis_type))
  bt[is.na(bt) | !nzchar(bt)] <- "(empty)"
  tbl <- d[, .N, by = .(basis_type = bt)][order(-N)]
  tbl[, provider := provider_labels[[k]]]
  tbl
}))
bt_wide <- dcast(bt_rows, basis_type ~ provider, value.var = "N", fill = 0L)
kable_styled(bt_wide, caption = "basis_type counts by provider (wide)")
```

### 6.4 Note on AusTraits / Scientific Data unit confidence {#austraits-note}

<div class="alert alert-warning" role="alert">
<strong>Pipeline labeling note - Scientific Data "none" unit confidence:</strong>
Scientific Data rows show <code>inferred_unit_confidence = "none"</code> because the simple-schema mapper used for this provider did not populate the <code>basis_type</code> or <code>value_type</code> metadata columns that the confidence scorer uses as inputs.
AusTraits (the dominant Scientific Data source) is a fully curated, peer-reviewed database (Falster et al. 2021, doi:10.1038/s41597-021-01006-6).
Its units were set by the original data contributors and validated during publication.
The <strong>"none" label reflects a pipeline labeling gap, not a unit quality failure.</strong>
</div>

---

## 7 Geographic Coverage - Interactive Map {#geographic-coverage}

```{r geo-coverage-table}
message(Sys.time(), " - loaded [geographic-coverage]")

geo_summary <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  list(
    Provider           = provider_labels[[k]],
    `Total Rows`       = nrow(d),
    `Rows with Coords` = sum(d$has_coord, na.rm = TRUE),
    `Coverage (%)`     = round(100 * mean(d$has_coord, na.rm = TRUE), 1)
  )
}))
kable_styled(geo_summary, caption = "7.1 Geographic coordinate coverage by provider")
```

### 7.2 Interactive map (sampled) {#interactive-map}

```{r leaflet-map, fig.height=6}
message(Sys.time(), " - loaded [leaflet-map]")

if (requireNamespace("leaflet", quietly = TRUE)) {
  coord_dat <- all_dat[has_coord == TRUE,
    .(lat_num, lon_num, provider, scrubbed_species_binomial,
      trait_name, trait_value, unit, source_title, country)]

  total_georef <- nrow(coord_dat)

  if (total_georef == 0L) {
    cat(
      "> **Map note:** No valid coordinate rows are available after coordinate filtering. Displaying a base map with no points.\n\n"
    )
    leaflet::leaflet() |>
      leaflet::addTiles()
  } else {

    set.seed(42)
    samp_list <- lapply(unique(coord_dat$provider), function(prov) {
      sub <- coord_dat[provider == prov]
      if (nrow(sub) > 10000) sub <- sub[sample(.N, 10000)]
      sub
    })
    map_dat <- rbindlist(samp_list, fill = TRUE)
    plotted_n <- nrow(map_dat)

    if (plotted_n == 0L || !all(c("lat_num", "lon_num", "provider") %in% names(map_dat))) {
      cat(
        "> **Map note:** No mappable sampled points were produced after filtering/sampling. Displaying a base map with no points.",
        "Total georeferenced records:", formatC(total_georef, format = "d", big.mark = ","),
        "| Plotted points:", formatC(plotted_n, format = "d", big.mark = ","), "\n\n"
      )
      leaflet::leaflet() |>
        leaflet::addTiles()
    } else {

      esc <- function(x) htmltools::htmlEscape(ifelse(is.na(x), "", as.character(x)))
      map_dat[, popup_text := paste0(
        "<b>", esc(scrubbed_species_binomial), "</b><br>",
        ifelse(is.na(trait_name), "", paste0(esc(trait_name), ": ")),
        esc(trait_value), " ", esc(unit), "<br>",
        "Source: ", esc(source_title), "<br>",
        "Country: ", esc(country)
      )]

      map_providers <- unique(map_dat$provider)
      pal <- leaflet::colorFactor(
        palette = c("DataDryad" = "#1f77b4", "Zenodo" = "#ff7f0e",
                    "Scientific Data" = "#2ca02c"),
        domain  = c("DataDryad", "Zenodo", "Scientific Data"),
        na.color = "#999999"
      )

      map_out <- leaflet::leaflet(map_dat) |>
        leaflet::addTiles() |>
        leaflet::addCircleMarkers(
          lng         = ~lon_num,
          lat         = ~lat_num,
          radius      = 4,
          color       = ~pal(provider),
          fillColor   = ~pal(provider),
          opacity     = 0.7,
          fillOpacity = 0.5,
          popup       = ~popup_text
        ) |>
        leaflet::addLegend(
          position = "bottomright",
          pal      = pal,
          values   = ~provider,
          title    = "Provider"
        )

      cat("> **Map note:** Displaying a random sample of up to 10,000 records per provider for performance.",
        "Total georeferenced records:", formatC(total_georef, format = "d", big.mark = ","),
        "| Plotted points:", formatC(plotted_n, format = "d", big.mark = ","), "\n\n")

      map_out
    }
  }
} else {
  cat("> **Note:** `leaflet` is not installed. Interactive map skipped.",
      "Install with `install.packages('leaflet')` and re-render.\n")
}
```

### 7.3 Elevation distribution {#elevation}

```{r fig-elevation, fig.height=4}
elev_vals <- if ("elevation_m" %in% names(all_dat)) {
  suppressWarnings(as.numeric(all_dat$elevation_m))
} else {
  numeric(0)
}
elev_valid <- elev_vals[is.finite(elev_vals) & elev_vals >= -500 & elev_vals <= 6000]
if (length(elev_valid) > 0) {
  elev_df <- data.frame(elevation_m = elev_valid)
  ggplot(elev_df, aes(x = elevation_m)) +
    geom_histogram(bins = 60, fill = "#76b7b2", color = "white", linewidth = 0.2) +
    labs(x = "Elevation (m)", y = "Records",
         title = "Records by elevation",
         caption = "Filtered to -500 to 6000 m range.") +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(face = "bold"))
} else {
  cat("> No valid elevation values found in the range -500 to 6000 m.\n")
}
```

---

## 8 BIEN Database Mapping Status {#bien-mapping}

```{r bien-mapping-setup}
message(Sys.time(), " - loaded [bien-mapping]")

bien_trait_map <- data.table(
  trait_canonical = c(
    "leaf area", "specific leaf area", "leaf nitrogen content per leaf dry mass",
    "leaf phosphorus content per leaf dry mass", "stem specific density",
    "plant height vegetative", "leaf dry mass", "seed mass",
    "leaf carbon content per leaf dry mass", "maximum height vegetative",
    "wood density", "leaf thickness", "root depth",
    "stomatal conductance", "leaf water content per leaf dry mass",
    "leaf lifespan", "leaf chlorophyll content",
    "bark thickness", "vessel diameter", "conduit diameter lumen"
  ),
  bien_name = c(
    "leaf area", "specific leaf area", "leaf nitrogen content per leaf dry mass",
    "leaf phosphorus content per leaf dry mass", "stem specific density",
    "plant height vegetative", "leaf dry mass", "seed mass",
    "leaf carbon content per leaf dry mass", "maximum height vegetative",
    "wood density", "leaf thickness", "root depth",
    "stomatal conductance", "leaf water content per leaf dry mass",
    "leaf lifespan", "leaf chlorophyll content",
    "bark thickness", "vessel diameter", "conduit diameter lumen"
  )
)

all_dat[, trait_name_lower := trimws(tolower(trait_name))]
bien_set <- tolower(trimws(bien_trait_map$trait_canonical))
all_dat[, is_bien_mappable := trait_name_lower %in% bien_set]
```

### 8.1-8.2 BIEN-mappable rows per provider {#bien-table}

```{r bien-rows-table}
bien_prov <- all_dat[is_bien_mappable == TRUE,
  .(Rows = .N), by = .(bien_trait = trait_name_lower, provider)]
bien_wide <- dcast(bien_prov, bien_trait ~ provider, value.var = "Rows", fill = 0L)
bien_wide[, Total := rowSums(.SD, na.rm = TRUE),
          .SDcols = setdiff(names(bien_wide), "bien_trait")]
setorder(bien_wide, -Total)
setnames(bien_wide, "bien_trait", "BIEN Trait")
kable_styled(bien_wide, caption = "8.2 BIEN-mappable rows by trait and provider (exact canonical match)")
```

### 8.3 BIEN-mappable rows by provider {#bien-bar}

```{r fig-bien-bar, fig.height=4}
bien_summary <- all_dat[, .(
  Total         = .N,
  BIEN_mappable = sum(is_bien_mappable, na.rm = TRUE),
  Pct_mappable  = round(100 * mean(is_bien_mappable, na.rm = TRUE), 1)
), by = provider]

ggplot(bien_summary, aes(x = reorder(provider, -BIEN_mappable),
                         y = BIEN_mappable / 1e3, fill = provider)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = paste0(formatC(BIEN_mappable, format = "d", big.mark = ","),
                               "\n(", Pct_mappable, "%)")),
            vjust = -0.3, size = 3.2) +
  scale_fill_manual(values = provider_colors) +
  labs(x = NULL, y = "BIEN-mappable rows (thousands)",
       title = "8.3 BIEN-mappable rows by provider") +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 8.4 Percentage of rows that are BIEN-mappable {#bien-pct}

```{r bien-pct-table}
kable_styled(
  bien_summary[, .(Provider = provider,
                   `Total Rows` = Total,
                   `BIEN-Mappable Rows` = BIEN_mappable,
                   `% Mappable` = Pct_mappable)],
  caption = "8.4 BIEN-mappable percentage by provider"
)
```

> **8.5 Coverage note:** Traits not yet in this lookup are not necessarily unmappable - they may require fuzzy matching or manual review. This table represents **exact canonical name matches only** (case-insensitive, whitespace-trimmed). Full BIEN trait vocabulary integration is a planned pipeline milestone.

---

## 9 QA Flag Summary {#qa-flags}

```{r qa-summary}
message(Sys.time(), " - loaded [qa-flags]")

qa_rows <- rbindlist(lapply(names(dat_list), function(k) {
  d <- dat_list[[k]]
  flagged <- sum(nzchar(trimws(d$qa_flags)), na.rm = TRUE)
  list(
    Provider         = provider_labels[[k]],
    `Total Rows`     = nrow(d),
    `Flagged Rows`   = flagged,
    `Unflagged Rows` = nrow(d) - flagged,
    `Flag Rate (%)`  = round(100 * flagged / nrow(d), 1)
  )
}))
kable_styled(qa_rows, caption = "9.1 QA flag prevalence by provider")
```

### 9.2 DataDryad top 15 QA flag codes {#qa-dryad-top}

```{r qa-dryad-top}
d_dry        <- dat_list[["dryad"]]
flagged_vals <- d_dry$qa_flags[nzchar(trimws(d_dry$qa_flags))]
flag_tokens  <- unlist(strsplit(flagged_vals, "\\|"))
flag_tokens  <- trimws(sub("\\[.*", "", flag_tokens))
flag_tokens  <- flag_tokens[nzchar(flag_tokens)]
flag_tbl_all <- sort(table(flag_tokens), decreasing = TRUE)
flag_top15   <- as.data.frame(
  flag_tbl_all[seq_len(min(15L, length(flag_tbl_all)))]
)
names(flag_top15) <- c("QA Flag Code", "Count")

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    flag_top15,
    extensions = "Scroller",
    options    = list(scrollY = 300, scroller = TRUE, scrollX = TRUE, dom = "t"),
    rownames   = FALSE,
    caption    = "9.2 DataDryad: top 15 QA flag codes (prefix only)"
  )
} else {
  kable_styled(flag_top15, caption = "9.2 DataDryad: top 15 QA flag codes")
}
```

### 9.3 Top 15 QA flag codes - bar chart {#qa-bar}

```{r fig-qa-bar, fig.height=4}
flag_df <- flag_top15
names(flag_df) <- c("flag_code", "count")
ggplot(flag_df, aes(x = reorder(flag_code, count), y = count / 1e3)) +
  geom_col(fill = "#e15759") +
  coord_flip() +
  labs(x = "QA flag code", y = "Occurrences (thousands)",
       title = "9.3 DataDryad - top 15 QA flag codes") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 10 Pending Sources — Manual Intake Queue {#pending-sources}

These are user-supplied sources queued for review and potential ingestion into the pipeline. They are tracked in `data/manual_source_intake.csv` and have not yet been formally harvested or deduplicated against existing provider outputs.

```{r pending-sources-table, message=FALSE}
message(Sys.time(), " - loaded [pending-sources-table]")
manual_intake <- data.table::fread(file.path(proj_root, "data", "manual_source_intake.csv"))
cols_show <- c("display_name", "source_group", "canonical_url_or_doi",
               "related_publication", "harvest_status", "next_action")
DT::datatable(manual_intake[, ..cols_show], rownames = FALSE, options = list(scrollX = TRUE))
```

All 6 sources are at status `pending_review`. Three are files supplied directly by Jesse Miller (data contact for the Spasojevic et al. 2016 Ozark trait study). One is the ShrubHub TraitHub GitHub repository (Bjorkman et al. 2018 Nature). One is email context from the same Jesse Miller. One is field data from Pablo Peri's BEPSP Patagonia network (Peri et al. 2016 J. Nature Conservation).

---

## 11 Session Info {#session-info}

```{r session-info}
message(Sys.time(), " - loaded [session-info]")

si <- sessionInfo()
cat("Rendered:    ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), "\n")
cat("R version:   ", si$R.version$version.string, "\n")
cat("data.table:  ", as.character(packageVersion("data.table")), "\n")
cat("ggplot2:     ", as.character(packageVersion("ggplot2")), "\n")
cat("leaflet:     ",
    tryCatch(as.character(packageVersion("leaflet")),    error = function(e) "not installed"), "\n")
cat("DT:          ",
    tryCatch(as.character(packageVersion("DT")),         error = function(e) "not installed"), "\n")
cat("kableExtra:  ",
    tryCatch(as.character(packageVersion("kableExtra")), error = function(e) "not installed"), "\n")
cat("scales:      ",
    tryCatch(as.character(packageVersion("scales")),     error = function(e) "not installed"), "\n")
cat("lubridate:   ",
    tryCatch(as.character(packageVersion("lubridate")),  error = function(e) "not installed"), "\n")
```
