---
title: "DryadPlantTraits: Trait Harvest Summary"
author: "DryadPlantTraits Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Detect project root from Rmd location
proj_root <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
  mustWork = FALSE
)

source(file.path(proj_root, "R", "dryad_api.R"))
source(file.path(proj_root, "R", "search_terms.R"))
source(file.path(proj_root, "R", "candidate_filter.R"))
source(file.path(proj_root, "R", "trait_dictionary.R"))
source(file.path(proj_root, "R", "standardize_records.R"))

kable_styled <- function(df, caption = NULL) {
  tbl <- knitr::kable(df, caption = caption, format = "html")
  result <- tryCatch(
    kableExtra::kable_styling(tbl, bootstrap_options = c("striped", "hover", "condensed"),
                              full_width = FALSE),
    error = function(e) tbl
  )
  result
}
```

## Overview

**DryadPlantTraits** is a pipeline for discovering, downloading, and standardising plant functional trait data from [Dryad](https://datadryad.org). It targets ten well-studied traits drawn from an AusTraits-inspired vocabulary (leaf area, SLA, LDMC, leaf N/P, wood density, stem specific density, seed mass, plant height, leaf thickness) and maps them to the BIEN observation model for direct upload into the BIEN plant trait database.

The pipeline proceeds in three stages:

1. **Discovery** — `discover_dryad_plant_traits.R` searches the Dryad API with keyword seed terms and scores candidate datasets using `R/candidate_filter.R`.
2. **Download & compile** — `compile_downloaded_traits.R` downloads scored datasets (requires `DRYAD_API_TOKEN`), extracts trait columns via `R/trait_dictionary.R`, and writes `output/compiled_trait_observations.csv` in the BIEN schema.
3. **Upload** — the compiled CSV is validated and submitted to BIEN using the schema defined in `output/bien_trait_upload_template.csv`.

---

## Trait Dictionary

```{r trait-dictionary}
trait_dict <- tryCatch(
  dryad_read_trait_dictionary(),
  error = function(e) {
    message("Could not load trait dictionary: ", conditionMessage(e))
    data.frame()
  }
)

if (nrow(trait_dict) > 0) {
  kable_styled(trait_dict, caption = "Trait dictionary (data/trait_dictionary_starter.csv)")
} else {
  cat("Trait dictionary unavailable.")
}
```

---

## Harvestable Traits by Functional Category

```{r traits-by-category}
trait_categories <- data.frame(
  `Functional Category` = c(
    "leaf_economics", "leaf_economics",
    "leaf_morphology", "leaf_morphology",
    "leaf_chemistry", "leaf_chemistry",
    "stem_structure", "stem_structure",
    "reproduction",
    "size_structure"
  ),
  `Trait Name` = c(
    "specific_leaf_area", "leaf_dry_matter_content",
    "leaf_area", "leaf_thickness",
    "leaf_n", "leaf_p",
    "wood_density", "stem_specific_density",
    "seed_mass",
    "plant_height"
  ),
  `Standard Unit` = c(
    "mm2/mg", "mg/g",
    "mm2", "mm",
    "mg/g", "mg/g",
    "g/cm3", "g/cm3",
    "mg",
    "m"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

# Add estimated counts from API (populated in next chunk; placeholder here)
trait_categories$`Estimated Dryad Dataset Count` <- NA_integer_

kable_styled(trait_categories, caption = "Harvestable traits by functional category")
```

---

## Dryad Search Coverage

```{r dryad-search-coverage}
seed_terms <- dryad_search_seed_terms()

counts <- vapply(seed_terms$query_term, function(term) {
  result <- tryCatch({
    parsed <- dryad_search_datasets(q = term, per_page = 1L, page = 1L)
    total <- parsed[["total"]]
    if (is.null(total)) NA_integer_ else as.integer(total)
  }, error = function(e) NA_integer_)
  Sys.sleep(0.5)
  result
}, integer(1))

coverage <- data.frame(
  `Query Term`              = seed_terms$query_term,
  `Theme`                   = seed_terms$theme,
  `Trait Focus`             = seed_terms$trait_focus,
  `Estimated Dataset Count` = counts,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

na_count <- sum(is.na(coverage$`Estimated Dataset Count`))
if (na_count > 0) {
  cat(sprintf(
    "_Note: %d of %d API queries returned NA — likely a network timeout or rate-limit; counts will populate when run with connectivity._\n\n",
    na_count, nrow(coverage)
  ))
}

kable_styled(coverage, caption = "Dryad API search coverage by seed term")
```

---

## BIEN Upload Schema

The primary compiled output is `output/compiled_trait_observations.csv`. It follows the BIEN observation model below. Use `output/bien_trait_upload_template.csv` as a zero-row header-only reference for downstream validation.

```{r bien-schema}
bien_schema <- data.frame(
  Column = c(
    "scrubbed_species_binomial",
    "latitude", "longitude",
    "date_collected",
    "country", "stateProvince", "county", "locality",
    "trait_name", "trait_value", "unit",
    "dataset", "datasource", "dataowner",
    "source_citation",
    "dryad_dataset_doi", "dryad_version_id", "dryad_file_id",
    "download_timestamp_utc",
    "source_file_path",
    "source_column_trait_name", "source_column_trait_value", "source_column_unit",
    "raw_scrubbed_species_binomial",
    "raw_latitude", "raw_longitude", "raw_date_collected",
    "raw_trait_name", "raw_trait_value", "raw_unit"
  ),
  Description = c(
    "TNRS-resolved binomial; primary taxonomic key",
    "Decimal degrees (WGS84)", "Decimal degrees (WGS84)",
    "ISO 8601 date of collection",
    "Country name", "State or province", "County", "Verbatim locality string",
    "Standardised trait name (AusTraits vocabulary)", "Numeric trait value (standardised unit)", "Standard unit string",
    "Dryad dataset title", "Dryad (datasource label)", "Dataset owner / PI",
    "Full bibliographic citation",
    "Dryad dataset DOI", "Dryad version integer ID", "Dryad file integer ID",
    "UTC timestamp of file download (ISO 8601)",
    "Relative path to source file within downloaded archive",
    "Original column name mapped to trait", "Original column name mapped to value", "Original column name mapped to unit",
    "Original (pre-TNRS) species string",
    "Original latitude", "Original longitude", "Original date string",
    "Original trait name string", "Original trait value string", "Original unit string"
  ),
  Source = c(
    "standardization",
    "source data", "source data",
    "source data",
    "source data", "source data", "source data", "source data",
    "standardization", "standardization", "standardization",
    "Dryad metadata", "Dryad metadata", "Dryad metadata",
    "Dryad metadata",
    "Dryad metadata", "Dryad metadata", "Dryad metadata",
    "pipeline",
    "pipeline",
    "standardization", "standardization", "standardization",
    "source data",
    "source data", "source data", "source data",
    "source data", "source data", "source data"
  ),
  stringsAsFactors = FALSE
)

kable_styled(bien_schema, caption = "BIEN observation model — 30-column upload schema")
```

---

## Next Steps

1. **Build candidate list** — run `scripts/discover_dryad_plant_traits.R` (no token needed). This writes `output/candidate_datasets.csv` with scored Dryad datasets.

2. **Download and compile** — set `DRYAD_API_TOKEN` (obtain from [datadryad.org/stash/user_account](https://datadryad.org/stash/user_account)), then run `scripts/compile_downloaded_traits.R`. This writes `output/compiled_trait_observations.csv`.

3. **Validate** — compare the compiled CSV columns against `output/bien_trait_upload_template.csv` and inspect record counts, trait coverage, and any missing species binomials before BIEN submission.
