---
title: "sPlotOpen v2.0 — Global Vegetation Plot Database Overview"
author: "splot-open-data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(leaflet)
  library(knitr)
  library(lubridate)
})

ZIP <- here::here("data", "iDiv Data Repository_3474_v55__20260429.zip")
# Fallback to absolute path if here is unavailable
if (!requireNamespace("here", quietly = TRUE)) {
  ZIP <- "/Users/brianjenquist/VSCode/splot-open-data/data/iDiv Data Repository_3474_v55__20260429.zip"
}
```

```{r load-data, cache=TRUE}
message("Loading header...")
header <- fread(
  cmd  = paste0("unzip -p '", ZIP, "' 'sPlotOpen_header(2).txt'"),
  sep  = "\t",
  na.strings = c("", "NA", "N/A")
)

message("Loading DT (species×plot records)...")
DT <- fread(
  cmd  = paste0("unzip -p '", ZIP, "' 'sPlotOpen_DT(1).txt'"),
  sep  = "\t",
  na.strings = c("", "NA", "N/A")
)

message("Loading CWM/CWV...")
CWM <- fread(
  cmd  = paste0("unzip -p '", ZIP, "' 'sPlotOpen_CWM_CWV(1).txt'"),
  sep  = "\t",
  na.strings = c("", "NA", "N/A")
)

# Derive year from Date_of_recording
header[, recording_year := year(suppressWarnings(ymd(Date_of_recording)))]
if (all(is.na(header$recording_year))) {
  header[, recording_year := as.integer(substr(as.character(Date_of_recording), 1, 4))]
}
```

---

## 1. Dataset Summary

sPlotOpen is an environmentally-balanced, open-access subset of the full sPlot vegetation-plot database (v2.1), designed to support global macroecological analyses. The dataset covers all major terrestrial biomes and continents, providing standardised species cover/abundance data alongside plot-level community-weighted means (CWMs) for six key plant functional traits.

```{r summary-stats}
n_plots    <- nrow(header)
n_obs      <- nrow(DT)
n_spp      <- uniqueN(DT[!is.na(Species) & Species != "", Species])
date_min   <- suppressWarnings(min(ymd(header$Date_of_recording), na.rm = TRUE))
date_max   <- suppressWarnings(max(ymd(header$Date_of_recording), na.rm = TRUE))
n_datasets <- uniqueN(header$Dataset)

summary_df <- data.frame(
  Metric = c(
    "Number of plots",
    "Species × plot records",
    "Unique species (binomials)",
    "Earliest recording",
    "Latest recording",
    "Contributing datasets"
  ),
  Value = c(
    format(n_plots, big.mark = ","),
    format(n_obs, big.mark = ","),
    format(n_spp, big.mark = ","),
    as.character(date_min),
    as.character(date_max),
    format(n_datasets, big.mark = ",")
  )
)

kable(summary_df, caption = "Table 1. sPlotOpen v2.0 top-level summary statistics.")
```

### Continental breakdown

```{r continent-table}
cont_tbl <- as.data.frame(sort(table(header$Continent), decreasing = TRUE))
names(cont_tbl) <- c("Continent", "Plot count")
cont_tbl[["Plot count"]] <- format(cont_tbl[["Plot count"]], big.mark = ",")
kable(cont_tbl, caption = "Table 2. Plot counts by continent.")
```

### Biome breakdown

```{r biome-table}
biome_tbl <- as.data.frame(sort(table(header$Biome), decreasing = TRUE))
names(biome_tbl) <- c("Biome", "Plot count")
biome_tbl[["Plot count"]] <- format(biome_tbl[["Plot count"]], big.mark = ",")
kable(biome_tbl, caption = "Table 3. Plot counts by biome (Whittaker classification).")
```

---

## 2. Global Plot Map

The map below shows the geographic distribution of all `r format(n_plots, big.mark=",")` sPlotOpen plots, coloured by continent. Marker clustering is applied at low zoom levels to reduce visual overload. The database spans all inhabited continents, with notable density in Europe (reflecting the large number of European vegetation surveys contributing to sPlot) and South America.

```{r leaflet-map}
map_dt <- header[!is.na(Latitude) & !is.na(Longitude),
                 .(Latitude, Longitude, Continent, Dataset)]

continents  <- sort(unique(map_dt$Continent))
cont_pal    <- colorFactor(
  palette = RColorBrewer::brewer.pal(min(length(continents), 8), "Set1"),
  domain  = continents
)

# Subsample for rendering speed (max 40 000 pts in leaflet)
set.seed(42)
plot_sample <- if (nrow(map_dt) > 40000) map_dt[sample(.N, 40000)] else map_dt

leaflet(plot_sample) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    lng        = ~Longitude,
    lat        = ~Latitude,
    color      = ~cont_pal(Continent),
    radius     = 3,
    stroke     = FALSE,
    fillOpacity = 0.6,
    clusterOptions = markerClusterOptions(),
    popup      = ~paste0("<b>Continent:</b> ", Continent,
                         "<br><b>Dataset:</b> ", Dataset)
  ) %>%
  addLegend(
    position = "bottomright",
    pal      = cont_pal,
    values   = ~Continent,
    title    = "Continent"
  ) %>%
  addControl(
    html     = "<b>sPlotOpen v2.0 — Global plot locations</b>",
    position = "topright"
  )
```

> **Note:** Up to 40,000 plots are shown for rendering performance; all `r format(n_plots, big.mark=",")` plots are included in analyses.

---

## 3. Distribution Histograms

### 3.1 Species richness per plot

Species richness (number of species per plot) is highly right-skewed, as is typical in vegetation databases. Most plots contain fewer than 50 species; however, high-richness outliers (tropical forests, species-rich grasslands) extend the tail considerably.

```{r richness-hist}
if ("Species_richness" %in% names(CWM)) {
  rich_dt <- CWM[!is.na(Species_richness), .(Species_richness)]
} else {
  rich_dt <- DT[Species != "" & !is.na(Species), .N, by = PlotObservationID]
  setnames(rich_dt, "N", "Species_richness")
}

ggplot(rich_dt, aes(x = Species_richness)) +
  geom_histogram(bins = 60, fill = "#2c7bb6", colour = "white", linewidth = 0.2) +
  scale_x_log10(labels = scales::comma) +
  labs(
    title = "Species richness per plot (log10 x-axis)",
    x     = "Species richness",
    y     = "Number of plots"
  ) +
  theme_minimal()
```

### 3.2 Year of recording

```{r year-hist}
year_dt <- header[!is.na(recording_year) & recording_year >= 1900 & recording_year <= 2030]

ggplot(year_dt, aes(x = recording_year)) +
  geom_histogram(binwidth = 2, fill = "#1a9641", colour = "white", linewidth = 0.2) +
  labs(
    title = "Year of recording",
    x     = "Year",
    y     = "Number of plots"
  ) +
  theme_minimal()
```

### 3.3 Latitude distribution

```{r lat-hist}
ggplot(header[!is.na(Latitude)], aes(x = Latitude)) +
  geom_histogram(bins = 72, fill = "#d7191c", colour = "white", linewidth = 0.2) +
  labs(
    title = "Latitudinal distribution of plots",
    x     = "Latitude (°)",
    y     = "Number of plots"
  ) +
  theme_minimal()
```

### 3.4 Elevation distribution

Elevation data are present for a subset of plots. The distribution reflects a combination of true montane plots and lowland plots, with a mode near sea level.

```{r elev-hist}
elev_dt <- header[!is.na(Elevation)]
ggplot(elev_dt, aes(x = Elevation)) +
  geom_histogram(bins = 60, fill = "#756bb1", colour = "white", linewidth = 0.2) +
  labs(
    title = paste0("Elevation distribution (n = ", format(nrow(elev_dt), big.mark = ","), " plots with elevation data)"),
    x     = "Elevation (m a.s.l.)",
    y     = "Number of plots"
  ) +
  theme_minimal()
```

### 3.5 Location uncertainty

```{r uncertainty-hist}
unc_dt <- header[!is.na(Location_uncertainty)]
ggplot(unc_dt, aes(x = Location_uncertainty)) +
  geom_histogram(bins = 60, fill = "#fd8d3c", colour = "white", linewidth = 0.2) +
  scale_x_log10(labels = scales::comma) +
  labs(
    title = paste0("Location uncertainty (n = ", format(nrow(unc_dt), big.mark = ","), " plots with uncertainty data; log10 x-axis)"),
    x     = "Location uncertainty (m)",
    y     = "Number of plots"
  ) +
  theme_minimal()
```

---

## 4. Biome Bar Chart

sPlotOpen covers all major terrestrial biomes. Temperate broadleaved and mixed forests dominate the dataset, reflecting the geographic concentration of long-running European and North American vegetation surveys. Tropical biomes are comparatively underrepresented, an important caveat for global analyses.

```{r biome-bar}
biome_counts <- header[, .N, by = Biome][order(-N)]
biome_counts[, Biome := factor(Biome, levels = rev(Biome))]

ggplot(biome_counts, aes(x = Biome, y = N)) +
  geom_col(fill = "#41b6c4") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Plot count by biome",
    x     = NULL,
    y     = "Number of plots"
  ) +
  theme_minimal()
```

---

## 5. Cover Layer Summary by Biome

Plot-level canopy cover data (tree, shrub, herb, and moss/lichen layers) were recorded where available. Mean cover values vary substantially across biomes, reflecting real vegetation-structural differences. Tropical and temperate forests show high tree-layer cover; grasslands and shrublands are characterised by herb and shrub dominance.

```{r cover-layers}
cover_cols <- c("Tree_layer", "Shrub_layer", "Herb_layer", "Moss_layer")
present_cols <- intersect(cover_cols, names(header))

if (length(present_cols) > 0) {
  cover_dt <- header[, c("Biome", present_cols), with = FALSE]
  cover_long <- melt(cover_dt, id.vars = "Biome",
                     measure.vars = present_cols,
                     variable.name = "Layer", value.name = "Cover")
  cover_long <- cover_long[!is.na(Cover)]
  cover_summary <- cover_long[, .(mean_cover = mean(Cover, na.rm = TRUE)), by = .(Biome, Layer)]

  # Shorten long biome names for readability
  cover_summary[, Biome_short := gsub("Tropical & Subtropical ", "Trop. ", Biome)]
  cover_summary[, Biome_short := gsub("Temperate ", "Temp. ", Biome_short)]
  cover_summary[, Biome_short := gsub(" Forests", " For.", Biome_short)]
  cover_summary[, Biome_short := gsub(" Grasslands, Savannas & Shrublands", " Grass./Shrub.", Biome_short)]

  ggplot(cover_summary, aes(x = Biome_short, y = mean_cover, fill = Layer)) +
    geom_col(position = "dodge") +
    coord_flip() +
    facet_wrap(~Layer, scales = "free_x", ncol = 2) +
    scale_fill_brewer(palette = "Set2", guide = "none") +
    labs(
      title = "Mean cover (%) per vegetation layer by biome",
      x     = NULL,
      y     = "Mean cover (%)"
    ) +
    theme_minimal(base_size = 9)
} else {
  cat("Cover layer columns not found in header; skipping this section.\n")
}
```

---

## 6. Trait Coverage (Community-Weighted Means)

Community-weighted means (CWMs) for key plant functional traits were computed from TRY trait data matched to species in each plot. Coverage is not 100% because not all species in a plot have trait data available in TRY. SLA and plant height tend to have the broadest coverage, while seed mass and leaf phosphorus have more limited coverage.

```{r trait-coverage}
trait_cols <- c("SLA_CWM", "PlantHeight_CWM", "LeafN_CWM",
                "LeafP_CWM", "SeedMass_CWM", "LeafArea_CWM", "LDMC_CWM")
present_traits <- intersect(trait_cols, names(CWM))

if (length(present_traits) > 0) {
  n_cwm <- nrow(CWM)
  trait_cov <- data.table(
    Trait   = present_traits,
    pct_non_na = sapply(present_traits, function(col) 100 * sum(!is.na(CWM[[col]])) / n_cwm)
  )
  trait_cov[, Trait := factor(Trait, levels = Trait[order(pct_non_na)])]

  ggplot(trait_cov, aes(x = Trait, y = pct_non_na)) +
    geom_col(fill = "#2ca25f") +
    coord_flip() +
    scale_y_continuous(limits = c(0, 100)) +
    labs(
      title = "Percentage of plots with non-NA CWM value",
      x     = NULL,
      y     = "% plots covered"
    ) +
    theme_minimal()
} else {
  cat("No recognised CWM columns found; skipping trait coverage chart.\n")
}
```

---

## 7. Resample Balance

sPlotOpen provides three resample sets (Resample_1, Resample_2, Resample_3) designed to achieve global environmental balance using a density-inverse weighting approach. `Resample_1_consensus` marks plots selected in the majority (≥2) of the three resample sets — the recommended subset for analyses requiring geographic balance.

```{r resample-balance}
if ("Resample_1_consensus" %in% names(header)) {
  rs_tbl <- header[, .(count = .N), by = .(Resample_1_consensus)]
  rs_tbl[, Resample_1_consensus := ifelse(is.na(Resample_1_consensus), "NA / not selected",
                                          ifelse(Resample_1_consensus, "TRUE (consensus selected)", "FALSE"))]
  setnames(rs_tbl, c("Resample_1_consensus", "count"), c("Resample_1_consensus", "Plot count"))
  rs_tbl[, `Plot count` := format(`Plot count`, big.mark = ",")]
  kable(rs_tbl, caption = "Table 4. Resample_1_consensus — geographically-balanced subset indicator.")
} else {
  cat("Resample_1_consensus column not found in header.\n")
}
```

---

## Session Info

```{r session-info}
sessionInfo()
```
