---
title: "Cacti Trait Data: Provenance and Current Status"
date: "2026-03-28"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

## Purpose

This report summarizes where current cacti trait data came from, what is available now, and how records are distributed across sources and traits.

## Data Inputs

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(knitr)
library(stringr)

run_summary <- read_csv("data_processed/run_summary.csv", show_col_types = FALSE)
non_bien_log <- read_csv("data_processed/non_bien_sources_log.csv", show_col_types = FALSE)
combined <- read_csv("data_processed/cacti_traits_combined.csv", show_col_types = FALSE)
bien_target <- read_csv("data_processed/bien_cacti_target_traits.csv", show_col_types = FALSE)
wd_raw <- read_csv("data_raw/wikidata_cacti_traits.csv", show_col_types = FALSE)
accepted_species <- read_csv("data_processed/bien_cactaceae_species_accepted.csv", show_col_types = FALSE)
```

## Where The Data Came From

### BIEN

- Source system: BIEN (Botanical Information and Ecology Network)
- Access method: BIEN R package
- Family backbone pulled: Cactaceae taxonomy and accepted species list
- Trait extraction focused on: body size (height/diameter/mass-like fields) and flower color

### Non-BIEN Internet Sources

```{r}
kable(non_bien_log, caption = "Non-BIEN source log (explicit provenance)")
```

## Current Coverage Snapshot

```{r}
kable(run_summary, caption = "Pipeline run summary")
```

```{r}
coverage <- tibble(
  metric = c(
    "Accepted cacti species in BIEN taxonomy",
    "Unique species with BIEN target-trait records",
    "Unique species in combined target-trait dataset",
    "Coverage of accepted species in combined dataset"
  ),
  value = c(
    nrow(accepted_species),
    bien_target %>% filter(!is.na(species), species != "") %>% distinct(species) %>% nrow(),
    combined %>% distinct(species) %>% nrow(),
    paste0(round(100 * (combined %>% distinct(species) %>% nrow()) / nrow(accepted_species), 2), "%")
  )
)

kable(coverage, caption = "Species-level coverage")
```

## Source Breakdown In Combined Dataset

```{r}
by_source <- combined %>%
  group_by(source_dataset) %>%
  summarise(
    records = n(),
    unique_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  arrange(desc(records))

kable(by_source, caption = "Records and species by source")
```

```{r}
by_source_trait <- combined %>%
  group_by(source_dataset, trait_group) %>%
  summarise(
    records = n(),
    unique_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  arrange(source_dataset, desc(records))

kable(by_source_trait, caption = "Records by source and trait group")
```

## Notes On Why Counts Differ

- BIEN taxonomy includes many accepted cacti species with no current records for the requested traits.
- BIEN target-trait records include rows with missing species names that are dropped when building the combined species-level file.
- Supplemental internet data (Wikidata) is currently sparse and mostly flower-color entries.

## Reproducibility

To refresh the underlying data files used by this report, run:

```bash
PATH="/opt/homebrew/bin:$PATH" Rscript cacti/scripts/fetch_cacti_traits.R
```

Then re-render this report.
