---
title: "Random BIEN Species: Multi-Species Climate and Multivariate Analysis"
author: "BIEN Climate Workflow"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

## Analysis Objective

This report samples several random species from BIEN, keeps species with at least 50 cleaned occurrence records, extracts WorldClim climate variables, and performs multivariate climate analyses.

The response model choice for constrained ordination follows gradient-length guidance from direct gradient analysis: use DCA axis-1 length as the diagnostic for linear (RDA) versus unimodal (CCA) methods.

- ter Braak, C.J.F. and Prentice, I.C. (1988). A theory of gradient analysis. Advances in Ecological Research 18:271-317. https://doi.org/10.1016/S0065-2504(08)60183-X
- ter Braak, C.J.F. (1986). Canonical Correspondence Analysis: a new eigenvector technique for multivariate direct gradient analysis. Ecology 67:1167-1179. https://doi.org/10.2307/1938672

## Data Sources and Method Citations

- BIEN data access:
  Maitner, B.S., et al. (2018). The BIEN R package: A tool to access the Botanical Information and Ecology Network (BIEN) database. Methods in Ecology and Evolution 9:373-379. https://doi.org/10.1111/2041-210X.12861
- WorldClim climate layers:
  Fick, S.E. and Hijmans, R.J. (2017). WorldClim 2: New 1-km spatial resolution climate surfaces for global land areas. International Journal of Climatology 37:4302-4315. https://doi.org/10.1002/joc.5086
- Hellinger transformation before linear ordination:
  Legendre, P. and Gallagher, E.D. (2001). Ecologically meaningful transformations for ordination of species data. Oecologia 129:271-280. https://doi.org/10.1007/s004420100716
- Permutation-based forward selection rationale:
  Blanchet, F.G., Legendre, P. and Borcard, D. (2008). Forward selection of explanatory variables. Ecology 89:2623-2632. https://doi.org/10.1890/07-0986.1

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(ggplot2)
library(tidyr)
library(vegan)

source("R/utils.R")
source("R/data_access.R")
source("R/cleaning.R")
source("R/climate.R")

cfg <- read_config("config.yml")
set.seed(as.integer(cfg$seed))

ensure_directory("outputs")
ensure_directory("outputs/multispecies")
```

## 1. Random Species Selection (>= 50 Cleaned Occurrences)

```{r select-species}
min_clean_records <- 50L
target_species_n <- 5L
max_attempts <- 80L

species_pool <- get_species_pool(cfg)
tried <- character(0)
selected_list <- list()
attempt_log <- list()

for (i in seq_len(max_attempts)) {
  available <- setdiff(species_pool, tried)
  if (length(available) == 0 || length(selected_list) >= target_species_n) break

  sp <- sample(available, size = 1)
  tried <- c(tried, sp)

  rec <- tryCatch({
    raw <- fetch_occurrences(sp, occurrence_limit = as.integer(cfg$occurrence_limit))
    cln <- qa_clean_occurrences(raw)
    list(ok = TRUE, raw = raw, cleaned = cln)
  }, error = function(e) {
    list(ok = FALSE, msg = conditionMessage(e))
  })

  if (!isTRUE(rec$ok)) {
    attempt_log[[length(attempt_log) + 1]] <- data.frame(
      attempt = i,
      species = sp,
      n_raw = 0L,
      n_clean = 0L,
      eligible = FALSE,
      message = rec$msg,
      stringsAsFactors = FALSE
    )
    next
  }

  n_raw <- nrow(rec$raw)
  n_clean <- nrow(rec$cleaned$data)
  eligible <- n_clean >= min_clean_records

  attempt_log[[length(attempt_log) + 1]] <- data.frame(
    attempt = i,
    species = sp,
    n_raw = n_raw,
    n_clean = n_clean,
    eligible = eligible,
    message = "ok",
    stringsAsFactors = FALSE
  )

  if (eligible) {
    rec$cleaned$data$species <- sp
    selected_list[[sp]] <- rec$cleaned$data
  }
}

attempt_df <- bind_rows(attempt_log)
write_csv(attempt_df, "outputs/multispecies/selection_attempts_multispecies.csv")

selected_species <- names(selected_list)
if (length(selected_species) < 3) {
  stop("Fewer than 3 eligible species found with >= 50 cleaned records. Increase candidate pool or attempts.")
}

selected_summary <- attempt_df %>%
  filter(species %in% selected_species, eligible) %>%
  group_by(species) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  arrange(desc(n_clean))

write_csv(selected_summary, "outputs/multispecies/selected_species_multispecies.csv")
selected_summary
```

## 2. Climate Extraction for Selected Species

```{r climate-extract}
occ_all <- bind_rows(selected_list)

bio_vars <- c(1, 4, 12, 15)
bio_names <- resolve_bio_var_names(bio_vars)

bio_rasters <- get_worldclim_bio(
  worldclim_res = cfg$worldclim_res,
  cache_dir = "outputs/worldclim_cache",
  bio_vars = bio_vars
)

occ_climate <- extract_climate_values(occ_all, bio_rasters)
occ_climate <- occ_climate %>%
  mutate(across(all_of(bio_names), as.numeric)) %>%
  filter(if_all(all_of(bio_names), is.finite))

write_csv(occ_climate, "outputs/multispecies/occurrences_with_climate_multispecies.csv")

occ_climate %>%
  count(species, name = "n_points") %>%
  arrange(desc(n_points))
```

## 3. Climate Plots for Several Random Species

```{r climate-plots, fig.width=11, fig.height=8}
p1 <- ggplot(occ_climate, aes(x = BIO1, y = BIO12, color = species)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_density_2d(alpha = 0.5) +
  facet_wrap(~ species, scales = "free") +
  theme_minimal(base_size = 11) +
  labs(
    title = "Realized Climate Space by Species",
    subtitle = "BIO1 (Annual Mean Temperature x10) vs BIO12 (Annual Precipitation)",
    x = "BIO1",
    y = "BIO12"
  ) +
  guides(color = "none")

p1

ggsave(
  filename = "outputs/multispecies/plot_climate_space_facet.png",
  plot = p1,
  width = 12,
  height = 8,
  dpi = 300
)
```

## 4. PCA of Climate Conditions Across Occurrences

```{r pca, fig.width=9, fig.height=7}
clim_mat <- as.data.frame(scale(occ_climate[, bio_names, drop = FALSE]))
pca_fit <- prcomp(clim_mat, center = FALSE, scale. = FALSE)

pca_scores <- data.frame(
  PC1 = pca_fit$x[, 1],
  PC2 = pca_fit$x[, 2],
  species = occ_climate$species
)

pca_var <- summary(pca_fit)$importance[2, 1:2]

p2 <- ggplot(pca_scores, aes(PC1, PC2, color = species)) +
  geom_point(alpha = 0.5, size = 1.0) +
  stat_ellipse(level = 0.80, linewidth = 0.6) +
  theme_minimal(base_size = 11) +
  labs(
    title = "PCA of Climate Variables at Occurrence Points",
    subtitle = sprintf("Explained variance: PC1 = %.1f%%, PC2 = %.1f%%", 100 * pca_var[1], 100 * pca_var[2])
  )

p2

ggsave(
  filename = "outputs/multispecies/plot_pca_climate.png",
  plot = p2,
  width = 10,
  height = 7,
  dpi = 300
)
```

## 5. Gradient-Length Check and Constrained Ordination (ter Braak Logic)

```{r dca-rda-cca, fig.width=9, fig.height=7}
# Community matrix: rows = occurrence sites, columns = species membership.
# Each row has one observed species in this workflow.
comm <- model.matrix(~ species - 1, data = occ_climate)
colnames(comm) <- sub("^species", "", colnames(comm))

# DCA axis length diagnostic
# Rule of thumb used:
#   >4 SD: unimodal methods (CCA)
#   2-4 SD: compare both
#   <2 SD: linear methods (RDA)
dca_fit <- decorana(comm)

axis1_length <- NA_real_
if (!is.null(dca_fit$rproj) && ncol(dca_fit$rproj) >= 1) {
  # In current vegan versions, summary.decorana is defunct.
  # We estimate gradient length as the observed range of DCA1 site scores.
  axis1_length <- diff(range(dca_fit$rproj[, 1], na.rm = TRUE))
}

env <- occ_climate[, bio_names, drop = FALSE]
for (v in names(env)) env[[v]] <- as.numeric(env[[v]])

comm_hell <- decostand(comm, method = "hellinger")

rda_fit <- rda(comm_hell ~ BIO1 + BIO4 + BIO12 + BIO15, data = env)
cca_fit <- cca(comm ~ BIO1 + BIO4 + BIO12 + BIO15, data = env)

anova_rda <- anova.cca(rda_fit, permutations = 999)
anova_cca <- anova.cca(cca_fit, permutations = 999)

r2_rda <- RsquareAdj(rda_fit)$adj.r.squared
r2_cca <- RsquareAdj(cca_fit)$adj.r.squared

method_decision <- if (is.finite(axis1_length)) {
  if (axis1_length > 4) {
    "CCA preferred (unimodal response expected)"
  } else if (axis1_length < 2) {
    "RDA preferred (linear response expected)"
  } else {
    "Intermediate (2-4): compare RDA and CCA"
  }
} else {
  "DCA axis length not available; compare RDA and CCA by permutation and adjusted R2"
}

method_table <- data.frame(
  metric = c("DCA axis-1 length (SD units)", "Decision", "RDA adjusted R2", "RDA permutation p", "CCA adjusted R2", "CCA permutation p"),
  value = c(
    ifelse(is.finite(axis1_length), sprintf("%.3f", axis1_length), "NA"),
    method_decision,
    sprintf("%.4f", r2_rda),
    sprintf("%.4f", anova_rda$`Pr(>F)`[1]),
    sprintf("%.4f", r2_cca),
    sprintf("%.4f", anova_cca$`Pr(>F)`[1])
  ),
  stringsAsFactors = FALSE
)

write_csv(method_table, "outputs/multispecies/method_decision_table.csv")
method_table
```

```{r ordination-plot, fig.width=9, fig.height=7}
# Plot the method selected by the DCA rule; if intermediate, use the model
# with higher adjusted R2 as the display model.
use_rda <- TRUE
if (is.finite(axis1_length)) {
  if (axis1_length > 4) {
    use_rda <- FALSE
  } else if (axis1_length >= 2 && axis1_length <= 4) {
    use_rda <- is.finite(r2_rda) && is.finite(r2_cca) && (r2_rda >= r2_cca)
  }
}

if (use_rda) {
  ord_fit <- rda_fit
  ord_name <- "RDA"
} else {
  ord_fit <- cca_fit
  ord_name <- "CCA"
}

site_scores <- as.data.frame(scores(ord_fit, display = "sites", scaling = 2))
site_scores$species <- occ_climate$species

axis_names <- names(site_scores)[1:2]

p_ord <- ggplot(site_scores, aes_string(x = axis_names[1], y = axis_names[2], color = "species")) +
  geom_point(alpha = 0.45, size = 1.0) +
  stat_ellipse(level = 0.80, linewidth = 0.6) +
  theme_minimal(base_size = 11) +
  labs(
    title = paste0(ord_name, " site scores (scaling = 2)"),
    subtitle = "Constrained ordination of species membership by climate predictors"
  )

p_ord

ggsave(
  filename = paste0("outputs/multispecies/plot_ordination_", tolower(ord_name), ".png"),
  plot = p_ord,
  width = 10,
  height = 7,
  dpi = 300
)
```

## 6. Notes on Interpretation and Limits

- This is a correlative climate-space analysis and constrained ordination exercise, not a causal model of species performance.
- Occurrence records may retain spatial bias and georeferencing uncertainty.
- In this workflow, one species is observed per record row; constrained ordination therefore reflects climate separation among species occurrence clouds rather than full co-occurrence community structure.

## Session Outputs

The analysis writes:

- `outputs/multispecies/selection_attempts_multispecies.csv`
- `outputs/multispecies/selected_species_multispecies.csv`
- `outputs/multispecies/occurrences_with_climate_multispecies.csv`
- `outputs/multispecies/method_decision_table.csv`
- `outputs/multispecies/plot_climate_space_facet.png`
- `outputs/multispecies/plot_pca_climate.png`
- `outputs/multispecies/plot_ordination_rda.png` or `outputs/multispecies/plot_ordination_cca.png`
