---
title: "BAAD: Above-Ground Biomass vs. Leaf Mass and Leaf Area Allometry"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(baad.data)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)
library(smatr)      # standardised major axis regression
library(ggrepel)    # label repulsion for family plots
library(ape)        # phylogenetics base
library(caper)      # PGLS
library(V.PhyloMaker2)  # backbone plant phylogeny

theme_set(theme_bw(base_size = 12))

# ── Load BAAD ─────────────────────────────────────────────────────────────────
baad <- baad.data::baad_data()
dat  <- baad$data

# ── Derive clade group from PFT code ─────────────────────────────────────────
# PFT codes: DA = Deciduous Angiosperm, DG = Deciduous Gymnosperm,
#             EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.
# Angiosperm/Gymnosperm is encoded in the trailing letter of the pft code.
dat <- dat %>%
  mutate(
    clade = case_when(
      grepl("A$", pft) ~ "Angiosperm",
      grepl("G$", pft) ~ "Gymnosperm",
      TRUE             ~ NA_character_
    )
  )

# ── Subset: records with m.to AND (m.lf or a.lf) ─────────────────────────────
d_lf <- dat %>% filter(m.to > 0, m.lf > 0)   # AGB + leaf mass
d_al <- dat %>% filter(m.to > 0, a.lf > 0)   # AGB + leaf area

# Helper: run sma() safely and return a tidy 1-row data.frame
sma_tidy <- function(df, response, predictor, group_label = "all") {
  tryCatch({
    fit <- sma(log10(df[[response]]) ~ log10(df[[predictor]]))
    s   <- summary(fit)$coef[[1]]
    data.frame(
      group    = group_label,
      n        = nrow(df),
      intercept = s[1, 1],
      slope     = s[2, 1],
      slope_lo  = s[2, 3],
      slope_hi  = s[2, 4],
      r2        = cor(log10(df[[response]]), log10(df[[predictor]]))^2,
      method   = "SMA"
    )
  }, error = function(e) {
    data.frame(group = group_label, n = nrow(df),
               intercept = NA, slope = NA, slope_lo = NA, slope_hi = NA, r2 = NA,
               method = "SMA")
  })
}

# Helper: run OLS and return a tidy 1-row data.frame (same structure as sma_tidy)
ols_tidy <- function(df, response, predictor, group_label = "all") {
  tryCatch({
    fit <- lm(log10(df[[response]]) ~ log10(df[[predictor]]))
    ci  <- confint(fit)
    data.frame(
      group     = group_label,
      n         = nrow(df),
      intercept = coef(fit)[1],
      slope     = coef(fit)[2],
      slope_lo  = ci[2, 1],
      slope_hi  = ci[2, 2],
      r2        = summary(fit)$r.squared,
      method    = "OLS"
    )
  }, error = function(e) {
    data.frame(group = group_label, n = nrow(df),
               intercept = NA, slope = NA, slope_lo = NA, slope_hi = NA, r2 = NA,
               method = "OLS")
  })
}
```

---

# Overview

This document focuses on the allometry between **total above-ground biomass**
(`m.to`, kg) and two leaf-related traits measured on the same individual:

| Response | Symbol | Units |
|---|---|---|
| Total leaf mass | `m.lf` | kg |
| Total leaf area | `a.lf` | m² |

Scaling relationships are estimated using **Standardised Major Axis (SMA)**
regression on log₁₀-transformed data, which is the statistically appropriate
method when both axes carry measurement error (Warton et al. 2006, *Biol. Rev.*).
The null prediction under West–Brown–Enquist metabolic scaling theory is a
3/4-power relationship between leaf mass and plant mass
(Enquist & Niklas 2002, *Nature*). The pipe model predicts leaf area scales
approximately linearly with sapwood area and thus sub-linearly with AGB.

**BAAD citation:** Falster DS et al. (2015). BAAD: a Biomass And Allometry
Database for woody plants. *Ecology* 96(5):1445.
DOI: [10.1890/14-1889.1](https://doi.org/10.1890/14-1889.1)

```{r data-summary}
kable(data.frame(
  Dataset      = c("m.lf ~ m.to", "a.lf ~ m.to"),
  N_records    = c(nrow(d_lf), nrow(d_al)),
  N_species    = c(n_distinct(d_lf$speciesMatched), n_distinct(d_al$speciesMatched)),
  N_families   = c(n_distinct(d_lf$family[!is.na(d_lf$family)]),
                   n_distinct(d_al$family[!is.na(d_al$family)])),
  N_angiosperms = c(sum(!is.na(d_lf$clade) & d_lf$clade == "Angiosperm"),
                    sum(!is.na(d_al$clade) & d_al$clade == "Angiosperm")),
  N_gymnosperms = c(sum(!is.na(d_lf$clade) & d_lf$clade == "Gymnosperm"),
                    sum(!is.na(d_al$clade) & d_al$clade == "Gymnosperm"))
), caption = "Record counts for the two focal relationships")
```

```{r pgls-prep, echo=TRUE, warning=FALSE, message=FALSE}
# ── PGLS: species-mean data frames ───────────────────────────────────────────
sp_lf <- d_lf %>%
  filter(!is.na(speciesMatched), !is.na(family)) %>%
  group_by(speciesMatched) %>%
  summarise(
    log_mto = mean(log10(m.to), na.rm = TRUE),
    log_mlf = mean(log10(m.lf), na.rm = TRUE),
    family  = first(family),
    .groups = "drop"
  ) %>%
  drop_na() %>%
  mutate(sp_phylo = gsub(" ", "_", speciesMatched))

sp_al <- d_al %>%
  filter(!is.na(speciesMatched), !is.na(family)) %>%
  group_by(speciesMatched) %>%
  summarise(
    log_mto = mean(log10(m.to), na.rm = TRUE),
    log_al  = mean(log10(a.lf), na.rm = TRUE),
    family  = first(family),
    .groups = "drop"
  ) %>%
  drop_na() %>%
  mutate(sp_phylo = gsub(" ", "_", speciesMatched))

# ── Build backbone phylogeny via V.PhyloMaker2 ────────────────────────────────
species_list <- bind_rows(
  sp_lf %>% dplyr::select(species = speciesMatched, family),
  sp_al %>% dplyr::select(species = speciesMatched, family)
) %>%
  distinct(species, .keep_all = TRUE) %>%
  mutate(genus = sub(" .*", "", species)) %>%
  dplyr::select(species, genus, family)

baad_tree <- tryCatch({
  phy_out <- V.PhyloMaker2::phylo.maker(
    sp.list   = species_list,
    tree      = GBOTB.extended.TPL,
    nodes     = nodes.info.1.TPL,
    scenarios = "S3"
  )
  phy_out$scenario.3
}, error = function(e) {
  message("Phylogeny construction failed: ", conditionMessage(e))
  NULL
})

if (!is.null(baad_tree)) {
  cat("Phylogeny built:", ape::Ntip(baad_tree), "tips\n")
} else {
  cat("baad_tree is NULL — PGLS will be skipped.\n")
}
```

---

# Interspecific scaling — all woody plants

## Leaf mass ~ AGB

```{r overall-leaf-mass, fig.height=6}
fit_lf <- sma(log10(m.lf) ~ log10(m.to), data = d_lf)
sl_lf  <- round(coef(fit_lf)[2], 3)
int_lf <- round(coef(fit_lf)[1], 3)
r2_lf  <- round(cor(log10(d_lf$m.lf), log10(d_lf$m.to))^2, 3)

# OLS fit
lm_lf      <- lm(log10(m.lf) ~ log10(m.to), data = d_lf)
ols_sl_lf  <- round(coef(lm_lf)[2], 3)
ols_int_lf <- round(coef(lm_lf)[1], 3)
ols_ci_lf  <- round(confint(lm_lf)[2, ], 3)
ols_r2_lf  <- round(summary(lm_lf)$r.squared, 3)

ggplot(d_lf, aes(x = m.to, y = m.lf, colour = pft)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_abline(intercept = int_lf, slope = sl_lf,
              colour = "black", linewidth = 1.2) +
  geom_abline(intercept = ols_int_lf, slope = ols_sl_lf,
              colour = "firebrick", linewidth = 1.0, linetype = "solid") +
  geom_abline(slope = 0.75, intercept = int_lf, linetype = "dashed",
              colour = "grey40", linewidth = 0.9) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_colour_brewer(palette = "Dark2", na.value = "grey70") +
  annotation_logticks() +
  labs(
    x = "Above-ground biomass, m.to (kg)",
    y = "Leaf mass, m.lf (kg)",
    colour = "PFT",
    title = paste0("Leaf mass ~ AGB  |  SMA β = ", sl_lf,
                   "  OLS β = ", ols_sl_lf,
                   "  |  n = ", format(nrow(d_lf), big.mark = ",")),
    subtitle = "Black: SMA fit.  Red: OLS fit.  Dashed: 3/4-power reference (WBE)"
  )
```

> **SMA result:** m.lf = 10^`r int_lf` × m.to^`r sl_lf`  
> (n = `r format(nrow(d_lf), big.mark=",")`; R² = `r r2_lf`)

```{r method-compare-lf, echo=TRUE, warning=FALSE, message=FALSE}
kable(data.frame(
  Method = c("OLS", "SMA"),
  Slope  = c(ols_sl_lf, sl_lf),
  CI_lo  = c(ols_ci_lf[1], round(fit_lf$coef[[1]][2, 2], 3)),
  CI_hi  = c(ols_ci_lf[2], round(fit_lf$coef[[1]][2, 3], 3)),
  R2     = c(ols_r2_lf, r2_lf)
),
col.names = c("Method", "Slope", "95% CI lo", "95% CI hi", "R²"),
caption   = "OLS vs. SMA comparison — leaf mass ~ AGB")
```

## Leaf area ~ AGB

```{r overall-leaf-area, fig.height=6}
fit_al <- sma(log10(a.lf) ~ log10(m.to), data = d_al)
sl_al  <- round(coef(fit_al)[2], 3)
int_al <- round(coef(fit_al)[1], 3)
r2_al  <- round(cor(log10(d_al$a.lf), log10(d_al$m.to))^2, 3)

# OLS fit
lm_al      <- lm(log10(a.lf) ~ log10(m.to), data = d_al)
ols_sl_al  <- round(coef(lm_al)[2], 3)
ols_int_al <- round(coef(lm_al)[1], 3)
ols_ci_al  <- round(confint(lm_al)[2, ], 3)
ols_r2_al  <- round(summary(lm_al)$r.squared, 3)

ggplot(d_al, aes(x = m.to, y = a.lf, colour = pft)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_abline(intercept = int_al, slope = sl_al,
              colour = "black", linewidth = 1.2) +
  geom_abline(intercept = ols_int_al, slope = ols_sl_al,
              colour = "firebrick", linewidth = 1.0, linetype = "solid") +
  geom_abline(slope = 0.75, intercept = int_al, linetype = "dashed",
              colour = "grey40", linewidth = 0.9) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_colour_brewer(palette = "Dark2", na.value = "grey70") +
  annotation_logticks() +
  labs(
    x = "Above-ground biomass, m.to (kg)",
    y = "Leaf area, a.lf (m²)",
    colour = "PFT",
    title = paste0("Leaf area ~ AGB  |  SMA β = ", sl_al,
                   "  OLS β = ", ols_sl_al,
                   "  |  n = ", format(nrow(d_al), big.mark = ",")),
    subtitle = "Black: SMA fit.  Red: OLS fit.  Dashed: 3/4-power reference"
  )
```

> **SMA result:** a.lf = 10^`r int_al` × m.to^`r sl_al`  
> (n = `r format(nrow(d_al), big.mark=",")`; R² = `r r2_al`)

```{r method-compare-al, echo=TRUE, warning=FALSE, message=FALSE}
kable(data.frame(
  Method = c("OLS", "SMA"),
  Slope  = c(ols_sl_al, sl_al),
  CI_lo  = c(ols_ci_al[1], round(fit_al$coef[[1]][2, 2], 3)),
  CI_hi  = c(ols_ci_al[2], round(fit_al$coef[[1]][2, 3], 3)),
  R2     = c(ols_r2_al, r2_al)
),
col.names = c("Method", "Slope", "95% CI lo", "95% CI hi", "R²"),
caption   = "OLS vs. SMA comparison — leaf area ~ AGB")
```

---

# Angiosperm vs. Gymnosperm

SMA regression fitted separately for each clade. Only records with a known `pft`
code (DA, DG, EA, or EG) are included; records with missing `pft` are excluded
from this section.

## Leaf mass ~ AGB by clade

```{r clade-leaf-mass, fig.height=6}
d_lf_c <- d_lf %>% filter(!is.na(clade))

# Fit SMA per clade
clade_lf <- d_lf_c %>%
  group_by(clade) %>%
  group_modify(~ {
    tryCatch({
      lx  <- log10(.x$m.to)
      ly  <- log10(.x$m.lf)
      fit <- sma(ly ~ lx)
      cm  <- fit$coef[[1]]
      data.frame(
        n         = nrow(.x),
        intercept = cm[1, 1],
        slope     = cm[2, 1],
        slope_lo  = cm[2, 2],
        slope_hi  = cm[2, 3],
        r2        = cor(lx, ly)^2
      )
    }, error = function(e) {
      data.frame(n = nrow(.x), intercept = NA, slope = NA,
                 slope_lo = NA, slope_hi = NA, r2 = NA)
    })
  }) %>%
  ungroup()

# OLS per clade — leaf mass
ols_clade_lf <- d_lf_c %>%
  group_by(clade) %>%
  group_modify(~ {
    tryCatch({
      fit <- lm(log10(.x$m.lf) ~ log10(.x$m.to))
      ci  <- confint(fit)
      data.frame(ols_slope = round(coef(fit)[2], 3),
                 ols_slope_lo = round(ci[2, 1], 3),
                 ols_slope_hi = round(ci[2, 2], 3))
    }, error = function(e) {
      data.frame(ols_slope = NA_real_, ols_slope_lo = NA_real_, ols_slope_hi = NA_real_)
    })
  }) %>%
  ungroup()
clade_lf <- clade_lf %>% left_join(ols_clade_lf, by = "clade")

# Build prediction lines across AGB range
pred_range <- seq(log10(min(d_lf_c$m.to)), log10(max(d_lf_c$m.to)), length.out = 200)

clade_lines_lf <- clade_lf %>%
  rowwise() %>%
  mutate(log_x = list(pred_range),
         log_y = list(intercept + slope * pred_range)) %>%
  unnest(c(log_x, log_y)) %>%
  mutate(x = 10^log_x, y = 10^log_y)

ggplot(d_lf_c, aes(x = m.to, y = m.lf, colour = clade)) +
  geom_point(alpha = 0.2, size = 0.7) +
  geom_line(data = clade_lines_lf, aes(x = x, y = y, colour = clade),
            linewidth = 1.3) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8")) +
  annotation_logticks() +
  labs(
    x = "Above-ground biomass, m.to (kg)",
    y = "Leaf mass, m.lf (kg)",
    colour = "Clade",
    title = "Leaf mass ~ AGB — Angiosperm vs. Gymnosperm (SMA)",
    subtitle = "Separate SMA lines per clade"
  )
```

```{r clade-leaf-mass-table}
kable(clade_lf %>%
        mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
        dplyr::select(clade, n, slope, slope_lo, slope_hi, ols_slope, ols_slope_lo, ols_slope_hi, r2),
      col.names = c("Clade", "N", "SMA slope", "SMA CI lo", "SMA CI hi",
                    "OLS slope", "OLS CI lo", "OLS CI hi", "R²"),
      caption = "SMA and OLS slopes for leaf mass ~ AGB by clade")
```

## Leaf area ~ AGB by clade

```{r clade-leaf-area, fig.height=6}
d_al_c <- d_al %>% filter(!is.na(clade))

clade_al <- d_al_c %>%
  group_by(clade) %>%
  group_modify(~ {
    tryCatch({
      lx  <- log10(.x$m.to)
      ly  <- log10(.x$a.lf)
      fit <- sma(ly ~ lx)
      cm  <- fit$coef[[1]]
      data.frame(
        n         = nrow(.x),
        intercept = cm[1, 1],
        slope     = cm[2, 1],
        slope_lo  = cm[2, 2],
        slope_hi  = cm[2, 3],
        r2        = cor(lx, ly)^2
      )
    }, error = function(e) {
      data.frame(n = nrow(.x), intercept = NA, slope = NA,
                 slope_lo = NA, slope_hi = NA, r2 = NA)
    })
  }) %>%
  ungroup()

# OLS per clade — leaf area
ols_clade_al <- d_al_c %>%
  group_by(clade) %>%
  group_modify(~ {
    tryCatch({
      fit <- lm(log10(.x$a.lf) ~ log10(.x$m.to))
      ci  <- confint(fit)
      data.frame(ols_slope = round(coef(fit)[2], 3),
                 ols_slope_lo = round(ci[2, 1], 3),
                 ols_slope_hi = round(ci[2, 2], 3))
    }, error = function(e) {
      data.frame(ols_slope = NA_real_, ols_slope_lo = NA_real_, ols_slope_hi = NA_real_)
    })
  }) %>%
  ungroup()
clade_al <- clade_al %>% left_join(ols_clade_al, by = "clade")

pred_range_al <- seq(log10(min(d_al_c$m.to)), log10(max(d_al_c$m.to)), length.out = 200)

clade_lines_al <- clade_al %>%
  rowwise() %>%
  mutate(log_x = list(pred_range_al),
         log_y = list(intercept + slope * pred_range_al)) %>%
  unnest(c(log_x, log_y)) %>%
  mutate(x = 10^log_x, y = 10^log_y)

ggplot(d_al_c, aes(x = m.to, y = a.lf, colour = clade)) +
  geom_point(alpha = 0.2, size = 0.7) +
  geom_line(data = clade_lines_al, aes(x = x, y = y, colour = clade),
            linewidth = 1.3) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8")) +
  annotation_logticks() +
  labs(
    x = "Above-ground biomass, m.to (kg)",
    y = "Leaf area, a.lf (m²)",
    colour = "Clade",
    title = "Leaf area ~ AGB — Angiosperm vs. Gymnosperm (SMA)",
    subtitle = "Separate SMA lines per clade"
  )
```

```{r clade-leaf-area-table}
kable(clade_al %>%
        mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
        dplyr::select(clade, n, slope, slope_lo, slope_hi, ols_slope, ols_slope_lo, ols_slope_hi, r2),
      col.names = c("Clade", "N", "SMA slope", "SMA CI lo", "SMA CI hi",
                    "OLS slope", "OLS CI lo", "OLS CI hi", "R²"),
      caption = "SMA and OLS slopes for leaf area ~ AGB by clade")
```

---

# Phylogenetic regression (PGLS)

PGLS models residual covariance using Brownian motion, accounting for shared
evolutionary history across species. Analysis is conducted on **species means**
(one row per species, computed in the `pgls-prep` chunk above). Pagel’s
λ is estimated by maximum likelihood. Backbone phylogeny from V.PhyloMaker2
(Jin & Qian 2022, *J. Biogeography* 49:1731–1744).

```{r pgls-analysis, echo=TRUE, warning=FALSE, message=FALSE}
# Initialise result holders in the global chunk environment
pgls_lf_fit    <- NULL
pgls_al_fit    <- NULL
pgls_lf_result <- NULL
pgls_al_result <- NULL

if (!is.null(baad_tree)) {
  tree_tips <- baad_tree$tip.label

  # --- Leaf mass ~ AGB PGLS ---
  sp_lf_match  <- sp_lf %>% filter(sp_phylo %in% tree_tips)
  n_matched_lf <- nrow(sp_lf_match)
  cat("Leaf mass: species matched to tree:", n_matched_lf, "of", nrow(sp_lf), "\n")
  if (n_matched_lf >= 10) {
    tryCatch({
      tree_lf      <- ape::drop.tip(baad_tree, setdiff(tree_tips, sp_lf_match$sp_phylo))
      df_lf_pgls   <- as.data.frame(sp_lf_match)
      rownames(df_lf_pgls) <- df_lf_pgls$sp_phylo
      cd_lf        <- caper::comparative.data(phy = tree_lf, data = df_lf_pgls,
                                              names.col = "sp_phylo", warn.dropped = FALSE)
      pgls_lf_fit  <- caper::pgls(log_mlf ~ log_mto, data = cd_lf, lambda = "ML")
      pgls_lf_result <- list(
        n       = n_matched_lf,
        slope   = round(coef(pgls_lf_fit)[2], 3),
        se      = round(summary(pgls_lf_fit)$coefficients[2, 2], 3),
        lambda  = round(pgls_lf_fit$param[["lambda"]], 3),
        p_value = round(summary(pgls_lf_fit)$coefficients[2, 4], 4)
      )
    }, error = function(e) message("Leaf mass PGLS failed: ", conditionMessage(e)))
  }

  # --- Leaf area ~ AGB PGLS ---
  sp_al_match  <- sp_al %>% filter(sp_phylo %in% tree_tips)
  n_matched_al <- nrow(sp_al_match)
  cat("Leaf area: species matched to tree:", n_matched_al, "of", nrow(sp_al), "\n")
  if (n_matched_al >= 10) {
    tryCatch({
      tree_al      <- ape::drop.tip(baad_tree, setdiff(tree_tips, sp_al_match$sp_phylo))
      df_al_pgls   <- as.data.frame(sp_al_match)
      rownames(df_al_pgls) <- df_al_pgls$sp_phylo
      cd_al        <- caper::comparative.data(phy = tree_al, data = df_al_pgls,
                                              names.col = "sp_phylo", warn.dropped = FALSE)
      pgls_al_fit  <- caper::pgls(log_al ~ log_mto, data = cd_al, lambda = "ML")
      pgls_al_result <- list(
        n       = n_matched_al,
        slope   = round(coef(pgls_al_fit)[2], 3),
        se      = round(summary(pgls_al_fit)$coefficients[2, 2], 3),
        lambda  = round(pgls_al_fit$param[["lambda"]], 3),
        p_value = round(summary(pgls_al_fit)$coefficients[2, 4], 4)
      )
    }, error = function(e) message("Leaf area PGLS failed: ", conditionMessage(e)))
  }
} else {
  cat("Phylogeny construction failed; PGLS skipped.\n")
}
```

## Leaf mass ~ AGB (PGLS)

```{r pgls-lf-results, echo=TRUE, warning=FALSE, message=FALSE}
if (!is.null(pgls_lf_result)) {
  kable(data.frame(
    Method  = "PGLS (Brownian motion, λ ML)",
    N_sp    = pgls_lf_result$n,
    Slope   = pgls_lf_result$slope,
    SE      = pgls_lf_result$se,
    Lambda  = pgls_lf_result$lambda,
    P_value = pgls_lf_result$p_value
  ),
  col.names = c("Method", "N species", "Slope", "SE", "Pagel's λ (ML)", "p-value"),
  caption = "PGLS result — leaf mass ~ AGB (species means)")
} else {
  cat("PGLS not available for leaf mass ~ AGB.\n")
}
```

```{r pgls-lf-plot, echo=TRUE, warning=FALSE, message=FALSE, fig.height=6}
if (nrow(sp_lf) > 0) {
  ols_sp_lf     <- lm(log_mlf ~ log_mto, data = sp_lf)
  sma_sp_lf     <- sma(log_mlf ~ log_mto, data = sp_lf)
  sma_sp_lf_int <- sma_sp_lf$coef[[1]][1, 1]
  sma_sp_lf_sl  <- sma_sp_lf$coef[[1]][2, 1]

  subtitle_lf <- paste0(
    "Black: SMA (β=", round(sma_sp_lf_sl, 3), ")",
    "  Red: OLS (β=", round(coef(ols_sp_lf)[2], 3), ")",
    if (!is.null(pgls_lf_result)) paste0("  Green: PGLS (β=", pgls_lf_result$slope, ")") else ""
  )

  p_pgls_lf <- ggplot(sp_lf, aes(x = log_mto, y = log_mlf)) +
    geom_point(aes(colour = family), alpha = 0.5, size = 1.5, show.legend = FALSE) +
    geom_abline(intercept = coef(ols_sp_lf)[1], slope = coef(ols_sp_lf)[2],
                colour = "firebrick", linewidth = 1.0) +
    geom_abline(intercept = sma_sp_lf_int, slope = sma_sp_lf_sl,
                colour = "black", linewidth = 1.0) +
    labs(x = "log₁₀ AGB (kg)", y = "log₁₀ leaf mass (kg)",
         title = "Species means — leaf mass ~ AGB", subtitle = subtitle_lf)

  if (!is.null(pgls_lf_fit)) {
    p_pgls_lf <- p_pgls_lf +
      geom_abline(intercept = coef(pgls_lf_fit)[1], slope = coef(pgls_lf_fit)[2],
                  colour = "darkgreen", linewidth = 1.0, linetype = "dashed")
  }
  print(p_pgls_lf)
}
```

## Leaf area ~ AGB (PGLS)

```{r pgls-al-results, echo=TRUE, warning=FALSE, message=FALSE}
if (!is.null(pgls_al_result)) {
  kable(data.frame(
    Method  = "PGLS (Brownian motion, λ ML)",
    N_sp    = pgls_al_result$n,
    Slope   = pgls_al_result$slope,
    SE      = pgls_al_result$se,
    Lambda  = pgls_al_result$lambda,
    P_value = pgls_al_result$p_value
  ),
  col.names = c("Method", "N species", "Slope", "SE", "Pagel's λ (ML)", "p-value"),
  caption = "PGLS result — leaf area ~ AGB (species means)")
} else {
  cat("PGLS not available for leaf area ~ AGB.\n")
}
```

```{r pgls-al-plot, echo=TRUE, warning=FALSE, message=FALSE, fig.height=6}
if (nrow(sp_al) > 0) {
  ols_sp_al     <- lm(log_al ~ log_mto, data = sp_al)
  sma_sp_al     <- sma(log_al ~ log_mto, data = sp_al)
  sma_sp_al_int <- sma_sp_al$coef[[1]][1, 1]
  sma_sp_al_sl  <- sma_sp_al$coef[[1]][2, 1]

  subtitle_al <- paste0(
    "Black: SMA (β=", round(sma_sp_al_sl, 3), ")",
    "  Red: OLS (β=", round(coef(ols_sp_al)[2], 3), ")",
    if (!is.null(pgls_al_result)) paste0("  Green: PGLS (β=", pgls_al_result$slope, ")") else ""
  )

  p_pgls_al <- ggplot(sp_al, aes(x = log_mto, y = log_al)) +
    geom_point(aes(colour = family), alpha = 0.5, size = 1.5, show.legend = FALSE) +
    geom_abline(intercept = coef(ols_sp_al)[1], slope = coef(ols_sp_al)[2],
                colour = "firebrick", linewidth = 1.0) +
    geom_abline(intercept = sma_sp_al_int, slope = sma_sp_al_sl,
                colour = "black", linewidth = 1.0) +
    labs(x = "log₁₀ AGB (kg)", y = "log₁₀ leaf area (m²)",
         title = "Species means — leaf area ~ AGB", subtitle = subtitle_al)

  if (!is.null(pgls_al_fit)) {
    p_pgls_al <- p_pgls_al +
      geom_abline(intercept = coef(pgls_al_fit)[1], slope = coef(pgls_al_fit)[2],
                  colour = "darkgreen", linewidth = 1.0, linetype = "dashed")
  }
  print(p_pgls_al)
}
```

---

# Family-level scaling

SMA regression fitted for each plant family with ≥ 30 individual records
having both `m.to` and the relevant leaf trait measured. Families are sorted
by slope. The **3/4-power reference line** is shown in grey.

## Leaf mass ~ AGB by family

```{r family-leaf-mass, fig.height=8, fig.width=11}
# ── Minimum records per family to estimate a regression ──────────────────────
MIN_N_FAM <- 30

fam_lf_counts <- d_lf %>%
  filter(!is.na(family)) %>%
  count(family) %>%
  filter(n >= MIN_N_FAM) %>%
  pull(family)

fam_lf_slopes <- d_lf %>%
  filter(!is.na(family), family %in% fam_lf_counts) %>%
  group_by(family) %>%
  group_modify(~ {
    tryCatch({
      lx  <- log10(.x$m.to)
      ly  <- log10(.x$m.lf)
      fit <- sma(ly ~ lx)
      cm  <- fit$coef[[1]]
      data.frame(
        n         = nrow(.x),
        n_sp      = n_distinct(.x$speciesMatched),
        intercept = cm[1, 1],
        slope     = cm[2, 1],
        slope_lo  = cm[2, 2],
        slope_hi  = cm[2, 3],
        r2        = cor(lx, ly)^2
      )
    }, error = function(e) {
      data.frame(n = nrow(.x), n_sp = n_distinct(.x$speciesMatched),
                 intercept = NA, slope = NA, slope_lo = NA, slope_hi = NA, r2 = NA)
    })
  }) %>%
  ungroup() %>%
  filter(!is.na(slope)) %>%
  arrange(slope)

# OLS per family — leaf mass
fam_lf_ols <- d_lf %>%
  filter(!is.na(family), family %in% fam_lf_counts) %>%
  group_by(family) %>%
  group_modify(~ {
    tryCatch({
      fit <- lm(log10(.x$m.lf) ~ log10(.x$m.to))
      data.frame(ols_slope = round(coef(fit)[2], 3))
    }, error = function(e) data.frame(ols_slope = NA_real_))
  }) %>%
  ungroup()
fam_lf_slopes <- fam_lf_slopes %>% left_join(fam_lf_ols, by = "family")

# Forest-style dot-range plot─
ggplot(fam_lf_slopes, aes(x = slope, y = reorder(family, slope))) +
  geom_vline(xintercept = 0.75, colour = "grey50", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = 1.00, colour = "grey70", linetype = "dotted", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = slope_lo, xmax = slope_hi),
                 height = 0.3, colour = "steelblue", linewidth = 0.7) +
  geom_point(aes(size = log10(n)), colour = "steelblue", fill = "white", shape = 21,
             stroke = 1.1) +
  scale_size_continuous(range = c(2, 6), name = "log₁₀(N records)") +
  labs(
    x = "SMA slope (log leaf mass ~ log AGB)",
    y = NULL,
    title = paste0("Leaf mass ~ AGB: family-level SMA slopes (families with ≥ ", MIN_N_FAM, " records)"),
    subtitle = "Dashed = 3/4-power (WBE); dotted = isometry (slope = 1). Points sized by log₁₀(N records)."
  ) +
  theme(panel.grid.major.y = element_line(colour = "grey92"),
        panel.grid.minor   = element_blank())
```

```{r family-leaf-mass-table}
kable(
  fam_lf_slopes %>%
    arrange(desc(n)) %>%
    mutate(across(c(slope, slope_lo, slope_hi, r2, intercept, ols_slope), ~ round(.x, 3))) %>%
    dplyr::select(family, n, n_sp, slope, slope_lo, slope_hi, ols_slope, r2),
  col.names = c("Family", "N records", "N species", "SMA slope",
                "SMA CI lo", "SMA CI hi", "OLS slope", "R²"),
  caption = paste0("Family-level SMA and OLS slopes — leaf mass ~ AGB (≥ ", MIN_N_FAM, " records)")
)
```

## Leaf area ~ AGB by family

```{r family-leaf-area, fig.height=8, fig.width=11}
fam_al_counts <- d_al %>%
  filter(!is.na(family)) %>%
  count(family) %>%
  filter(n >= MIN_N_FAM) %>%
  pull(family)

fam_al_slopes <- d_al %>%
  filter(!is.na(family), family %in% fam_al_counts) %>%
  group_by(family) %>%
  group_modify(~ {
    tryCatch({
      lx  <- log10(.x$m.to)
      ly  <- log10(.x$a.lf)
      fit <- sma(ly ~ lx)
      cm  <- fit$coef[[1]]
      data.frame(
        n         = nrow(.x),
        n_sp      = n_distinct(.x$speciesMatched),
        intercept = cm[1, 1],
        slope     = cm[2, 1],
        slope_lo  = cm[2, 2],
        slope_hi  = cm[2, 3],
        r2        = cor(lx, ly)^2
      )
    }, error = function(e) {
      data.frame(n = nrow(.x), n_sp = n_distinct(.x$speciesMatched),
                 intercept = NA, slope = NA, slope_lo = NA, slope_hi = NA, r2 = NA)
    })
  }) %>%
  ungroup() %>%
  filter(!is.na(slope)) %>%
  arrange(slope)

# OLS per family — leaf area
fam_al_ols <- d_al %>%
  filter(!is.na(family), family %in% fam_al_counts) %>%
  group_by(family) %>%
  group_modify(~ {
    tryCatch({
      fit <- lm(log10(.x$a.lf) ~ log10(.x$m.to))
      data.frame(ols_slope = round(coef(fit)[2], 3))
    }, error = function(e) data.frame(ols_slope = NA_real_))
  }) %>%
  ungroup()
fam_al_slopes <- fam_al_slopes %>% left_join(fam_al_ols, by = "family")

ggplot(fam_al_slopes, aes(x = slope, y = reorder(family, slope))) +
  geom_vline(xintercept = 0.75, colour = "grey50", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = 1.00, colour = "grey70", linetype = "dotted", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = slope_lo, xmax = slope_hi),
                 height = 0.3, colour = "#e5691f", linewidth = 0.7) +
  geom_point(aes(size = log10(n)), colour = "#e5691f", fill = "white", shape = 21,
             stroke = 1.1) +
  scale_size_continuous(range = c(2, 6), name = "log₁₀(N records)") +
  labs(
    x = "SMA slope (log leaf area ~ log AGB)",
    y = NULL,
    title = paste0("Leaf area ~ AGB: family-level SMA slopes (families with ≥ ", MIN_N_FAM, " records)"),
    subtitle = "Dashed = 3/4-power (WBE); dotted = isometry (slope = 1). Points sized by log₁₀(N records)."
  ) +
  theme(panel.grid.major.y = element_line(colour = "grey92"),
        panel.grid.minor   = element_blank())
```

```{r family-leaf-area-table}
kable(
  fam_al_slopes %>%
    arrange(desc(n)) %>%
    mutate(across(c(slope, slope_lo, slope_hi, r2, intercept, ols_slope), ~ round(.x, 3))) %>%
    dplyr::select(family, n, n_sp, slope, slope_lo, slope_hi, ols_slope, r2),
  col.names = c("Family", "N records", "N species", "SMA slope",
                "SMA CI lo", "SMA CI hi", "OLS slope", "R²"),
  caption = paste0("Family-level SMA and OLS slopes — leaf area ~ AGB (≥ ", MIN_N_FAM, " records)")
)
```

---

# Intraspecific scaling

Within-species (intraspecific) scaling slopes are estimated for species with
**≥ 8 individual records** having both AGB and the leaf trait measured.
Intraspecific slopes are compared to the interspecific SMA slope (shown as a
vertical dashed line).

**Interpretation caveat:** BAAD individuals within a species often come from a
single study or a narrow size range. Intraspecific slopes should be treated as
within-study estimates and may reflect ontogenetic or site-level effects rather
than true species-level allometries.

## Leaf mass ~ AGB: intraspecific slope distribution

```{r intra-leaf-mass-hist, fig.height=5}
MIN_N_INTRA <- 8

intra_lf <- d_lf %>%
  filter(!is.na(speciesMatched)) %>%
  group_by(speciesMatched) %>%
  group_modify(~ {
    if (nrow(.x) < MIN_N_INTRA) return(data.frame())
    tryCatch({
      lx   <- log10(.x$m.to)
      ly   <- log10(.x$m.lf)
      fit  <- sma(ly ~ lx)
      cm   <- fit$coef[[1]]
      data.frame(
        n         = nrow(.x),
        family    = .x$family[1],
        clade     = .x$clade[1],
        slope     = cm[2, 1],
        slope_lo  = cm[2, 2],
        slope_hi  = cm[2, 3],
        r2        = cor(lx, ly)^2,
        agb_range = diff(range(lx))
      )
    }, error = function(e) data.frame())
  }) %>%
  ungroup()

cat("Species with ≥", MIN_N_INTRA, "records (m.lf + m.to):", nrow(intra_lf), "\n")

ggplot(intra_lf, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.05, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = sl_lf,  linetype = "solid",  colour = "black", linewidth = 1) +
  geom_vline(xintercept = 0.75,   linetype = "dashed", colour = "grey40", linewidth = 0.9) +
  geom_vline(xintercept = 1.00,   linetype = "dotted", colour = "grey60", linewidth = 0.7) +
  scale_fill_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                    na.value = "grey60", name = "Clade") +
  labs(
    x = "Intraspecific SMA slope",
    y = "Number of species",
    title = paste0("Distribution of intraspecific slopes — leaf mass ~ AGB  (≥ ", MIN_N_INTRA, " records/species)"),
    subtitle = paste0("Solid = interspecific SMA slope (", sl_lf, "); dashed = 3/4; dotted = 1.0")
  )
```

```{r intra-leaf-mass-summary}
kable(
  intra_lf %>%
    group_by(clade) %>%
    summarise(
      n_species   = n(),
      median_slope = round(median(slope, na.rm = TRUE), 3),
      mean_slope   = round(mean(slope, na.rm = TRUE), 3),
      sd_slope     = round(sd(slope, na.rm = TRUE), 3),
      pct_lt_075   = round(mean(slope < 0.75, na.rm = TRUE) * 100, 1),
      pct_gt_1     = round(mean(slope > 1.00, na.rm = TRUE) * 100, 1)
    ),
  col.names = c("Clade", "N species", "Median slope", "Mean slope", "SD slope",
                "% slopes < 0.75", "% slopes > 1.0"),
  caption = "Summary of intraspecific leaf mass ~ AGB slopes by clade"
)
```

## Top species by record count — leaf mass ~ AGB

Scatter plots for the 16 most data-rich species, each with its own SMA line.

```{r intra-leaf-mass-topsp, fig.height=10, fig.width=12}
top_sp_lf <- intra_lf %>%
  arrange(desc(n)) %>%
  slice_head(n = 16) %>%
  pull(speciesMatched)

d_lf %>%
  filter(speciesMatched %in% top_sp_lf) %>%
  ggplot(aes(x = m.to, y = m.lf)) +
  geom_point(aes(colour = clade), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", formula = y ~ x,   # lm on log scale = SMA direction visual
              colour = "black", se = TRUE, linewidth = 0.9) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                      na.value = "grey60") +
  facet_wrap(~ speciesMatched, scales = "free", ncol = 4) +
  annotation_logticks() +
  labs(
    x = "AGB m.to (kg)", y = "Leaf mass m.lf (kg)", colour = "Clade",
    title = "Intraspecific leaf mass ~ AGB: top 16 species by record count"
  ) +
  theme(strip.text = element_text(size = 7, face = "italic"),
        legend.position = "bottom")
```

## Leaf area ~ AGB: intraspecific slope distribution

```{r intra-leaf-area-hist, fig.height=5}
intra_al <- d_al %>%
  filter(!is.na(speciesMatched)) %>%
  group_by(speciesMatched) %>%
  group_modify(~ {
    if (nrow(.x) < MIN_N_INTRA) return(data.frame())
    tryCatch({
      lx   <- log10(.x$m.to)
      ly   <- log10(.x$a.lf)
      fit  <- sma(ly ~ lx)
      cm   <- fit$coef[[1]]
      data.frame(
        n         = nrow(.x),
        family    = .x$family[1],
        clade     = .x$clade[1],
        slope     = cm[2, 1],
        slope_lo  = cm[2, 2],
        slope_hi  = cm[2, 3],
        r2        = cor(lx, ly)^2,
        agb_range = diff(range(lx))
      )
    }, error = function(e) data.frame())
  }) %>%
  ungroup()

cat("Species with ≥", MIN_N_INTRA, "records (a.lf + m.to):", nrow(intra_al), "\n")

ggplot(intra_al, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.05, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = sl_al,  linetype = "solid",  colour = "black", linewidth = 1) +
  geom_vline(xintercept = 0.75,   linetype = "dashed", colour = "grey40", linewidth = 0.9) +
  geom_vline(xintercept = 1.00,   linetype = "dotted", colour = "grey60", linewidth = 0.7) +
  scale_fill_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                    na.value = "grey60", name = "Clade") +
  labs(
    x = "Intraspecific SMA slope",
    y = "Number of species",
    title = paste0("Distribution of intraspecific slopes — leaf area ~ AGB  (≥ ", MIN_N_INTRA, " records/species)"),
    subtitle = paste0("Solid = interspecific SMA slope (", sl_al, "); dashed = 3/4; dotted = 1.0")
  )
```

```{r intra-leaf-area-summary}
kable(
  intra_al %>%
    group_by(clade) %>%
    summarise(
      n_species   = n(),
      median_slope = round(median(slope, na.rm = TRUE), 3),
      mean_slope   = round(mean(slope, na.rm = TRUE), 3),
      sd_slope     = round(sd(slope, na.rm = TRUE), 3),
      pct_lt_075   = round(mean(slope < 0.75, na.rm = TRUE) * 100, 1),
      pct_gt_1     = round(mean(slope > 1.00, na.rm = TRUE) * 100, 1)
    ),
  col.names = c("Clade", "N species", "Median slope", "Mean slope", "SD slope",
                "% slopes < 0.75", "% slopes > 1.0"),
  caption = "Summary of intraspecific leaf area ~ AGB slopes by clade"
)
```

## Top species by record count — leaf area ~ AGB

```{r intra-leaf-area-topsp, fig.height=10, fig.width=12}
top_sp_al <- intra_al %>%
  arrange(desc(n)) %>%
  slice_head(n = 16) %>%
  pull(speciesMatched)

d_al %>%
  filter(speciesMatched %in% top_sp_al) %>%
  ggplot(aes(x = m.to, y = a.lf)) +
  geom_point(aes(colour = clade), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", formula = y ~ x,
              colour = "black", se = TRUE, linewidth = 0.9) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                      na.value = "grey60") +
  facet_wrap(~ speciesMatched, scales = "free", ncol = 4) +
  annotation_logticks() +
  labs(
    x = "AGB m.to (kg)", y = "Leaf area a.lf (m²)", colour = "Clade",
    title = "Intraspecific leaf area ~ AGB: top 16 species by record count"
  ) +
  theme(strip.text = element_text(size = 7, face = "italic"),
        legend.position = "bottom")
```

---

# Intraspecific vs. interspecific slopes

A key question in allometry is whether slopes measured within species (ontogenetic
or population-level variation) match those estimated across species (evolutionary
divergence). The scatter below plots intraspecific SMA slope against the log₁₀
of sample size within each species.

## Leaf mass ~ AGB

```{r slope-vs-n-lf, fig.height=5}
ggplot(intra_lf, aes(x = n, y = slope, colour = clade)) +
  geom_hline(yintercept = sl_lf,  linetype = "solid",  colour = "grey30", linewidth = 0.8) +
  geom_hline(yintercept = 0.75,   linetype = "dashed", colour = "grey50", linewidth = 0.7) +
  geom_point(alpha = 0.6, size = 2) +
  geom_errorbar(aes(ymin = slope_lo, ymax = slope_hi), width = 0, alpha = 0.25) +
  scale_x_log10() +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                      na.value = "grey60") +
  labs(
    x = "N records per species (log scale)",
    y = "Intraspecific SMA slope",
    colour = "Clade",
    title = "Intraspecific slope vs. sample size — leaf mass ~ AGB",
    subtitle = paste0("Solid = interspecific slope (", sl_lf, "); dashed = 3/4")
  )
```

## Leaf area ~ AGB

```{r slope-vs-n-al, fig.height=5}
ggplot(intra_al, aes(x = n, y = slope, colour = clade)) +
  geom_hline(yintercept = sl_al,  linetype = "solid",  colour = "grey30", linewidth = 0.8) +
  geom_hline(yintercept = 0.75,   linetype = "dashed", colour = "grey50", linewidth = 0.7) +
  geom_point(alpha = 0.6, size = 2) +
  geom_errorbar(aes(ymin = slope_lo, ymax = slope_hi), width = 0, alpha = 0.25) +
  scale_x_log10() +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                      na.value = "grey60") +
  labs(
    x = "N records per species (log scale)",
    y = "Intraspecific SMA slope",
    colour = "Clade",
    title = "Intraspecific slope vs. sample size — leaf area ~ AGB",
    subtitle = paste0("Solid = interspecific slope (", sl_al, "); dashed = 3/4")
  )
```

---

# Summary table

```{r final-summary-table}
kable(data.frame(
  Relationship        = c("Leaf mass ~ AGB (all)",
                           "Leaf mass ~ AGB (Angiosperm)",
                           "Leaf mass ~ AGB (Gymnosperm)",
                           "Leaf area ~ AGB (all)",
                           "Leaf area ~ AGB (Angiosperm)",
                           "Leaf area ~ AGB (Gymnosperm)"),
  N                   = c(nrow(d_lf),
                           clade_lf$n[clade_lf$clade == "Angiosperm"],
                           clade_lf$n[clade_lf$clade == "Gymnosperm"],
                           nrow(d_al),
                           clade_al$n[clade_al$clade == "Angiosperm"],
                           clade_al$n[clade_al$clade == "Gymnosperm"]),
  SMA_slope           = round(c(sl_lf,
                                 clade_lf$slope[clade_lf$clade == "Angiosperm"],
                                 clade_lf$slope[clade_lf$clade == "Gymnosperm"],
                                 sl_al,
                                 clade_al$slope[clade_al$clade == "Angiosperm"],
                                 clade_al$slope[clade_al$clade == "Gymnosperm"]), 3),
  CI_lo               = round(c(fit_lf$coef[[1]][2, 2],
                                 clade_lf$slope_lo[clade_lf$clade == "Angiosperm"],
                                 clade_lf$slope_lo[clade_lf$clade == "Gymnosperm"],
                                 fit_al$coef[[1]][2, 2],
                                 clade_al$slope_lo[clade_al$clade == "Angiosperm"],
                                 clade_al$slope_lo[clade_al$clade == "Gymnosperm"]), 3),
  CI_hi               = round(c(fit_lf$coef[[1]][2, 3],
                                 clade_lf$slope_hi[clade_lf$clade == "Angiosperm"],
                                 clade_lf$slope_hi[clade_lf$clade == "Gymnosperm"],
                                 fit_al$coef[[1]][2, 3],
                                 clade_al$slope_hi[clade_al$clade == "Angiosperm"],
                                 clade_al$slope_hi[clade_al$clade == "Gymnosperm"]), 3),
  OLS_slope           = round(c(ols_sl_lf,
                                 clade_lf$ols_slope[clade_lf$clade == "Angiosperm"],
                                 clade_lf$ols_slope[clade_lf$clade == "Gymnosperm"],
                                 ols_sl_al,
                                 clade_al$ols_slope[clade_al$clade == "Angiosperm"],
                                 clade_al$ols_slope[clade_al$clade == "Gymnosperm"]), 3),
  PGLS_slope          = c(if (!is.null(pgls_lf_result)) pgls_lf_result$slope else NA,
                           NA, NA,
                           if (!is.null(pgls_al_result)) pgls_al_result$slope else NA,
                           NA, NA),
  R2                  = round(c(r2_lf,
                                 clade_lf$r2[clade_lf$clade == "Angiosperm"],
                                 clade_lf$r2[clade_lf$clade == "Gymnosperm"],
                                 r2_al,
                                 clade_al$r2[clade_al$clade == "Angiosperm"],
                                 clade_al$r2[clade_al$clade == "Gymnosperm"]), 3)
),
col.names = c("Relationship", "N records", "SMA slope", "SMA CI lo", "SMA CI hi",
              "OLS slope", "PGLS slope", "R²"),
caption = "Summary of interspecific slopes by method (all and by clade); PGLS on species means only")
```

---

# Methods note

- **Data source:** BAAD v1 (Falster et al. 2015), accessed via the `baad.data` R
  package. All measurements are individual-level.
- **OLS:** `lm()` on log₁₀-transformed data. Assumes all error is in the response
  variable; appropriate for prediction but biased for characterising bivariate
  allometric relationships when both axes carry measurement error.
- **SMA (Standardised Major Axis):** `smatr::sma()` (Warton et al. 2012,
  *Methods Ecol. Evol.* 3:257–259; Warton et al. 2006, *Biol. Rev.* 81:259–282).
  Appropriate when both axes carry error and the goal is to characterise the
  bivariate relationship rather than predict one from the other; standard for
  allometry.
- **PGLS (Phylogenetic Generalised Least Squares):** `caper::pgls()` fitted to
  species means; Pagel’s λ estimated by maximum likelihood; accounts for
  shared evolutionary history (Freckleton et al. 2002, *Am. Nat.* 160:712–726).
  Backbone phylogeny from V.PhyloMaker2 (Jin & Qian 2022, *J. Biogeography*
  49:1731–1744; Qian & Jin 2016, *J. Systematics & Evolution* 54:380–388).
- **Angio/Gymno classification:** derived from the BAAD `pft` field (trailing
  letter A = Angiosperm, G = Gymnosperm). Records with missing `pft` are excluded
  from the clade-level comparisons.
- **Family-level analysis:** families with ≥ 30 individual records.
- **Intraspecific analysis:** species with ≥ 8 individual records having both
  AGB and the focal leaf trait. Slopes reflect ontogenetic or population-level
  size variation within each dataset contributing records for that species.
- **Theoretical benchmarks:** WBE/MST theory (Enquist & Niklas 2002) predicts
  leaf mass ∝ AGB^(3/4); isometry (slope = 1.0) would indicate constant leaf
  mass fraction regardless of plant size.
