---
title: "BAAD: Biomass And Allometry Database — Detailed Summary"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
library(baad.data)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)

theme_set(theme_bw(base_size = 12))

# Load BAAD once
baad   <- baad.data::baad_data()
dat    <- baad$data
refs   <- baad$references
dict   <- baad$dictionary
meta   <- baad$metadata
```

---

# Overview

The **Biomass And Allometry Database (BAAD)** is a curated, open-access compilation of
individual-level plant structural measurements drawn from the primary literature.
It was assembled and published by Falster et al. (2015) and is freely accessible
via the `baad.data` R package (GitHub: traitecoevo/baad.data).

> **Reference:** Falster DS, Duursma RA, Ishihara MI, et al. (2015). BAAD: a Biomass
> And Allometry Database for woody plants. *Ecology*, 96(5):1445.
> DOI: [10.1890/14-1889.1](https://doi.org/10.1890/14-1889.1)

```{r top-level-counts}
n_rows    <- nrow(dat)
n_sp      <- length(unique(dat$speciesMatched))
n_studies <- length(unique(dat$studyName))
n_geo     <- sum(!is.na(dat$latitude))

kable(data.frame(
  Metric  = c("Total individual records", "Species (matched)", "Source studies",
               "Records with coordinates"),
  Value   = format(c(n_rows, n_sp, n_studies, n_geo), big.mark = ",")
), caption = "Top-level database summary")
```

---

# Data structure

## Column dictionary

All trait columns have consistent units. The table below reproduces the BAAD
data dictionary — the canonical reference for variable meanings and units.

```{r dictionary}
kable(dict, caption = "BAAD data dictionary (all variables)")
```

## Column-level data availability

Not all traits are measured for every individual. The chart below shows the
number of non-missing values per column.

```{r availability, fig.height=7}
avail <- dat %>%
  summarise(across(where(is.numeric), ~ sum(!is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n") %>%
  left_join(dict %>% select(variable, units, group), by = "variable") %>%
  arrange(desc(n))

ggplot(avail, aes(x = reorder(variable, n), y = n, fill = group)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Non-missing records", fill = "Group",
       title = "Data availability per trait column in BAAD",
       subtitle = paste0("Total rows = ", format(n_rows, big.mark = ","))) +
  scale_fill_brewer(palette = "Set2")
```

---

# Taxonomic coverage

## Family breakdown

```{r family-bar, fig.height=6}
fam <- dat %>%
  filter(!is.na(family)) %>%
  count(family, sort = TRUE)

ggplot(fam[1:25, ], aes(x = reorder(family, n), y = n)) +
  geom_col(fill = "#2c7bb6") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Number of records",
       title = "Top 25 plant families in BAAD")
```

```{r family-table}
kable(fam, col.names = c("Family", "N records"),
      caption = paste0("All ", nrow(fam), " families in BAAD"))
```

## Top species by record count

```{r species-table}
sp_counts <- dat %>%
  filter(!is.na(speciesMatched)) %>%
  count(speciesMatched, family, sort = TRUE)

kable(head(sp_counts, 40),
      col.names = c("Species (matched)", "Family", "N records"),
      caption = "Top 40 species by individual record count")
```

## Plant functional types (PFT)

```{r pft-bar}
pft_labels <- c(DA = "Deciduous Angiosperm", DG = "Deciduous Gymnosperm",
                EA = "Evergreen Angiosperm", EG = "Evergreen Gymnosperm")

dat %>%
  filter(!is.na(pft)) %>%
  count(pft) %>%
  mutate(label = pft_labels[pft]) %>%
  ggplot(aes(x = reorder(label, n), y = n, fill = pft)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = NULL, y = "Records", title = "Records by plant functional type")
```

---

# Geographic coverage

## Map of sampling locations

```{r world-map, fig.width=10, fig.height=5}
world <- map_data("world")

site_pts <- dat %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  distinct(studyName, latitude, longitude, vegetation)

ggplot() +
  geom_polygon(data = world, aes(x = long, y = lat, group = group),
               fill = "grey85", color = "grey60", linewidth = 0.2) +
  geom_point(data = site_pts, aes(x = longitude, y = latitude, color = vegetation),
             alpha = 0.7, size = 1.8) +
  coord_fixed(1.3) +
  scale_color_brewer(palette = "Set1", na.value = "grey50") +
  labs(title = "BAAD sampling locations by vegetation type",
       x = "Longitude", y = "Latitude", color = "Vegetation") +
  theme(legend.position = "bottom")
```

## Vegetation type breakdown

```{r veg-bar}
veg_labels <- c(TempF  = "Temperate Forest", TropRF = "Tropical Rainforest",
                BorF   = "Boreal Forest",    TropSF = "Tropical Seasonal Forest",
                Wo     = "Woodland",         TempRF = "Temperate Rainforest",
                Sav    = "Savanna",          Sh     = "Shrubland",  Gr = "Grassland")

dat %>%
  filter(!is.na(vegetation)) %>%
  count(vegetation, sort = TRUE) %>%
  mutate(label = veg_labels[vegetation]) %>%
  ggplot(aes(x = reorder(label, n), y = n, fill = label)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Paired") +
  labs(x = NULL, y = "Records", title = "Records by vegetation biome")
```

## Growing conditions

```{r growing-condition}
gc_labels <- c(CG = "Common Garden", FE = "Field – Even-aged",
               FW = "Field – Wide-aged", GH = "Greenhouse",
               PM = "Plantation – Monoculture", PU = "Plantation – Uneven-aged")

dat %>%
  filter(!is.na(growingCondition)) %>%
  count(growingCondition, sort = TRUE) %>%
  mutate(label = ifelse(growingCondition %in% names(gc_labels),
                        gc_labels[growingCondition], growingCondition)) %>%
  ggplot(aes(x = reorder(label, n), y = n, fill = label)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set3") +
  labs(x = NULL, y = "Records", title = "Records by growing condition")
```

---

# Scaling relationships

All mass variables are in **kg**; all length/diameter variables are in **m**;
all area variables are in **m²**. Scaling is assessed on log-transformed data
following standard allometric practice. A regression slope $\beta$ in log-log
space gives the power-law exponent: $Y = \alpha \cdot X^\beta$.

## Total biomass vs. height

```{r mto-ht, fig.width=8, fig.height=5}
d_mto_ht <- dat %>% filter(m.to > 0, h.t > 0)

lm1 <- lm(log10(m.to) ~ log10(h.t), data = d_mto_ht)
b1  <- round(coef(lm1)[2], 3)
a1  <- round(10^coef(lm1)[1], 4)
n1  <- nrow(d_mto_ht)

ggplot(d_mto_ht, aes(x = h.t, y = m.to, color = pft)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE,
              aes(group = 1), linewidth = 1) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_color_brewer(palette = "Dark2", na.value = "grey60") +
  labs(x = "Height h.t (m)", y = "Total biomass m.to (kg)", color = "PFT",
       title = paste0("Total biomass ~ height  |  β = ", b1,
                      "  |  n = ", format(n1, big.mark = ","))) +
  annotation_logticks()
```

> OLS in log-log space: m.to = `r a1` × h.t^`r b1`
> (n = `r format(n1, big.mark=",")`). Note: OLS on log-transformed data
> is exploratory; use SMA regression for allometric inference.

## Total biomass vs. stem diameter at breast height (DBH)

```{r mto-dbh, fig.width=8, fig.height=5}
d_mto_dbh <- dat %>% filter(m.to > 0, d.bh > 0)

lm2 <- lm(log10(m.to) ~ log10(d.bh), data = d_mto_dbh)
b2  <- round(coef(lm2)[2], 3)
a2  <- round(10^coef(lm2)[1], 4)
n2  <- nrow(d_mto_dbh)

ggplot(d_mto_dbh, aes(x = d.bh * 100, y = m.to, color = pft)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE,
              aes(group = 1), linewidth = 1) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_color_brewer(palette = "Dark2", na.value = "grey60") +
  labs(x = "DBH (cm)", y = "Total biomass m.to (kg)", color = "PFT",
       title = paste0("Total biomass ~ DBH  |  β = ", b2,
                      "  |  n = ", format(n2, big.mark = ","))) +
  annotation_logticks()
```

> OLS in log-log space: m.to = `r a2` × d.bh^`r b2`
> (n = `r format(n2, big.mark=",")`). Diameter in metres for regression.

## Leaf mass vs. total biomass

```{r mlf-mto}
d_lf_to <- dat %>% filter(m.lf > 0, m.to > 0)

lm3 <- lm(log10(m.lf) ~ log10(m.to), data = d_lf_to)
b3  <- round(coef(lm3)[2], 3)
a3  <- round(10^coef(lm3)[1], 5)
n3  <- nrow(d_lf_to)

ggplot(d_lf_to, aes(x = m.to, y = m.lf, color = pft)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE,
              aes(group = 1), linewidth = 1) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_color_brewer(palette = "Dark2", na.value = "grey60") +
  labs(x = "Total biomass m.to (kg)", y = "Leaf mass m.lf (kg)", color = "PFT",
       title = paste0("Leaf mass ~ total biomass  |  β = ", b3,
                      "  |  n = ", format(n3, big.mark = ","))) +
  annotation_logticks()
```

## Stem mass vs. total biomass

```{r mst-mto}
d_st_to <- dat %>% filter(m.st > 0, m.to > 0)

lm4 <- lm(log10(m.st) ~ log10(m.to), data = d_st_to)
b4  <- round(coef(lm4)[2], 3)
n4  <- nrow(d_st_to)

ggplot(d_st_to, aes(x = m.to, y = m.st, color = pft)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE,
              aes(group = 1), linewidth = 1) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_color_brewer(palette = "Dark2", na.value = "grey60") +
  labs(x = "Total biomass m.to (kg)", y = "Stem mass m.st (kg)", color = "PFT",
       title = paste0("Stem mass ~ total biomass  |  β = ", b4,
                      "  |  n = ", format(n4, big.mark = ","))) +
  annotation_logticks()
```

## Root mass vs. total biomass

```{r mrt-mto}
d_rt_to <- dat %>% filter(m.rt > 0, m.to > 0)

lm5 <- lm(log10(m.rt) ~ log10(m.to), data = d_rt_to)
b5  <- round(coef(lm5)[2], 3)
n5  <- nrow(d_rt_to)

ggplot(d_rt_to, aes(x = m.to, y = m.rt, color = pft)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE,
              aes(group = 1), linewidth = 1) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_color_brewer(palette = "Dark2", na.value = "grey60") +
  labs(x = "Total biomass m.to (kg)", y = "Root mass m.rt (kg)", color = "PFT",
       title = paste0("Root mass ~ total biomass  |  β = ", b5,
                      "  |  n = ", format(n5, big.mark = ","))) +
  annotation_logticks()
```

## Leaf area vs. total biomass

```{r alf-mto}
d_alf_to <- dat %>% filter(a.lf > 0, m.to > 0)

lm6 <- lm(log10(a.lf) ~ log10(m.to), data = d_alf_to)
b6  <- round(coef(lm6)[2], 3)
n6  <- nrow(d_alf_to)

ggplot(d_alf_to, aes(x = m.to, y = a.lf, color = pft)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE,
              aes(group = 1), linewidth = 1) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_color_brewer(palette = "Dark2", na.value = "grey60") +
  labs(x = "Total biomass m.to (kg)", y = "Leaf area a.lf (m²)", color = "PFT",
       title = paste0("Leaf area ~ total biomass  |  β = ", b6,
                      "  |  n = ", format(n6, big.mark = ","))) +
  annotation_logticks()
```

## Summary of scaling exponents (interspecific, OLS)

All regressions are OLS on log10-transformed data and are **exploratory only**.
Standardized major axis (SMA) regression is recommended for allometric inference
and will be applied in the formal analysis phase.

```{r exponent-summary}
exponents <- data.frame(
  Relationship      = c("m.to ~ h.t", "m.to ~ d.bh",
                        "m.lf ~ m.to", "m.st ~ m.to",
                        "m.rt ~ m.to", "a.lf ~ m.to"),
  Y_variable        = c("Total biomass (kg)", "Total biomass (kg)",
                        "Leaf mass (kg)", "Stem mass (kg)",
                        "Root mass (kg)", "Leaf area (m²)"),
  X_variable        = c("Height (m)", "DBH (m)",
                        "Total biomass (kg)", "Total biomass (kg)",
                        "Total biomass (kg)", "Total biomass (kg)"),
  OLS_exponent_beta = c(b1, b2, b3, b4, b5, b6),
  n                 = c(n1, n2, n3, n4, n5, n6)
)

kable(exponents, caption = "Interspecific OLS scaling exponents (log-log, all PFTs pooled). Exploratory only — see Methods note.")
```

---

# Source studies

## Studies sorted by record count

```{r studies-table}
study_counts <- dat %>%
  count(studyName, sort = TRUE) %>%
  left_join(refs %>% select(studyName, citation, doi), by = "studyName")

kable(study_counts, col.names = c("Study", "Records", "Citation", "DOI"),
      caption = paste0("All ", nrow(study_counts), " source studies in BAAD"))
```

---

# Key caveats and notes for downstream analysis

1. **OLS vs SMA**: Log-log OLS exponents reported here underestimate true
   allometric slopes when both variables have measurement error. Standardised
   Major Axis (SMA) regression using the `smatr` package is required for
   unbiased allometric exponent estimation.

2. **Intra- vs interspecific**: All plots above pool all species. Intraspecific
   regressions are expected to differ from interspecific ones and must be
   assessed separately.

3. **Growing conditions**: Field-grown, greenhouse, and plantation individuals
   span very different ontogenetic stages; analyses should stratify or account
   for `growingCondition`.

4. **Leaf mass dominance**: `m.lf` has the highest data availability (n = 17,069)
   but `m.to` coverage is limited (n = 8,878). Biomass allocation fractions
   should only be computed where all required organ masses are present.

5. **Unit consistency**: All BAAD variables use SI units (kg, m, m²) as specified
   in the dictionary. No unit conversion is required within BAAD itself.

6. **Taxonomy**: `speciesMatched` is the harmonized species name column; use this
   rather than the raw `species` field for cross-study comparisons.

---

# Session info

```{r session}
sessionInfo()
```
