---
title: "BAAD: Intraspecific Leaf Scaling — OLS and SMA Slopes per Species"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(baad.data)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)
library(smatr)

theme_set(theme_bw(base_size = 12))

# ─── Load BAAD ────────────────────────────────────────────────────────────────
baad <- baad.data::baad_data()
dat  <- baad$data

dat <- dat %>%
  mutate(
    clade = case_when(
      grepl("A$", pft) ~ "Angiosperm",
      grepl("G$", pft) ~ "Gymnosperm",
      TRUE             ~ NA_character_
    )
  )

# ─── Analysis constants ───────────────────────────────────────────────────────
MIN_N <- 10   # minimum records per species for regression

# ─── Six focal allometric combinations ───────────────────────────────────────
# Predictor: basal diameter (d.ba, m), total plant biomass (m.to, kg),
#            stem mass (m.st, kg — AGB excluding leaves)
# Response:  leaf mass (m.lf, kg), leaf area (a.lf, m²)
#
# Note: BAAD m.to = m.rt + m.so (includes root mass); it is the best available
# whole-plant size measure in BAAD but is NOT purely above-ground. m.st = all
# stem tissue (branches, bark) and approximates AGB without leaves.

# wbe_ref: WBE-predicted exponent for each combination.
#   d.ba combos : 2.0  (area-preserving branching → leaf area/mass ~ d^2)
#   m.to combos : 0.75 (WBE metabolic scaling → leaf ~ total mass^(3/4))
#   m.st combos : NA   (no explicit WBE prediction for leaf ~ stem mass)
combos <- list(
  list(x = "d.ba",  y = "m.lf", wbe_ref = 2.00,
       xlab = "Basal diameter d.ba (m)",        ylab = "Leaf mass m.lf (kg)",
       xlab_log = "log₁₀ basal diameter (m)",   ylab_log = "log₁₀ leaf mass (kg)",
       title_short = "Leaf mass ~ basal diameter"),
  list(x = "d.ba",  y = "a.lf", wbe_ref = 2.00,
       xlab = "Basal diameter d.ba (m)",        ylab = "Leaf area a.lf (m²)",
       xlab_log = "log₁₀ basal diameter (m)",   ylab_log = "log₁₀ leaf area (m²)",
       title_short = "Leaf area ~ basal diameter"),
  list(x = "m.to",  y = "m.lf", wbe_ref = 0.75,
       xlab = "Total plant biomass m.to (kg)",  ylab = "Leaf mass m.lf (kg)",
       xlab_log = "log₁₀ total biomass (kg)",   ylab_log = "log₁₀ leaf mass (kg)",
       title_short = "Leaf mass ~ total plant biomass"),
  list(x = "m.to",  y = "a.lf", wbe_ref = 0.75,
       xlab = "Total plant biomass m.to (kg)",  ylab = "Leaf area a.lf (m²)",
       xlab_log = "log₁₀ total biomass (kg)",   ylab_log = "log₁₀ leaf area (m²)",
       title_short = "Leaf area ~ total plant biomass"),
  list(x = "m.st",  y = "m.lf", wbe_ref = NA_real_,
       xlab = "Stem mass m.st (kg)",            ylab = "Leaf mass m.lf (kg)",
       xlab_log = "log₁₀ stem mass (kg)",       ylab_log = "log₁₀ leaf mass (kg)",
       title_short = "Leaf mass ~ stem mass"),
  list(x = "m.st",  y = "a.lf", wbe_ref = NA_real_,
       xlab = "Stem mass m.st (kg)",            ylab = "Leaf area a.lf (m²)",
       xlab_log = "log₁₀ stem mass (kg)",       ylab_log = "log₁₀ leaf area (m²)",
       title_short = "Leaf area ~ stem mass")
)

# ─── Per-species regression helper ───────────────────────────────────────────
# Returns per-species OLS + SMA slopes for one (x, y) combo.
# Minimum MIN_N records per species required.
fit_intra <- function(data, xcol, ycol, min_n = MIN_N) {
  sub <- data %>%
    filter(!is.na(speciesMatched),
           !is.na(.data[[xcol]]), .data[[xcol]] > 0,
           !is.na(.data[[ycol]]), .data[[ycol]] > 0)

  sub %>%
    group_by(speciesMatched) %>%
    group_modify(~ {
      if (nrow(.x) < min_n) return(data.frame())
      lx <- log10(.x[[xcol]])
      ly <- log10(.x[[ycol]])
      rng <- diff(range(lx))
      if (rng == 0) return(data.frame())  # zero size range → no slope

      # OLS
      ols_fit <- lm(ly ~ lx)
      ols_sl  <- coef(ols_fit)[2]
      ols_ci  <- tryCatch(confint(ols_fit)[2, ], error = function(e) c(NA, NA))

      # SMA
      sma_res <- tryCatch({
        s  <- sma(ly ~ lx)
        cm <- s$coef[[1]]
        c(slope = cm[2, 1], lo = cm[2, 2], hi = cm[2, 3])
      }, error = function(e) c(slope = NA, lo = NA, hi = NA))

      data.frame(
        n          = nrow(.x),
        family     = .x$family[1],
        clade      = .x$clade[1],
        study      = .x$studyName[1],
        ols_slope  = ols_sl,
        ols_lo     = ols_ci[1],
        ols_hi     = ols_ci[2],
        sma_slope  = sma_res["slope"],
        sma_lo     = sma_res["lo"],
        sma_hi     = sma_res["hi"],
        r2         = cor(lx, ly)^2,
        log_x_min  = min(lx),
        log_x_max  = max(lx),
        log_x_range = rng
      )
    }) %>%
    ungroup() %>%
    filter(!is.na(ols_slope))
}

# ─── Compute all six datasets ─────────────────────────────────────────────────
cat("Computing intraspecific regressions for 6 combinations...\n")
results <- lapply(combos, function(co) {
  fit_intra(dat, co$x, co$y)
})
names(results) <- sapply(combos, function(co) paste0(co$y, "~", co$x))
cat("Done.\n")
```

---

# Overview

This document reports **intraspecific** (within-species) scaling relationships between
leaf traits and three plant size proxies across all woody plant species in the BAAD
database with ≥ `r MIN_N` individual records per species.

**Six allometric combinations analysed:**

| Response | Predictor | Symbol | Units |
|---|---|---|---|
| Leaf mass | Basal stem diameter | `m.lf ~ d.ba` | kg ~ m |
| Leaf area | Basal stem diameter | `a.lf ~ d.ba` | m² ~ m |
| Leaf mass | Total plant biomass | `m.lf ~ m.to` | kg ~ kg |
| Leaf area | Total plant biomass | `a.lf ~ m.to` | m² ~ kg |
| Leaf mass | Stem mass | `m.lf ~ m.st` | kg ~ kg |
| Leaf area | Stem mass | `a.lf ~ m.st` | m² ~ kg |

> **WBE predictions tested:**
> - Leaf area and leaf mass vs. **basal diameter**: exponent = **2** (area-preserving branching: conduit cross-sectional area is conserved across branching levels, so total leaf area ∝ d²; Enquist et al. 1998; West et al. 1999).
> - Leaf area and leaf mass vs. **total plant biomass**: exponent = **3/4** (metabolic scaling: whole-plant resource flux ∝ M^(3/4), so leaf area, which sets gas exchange capacity, also scales as M^(3/4); Enquist & Niklas 2001).
> - Leaf area and leaf mass vs. **stem mass**: no explicit WBE exponent stated here; isometry (1.0) shown as a reference.
>
> **Variable notes:**  
> `d.ba` = stem diameter at base [m] (more commonly measured in BAAD than DBH).  
> `m.to` = total plant biomass including roots (= `m.rt + m.so`); used as the best available whole-plant size indicator.  
> `m.st` = all stem tissue including branches; approximates above-ground structural biomass (AGB excluding leaves and roots).  
> Both **OLS** and **SMA** regression slopes are reported. OLS minimises residuals in Y and is appropriate for prediction; SMA minimises residuals in both directions and is the standard for allometric inference (Warton et al. 2006).

```{r data-coverage}
cov <- data.frame(
  Combo     = names(results),
  N_records = sapply(seq_along(combos), function(i) {
    co <- combos[[i]]
    sum(!is.na(dat[[co$x]]) & dat[[co$x]] > 0 &
        !is.na(dat[[co$y]]) & dat[[co$y]] > 0, na.rm = TRUE)
  }),
  N_species = sapply(results, nrow),
  N_angio   = sapply(results, function(r) sum(r$clade == "Angiosperm", na.rm = TRUE)),
  N_gymno   = sapply(results, function(r) sum(r$clade == "Gymnosperm", na.rm = TRUE))
)

kable(cov,
      col.names = c("Combination", "Total records in BAAD",
                    paste0("Species (≥", MIN_N, " records)"),
                    "Angiosperms", "Gymnosperms"),
      caption = "Data coverage for the six focal allometric combinations")
```

---

```{r shared-plot-fns, include=FALSE}
# ── Shared helper: size-range plot ────────────────────────────────────────────
# Horizontal range bars per species, sorted by log_x_range (descending).
# Top max_sp species by record count shown if too many.
plot_size_range <- function(res, co, max_sp = 60) {
  if (nrow(res) == 0) { plot.new(); return(invisible(NULL)) }
  top <- res %>% arrange(desc(n)) %>% slice_head(n = max_sp)
  top <- top %>% arrange(log_x_range)
  top$speciesMatched <- factor(top$speciesMatched,
                               levels = top$speciesMatched)
  ggplot(top, aes(y = speciesMatched, colour = clade)) +
    geom_segment(aes(x = log_x_min, xend = log_x_max,
                     yend = speciesMatched), linewidth = 1, alpha = 0.8) +
    geom_point(aes(x = log_x_min), size = 1.5, shape = 16) +
    geom_point(aes(x = log_x_max), size = 1.5, shape = 16) +
    scale_colour_manual(values = c(Angiosperm = "#e41a1c",
                                   Gymnosperm = "#377eb8"),
                        na.value = "grey60", name = "Clade") +
    labs(
      x = paste0("log₁₀ ", co$xlab),
      y = NULL,
      title = paste0("Body size range per species — ", co$title_short),
      subtitle = paste0("Top ", min(max_sp, nrow(res)),
                        " species by record count; bars span [min, max] of ",
                        co$x, " in log₁₀ space")
    ) +
    theme(axis.text.y  = element_text(face = "italic", size = 6),
          panel.grid.major.y = element_line(colour = "grey95"))
}

# ── Shared helper: exponent vs. size range ────────────────────────────────────
plot_slope_vs_range <- function(res, co) {
  if (nrow(res) == 0) { plot.new(); return(invisible(NULL)) }
  # pivot OLS and SMA long
  long <- res %>%
    select(speciesMatched, clade, n, log_x_range,
           ols_slope, ols_lo, ols_hi,
           sma_slope, sma_lo, sma_hi) %>%
    pivot_longer(
      cols      = c(ols_slope, sma_slope),
      names_to  = "method",
      values_to = "slope"
    ) %>%
    mutate(
      lo     = ifelse(method == "ols_slope", ols_lo, sma_lo),
      hi     = ifelse(method == "ols_slope", ols_hi, sma_hi),
      method = recode(method, ols_slope = "OLS", sma_slope = "SMA")
    )

  wbe_label <- if (!is.na(co$wbe_ref))
    paste0("Dashed = ", co$wbe_ref, " (WBE); dotted = 1.0 (isometry). ")
  else
    "Dotted = 1.0 (isometry); no specific WBE prediction for this predictor. "

  p <- ggplot(long, aes(x = log_x_range, y = slope,
                   colour = clade, shape = method)) +
    geom_hline(yintercept = 1.00, linetype = "dotted",
               colour = "grey70", linewidth = 0.6) +
    geom_errorbar(aes(ymin = lo, ymax = hi),
                  width = 0, alpha = 0.2, linewidth = 0.5) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(aes(group = method), method = "lm",
                se = TRUE, colour = "black",
                linewidth = 0.8, alpha = 0.15) +
    scale_colour_manual(values = c(Angiosperm = "#e41a1c",
                                   Gymnosperm = "#377eb8"),
                        na.value = "grey60", name = "Clade") +
    scale_shape_manual(values = c(OLS = 16, SMA = 17), name = "Method") +
    facet_wrap(~ method) +
    labs(
      x = paste0("log₁₀ size range  [log₁₀(max) − log₁₀(min) of ", co$x, "]"),
      y = "Intraspecific slope",
      title = paste0("Slope vs. size range — ", co$title_short),
      subtitle = paste0(wbe_label, "Black line: OLS trend across species.")
    )
  if (!is.na(co$wbe_ref))
    p <- p + geom_hline(yintercept = co$wbe_ref, linetype = "dashed",
                        colour = "grey50", linewidth = 0.8)
  p
}

# ── Shared helper: slopes table ───────────────────────────────────────────────
table_slopes <- function(res, caption_txt) {
  if (nrow(res) == 0) {
    return(kable(data.frame(Note = "No species met the minimum record threshold.")))
  }
  res %>%
    arrange(desc(n)) %>%
    mutate(across(c(ols_slope, ols_lo, ols_hi,
                    sma_slope, sma_lo, sma_hi,
                    r2, log_x_range), ~ round(.x, 3)),
           speciesMatched = paste0("*", speciesMatched, "*")) %>%
    select(speciesMatched, family, clade, n,
           ols_slope, ols_lo, ols_hi,
           sma_slope, sma_lo, sma_hi,
           r2, log_x_range) %>%
    kable(
      col.names = c("Species", "Family", "Clade", "N",
                    "OLS slope", "OLS 95% lo", "OLS 95% hi",
                    "SMA slope", "SMA 95% lo", "SMA 95% hi",
                    "R²", "log₁₀ size range"),
      caption   = caption_txt,
      escape    = FALSE
    )
}
```

---

# Leaf mass ~ Basal stem diameter (`m.lf ~ d.ba`)

```{r sec1-summary}
r1  <- results[["m.lf~d.ba"]]
co1 <- combos[[1]]
cat("Species qualifying (≥", MIN_N, "records):", nrow(r1), "\n")
cat("Median OLS slope:", round(median(r1$ols_slope, na.rm=TRUE), 3), "\n")
cat("Median SMA slope:", round(median(r1$sma_slope, na.rm=TRUE), 3), "\n")
```

## Per-species slopes table {.tabset}

### All species (sorted by N)

```{r sec1-table, results='asis'}
table_slopes(r1, paste0("Leaf mass ~ basal diameter: per-species OLS and SMA slopes (≥ ",
                         MIN_N, " records per species, sorted by N)"))
```

## Body size range per species

```{r sec1-range, fig.height=14}
plot_size_range(r1, co1)
```

## Scaling slope vs. size range

```{r sec1-slope-range, fig.height=6}
plot_slope_vs_range(r1, co1)
```

## Slope distributions

```{r sec1-dist, fig.height=5}
r1_long <- r1 %>%
  pivot_longer(cols = c(ols_slope, sma_slope),
               names_to = "method", values_to = "slope") %>%
  mutate(method = recode(method, ols_slope = "OLS", sma_slope = "SMA"))

ggplot(r1_long, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = 2.00, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "grey60") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(Angiosperm="#e41a1c", Gymnosperm="#377eb8"),
                    na.value="grey60", name="Clade") +
  labs(x = "Intraspecific slope", y = "N species",
       title = "Leaf mass ~ basal diameter: intraspecific slope distribution",
       subtitle = "Dashed = 2 (WBE); dotted = 1.0 (isometry)")
```

---

# Leaf area ~ Basal stem diameter (`a.lf ~ d.ba`)

```{r sec2-summary}
r2  <- results[["a.lf~d.ba"]]
co2 <- combos[[2]]
cat("Species qualifying (≥", MIN_N, "records):", nrow(r2), "\n")
cat("Median OLS slope:", round(median(r2$ols_slope, na.rm=TRUE), 3), "\n")
cat("Median SMA slope:", round(median(r2$sma_slope, na.rm=TRUE), 3), "\n")
```

## Per-species slopes table

```{r sec2-table, results='asis'}
table_slopes(r2, paste0("Leaf area ~ basal diameter: per-species OLS and SMA slopes (≥ ",
                         MIN_N, " records per species, sorted by N)"))
```

## Body size range per species

```{r sec2-range, fig.height=14}
plot_size_range(r2, co2)
```

## Scaling slope vs. size range

```{r sec2-slope-range, fig.height=6}
plot_slope_vs_range(r2, co2)
```

## Slope distributions

```{r sec2-dist, fig.height=5}
r2_long <- r2 %>%
  pivot_longer(cols = c(ols_slope, sma_slope),
               names_to = "method", values_to = "slope") %>%
  mutate(method = recode(method, ols_slope = "OLS", sma_slope = "SMA"))

ggplot(r2_long, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = 2.00, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "grey60") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(Angiosperm="#e41a1c", Gymnosperm="#377eb8"),
                    na.value="grey60", name="Clade") +
  labs(x = "Intraspecific slope", y = "N species",
       title = "Leaf area ~ basal diameter: intraspecific slope distribution",
       subtitle = "Dashed = 2 (WBE); dotted = 1.0 (isometry)")
```

---

# Leaf mass ~ Total plant biomass (`m.lf ~ m.to`)

```{r sec3-summary}
r3  <- results[["m.lf~m.to"]]
co3 <- combos[[3]]
cat("Species qualifying (≥", MIN_N, "records):", nrow(r3), "\n")
cat("Median OLS slope:", round(median(r3$ols_slope, na.rm=TRUE), 3), "\n")
cat("Median SMA slope:", round(median(r3$sma_slope, na.rm=TRUE), 3), "\n")
```

## Per-species slopes table

```{r sec3-table, results='asis'}
table_slopes(r3, paste0("Leaf mass ~ total biomass: per-species OLS and SMA slopes (≥ ",
                         MIN_N, " records per species, sorted by N)"))
```

## Body size range per species

```{r sec3-range, fig.height=14}
plot_size_range(r3, co3)
```

## Scaling slope vs. size range

```{r sec3-slope-range, fig.height=6}
plot_slope_vs_range(r3, co3)
```

## Slope distributions

```{r sec3-dist, fig.height=5}
r3_long <- r3 %>%
  pivot_longer(cols = c(ols_slope, sma_slope),
               names_to = "method", values_to = "slope") %>%
  mutate(method = recode(method, ols_slope = "OLS", sma_slope = "SMA"))

ggplot(r3_long, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "grey60") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(Angiosperm="#e41a1c", Gymnosperm="#377eb8"),
                    na.value="grey60", name="Clade") +
  labs(x = "Intraspecific slope", y = "N species",
       title = "Leaf mass ~ total plant biomass: intraspecific slope distribution",
       subtitle = "Dashed = 3/4 (WBE prediction); dotted = 1.0 (isometry)")
```

---

# Leaf area ~ Total plant biomass (`a.lf ~ m.to`)

```{r sec4-summary}
r4  <- results[["a.lf~m.to"]]
co4 <- combos[[4]]
cat("Species qualifying (≥", MIN_N, "records):", nrow(r4), "\n")
cat("Median OLS slope:", round(median(r4$ols_slope, na.rm=TRUE), 3), "\n")
cat("Median SMA slope:", round(median(r4$sma_slope, na.rm=TRUE), 3), "\n")
```

## Per-species slopes table

```{r sec4-table, results='asis'}
table_slopes(r4, paste0("Leaf area ~ total biomass: per-species OLS and SMA slopes (≥ ",
                         MIN_N, " records per species, sorted by N)"))
```

## Body size range per species

```{r sec4-range, fig.height=14}
plot_size_range(r4, co4)
```

## Scaling slope vs. size range

```{r sec4-slope-range, fig.height=6}
plot_slope_vs_range(r4, co4)
```

## Slope distributions

```{r sec4-dist, fig.height=5}
r4_long <- r4 %>%
  pivot_longer(cols = c(ols_slope, sma_slope),
               names_to = "method", values_to = "slope") %>%
  mutate(method = recode(method, ols_slope = "OLS", sma_slope = "SMA"))

ggplot(r4_long, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "grey60") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(Angiosperm="#e41a1c", Gymnosperm="#377eb8"),
                    na.value="grey60", name="Clade") +
  labs(x = "Intraspecific slope", y = "N species",
       title = "Leaf area ~ total plant biomass: intraspecific slope distribution",
       subtitle = "Dashed = 3/4 (WBE prediction); dotted = 1.0 (isometry)")
```

---

# Leaf mass ~ Stem mass (`m.lf ~ m.st`)

```{r sec5-summary}
r5  <- results[["m.lf~m.st"]]
co5 <- combos[[5]]
cat("Species qualifying (≥", MIN_N, "records):", nrow(r5), "\n")
cat("Median OLS slope:", round(median(r5$ols_slope, na.rm=TRUE), 3), "\n")
cat("Median SMA slope:", round(median(r5$sma_slope, na.rm=TRUE), 3), "\n")
```

## Per-species slopes table

```{r sec5-table, results='asis'}
table_slopes(r5, paste0("Leaf mass ~ stem mass: per-species OLS and SMA slopes (≥ ",
                         MIN_N, " records per species, sorted by N)"))
```

## Body size range per species

```{r sec5-range, fig.height=14}
plot_size_range(r5, co5)
```

## Scaling slope vs. size range

```{r sec5-slope-range, fig.height=6}
plot_slope_vs_range(r5, co5)
```

## Slope distributions

```{r sec5-dist, fig.height=5}
r5_long <- r5 %>%
  pivot_longer(cols = c(ols_slope, sma_slope),
               names_to = "method", values_to = "slope") %>%
  mutate(method = recode(method, ols_slope = "OLS", sma_slope = "SMA"))

ggplot(r5_long, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "grey60") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(Angiosperm="#e41a1c", Gymnosperm="#377eb8"),
                    na.value="grey60", name="Clade") +
  labs(x = "Intraspecific slope", y = "N species",
       title = "Leaf mass ~ stem mass: intraspecific slope distribution",
       subtitle = "Dotted = 1.0 (isometry); no explicit WBE prediction for leaf ~ stem mass")
```

---

# Leaf area ~ Stem mass (`a.lf ~ m.st`)

```{r sec6-summary}
r6  <- results[["a.lf~m.st"]]
co6 <- combos[[6]]
cat("Species qualifying (≥", MIN_N, "records):", nrow(r6), "\n")
cat("Median OLS slope:", round(median(r6$ols_slope, na.rm=TRUE), 3), "\n")
cat("Median SMA slope:", round(median(r6$sma_slope, na.rm=TRUE), 3), "\n")
```

## Per-species slopes table

```{r sec6-table, results='asis'}
table_slopes(r6, paste0("Leaf area ~ stem mass: per-species OLS and SMA slopes (≥ ",
                         MIN_N, " records per species, sorted by N)"))
```

## Body size range per species

```{r sec6-range, fig.height=14}
plot_size_range(r6, co6)
```

## Scaling slope vs. size range

```{r sec6-slope-range, fig.height=6}
plot_slope_vs_range(r6, co6)
```

## Slope distributions

```{r sec6-dist, fig.height=5}
r6_long <- r6 %>%
  pivot_longer(cols = c(ols_slope, sma_slope),
               names_to = "method", values_to = "slope") %>%
  mutate(method = recode(method, ols_slope = "OLS", sma_slope = "SMA"))

ggplot(r6_long, aes(x = slope, fill = clade)) +
  geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.75, position = "stack") +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "grey60") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(Angiosperm="#e41a1c", Gymnosperm="#377eb8"),
                    na.value="grey60", name="Clade") +
  labs(x = "Intraspecific slope", y = "N species",
       title = "Leaf area ~ stem mass: intraspecific slope distribution",
       subtitle = "Dotted = 1.0 (isometry); no explicit WBE prediction for leaf ~ stem mass")
```

---

# Cross-combination summary

## Median slopes and spread across all six relationships

```{r cross-summary-table}
all_res <- bind_rows(
  lapply(seq_along(combos), function(i) {
    results[[i]] %>%
      mutate(combo     = names(results)[i],
             response  = combos[[i]]$y,
             predictor = combos[[i]]$x,
             wbe_ref   = combos[[i]]$wbe_ref)
  })
)

cross_sum <- all_res %>%
  group_by(combo) %>%
  summarise(
    n_species        = n(),
    wbe_pred         = ifelse(is.na(wbe_ref[1]), "none", as.character(wbe_ref[1])),
    med_ols          = round(median(ols_slope, na.rm = TRUE), 3),
    iqr_ols          = round(IQR(ols_slope, na.rm = TRUE), 3),
    pct_ols_lt_wbe   = round(mean(ols_slope < ifelse(is.na(wbe_ref[1]), Inf, wbe_ref[1]),
                                  na.rm = TRUE) * 100, 1),
    pct_ols_gt_iso   = round(mean(ols_slope > 1.00, na.rm = TRUE) * 100, 1),
    med_sma          = round(median(sma_slope, na.rm = TRUE), 3),
    iqr_sma          = round(IQR(sma_slope, na.rm = TRUE), 3),
    med_r2           = round(median(r2, na.rm = TRUE), 3),
    med_range        = round(median(log_x_range, na.rm = TRUE), 3),
    .groups = "drop"
  )

kable(cross_sum,
      col.names = c("Combination", "WBE exponent", "N species",
                    "Median OLS slope", "IQR (OLS)",
                    "% OLS < WBE", "% OLS > 1.0",
                    "Median SMA slope", "IQR (SMA)",
                    "Median R²", "Median log₁₀ range"),
      caption = paste0("Cross-combination summary of intraspecific slopes (≥ ", MIN_N,
                       " records/species). WBE exponent: 2.0 for diameter predictors,",
                       " 3/4 for total biomass, none stated for stem mass."))
```

## OLS vs. SMA slope comparison — all combinations

```{r ols-vs-sma, fig.height=7, fig.width=10}
all_res %>%
  filter(!is.na(ols_slope), !is.na(sma_slope)) %>%
  ggplot(aes(x = ols_slope, y = sma_slope, colour = clade)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50", linewidth = 0.8) +
  geom_point(aes(size = log10(n)), alpha = 0.5) +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                      na.value = "grey60", name = "Clade") +
  scale_size_continuous(range = c(1, 4), name = "log₁₀(N records)") +
  facet_wrap(~ combo, scales = "free", ncol = 3) +
  labs(
    x = "OLS slope", y = "SMA slope",
    title = "OLS vs. SMA intraspecific slopes — all six combinations",
    subtitle = "Dashed line = 1:1. SMA slopes are systematically steeper when R² < 1."
  )
```

## Size range vs. R² — all combinations

```{r range-r2, fig.height=7, fig.width=10}
all_res %>%
  filter(!is.na(r2)) %>%
  ggplot(aes(x = log_x_range, y = r2, colour = clade)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(aes(group = 1), method = "lm", colour = "black",
              se = TRUE, linewidth = 0.7, alpha = 0.15) +
  scale_colour_manual(values = c(Angiosperm = "#e41a1c", Gymnosperm = "#377eb8"),
                      na.value = "grey60", name = "Clade") +
  facet_wrap(~ combo, ncol = 3) +
  labs(
    x = "log₁₀ size range [log₁₀(max) − log₁₀(min)]",
    y = "R² (OLS)",
    title = "R² vs. body size range — all six combinations",
    subtitle = "Wider size range generally yields higher R²; slope estimates more reliable when R² > 0.7."
  )
```

## Slope vs. size range — all six combinations

```{r all-slope-range, fig.height=10, fig.width=11}
all_res %>%
  pivot_longer(cols = c(ols_slope, sma_slope),
               names_to = "method", values_to = "slope") %>%
  mutate(method = recode(method, ols_slope = "OLS", sma_slope = "SMA")) %>%
  filter(!is.na(slope)) %>%
  ggplot(aes(x = log_x_range, y = slope, colour = method)) +
  geom_hline(yintercept = 1.00, linetype = "dotted",
             colour = "grey80", linewidth = 0.5) +
  # Per-combo WBE reference lines (different exponents for d.ba vs m.to)
  geom_hline(data = data.frame(
               combo   = c("m.lf~d.ba", "a.lf~d.ba"),
               wbe_ref = 2.00),
             aes(yintercept = wbe_ref), inherit.aes = FALSE,
             linetype = "dashed", colour = "#e66101", linewidth = 0.7) +
  geom_hline(data = data.frame(
               combo   = c("m.lf~m.to", "a.lf~m.to"),
               wbe_ref = 0.75),
             aes(yintercept = wbe_ref), inherit.aes = FALSE,
             linetype = "dashed", colour = "#5e3c99", linewidth = 0.7) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, alpha = 0.9) +
  scale_colour_manual(values = c(OLS = "#1b7837", SMA = "#762a83"),
                      name = "Method") +
  facet_wrap(~ combo, ncol = 3) +
  labs(
    x = "log₁₀ size range",
    y = "Intraspecific slope",
    title = "Intraspecific slope vs. log₁₀ body size range — all six combinations",
    subtitle = "OLS slopes are typically compressed toward zero at small size ranges (regression dilution).\nSMA slopes are more consistent across size range.\nOrange dashed = 2.0 (WBE, d.ba); purple dashed = 3/4 (WBE, m.to); dotted = 1.0 (isometry)."
  ) +
  theme(strip.text = element_text(size = 9))
```

---

# Methods

- **Data:** BAAD v1 (Falster et al. 2015, *Ecology* 96:1445). Individual-level
  plant measurements from >170 primary studies.
- **Minimum threshold:** ≥ `r MIN_N` individual records per species per combination.
  Species with zero size range (all individuals same size) excluded.
- **OLS regression:** `lm(log10 Y ~ log10 X)`. Slope = predicted change in
  log₁₀ Y per unit log₁₀ X. 95% CIs via `confint()`. Appropriate when predicting Y
  from X; known to underestimate the true allometric slope when R² < 1 (regression
  dilution / attenuation bias).
- **SMA regression:** Standardised Major Axis via `smatr::sma(Y ~ X)` on
  log₁₀-transformed variables. 95% CIs from the coefficient matrix. SMA minimises
  residuals in both X and Y directions and gives unbiased slope estimates for allometric
  inference regardless of R². SMA slope = OLS slope / r (Pearson correlation), so
  SMA ≥ OLS always.
- **Size range:** log₁₀(max predictor) − log₁₀(min predictor) within each species.
  A key diagnostic: OLS slopes estimated from narrow size ranges are strongly biased
  downward; SMA slopes are less affected but CIs widen with small range.
- **Reference slopes — two distinct WBE predictions:**
  1. **Leaf area/mass vs. basal diameter (`d.ba`):** WBE exponent = **2** (dashed orange).
     Under area-preserving branching, conduit cross-section is conserved across branching
     levels, so total leaf area ∝ d² (West et al. 1999; Enquist et al. 1998).
  2. **Leaf area/mass vs. total plant biomass (`m.to`):** WBE exponent = **3/4** (dashed
     purple). Whole-plant metabolic rate ∝ M^(3/4), and leaf area sets gas exchange
     capacity, so leaf area ∝ M^(3/4) (Enquist & Niklas 2001; Enquist et al. 2009).
  3. **Leaf area/mass vs. stem mass (`m.st`):** No explicit WBE exponent; dotted line
     at 1.0 (isometry) shown for reference only.
  - 1.0 (dotted, all panels) = isometric (geometric) scaling.
- **BAAD variable notes:**
  - `d.ba` = diameter at base of stem [m]; more commonly measured than DBH in BAAD.
  - `m.to` = total plant biomass [kg] (= `m.rt + m.so`; includes root mass).
  - `m.st` = all stem tissue [kg] (branches, bark; excludes leaves and roots).
