---
title: "BAAD Stem Diameter vs Above-Ground Leaf Biomass Scaling"
subtitle: "Interspecific, Intraspecific, Functional Group, and Family-Level Analyses"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5
)

safe_require <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(sprintf("Package '%s' is required but not installed.", pkg), call. = FALSE)
  }
}

required_pkgs <- c("baad.data", "dplyr", "ggplot2", "tidyr", "purrr", "tibble", "knitr", "scales")
invisible(lapply(required_pkgs, safe_require))

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(knitr)
  library(scales)
})

theme_set(theme_bw(base_size = 12))

HAS_SMATR <- requireNamespace("smatr", quietly = TRUE)
HAS_LMODEL2 <- requireNamespace("lmodel2", quietly = TRUE)

if (!HAS_SMATR && !HAS_LMODEL2) {
  stop("Need either 'smatr' or 'lmodel2' for RMA/SMA fits. Install one of them.", call. = FALSE)
}

if (HAS_SMATR) {
  suppressPackageStartupMessages(library(smatr))
}
if (HAS_LMODEL2) {
  suppressPackageStartupMessages(library(lmodel2))
}

extract_sma_stats <- function(fit_obj) {
  out <- list(slope = NA_real_, slope_lwr = NA_real_, slope_upr = NA_real_, r2 = NA_real_)

  coerce_coef_table <- function(x) {
    if (is.null(x)) return(NULL)
    if (is.matrix(x) || is.data.frame(x)) return(as.data.frame(x, stringsAsFactors = FALSE))
    if (is.list(x) && length(x) > 0L) {
      for (elt in x) {
        tbl <- coerce_coef_table(elt)
        if (!is.null(tbl)) return(tbl)
      }
    }
    NULL
  }

  coef_sources <- list(
    tryCatch(fit_obj$coef, error = function(e) NULL),
    tryCatch(fit_obj$coefficients, error = function(e) NULL)
  )

  coef_tbl <- NULL
  for (src in coef_sources) {
    coef_tbl <- coerce_coef_table(src)
    if (!is.null(coef_tbl)) break
  }

  if (!is.null(coef_tbl) && nrow(coef_tbl) > 0L && ncol(coef_tbl) > 0L) {
    rn <- tolower(trimws(rownames(coef_tbl)))
    slope_row <- which(rn %in% c("slope", "coef", "b", "beta"))
    if (length(slope_row) == 0L) slope_row <- if (nrow(coef_tbl) >= 2L) 2L else 1L
    slope_row <- slope_row[1]

    coln <- tolower(trimws(colnames(coef_tbl)))
    numeric_cols <- which(vapply(coef_tbl, function(z) is.numeric(suppressWarnings(as.numeric(z))), logical(1)))
    slope_col <- which(grepl("slope|coef|estimate|est", coln))
    if (length(slope_col) == 0L) slope_col <- numeric_cols
    if (length(slope_col) > 0L) {
      out$slope <- suppressWarnings(as.numeric(coef_tbl[slope_row, slope_col[1]]))
    }

    ci_l <- which(grepl("lower|lcl|lwr|low", coln))
    ci_u <- which(grepl("upper|ucl|upr|high", coln))
    if (length(ci_l) > 0L) out$slope_lwr <- suppressWarnings(as.numeric(coef_tbl[slope_row, ci_l[1]]))
    if (length(ci_u) > 0L) out$slope_upr <- suppressWarnings(as.numeric(coef_tbl[slope_row, ci_u[1]]))
  }

  r2_sources <- list(
    tryCatch(fit_obj$r2, error = function(e) NULL),
    tryCatch(fit_obj$rsquared, error = function(e) NULL)
  )
  for (r2_src in r2_sources) {
    if (is.null(r2_src)) next
    if (is.list(r2_src) && length(r2_src) > 0L) {
      r2_vals <- unlist(r2_src, use.names = FALSE)
    } else {
      r2_vals <- r2_src
    }
    r2_num <- suppressWarnings(as.numeric(r2_vals))
    r2_num <- r2_num[is.finite(r2_num)]
    if (length(r2_num) > 0L) {
      out$r2 <- r2_num[1]
      break
    }
  }

  out
}

fit_scaling_models <- function(df, x = "log_d", y = "log_mlf") {
  dat <- df %>% filter(is.finite(.data[[x]]), is.finite(.data[[y]]))
  n <- nrow(dat)

  if (n < 3) {
    return(tibble(
      n = n,
      model = c("OLS", "RMA"),
      method = c("lm", if (HAS_SMATR) "smatr::sma" else if (HAS_LMODEL2) "lmodel2" else "unavailable"),
      slope = NA_real_,
      slope_lwr = NA_real_,
      slope_upr = NA_real_,
      intercept = NA_real_,
      r2 = NA_real_,
      se = NA_real_,
      t_stat = NA_real_,
      p_beta_eq_2 = NA_real_,
      note = "Insufficient data (n < 3)"
    ))
  }

  fml <- as.formula(sprintf("%s ~ %s", y, x))

  ols <- lm(fml, data = dat)
  ols_sum <- summary(ols)
  ols_ci <- tryCatch(confint(ols), error = function(e) matrix(c(NA_real_, NA_real_), nrow = 2))

  out_ols <- tibble(
    n = n,
    model = "OLS",
    method = "lm",
    slope = unname(coef(ols)[2]),
    slope_lwr = unname(ols_ci[2, 1]),
    slope_upr = unname(ols_ci[2, 2]),
    intercept = unname(coef(ols)[1]),
    r2 = unname(ols_sum$r.squared),
    se = unname(ols_sum$coefficients[2, 2]),
    t_stat = unname((coef(ols)[2] - 2) / ols_sum$coefficients[2, 2]),
    p_beta_eq_2 = unname(2 * pt(-abs((coef(ols)[2] - 2) / ols_sum$coefficients[2, 2]), df = ols$df.residual)),
    note = NA_character_
  )

  out_rma <- NULL
  smatr_note <- NULL
  if (HAS_SMATR) {
    sma_fit <- tryCatch(smatr::sma(fml, data = dat), error = function(e) NULL)
    if (!is.null(sma_fit)) {
      ss <- extract_sma_stats(sma_fit)
      if (is.finite(ss$slope)) {
        out_rma <- tibble(
          n = n,
          model = "RMA",
          method = "smatr::sma",
          slope = ss$slope,
          slope_lwr = ss$slope_lwr,
          slope_upr = ss$slope_upr,
          intercept = NA_real_,
          r2 = ss$r2,
          se = NA_real_,
          t_stat = NA_real_,
          p_beta_eq_2 = NA_real_,
          note = "RMA beta=2 p-value unavailable; interpret CI"
        )
      } else {
        smatr_note <- "smatr fit parsed but slope was non-finite"
      }
    }
  }

  if (is.null(out_rma) && HAS_LMODEL2) {
    l2 <- tryCatch(lmodel2::lmodel2(fml, data = dat), error = function(e) NULL)
    if (!is.null(l2) && is.data.frame(l2$regression.results)) {
      rr <- l2$regression.results
      rma_row <- rr[rr$Method %in% c("RMA", "SMA"), , drop = FALSE]
      if (nrow(rma_row) > 0L) {
        out_rma <- tibble(
          n = n,
          model = "RMA",
          method = "lmodel2",
          slope = rma_row$slope[1],
          slope_lwr = NA_real_,
          slope_upr = NA_real_,
          intercept = rma_row$Intercept[1],
          r2 = rma_row$r[1]^2,
          se = NA_real_,
          t_stat = NA_real_,
          p_beta_eq_2 = NA_real_,
          note = if (!is.null(smatr_note)) paste0(smatr_note, "; RMA CI/test unavailable from lmodel2") else "RMA CI/test unavailable from lmodel2"
        )
      }
    }
  }

  if (is.null(out_rma)) {
    out_rma <- tibble(
      n = n,
      model = "RMA",
      method = "unavailable",
      slope = NA_real_,
      slope_lwr = NA_real_,
      slope_upr = NA_real_,
      intercept = NA_real_,
      r2 = NA_real_,
      se = NA_real_,
      t_stat = NA_real_,
      p_beta_eq_2 = NA_real_,
      note = if (!is.null(smatr_note)) paste0(smatr_note, "; RMA fitting unavailable") else "RMA fitting unavailable"
    )
  }

  bind_rows(out_ols, out_rma)
}

baad_raw <- baad.data::baad_data()$data

dat0 <- baad_raw %>%
  transmute(
    species = speciesMatched,
    family = family,
    pft = pft,
    d_bh_m = d.bh,
    m_lf_kg = m.lf
  )

dat1 <- dat0 %>% filter(!is.na(d_bh_m), !is.na(m_lf_kg))
dat2 <- dat1 %>% filter(d_bh_m > 0, m_lf_kg > 0)

dat <- dat2 %>%
  mutate(
    pft_norm = toupper(trimws(as.character(pft))),
    pft_group = case_when(
      grepl("A$", pft_norm) ~ "Angiosperm",
      grepl("G$", pft_norm) ~ "Gymnosperm",
      TRUE ~ NA_character_
    ),
    log_d = log10(d_bh_m),
    log_mlf = log10(m_lf_kg)
  )

attrition <- tibble(
  step = c(
    "Raw BAAD records",
    "Non-missing d.bh and m.lf",
    "Positive d.bh and m.lf",
    "Positive + non-missing species",
    "Positive + non-missing family",
    "Positive + mapped pft_group"
  ),
  n_records = c(
    nrow(dat0),
    nrow(dat1),
    nrow(dat2),
    sum(dat2$species != "" & !is.na(dat2$species)),
    sum(dat2$family != "" & !is.na(dat2$family)),
    sum(!is.na(dat$pft_group))
  )
) %>%
  mutate(removed_from_previous = c(NA_integer_, diff(n_records) * -1L))

analysis_dat <- dat %>%
  filter(!is.na(species), species != "")

species_label_checks <- analysis_dat %>%
  group_by(species) %>%
  summarise(
    family_nonmissing_n = n_distinct(family[!is.na(family) & family != ""]),
    pft_nonmissing_n = n_distinct(pft_group[!is.na(pft_group) & pft_group != ""]),
    family_conflict = family_nonmissing_n > 1,
    pft_conflict = pft_nonmissing_n > 1,
    .groups = "drop"
  )

species_label_diagnostics <- tibble(
  metric = c(
    "Species with conflicting non-missing family labels",
    "Species with conflicting non-missing pft_group labels",
    "Species with any conflicting non-missing label"
  ),
  n_species = c(
    sum(species_label_checks$family_conflict, na.rm = TRUE),
    sum(species_label_checks$pft_conflict, na.rm = TRUE),
    sum(species_label_checks$family_conflict | species_label_checks$pft_conflict, na.rm = TRUE)
  )
)

species_means <- analysis_dat %>%
  group_by(species) %>%
  summarise(
    family_raw = first(na.omit(family)),
    pft_group_raw = first(na.omit(pft_group)),
    n_individuals = n(),
    log_d_mean = mean(log_d, na.rm = TRUE),
    log_mlf_mean = mean(log_mlf, na.rm = TRUE),
    log_d_range = diff(range(log_d, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(species_label_checks %>% select(species, family_conflict, pft_conflict), by = "species") %>%
  mutate(
    family = if_else(family_conflict, NA_character_, family_raw),
    pft_group = if_else(pft_conflict, NA_character_, pft_group_raw)
  ) %>%
  select(-family_raw, -pft_group_raw, -family_conflict, -pft_conflict) %>%
  filter(is.finite(log_d_mean), is.finite(log_mlf_mean))
```

# Scope and Data Definition

This report evaluates the allometric relation between stem diameter and above-ground leaf biomass in BAAD using:

- Predictor: d.bh (m)
- Response: m.lf (kg)
- Species identifier: speciesMatched
- Family: family
- Functional group mapping from pft:
  - DA, EA -> Angiosperm
  - DG, EG -> Gymnosperm

The target relation is written as a power law:

$$
m_{lf} = \alpha \cdot d_{bh}^{\beta}
$$

or, in log10 space:

$$
\log_{10}(m_{lf}) = \log_{10}(\alpha) + \beta \log_{10}(d_{bh})
$$

A formal contrast to $\beta = 2$ is included where standard errors are available.

# Filtering and Attrition

```{r attrition-table}
kable(attrition, digits = 0, caption = "Explicit filtering attrition table")
```

```{r threshold-declarations}
thresholds <- tibble(
  component = c(
    "Intraspecific species inclusion",
    "Intraspecific species inclusion",
    "Family inclusion",
    "Family inclusion"
  ),
  threshold = c(
    "n >= 11 individuals",
    "log10(d.bh) range >= 0.7",
    "n >= 30 individual records",
    ">= 5 species"
  )
)
kable(thresholds, caption = "Declared thresholds used in this report")
```

```{r species-label-diagnostics}
kable(species_label_diagnostics, digits = 0, caption = "Species-level label consistency diagnostics")
```

# Guidance Consultation and Implementation

This analysis was cross-checked against enhanced-theory and ecology-user guidance to keep modeling assumptions explicit, ecologically interpretable, and reproducible.

- Model formalism: the core model is declared as a diameter-leaf-biomass power law in raw units and log10-linearized form, with OLS and RMA estimated for complementary Model I and Model II interpretation.
- Declared thresholds: all inclusion rules are pre-declared and tabulated (intraspecific: n >= 11 and log10 diameter range >= 0.7; family: n >= 30 records and >= 5 species).
- Uncertainty and caveats: approximation limits are made explicit, including package-dependent RMA uncertainty reporting and approximate tests versus beta = 2.
- Reproducible workflow structure: setup, filtering attrition, threshold declaration, interspecific and stratified fits, diagnostics, caveats, and session metadata are organized as an auditable sequence.

# Theory Framing and Testable Predictions

Leaf biomass should increase with stem diameter as an increasing power function. The exponent $\beta$ describes how rapidly foliage mass scales with stem size. A value of $\beta = 2$ implies a very steep relation where leaf mass scales approximately with area-like growth in diameter terms.

To anchor interpretation, this report estimates OLS and RMA exponents and tests whether observed slopes differ from 2 where uncertainty can be quantified.

# Interspecific Analysis (Primary)

Primary interspecific inference uses species-level means in log space to reduce pseudo-replication from uneven within-species sampling.

```{r interspecific-fits}
inter_df <- species_means %>%
  transmute(log_d = log_d_mean, log_mlf = log_mlf_mean)

inter_fit <- fit_scaling_models(inter_df)
inter_fit <- inter_fit %>% mutate(beta0 = 2)

kable(inter_fit, digits = 3, caption = "Interspecific (species means) OLS and RMA fits")
```

```{r interspecific-plot, fig.height=6}
ggplot(species_means, aes(log_d_mean, log_mlf_mean)) +
  geom_point(alpha = 0.6, color = "#2c7fb8") +
  geom_smooth(method = "lm", se = TRUE, color = "#d95f0e", linewidth = 1) +
  labs(
    x = "Species mean log10(d.bh [m])",
    y = "Species mean log10(m.lf [kg])",
    title = "Primary interspecific scaling (species means)",
    subtitle = paste0("n species = ", comma(nrow(species_means)))
  )
```

```{r pseudo-replication-comparison}
ind_fit <- fit_scaling_models(analysis_dat)
comp_tab <- bind_rows(
  inter_fit %>% mutate(context = "Species means"),
  ind_fit %>% mutate(context = "Individual records")
) %>%
  select(context, model, method, n, slope, slope_lwr, slope_upr, r2, note)

kable(comp_tab, digits = 3, caption = "Comparison: interspecific species-means vs individual-record fit")
```

# Angiosperm vs Gymnosperm Split

## Individual-level context

```{r pft-individual-fits}
pft_ind <- analysis_dat %>%
  filter(!is.na(pft_group)) %>%
  group_split(pft_group) %>%
  map_df(function(dg) {
    grp <- unique(dg$pft_group)[1]
    fit_scaling_models(dg) %>% mutate(group = grp, context = "Individual")
  }) %>%
  select(context, group, everything())

kable(pft_ind, digits = 3, caption = "Angiosperm vs gymnosperm fits: individual-level")
```

```{r pft-individual-plot, fig.height=6}
ggplot(analysis_dat %>% filter(!is.na(pft_group)), aes(log_d, log_mlf, color = pft_group)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(
    x = "log10(d.bh [m])",
    y = "log10(m.lf [kg])",
    title = "Individual-level scaling by major functional group",
    color = "Group"
  )
```

## Interspecific species-mean context

```{r pft-species-fits}
pft_species <- species_means %>%
  filter(!is.na(pft_group)) %>%
  group_split(pft_group) %>%
  map_df(function(ds) {
    grp <- unique(ds$pft_group)[1]
    fit_scaling_models(ds %>% transmute(log_d = log_d_mean, log_mlf = log_mlf_mean)) %>%
      mutate(group = grp, context = "Species means")
  }) %>%
  select(context, group, everything())

kable(pft_species, digits = 3, caption = "Angiosperm vs gymnosperm fits: species means")
```

```{r pft-species-plot, fig.height=6}
ggplot(species_means %>% filter(!is.na(pft_group)), aes(log_d_mean, log_mlf_mean, color = pft_group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(
    x = "Species mean log10(d.bh [m])",
    y = "Species mean log10(m.lf [kg])",
    title = "Species-mean scaling by major functional group",
    color = "Group"
  )
```

# Intraspecific Scaling (Species with n >= 11 and log10 range >= 0.7)

```{r intraspecific-screen}
intra_screen <- analysis_dat %>%
  group_by(species) %>%
  summarise(
    n = n(),
    log_d_range = diff(range(log_d, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(species_means %>% select(species, family, pft_group), by = "species") %>%
  mutate(
    pass_n = n >= 11,
    pass_range = log_d_range >= 0.7,
    eligible = pass_n & pass_range
  )

kable(
  intra_screen %>%
    summarise(
      species_total = n(),
      species_n_ge_11 = sum(pass_n),
      species_range_ge_0_7 = sum(pass_range),
      species_eligible = sum(eligible)
    ),
  caption = "Intraspecific screening summary"
)
```

```{r intraspecific-fits}
eligible_species <- intra_screen %>% filter(eligible) %>% pull(species)

intra_fits <- analysis_dat %>%
  filter(species %in% eligible_species) %>%
  group_by(species) %>%
  group_split() %>%
  map_df(function(ds) {
    sp <- unique(ds$species)[1]
    fit_scaling_models(ds) %>%
      mutate(species = sp)
  }) %>%
  left_join(species_means %>% select(species, family, pft_group), by = "species") %>%
  relocate(species, family, pft_group, model)

kable(
  intra_fits %>% arrange(desc(n), species, model),
  digits = 3,
  caption = "Per-species intraspecific OLS and RMA fits"
)
```

```{r intraspecific-top-table}
top_species <- intra_screen %>%
  filter(eligible) %>%
  arrange(desc(n), desc(log_d_range)) %>%
  slice_head(n = 20)

kable(top_species, digits = 3, caption = "Top eligible intraspecific species by sample size")
```

```{r intraspecific-forest, fig.height=8}
forest_df <- intra_fits %>%
  filter(species %in% (top_species %>% pull(species))) %>%
  mutate(species = reorder(species, n))

ggplot(forest_df, aes(x = slope, y = species, color = model)) +
  geom_vline(xintercept = 2, linetype = "dashed", color = "gray40") +
  geom_errorbarh(aes(xmin = slope_lwr, xmax = slope_upr), height = 0.18, alpha = 0.7,
                 position = position_dodge(width = 0.5), na.rm = TRUE) +
  geom_point(size = 2.1, position = position_dodge(width = 0.5)) +
  labs(
    x = "Estimated slope (beta)",
    y = NULL,
    title = "Intraspecific slope forest plot (top species)",
    subtitle = "Dashed line marks beta = 2"
  )
```

# Family-Level Analysis

Well-sampled family rule: at least 30 individual records and at least 5 species.

```{r family-screen}
family_screen <- analysis_dat %>%
  filter(!is.na(family), family != "") %>%
  group_by(family) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  mutate(eligible = n_records >= 30 & n_species >= 5)

kable(family_screen %>% arrange(desc(n_records)), caption = "Family screening table", digits = 0)
```

```{r family-fits}
eligible_families <- family_screen %>% filter(eligible) %>% pull(family)

family_fits <- analysis_dat %>%
  filter(family %in% eligible_families) %>%
  group_by(family) %>%
  group_split() %>%
  map_df(function(dfam) {
    fam <- unique(dfam$family)[1]
    fit_scaling_models(dfam) %>%
      mutate(family = fam)
  }) %>%
  left_join(family_screen, by = "family") %>%
  relocate(family, n_records, n_species, model)

kable(family_fits %>% arrange(desc(n_records), family, model), digits = 3,
      caption = "Family-level OLS and RMA fits for well-sampled families")
```

```{r family-forest, fig.height=7}
ggplot(family_fits, aes(x = slope, y = reorder(family, n_records), color = model)) +
  geom_vline(xintercept = 2, linetype = "dashed", color = "gray40") +
  geom_errorbarh(aes(xmin = slope_lwr, xmax = slope_upr), height = 0.18,
                 position = position_dodge(width = 0.5), na.rm = TRUE) +
  geom_point(size = 2.2, position = position_dodge(width = 0.5)) +
  labs(
    x = "Estimated slope (beta)",
    y = NULL,
    title = "Family-level slope comparison",
    subtitle = "Families with n >= 30 and >= 5 species"
  )
```

# Diagnostics and Caveats

```{r diagnostics-ols}
inter_ols <- lm(log_mlf_mean ~ log_d_mean, data = species_means)
inter_aug <- species_means %>%
  mutate(
    fitted = fitted(inter_ols),
    resid = resid(inter_ols),
    std_resid = rstandard(inter_ols),
    cook = cooks.distance(inter_ols)
  )

cook_cut <- 4 / nrow(species_means)

diagnostic_summary <- tibble(
  metric = c(
    "Species-mean OLS n",
    "Species-mean OLS R2",
    "Species with |standardized residual| > 2",
    "Species with Cook's distance > 4/n"
  ),
  value = c(
    nrow(species_means),
    summary(inter_ols)$r.squared,
    sum(abs(inter_aug$std_resid) > 2, na.rm = TRUE),
    sum(inter_aug$cook > cook_cut, na.rm = TRUE)
  )
)

kable(diagnostic_summary, digits = 3, caption = "Global species-mean OLS diagnostics")
```

```{r diagnostics-plots, fig.height=4.8, fig.width=11}
par(mfrow = c(1, 2))
plot(inter_aug$fitted, inter_aug$resid,
     xlab = "Fitted log10(m.lf)", ylab = "Residual",
     main = "Residual vs fitted")
abline(h = 0, col = "red", lty = 2)
qqnorm(inter_aug$resid, main = "Q-Q plot of residuals")
qqline(inter_aug$resid, col = "red", lty = 2)
par(mfrow = c(1, 1))
```

```{r leverage-table}
lev_tab <- inter_aug %>%
  arrange(desc(cook)) %>%
  select(species, n_individuals, cook, std_resid) %>%
  slice_head(n = 15)

kable(lev_tab, digits = 4, caption = "Top potentially influential species (Cook's distance)")
```

```{r sampling-bias-figure, fig.height=5.5}
ggplot(species_means, aes(n_individuals)) +
  geom_histogram(bins = 30, fill = "#3182bd", color = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    x = "Individuals per species",
    y = "Number of species",
    title = "Sampling imbalance across species"
  )
```

## Caveat Reporting

1. BAAD combines studies with heterogeneous protocols, sites, and taxonomic representation.
2. Many species are sparsely sampled, motivating species-mean primary inference to reduce pseudo-replication.
3. RMA uncertainty can depend on package implementation details; smatr provides preferred CIs, while lmodel2 fallback may omit CIs.
4. The intraspecific threshold (n >= 11 and log10 range >= 0.7) improves slope identifiability but excludes narrow-range species.
5. Family-level slopes can still reflect compositional effects and uneven within-family sampling.
6. Hypothesis tests against beta = 2 are approximate where standard errors are inferred from CIs.

# Reproducibility Metadata

```{r session-info}
si <- sessionInfo()
print(si$R.version$version.string)
print(si$platform)
print(c(
  HAS_SMATR = HAS_SMATR,
  HAS_LMODEL2 = HAS_LMODEL2
))
```

```{r analysis-summaries}
summary_block <- tibble(
  item = c(
    "Records analyzed (positive d.bh and m.lf, with species)",
    "Species in interspecific analysis",
    "Eligible intraspecific species",
    "Eligible families"
  ),
  value = c(
    nrow(analysis_dat),
    nrow(species_means),
    length(eligible_species),
    length(eligible_families)
  )
)

kable(summary_block, caption = "Analysis coverage summary")
```
