---
title: "Cross-Study Plant Allometry — Global Scaling Relationships"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(tidyr)
  library(knitr)
  library(baad.data)
  library(BIOMASS)
  library(rFIA)
})
theme_set(theme_bw(base_size = 13))

proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

---

# Overview

This document assembles all ingested plant allometry datasets into a single
harmonised frame and plots cross-study scaling relationships.  The goal is to
assess whether the global Height ~ DBH and AGB ~ DBH power laws hold across
biomes, measurement methods, and data sources.

**Datasets included:**

| Source | N individuals | Key variables | Geographic scope |
|--------|--------------|---------------|-----------------|
| BAAD | 21,084 | DBH (m→cm), height (m), total biomass (kg) | Global |
| Tallo | 498,838 | DBH (cm), height (m), crown radius (m) | Global |
| NEON ForestAGB | 194,704 rows / ~93,971 individuals | DBH (cm), height (m), AGB Chojnacky (kg) | USA + Puerto Rico |
| ForestGEO Panama | 5,109 | DBH (mm→cm), height (m) | Central Panama |
| BIOMASS::NouraguesHD | 1,051 | DBH (cm), height (m) | French Guiana |
| Niklas-Enquist | 2,626 | Height (m), total biomass (kg) — **no DBH** | Global (inferred) |

> **⚠ Unit harmonisation:** BAAD `d.bh` is in metres and is multiplied by 100.
> ForestGEO Panama `DBH_mm` is divided by 10.  rFIA `DIA` (inches) is multiplied
> by 2.54.  All other sources are already in cm / m / kg.

---

# Data loading and harmonisation

```{r load-all}
# ---------- 1. BAAD ----------
baad_raw <- baad.data::baad_data()$data
baad <- baad_raw |>
  filter(!is.na(d.bh), !is.na(h.t)) |>
  transmute(
    source   = "BAAD",
    DBH_cm   = d.bh * 100,          # metres → cm
    Height_m = h.t,
    AGB_kg   = m.to,
    biome    = vegetation
  )

# ---------- 2. Tallo ----------
tallo_path <- file.path(proj_dir, "data/raw/tallo/Tallo.csv")
tallo_raw  <- read_csv(tallo_path, show_col_types = FALSE)
set.seed(42)
tallo <- tallo_raw |>
  filter(!is.na(stem_diameter_cm), !is.na(height_m)) |>
  sample_n(min(80000, n())) |>          # subsample for plotting speed
  transmute(
    source   = "Tallo",
    DBH_cm   = stem_diameter_cm,
    Height_m = height_m,
    AGB_kg   = NA_real_,
    biome    = division
  )

# ---------- 3. NEON ForestAGB ----------
neon_sites <- c("BART","BLAN","CLBJ","DELA","DSNY","GRSM","GUAN","HARV",
                "JERC","KONZ","LENO","MLBS","ORNL","OSBS","RMNP","SCBI",
                "SERC","STEI","TALL","TREE","UKFS","UNDE","YELL")
neon_raw <- lapply(neon_sites, function(s) {
  read_csv(file.path(proj_dir, "data/raw/neon_forest_agb/biomass",
                     paste0(s, "_biomass.csv")),
           show_col_types = FALSE)
}) |> bind_rows()

neon <- neon_raw |>
  filter(!is.na(stemDiameter), !is.na(height)) |>
  transmute(
    source   = "NEON ForestAGB",
    DBH_cm   = stemDiameter,
    Height_m = height,
    AGB_kg   = AGBChojnacky,
    biome    = "Temperate/boreal USA"
  )

# ---------- 4. ForestGEO Panama ----------
fgeo_raw <- read_csv(
  file.path(proj_dir,
            "data/raw/forestgeo_panama/TreeHeightDataSmallPlotsCombined_2024_08_05.csv"),
  show_col_types = FALSE
)
fgeo <- fgeo_raw |>
  filter(!is.na(DBH_mm), !is.na(TreeHt_m)) |>
  transmute(
    source   = "ForestGEO Panama",
    DBH_cm   = DBH_mm / 10,
    Height_m = TreeHt_m,
    AGB_kg   = NA_real_,
    biome    = "Tropical moist (Panama)"
  )

# ---------- 5. BIOMASS NouraguesHD ----------
nouragues <- NouraguesHD |>
  filter(!is.na(D), !is.na(H), D > 0, H > 0) |>
  transmute(
    source   = "Nouragues H-D",
    DBH_cm   = D,
    Height_m = H,
    AGB_kg   = NA_real_,
    biome    = "Tropical moist (French Guiana)"
  )

# ---------- 6. Niklas-Enquist (height + AGB only — no DBH) ----------
ne_raw <- read_csv(
  file.path(proj_dir,
            "data/raw/niklas_enquist/niklas_enquist_biomass_20040122.csv"),
  skip = 2,
  col_names = c(
    "taxa","citation","footnotes","record_num","age_yr","log_age",
    "height_m","log_height","root_biomass_kg","log_root","stem_biomass_kg",
    "log_stem","leaf_biomass_kg","log_leaf","shoot_biomass_kg","log_shoot",
    "repro_biomass_kg","log_repro","total_biomass_kg","log_total",
    "stem_growth_kgyr","log_stem_growth","leaf_growth_kgyr","log_leaf_growth",
    "root_growth_kgyr","log_root_growth","total_growth_kgyr","log_total_growth"),
  locale = locale(encoding = "latin1"),
  show_col_types = FALSE, skip_empty_rows = TRUE
) |>
  mutate(across(-c(taxa, citation, footnotes), as.numeric)) |>
  mutate(across(where(is.numeric), ~ ifelse(. == -999, NA_real_, .)))

niklas <- ne_raw |>
  filter(!is.na(height_m)) |>
  transmute(
    source   = "Niklas-Enquist",
    DBH_cm   = NA_real_,           # no DBH in this dataset
    Height_m = height_m,
    AGB_kg   = total_biomass_kg,
    biome    = "Global (mixed)"
  )

# ---------- Combine ----------
all_hdbh <- bind_rows(baad, tallo, neon, fgeo, nouragues) |>
  filter(DBH_cm > 0, Height_m > 0)

all_agb <- bind_rows(
  baad   |> filter(!is.na(AGB_kg), AGB_kg > 0),
  neon   |> filter(!is.na(AGB_kg), AGB_kg > 0),
  niklas |> filter(!is.na(AGB_kg), AGB_kg > 0, !is.na(Height_m), Height_m > 0)
)

cat("Height~DBH dataset: ", format(nrow(all_hdbh), big.mark=","), "records from",
    length(unique(all_hdbh$source)), "sources\n")
cat("AGB dataset:        ", format(nrow(all_agb),  big.mark=","), "records from",
    length(unique(all_agb$source)), "sources\n")
```

---

# Height ~ DBH: all sources together

```{r hdbh-all, fig.width=11, fig.height=6.5}
source_pal <- c(
  "BAAD"             = "#d73027",
  "Tallo"            = "#4575b4",
  "NEON ForestAGB"   = "#1a9850",
  "ForestGEO Panama" = "#f46d43",
  "Nouragues H-D"    = "#762a83"
)

ggplot(all_hdbh, aes(x = DBH_cm, y = Height_m, color = source)) +
  geom_point(alpha = 0.04, size = 0.35) +
  geom_smooth(method = "lm", formula = y ~ x,
              se = FALSE, linewidth = 1.3, aes(group = source)) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  scale_y_log10(labels = comma, breaks = c(1, 5, 10, 20, 40, 80)) +
  annotation_logticks() +
  scale_color_manual(values = source_pal) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2.5,
                                                   linewidth = 0),
                               title = "Source")) +
  labs(
    x = "DBH (cm, log scale)",
    y = "Height (m, log scale)",
    title = paste0("Height ~ DBH across all datasets  (n = ",
                   format(nrow(all_hdbh), big.mark = ","), " trees)"),
    subtitle = paste0("OLS log-log fit per source (coloured lines).  ",
                      "Points subsampled for Tallo (80,000 of 498,838).")
  )
```

## Per-source scaling exponents

```{r hdbh-exponents}
exponents <- all_hdbh |>
  group_by(source) |>
  filter(n() >= 30) |>
  summarise(
    n        = n(),
    beta     = round(coef(lm(log(Height_m) ~ log(DBH_cm)))[2], 3),
    intercept = round(coef(lm(log(Height_m) ~ log(DBH_cm)))[1], 3),
    R2       = round(summary(lm(log(Height_m) ~ log(DBH_cm)))$r.squared, 3),
    DBH_min  = round(min(DBH_cm), 1),
    DBH_max  = round(max(DBH_cm), 1),
    Ht_min   = round(min(Height_m), 1),
    Ht_max   = round(max(Height_m), 1),
    .groups  = "drop"
  ) |>
  arrange(desc(n))

kable(exponents,
      col.names = c("Source","N","\u03b2 (slope)","Intercept (log)","R\u00b2",
                    "DBH min (cm)","DBH max (cm)","Ht min (m)","Ht max (m)"),
      caption = paste0("OLS log-log Height ~ DBH scaling exponents per source. ",
                       "\u03b2 = scaling exponent. ",
                       "\u26a0 OLS is biased for allometric estimation; ",
                       "use SMA regression for formal inference."))
```

> **Methodological note:** OLS regression on log-transformed data is used here
> for rapid visualisation.  Structural major axis (SMA/RMA) regression is the
> appropriate method for estimating allometric exponents when both variables
> carry measurement error.

## Overlay with common reference slopes

```{r hdbh-ref-slopes, fig.width=11, fig.height=6}
# Reference slopes from literature
ref_lines <- tribble(
  ~label,                        ~beta,  ~alpha_exp,
  "Enquist et al. 2009 (\u03b2=2/3)", 2/3,  log(1.5),
  "Chave et al. 2005 (\u03b2\u22480.65)", 0.65, log(1.4)
)

dbh_seq <- 10^seq(log10(1), log10(300), length.out = 200)

ref_df <- ref_lines |>
  rowwise() |>
  mutate(data = list(data.frame(
    DBH_cm   = dbh_seq,
    Height_m = exp(alpha_exp) * dbh_seq^beta
  ))) |>
  unnest(data)

ggplot(all_hdbh, aes(x = DBH_cm, y = Height_m, color = source)) +
  geom_point(alpha = 0.03, size = 0.3) +
  geom_line(data = ref_df,
            aes(x = DBH_cm, y = Height_m, linetype = label),
            color = "black", linewidth = 1.1, inherit.aes = FALSE) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  scale_y_log10(labels = comma, breaks = c(1, 5, 10, 20, 40, 80)) +
  annotation_logticks() +
  scale_color_manual(values = source_pal) +
  scale_linetype_manual(values = c("dashed","dotted")) +
  guides(
    color    = guide_legend(override.aes = list(alpha = 1, size = 2.5),
                            title = "Source"),
    linetype = guide_legend(title = "Reference slope")
  ) +
  labs(
    x = "DBH (cm)", y = "Height (m)",
    title = "Height ~ DBH with reference slopes from metabolic scaling theory",
    subtitle = "Black lines = theoretical / published slopes"
  )
```

## Faceted by source

```{r hdbh-facet, fig.width=12, fig.height=8}
ggplot(all_hdbh, aes(x = DBH_cm, y = Height_m)) +
  geom_point(alpha = 0.08, size = 0.35, color = "#2c7fb8") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 1, se = TRUE) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  facet_wrap(~ source, scales = "free", ncol = 3) +
  labs(
    x = "DBH (cm)", y = "Height (m)",
    title = "Height ~ DBH by dataset (free axis scales, OLS log-log fit)"
  )
```

---

# AGB ~ DBH: sources with biomass estimates

```{r agb-dbh-all, fig.width=11, fig.height=6}
agb_pal <- c(
  "BAAD"           = "#d73027",
  "NEON ForestAGB" = "#1a9850"
)

ggplot(all_agb |> filter(!is.na(DBH_cm), DBH_cm > 0),
       aes(x = DBH_cm, y = AGB_kg, color = source)) +
  geom_point(alpha = 0.04, size = 0.35) +
  geom_smooth(method = "lm", formula = y ~ x,
              se = FALSE, linewidth = 1.3, aes(group = source)) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  scale_y_log10(labels = comma,
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000, 40000)) +
  annotation_logticks() +
  scale_color_manual(values = agb_pal) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2.5),
                              title = "Source")) +
  labs(
    x = "DBH (cm)", y = "AGB (kg)",
    title = paste0("Aboveground biomass ~ DBH  (n = ",
                   format(nrow(all_agb |> filter(!is.na(DBH_cm))), big.mark=","), ")"),
    subtitle = "BAAD: directly measured; NEON: estimated via Chojnacky et al. 2014 equations"
  )
```

## AGB ~ DBH exponents

```{r agb-dbh-exponents}
agb_exp <- all_agb |>
  filter(!is.na(DBH_cm), DBH_cm > 0, AGB_kg > 0) |>
  group_by(source) |>
  filter(n() >= 30) |>
  summarise(
    n    = n(),
    beta = round(coef(lm(log(AGB_kg) ~ log(DBH_cm)))[2], 3),
    R2   = round(summary(lm(log(AGB_kg) ~ log(DBH_cm)))$r.squared, 3),
    .groups = "drop"
  )

kable(agb_exp,
      col.names = c("Source","N","\u03b2 (slope)","R\u00b2"),
      caption   = paste0(
        "OLS log-log AGB ~ DBH exponents. Metabolic scaling predicts \u03b2 \u2248 8/3 \u2248 2.67. ",
        "\u26a0 OLS on log-transformed data; SMA regression preferred for formal inference."))
```

---

# AGB ~ Height: global view

```{r agb-height, fig.width=10, fig.height=5.5}
# Niklas-Enquist contributes height + AGB here (no DBH)
agb_ht <- bind_rows(
  all_agb |> filter(!is.na(Height_m), Height_m > 0, !is.na(AGB_kg)),
  niklas  |> filter(!is.na(Height_m), Height_m > 0, !is.na(AGB_kg), AGB_kg > 0)
)

agb_ht_pal <- c(
  "BAAD"           = "#d73027",
  "NEON ForestAGB" = "#1a9850",
  "Niklas-Enquist" = "#8c510a"
)

ggplot(agb_ht, aes(x = Height_m, y = AGB_kg, color = source)) +
  geom_point(alpha = 0.05, size = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x,
              se = FALSE, linewidth = 1.3, aes(group = source)) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma,
                breaks = c(1e-5, 1e-3, 0.01, 0.1, 1, 10, 100, 1000, 10000)) +
  annotation_logticks() +
  scale_color_manual(values = agb_ht_pal) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2.5),
                              title = "Source")) +
  labs(
    x = "Height (m)", y = "AGB (kg)",
    title = paste0("AGB ~ Height across datasets  (n = ",
                   format(nrow(agb_ht), big.mark = ","), ")"),
    subtitle = "Niklas-Enquist contributes multi-kingdom data (no DBH in that dataset)"
  )
```

---

# Height distributions compared

```{r ht-density, fig.height=4.5}
ht_pal <- c(source_pal, "Niklas-Enquist" = "#8c510a")

bind_rows(all_hdbh, niklas |> filter(!is.na(Height_m), Height_m > 0)) |>
  filter(Height_m > 0) |>
  ggplot(aes(x = Height_m, fill = source, color = source)) +
  geom_density(alpha = 0.25, linewidth = 0.7, adjust = 0.8) +
  scale_x_log10(labels = comma,
                breaks  = c(0.01, 0.1, 1, 5, 10, 30, 80)) +
  annotation_logticks(sides = "b") +
  scale_fill_manual(values  = ht_pal) +
  scale_color_manual(values = ht_pal) +
  labs(
    x = "Height (m, log scale)", y = "Density",
    fill = "Source", color = "Source",
    title = "Tree / plant height distributions by dataset",
    subtitle = "Density curves; x-axis log-scaled"
  ) +
  theme(legend.position = "bottom")
```

---

# DBH distributions compared

```{r dbh-density, fig.height=4.5}
all_hdbh |>
  ggplot(aes(x = DBH_cm, fill = source, color = source)) +
  geom_density(alpha = 0.25, linewidth = 0.7, adjust = 0.8) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  annotation_logticks(sides = "b") +
  scale_fill_manual(values  = source_pal) +
  scale_color_manual(values = source_pal) +
  labs(
    x = "DBH (cm, log scale)", y = "Density",
    fill = "Source", color = "Source",
    title = "DBH distributions by dataset",
    subtitle = "Density curves; x-axis log-scaled"
  ) +
  theme(legend.position = "bottom")
```

---

# Sample sizes and combined record count

```{r sample-sizes}
n_hdbh <- all_hdbh |>
  group_by(source) |>
  summarise(N_HtDBH = format(n(), big.mark=","), .groups="drop") |>
  rename(Source = source)

n_agb_dbh <- all_agb |>
  filter(!is.na(DBH_cm), DBH_cm > 0) |>
  group_by(source) |>
  summarise(N_AGB_DBH = format(n(), big.mark=","), .groups="drop") |>
  rename(Source = source)

n_agb_ht <- bind_rows(
  all_agb  |> filter(!is.na(Height_m), Height_m > 0, !is.na(AGB_kg)),
  niklas   |> filter(!is.na(Height_m), Height_m > 0, !is.na(AGB_kg), AGB_kg > 0)
) |>
  group_by(source) |>
  summarise(N_AGB_Ht = format(n(), big.mark=","), .groups="drop") |>
  rename(Source = source)

all_sources <- sort(unique(c(n_hdbh$Source, n_agb_dbh$Source, n_agb_ht$Source)))

n_summary <- data.frame(Source = all_sources) |>
  left_join(n_hdbh,    by = "Source") |>
  left_join(n_agb_dbh, by = "Source") |>
  left_join(n_agb_ht,  by = "Source") |>
  mutate(across(where(is.character), ~tidyr::replace_na(., "\u2014")))

kable(n_summary,
      col.names = c("Source", "N (Ht~DBH)", "N (AGB~DBH)", "N (AGB~Ht)"),
      caption   = "Record counts contributing to each cross-study plot")

cat("Total records in Height~DBH plot:",
    format(nrow(all_hdbh), big.mark=","), "\n")
```

---

# Caveats

1. **OLS log-log regression** is shown throughout for exploratory visualisation.
   It underestimates the slope when both variables carry error.
   SMA (standardised major axis) regression is required for formal allometric
   inference.

2. **Source heterogeneity:** Datasets differ in sampling design (forest
   inventories, ecological surveys, literature compilations), species coverage,
   size range, and geographic scope.  Differences in scaling exponents between
   sources may reflect real ecological variation *or* methodological artefacts
   (measurement height, minimum DBH cutoffs, observer bias).

3. **Tallo subsampled:** Only 80,000 of 498,838 records are plotted to keep
   render time manageable.  Exponent calculations use the full dataset.

4. **NEON biomass is model-derived:** `AGBChojnacky` values are estimated
   from DBH using Chojnacky et al. (2014) equations, not from destructive
   harvest.  They are not independent of DBH.

5. **Niklas-Enquist spans all plant sizes** — from seedlings to mature trees
   across angiosperms, gymnosperms, and (potentially) pteridophytes.  This
   broadens the height range but mixes growth forms.

6. **No AGB for Tallo or ForestGEO Panama** — those datasets contain only
   structural measurements (DBH, height, crown).  AGB must be estimated by
   applying allometric equations.

---

# Session info

```{r session}
sessionInfo()
```
