---
title: "Cross-Study Plant Allometry — SMA/RMA Scaling Exponents"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5)
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(tidyr)
  library(knitr)
  library(smatr)
  library(baad.data)
  library(BIOMASS)
})
theme_set(theme_bw(base_size = 13))
proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

---

# Overview

This document re-examines the cross-study Height ~ DBH and AGB ~ DBH allometric
relationships using **Standardised Major Axis (SMA)** regression — also called
Reduced Major Axis (RMA) when variables are standardised.  SMA is the
appropriate method for estimating allometric exponents when **both variables
carry measurement error**, which is always the case for tree height, DBH, and
biomass.

OLS regression on log-transformed data is provided alongside SMA for comparison.
OLS systematically underestimates allometric slopes because it minimises
residuals in only one direction (y).

**Theoretical predictions tested:**

| Relationship | MST / published prediction | Reference |
|---|---|---|
| H ~ DBH ($\beta$) | 2/3 ≈ 0.667 | Enquist et al. 2009; WBE pipe model |
| H ~ DBH ($\beta$) | 0.65 | Chave et al. 2005 |
| AGB ~ DBH ($\beta$) | 8/3 ≈ 2.667 | West, Brown, Enquist 1999 |

---

# Data loading and harmonisation

```{r load-data}
# ---------- 1. BAAD ----------
baad_raw <- baad.data::baad_data()$data
baad <- baad_raw |>
  filter(!is.na(d.bh), !is.na(h.t)) |>
  transmute(
    source   = "BAAD",
    DBH_cm   = d.bh * 100,
    Height_m = h.t,
    AGB_kg   = m.to
  )

# ---------- 2. Tallo ----------
tallo_raw <- read_csv(
  file.path(proj_dir, "data/raw/tallo/Tallo.csv"),
  show_col_types = FALSE
)
set.seed(42)
tallo <- tallo_raw |>
  filter(!is.na(stem_diameter_cm), !is.na(height_m)) |>
  sample_n(min(80000, n())) |>
  transmute(
    source   = "Tallo",
    DBH_cm   = stem_diameter_cm,
    Height_m = height_m,
    AGB_kg   = NA_real_
  )

# ---------- 3. NEON ForestAGB ----------
neon_sites <- c("BART","BLAN","CLBJ","DELA","DSNY","GRSM","GUAN","HARV",
                "JERC","KONZ","LENO","MLBS","ORNL","OSBS","RMNP","SCBI",
                "SERC","STEI","TALL","TREE","UKFS","UNDE","YELL")
neon_raw <- lapply(neon_sites, function(s) {
  read_csv(file.path(proj_dir, "data/raw/neon_forest_agb/biomass",
                     paste0(s, "_biomass.csv")),
           show_col_types = FALSE)
}) |> bind_rows()

neon <- neon_raw |>
  filter(!is.na(stemDiameter), !is.na(height)) |>
  transmute(
    source   = "NEON ForestAGB",
    DBH_cm   = stemDiameter,
    Height_m = height,
    AGB_kg   = AGBChojnacky
  )

# ---------- 4. ForestGEO Panama ----------
fgeo_raw <- read_csv(
  file.path(proj_dir,
    "data/raw/forestgeo_panama/TreeHeightDataSmallPlotsCombined_2024_08_05.csv"),
  show_col_types = FALSE
)
fgeo <- fgeo_raw |>
  filter(!is.na(DBH_mm), !is.na(TreeHt_m)) |>
  transmute(
    source   = "ForestGEO Panama",
    DBH_cm   = DBH_mm / 10,
    Height_m = TreeHt_m,
    AGB_kg   = NA_real_
  )

# ---------- 5. Nouragues H-D ----------
nouragues <- NouraguesHD |>
  filter(!is.na(D), !is.na(H), D > 0, H > 0) |>
  transmute(
    source   = "Nouragues H-D",
    DBH_cm   = D,
    Height_m = H,
    AGB_kg   = NA_real_
  )

# ---------- 6. Niklas-Enquist (height + AGB; no DBH) ----------
ne_raw <- read_csv(
  file.path(proj_dir,
    "data/raw/niklas_enquist/niklas_enquist_biomass_20040122.csv"),
  skip = 2,
  col_names = c(
    "taxa","citation","footnotes","record_num","age_yr","log_age",
    "height_m","log_height","root_biomass_kg","log_root","stem_biomass_kg",
    "log_stem","leaf_biomass_kg","log_leaf","shoot_biomass_kg","log_shoot",
    "repro_biomass_kg","log_repro","total_biomass_kg","log_total",
    "stem_growth_kgyr","log_stem_growth","leaf_growth_kgyr","log_leaf_growth",
    "root_growth_kgyr","log_root_growth","total_growth_kgyr","log_total_growth"),
  locale = locale(encoding = "latin1"),
  show_col_types = FALSE, skip_empty_rows = TRUE
) |>
  mutate(across(-c(taxa, citation, footnotes), as.numeric)) |>
  mutate(across(where(is.numeric), ~ ifelse(. == -999, NA_real_, .)))

niklas <- ne_raw |>
  filter(!is.na(height_m)) |>
  transmute(
    source   = "Niklas-Enquist",
    DBH_cm   = NA_real_,
    Height_m = height_m,
    AGB_kg   = total_biomass_kg
  )

# ---------- Combined frames ----------
all_hdbh <- bind_rows(baad, tallo, neon, fgeo, nouragues) |>
  filter(DBH_cm > 0, Height_m > 0) |>
  mutate(log_DBH = log(DBH_cm), log_H = log(Height_m))

all_agb <- bind_rows(
    baad |> filter(!is.na(AGB_kg), AGB_kg > 0),
    neon |> filter(!is.na(AGB_kg), AGB_kg > 0)
  ) |>
  filter(DBH_cm > 0) |>
  mutate(log_DBH = log(DBH_cm), log_AGB = log(AGB_kg))

niklas_agb_ht <- niklas |>
  filter(!is.na(Height_m), Height_m > 0, !is.na(AGB_kg), AGB_kg > 0) |>
  mutate(log_H = log(Height_m), log_AGB = log(AGB_kg))

cat("H~DBH dataset: ", format(nrow(all_hdbh), big.mark=","),
    "records /", length(unique(all_hdbh$source)), "sources\n")
cat("AGB~DBH dataset:", format(nrow(all_agb),  big.mark=","),
    "records /", length(unique(all_agb$source)), "sources\n")
```

---

# Height ~ DBH: SMA exponents per source

SMA fits `log(H) ~ log(DBH)` with `smatr::sma(method = "SMA")`.
For each source we test whether the slope equals the two key theoretical
predictions: β = 2/3 (WBE) and β = 0.65 (Chave et al. 2005), using
`sma(..., slope.test = value)`.

```{r hdbh-sma-table}
sources_hdbh <- sort(unique(all_hdbh$source))

fit_rows_hdbh <- lapply(sources_hdbh, function(src) {
  d <- all_hdbh |> filter(source == src)
  if (nrow(d) < 30) return(NULL)

  fit_wbe <- sma(log_H ~ log_DBH, data = d, method = "SMA",
                 slope.test = 2/3)
  fit_chv <- sma(log_H ~ log_DBH, data = d, method = "SMA",
                 slope.test = 0.65)
  fit_ols <- lm(log_H ~ log_DBH, data = d)

  cf <- fit_wbe$coef[[1]]

  data.frame(
    source     = src,
    n          = nrow(d),
    SMA_slope  = round(cf["slope", "coef(SMA)"], 3),
    CI_lo      = round(cf["slope", "lower limit"], 3),
    CI_hi      = round(cf["slope", "upper limit"], 3),
    OLS_slope  = round(coef(fit_ols)[2], 3),
    R2         = round(summary(fit_ols)$r.squared, 3),
    p_vs_WBE   = round(fit_wbe$slopetest[[1]]$p, 4),
    p_vs_Chave = round(fit_chv$slopetest[[1]]$p, 4),
    stringsAsFactors = FALSE
  )
})

sma_hdbh_tab <- bind_rows(fit_rows_hdbh)

kable(sma_hdbh_tab,
  col.names = c("Source","N",
                "SMA \u03b2","95% CI lo","95% CI hi",
                "OLS \u03b2","R\u00b2",
                "p (vs 2/3)","p (vs 0.65)"),
  caption = paste0(
    "SMA log-log H ~ DBH exponents by dataset.  ",
    "95% CIs from smatr coef table.  ",
    "OLS slope shown for comparison (biased downward).  ",
    "p-values test H\u2080: slope = theoretical value."
  )
)
```

## Dot plot: SMA slopes vs theoretical predictions

```{r hdbh-sma-dotplot, fig.height=4.5}
source_pal <- c(
  "BAAD"             = "#d73027",
  "Tallo"            = "#4575b4",
  "NEON ForestAGB"   = "#1a9850",
  "ForestGEO Panama" = "#f46d43",
  "Nouragues H-D"    = "#762a83"
)

sma_hdbh_tab |>
  mutate(source = factor(source, levels = rev(sort(unique(source))))) |>
  ggplot(aes(x = SMA_slope, y = source, color = source)) +
  geom_vline(xintercept = 2/3,  linetype = "dashed", color = "#4575b4",
             linewidth = 0.8) +
  geom_vline(xintercept = 0.65, linetype = "dotted", color = "#8c510a",
             linewidth = 0.8) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.25, linewidth = 1) +
  geom_point(aes(size = log10(n)), alpha = 0.9) +
  scale_color_manual(values = source_pal, guide = "none") +
  scale_size_continuous(range = c(3, 8),
                        name  = "log\u2081\u2080(N)",
                        breaks = c(2, 3, 4, 5)) +
  annotate("text", x = 2/3  + 0.01, y = 0.5,
           label = "WBE 2/3",    color = "#4575b4", hjust = 0, size = 3.5) +
  annotate("text", x = 0.65 + 0.01, y = 1.1,
           label = "Chave 0.65", color = "#8c510a", hjust = 0, size = 3.5) +
  labs(
    x = "SMA slope (log H ~ log DBH)",
    y = NULL,
    title = "SMA scaling exponent \u03b2 for H ~ DBH by dataset",
    subtitle = "Error bars = 95% CI from smatr"
  )
```

## OLS vs SMA slope comparison

```{r hdbh-ols-vs-sma, fig.height=4}
sma_hdbh_tab |>
  select(source, OLS_slope, SMA_slope, n) |>
  pivot_longer(cols = c(OLS_slope, SMA_slope),
               names_to = "method", values_to = "slope") |>
  mutate(method = recode(method, OLS_slope = "OLS", SMA_slope = "SMA"),
         source = factor(source, levels = sort(unique(source)))) |>
  ggplot(aes(x = source, y = slope, color = method, group = method)) +
  geom_hline(yintercept = 2/3, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 0.6, alpha = 0.5) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = c(OLS = "#d73027", SMA = "#4575b4"),
                     name = "Method") +
  annotate("text", x = 0.6, y = 2/3 + 0.012,
           label = "WBE 2/3", color = "grey40", hjust = 0, size = 3.2) +
  labs(
    x = NULL, y = "log-log slope",
    title = "OLS vs SMA slope for H ~ DBH: OLS is consistently lower",
    subtitle = "Dashed line = WBE prediction (2/3)"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Scatter with SMA fits overlaid

```{r hdbh-scatter-sma, fig.width=11, fig.height=6.5}
sma_lines_hdbh <- lapply(sources_hdbh, function(src) {
  d <- all_hdbh |> filter(source == src)
  if (nrow(d) < 30) return(NULL)
  fit <- sma(log_H ~ log_DBH, data = d, method = "SMA")
  rng <- range(d$log_DBH)
  xs  <- seq(rng[1], rng[2], length.out = 150)
  cf  <- fit$coef[[1]]
  ys  <- cf["elevation","coef(SMA)"] + cf["slope","coef(SMA)"] * xs
  data.frame(source = src, DBH_cm = exp(xs), Height_m = exp(ys))
}) |> bind_rows()

set.seed(42)
sub_plot <- all_hdbh |>
  group_by(source) |> sample_n(min(15000, n())) |> ungroup()

ggplot(sub_plot, aes(x = DBH_cm, y = Height_m, color = source)) +
  geom_point(alpha = 0.04, size = 0.35) +
  geom_line(data = sma_lines_hdbh,
            aes(x = DBH_cm, y = Height_m, color = source),
            linewidth = 1.4, inherit.aes = FALSE) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  scale_y_log10(labels = comma, breaks = c(1, 5, 10, 20, 40, 80)) +
  annotation_logticks() +
  scale_color_manual(values = source_pal) +
  guides(color = guide_legend(
    override.aes = list(alpha = 1, size = 2.5, linewidth = 1.5),
    title = "Source")) +
  labs(
    x = "DBH (cm, log scale)", y = "Height (m, log scale)",
    title = "H ~ DBH: SMA fits per dataset",
    subtitle = paste0("Solid lines = SMA log-log fits  (n = ",
                      format(nrow(all_hdbh), big.mark=","), " trees; ",
                      "Tallo subsampled to 80,000)")
  )
```

---

# Height ~ DBH: pooled cross-study SMA

```{r hdbh-pooled-sma}
fit_pool_wbe <- sma(log_H ~ log_DBH, data = all_hdbh, method = "SMA",
                    slope.test = 2/3)
fit_pool_chv <- sma(log_H ~ log_DBH, data = all_hdbh, method = "SMA",
                    slope.test = 0.65)
pool_ols     <- lm(log_H ~ log_DBH, data = all_hdbh)
cf_pool      <- fit_pool_wbe$coef[[1]]

kable(data.frame(
  Method   = c("SMA (pooled)", "OLS (pooled)"),
  N        = format(nrow(all_hdbh), big.mark=","),
  slope    = c(round(cf_pool["slope","coef(SMA)"], 4),
               round(coef(pool_ols)[2], 4)),
  CI_lo    = c(round(cf_pool["slope","lower limit"], 4), NA),
  CI_hi    = c(round(cf_pool["slope","upper limit"], 4), NA),
  R2       = round(summary(pool_ols)$r.squared, 3),
  p_vs_WBE = c(round(fit_pool_wbe$slopetest[[1]]$p, 6), NA),
  p_vs_Chv = c(round(fit_pool_chv$slopetest[[1]]$p, 6), NA),
  stringsAsFactors = FALSE
),
col.names = c("Method","N","Slope","95% CI lo","95% CI hi","R\u00b2",
              "p (vs 2/3)","p (vs 0.65)"),
caption = "Pooled cross-study H ~ DBH SMA and OLS results")
```

> **Note:** With > 700,000 records the pooled test has enormous power; even
> trivial deviations from the theoretical slope will be significant.  The
> magnitude of departure (SMA slope minus the theoretical 2/3 or 0.65) is
> the primary effect-size metric.

---

# AGB ~ DBH: SMA exponents per source

Available sources with both AGB and DBH: BAAD (direct measurement) and
NEON ForestAGB (Chojnacky model-derived).

> **AGB note:** NEON AGB values are estimated from DBH via Chojnacky et al.
> (2014).  The AGB~DBH relationship for NEON largely reflects the embedded
> allometric equations, not independent biomass data.

```{r agb-sma-table}
sources_agb <- sort(unique(all_agb$source))

fit_rows_agb <- lapply(sources_agb, function(src) {
  d <- all_agb |> filter(source == src)
  if (nrow(d) < 30) return(NULL)

  fit_mst <- sma(log_AGB ~ log_DBH, data = d, method = "SMA",
                 slope.test = 8/3)
  fit_ols <- lm(log_AGB ~ log_DBH, data = d)
  cf      <- fit_mst$coef[[1]]

  data.frame(
    source    = src,
    n         = nrow(d),
    SMA_slope = round(cf["slope","coef(SMA)"], 3),
    CI_lo     = round(cf["slope","lower limit"], 3),
    CI_hi     = round(cf["slope","upper limit"], 3),
    OLS_slope = round(coef(fit_ols)[2], 3),
    R2        = round(summary(fit_ols)$r.squared, 3),
    p_vs_MST  = round(fit_mst$slopetest[[1]]$p, 4),
    stringsAsFactors = FALSE
  )
}) |> bind_rows()

kable(fit_rows_agb,
  col.names = c("Source","N","SMA \u03b2","95% CI lo","95% CI hi",
                "OLS \u03b2","R\u00b2","p (vs 8/3 \u2248 2.667)"),
  caption = paste0(
    "SMA log-log AGB ~ DBH exponents. ",
    "MST prediction: \u03b2 = 8/3 \u2248 2.667 (West et al. 1999). ",
    "\u26a0 NEON AGB is model-derived from DBH (not independent)."
  )
)
```

## AGB ~ DBH: SMA scatter

```{r agb-sma-scatter, fig.height=6}
agb_pal <- c("BAAD" = "#d73027", "NEON ForestAGB" = "#1a9850")

sma_lines_agb <- lapply(sources_agb, function(src) {
  d <- all_agb |> filter(source == src)
  if (nrow(d) < 30) return(NULL)
  fit <- sma(log_AGB ~ log_DBH, data = d, method = "SMA")
  rng <- range(d$log_DBH)
  xs  <- seq(rng[1], rng[2], length.out = 150)
  cf  <- fit$coef[[1]]
  ys  <- cf["elevation","coef(SMA)"] + cf["slope","coef(SMA)"] * xs
  data.frame(source = src, DBH_cm = exp(xs), AGB_kg = exp(ys))
}) |> bind_rows()

set.seed(42)
agb_sub <- all_agb |>
  group_by(source) |> sample_n(min(20000, n())) |> ungroup()

ggplot(agb_sub, aes(x = DBH_cm, y = AGB_kg, color = source)) +
  geom_point(alpha = 0.05, size = 0.4) +
  geom_line(data = sma_lines_agb,
            aes(x = DBH_cm, y = AGB_kg, color = source),
            linewidth = 1.4, inherit.aes = FALSE) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  scale_y_log10(labels = comma,
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000, 40000)) +
  annotation_logticks() +
  scale_color_manual(values = agb_pal) +
  guides(color = guide_legend(
    override.aes = list(alpha = 1, size = 2.5, linewidth = 1.5),
    title = "Source")) +
  labs(
    x = "DBH (cm, log scale)", y = "AGB (kg, log scale)",
    title = "AGB ~ DBH: SMA fits per dataset",
    subtitle = "Solid lines = SMA log-log fits.  NEON: Chojnacky-derived AGB (not independent)."
  )
```

---

# AGB ~ Height: Niklas-Enquist + BAAD + NEON

Niklas-Enquist lacks DBH; height is the shared predictor across all three sources.

```{r agb-ht-sma-table}
agb_ht_all <- bind_rows(
    all_agb |> filter(!is.na(Height_m), Height_m > 0) |>
      mutate(log_H = log(Height_m)),
    niklas_agb_ht
  )

agb_ht_rows <- agb_ht_all |>
  group_by(source) |>
  filter(n() >= 30) |>
  group_split() |>
  lapply(function(d) {
    src     <- unique(d$source)
    fit3    <- sma(log_AGB ~ log_H, data = d, method = "SMA", slope.test = 3)
    fit267  <- sma(log_AGB ~ log_H, data = d, method = "SMA", slope.test = 8/3)
    fit_ols <- lm(log_AGB ~ log_H, data = d)
    cf      <- fit3$coef[[1]]
    data.frame(
      source    = src,
      n         = nrow(d),
      SMA_slope = round(cf["slope","coef(SMA)"], 3),
      CI_lo     = round(cf["slope","lower limit"], 3),
      CI_hi     = round(cf["slope","upper limit"], 3),
      OLS_slope = round(coef(fit_ols)[2], 3),
      R2        = round(summary(fit_ols)$r.squared, 3),
      p_vs_3    = round(fit3$slopetest[[1]]$p, 4),
      p_vs_267  = round(fit267$slopetest[[1]]$p, 4),
      stringsAsFactors = FALSE
    )
  }) |> bind_rows()

kable(agb_ht_rows,
  col.names = c("Source","N","SMA \u03b2","95% CI lo","95% CI hi",
                "OLS \u03b2","R\u00b2","p (vs 3)","p (vs 8/3)"),
  caption = "SMA log-log AGB ~ Height exponents. Tests: \u03b2 = 3 (isometric) and \u03b2 = 8/3."
)
```

---

# Summary: all SMA exponents

```{r all-exponents-summary}
hdbh_summ <- sma_hdbh_tab |>
  transmute(relationship = "H ~ DBH", source, n,
            SMA_slope, CI_lo, CI_hi, OLS_slope, R2,
            theoretical = "2/3 (WBE)", p_ref = p_vs_WBE)

agb_dbh_summ <- fit_rows_agb |>
  transmute(relationship = "AGB ~ DBH", source, n,
            SMA_slope, CI_lo, CI_hi, OLS_slope, R2,
            theoretical = "8/3 (MST)", p_ref = p_vs_MST)

summary_all <- bind_rows(hdbh_summ, agb_dbh_summ) |>
  arrange(relationship, desc(n))

kable(summary_all,
  col.names = c("Relationship","Source","N","SMA \u03b2",
                "CI lo","CI hi","OLS \u03b2","R\u00b2",
                "Theoretical ref.","p (vs ref)"),
  caption = "All SMA allometric exponents \u2014 cross-study summary"
)
```

## Forest plot

```{r forest-plot, fig.height=6}
summary_all |>
  mutate(
    label = paste0(source, "\n(", relationship, ")"),
    sig   = ifelse(p_ref < 0.05, "p < 0.05", "n.s.")
  ) |>
  ggplot(aes(x = SMA_slope, y = reorder(label, SMA_slope),
             color = relationship, shape = sig)) +
  geom_vline(data = data.frame(relationship = c("H ~ DBH","AGB ~ DBH"),
                                xint = c(2/3, 8/3)),
             aes(xintercept = xint, color = relationship),
             linetype = "dashed", linewidth = 0.8, inherit.aes = FALSE) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.35, linewidth = 1) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c("H ~ DBH" = "#4575b4", "AGB ~ DBH" = "#d73027"),
                     name = "Relationship") +
  scale_shape_manual(values = c("p < 0.05" = 16, "n.s." = 1),
                     name = "vs theory") +
  facet_wrap(~ relationship, scales = "free_x", ncol = 2) +
  labs(
    x = "SMA \u03b2 (95% CI)", y = NULL,
    title = "Forest plot: SMA scaling exponents vs theoretical predictions",
    subtitle = "Dashed lines = WBE (2/3) and MST (8/3) predictions; filled = p < 0.05"
  )
```

---

# Caveats

1. **SMA vs RMA:** `smatr::sma()` implements SMA (Standardised Major Axis).
   True RMA (Reduced Major Axis) assumes equal error variances in both
   variables; SMA does not.  For allometric work on log-transformed data, SMA
   is the standard choice (Warton et al. 2006).

2. **OLS vs SMA:** OLS minimises only y-residuals and underestimates the slope
   when both variables carry error — typically by a factor of $1/\sqrt{R^2}$
   relative to the true slope.  For low-R² allometries the difference is
   substantial.

3. **Source heterogeneity:** Mixing sources with different measurement
   protocols, size ranges, and biomes inflates variance and may shift the
   pooled SMA slope.

4. **NEON AGB is model-derived:** AGB ~ DBH for NEON reflects the Chojnacky
   equations, not independent data.

5. **Tallo subsampled:** 80,000 of 498,838 records used for fitting; seed 42.

6. **Pooled test p-values:** With large N, even trivial deviations from
   theoretical predictions will be statistically significant.

---

# References

- Warton, D.I., Wright, I.J., Falster, D.S. & Westoby, M. (2006). Bivariate
  line-fitting methods for allometry. *Biological Reviews* 81, 259–291.
- West, G.B., Brown, J.H. & Enquist, B.J. (1999). A general model for the
  structure and allometry of plant vascular systems. *Nature* 400, 664–667.
- Enquist, B.J. et al. (2009). Biological scaling: Does the exception prove the
  rule? *Nature* 445, E9–E10.
- Chave, J. et al. (2005). Tree allometry and improved estimation of carbon
  stocks and balance in tropical forests. *Oecologia* 145, 87–99.
- Chojnacky, D.C., Heath, L.S. & Jenkins, J.C. (2014). Updated generalized
  biomass equations for North American tree species. *Forestry* 87, 129–151.

---

# Session info

```{r session}
sessionInfo()
```
