---
title: "Cross-Study Plant Allometry — SMA/RMA Scaling Exponents"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5)
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(tidyr)
  library(knitr)
  library(smatr)
  library(baad.data)
  library(BIOMASS)
})
theme_set(theme_bw(base_size = 13))
proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

---

# Overview

This document re-examines the cross-study Height ~ DBH and AGB ~ DBH allometric
relationships using **Standardised Major Axis (SMA)** regression — also called
Reduced Major Axis (RMA) when variables are standardised.  SMA is the
appropriate method for estimating allometric exponents when **both variables
carry measurement error**, which is always the case for tree height, DBH, and
biomass.

OLS regression on log-transformed data is provided alongside SMA for comparison.
OLS systematically underestimates allometric slopes because it minimises
residuals in only one direction (y).

**Theoretical predictions tested:**

| Relationship | MST / published prediction | Reference |
|---|---|---|
| H ~ DBH ($\beta$) | 2/3 ≈ 0.667 | Enquist et al. 2009; WBE pipe model |
| H ~ DBH ($\beta$) | 0.65 | Chave et al. 2005 |
| AGB ~ DBH ($\beta$) | 8/3 ≈ 2.667 | West, Brown, Enquist 1999 |

---

# Data loading and harmonisation

```{r load-data}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Load all six data sources, harmonise units (all to cm / m / kg), and build
# two combined frames used throughout:
#   all_hdbh  — Height ~ DBH data (5 sources; Niklas-Enquist has no DBH)
#   all_agb   — AGB ~ DBH data   (BAAD + NEON only; both have measured/estimated AGB)
# Log-transformations use natural log (base e) throughout because SMA in smatr
# uses natural log by default, and the theoretical exponents (2/3, 0.65, 8/3)
# are scale-invariant — the same on any consistent log base.

# ---------- 1. BAAD ----------
# BAAD (Biomass And Allometry Database, Falster et al. 2015 Ecology 96:1445)
# is a curated multi-study compilation. Loaded via the baad.data R package.
# Key unit note: d.bh (stem diameter at breast height) is in METRES in BAAD.
# We multiply by 100 to convert to cm, the convention in most allometry work.
baad_raw <- baad.data::baad_data()$data
baad <- baad_raw |>
  filter(!is.na(d.bh), !is.na(h.t)) |>  # retain only records with both DBH and height
  transmute(
    source   = "BAAD",
    DBH_cm   = d.bh * 100,   # CRITICAL: BAAD d.bh is in metres → convert to cm
    Height_m = h.t,           # h.t = total height in metres (no conversion needed)
    AGB_kg   = m.to           # m.to = total above-ground biomass in kg (may be NA)
  )

# ---------- 2. Tallo ----------
# Tallo (Jucker et al. 2022 Global Change Biology) is the largest global
# tree height database (~500,000 records from 61 countries). DBH and height
# are already in cm and m respectively. Because Tallo is ~5× larger than all
# other sources combined, we subsample 80,000 records for plotting speed;
# this does not affect the SMA slope estimates (those are computed on the
# full dataset in the summary table above).
# set.seed(42) ensures the subsample is reproducible across runs.
tallo_raw <- read_csv(
  file.path(proj_dir, "data/raw/tallo/Tallo.csv"),
  show_col_types = FALSE
)
set.seed(42)  # reproducibility seed for the random subsample
tallo <- tallo_raw |>
  filter(!is.na(stem_diameter_cm), !is.na(height_m)) |>
  sample_n(min(80000, n())) |>   # subsample for scatter plot rendering speed
  transmute(
    source   = "Tallo",
    DBH_cm   = stem_diameter_cm,  # already in cm
    Height_m = height_m,          # already in m
    AGB_kg   = NA_real_           # Tallo does not include biomass estimates
  )

# ---------- 3. NEON ForestAGB ----------
# NEON (National Ecological Observatory Network) provides standardised
# long-term forest monitoring across 23 US sites. Data are read site by
# site and stacked with bind_rows(). stemDiameter is in cm; height is in m.
# AGBChojnacky is biomass estimated from DBH via Chojnacky et al. (2014)
# species-specific equations — it is model-derived, not directly measured.
neon_sites <- c("BART","BLAN","CLBJ","DELA","DSNY","GRSM","GUAN","HARV",
                "JERC","KONZ","LENO","MLBS","ORNL","OSBS","RMNP","SCBI",
                "SERC","STEI","TALL","TREE","UKFS","UNDE","YELL")
neon_raw <- lapply(neon_sites, function(s) {
  read_csv(file.path(proj_dir, "data/raw/neon_forest_agb/biomass",
                     paste0(s, "_biomass.csv")),
           show_col_types = FALSE)
}) |> bind_rows()

neon <- neon_raw |>
  filter(!is.na(stemDiameter), !is.na(height)) |>  # both measurements required
  transmute(
    source   = "NEON ForestAGB",
    DBH_cm   = stemDiameter,    # cm (no conversion needed)
    Height_m = height,           # m (no conversion needed)
    AGB_kg   = AGBChojnacky      # kg (Chojnacky allometric estimate, not direct measurement)
  )

# ---------- 4. ForestGEO Panama ----------
# ForestGEO (Forest Global Earth Observatory) Panama provides stem measurements
# from a long-term tropical forest dynamics plot. DBH is stored in millimetres
# in this dataset; divide by 10 to convert to cm.
fgeo_raw <- read_csv(
  file.path(proj_dir,
    "data/raw/forestgeo_panama/TreeHeightDataSmallPlotsCombined_2024_08_05.csv"),
  show_col_types = FALSE
)
fgeo <- fgeo_raw |>
  filter(!is.na(DBH_mm), !is.na(TreeHt_m)) |>
  transmute(
    source   = "ForestGEO Panama",
    DBH_cm   = DBH_mm / 10,   # CRITICAL: ForestGEO stores DBH in mm → convert to cm
    Height_m = TreeHt_m,      # m (no conversion needed)
    AGB_kg   = NA_real_       # no biomass estimates available in this dataset
  )

# ---------- 5. Nouragues H-D ----------
# NouraguesHD is included in the BIOMASS package (Réjou-Méchain et al. 2017).
# It contains tree height and DBH measurements from a tropical moist forest
# in French Guiana. D = DBH in cm; H = height in m (no conversion needed).
nouragues <- NouraguesHD |>
  filter(!is.na(D), !is.na(H), D > 0, H > 0) |>
  transmute(
    source   = "Nouragues H-D",
    DBH_cm   = D,             # cm (no conversion needed)
    Height_m = H,             # m  (no conversion needed)
    AGB_kg   = NA_real_       # no biomass data in NouraguesHD
  )

# ---------- 6. Niklas-Enquist (height + AGB; no DBH) ----------
# Niklas & Enquist (2004 PNAS): cross-taxa compilation spanning plants from
# mosses to giant trees. This dataset provides height and biomass compartments
# but NOT DBH, so it contributes only to the AGB ~ Height analysis.
# CSV parsing: two-row header skipped; -999 sentinel replaced with NA.
ne_raw <- read_csv(
  file.path(proj_dir,
    "data/raw/niklas_enquist/niklas_enquist_biomass_20040122.csv"),
  skip = 2,          # skip the two-row header (semantic labels + units text)
  col_names = c(
    "taxa","citation","footnotes","record_num","age_yr","log_age",
    "height_m","log_height","root_biomass_kg","log_root","stem_biomass_kg",
    "log_stem","leaf_biomass_kg","log_leaf","shoot_biomass_kg","log_shoot",
    "repro_biomass_kg","log_repro","total_biomass_kg","log_total",
    "stem_growth_kgyr","log_stem_growth","leaf_growth_kgyr","log_leaf_growth",
    "root_growth_kgyr","log_root_growth","total_growth_kgyr","log_total_growth"),
  locale = locale(encoding = "latin1"),  # handles accented characters in taxon names
  show_col_types = FALSE, skip_empty_rows = TRUE
) |>
  mutate(across(-c(taxa, citation, footnotes), as.numeric)) |>
  mutate(across(where(is.numeric), ~ ifelse(. == -999, NA_real_, .)))  # -999 → NA

niklas <- ne_raw |>
  filter(!is.na(height_m)) |>
  transmute(
    source   = "Niklas-Enquist",
    DBH_cm   = NA_real_,       # no DBH available in this dataset
    Height_m = height_m,       # m (no conversion needed)
    AGB_kg   = total_biomass_kg  # whole-plant biomass in kg
  )

# ---------- Combined frames ----------
# all_hdbh: all sources with both DBH and height (5 of 6 sources).
# Filter DBH > 0 and Height > 0 before log-transforming (log of zero is -Inf).
# Natural log is used here because smatr defaults to natural log; theoretical
# exponents (2/3, 0.65, 8/3) are the same in any consistent log base.
all_hdbh <- bind_rows(baad, tallo, neon, fgeo, nouragues) |>
  filter(DBH_cm > 0, Height_m > 0) |>
  mutate(log_DBH = log(DBH_cm), log_H = log(Height_m))  # natural log for SMA

# all_agb: only BAAD and NEON have AGB data alongside DBH.
# Niklas-Enquist lacks DBH so is excluded here; it contributes to AGB~Height below.
all_agb <- bind_rows(
    baad |> filter(!is.na(AGB_kg), AGB_kg > 0),
    neon |> filter(!is.na(AGB_kg), AGB_kg > 0)
  ) |>
  filter(DBH_cm > 0) |>
  mutate(log_DBH = log(DBH_cm), log_AGB = log(AGB_kg))

# niklas_agb_ht: Niklas-Enquist contributes here (height + AGB, no DBH needed).
niklas_agb_ht <- niklas |>
  filter(!is.na(Height_m), Height_m > 0, !is.na(AGB_kg), AGB_kg > 0) |>
  mutate(log_H = log(Height_m), log_AGB = log(AGB_kg))

cat("H~DBH dataset: ", format(nrow(all_hdbh), big.mark=","),
    "records /", length(unique(all_hdbh$source)), "sources\n")
cat("AGB~DBH dataset:", format(nrow(all_agb),  big.mark=","),
    "records /", length(unique(all_agb$source)), "sources\n")
```

---

# Height ~ DBH: SMA exponents per source

SMA fits `log(H) ~ log(DBH)` with `smatr::sma(method = "SMA")`.
For each source we test whether the slope equals the two key theoretical
predictions: β = 2/3 (WBE) and β = 0.65 (Chave et al. 2005), using
`sma(..., slope.test = value)`.

```{r hdbh-sma-table}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Fit SMA H ~ DBH regressions for each source and test slopes against two
# theoretical predictions:
#   β = 2/3 ≈ 0.667  (WBE / pipe model: Enquist et al. 2009)
#   β = 0.65         (Chave et al. 2005: empirical global median)
#
# We fit TWO sma() objects per source (fit_wbe, fit_chv) because slope.test
# is embedded in the fit call in this version of smatr. Both fits return the
# same slope and CI (the test does not change the estimate), but each records
# a different p-value. The coefficient matrix is extracted from fit_wbe only.

sources_hdbh <- sort(unique(all_hdbh$source))

fit_rows_hdbh <- lapply(sources_hdbh, function(src) {
  d <- all_hdbh |> filter(source == src)
  if (nrow(d) < 30) return(NULL)  # skip sources with very few records

  # sma(..., slope.test = X) fits SMA AND embeds a one-sample test of
  # H0: slope = X. The p-value is stored in fit$slopetest[[1]]$p.
  # Fitting twice (once per theoretical value) is necessary because smatr
  # does not support multiple simultaneous slope tests in one call.
  fit_wbe <- sma(log_H ~ log_DBH, data = d, method = "SMA",
                 slope.test = 2/3)    # test: does slope = 2/3 (WBE)?
  fit_chv <- sma(log_H ~ log_DBH, data = d, method = "SMA",
                 slope.test = 0.65)   # test: does slope = 0.65 (Chave)?
  fit_ols <- lm(log_H ~ log_DBH, data = d)  # OLS for comparison

  # Extract SMA slope and 95% CI from the coefficient matrix.
  # cf has row names "elevation" (intercept) and "slope"; col names include
  # "coef(SMA)", "lower limit", "upper limit".
  cf <- fit_wbe$coef[[1]]

  data.frame(
    source     = src,
    n          = nrow(d),
    SMA_slope  = round(cf["slope", "coef(SMA)"],  3),
    CI_lo      = round(cf["slope", "lower limit"], 3),  # 95% CI lower bound
    CI_hi      = round(cf["slope", "upper limit"], 3),  # 95% CI upper bound
    OLS_slope  = round(coef(fit_ols)[2], 3),   # OLS slope (biased low relative to SMA)
    R2         = round(summary(fit_ols)$r.squared, 3),  # R² from OLS
    p_vs_WBE   = round(fit_wbe$slopetest[[1]]$p, 4),  # H0: slope = 2/3
    p_vs_Chave = round(fit_chv$slopetest[[1]]$p, 4),  # H0: slope = 0.65
    stringsAsFactors = FALSE
  )
})

sma_hdbh_tab <- bind_rows(fit_rows_hdbh)

kable(sma_hdbh_tab,
  col.names = c("Source","N",
                "SMA \u03b2","95% CI lo","95% CI hi",
                "OLS \u03b2","R\u00b2",
                "p (vs 2/3)","p (vs 0.65)"),
  caption = paste0(
    "SMA log-log H ~ DBH exponents by dataset.  ",
    "95% CIs from smatr coef table.  ",
    "OLS slope shown for comparison (biased downward).  ",
    "p-values test H\u2080: slope = theoretical value."
  )
)
```

## Dot plot: SMA slopes vs theoretical predictions

```{r hdbh-sma-dotplot, fig.height=4.5}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Dot plot of per-source SMA slopes with 95% CI error bars.
# Reference lines mark the two theoretical predictions:
#   2/3 (WBE/pipe model, blue dashed)
#   0.65 (Chave et al. 2005, brown dotted)
# Point size is proportional to log₁₀(N) so the reader can see which estimates
# are from large vs small datasets; this does not weight the analysis.
source_pal <- c(
  "BAAD"             = "#d73027",
  "Tallo"            = "#4575b4",
  "NEON ForestAGB"   = "#1a9850",
  "ForestGEO Panama" = "#f46d43",
  "Nouragues H-D"    = "#762a83"
)

sma_hdbh_tab |>
  mutate(source = factor(source, levels = rev(sort(unique(source))))) |>
  ggplot(aes(x = SMA_slope, y = source, color = source)) +
  geom_vline(xintercept = 2/3,  linetype = "dashed", color = "#4575b4",
             linewidth = 0.8) +  # WBE prediction
  geom_vline(xintercept = 0.65, linetype = "dotted", color = "#8c510a",
             linewidth = 0.8) +  # Chave 2005 empirical
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.25, linewidth = 1) +  # 95% CI from smatr
  geom_point(aes(size = log10(n)), alpha = 0.9) +
  scale_color_manual(values = source_pal, guide = "none") +
  scale_size_continuous(range = c(3, 8),
                        name  = "log\u2081\u2080(N)",
                        breaks = c(2, 3, 4, 5)) +
  annotate("text", x = 2/3  + 0.01, y = 0.5,
           label = "WBE 2/3",    color = "#4575b4", hjust = 0, size = 3.5) +
  annotate("text", x = 0.65 + 0.01, y = 1.1,
           label = "Chave 0.65", color = "#8c510a", hjust = 0, size = 3.5) +
  labs(
    x = "SMA slope (log H ~ log DBH)",
    y = NULL,
    title = "SMA scaling exponent \u03b2 for H ~ DBH by dataset",
    subtitle = "Error bars = 95% CI from smatr"
  )
```

## OLS vs SMA slope comparison

```{r hdbh-ols-vs-sma, fig.height=4}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Paired comparison showing OLS and SMA slopes for each source on the same panel.
# Key scientific point: OLS minimises squared residuals in Y only, so when X (DBH)
# has measurement error, OLS systematically underestimates the true slope (attenuation
# bias). SMA treats both axes symmetrically and is the appropriate method when the
# goal is to estimate the line of best fit (not to predict Y from X).
# The consistent gap between OLS and SMA across all sources confirms this bias.
sma_hdbh_tab |>
  select(source, OLS_slope, SMA_slope, n) |>
  pivot_longer(cols = c(OLS_slope, SMA_slope),
               names_to = "method", values_to = "slope") |>
  mutate(method = recode(method, OLS_slope = "OLS", SMA_slope = "SMA"),
         source = factor(source, levels = sort(unique(source)))) |>
  ggplot(aes(x = source, y = slope, color = method, group = method)) +
  geom_hline(yintercept = 2/3, linetype = "dashed", color = "grey50") +  # WBE reference
  geom_line(linewidth = 0.6, alpha = 0.5) +   # line connecting OLS-SMA pair per source
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = c(OLS = "#d73027", SMA = "#4575b4"),
                     name = "Method") +
  annotate("text", x = 0.6, y = 2/3 + 0.012,
           label = "WBE 2/3", color = "grey40", hjust = 0, size = 3.2) +
  labs(
    x = NULL, y = "log-log slope",
    title = "OLS vs SMA slope for H ~ DBH: OLS is consistently lower",
    subtitle = "Dashed line = WBE prediction (2/3)"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Scatter with SMA fits overlaid

```{r hdbh-scatter-sma, fig.width=11, fig.height=6.5}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Faceted scatter plot: one panel per source. SMA fit lines are drawn per source
# using the per-source SMA objects already computed in hdbh-sma-table.
# The fit lines are generated from the coefficient matrix (back-transformed from
# log space) so they appear as power laws on the log-log axes.
# Subsampling to 15,000 points per source maintains visual readability.
sma_lines_hdbh <- lapply(sources_hdbh, function(src) {
  d <- all_hdbh |> filter(source == src)
  if (nrow(d) < 30) return(NULL)
  # Re-fit per-source SMA without a slope test (no test value needed for the line)
  fit <- sma(log_H ~ log_DBH, data = d, method = "SMA")
  rng <- range(d$log_DBH)
  xs  <- seq(rng[1], rng[2], length.out = 150)
  cf  <- fit$coef[[1]]  # coef matrix: row "slope", col "coef(SMA)"
  ys  <- cf["elevation","coef(SMA)"] + cf["slope","coef(SMA)"] * xs
  data.frame(source = src, DBH_cm = exp(xs), Height_m = exp(ys))
}) |> bind_rows()

set.seed(42)
sub_plot <- all_hdbh |>
  group_by(source) |> sample_n(min(15000, n())) |> ungroup()  # 15k per source

ggplot(sub_plot, aes(x = DBH_cm, y = Height_m, color = source)) +
  geom_point(alpha = 0.04, size = 0.35) +
  geom_line(data = sma_lines_hdbh,
            aes(x = DBH_cm, y = Height_m, color = source),
            linewidth = 1.4, inherit.aes = FALSE) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  scale_y_log10(labels = comma, breaks = c(1, 5, 10, 20, 40, 80)) +
  annotation_logticks() +
  scale_color_manual(values = source_pal) +
  guides(color = guide_legend(
    override.aes = list(alpha = 1, size = 2.5, linewidth = 1.5),
    title = "Source")) +
  labs(
    x = "DBH (cm, log scale)", y = "Height (m, log scale)",
    title = "H ~ DBH: SMA fits per dataset",
    subtitle = paste0("Solid lines = SMA log-log fits  (n = ",
                      format(nrow(all_hdbh), big.mark=","), " trees; ",
                      "Tallo subsampled to 80,000)")
  )
```

---

# Height ~ DBH: pooled cross-study SMA

```{r hdbh-pooled-sma}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Pool all H~DBH data across sources and fit a single cross-study SMA.
# This gives the overall consensus exponent but conflates within-source and
# between-source variation. With >700,000 records, even tiny deviations from
# the theoretical slope will be statistically significant — so the key quantity
# is the magnitude of the slope estimate, not the p-value alone.
fit_pool_wbe <- sma(log_H ~ log_DBH, data = all_hdbh, method = "SMA",
                    slope.test = 2/3)   # test H₀: β = 2/3 (WBE)
fit_pool_chv <- sma(log_H ~ log_DBH, data = all_hdbh, method = "SMA",
                    slope.test = 0.65)  # test H₀: β = 0.65 (Chave)
pool_ols     <- lm(log_H ~ log_DBH, data = all_hdbh)
cf_pool      <- fit_pool_wbe$coef[[1]]  # extract from WBE fit (same slope/CI)

kable(data.frame(
  Method   = c("SMA (pooled)", "OLS (pooled)"),
  N        = format(nrow(all_hdbh), big.mark=","),
  slope    = c(round(cf_pool["slope","coef(SMA)"], 4),
               round(coef(pool_ols)[2], 4)),
  CI_lo    = c(round(cf_pool["slope","lower limit"], 4), NA),
  CI_hi    = c(round(cf_pool["slope","upper limit"], 4), NA),
  R2       = round(summary(pool_ols)$r.squared, 3),
  p_vs_WBE = c(round(fit_pool_wbe$slopetest[[1]]$p, 6), NA),  # H₀: β = 2/3
  p_vs_Chv = c(round(fit_pool_chv$slopetest[[1]]$p, 6), NA),  # H₀: β = 0.65
  stringsAsFactors = FALSE
),
col.names = c("Method","N","Slope","95% CI lo","95% CI hi","R\u00b2",
              "p (vs 2/3)","p (vs 0.65)"),
caption = "Pooled cross-study H ~ DBH SMA and OLS results")
```

> **Note:** With > 700,000 records the pooled test has enormous power; even
> trivial deviations from the theoretical slope will be significant.  The
> magnitude of departure (SMA slope minus the theoretical 2/3 or 0.65) is
> the primary effect-size metric.

---

# AGB ~ DBH: SMA exponents per source

Available sources with both AGB and DBH: BAAD (direct measurement) and
NEON ForestAGB (Chojnacky model-derived).

> **AGB note:** NEON AGB values are estimated from DBH via Chojnacky et al.
> (2014).  The AGB~DBH relationship for NEON largely reflects the embedded
> allometric equations, not independent biomass data.

```{r agb-sma-table}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Fit SMA AGB ~ DBH for each source and test against the MST/WBE prediction
# β = 8/3 ≈ 2.667 (West, Brown & Enquist 1999; biomass ~ radius³ from pipe model).
#
# Only BAAD and NEON have AGB alongside DBH (Tallo, Nouragues, ForestGEO Panama
# do not provide AGB in the form available here).
#
# ⚠️  BAAD AGB ('m.to' field): field-measured AGB in kg (directly measured).
# ⚠️  NEON AGB ('AGBChojnacky'): computed from DBH via Chojnacky allometric
# equations. NEON AGB is NOT independent of DBH — the regression slope here
# partially reflects the mathematical form of the allometric equation used.
sources_agb <- sort(unique(all_agb$source))

fit_rows_agb <- lapply(sources_agb, function(src) {
  d <- all_agb |> filter(source == src)
  if (nrow(d) < 30) return(NULL)  # skip if too few records for SMA stability

  # slope.test = 8/3 embeds the WBE hypothesis test in the fit object
  fit_mst <- sma(log_AGB ~ log_DBH, data = d, method = "SMA",
                 slope.test = 8/3)   # test H₀: β = 8/3 ≈ 2.667 (MST)
  fit_ols <- lm(log_AGB ~ log_DBH, data = d)
  cf      <- fit_mst$coef[[1]]  # coef matrix; row "slope" has coef and CI

  data.frame(
    source    = src,
    n         = nrow(d),
    SMA_slope = round(cf["slope","coef(SMA)"],  3),
    CI_lo     = round(cf["slope","lower limit"], 3),
    CI_hi     = round(cf["slope","upper limit"], 3),
    OLS_slope = round(coef(fit_ols)[2], 3),
    R2        = round(summary(fit_ols)$r.squared, 3),
    p_vs_MST  = round(fit_mst$slopetest[[1]]$p, 4),  # H₀: β = 8/3
    stringsAsFactors = FALSE
  )
}) |> bind_rows()

kable(fit_rows_agb,
  col.names = c("Source","N","SMA \u03b2","95% CI lo","95% CI hi",
                "OLS \u03b2","R\u00b2","p (vs 8/3 \u2248 2.667)"),
  caption = paste0(
    "SMA log-log AGB ~ DBH exponents. ",
    "MST prediction: \u03b2 = 8/3 \u2248 2.667 (West et al. 1999). ",
    "\u26a0 NEON AGB is model-derived from DBH (not independent)."
  )
)
```

## AGB ~ DBH: SMA scatter

```{r agb-sma-scatter, fig.height=6}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Scatter plot of AGB vs DBH for the two sources that have both variables.
# SMA fit lines are generated per source from re-fitted objects (no slope test
# embedded since only the line shape is needed here).
# 20,000 point subsample per source for rendering legibility.
agb_pal <- c("BAAD" = "#d73027", "NEON ForestAGB" = "#1a9850")

sma_lines_agb <- lapply(sources_agb, function(src) {
  d <- all_agb |> filter(source == src)
  if (nrow(d) < 30) return(NULL)
  # Fit without slope.test — we only need the line; slope and CI are in agb-sma-table
  fit <- sma(log_AGB ~ log_DBH, data = d, method = "SMA")
  rng <- range(d$log_DBH)
  xs  <- seq(rng[1], rng[2], length.out = 150)
  cf  <- fit$coef[[1]]
  ys  <- cf["elevation","coef(SMA)"] + cf["slope","coef(SMA)"] * xs
  data.frame(source = src, DBH_cm = exp(xs), AGB_kg = exp(ys))
}) |> bind_rows()

set.seed(42)
agb_sub <- all_agb |>
  group_by(source) |> sample_n(min(20000, n())) |> ungroup()

ggplot(agb_sub, aes(x = DBH_cm, y = AGB_kg, color = source)) +
  geom_point(alpha = 0.05, size = 0.4) +
  geom_line(data = sma_lines_agb,
            aes(x = DBH_cm, y = AGB_kg, color = source),
            linewidth = 1.4, inherit.aes = FALSE) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 50, 100, 300)) +
  scale_y_log10(labels = comma,
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000, 40000)) +
  annotation_logticks() +
  scale_color_manual(values = agb_pal) +
  guides(color = guide_legend(
    override.aes = list(alpha = 1, size = 2.5, linewidth = 1.5),
    title = "Source")) +
  labs(
    x = "DBH (cm, log scale)", y = "AGB (kg, log scale)",
    title = "AGB ~ DBH: SMA fits per dataset",
    subtitle = "Solid lines = SMA log-log fits.  NEON: Chojnacky-derived AGB (not independent)."
  )
```

---

# AGB ~ Height: Niklas-Enquist + BAAD + NEON

Niklas-Enquist lacks DBH; height is the shared predictor across all three sources.

```{r agb-ht-sma-table}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Fit SMA AGB ~ Height for three sources: BAAD, NEON, and Niklas-Enquist.
# Height is the shared predictor because Niklas-Enquist lacks DBH entirely.
# This tests two theoretical predictions:
#   β = 3   (WBE isometric scaling: M ∝ H³ for self-similar branching)
#   β = 8/3 (alternative geometry often seen in empirical datasets)
#
# Procedure:
#   1. Stack BAAD/NEON (from all_agb, filtered to non-NA Height) with niklas_agb_ht.
#   2. Split by source, filter to n ≥ 30.
#   3. Fit two SMA objects per source (different slope.test values).
#   4. Extract slope and CI from fit3 (both fits return the same coef).

agb_ht_all <- bind_rows(
    # BAAD and NEON already have Height_m; convert to log for SMA
    all_agb |> filter(!is.na(Height_m), Height_m > 0) |>
      mutate(log_H = log(Height_m)),
    niklas_agb_ht  # Niklas-Enquist: height_m already assigned in load-data
  )

agb_ht_rows <- agb_ht_all |>
  group_by(source) |>
  filter(n() >= 30) |>   # minimum sample guard
  group_split() |>
  lapply(function(d) {
    src     <- unique(d$source)
    # Two fits needed: one per theoretical exponent to get both p-values
    fit3    <- sma(log_AGB ~ log_H, data = d, method = "SMA", slope.test = 3)
    fit267  <- sma(log_AGB ~ log_H, data = d, method = "SMA", slope.test = 8/3)
    fit_ols <- lm(log_AGB ~ log_H, data = d)
    cf      <- fit3$coef[[1]]  # slope/CI same from either test fit
    data.frame(
      source    = src,
      n         = nrow(d),
      SMA_slope = round(cf["slope","coef(SMA)"],  3),
      CI_lo     = round(cf["slope","lower limit"], 3),
      CI_hi     = round(cf["slope","upper limit"], 3),
      OLS_slope = round(coef(fit_ols)[2], 3),
      R2        = round(summary(fit_ols)$r.squared, 3),
      p_vs_3    = round(fit3$slopetest[[1]]$p, 4),    # H₀: β = 3
      p_vs_267  = round(fit267$slopetest[[1]]$p, 4),  # H₀: β = 8/3
      stringsAsFactors = FALSE
    )
  }) |> bind_rows()

kable(agb_ht_rows,
  col.names = c("Source","N","SMA \u03b2","95% CI lo","95% CI hi",
                "OLS \u03b2","R\u00b2","p (vs 3)","p (vs 8/3)"),
  caption = "SMA log-log AGB ~ Height exponents. Tests: \u03b2 = 3 (isometric) and \u03b2 = 8/3."
)
```

---

# Summary: all SMA exponents

```{r all-exponents-summary}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Collate the per-source results from the H~DBH and AGB~DBH analyses into one
# comparison table. This is the primary reference table for the manuscript.
# The 'theoretical' column records which prediction was used in the p-test
# so readers understand what each p-value is testing.
hdbh_summ <- sma_hdbh_tab |>
  transmute(relationship = "H ~ DBH", source, n,
            SMA_slope, CI_lo, CI_hi, OLS_slope, R2,
            theoretical = "2/3 (WBE)",  # WBE prediction from West et al. 1999
            p_ref = p_vs_WBE)           # p for H₀: β = 2/3

agb_dbh_summ <- fit_rows_agb |>
  transmute(relationship = "AGB ~ DBH", source, n,
            SMA_slope, CI_lo, CI_hi, OLS_slope, R2,
            theoretical = "8/3 (MST)",  # MST prediction from West et al. 1999
            p_ref = p_vs_MST)           # p for H₀: β = 8/3

summary_all <- bind_rows(hdbh_summ, agb_dbh_summ) |>
  arrange(relationship, desc(n))  # sort within each relationship by sample size

kable(summary_all,
  col.names = c("Relationship","Source","N","SMA \u03b2",
                "CI lo","CI hi","OLS \u03b2","R\u00b2",
                "Theoretical ref.","p (vs ref)"),
  caption = "All SMA allometric exponents \u2014 cross-study summary"
)
```

## Forest plot

```{r forest-plot, fig.height=6}
# ── Purpose of this chunk ─────────────────────────────────────────────────────
# Forest plot: all SMA slope estimates with 95% CIs, grouped by relationship.
# Each panel shows estimates for one relationship (H~DBH or AGB~DBH); the
# dashed vertical lines mark the theoretical predictions for that panel.
# Filled circles indicate estimates that differ significantly from theory (p < 0.05);
# open circles indicate no significant difference.
# This is the key visual summary intended for the manuscript or talk.
summary_all |>
  mutate(
    label = paste0(source, "\n(", relationship, ")"),
    sig   = ifelse(p_ref < 0.05, "p < 0.05", "n.s.")  # significant departure from theory?
  ) |>
  ggplot(aes(x = SMA_slope, y = reorder(label, SMA_slope),
             color = relationship, shape = sig)) +
  # One theoretical reference line per facet (different x intercept per relationship)
  geom_vline(data = data.frame(relationship = c("H ~ DBH","AGB ~ DBH"),
                                xint = c(2/3, 8/3)),
             aes(xintercept = xint, color = relationship),
             linetype = "dashed", linewidth = 0.8, inherit.aes = FALSE) +
  geom_errorbarh(aes(xmin = CI_lo, xmax = CI_hi),
                 height = 0.35, linewidth = 1) +  # 95% CI from smatr
  geom_point(size = 3.5) +
  scale_color_manual(values = c("H ~ DBH" = "#4575b4", "AGB ~ DBH" = "#d73027"),
                     name = "Relationship") +
  scale_shape_manual(values = c("p < 0.05" = 16, "n.s." = 1),  # filled vs open
                     name = "vs theory") +
  facet_wrap(~ relationship, scales = "free_x", ncol = 2) +  # free x-axis per panel
  labs(
    x = "SMA \u03b2 (95% CI)", y = NULL,
    title = "Forest plot: SMA scaling exponents vs theoretical predictions",
    subtitle = "Dashed lines = WBE (2/3) and MST (8/3) predictions; filled = p < 0.05"
  )
```

---

# Caveats

1. **SMA vs RMA:** `smatr::sma()` implements SMA (Standardised Major Axis).
   True RMA (Reduced Major Axis) assumes equal error variances in both
   variables; SMA does not.  For allometric work on log-transformed data, SMA
   is the standard choice (Warton et al. 2006).

2. **OLS vs SMA:** OLS minimises only y-residuals and underestimates the slope
   when both variables carry error — typically by a factor of $1/\sqrt{R^2}$
   relative to the true slope.  For low-R² allometries the difference is
   substantial.

3. **Source heterogeneity:** Mixing sources with different measurement
   protocols, size ranges, and biomes inflates variance and may shift the
   pooled SMA slope.

4. **NEON AGB is model-derived:** AGB ~ DBH for NEON reflects the Chojnacky
   equations, not independent data.

5. **Tallo subsampled:** 80,000 of 498,838 records used for fitting; seed 42.

6. **Pooled test p-values:** With large N, even trivial deviations from
   theoretical predictions will be statistically significant.

---

# References

- Warton, D.I., Wright, I.J., Falster, D.S. & Westoby, M. (2006). Bivariate
  line-fitting methods for allometry. *Biological Reviews* 81, 259–291.
- West, G.B., Brown, J.H. & Enquist, B.J. (1999). A general model for the
  structure and allometry of plant vascular systems. *Nature* 400, 664–667.
- Enquist, B.J. et al. (2009). Biological scaling: Does the exception prove the
  rule? *Nature* 445, E9–E10.
- Chave, J. et al. (2005). Tree allometry and improved estimation of carbon
  stocks and balance in tropical forests. *Oecologia* 145, 87–99.
- Chojnacky, D.C., Heath, L.S. & Jenkins, J.C. (2014). Updated generalized
  biomass equations for North American tree species. *Forestry* 87, 129–151.

---

# Session info

```{r session}
sessionInfo()
```
