---
title: "ForestGEO Panama — Tree Height Allometry, Small Plots 2011–2020"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
library(readr)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)
library(maps)

theme_set(theme_bw(base_size = 12))

raw_dir <- file.path(dirname(knitr::current_input(dir = TRUE)),
                     "data/raw/forestgeo_panama")

dat <- read_csv(
  file.path(raw_dir, "TreeHeightDataSmallPlotsCombined_2024_08_05.csv"),
  show_col_types = FALSE
)

# Unit-convert DBH for display (source is mm, all plots use mm)
dat <- dat |> mutate(DBH_cm = DBH_mm / 10)

# Load data dictionary
dd <- read_excel(
  file.path(raw_dir, "Data_dictionary_for_smallplots_allometry_2023_08_05.xlsx"),
  sheet = 1
)
```

---

# Overview

**ForestGEO Panama small-plots allometry** is a dataset of individual tree
height and diameter measurements collected across 24 small (mostly 1-ha) forest
plots in central Panama between 2011 and 2020. It was compiled as part of the
Smithsonian ForestGEO network to support allometric model development for
tropical forests.

> Ramos P, Villareal P, Muller-Landau HC (2024).
> *Tree height allometry data for 5,109 trees in twenty-four small ForestGEO
> plots in Panama, 2011–2020.*
> Smithsonian Figshare.
> doi: [10.25573/data.24954204.v1](https://doi.org/10.25573/data.24954204.v1)
> License: CC BY 4.0

> **⚠ Citation note:** The DOI and author list above are taken directly from
> the Figshare landing page and should be verified against the source before
> use in any manuscript.

---

# Top-level summary

```{r top-counts}
n_spp   <- length(unique(dat$LatinBinomial[!is.na(dat$LatinBinomial)]))
n_plots <- length(unique(dat$Plotname))

kable(data.frame(
  Metric = c(
    "Total individual trees",
    "Unique plots",
    "Unique species (LatinBinomial)",
    "Measurement years",
    "DBH range (display in cm; raw in mm)",
    "Height range (m)",
    "Height-of-measurement (HOM) range (m)",
    "Global lat/lon coverage",
    "AGB estimates in source file",
    "Data license",
    "Source"
  ),
  Value = c(
    format(nrow(dat), big.mark = ","),
    n_plots,
    n_spp,
    paste0(min(dat$YYYY), "\u20132020"),
    paste0(round(min(dat$DBH_cm), 1), "\u2013",
           round(max(dat$DBH_cm), 1), " cm"),
    paste0(round(min(dat$TreeHt_m, na.rm = TRUE), 1), "\u2013",
           round(max(dat$TreeHt_m, na.rm = TRUE), 1), " m"),
    paste0(round(min(dat$HOM_m), 1), "\u2013",
           round(max(dat$HOM_m), 1),
           " m  (non-standard; elevated above buttresses)"),
    paste0("None \u2014 plot-local Cartesian coords only (PX_m, PY_m). ",
           "Plots in central Panama."),
    "None \u2014 raw diameters and heights only; no AGB in source file",
    "CC BY 4.0",
    "Smithsonian Figshare doi:10.25573/data.24954204.v1"
  ),
  stringsAsFactors = FALSE
), caption = "ForestGEO Panama small-plots top-level summary")
```

---

# Data dictionary

```{r data-dict}
kable(dd, caption = "Variable definitions (from source xlsx data dictionary)")
```

> **Note:** Column `B` appears in the data file (25 columns total) but has no
> entry in the 24-row data dictionary.  All 5,109 values are `FALSE`.  It may
> represent a broken-below-breast-height flag analogous to `Q` (broken *above*
> 1.3 m); the exact definition is undocumented in the deposited files.

---

# Geographic context

```{r geo-map, fig.width=10, fig.height=4.5}
# No global lat/lon in the dataset — plots use local Cartesian coordinates.
# We show the approximate study region in Panama.
world <- map_data("world")

panama_pt <- data.frame(
  label = "24 ForestGEO small plots\n(central Panama)",
  lon   = -79.8,
  lat   =   9.1
)

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat, group = group),
               fill = "grey88", color = "grey70", linewidth = 0.2) +
  geom_point(data = panama_pt,
             aes(x = lon, y = lat),
             color = "#e41a1c", size = 5) +
  geom_text(data = panama_pt,
            aes(x = lon, y = lat, label = label),
            hjust = -0.08, vjust = 0.5, size = 3.5, color = "#e41a1c") +
  coord_fixed(1.3, xlim = c(-120, -30), ylim = c(-15, 35)) +
  labs(
    title    = "ForestGEO Panama small plots \u2014 approximate study region",
    subtitle = paste0("\u26a0 No lat/lon in source file. ",
                      "Exact plot coordinates are in the ForestGEO site registry ",
                      "(https://forestgeo.si.edu/sites)."),
    x = NULL, y = NULL
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```

---

# Per-plot tree counts

```{r per-plot, fig.height=6}
plot_counts <- dat |>
  count(Plotname, name = "n_trees") |>
  arrange(desc(n_trees))

ggplot(plot_counts, aes(x = reorder(Plotname, n_trees), y = n_trees,
                        fill = n_trees)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n_trees), hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_fill_viridis_c(option = "D", direction = -1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "N trees", title = "Trees per plot (24 plots)")

kable(plot_counts, caption = "Trees per plot")
```

---

# Diameter distribution

```{r dbh-dist, fig.height=4}
dat |>
  ggplot(aes(x = DBH_cm)) +
  geom_histogram(bins = 60, fill = "#4393c3", color = "white", linewidth = 0.2) +
  scale_x_log10(labels = comma) +
  annotation_logticks(sides = "b") +
  scale_y_continuous(labels = comma) +
  labs(
    x = "DBH (cm, log scale)", y = "N trees",
    title = "DBH distribution \u2014 ForestGEO Panama (n = 5,109)",
    subtitle = "Raw data are in mm; converted to cm here (DBH_cm = DBH_mm / 10)"
  )
```

---

# Height distribution

```{r ht-dist, fig.height=4}
dat |>
  filter(!is.na(TreeHt_m)) |>
  ggplot(aes(x = TreeHt_m)) +
  geom_histogram(bins = 50, fill = "#2ca25f", color = "white", linewidth = 0.2) +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Tree height (m)", y = "N trees",
    title = paste0("Height distribution \u2014 ForestGEO Panama  (n = ",
                   sum(!is.na(dat$TreeHt_m)), ")")
  )
```

---

# Height ~ DBH scaling

```{r hdbh-scaling, fig.height=5}
sub <- dat |> filter(!is.na(TreeHt_m), DBH_cm > 0, TreeHt_m > 0)
lm_fit <- lm(log(TreeHt_m) ~ log(DBH_cm), data = sub)
beta <- round(coef(lm_fit)[2], 3)
r2   <- round(summary(lm_fit)$r.squared, 3)

sub |>
  ggplot(aes(x = DBH_cm, y = TreeHt_m)) +
  geom_point(alpha = 0.2, size = 0.7, color = "#2ca25f") +
  geom_smooth(method = "lm", color = "black", linewidth = 1.2, se = TRUE) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(
    x = "DBH (cm)", y = "Height (m)",
    title = "Height \u223c DBH  (OLS log-log regression)",
    subtitle = paste0("\u03b2 = ", beta, "   R\u00b2 = ", r2,
                      "   n = ", nrow(sub))
  )
```

> OLS log-log regression on all individuals with valid DBH and height.
> $\beta \approx 0.6$–$0.7$ is typical for tropical moist forest.
> ⚠ This regression pools all plots and species and should not be treated as a
> formal allometric model — use SMA regression and species-level grouping for
> inference.

---

# Height of measurement (HOM)

```{r hom-dist, fig.height=4}
pct_standard <- round(100 * mean(dat$HOM_m == 1.3), 1)

dat |>
  ggplot(aes(x = HOM_m)) +
  geom_histogram(binwidth = 0.1, fill = "#d6604d", color = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 1.3, linetype = "dashed", color = "black",
             linewidth = 0.9) +
  annotate("text", x = 1.42, y = Inf, label = "Standard 1.3 m BH",
           hjust = 0, vjust = 1.5, size = 3.5) +
  scale_y_continuous(labels = comma) +
  labs(
    x = "HOM (m)", y = "N trees",
    title = "Height of measurement (HOM) distribution",
    subtitle = paste0(pct_standard, "% at standard 1.3 m; remainder elevated above buttresses")
  )
```

**Key caveat:** Diameter is measured at a paint-marked point (`HOM_m`) ranging
from 0.3 to 4.8 m.  Trees with `HOM_m > 1.3` are typically buttressed
individuals measured above the buttress.  This **must be accounted for**
in any allometric comparison that assumes DBH at breast height (1.3 m).

---

# Quality and condition flags

```{r flags}
flags <- data.frame(
  Flag = c(
    "Q — stem broken above 1.3 m (Quebrado)",
    "L — leaning > 45\u00b0",
    "PP — tag missing at measurement date (Placa Perdida)",
    "R — resprout (stem broken below 1.3 m)",
    "B — undocumented flag (all FALSE; not in data dictionary)",
    "MultiStem — tree has multiple stems in dataset"
  ),
  N_TRUE = c(
    sum(dat$Q,         na.rm = TRUE),
    sum(dat$L,         na.rm = TRUE),
    sum(dat$PP,        na.rm = TRUE),
    sum(dat$R,         na.rm = TRUE),
    sum(dat$B,         na.rm = TRUE),
    sum(dat$MultiStem, na.rm = TRUE)
  ),
  Pct = paste0(round(100 * c(
    mean(dat$Q,         na.rm = TRUE),
    mean(dat$L,         na.rm = TRUE),
    mean(dat$PP,        na.rm = TRUE),
    mean(dat$R,         na.rm = TRUE),
    mean(dat$B,         na.rm = TRUE),
    mean(dat$MultiStem, na.rm = TRUE)
  ), 1), "%"),
  stringsAsFactors = FALSE
)
kable(flags, caption = "Quality and condition flag counts (n = 5,109 trees)")
```

---

# Crown intactness

```{r crown-intactness, fig.height=3.5}
dat |>
  filter(!is.na(Crown)) |>
  count(Crown) |>
  ggplot(aes(x = factor(Crown), y = n, fill = factor(Crown))) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "YlGn") +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Crown intactness score\n(1 = 0\u201325% intact  \u2026  4 = 75\u2013100% intact)",
    y = "N trees",
    title = "Crown intactness (score 1\u20134)"
  )
```

---

# Crown illumination

```{r crown-light, fig.height=3.5}
dat |>
  filter(!is.na(Light)) |>
  count(Light) |>
  ggplot(aes(x = factor(Light), y = n, fill = factor(Light))) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(labels = comma) +
  labs(
    x = paste0("Crown illumination index\n",
               "(1 = no direct light  \u2026  5 = fully exposed)"),
    y = "N trees",
    title = "Crown illumination index (score 1\u20135)"
  )
```

---

# Liana infestation

```{r lianas, fig.height=3.5}
dat |>
  filter(!is.na(Lianas)) |>
  count(Lianas) |>
  ggplot(aes(x = factor(Lianas), y = n, fill = factor(Lianas))) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Oranges") +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Liana cover score\n(0 = none  \u2026  4 = 76\u2013100% covered)",
    y = "N trees",
    title = "Liana infestation of crown (score 0\u20134)"
  )
```

---

# Top 20 most abundant species

```{r top-spp, fig.height=5}
dat |>
  count(LatinBinomial, sort = TRUE) |>
  head(20) |>
  ggplot(aes(x = reorder(LatinBinomial, n), y = n)) +
  geom_col(fill = "#74c476") +
  geom_text(aes(label = n), hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    x = NULL, y = "N individuals",
    title = "Top 20 most abundant species \u2014 ForestGEO Panama"
  )
```

---

# Measurements per year

```{r years, fig.height=3.5}
dat |>
  count(YYYY) |>
  ggplot(aes(x = YYYY, y = n)) +
  geom_col(fill = "#9ecae1") +
  geom_text(aes(label = n), vjust = -0.4, size = 3.2) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  scale_x_continuous(breaks = 2011:2020) +
  labs(x = "Year", y = "N measurements",
       title = "Tree measurements per year (2011\u20132020)")
```

---

# Data completeness

```{r completeness, fig.height=4}
key_cols <- c("DBH_mm", "TreeHt_m", "HOM_m", "LatinBinomial",
              "Plotname", "Crown", "Light", "Lianas", "YYYY", "Observer")

compl_df <- data.frame(
  Column       = key_cols,
  N_nonNA      = sapply(key_cols, function(c) sum(!is.na(dat[[c]]))),
  Pct_complete = paste0(
    sapply(key_cols, function(c) round(100 * mean(!is.na(dat[[c]])), 1)),
    "%"
  ),
  stringsAsFactors = FALSE
)
kable(compl_df, caption = "Completeness of key variables (n = 5,109 trees)")

data.frame(
  Column = key_cols,
  pct    = sapply(key_cols, function(c) round(100 * mean(!is.na(dat[[c]])), 1))
) |>
  ggplot(aes(x = reorder(Column, pct), y = pct)) +
  geom_col(fill = "#9ecae1") +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 108)) +
  labs(x = NULL, y = "% complete",
       title = "Data completeness by key column")
```

---

# Notes for allometric merging

1. **DBH units:** Raw data are in **mm** — convert to cm (`DBH_cm = DBH_mm / 10`)
   before merging with BAAD, Tallo, Niklas-Enquist, or NEON (all use cm or the
   equivalent SI unit).

2. **Non-standard HOM:** Diameter is measured at a paint-marked point varying
   from 0.3 to 4.8 m.  Trees with `HOM_m > 1.3 m` are buttressed individuals.
   Allometries that assume DBH at 1.3 m BH should flag these or apply a buttress
   correction (e.g., Cushman et al. 2014, *For. Ecol. Manage.*).

3. **No global lat/lon:** The dataset uses plot-local Cartesian coordinates
   (`PX_m`, `PY_m`).  Global coordinates must be retrieved from the ForestGEO
   site registry (<https://forestgeo.si.edu/sites>) before any spatial analysis
   or merging with georeferenced datasets.

4. **No AGB in source:** Only diameters and heights are provided.  AGB must be
   estimated by applying an allometric equation (e.g., Chave et al. 2014
   pantropical equation), which requires a wood density lookup (Zanne et al.
   2009 backbone via BIOMASS R package).

5. **Taxonomy:** Species are identified to `LatinBinomial` with no authority or
   taxonomic backbone.  TNRS or GBIF name matching is recommended before any
   cross-source taxonomic merge.

6. **Undocumented `B` column:** All 5,109 values are `FALSE`.  No definition
   appears in the deposited data dictionary; confirm with data depositors if
   this flag is relevant before excluding or retaining records.

---

# Session info

```{r session}
sessionInfo()
```
