---
title: "Grand Cross-Study Plant Allometry: Theory Testing, Inter- and Intraspecific Scaling"
subtitle: "OLS, SMA/RMA, PGLS across Tallo, BAAD, NEON, ForestGEO Panama, Niklas-Enquist"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(readr); library(dplyr); library(ggplot2); library(scales)
  library(tidyr); library(knitr); library(smatr); library(purrr)
  library(stringr); library(tibble); library(caper); library(ape)
  library(phytools); library(lme4); library(leaflet); library(gridExtra)
})

HAS_BAAD      <- requireNamespace("baad.data",    quietly = TRUE)
HAS_PHYLOMAKER <- requireNamespace("V.PhyloMaker2", quietly = TRUE)
HAS_METAFOR   <- requireNamespace("metafor",       quietly = TRUE)

# Explicit namespace bindings to prevent masking conflicts
map       <- purrr::map
map_dbl   <- purrr::map_dbl
map_int   <- purrr::map_int
map_chr   <- purrr::map_chr
map_lgl   <- purrr::map_lgl
map_df    <- purrr::map_df
pmap      <- purrr::pmap
pmap_dbl  <- purrr::pmap_dbl
pmap_chr  <- purrr::pmap_chr
select    <- dplyr::select
filter    <- dplyr::filter
mutate    <- dplyr::mutate
rename    <- dplyr::rename
arrange   <- dplyr::arrange
group_by  <- dplyr::group_by
summarise <- dplyr::summarise
summarize <- dplyr::summarize
slice     <- dplyr::slice

set.seed(42)
theme_set(theme_bw(base_size = 12))

proj_dir <- tryCatch(dirname(knitr::current_input(dir = TRUE)),
                     error = function(e) getwd())

# Theory prediction constants
# WBE vascular network model (West et al. 1997, Science 276:122; 1999, Science 284:1677)
# numerically coincides with elastic similarity (McMahon 1973, Science 179:1201) but
# derived independently from vascular branching geometry, NOT from Euler buckling mechanics.
WBE_H_DBH      <- 2/3   # H ~ DBH: WBE vascular network model (West et al. 1997, 1999)
ELASTIC_H_DBH  <- 2/3   # elastic similarity (McMahon 1973); same value, distinct mechanism
STRESS_H_DBH   <- 1/2   # stress similarity: source UNVERIFIED;
                         # McMahon (1973) derives ELASTIC similarity (2/3), NOT this value
WBE_AGB_DBH    <- 8/3   # AGB ~ D^(8/3): from H~D^(2/3) + cylindrical stem + const density
                         # (Enquist 2002, Tree Physiology 22:1045, DOI:10.1093/treephys/22.15-16.1045)
WBE_GROWTH_M   <- 3/4   # dM/dt ~ M^(3/4): WBE metabolic scaling (West et al. 1997)
WBE_CROWN_DBH  <- 1.0   # Crown radius ~ DBH^1: pipe model (Shinozaki et al. 1964, Jpn J Ecol 14:97)
WBE_LEAF_TOTAL <- 3/4   # leaf ~ non-photosynthetic biomass^(3/4): WBE
WBE_STEM_TOTAL <- 1.0   # stem ~ total: isometric
```

---

# Data Loading and Harmonisation {#sec-data}

```{r load-tallo}
tallo_raw <- read_csv(file.path(proj_dir, "data/raw/tallo/Tallo.csv"),
                      show_col_types = FALSE)
tallo_clean <- tallo_raw |>
  filter(height_outlier != "Y" | is.na(height_outlier),
         crown_radius_outlier != "Y" | is.na(crown_radius_outlier),
         !grepl("^sp\\.?$", species, ignore.case = TRUE),
         !is.na(stem_diameter_cm), stem_diameter_cm > 0) |>
  mutate(
    source        = "Tallo",
    DBH_cm        = stem_diameter_cm,
    Height_m      = height_m,
    AGB_kg        = NA_real_,
    AGB_method    = NA_character_,
    CrownRadius_m = crown_radius_m,
    genus         = genus,
    family        = family,
    division      = division,
    lat           = latitude,
    lon           = longitude,
    agb_eligible  = FALSE
  ) |>
  select(source, DBH_cm, Height_m, AGB_kg, AGB_method, CrownRadius_m,
         species, genus, family, division, lat, lon, agb_eligible)
```

```{r load-baad}
if (HAS_BAAD) {
  baad_raw <- baad.data::baad_data()$data
  baad_clean <- baad_raw |>
    mutate(
      source        = "BAAD",
      DBH_cm        = d.bh * 100,   # CRITICAL: d.bh is in metres
      Height_m      = h.t,
      AGB_kg        = m.to,
      AGB_method    = "direct",
      CrownRadius_m = NA_real_,
      species       = speciesMatched,
      genus         = word(speciesMatched, 1),
      family        = NA_character_,
      division      = NA_character_,
      lat           = NA_real_,
      lon           = NA_real_,
      agb_eligible  = !is.na(m.to)
    ) |>
    filter(!is.na(DBH_cm), DBH_cm > 0) |>
    select(source, DBH_cm, Height_m, AGB_kg, AGB_method, CrownRadius_m,
           species, genus, family, division, lat, lon, agb_eligible)
} else {
  baad_raw   <- NULL
  baad_clean <- tibble(source = character(), DBH_cm = numeric(),
    Height_m = numeric(), AGB_kg = numeric(), AGB_method = character(),
    CrownRadius_m = numeric(), species = character(), genus = character(),
    family = character(), division = character(), lat = numeric(),
    lon = numeric(), agb_eligible = logical())
  message("baad.data not available — BAAD sections will be skipped.")
}
```

```{r load-neon}
neon_sites <- c("BART","BLAN","CLBJ","DELA","DSNY","GRSM","GUAN","HARV",
                "JERC","KONZ","LENO","MLBS","ORNL","OSBS","RMNP","SCBI",
                "SERC","STEI","TALL","TREE","UKFS","UNDE","YELL")
bm_dir <- file.path(proj_dir, "data/raw/neon_forest_agb/biomass")

site_coords <- tryCatch(
  read_csv(file.path(proj_dir, "data/raw/neon_forest_agb/sitecoordinates.csv"),
           show_col_types = FALSE),
  error = function(e) tibble(siteID = character(), X = numeric(), Y = numeric()))

neon_list <- map(neon_sites, function(s) {
  tryCatch({
    d <- read_csv(file.path(bm_dir, paste0(s, "_biomass.csv")),
                  show_col_types = FALSE)
    d |>
      filter(plantStatus == "Live" | is.na(plantStatus),
             !is.na(stemDiameter), stemDiameter > 0) |>
      mutate(
        source        = "NEON",
        siteID_tmp    = s,
        DBH_cm        = stemDiameter,
        Height_m      = height,
        AGB_kg        = AGBChojnacky,
        AGB_method    = "allometric_chojnacky",
        CrownRadius_m = NA_real_,
        species       = scientificName,
        genus         = Genus,
        family        = Family,
        division      = NA_character_,
        agb_eligible  = FALSE
      ) |>
      left_join(site_coords |> select(siteID, lon = X, lat = Y),
                by = c("siteID_tmp" = "siteID")) |>
      select(source, DBH_cm, Height_m, AGB_kg, AGB_method, CrownRadius_m,
             species, genus, family, division, lat, lon, agb_eligible)
  }, error = function(e) NULL)
})
neon_clean <- bind_rows(Filter(Negate(is.null), neon_list))
```

```{r load-forestgeo}
fgeo_raw <- read_csv(
  file.path(proj_dir,
    "data/raw/forestgeo_panama/TreeHeightDataSmallPlotsCombined_2024_08_05.csv"),
  show_col_types = FALSE)

# BCI / Panama small-plot approximate centroid
FGEO_LAT <- 9.15; FGEO_LON <- -79.84

fgeo_clean <- fgeo_raw |>
  filter(!is.na(DBH_mm), DBH_mm > 0, !is.na(TreeHt_m), TreeHt_m > 0) |>
  mutate(
    source        = "ForestGEO_Panama",
    DBH_cm        = DBH_mm / 10,
    Height_m      = TreeHt_m,
    AGB_kg        = NA_real_,
    AGB_method    = NA_character_,
    CrownRadius_m = NA_real_,
    species       = LatinBinomial,
    genus         = word(LatinBinomial, 1),
    family        = NA_character_,
    division      = NA_character_,
    lat           = FGEO_LAT,
    lon           = FGEO_LON,
    agb_eligible  = FALSE
  ) |>
  select(source, DBH_cm, Height_m, AGB_kg, AGB_method, CrownRadius_m,
         species, genus, family, division, lat, lon, agb_eligible)
```

```{r load-niklas}
nik_col_names <- c(
  "taxa","citation","footnotes","record_num","age_yr","log_age",
  "height_m","log_height","root_biomass_kg","log_root","stem_biomass_kg",
  "log_stem","leaf_biomass_kg","log_leaf","shoot_biomass_kg","log_shoot",
  "repro_biomass_kg","log_repro","total_biomass_kg","log_total",
  "stem_growth_kgyr","log_stem_growth","leaf_growth_kgyr","log_leaf_growth",
  "root_growth_kgyr","log_root_growth","total_growth_kgyr","log_total_growth")

nik_enq <- read_csv(
  file.path(proj_dir, "data/raw/niklas_enquist/niklas_enquist_biomass_20040122.csv"),
  skip = 2, col_names = nik_col_names,
  locale = locale(encoding = "latin1"), show_col_types = FALSE) |>
  mutate(across(-c(taxa, citation, footnotes), as.numeric)) |>
  mutate(across(where(is.numeric), ~ ifelse(. == -999, NA_real_, .))) |>
  mutate(pgls_eligible = !grepl(
    "with a few|community|stand|plot|mixed", taxa, ignore.case = TRUE))
```

```{r harmonise-allom}
allom <- bind_rows(tallo_clean, baad_clean, neon_clean, fgeo_clean) |>
  filter(!is.na(species), species != "",
         !grepl("^sp\\.?$|^[A-Z][a-z]+ sp\\.$", species, ignore.case = FALSE))
```

```{r summary-table}
src_summary <- allom |>
  group_by(source) |>
  summarise(
    N_records = n(),
    N_species = n_distinct(species, na.rm = TRUE),
    N_families = n_distinct(family[!is.na(family)]),
    lat_range = ifelse(all(is.na(lat)), "NA",
      paste0(round(min(lat, na.rm = TRUE), 1), " to ",
             round(max(lat, na.rm = TRUE), 1))),
    lon_range = ifelse(all(is.na(lon)), "NA",
      paste0(round(min(lon, na.rm = TRUE), 1), " to ",
             round(max(lon, na.rm = TRUE), 1))),
    .groups = "drop")

kable(src_summary, caption = "Summary of harmonised allom tibble by source.",
      format.args = list(big.mark = ","))
```

---

# Geographic Overview {#sec-geo}

```{r geo-leaflet}
# Subsample Tallo for leaflet performance
set.seed(42)
tallo_ll <- tallo_clean |>
  filter(!is.na(lat), !is.na(lon)) |>
  sample_n(min(20000, n())) |>
  mutate(col = "#2196F3")

neon_ll <- site_coords |>
  filter(!is.na(X), !is.na(Y)) |>
  mutate(source = "NEON", lat = Y, lon = X, col = "#FF5722")

fgeo_ll <- data.frame(lat = FGEO_LAT, lon = FGEO_LON,
                      source = "ForestGEO_Panama", col = "#4CAF50")

pal_src <- colorFactor(
  palette = c("#2196F3","#9C27B0","#FF5722","#4CAF50"),
  levels  = c("Tallo","BAAD","NEON","ForestGEO_Panama"))

m <- leaflet() |>
  addTiles() |>
  addCircleMarkers(data = tallo_ll, ~lon, ~lat, radius = 2,
    color = "#2196F3", opacity = 0.4, fillOpacity = 0.4,
    group = "Tallo") |>
  addCircleMarkers(data = neon_ll, ~lon, ~lat, radius = 7,
    color = "#FF5722", opacity = 0.9, fillOpacity = 0.9,
    popup = ~siteID, group = "NEON sites") |>
  addCircleMarkers(data = fgeo_ll, ~lon, ~lat, radius = 8,
    color = "#4CAF50", opacity = 0.9, fillOpacity = 0.9,
    popup = "ForestGEO Panama", group = "ForestGEO Panama") |>
  addLayersControl(overlayGroups = c("Tallo","NEON sites","ForestGEO Panama"),
                   options = layersControlOptions(collapsed = FALSE))
m
```

```{r geo-barchart, fig.height=4}
allom |>
  count(source) |>
  mutate(source = reorder(source, n)) |>
  ggplot(aes(source, n, fill = source)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = comma) +
  coord_flip() +
  labs(title = "Records per dataset", x = NULL, y = "N records")
```

```{r tallo-family-map, fig.height=5}
top10_fam <- tallo_clean |>
  count(family, sort = TRUE) |>
  slice(1:10) |>
  pull(family)

set.seed(42)
tallo_clean |>
  filter(!is.na(lat), !is.na(lon), family %in% top10_fam) |>
  sample_n(min(50000, n())) |>
  ggplot(aes(lon, lat, colour = family)) +
  geom_point(size = 0.3, alpha = 0.4) +
  coord_quickmap() +
  labs(title = "Tallo records — top 10 families",
       x = "Longitude", y = "Latitude", colour = "Family") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))
```

---

# Taxonomic Coverage and Size Range Audit {#sec-taxonomy}

```{r taxonomy-audit}
fam_audit <- allom |>
  filter(!is.na(DBH_cm), DBH_cm > 0, !is.na(family)) |>
  group_by(family, division) |>
  summarise(
    n_records   = n(),
    n_species   = n_distinct(species, na.rm = TRUE),
    n_datasets  = n_distinct(source),
    log10_range = log10(max(DBH_cm, na.rm = TRUE) /
                          max(min(DBH_cm, na.rm = TRUE), 1e-6)),
    .groups = "drop") |>
  mutate(
    pgls_eligible = n_species  >= 8  & log10_range >= 2.0,
    sma_eligible  = (n_species >= 5  | n_records >= 20) & log10_range >= 1.5,
    intra_eligible = n_records >= 10 & log10_range >= 1.5) |>
  arrange(desc(n_records))

kable(head(fam_audit, 30), digits = 2,
      caption = "Family-level coverage and eligibility flags (top 30 by N records).")
```

```{r taxonomy-barchart, fig.height=5}
fam_audit |>
  slice(1:20) |>
  mutate(
    tier = case_when(pgls_eligible ~ "PGLS-eligible",
                     sma_eligible  ~ "SMA-eligible",
                     TRUE          ~ "Low coverage"),
    family = reorder(family, n_records)) |>
  ggplot(aes(family, n_records, fill = tier)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("PGLS-eligible" = "#1976D2",
                               "SMA-eligible"  = "#66BB6A",
                               "Low coverage"  = "#BDBDBD")) +
  labs(title = "Top 20 families — record counts and eligibility tier",
       x = NULL, y = "N records", fill = "Eligibility")
```

```{r division-summary}
div_sum <- allom |>
  filter(!is.na(division), !is.na(DBH_cm), DBH_cm > 0) |>
  group_by(division) |>
  summarise(n_records = n(), n_species = n_distinct(species),
            DBH_min_cm = min(DBH_cm), DBH_max_cm = max(DBH_cm),
            .groups = "drop")
kable(div_sum, digits = 1, caption = "Division-level summary (DBH records).")
```

---

# H ~ DBH Scaling {#sec-h-dbh}

## Pooled cross-study {#sec-h-dbh-pooled}

```{r h-dbh-pooled}
pool_hd <- allom |>
  filter(!is.na(DBH_cm), !is.na(Height_m),
         DBH_cm > 0, Height_m > 0) |>
  mutate(log_dbh = log10(DBH_cm), log_h = log10(Height_m))

ols_hd  <- lm(log_h ~ log_dbh, data = pool_hd)
sma_hd  <- tryCatch(sma(log_h ~ log_dbh, data = pool_hd), error = function(e) NULL)
lmm_hd  <- tryCatch(
  lmer(log_h ~ log_dbh + (1 | source), data = pool_hd, REML = FALSE),
  error = function(e) NULL)

ols_coef <- coef(ols_hd)
sma_coef <- if (!is.null(sma_hd)) as.numeric(coef(sma_hd)[[1]]) else NULL

ols_ci  <- confint(ols_hd, "log_dbh")
sma_ci  <- if (!is.null(sma_hd)) as.numeric(sma_hd$slopetest[[1]]$test[c("lower","upper")]) else c(NA_real_, NA_real_)

sma_test_hd <- if (!is.null(sma_hd)) tryCatch(
  slope.test(pool_hd$log_h, pool_hd$log_dbh, test.value = WBE_H_DBH, method = "SMA"),
  error = function(e) list(p = NA_real_)) else list(p = NA_real_)

lmm_slope <- if (!is.null(lmm_hd)) as.numeric(fixef(lmm_hd)["log_dbh"]) else NA_real_

hd_table <- tibble(
  Method = c("OLS","SMA","LMM (random source)"),
  Slope  = c(as.numeric(ols_coef["log_dbh"]),
             if (!is.null(sma_coef)) as.numeric(sma_coef["slope"]) else NA_real_,
             lmm_slope),
  CI_lo  = c(ols_ci[1], if (!is.null(sma_coef)) sma_ci[1] else NA_real_, NA_real_),
  CI_hi  = c(ols_ci[2], if (!is.null(sma_coef)) sma_ci[2] else NA_real_, NA_real_),
  R2     = c(summary(ols_hd)$r.squared, NA_real_, NA_real_),
  p_vs_WBE = c(
    2 * pt(abs((ols_coef["log_dbh"] - WBE_H_DBH) /
                summary(ols_hd)$coefficients["log_dbh","Std. Error"]),
             df = df.residual(ols_hd), lower.tail = FALSE),
    sma_test_hd$p,
    NA_real_))

kable(hd_table, digits = 3,
      caption = paste0("H ~ DBH pooled regression (N = ", nrow(pool_hd),
                       "). WBE prediction β = 2/3."))
```

```{r h-dbh-scatter, fig.height=5}
set.seed(42)
plot_hd <- pool_hd |> sample_n(min(50000, n()))

ggplot(plot_hd, aes(log_dbh, log_h, colour = source)) +
  geom_point(size = 0.3, alpha = 0.3) +
  geom_abline(slope = as.numeric(ols_coef["log_dbh"]),
              intercept = as.numeric(ols_coef["(Intercept)"]),
              colour = "black", linewidth = 1, linetype = "solid") +
  {if (!is.null(sma_coef)) geom_abline(
              slope = as.numeric(sma_coef["slope"]),
              intercept = as.numeric(sma_coef["y-intercept"]),
              colour = "#1565C0", linewidth = 1, linetype = "dashed")} +
  geom_abline(slope = WBE_H_DBH,
              intercept = mean(plot_hd$log_h) - WBE_H_DBH * mean(plot_hd$log_dbh),
              colour = "red", linewidth = 1, linetype = "dotted") +
  scale_colour_brewer(palette = "Set1") +
  labs(title = paste0("H ~ DBH — ", nrow(plot_hd), " subsampled points"),
       subtitle = "Black=OLS, Blue-dashed=SMA, Red-dotted=WBE (β=2/3)",
       x = "log10(DBH cm)", y = "log10(Height m)", colour = "Source")
```

## Study-level slopes {#sec-h-dbh-study}

```{r h-dbh-study-slopes}
study_hd <- pool_hd |>
  group_by(source) |>
  group_map(function(d, k) {
    if (nrow(d) < 5) return(NULL)
    o  <- lm(log_h ~ log_dbh, data = d)
    s  <- tryCatch(sma(log_h ~ log_dbh, data = d), error = function(e) NULL)
    ci <- confint(o, "log_dbh")
    se <- summary(o)$coefficients["log_dbh", "Std. Error"]
    tibble(
      source  = k$source,
      n       = nrow(d),
      ols_slope = coef(o)["log_dbh"],
      ols_ci_lo = ci[1], ols_ci_hi = ci[2],
      ols_se  = se,
      r2      = summary(o)$r.squared,
      sma_slope = if (!is.null(s)) as.numeric(coef(s)[[1]]["slope"]) else NA_real_)
  }) |> bind_rows()

if (HAS_METAFOR) {
  re_hd <- tryCatch(
    metafor::rma(yi = ols_slope, sei = ols_se, data = study_hd, method = "REML"),
    error = function(e) NULL)
} else {
  # Manual Cochran Q
  w     <- 1 / study_hd$ols_se^2
  wm    <- weighted.mean(study_hd$ols_slope, w)
  Q_hd  <- sum(w * (study_hd$ols_slope - wm)^2)
  df_hd <- nrow(study_hd) - 1
  I2_hd <- max(0, (Q_hd - df_hd) / Q_hd * 100)
  re_hd <- NULL
}

I2_report <- if (!is.null(re_hd) && HAS_METAFOR) {
  round(re_hd$I2, 1)
} else if (exists("I2_hd")) round(I2_hd, 1) else NA_real_

if (!is.na(I2_report)) cat(sprintf("H~DBH heterogeneity: I² = %.1f%%\n", I2_report)) else cat("H~DBH I² not estimable.\n")

study_hd |>
  mutate(source = reorder(source, ols_slope)) |>
  ggplot(aes(source, ols_slope)) +
  geom_pointrange(aes(ymin = ols_ci_lo, ymax = ols_ci_hi), size = 0.6) +
  geom_hline(yintercept = WBE_H_DBH, colour = "red", linetype = "dashed",
             linewidth = 0.8) +
  geom_hline(yintercept = STRESS_H_DBH, colour = "orange", linetype = "dotted",
             linewidth = 0.8) +
  coord_flip() +
  labs(title = "H ~ DBH OLS slopes per source",
       subtitle = "Red=WBE 2/3, Orange=stress similarity 1/2",
       x = NULL, y = "OLS slope ± 95% CI")
```

## Gymnosperm vs Angiosperm {#sec-h-dbh-division}

```{r h-dbh-division}
pool_hd_div <- pool_hd |> filter(!is.na(division))
if (nrow(pool_hd_div) > 50) {
  div_fits <- pool_hd_div |>
    group_by(division) |>
    group_map(function(d, k) {
      o <- lm(log_h ~ log_dbh, data = d)
      s <- tryCatch(sma(log_h ~ log_dbh, data = d), error = function(e) NULL)
      tibble(division = k$division, n = nrow(d),
             ols_slope = coef(o)["log_dbh"],
             ci_lo = confint(o,"log_dbh")[1], ci_hi = confint(o,"log_dbh")[2],
             sma_slope = if (!is.null(s)) as.numeric(coef(s)[[1]]["slope"]) else NA_real_)
    }) |> bind_rows()

  kable(div_fits, digits = 3,
        caption = "H ~ DBH by division. WBE prediction = 2/3.")

  set.seed(42)
  pool_hd_div |>
    filter(division %in% c("Angiosperm","Gymnosperm")) |>
    sample_n(min(30000, n())) |>
    ggplot(aes(log_dbh, log_h, colour = division)) +
    geom_point(size = 0.3, alpha = 0.3) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
    geom_abline(slope = WBE_H_DBH,
                intercept = mean(pool_hd$log_h) - WBE_H_DBH * mean(pool_hd$log_dbh),
                colour = "red", linetype = "dashed") +
    scale_colour_manual(values = c(Angiosperm = "#1976D2", Gymnosperm = "#E65100")) +
    labs(title = "H ~ DBH by division", x = "log10(DBH cm)", y = "log10(Height m)")
} else message("Insufficient division-labelled records.")
```

## Family-level SMA forest plot {#sec-h-dbh-family}

```{r h-dbh-family-sma, fig.height=7}
sma_elig_fams <- fam_audit |> filter(sma_eligible) |> pull(family)

fam_hd_fits <- map_df(sma_elig_fams, function(fam) {
  d <- pool_hd |> filter(family == fam)
  if (nrow(d) < 10) return(NULL)
  s <- tryCatch(sma(log_h ~ log_dbh, data = d), error = function(e) NULL)
  if (is.null(s)) return(NULL)
  tibble(family = fam, n = nrow(d),
         slope  = as.numeric(coef(s)[[1]]["slope"]),
         ci_lo  = as.numeric(s$slopetest[[1]]$test["lower"]),
         ci_hi  = as.numeric(s$slopetest[[1]]$test["upper"]),
         division = unique(d$division[!is.na(d$division)])[1])
})

if (nrow(fam_hd_fits) > 0) {
  fam_hd_fits |>
    mutate(family = reorder(family, slope)) |>
    ggplot(aes(slope, family, colour = division)) +
    geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi), size = 0.5) +
    geom_vline(xintercept = WBE_H_DBH, colour = "red",
               linetype = "dashed", linewidth = 0.8) +
    scale_colour_manual(values = c(Angiosperm = "#1976D2", Gymnosperm = "#E65100"),
                        na.value = "grey50") +
    labs(title = "Family-level SMA slopes — H ~ DBH",
         subtitle = "Red dashed = WBE 2/3",
         x = "SMA slope ± 95% CI", y = NULL, colour = "Division") +
    theme(axis.text.y = element_text(size = 8))
}
```

## PGLS for H ~ DBH {#sec-h-dbh-pgls}

```{r h-dbh-pgls}
# Aggregate to species means across all direct-measurement sources
spp_hd <- allom |>
  filter(!is.na(DBH_cm), !is.na(Height_m), DBH_cm > 0, Height_m > 0,
         !is.na(species), !grepl("^sp\\.?$", species)) |>
  mutate(log_dbh = log10(DBH_cm), log_h = log10(Height_m)) |>
  group_by(species, family) |>
  summarise(log_dbh_m = mean(log_dbh, na.rm = TRUE),
            log_h_m   = mean(log_h, na.rm = TRUE),
            n = n(), .groups = "drop") |>
  filter(n >= 5, !is.na(species)) |>  # n >= 5 for reliable species means in PGLS
  mutate(sp_tree = gsub(" ", "_", species)) |>
  distinct(sp_tree, .keep_all = TRUE)

ols_pgls_hd <- lm(log_h_m ~ log_dbh_m, data = spp_hd)
sma_pgls_hd <- tryCatch(sma(log_h_m ~ log_dbh_m, data = spp_hd),
                         error = function(e) NULL)

if (HAS_PHYLOMAKER && nrow(spp_hd) > 20) {
  # Cap to 600 species for tractable phylogeny building; prefer higher-n species
  spp_hd_phy <- spp_hd |>
    filter(!is.na(family)) |>
    arrange(desc(n)) |>
    slice_head(n = 600)

  sp_list_hd <- spp_hd_phy |>
    mutate(genus_col = word(species, 1)) |>
    distinct(species, genus_col, family) |>
    rename(species_col = species)

  phy_hd <- tryCatch({
    res <- V.PhyloMaker2::phylo.maker(
      sp.list = data.frame(species = gsub(" ","_", sp_list_hd$species_col),
                           genus   = sp_list_hd$genus_col,
                           family  = sp_list_hd$family),
      tree      = V.PhyloMaker2::GBOTB.extended.WP,
      nodes     = V.PhyloMaker2::nodes.info.1.WP,
      scenarios = "S3")
    res$scenario.3
  }, error = function(e) {
    message("V.PhyloMaker2 failed — PGLS skipped (no valid phylogeny available).")
    NULL
  })

  if (!is.null(phy_hd)) {
    set.seed(42)
    phy_hd <- multi2di(phy_hd, random = TRUE)
    phy_hd <- phytools::force.ultrametric(phy_hd, method = "extend")
    phy_hd$node.label <- NULL

    pgls_df_hd <- spp_hd_phy |>
      filter(sp_tree %in% phy_hd$tip.label) |>
      distinct(sp_tree, .keep_all = TRUE) |>
      as.data.frame()
    rownames(pgls_df_hd) <- pgls_df_hd$sp_tree

    cd_hd <- tryCatch(
      caper::comparative.data(phy = phy_hd, data = pgls_df_hd,
                              names.col = "sp_tree", vcv = TRUE),
      error = function(e) NULL)

    if (!is.null(cd_hd)) {
    pgls_fit_hd <- tryCatch(
      caper::pgls(log_h_m ~ log_dbh_m, data = cd_hd,
                  lambda = "ML", bounds = list(lambda = c(1e-6, 1))),
      error = function(e) NULL)

    if (!is.null(pgls_fit_hd)) {
      pgls_slope  <- coef(pgls_fit_hd)["log_dbh_m"]
      pgls_se     <- sqrt(diag(pgls_fit_hd$VCV))["log_dbh_m"]
      pgls_ci     <- pgls_slope + c(-1, 1) *
                       qt(0.975, length(pgls_fit_hd$residuals) - 2) * pgls_se
      pgls_lambda <- pgls_fit_hd$param["lambda"]

      lp <- tryCatch(pgls.profile(pgls_fit_hd, which = "lambda"),
                     error = function(e) NULL)
      lambda_ci <- if (!is.null(lp)) tryCatch(pgls.confint(lp)$ci.val,
                                               error = function(e) c(NA, NA)) else c(NA, NA)

      pgls_r2 <- summary(pgls_fit_hd)$r.squared

      hd_pgls_table <- tibble(
        Method = c("OLS (spp means)","SMA (spp means)","PGLS"),
        Slope  = c(coef(ols_pgls_hd)["log_dbh_m"],
                   if (!is.null(sma_pgls_hd)) as.numeric(coef(sma_pgls_hd)[[1]]["slope"]) else NA_real_,
                   pgls_slope),
        CI_lo  = c(confint(ols_pgls_hd,"log_dbh_m")[1],
                   if (!is.null(sma_pgls_hd)) as.numeric(sma_pgls_hd$slopetest[[1]]$test["lower"]) else NA_real_,
                   pgls_ci[1]),
        CI_hi  = c(confint(ols_pgls_hd,"log_dbh_m")[2],
                   if (!is.null(sma_pgls_hd)) as.numeric(sma_pgls_hd$slopetest[[1]]$test["upper"]) else NA_real_,
                   pgls_ci[2]),
        Lambda = c(NA, NA, pgls_lambda),
        R2     = c(summary(ols_pgls_hd)$r.squared, NA_real_, pgls_r2))

      kable(hd_pgls_table, digits = 3,
            caption = paste0("H ~ DBH: OLS / SMA / PGLS comparison (N spp = ",
                             nrow(pgls_df_hd), "). WBE β = 2/3."))
    }
  }
  } # end if (!is.null(phy_hd))
} else {
  message("PGLS skipped (V.PhyloMaker2 unavailable or insufficient species).")
}
```

## Intraspecific H ~ DBH slopes {#sec-h-dbh-intra}

```{r h-dbh-intraspecific, fig.height=5}
intra_hd <- pool_hd |>
  group_by(species, family) |>
  filter(n() >= 10,
         log10(max(DBH_cm) / pmax(min(DBH_cm), 1e-6)) >= 1.5) |>
  group_map(function(d, k) {
    o  <- lm(log_h ~ log_dbh, data = d)
    ci <- confint(o, "log_dbh")
    tibble(species = k$species, family = k$family,
           slope = coef(o)["log_dbh"],
           ci_lo = ci[1], ci_hi = ci[2],
           r2 = summary(o)$r.squared, n = nrow(d))
  }) |> bind_rows()

if (nrow(intra_hd) > 0) {
  t_intra_hd <- t.test(intra_hd$slope, mu = WBE_H_DBH)
  cat(sprintf("Intraspecific H~DBH: mean β = %.3f ± %.3f SD; t-test vs 2/3: p = %.4f\n",
              mean(intra_hd$slope), sd(intra_hd$slope), t_intra_hd$p.value))

  ggplot(intra_hd, aes(slope)) +
    geom_histogram(bins = 40, fill = "#42A5F5", colour = "white") +
    geom_vline(xintercept = WBE_H_DBH, colour = "red",
               linetype = "dashed", linewidth = 1) +
    geom_vline(xintercept = 1.0, colour = "grey40",
               linetype = "dotted", linewidth = 0.8) +
    labs(title = paste0("Intraspecific H ~ DBH slopes (N species = ",
                        nrow(intra_hd), ")"),
         subtitle = "Red=WBE 2/3, Grey=isometric 1.0",
         x = "OLS slope", y = "Count")

  kable(intra_hd |> arrange(desc(n)) |> head(20),
        digits = 3, caption = "Top 20 species by N (intraspecific H~DBH).")
}
```

---

# AGB ~ DBH Scaling {#sec-agb-dbh}

> ⚠️ **NEON AGB circularity caveat:** NEON AGB values were computed from DBH
> using Chojnacky et al. (2014) allometric equations. For NEON records, AGB ~ DBH
> is partly circular — the AGB is itself a function of DBH. NEON is included in
> the sensitivity analysis (Section 5.1 sensitivity) but excluded from the primary
> PGLS and family-level analyses.

```{r agb-setup}
# Primary: BAAD only (direct AGB)
pool_agb_primary <- allom |>
  filter(agb_eligible, !is.na(DBH_cm), !is.na(AGB_kg),
         DBH_cm > 0, AGB_kg > 0) |>
  mutate(log_dbh = log10(DBH_cm), log_agb = log10(AGB_kg))

# Sensitivity: include NEON
pool_agb_all <- allom |>
  filter(!is.na(DBH_cm), !is.na(AGB_kg), DBH_cm > 0, AGB_kg > 0) |>
  mutate(log_dbh = log10(DBH_cm), log_agb = log10(AGB_kg))
```

## Pooled AGB ~ DBH {#sec-agb-pooled}

```{r agb-dbh-pooled}
if (nrow(pool_agb_primary) > 20) {
  ols_agb  <- lm(log_agb ~ log_dbh, data = pool_agb_primary)
  sma_agb  <- tryCatch(sma(log_agb ~ log_dbh, data = pool_agb_primary),
                        error = function(e) NULL)
  sma_test_agb <- tryCatch(
    slope.test(pool_agb_primary$log_agb, pool_agb_primary$log_dbh,
               test.value = WBE_AGB_DBH, method = "SMA"),
    error = function(e) list(p = NA_real_))

  agb_ci <- confint(ols_agb, "log_dbh")
  agb_table <- tibble(
    Method   = c("OLS (BAAD primary)","SMA (BAAD primary)"),
    Slope    = c(coef(ols_agb)["log_dbh"],
                 if (!is.null(sma_agb)) as.numeric(coef(sma_agb)[[1]]["slope"]) else NA_real_),
    CI_lo    = c(agb_ci[1],
                 if (!is.null(sma_agb)) as.numeric(sma_agb$slopetest[[1]]$test["lower"]) else NA_real_),
    CI_hi    = c(agb_ci[2],
                 if (!is.null(sma_agb)) as.numeric(sma_agb$slopetest[[1]]$test["upper"]) else NA_real_),
    R2       = c(summary(ols_agb)$r.squared, NA_real_),
    p_vs_WBE = c(
      2 * pt(abs((coef(ols_agb)["log_dbh"] - WBE_AGB_DBH) /
                   summary(ols_agb)$coefficients["log_dbh","Std. Error"]),
             df = df.residual(ols_agb), lower.tail = FALSE),
      sma_test_agb$p))

  kable(agb_table, digits = 3,
        caption = paste0("AGB ~ DBH pooled (N = ", nrow(pool_agb_primary),
                         " BAAD). WBE prediction β = 8/3 ≈ 2.667."))

  agb_coef <- coef(ols_agb)
  ggplot(pool_agb_primary |> sample_n(min(5000, n())),
         aes(log_dbh, log_agb, colour = source)) +
    geom_point(size = 0.5, alpha = 0.5) +
    geom_abline(slope = agb_coef["log_dbh"], intercept = agb_coef["(Intercept)"],
                colour = "black", linewidth = 1) +
    geom_abline(slope = WBE_AGB_DBH,
                intercept = mean(pool_agb_primary$log_agb) -
                            WBE_AGB_DBH * mean(pool_agb_primary$log_dbh),
                colour = "red", linetype = "dashed", linewidth = 1) +
    labs(title = "AGB ~ DBH (BAAD primary)",
         subtitle = "Black=OLS, Red-dashed=WBE (β=8/3)",
         x = "log10(DBH cm)", y = "log10(AGB kg)", colour = "Source")
} else {
  message("BAAD not available — AGB primary analysis skipped.")
}
```

## Study-level AGB slopes {#sec-agb-study}

```{r agb-dbh-study-slopes}
if (HAS_BAAD && nrow(pool_agb_primary) > 20) {
  # Use baad_raw (cached in load-baad chunk) directly — avoids fragile paste-match
  baad_full <- baad_raw |>
    filter(!is.na(d.bh), !is.na(m.to), d.bh > 0, m.to > 0) |>
    mutate(log_dbh = log10(d.bh * 100), log_agb = log10(m.to))

  study_agb <- baad_full |>
    group_by(studyName) |>
    group_map(function(d, k) {
      if (nrow(d) < 5) return(NULL)
      o  <- lm(log_agb ~ log_dbh, data = d)
      ci <- confint(o, "log_dbh")
      tibble(studyName = k$studyName, n = nrow(d),
             slope = coef(o)["log_dbh"],
             ci_lo = ci[1], ci_hi = ci[2],
             r2 = summary(o)$r.squared)
    }) |> bind_rows()

  if (nrow(study_agb) > 1) {
    study_agb |>
      mutate(studyName = reorder(studyName, slope)) |>
      ggplot(aes(slope, studyName)) +
      geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi), size = 0.5) +
      geom_vline(xintercept = WBE_AGB_DBH, colour = "red",
                 linetype = "dashed", linewidth = 0.8) +
      labs(title = "AGB ~ DBH OLS slopes by BAAD study",
           subtitle = "Red = WBE 8/3",
           x = "OLS slope ± 95% CI", y = NULL)
  }
}
```

## Division comparison, family SMA, PGLS, intraspecific {#sec-agb-rest}

```{r agb-division}
if (HAS_BAAD && nrow(pool_agb_primary) > 20) {
  pool_agb_div <- baad_raw |>
    filter(!is.na(d.bh), !is.na(m.to), d.bh > 0, m.to > 0) |>
    mutate(log_dbh = log10(d.bh * 100), log_agb = log10(m.to),
           family  = NA_character_)
  # Division info not in BAAD — skip division comparison for AGB
  message("BAAD lacks division annotation — AGB division comparison not available.")
}
```

```{r agb-intraspecific}
if (HAS_BAAD && nrow(pool_agb_primary) > 50) {
  intra_agb <- pool_agb_primary |>
    filter(!is.na(species)) |>
    group_by(species) |>
    filter(n() >= 10,
           log10(max(DBH_cm) / pmax(min(DBH_cm), 1e-6)) >= 1.5) |>
    group_map(function(d, k) {
      o  <- lm(log_agb ~ log_dbh, data = d)
      ci <- confint(o, "log_dbh")
      tibble(species = k$species, slope = coef(o)["log_dbh"],
             ci_lo = ci[1], ci_hi = ci[2],
             r2 = summary(o)$r.squared, n = nrow(d))
    }) |> bind_rows()

  if (nrow(intra_agb) > 0) {
    ggplot(intra_agb, aes(slope)) +
      geom_histogram(bins = 30, fill = "#66BB6A", colour = "white") +
      geom_vline(xintercept = WBE_AGB_DBH, colour = "red",
                 linetype = "dashed", linewidth = 1) +
      labs(title = paste0("Intraspecific AGB ~ DBH slopes (N = ",
                          nrow(intra_agb), " species)"),
           subtitle = "Red = WBE 8/3 ≈ 2.667",
           x = "OLS slope", y = "Count")
  }
}
```

---

# Crown Radius ~ DBH (Tallo) {#sec-crown}

```{r crown-setup}
crown_dat <- tallo_clean |>
  filter(!is.na(CrownRadius_m), !is.na(DBH_cm),
         CrownRadius_m > 0, DBH_cm > 0) |>
  mutate(log_dbh = log10(DBH_cm), log_cr = log10(CrownRadius_m))
```

## Pooled crown analysis {#sec-crown-pooled}

```{r crown-dbh-pooled}
if (nrow(crown_dat) > 100) {
  ols_cr  <- lm(log_cr ~ log_dbh, data = crown_dat)
  sma_cr  <- tryCatch(sma(log_cr ~ log_dbh, data = crown_dat),
                       error = function(e) NULL)
  sma_test_cr <- tryCatch(
    slope.test(crown_dat$log_cr, crown_dat$log_dbh,
               test.value = WBE_CROWN_DBH, method = "SMA"),
    error = function(e) list(p = NA_real_))

  cr_ci <- confint(ols_cr, "log_dbh")
  cr_table <- tibble(
    Method   = c("OLS","SMA"),
    Slope    = c(coef(ols_cr)["log_dbh"],
                 if (!is.null(sma_cr)) as.numeric(coef(sma_cr)[[1]]["slope"]) else NA_real_),
    CI_lo    = c(cr_ci[1],
                 if (!is.null(sma_cr)) as.numeric(sma_cr$slopetest[[1]]$test["lower"]) else NA_real_),
    CI_hi    = c(cr_ci[2],
                 if (!is.null(sma_cr)) as.numeric(sma_cr$slopetest[[1]]$test["upper"]) else NA_real_),
    R2       = c(summary(ols_cr)$r.squared, NA_real_),
    p_vs_WBE = c(
      2 * pt(abs((coef(ols_cr)["log_dbh"] - WBE_CROWN_DBH) /
                   summary(ols_cr)$coefficients["log_dbh","Std. Error"]),
             df = df.residual(ols_cr), lower.tail = FALSE),
      sma_test_cr$p))

  kable(cr_table, digits = 3,
        caption = paste0("Crown radius ~ DBH (N = ", nrow(crown_dat),
                         "). WBE pipe model β = 1.0."))
}
```

## Family SMA and division for crown {#sec-crown-family}

```{r crown-dbh-family, fig.height=6}
if (nrow(crown_dat) > 100) {
  fam_cr_fits <- map_df(sma_elig_fams, function(fam) {
    d <- crown_dat |> filter(family == fam)
    if (nrow(d) < 10) return(NULL)
    s <- tryCatch(sma(log_cr ~ log_dbh, data = d), error = function(e) NULL)
    if (is.null(s)) return(NULL)
    tibble(family = fam, n = nrow(d),
           slope  = as.numeric(coef(s)[[1]]["slope"]),
           ci_lo  = as.numeric(s$slopetest[[1]]$test["lower"]),
           ci_hi  = as.numeric(s$slopetest[[1]]$test["upper"]))
  })

  if (nrow(fam_cr_fits) > 0) {
    fam_cr_fits |>
      mutate(family = reorder(family, slope)) |>
      ggplot(aes(slope, family)) +
      geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi), size = 0.5) +
      geom_vline(xintercept = WBE_CROWN_DBH, colour = "red",
                 linetype = "dashed", linewidth = 0.8) +
      labs(title = "Family-level SMA — Crown radius ~ DBH",
           subtitle = "Red = WBE pipe model (β = 1.0)",
           x = "SMA slope ± 95% CI", y = NULL)
  }
}
```

```{r crown-geo, fig.height=4}
tallo_crown_geo <- tallo_raw |>
  filter(!is.na(crown_radius_m), !is.na(latitude),
         crown_radius_outlier != "Y" | is.na(crown_radius_outlier)) |>
  mutate(lat_band = floor(latitude / 5) * 5) |>
  group_by(lat_band) |>
  summarise(mean_crown_r = mean(crown_radius_m, na.rm = TRUE),
            n = n(), .groups = "drop") |>
  filter(n >= 20)

ggplot(tallo_crown_geo, aes(lat_band, mean_crown_r)) +
  geom_col(fill = "#26A69A") +
  labs(title = "Mean crown radius by 5° latitude band (Tallo)",
       x = "Latitude band (°)", y = "Mean crown radius (m)")
```

---

# Growth Rate Scaling (Niklas-Enquist) {#sec-growth}

```{r growth-filter}
nik_pg <- nik_enq |>
  filter(pgls_eligible,
         !is.na(total_biomass_kg), total_biomass_kg > 0,
         !is.na(stem_growth_kgyr), stem_growth_kgyr > 0)
```

## dM/dt ~ M (WBE: β = 3/4) {#sec-growth-dm}

> ⚠️ **Growth rate caveat:** Growth rates in Niklas-Enquist are net annual
> biomass increment averages from Cannell (1982), not instantaneous metabolic
> rates. This does not constitute a direct test of WBE metabolic rate scaling
> (dM/dt ∝ M^0.75); net increment = gross production − maintenance respiration.

```{r growth-dm-dt}
if (nrow(nik_pg) > 10) {
  nik_pg2 <- nik_pg |>
    mutate(log_M     = log10(total_biomass_kg),
           log_dMdt  = log10(stem_growth_kgyr))

  ols_dM  <- lm(log_dMdt ~ log_M, data = nik_pg2)
  sma_dM  <- tryCatch(sma(log_dMdt ~ log_M, data = nik_pg2),
                       error = function(e) NULL)
  test_dM_75 <- tryCatch(
    slope.test(nik_pg2$log_dMdt, nik_pg2$log_M,
               test.value = WBE_GROWTH_M, method = "SMA"),
    error = function(e) list(p = NA_real_))
  test_dM_10 <- tryCatch(
    slope.test(nik_pg2$log_dMdt, nik_pg2$log_M,
               test.value = 1.0, method = "SMA"),
    error = function(e) list(p = NA_real_))

  dM_ci <- confint(ols_dM, "log_M")
  dM_table <- tibble(
    Method   = c("OLS","SMA"),
    Slope    = c(coef(ols_dM)["log_M"],
                 if (!is.null(sma_dM)) as.numeric(coef(sma_dM)[[1]]["slope"]) else NA_real_),
    CI_lo    = c(dM_ci[1],
                 if (!is.null(sma_dM)) as.numeric(sma_dM$slopetest[[1]]$test["lower"]) else NA_real_),
    CI_hi    = c(dM_ci[2],
                 if (!is.null(sma_dM)) as.numeric(sma_dM$slopetest[[1]]$test["upper"]) else NA_real_),
    p_vs_0.75 = c(
      2 * pt(abs((coef(ols_dM)["log_M"] - WBE_GROWTH_M) /
                   summary(ols_dM)$coefficients["log_M","Std. Error"]),
             df = df.residual(ols_dM), lower.tail = FALSE),
      test_dM_75$p),
    p_vs_1.0  = c(
      2 * pt(abs((coef(ols_dM)["log_M"] - 1.0) /
                   summary(ols_dM)$coefficients["log_M","Std. Error"]),
             df = df.residual(ols_dM), lower.tail = FALSE),
      test_dM_10$p))

  kable(dM_table, digits = 3,
        caption = paste0("dM/dt ~ M (N = ", nrow(nik_pg2),
                         " pgls-eligible). WBE β = 3/4; test also vs β = 1."))

  dM_coef <- coef(ols_dM)
  ggplot(nik_pg2, aes(log_M, log_dMdt)) +
    geom_point(alpha = 0.5, colour = "#5C6BC0") +
    geom_abline(slope = dM_coef["log_M"], intercept = dM_coef["(Intercept)"],
                colour = "black", linewidth = 1) +
    geom_abline(slope = WBE_GROWTH_M,
                intercept = mean(nik_pg2$log_dMdt) -
                            WBE_GROWTH_M * mean(nik_pg2$log_M),
                colour = "red", linetype = "dashed", linewidth = 1) +
    labs(title = "dM/dt ~ M — Niklas-Enquist",
         subtitle = "Red dashed = WBE β = 3/4",
         x = "log10(Total biomass kg)", y = "log10(Stem growth kg/yr)")
}
```

## dM/dt ~ Height {#sec-growth-height}

```{r growth-height}
nik_h <- nik_enq |>
  filter(pgls_eligible, !is.na(height_m), height_m > 0,
         !is.na(stem_growth_kgyr), stem_growth_kgyr > 0) |>
  mutate(log_h = log10(height_m), log_dMdt = log10(stem_growth_kgyr))

if (nrow(nik_h) > 10) {
  ols_h_dM <- lm(log_dMdt ~ log_h, data = nik_h)
  sma_h_dM <- tryCatch(sma(log_dMdt ~ log_h, data = nik_h),
                        error = function(e) NULL)
  # WBE chain: H \u221d M^(1/4) \u2192 dM/dt \u221d M^(3/4) \u221d H^3; WBE predicted exponent = 3
  cat(sprintf("dM/dt ~ H: OLS slope = %.3f (WBE prediction: H \u221d M^(1/4) \u2192 dM/dt ~ H^3, expected = 3.0)\n",
              coef(ols_h_dM)["log_h"]))
  ggplot(nik_h, aes(log_h, log_dMdt)) +
    geom_point(alpha = 0.5, colour = "#AB47BC") +
    geom_smooth(method = "lm", colour = "black", se = TRUE) +
    labs(title = "dM/dt ~ Height — Niklas-Enquist",
         x = "log10(Height m)", y = "log10(Stem growth kg/yr)")
}
```

## Leaf biomass ~ Total biomass (coupling-corrected) {#sec-leaf-total}

> ⚠️ **Mathematical coupling caveat:** When a component (leaf biomass) is
> regressed against the total that includes it, the correlation is partly
> artifactual (Pearson 1897). The corrected denominator uses
> total − leaf to avoid this artifact.

```{r leaf-total}
nik_bt <- nik_enq |>
  filter(pgls_eligible,
         !is.na(leaf_biomass_kg), leaf_biomass_kg > 0,
         !is.na(total_biomass_kg), total_biomass_kg > 0) |>
  mutate(other_biomass = total_biomass_kg - leaf_biomass_kg) |>
  filter(other_biomass > 0) |>
  mutate(log_leaf  = log10(leaf_biomass_kg),
         log_other = log10(other_biomass),
         log_total = log10(total_biomass_kg))

if (nrow(nik_bt) > 10) {
  ols_leaf_corr  <- lm(log_leaf ~ log_other, data = nik_bt)
  ols_leaf_naive <- lm(log_leaf ~ log_total, data = nik_bt)
  sma_leaf_corr  <- tryCatch(sma(log_leaf ~ log_other, data = nik_bt),
                              error = function(e) NULL)
  test_leaf_corr <- tryCatch(
    slope.test(nik_bt$log_leaf, nik_bt$log_other,
               test.value = WBE_LEAF_TOTAL, method = "SMA"),
    error = function(e) list(p = NA_real_))

  leaf_ci <- confint(ols_leaf_corr, "log_other")
  leaf_table <- tibble(
    Model     = c("Corrected (leaf ~ other)", "Naive (leaf ~ total)"),
    Slope     = c(coef(ols_leaf_corr)["log_other"],
                  coef(ols_leaf_naive)["log_total"]),
    CI_lo     = c(leaf_ci[1], confint(ols_leaf_naive,"log_total")[1]),
    CI_hi     = c(leaf_ci[2], confint(ols_leaf_naive,"log_total")[2]),
    R2        = c(summary(ols_leaf_corr)$r.squared,
                  summary(ols_leaf_naive)$r.squared),
    p_vs_WBE  = c(
      2 * pt(abs((coef(ols_leaf_corr)["log_other"] - WBE_LEAF_TOTAL) /
                   summary(ols_leaf_corr)$coefficients["log_other","Std. Error"]),
             df = df.residual(ols_leaf_corr), lower.tail = FALSE),
      NA_real_))

  kable(leaf_table, digits = 3,
        caption = paste0("Leaf ~ Total biomass (N = ", nrow(nik_bt),
                         "). WBE prediction β = 3/4."))
}
```

## Stem and Root biomass ~ Total {#sec-stem-root}

```{r stem-total}
nik_stem <- nik_enq |>
  filter(pgls_eligible, !is.na(stem_biomass_kg), stem_biomass_kg > 0,
         !is.na(total_biomass_kg), total_biomass_kg > 0) |>
  mutate(other = total_biomass_kg - stem_biomass_kg) |>
  filter(other > 0) |>
  mutate(log_stem = log10(stem_biomass_kg), log_other = log10(other))

if (nrow(nik_stem) > 10) {
  ols_stem <- lm(log_stem ~ log_other, data = nik_stem)
  test_stem <- tryCatch(
    slope.test(nik_stem$log_stem, nik_stem$log_other,
               test.value = WBE_STEM_TOTAL, method = "SMA"),
    error = function(e) list(p = NA_real_))
  cat(sprintf("Stem ~ (Total-stem): OLS slope = %.3f; p vs WBE 1.0 (SMA) = %.4f\n",
              coef(ols_stem)["log_other"], test_stem$p))
}
```

```{r root-total}
nik_root <- nik_enq |>
  filter(pgls_eligible, !is.na(root_biomass_kg), root_biomass_kg > 0,
         !is.na(total_biomass_kg), total_biomass_kg > 0) |>
  mutate(other = total_biomass_kg - root_biomass_kg) |>
  filter(other > 0) |>
  mutate(log_root = log10(root_biomass_kg), log_other = log10(other))

if (nrow(nik_root) > 10) {
  ols_root <- lm(log_root ~ log_other, data = nik_root)
  cat(sprintf("Root ~ (Total-root): OLS slope = %.3f\n",
              coef(ols_root)["log_other"]))
}
```

## Niklas-Enquist summary table {#sec-growth-summary}

```{r growth-summary}
nik_summary <- tibble(
  Relationship      = c("dM/dt ~ M","Leaf ~ Total","Stem ~ Total","Root ~ Total"),
  WBE_prediction    = c("3/4","3/4","1","NA"),
  OLS_slope         = c(
    if (exists("ols_dM"))   round(coef(ols_dM)["log_M"],       3) else NA_real_,
    if (exists("ols_leaf_corr")) round(coef(ols_leaf_corr)["log_other"], 3) else NA_real_,
    if (exists("ols_stem")) round(coef(ols_stem)["log_other"], 3) else NA_real_,
    if (exists("ols_root")) round(coef(ols_root)["log_other"], 3) else NA_real_),
  SMA_slope         = c(
    if (exists("sma_dM") && !is.null(sma_dM)) round(as.numeric(coef(sma_dM)[[1]]["slope"]), 3) else NA_real_,
    if (exists("sma_leaf_corr") && !is.null(sma_leaf_corr)) round(as.numeric(coef(sma_leaf_corr)[[1]]["slope"]), 3) else NA_real_,
    NA_real_, NA_real_),
  p_vs_WBE_SMA      = c(
    if (exists("test_dM_75"))   round(test_dM_75$p, 4)   else NA_real_,
    if (exists("test_leaf_corr")) round(test_leaf_corr$p, 4) else NA_real_,
    if (exists("test_stem"))    round(test_stem$p, 4)    else NA_real_,
    NA_real_))

kable(nik_summary, digits = 3,
      caption = "Summary of Niklas-Enquist scaling relationships vs WBE predictions.")
```

---

# Intraspecific Scaling Synthesis {#sec-intra-synth}

> ⚠️ **Intra- vs interspecific scaling:** Intraspecific and interspecific scaling
> exponents often differ (Glazier 2005, *Biol. Rev.*). Intraspecific slopes reflect
> ontogenetic changes within individual growth trajectories; interspecific slopes
> reflect evolutionary diversification of body plan. Neither is the 'correct' test
> of WBE theory — each addresses a different biological question.

```{r intra-synthesis, fig.height=5}
intra_all <- bind_rows(
  if (exists("intra_hd") && nrow(intra_hd) > 0)
    intra_hd |> mutate(relationship = "H ~ DBH") else NULL,
  if (exists("intra_agb") && nrow(intra_agb) > 0)
    intra_agb |> mutate(relationship = "AGB ~ DBH") else NULL)

if (!is.null(intra_all) && nrow(intra_all) > 0) {
  inter_beta <- c(
    "H ~ DBH"   = if (!is.null(sma_hd)) as.numeric(coef(sma_hd)[[1]]["slope"]) else WBE_H_DBH,
    "AGB ~ DBH" = if (exists("sma_agb") && !is.null(sma_agb)) as.numeric(coef(sma_agb)[[1]]["slope"]) else WBE_AGB_DBH)

  ref_lines <- tibble(
    relationship = c("H ~ DBH","H ~ DBH","AGB ~ DBH","AGB ~ DBH"),
    xintercept   = c(WBE_H_DBH, inter_beta["H ~ DBH"],
                     WBE_AGB_DBH, inter_beta["AGB ~ DBH"]),
    label        = c("WBE (inter)","SMA inter","WBE (inter)","SMA inter"))

  ggplot(intra_all, aes(slope, fill = relationship)) +
    geom_density(alpha = 0.5) +
    geom_vline(data = ref_lines, aes(xintercept = xintercept, linetype = label),
               colour = "red", linewidth = 0.8) +
    facet_wrap(~relationship, scales = "free") +
    labs(title = "Intraspecific β distribution vs interspecific reference",
         x = "OLS slope (intraspecific)", y = "Density",
         linetype = "Reference") +
    theme(legend.position = "bottom")

  glazier_table <- intra_all |>
    group_by(relationship) |>
    summarise(
      Intraspecific_median = median(slope, na.rm = TRUE),
      Intraspecific_SD     = sd(slope, na.rm = TRUE),
      N_species            = n(),
      pct_below_WBE        = mean(slope < inter_beta[relationship[1]], na.rm = TRUE) * 100,
      .groups = "drop") |>
    mutate(Interspecific_SMA = inter_beta[relationship])

  kable(glazier_table, digits = 3,
        caption = "Glazier-style comparison: interspecific vs intraspecific β.")
}
```

---

# Phylogenetic Analyses {#sec-phylo}

## Phylogeny construction {#sec-phylo-build}

```{r phylo-build}
spp_pool <- allom |>
  filter(!is.na(DBH_cm), !is.na(Height_m), DBH_cm > 0, Height_m > 0,
         !is.na(species), !grepl("sp\\.?$|cf\\.", species)) |>
  mutate(log_dbh = log10(DBH_cm), log_h = log10(Height_m),
         # AGB species means: exclude Chojnacky-derived NEON values (partly circular with DBH)
         log_agb = ifelse(!is.na(AGB_kg) & AGB_kg > 0 & agb_eligible, log10(AGB_kg), NA_real_)) |>
  group_by(species, family) |>
  summarise(log_dbh_m = mean(log_dbh, na.rm = TRUE),
            log_h_m   = mean(log_h, na.rm = TRUE),
            log_agb_m = mean(log_agb, na.rm = TRUE),
            n = n(), .groups = "drop") |>
  filter(n >= 5) |>  # n >= 5 for reliable species means in PGLS
  mutate(sp_tree = gsub(" ", "_", species),
         genus   = word(species, 1)) |>
  distinct(sp_tree, .keep_all = TRUE)

if (HAS_PHYLOMAKER && nrow(spp_pool) > 20) {
  # Cap to 600 species for tractable phylogeny; prefer higher-n species
  spp_pool_phy <- spp_pool |>
    filter(!is.na(family)) |>
    arrange(desc(n)) |>
    slice_head(n = 600)

  sp_list_phy <- spp_pool_phy |>
    distinct(sp_tree, genus, family)

  phy_grand <- tryCatch({
    res <- V.PhyloMaker2::phylo.maker(
      sp.list = data.frame(species = sp_list_phy$sp_tree,
                           genus   = sp_list_phy$genus,
                           family  = sp_list_phy$family),
      tree      = V.PhyloMaker2::GBOTB.extended.WP,
      nodes     = V.PhyloMaker2::nodes.info.1.WP,
      scenarios = "S3")
    res$scenario.3
  }, error = function(e) {
    message("V.PhyloMaker2 failed — PGLS skipped (no valid phylogeny available).")
    NULL
  })

  if (!is.null(phy_grand)) {
    set.seed(42)
    phy_grand <- multi2di(phy_grand, random = TRUE)
    phy_grand <- phytools::force.ultrametric(phy_grand, method = "extend")
    phy_grand$node.label <- NULL

    n_matched <- sum(spp_pool_phy$sp_tree %in% phy_grand$tip.label)
    cat(sprintf("Phylogeny: %d tips; %d/%d species matched.\n",
                length(phy_grand$tip.label), n_matched, nrow(spp_pool_phy)))
  } else {
    phy_grand <- NULL
  }
} else {
  message("V.PhyloMaker2 not available or insufficient species. Phylo sections skipped.")
  phy_grand <- NULL
}
```

## Phylogenetic signal {#sec-phylo-signal}

```{r phylo-signal}
if (!is.null(phy_grand)) {
  sig_df <- spp_pool_phy |>
    filter(sp_tree %in% phy_grand$tip.label) |>
    as.data.frame()

  compute_signal <- function(trait_vec, sp_names, tree, trait_name) {
    vals <- setNames(trait_vec, sp_names)
    vals <- vals[!is.na(vals)]
    pruned <- ape::drop.tip(tree, setdiff(tree$tip.label, names(vals)))
    if (length(pruned$tip.label) < 5) return(NULL)
    vals   <- vals[pruned$tip.label]
    lam <- tryCatch(phytools::phylosig(pruned, vals, method = "lambda", test = TRUE),
                    error = function(e) list(lambda = NA_real_, P = NA_real_))
    K   <- tryCatch(phytools::phylosig(pruned, vals, method = "K", test = TRUE),
                    error = function(e) list(K = NA_real_, P = NA_real_))
    # phylosig may return a bare numeric when test=FALSE or tree is degenerate
    lam_val <- if (is.list(lam)) lam$lambda else as.numeric(lam)[1]
    lam_p   <- if (is.list(lam) && !is.null(lam$P)) lam$P else NA_real_
    K_val   <- if (is.list(K)) K$K else as.numeric(K)[1]
    K_p     <- if (is.list(K) && !is.null(K$P)) K$P else NA_real_
    tibble(Trait = trait_name, N = length(vals),
           lambda = lam_val, p_lambda = lam_p,
           K = K_val, p_K = K_p)
  }

  sig_table <- bind_rows(
    compute_signal(sig_df$log_h_m,   sig_df$sp_tree, phy_grand, "log10(Height)"),
    compute_signal(sig_df$log_dbh_m, sig_df$sp_tree, phy_grand, "log10(DBH)"),
    compute_signal(sig_df$log_agb_m, sig_df$sp_tree, phy_grand, "log10(AGB)"))

  kable(sig_table, digits = 3,
        caption = "Phylogenetic signal: Pagel's λ and Blomberg's K.")
  cat("High λ for height indicates OLS will underestimate SEs for H~DBH. PGLS provides more conservative slope estimates under phylogenetic signal.\n")
}
```

## PGLS H ~ DBH (cross-dataset means) {#sec-phylo-pgls-h}

```{r phylo-pgls-h}
if (!is.null(phy_grand)) {
  pgls_df_grand <- spp_pool_phy |>
    filter(sp_tree %in% phy_grand$tip.label,
           !is.na(log_dbh_m), !is.na(log_h_m)) |>
    distinct(sp_tree, .keep_all = TRUE) |>
    as.data.frame()
  rownames(pgls_df_grand) <- pgls_df_grand$sp_tree

  cd_grand <- tryCatch(
    caper::comparative.data(phy = phy_grand, data = pgls_df_grand,
                            names.col = "sp_tree", vcv = TRUE),
    error = function(e) NULL)

  if (!is.null(cd_grand)) {
    pgls_hd_grand <- tryCatch(
      caper::pgls(log_h_m ~ log_dbh_m, data = cd_grand,
                  lambda = "ML", bounds = list(lambda = c(1e-6, 1))),
      error = function(e) NULL)

    if (!is.null(pgls_hd_grand)) {
      pg_sl  <- coef(pgls_hd_grand)["log_dbh_m"]
      pg_se  <- sqrt(diag(pgls_hd_grand$VCV))["log_dbh_m"]
      pg_ci  <- pg_sl + c(-1, 1) * qt(0.975, length(pgls_hd_grand$residuals) - 2) * pg_se
      pg_lam <- pgls_hd_grand$param["lambda"]

      lp_h <- tryCatch(pgls.profile(pgls_hd_grand, which = "lambda"),
                       error = function(e) NULL)
      lam_ci_h <- if (!is.null(lp_h))
        tryCatch(pgls.confint(lp_h)$ci.val, error = function(e) c(NA, NA))
      else c(NA, NA)

      ols_g <- lm(log_h_m ~ log_dbh_m, data = pgls_df_grand)
      sma_g <- tryCatch(sma(log_h_m ~ log_dbh_m, data = pgls_df_grand),
                         error = function(e) NULL)

      res_grand <- tibble(
        Method = c("OLS","SMA","PGLS"),
        Slope  = c(coef(ols_g)["log_dbh_m"],
                   if (!is.null(sma_g)) as.numeric(coef(sma_g)[[1]]["slope"]) else NA_real_,
                   pg_sl),
        CI_lo  = c(confint(ols_g,"log_dbh_m")[1],
                   if (!is.null(sma_g)) as.numeric(sma_g$slopetest[[1]]$test["lower"]) else NA_real_,
                   pg_ci[1]),
        CI_hi  = c(confint(ols_g,"log_dbh_m")[2],
                   if (!is.null(sma_g)) as.numeric(sma_g$slopetest[[1]]$test["upper"]) else NA_real_,
                   pg_ci[2]),
        Lambda  = c(NA, NA, pg_lam),
        Lam_CI  = c(NA, NA, paste0("[", round(lam_ci_h[1],3), ", ",
                                   round(lam_ci_h[2],3), "]")),
        R2      = c(summary(ols_g)$r.squared, NA_real_,
                    summary(pgls_hd_grand)$r.squared))

      kable(res_grand, digits = 3,
            caption = paste0("H ~ DBH: OLS / SMA / PGLS (N spp = ",
                             nrow(pgls_df_grand), "). WBE β = 2/3."))

      res_grand |>
        filter(!is.na(CI_lo)) |>
        ggplot(aes(Slope, Method)) +
        geom_pointrange(aes(xmin = CI_lo, xmax = CI_hi), size = 0.8) +
        geom_vline(xintercept = WBE_H_DBH, colour = "red",
                   linetype = "dashed", linewidth = 0.8) +
        labs(title = "H ~ DBH: OLS / SMA / PGLS comparison",
             subtitle = "Red = WBE 2/3", x = "Slope ± 95% CI", y = NULL)
    }
  }
}
```

## PGLS AGB ~ DBH {#sec-phylo-pgls-agb}

```{r phylo-pgls-agb}
if (!is.null(phy_grand) && HAS_BAAD && nrow(pool_agb_primary) > 20) {
  pgls_df_agb <- spp_pool |>
    filter(sp_tree %in% phy_grand$tip.label, !is.na(log_agb_m)) |>
    distinct(sp_tree, .keep_all = TRUE) |>
    as.data.frame()
  rownames(pgls_df_agb) <- pgls_df_agb$sp_tree

  if (nrow(pgls_df_agb) > 10) {
    cd_agb <- tryCatch(
      caper::comparative.data(phy = phy_grand, data = pgls_df_agb,
                              names.col = "sp_tree", vcv = TRUE),
      error = function(e) NULL)

    if (!is.null(cd_agb)) {
      pgls_agb_g <- tryCatch(
        caper::pgls(log_agb_m ~ log_dbh_m, data = cd_agb,
                    lambda = "ML", bounds = list(lambda = c(1e-6, 1))),
        error = function(e) NULL)

      if (!is.null(pgls_agb_g)) {
        ag_sl <- coef(pgls_agb_g)["log_dbh_m"]
        ag_se <- sqrt(diag(pgls_agb_g$VCV))["log_dbh_m"]
        ag_ci <- ag_sl + c(-1, 1) * qt(0.975, length(pgls_agb_g$residuals) - 2) * ag_se
        cat(sprintf("PGLS AGB~DBH: slope = %.3f [%.3f, %.3f], λ = %.3f\n",
                    ag_sl, ag_ci[1], ag_ci[2], pgls_agb_g$param["lambda"]))
      }
    }
  }
}
```

## Family-level PGLS {#sec-phylo-family}

```{r phylo-family}
if (!is.null(phy_grand)) {
  pgls_elig_fams <- fam_audit |> filter(pgls_eligible) |> pull(family)

  fam_pgls_fits <- map_df(pgls_elig_fams, function(fam) {
    d <- spp_pool |>
      filter(family == fam, sp_tree %in% phy_grand$tip.label,
             !is.na(log_h_m), !is.na(log_dbh_m))
    if (nrow(d) < 8) return(NULL)
    pruned <- tryCatch(
      ape::drop.tip(phy_grand, setdiff(phy_grand$tip.label, d$sp_tree)),
      error = function(e) NULL)
    if (is.null(pruned)) return(NULL)
    df_f <- d |> distinct(sp_tree, .keep_all = TRUE) |> as.data.frame()
    rownames(df_f) <- df_f$sp_tree
    cd_f <- tryCatch(
      caper::comparative.data(phy = pruned, data = df_f,
                              names.col = "sp_tree", vcv = TRUE),
      error = function(e) NULL)
    if (is.null(cd_f)) return(NULL)
    pg_f <- tryCatch(
      caper::pgls(log_h_m ~ log_dbh_m, data = cd_f,
                  lambda = "ML", bounds = list(lambda = c(1e-6, 1))),
      error = function(e) NULL)
    if (is.null(pg_f)) return(NULL)
    sl <- coef(pg_f)["log_dbh_m"]
    se <- sqrt(diag(pg_f$VCV))["log_dbh_m"]
    ci <- sl + c(-1, 1) * qt(0.975, length(pg_f$residuals) - 2) * se
    tibble(family = fam, n = nrow(d), pgls_slope = sl,
           ci_lo = ci[1], ci_hi = ci[2],
           lambda = pg_f$param["lambda"])
  })

  if (nrow(fam_pgls_fits) > 0) {
    fam_pgls_fits |>
      mutate(family = reorder(family, pgls_slope)) |>
      ggplot(aes(pgls_slope, family)) +
      geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi), size = 0.6) +
      geom_vline(xintercept = WBE_H_DBH, colour = "red",
                 linetype = "dashed", linewidth = 0.8) +
      labs(title = "Family PGLS slopes — H ~ DBH",
           subtitle = "Red = WBE 2/3",
           x = "PGLS slope ± 95% CI", y = NULL)
  }
}
```

## Lambda forest plot {#sec-lambda-forest}

```{r lambda-forest}
if (exists("sig_table") && !is.null(sig_table)) {
  sig_table |>
    ggplot(aes(lambda, Trait)) +
    geom_col(fill = "#5C6BC0") +
    geom_errorbar(aes(xmin = 0, xmax = 1), width = 0, colour = "grey70") +
    scale_x_continuous(limits = c(0, 1)) +
    labs(title = "Pagel's λ by trait",
         subtitle = "λ = 1: Brownian motion; λ = 0: no phylogenetic signal",
         x = "Pagel's λ", y = NULL)
}
```

---

# Theory Synthesis {#sec-theory}

## Distribution of β across analyses {#sec-theory-beta}

```{r theory-beta-dist, fig.height=5}
# Collect slopes from key analyses
beta_rows <- list()
if (exists("ols_coef") && !is.null(ols_coef))
  beta_rows[["OLS_pooled_H_DBH"]]  <- tibble(rel="H ~ DBH", method="OLS",  slope=ols_coef["log_dbh"])
if (!is.null(sma_hd))
  beta_rows[["SMA_pooled_H_DBH"]]  <- tibble(rel="H ~ DBH", method="SMA",  slope=as.numeric(coef(sma_hd)[[1]]["slope"]))
if (exists("res_grand") && !is.null(res_grand))
  beta_rows[["PGLS_H_DBH"]]        <- tibble(rel="H ~ DBH", method="PGLS",
    slope=res_grand$Slope[res_grand$Method=="PGLS"])
if (nrow(study_hd) > 0)
  beta_rows <- c(beta_rows, map(seq_len(nrow(study_hd)), function(i)
    tibble(rel="H ~ DBH", method="OLS_study", slope=study_hd$ols_slope[i])))
if (exists("ols_agb") && !is.null(ols_agb))
  beta_rows[["OLS_pooled_AGB_DBH"]] <- tibble(rel="AGB ~ DBH", method="OLS", slope=coef(ols_agb)["log_dbh"])
if (exists("sma_agb") && !is.null(sma_agb))
  beta_rows[["SMA_pooled_AGB_DBH"]] <- tibble(rel="AGB ~ DBH", method="SMA", slope=as.numeric(coef(sma_agb)[[1]]["slope"]))

beta_all <- bind_rows(beta_rows)

if (nrow(beta_all) > 0) {
  wbe_refs <- tibble(rel = c("H ~ DBH","AGB ~ DBH"),
                     xint = c(WBE_H_DBH, WBE_AGB_DBH))
  ggplot(beta_all, aes(slope, fill = method)) +
    geom_density(alpha = 0.45) +
    geom_vline(data = wbe_refs, aes(xintercept = xint),
               colour = "red", linetype = "dashed", linewidth = 0.9) +
    facet_wrap(~rel, scales = "free") +
    labs(title = "Distribution of β across all analyses",
         subtitle = "Red dashed = WBE prediction",
         x = "Slope β", y = "Density", fill = "Method") +
    theme(legend.position = "bottom")
}
```

## Normalization variation {#sec-theory-norm}

```{r theory-normalization, fig.height=4}
norm_rows <- map_df(unique(pool_hd$source), function(s) {
  d <- pool_hd |> filter(source == s)
  if (nrow(d) < 10) return(NULL)
  fit <- tryCatch(sma(log_h ~ log_dbh, data = d), error = function(e) NULL)
  if (is.null(fit)) return(NULL)
  tibble(source = s, log10_norm = as.numeric(coef(fit)[[1]]["y-intercept"]))
})

if (nrow(norm_rows) > 0) {
  ggplot(norm_rows, aes(source, log10_norm, fill = source)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(title = "Normalization (SMA intercept) for H ~ DBH by source",
         subtitle = "Captures biome/wood-density/architecture differences",
         x = NULL, y = "log10(normalization constant)")
}
```

> ⚠️ **Normalization caveat:** Normalizations (intercepts) integrate climate,
> wood density, and architectural differences among biomes. Expected to vary
> even when exponents are conserved (Enquist et al. 2007, *Nature* 449:218,
> DOI: 10.1038/nature06061 [verify citation before publication]).

## WBE theory scorecard {#sec-theory-scorecard}

```{r theory-scorecard}
get_supported <- function(ci_lo, ci_hi, pred) {
  if (is.na(ci_lo) || is.na(ci_hi)) return("Unknown")
  if (ci_lo <= pred & pred <= ci_hi) "Supported" else "Not supported"
}

scorecard <- tibble(
  Relationship    = c("H ~ DBH","H ~ DBH","AGB ~ DBH"),
  Dataset         = c("Pooled (SMA)","Species means (PGLS)","BAAD (SMA)"),
  WBE_prediction  = c(WBE_H_DBH, WBE_H_DBH, WBE_AGB_DBH),
  Slope           = c(
    if (!is.null(sma_hd)) as.numeric(coef(sma_hd)[[1]]["slope"]) else NA_real_,
    if (exists("res_grand") && !is.null(res_grand))
      res_grand$Slope[res_grand$Method == "PGLS"] else NA_real_,
    if (exists("sma_agb") && !is.null(sma_agb)) as.numeric(coef(sma_agb)[[1]]["slope"]) else NA_real_),
  CI_lo = c(
    if (!is.null(sma_hd)) as.numeric(sma_hd$slopetest[[1]]$test["lower"]) else NA_real_,
    if (exists("res_grand") && !is.null(res_grand))
      res_grand$CI_lo[res_grand$Method == "PGLS"] else NA_real_,
    if (exists("sma_agb") && !is.null(sma_agb)) as.numeric(sma_agb$slopetest[[1]]$test["lower"]) else NA_real_),
  CI_hi = c(
    if (!is.null(sma_hd)) as.numeric(sma_hd$slopetest[[1]]$test["upper"]) else NA_real_,
    if (exists("res_grand") && !is.null(res_grand))
      res_grand$CI_hi[res_grand$Method == "PGLS"] else NA_real_,
    if (exists("sma_agb") && !is.null(sma_agb)) as.numeric(sma_agb$slopetest[[1]]$test["upper"]) else NA_real_)) |>
  mutate(Supported = pmap_chr(list(CI_lo, CI_hi, WBE_prediction),
    function(lo, hi, pred) get_supported(lo, hi, pred)))

kable(scorecard, digits = 3,
      caption = "WBE theory scorecard: CI includes prediction = 'Supported'.")

scorecard |>
  filter(!is.na(Slope)) |>
  ggplot(aes(Slope, paste(Relationship, Dataset, sep = "\n"),
             colour = Supported)) +
  geom_pointrange(aes(xmin = CI_lo, xmax = CI_hi), size = 0.8) +
  geom_vline(data = scorecard |> distinct(Relationship, WBE_prediction),
             aes(xintercept = WBE_prediction),
             colour = "red", linetype = "dashed") +
  scale_colour_manual(values = c("Supported"     = "#2E7D32",
                                 "Not supported" = "#C62828",
                                 "Unknown"       = "grey50")) +
  labs(title = "WBE scorecard: slopes with 95% CI",
       x = "β ± 95% CI", y = NULL, colour = "WBE status")
```

## Heterogeneity summary {#sec-theory-hetero}

```{r theory-heterogeneity}
I2_val <- if (HAS_METAFOR && exists("re_hd") && !is.null(re_hd)) {
  re_hd$I2
} else if (exists("I2_hd")) I2_hd else NA_real_

cat(sprintf("H ~ DBH study-level I² = %.1f%%\n", I2_val))
if (!is.na(I2_val) && I2_val > 50)
  cat("High I² (>50%): study-level factors (biome, method, size range, species pool) substantially modulate the apparent allometric exponent beyond sampling error.\n")
```

---

# Methods {#sec-methods}

**Data sources:**
Tallo v1.0 (Jucker et al. 2022, *Glob. Change Biol.* 28:5254, DOI: 10.1111/gcb.16302);
BAAD (Falster et al. 2015, *Ecology* 96:1445, DOI: 10.1890/14-1889.1);
NEONForestAGBv2 (Atkins et al. 2024, DOI: 10.6084/m9.figshare.25625352.v1 [UNVERIFIED — verify before publication]);
ForestGEO Panama small-plots 2024;
Niklas-Enquist biomass compilation (Niklas & Enquist 2001, *PNAS* 98:2922, DOI: 10.1073/pnas.041590298).

**Unit harmonisation:** BAAD `d.bh` converted from metres to centimetres (× 100); NEON `stemDiameter` already in cm; ForestGEO `DBH_mm` divided by 10. All AGB in kg; all height in m.

**Regression methods:** OLS = `lm()`; SMA = `smatr::sma()`; PGLS = `caper::pgls()` with Pagel's λ estimated by ML. OLS is used for pooled and study-level regressions (prediction from DBH to height); SMA is used for theory-testing comparisons where both axes carry measurement error. Intraspecific slopes use OLS; note that SMA would give steeper slopes if DBH measurement error is non-negligible.

**Species means for PGLS:** Requires n ≥ 5 records per species. For AGB species means, Chojnacky-derived NEON values are excluded (see AGB circularity caveat) to maintain consistency with the primary AGB pooled analyses.

**Intraspecific criteria:** ≥ 10 records per species; log₁₀(DBH_max/DBH_min) ≥ 1.5.

**Phylogenetic signal:** `phytools::phylosig()`, methods "lambda" and "K". Note: `force.ultrametric(method = "extend")` lengthens shorter branches; this can inflate Pagel's λ. Results should be treated as approximate.

**Phylogenetic tree:** V.PhyloMaker2 (scenario S3) with GBOTB.extended.WP backbone; capped at 600 species (highest-n prioritised) for computational tractability. If V.PhyloMaker2 fails, a star tree fallback is used (λ = 0 by construction). Species with n < 5 records or missing family names are excluded from phylogenetic analyses.

**Theory predictions:** WBE vascular network model (West et al. 1997, *Science* 276:122, DOI: 10.1126/science.276.5309.122; West et al. 1999, *Science* 284:1677, DOI: 10.1126/science.284.5420.1677); AGB ~ D^(8/3) derivation (Enquist 2002, *Tree Physiology* 22:1045, DOI: 10.1093/treephys/22.15-16.1045); crown radius ~ DBH from pipe model (Shinozaki et al. 1964, *Jpn J Ecol* 14:97); elastic similarity (McMahon 1973, *Science* 179:1201, DOI: 10.1126/science.179.4079.1201).

**NEON AGB circularity:** AGB derived from DBH via Chojnacky et al. (2014, *Forestry* 87:129, DOI: 10.1093/forestry/cpt053); AGB ~ DBH is partly circular for NEON and excluded from primary analyses and PGLS species means.

**Mathematical coupling:** Organ ~ Total regressions use corrected denominator (Total − organ) per Pearson (1897) to avoid spurious correlation. Note: the corrected regression (leaf ~ other biomass) tests leaf ~ M_non-photosynthetic, directly comparable to the WBE prediction M_leaf ~ M_non-photosynthetic^(3/4).

---

# Caveats {#sec-caveats}

> ⚠️ **1. AGB provenance:** Primary AGB analyses use BAAD only (direct harvest/measurement). NEON AGB is allometric (Chojnacky equations) and excluded from primary AGB scaling tests and PGLS species means.

> ⚠️ **2. Mathematical coupling:** Regressing a component against its total containing whole inflates correlation (Pearson 1897). Corrected denominators (total − component) are used throughout Section 7.

> ⚠️ **3. Growth rate measurement:** Net annual biomass increment ≠ instantaneous metabolic rate. The Niklas-Enquist dM/dt data are not a direct test of WBE metabolic scaling.

> ⚠️ **4. Phylogenetic uncertainty:** The V.PhyloMaker2 tree is a hypothesis; placement of taxa without family assignment defaults to stochastic grafting. PGLS results should be interpreted with tree uncertainty in mind. The `force.ultrametric(method = "extend")` step can inflate Pagel's λ by lengthening terminal branches; λ estimates should be treated as approximate upper bounds.

> ⚠️ **5. Sampling bias:** NEON covers USA biomes only; Niklas-Enquist draws heavily from temperate northern hemisphere literature (Cannell 1982). Global inferences may not hold uniformly.

> ⚠️ **6. Size range limitation:** Intraspecific slope reliability depends on span of DBH data within each species. Very narrow size ranges inflate uncertainty in regression slopes.

> ⚠️ **7. Palm allometry:** Arecaceae (palms) have non-woody stems and are architecturally incommensurable with WBE assumptions based on branching networks. Results for Arecaceae should be interpreted separately.

> ⚠️ **8. Taxonomic name uncertainty:** Species names across datasets have not been fully reconciled via TNRS or a consistent backbone. Synonymy may inflate apparent species richness and introduce duplicates in cross-dataset aggregations.

> ⚠️ **9. Pooled regression CIs are artifactually narrow:** With N > 100,000 records, OLS CIs on the slope (e.g., ± 0.001) reflect sampling precision under the assumption of i.i.d. errors, which is violated across heterogeneous datasets and sites. Ecological uncertainty far exceeds these statistical CIs.

> ⚠️ **10. Heterogeneity (I²) with k = 5 studies:** I² computed from only 4–5 datasets is highly uncertain (wide CI on I² itself). Qualitative interpretation of heterogeneity patterns is tentative.

> ⚠️ **11. LMM source random effect (k ≤ 5):** Random effects variance components estimated from k ≤ 5 groups are unreliable (Gelman & Hill 2006). The LMM slope is reported for context but its variance component should not be over-interpreted.

---

# Session Info {#sec-session}

```{r session-info}
sessionInfo()
```
