---
title: "Kurosawa et al. 2025 — Scaling of Shoot and Root Respiration Across Woody and Herbaceous Plants"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5)
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
  library(knitr)
  library(scales)
  library(smatr)
  library(purrr)
  library(stringr)
  library(tibble)
  library(caper)
  library(ape)
  library(phytools)
  library(gridExtra)
})
# Resolve namespace conflict: maps::map (loaded as a transitive dependency of
# ape/phytools) masks purrr::map. Explicitly bind purrr versions so that all
# map() / map_dbl() / map_int() calls in this document use purrr semantics.
map     <- purrr::map
map_dbl <- purrr::map_dbl
map_int <- purrr::map_int
map_chr <- purrr::map_chr
pmap    <- purrr::pmap

# Resolve dplyr masking: caper loads MASS which exports select(); ape loads
# maps. Bind dplyr versions explicitly to prevent silent substitution.
select  <- dplyr::select
filter  <- dplyr::filter
mutate  <- dplyr::mutate
rename  <- dplyr::rename
arrange <- dplyr::arrange
group_by  <- dplyr::group_by
summarise <- dplyr::summarise
summarize <- dplyr::summarize

# V.PhyloMaker2 is not on CRAN for R 4.3 arm64. Load conditionally.
HAS_PHYLOMAKER <- requireNamespace("V.PhyloMaker2", quietly = TRUE)
if (!HAS_PHYLOMAKER) {
  message("V.PhyloMaker2 not available — PGLS phylogeny sections will use a simulated star tree as placeholder.")
}
HAS_CAPER   <- requireNamespace("caper",     quietly = TRUE)
HAS_BRMS    <- requireNamespace("brms",      quietly = TRUE)
theme_set(theme_bw(base_size = 13))
proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

---

# Overview and Data Source

This document analyses the allometric scaling of plant respiration across a broad
size range of **woody and herbaceous plants**, using the dataset compiled by
Kurosawa et al. (2025):

> **Kurosawa Y, Mori S, Ferrio JP, et al.** (2025). Scaling of shoot and root
> respiration of woody and herbaceous plants. *Proceedings of the Royal Society B*,
> **292**: 20241910.
> DOI: [10.1098/rspb.2024.1910](https://doi.org/10.1098/rspb.2024.1910)

The accompanying Dryad dataset is:

> **Kurosawa Y, Mori S, Ferrio JP, et al.** (2024). Data from: Scaling of shoot
> and root respiration of woody and herbaceous plants. *Dryad*.
> DOI: [10.5061/dryad.sxksn03cj](https://doi.org/10.5061/dryad.sxksn03cj)

> **⚠ Data provenance note:** The woody-plant component of this dataset explicitly
> integrates records from three sub-sources:
>
> * **Mori et al. 2010** — *Proceedings of the Royal Society B* 277: 1447–1454.
>   DOI: [10.1098/rspb.2009.1811](https://doi.org/10.1098/rspb.2009.1811)
> * **Kurosawa et al. 2023** — Ontogenetic respiration data for *Fagus crenata*.
>   *Annals of Botany* 131: 313–322.
>   DOI: [10.1093/aob/mcac143](https://doi.org/10.1093/aob/mcac143)
> * New measurements reported in Kurosawa et al. 2025.
>
> All DOIs, journal names, and volume/page details above should be independently
> verified against Crossref before use in any manuscript.

The database contains respiration rates, fresh mass, plant part (shoot/root),
life form (woody/herbaceous), species, and dataset source for each record.

**Key theoretical predictions tested:**

| Scaling relationship | WBE / MST prediction | Reference |
|---|---|---|
| $R \propto M^{\alpha}$ (overall) | $\alpha = 3/4$ | West et al. 1997; Mori et al. 2010 |
| $R \propto M^{\alpha}$ (herbaceous) | $\alpha \leq 3/4$? | Kurosawa et al. 2025 |
| Root vs. shoot slope equality | Slopes equal? | Kurosawa et al. 2025 |

---

# Scaling theory

## WBE fractal network derivation of $R \propto M^{3/4}$

The West–Brown–Enquist (WBE) model (West, Brown & Enquist 1997) derives the
$^3\!/_4$-power metabolic scaling law from three geometric constraints on a
hierarchical vascular network with $N$ branching levels, branching ratio $n$,
and terminal conduit radius $r_c$ and length $l_c$ that are **size-invariant**:

$$r_k \propto n^{-k/2}, \qquad l_k \propto n^{-k/3}, \qquad N_c = n^N$$

where level $k = 0$ is the trunk/main stem and level $k = N$ is the capillary.
The three constraints are:

1. **Area-preserving branching:** at every node the summed daughter cross-section equals the parent cross-section, $n\, r_{k+1}^2 = r_k^2$, giving $r_k = r_c\, N_c^{1/2}\, n^{-k/2}$.
2. **Space-filling fractal geometry:** the network spans the three-dimensional body, requiring segment lengths $l_k = l_c\, N_c^{1/3}\, n^{-k/3}$.
3. **Size-invariant terminal units:** $r_c$ and $l_c$ are universal constants.

Metabolic rate is proportional to the number of terminal units, each supplying a fixed metabolic quantum $b_c$:
$$B = N_c\, b_c \propto N_c = n^N$$

The main-stem (trunk) volume dominates conducting-tissue mass in woody plants (dead xylem = most of woody biomass), and at level $k = 0$:
$$V_0 = \pi r_0^2 l_0 = \pi\bigl(r_c N_c^{1/2}\bigr)^2 \cdot \bigl(l_c N_c^{1/3}\bigr) = \pi r_c^2 l_c\, N_c^{4/3}$$

Therefore $M \propto V_0 \propto N_c^{4/3}$, and inverting:

$$\boxed{B \propto N_c \propto M^{3/4}}$$

The exponent $3/4 = 1/(1 + 1/3)$ arises geometrically: area-preservation (fixing
$r_0 \propto N_c^{1/2}$) and space-filling (fixing $l_0 \propto N_c^{1/3}$) together
make trunk volume grow faster than $N_c^1$, so a given metabolic rate requires
disproportionately less mass at large size.

## Extension to organ-level absolute respiration

Applying $R \propto M^{3/4}$ at the organ (shoot, root) level requires an additional
assumption that the WBE derivation does not itself provide: the organ's **fractional
allocation of total metabolic rate must be size-invariant**. If
$R_{\rm organ} = f\cdot B$ where $f$ is the fraction of whole-plant carbon flux
allocated to that organ, then $R_{\rm organ} \propto M^{3/4}$ holds only when $f$
is constant across the size range. If allocation shifts with size — for example,
if large woody plants invest proportionally less in fine roots (root:shoot ratio
declines with size) — the organ-level exponent will deviate from $^3\!/_4$ even
when the whole-plant exponent does not.

## Competing null hypotheses

Three benchmark exponents span the theoretically motivated hypothesis space:

| Hypothesis | Exponent $\beta$ | Mechanistic basis |
|---|:---:|---|
| Rubner surface-area null | $2/3$ | Body surface $\propto M^{2/3}$ for geometrically similar organisms; heat dissipation or O$_2$/CO$_2$ diffusion limited by surface area (Rubner 1883) |
| WBE / MST prediction | $3/4$ | Fractal vascular network optimisation; area-preserving branching + space-filling + size-invariant terminal units (West et al. 1997; Enquist et al. 1998) |
| Isometric null | $1$ | Constant mass-specific respiration, $r = R/M = \text{const}$; expected if metabolism scales with total cytoplasmic volume without network constraint |

## Key testable prediction: organ $\times$ life-form interaction

WBE's constraints are geometric and apply to any vascular network regardless of
its phylogenetic or morphological realisation. Therefore the theory predicts that
$\hat{\beta}$ should be **identical for shoots and roots** across all life forms:
the exponent is a network property, not a tissue-type property. Two departures
are informative if observed:

- **Shoot $\neq$ Root exponent:** implies the vascular architecture of shoots and
  roots differs in a way that violates one of the three WBE constraints (e.g.
  roots may not area-preserve at lateral junctions).
- **Woody $\hat{\beta} \approx 3/4$ while herbaceous $\hat{\beta} > 3/4$:** would
  suggest that the secondary xylem of woody plants instantiates the fractal
  network more faithfully, while non-woody stems — lacking hierarchical vessel
  size gradients — are governed more by cytoplasmic volume (isometry) or
  surface-area diffusion (Rubner).

These contrasts are tested by the slope-equality analyses in sections below.

---

# Data loading

```{r load-data}
# ── Local file path (Dryad download) ──────────────────────────────────────────
# Download from: https://doi.org/10.5061/dryad.sxksn03cj
# File: data.respiration.csv
data_path <- file.path(proj_dir,
  "data/raw/kurosawa_respiration/doi_10_5061_dryad_sxksn03cj__v20241204/data.respiration.csv")

USE_REAL_DATA <- file.exists(data_path) && file.info(data_path)$size > 1000

if (USE_REAL_DATA) {
  raw <- read_csv(data_path, show_col_types = FALSE)

  # ── Flexible column detection ─────────────────────────────────────────────
  # Normalise column names to lowercase for matching
  names(raw) <- tolower(trimws(names(raw)))
  names(raw) <- gsub("[ .]", "_", names(raw))
  cat("Columns detected:", paste(names(raw), collapse = ", "), "\n")
  cat("Rows:", nrow(raw), "\n")

  # Detect mass column (g fresh mass)
  mass_col <- names(raw)[which(grepl("mass|fm|fresh", names(raw)) &
                                !grepl("log", names(raw)))[1]]
  # Detect respiration column (nmol or µmol CO2 g-1 s-1 or similar)
  resp_col  <- names(raw)[which(grepl("resp|rate|r_", names(raw)) &
                                 !grepl("log", names(raw)))[1]]
  # Detect life-form and plant-part columns
  lf_col    <- names(raw)[which(grepl("life|form|type|woody|herb|group",
                                       names(raw)))[1]]
  part_col  <- names(raw)[which(grepl("part|organ|shoot|root",
                                       names(raw)))[1]]
  sp_col    <- names(raw)[which(grepl("speci|taxa|name", names(raw)))[1]]
  ds_col    <- names(raw)[which(grepl("dataset|source|study|ref",
                                       names(raw)))[1]]
  temp_col  <- names(raw)[which(grepl("temp", names(raw)) &
                                  !grepl("log", names(raw)))[1]]

  # Build harmonised frame
  dat <- raw |>
    rename_with(~ "fresh_mass_kg",   all_of(mass_col)) |>
    rename_with(~ "respiration",     all_of(resp_col)) |>
    rename_with(~ "life_form",       any_of(lf_col))  |>
    rename_with(~ "plant_part",      any_of(part_col)) |>
    rename_with(~ "species",         any_of(sp_col))   |>
    rename_with(~ "dataset_source",  any_of(ds_col))   |>
    filter(!is.na(fresh_mass_kg), !is.na(respiration),
           fresh_mass_kg > 0,  respiration > 0) |>
    mutate(
      # Units from Dryad README:
      #   Fresh mass: kg   — convert to g for consistency with body-mass scaling literature
      #   Respiration: μmol CO2 s⁻¹  (absolute whole-organ rate, not mass-specific)
      fresh_mass_g = fresh_mass_kg * 1000,
      log_mass = log10(fresh_mass_g),
      log_resp = log10(respiration),
      # Standardise life_form to title case used throughout this document
      life_form  = dplyr::recode(as.character(life_form),
                                 "woody" = "Woody",
                                 "herb"  = "Herbaceous",
                                 .default = as.character(life_form)),
      # Standardise plant_part to title case
      plant_part = stringr::str_to_title(as.character(plant_part))
    )

  # ── Mass-range QA ──────────────────────────────────────────────────────────
  # The Dryad README declares units as "kg". A ×1000 conversion is applied here.
  # Records > 1 tonne (1e6 g) per organ are flagged for manual verification.
  mass_outliers <- dat |>
    filter(fresh_mass_g > 1e6) |>
    dplyr::select(species, life_form, plant_part, fresh_mass_kg, fresh_mass_g, dataset_source)
  if (nrow(mass_outliers) > 0) {
    cat("⚠ Mass outliers (fresh_mass_g > 1,000,000 g = 1 tonne) — verify units against original study:\n")
    print(mass_outliers)
  } else {
    cat("Mass QA: no records > 1 tonne. Max fresh mass:", round(max(dat$fresh_mass_g, na.rm=TRUE) / 1000, 1), "kg.\n")
  }
  if (!is.null(temp_col) && !is.na(temp_col) && temp_col %in% names(raw)) {
    dat <- left_join(dat, raw |>
                       rename(meas_temp_c = all_of(temp_col)) |>
                       select(meas_temp_c), by = character())
  }

} else {
  stop(
    "\n\n=== REAL DATA REQUIRED — ANALYSIS HALTED ===\n",
    "data.respiration.csv is missing or empty.\n",
    "Path checked: ", data_path, "\n\n",
    "Download from: https://doi.org/10.5061/dryad.sxksn03cj\n",
    "  1. Go to the Dryad page above\n",
    "  2. Download 'data.respiration.csv'\n",
    "  3. Save to: plant_scaling_data/data/raw/kurosawa_respiration/data.respiration.csv\n",
    "  4. Re-render this Rmd\n\n",
    "Fabricated data will NEVER be used as a substitute for real measurements.\n"
  )
  # ── Synthetic data fallback permanently removed (2026-05-02, hardened 2026-05-02) ──
  # stop() above halts all rendering. No fallback code is retained.
  # Audit: scholarly-rigor-reviewer confirmed that retaining functional synthetic code
  # after stop() is a latent integrity risk (bypass via error=TRUE). All dead code removed.
}
# Add placeholder family column if real data loaded without it (will be filled in Section 6)
if (!"family" %in% names(dat)) {
  dat$family <- NA_character_
}
```

```{r data-summary-table}
kable(
  dat |>
    count(life_form, plant_part, dataset_source) |>
    arrange(life_form, plant_part, dataset_source),
  caption = "Record counts by life form, plant part, and dataset source"
)
```

---

# Geographic distribution of observations

```{r geo-check}
# ── Check for coordinate columns in loaded data ───────────────────────────────
# The Kurosawa et al. 2025 Dryad dataset (doi:10.5061/dryad.sxksn03cj) contains
# allometric measurements compiled across Japanese forest sites (Mori 2010;
# Kurosawa 2023) and European sites (Ferrio JP, Spain). The Dryad CSV does NOT
# include per-observation lat/lon; coordinates are at the site / study level
# and must be sourced from the original study metadata or supplementary tables.
#
# ⚠ DATA STATUS (current session):
#   - data.respiration.csv is EMPTY — real data not yet downloaded from Dryad.
#   - All analyses above use RECONSTRUCTED DEMONSTRATION DATA (no coordinates).
#   - The map below shows APPROXIMATE LITERATURE-DERIVED study regions only.
#     Individual measurement coordinates are NOT available in either the demo
#     data or the Dryad dataset as currently structured.

HAS_COORDS <- USE_REAL_DATA &&
  any(grepl("lat|latitude",  names(dat), ignore.case = TRUE)) &&
  any(grepl("lon|longitude", names(dat), ignore.case = TRUE))

if (USE_REAL_DATA) {
  coord_cols <- names(dat)[grepl("lat|lon|latitude|longitude", names(dat),
                                 ignore.case = TRUE)]
  if (length(coord_cols) > 0) {
    cat("Real data loaded. Coordinate columns found:", paste(coord_cols, collapse = ", "), "\n")
    cat("Records with non-NA coordinates:",
        sum(!is.na(dat[[coord_cols[grepl("lat", coord_cols, ignore.case=TRUE)[1]]]])), "/", nrow(dat), "\n")
  } else {
    cat("Real data loaded but NO coordinate columns detected.\n")
    cat("The Dryad CSV does not appear to contain per-observation lat/lon.\n")
    cat("Site-level coordinates must be appended from original study metadata.\n")
  }
} else {
  cat("DEMO DATA active — no observation coordinates available.\n")
  cat("Map below shows approximate study regions from published literature.\n")
}
```

```{r geo-map, fig.cap="Approximate geographic distribution of study regions contributing to the Kurosawa et al. 2025 dataset. Points represent known measurement site locations sourced from published study descriptions, NOT per-observation coordinates from the Dryad CSV. All coordinates are literature-derived approximations and should be verified against original publications before any spatial analysis.", fig.width=10, fig.height=5.5}
library(maps)
library(ggrepel)

# ── Literature-derived approximate site locations ─────────────────────────────
# Sources: Mori et al. 2010 (RSPB 277:1447-1454); Kurosawa et al. 2023 (Ann Bot
# 131:313-322); Kurosawa et al. 2025 (RSPB 292:20241910).
# Coordinates are approximate centre-points of the named forest sites described
# in those papers. They are NOT from the Dryad dataset.
# ⚠ Verify all coordinates against original supplementary material before use.
approx_sites <- tribble(
  ~site_name,                     ~lat,    ~lon,   ~dataset_source,  ~region,
  "Miyagi Exp. Forest (Tohoku)",  38.15,  140.72,  "Mori2010",       "Japan",
  "Aomori Research Forest",       40.60,  141.10,  "Mori2010",       "Japan",
  "Hokkaido Univ. Forest (N)",    44.02,  142.40,  "Mori2010",       "Japan",
  "Chichibu (Saitama)",           35.90,  138.90,  "Mori2010",       "Japan",
  "Wakayama Forest (Osaka CU)",   33.92,  135.54,  "Mori2010",       "Japan",
  "Takayama (Gifu)",              36.13,  137.42,  "New",            "Japan",
  "Fuji Oshima Research Forest",  35.40,  138.72,  "Kurosawa2023",   "Japan",
  "Lleida / Pyrenees (Spain)",    41.62,    0.62,  "New",            "Europe"
)

world_map <- map_data("world")

ggplot() +
  geom_polygon(data = world_map |>
                 filter(long > -20, long < 160, lat > 20, lat < 60),
               aes(x = long, y = lat, group = group),
               fill = "grey92", colour = "grey70", linewidth = 0.2) +
  geom_point(data = approx_sites,
             aes(x = lon, y = lat, colour = dataset_source, shape = region),
             size = 3.5, alpha = 0.85) +
  geom_label_repel(data = approx_sites,
                   aes(x = lon, y = lat, label = site_name,
                       colour = dataset_source),
                   size = 2.8, max.overlaps = 20,
                   box.padding = 0.35, point.padding = 0.2,
                   segment.colour = "grey50", show.legend = FALSE) +
  scale_colour_brewer(palette = "Set1", name = "Dataset source") +
  scale_shape_manual(values = c(Japan = 16, Europe = 17), name = "Region") +
  coord_fixed(ratio = 1.3, xlim = c(-10, 155), ylim = c(25, 55)) +
  labs(
    title = "Approximate study regions — Kurosawa et al. 2025 dataset",
    subtitle = paste0(
      if (USE_REAL_DATA) "Real data loaded. " else "⚠ DEMO DATA — real data.respiration.csv not downloaded. ",
      "Points are literature-derived site approximations, NOT per-observation coordinates."
    ),
    x = "Longitude", y = "Latitude"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right",
        plot.subtitle = element_text(size = 9, colour = "firebrick"))
```

```{r geo-site-table}
kable(
  approx_sites |>
    arrange(region, dataset_source) |>
    rename(`Site name` = site_name, Latitude = lat, Longitude = lon,
           `Dataset source` = dataset_source, Region = region),
  digits = 2,
  caption = paste0(
    "Literature-derived approximate study site coordinates. ",
    "Sources: Mori et al. 2010 (RSPB); Kurosawa et al. 2023 (Ann Bot); ",
    "Kurosawa et al. 2025 (RSPB). ",
    "⚠ Coordinates are approximate and must be verified against original supplementary material."
  )
)
```

---

# Descriptive overview

```{r mass-distribution, fig.cap="Fresh-mass distribution by life form and plant part (log10 scale)."}
ggplot(dat, aes(x = log_mass, fill = life_form)) +
  geom_histogram(bins = 40, alpha = 0.75, colour = "white") +
  facet_grid(plant_part ~ life_form, scales = "free_y") +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                               unknown = "grey60")) +
  labs(x = expression(log[10]~"Fresh mass (g)"),
       y = "Count",
       fill = "Life form",
       title = "Fresh-mass distribution") +
  theme(legend.position = "none")
```

```{r resp-distribution, fig.cap="Respiration rate distribution by life form and plant part (log10 scale)."}
ggplot(dat, aes(x = log_resp, fill = life_form)) +
  geom_histogram(bins = 40, alpha = 0.75, colour = "white") +
  facet_grid(plant_part ~ life_form, scales = "free_y") +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                               unknown = "grey60")) +
  labs(x = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
       y = "Count",
       fill = "Life form",
       title = "Respiration rate distribution") +
  theme(legend.position = "none")
```

---

# SMA scaling analyses

## Overall scaling: R ~ Mass (pooled)

```{r sma-pooled}
# ── OLS ───────────────────────────────────────────────────────────────────────
fit_ols_pooled <- lm(log_resp ~ log_mass, data = dat)
ols_cf_p       <- coef(fit_ols_pooled)
ols_ci_p       <- confint(fit_ols_pooled)
ols_r2_p       <- summary(fit_ols_pooled)$r.squared
ols_se_p       <- summary(fit_ols_pooled)$coefficients[2, 2]
ols_n_p        <- nrow(model.frame(fit_ols_pooled))
ols_p075_p     <- 2 * pt(abs((ols_cf_p[2] - 0.75) / ols_se_p), ols_n_p - 2, lower.tail = FALSE)
ols_p100_p     <- 2 * pt(abs((ols_cf_p[2] - 1.00) / ols_se_p), ols_n_p - 2, lower.tail = FALSE)

# ── SMA ───────────────────────────────────────────────────────────────────────
fit_pooled  <- sma(log_resp ~ log_mass, data = dat, method = "SMA")
cf_pooled   <- coef(fit_pooled)
sma_r2_p    <- fit_pooled$r2[[1]]
test_pooled <- slope.test(dat$log_resp, dat$log_mass, test.value = 0.75, method = "SMA")
test_1      <- slope.test(dat$log_resp, dat$log_mass, test.value = 1.0,  method = "SMA")

# ── Print comparison ──────────────────────────────────────────────────────────
cat(sprintf("=== Pooled scaling: R ~ Mass (N = %d individuals) ===\n", ols_n_p))
cat(sprintf("OLS:  slope = %.3f  95%% CI [%.3f, %.3f]  R2 = %.3f\n",
            ols_cf_p[2], ols_ci_p[2,1], ols_ci_p[2,2], ols_r2_p))
cat(sprintf("  p (slope = 0.75) = %.4f  |  p (slope = 1.00) = %.4f\n",
            ols_p075_p, ols_p100_p))
cat(sprintf("SMA:  slope = %.3f  95%% CI [%.3f, %.3f]  R2 = %.3f\n",
            cf_pooled[2], fit_pooled$coef[[1]][2,2], fit_pooled$coef[[1]][2,3], sma_r2_p))
cat(sprintf("  p (slope = 0.75) = %.4f  |  p (slope = 1.00) = %.4f\n",
            test_pooled$p, test_1$p))

# ── Summary table ─────────────────────────────────────────────────────────────
kable(
  tibble(
    Method       = c("OLS", "SMA"),
    N            = ols_n_p,
    Slope        = round(c(ols_cf_p[2],           cf_pooled[2]),              3),
    `CI lo`      = round(c(ols_ci_p[2,1],         fit_pooled$coef[[1]][2,2]), 3),
    `CI hi`      = round(c(ols_ci_p[2,2],         fit_pooled$coef[[1]][2,3]), 3),
    R2           = round(c(ols_r2_p,              sma_r2_p),                  3),
    `p(b=0.75)`  = signif(c(ols_p075_p,           test_pooled$p),             3),
    `p(b=1.00)`  = signif(c(ols_p100_p,           test_1$p),                  3)
  ),
  caption = "Pooled respiration scaling exponents: OLS vs SMA (all individuals, log R ~ log M). p values are two-tailed tests against WBE 3/4 and isometric 1.0."
)
```

```{r sma-pooled-plot, fig.cap="Pooled OLS (blue dashed) and SMA (black solid) regressions of log respiration vs. log fresh mass. Grey dashed = WBE 3/4 reference."}
ggplot(dat, aes(x = log_mass, y = log_resp, colour = life_form)) +
  geom_point(alpha = 0.25, size = 1.2) +
  geom_abline(intercept = ols_cf_p[1], slope = ols_cf_p[2],
              linewidth = 1.0, colour = "#2166ac", linetype = "dashed") +
  geom_abline(intercept = cf_pooled[1], slope = cf_pooled[2],
              linewidth = 1.1, colour = "black") +
  geom_abline(intercept = mean(dat$log_resp, na.rm = TRUE) - 0.75 * mean(dat$log_mass, na.rm = TRUE),
              slope = 0.75, linetype = "dashed", colour = "grey40", linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.8, size = 3.5,
           label = sprintf("SMA slope = %.3f\nOLS slope = %.3f", cf_pooled[2], ols_cf_p[2])) +
  scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                                 unknown = "grey60")) +
  labs(x = expression(log[10]~"Fresh mass (g)"),
       y = expression(log[10]~"Respiration"),
       colour = "Life form",
       title = "Pooled respiration scaling: OLS (blue dashed) vs SMA (black solid)",
       subtitle = "Grey dashed = WBE 3/4 reference") +
  theme(legend.position = "bottom")
```

## By life form and plant part

```{r sma-by-group}
groups <- dat |>
  filter(life_form != "unknown", plant_part != "unknown") |>
  mutate(group = paste(life_form, plant_part, sep = " — "))

group_fits <- groups |>
  group_by(group, life_form, plant_part) |>
  nest() |>
  mutate(
    # SMA
    sma_fit   = map(data, ~ sma(log_resp ~ log_mass, data = .x, method = "SMA")),
    sma_slope = map_dbl(sma_fit, ~ coef(.x)[2]),
    intercept = map_dbl(sma_fit, ~ coef(.x)[1]),
    sma_ci_lo = map_dbl(sma_fit, ~ .x$coef[[1]][2,2]),
    sma_ci_hi = map_dbl(sma_fit, ~ .x$coef[[1]][2,3]),
    sma_r2    = map_dbl(sma_fit, ~ .x$r2[[1]]),
    sma_p_075 = map_dbl(data, ~ tryCatch(
      slope.test(.x$log_resp, .x$log_mass, test.value = 0.75, method = "SMA")$p,
      error = function(e) NA_real_)),
    sma_p_100 = map_dbl(data, ~ tryCatch(
      slope.test(.x$log_resp, .x$log_mass, test.value = 1.00, method = "SMA")$p,
      error = function(e) NA_real_)),
    # OLS
    ols_fit   = map(data, ~ lm(log_resp ~ log_mass, data = .x)),
    ols_slope = map_dbl(ols_fit, ~ coef(.x)[2]),
    ols_ci_lo = map_dbl(ols_fit, ~ tryCatch(confint(.x)[2,1], error=function(e) NA_real_)),
    ols_ci_hi = map_dbl(ols_fit, ~ tryCatch(confint(.x)[2,2], error=function(e) NA_real_)),
    ols_r2    = map_dbl(ols_fit, ~ summary(.x)$r.squared),
    ols_se    = map_dbl(ols_fit, ~ summary(.x)$coefficients[2,2]),
    n         = map_int(data, nrow),
    ols_p_075 = map_dbl(ols_fit, ~ {
      s <- summary(.x)$coefficients[2, ]
      2 * pt(abs((s[1] - 0.75) / s[2]), df = .x$df.residual, lower.tail = FALSE)
    }),
    ols_p_100 = map_dbl(ols_fit, ~ {
      s <- summary(.x)$coefficients[2, ]
      2 * pt(abs((s[1] - 1.00) / s[2]), df = .x$df.residual, lower.tail = FALSE)
    })
  )

# ── Print per-group results ───────────────────────────────────────────────────
for (i in seq_len(nrow(group_fits))) {
  g <- group_fits$group[i]
  cat(sprintf("\n--- %s  (N=%d) ---\n", g, group_fits$n[i]))
  cat(sprintf("  OLS:  slope = %.3f [%.3f, %.3f]  R\u00b2 = %.3f  p(\u03b2=0.75) = %.4f  p(\u03b2=1.00) = %.4f\n",
              group_fits$ols_slope[i], group_fits$ols_ci_lo[i], group_fits$ols_ci_hi[i],
              group_fits$ols_r2[i], group_fits$ols_p_075[i], group_fits$ols_p_100[i]))
  cat(sprintf("  SMA:  slope = %.3f [%.3f, %.3f]  R\u00b2 = %.3f  p(\u03b2=0.75) = %.4f  p(\u03b2=1.00) = %.4f\n",
              group_fits$sma_slope[i], group_fits$sma_ci_lo[i], group_fits$sma_ci_hi[i],
              group_fits$sma_r2[i], group_fits$sma_p_075[i], group_fits$sma_p_100[i]))
}

# ── Summary table ─────────────────────────────────────────────────────────────
kable(
  group_fits |>
    select(group, n,
           ols_slope, ols_ci_lo, ols_ci_hi, ols_r2, ols_p_075, ols_p_100,
           sma_slope, sma_ci_lo, sma_ci_hi, sma_r2, sma_p_075, sma_p_100) |>
    rename(Group = group, N = n,
           `OLS slope` = ols_slope, `OLS CI lo` = ols_ci_lo, `OLS CI hi` = ols_ci_hi,
           `OLS R2` = ols_r2, `OLS p(b=.75)` = ols_p_075, `OLS p(b=1)` = ols_p_100,
           `SMA slope` = sma_slope, `SMA CI lo` = sma_ci_lo, `SMA CI hi` = sma_ci_hi,
           `SMA R2` = sma_r2, `SMA p(b=.75)` = sma_p_075, `SMA p(b=1)` = sma_p_100) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "OLS and SMA scaling exponents by life form and plant part (individual-level data, log R ~ log M). SMA slopes are always >= OLS slopes when R-squared < 1."
)
```

```{r sma-by-group-plot, fig.cap="Observed data with OLS (blue dashed) and SMA (solid) regression lines per group. Grey dashed = WBE 3/4 reference."}
ablines_df <- group_fits |>
  select(group, intercept, sma_slope, ols_slope, life_form, plant_part)

ggplot(groups, aes(x = log_mass, y = log_resp, colour = life_form)) +
  geom_point(alpha = 0.20, size = 1.1) +
  geom_abline(data = ablines_df,
              aes(intercept = intercept, slope = ols_slope, colour = life_form),
              linewidth = 0.9, linetype = "dashed") +
  geom_abline(data = ablines_df,
              aes(intercept = intercept, slope = sma_slope, colour = life_form),
              linewidth = 1.2) +
  facet_grid(plant_part ~ life_form, scales = "free") +
  scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                                 unknown = "grey60")) +
  labs(x = expression(log[10]~"Fresh mass (g)"),
       y = expression(log[10]~"Respiration (\u03bcmol CO"[2]~"s"^{-1}~")"),
       colour = "Life form",
       title = "Respiration scaling by life form and plant part",
       subtitle = "Solid = SMA; dashed = OLS") +
  theme(legend.position = "bottom")
```

## Slope comparison: woody vs. herbaceous

```{r slope-comparison-lf}
dat_lf <- filter(dat, life_form %in% c("Woody", "Herbaceous"))

# ── OLS: slope comparison via interaction model ───────────────────────────────
ols_lf_int  <- lm(log_resp ~ log_mass * life_form, data = dat_lf)
ols_lf_summ <- summary(ols_lf_int)
cat("=== OLS slope comparison: Woody vs. Herbaceous ===\n")
cat("Interaction model: log_resp ~ log_mass * life_form\n")
print(round(ols_lf_summ$coefficients, 4))
ols_lf_p <- ols_lf_summ$coefficients["log_mass:life_formWoody", "Pr(>|t|)"]
if (is.na(ols_lf_p)) {
  # try the other level
  coef_names <- rownames(ols_lf_summ$coefficients)
  int_name   <- coef_names[grepl("log_mass:life_form", coef_names)]
  ols_lf_p   <- if (length(int_name)) ols_lf_summ$coefficients[int_name, "Pr(>|t|)"] else NA_real_
}
cat(sprintf("\nOLS interaction p (different slopes?) = %.4f\n", ols_lf_p))

# Per-group OLS slopes for reporting
ols_lf_grp <- dat_lf |>
  group_by(life_form) |>
  nest() |>
  mutate(
    fit     = map(data, ~ lm(log_resp ~ log_mass, data = .x)),
    slope   = map_dbl(fit, ~ coef(.x)[2]),
    ci_lo   = map_dbl(fit, ~ confint(.x)[2,1]),
    ci_hi   = map_dbl(fit, ~ confint(.x)[2,2]),
    r2      = map_dbl(fit, ~ summary(.x)$r.squared),
    n       = map_int(data, nrow)
  )
cat("\nPer-group OLS slopes:\n")
for (i in seq_len(nrow(ols_lf_grp))) {
  cat(sprintf("  %s: slope = %.3f [%.3f, %.3f]  R\u00b2 = %.3f  n = %d\n",
              ols_lf_grp$life_form[i], ols_lf_grp$slope[i],
              ols_lf_grp$ci_lo[i], ols_lf_grp$ci_hi[i],
              ols_lf_grp$r2[i], ols_lf_grp$n[i]))
}

# ── SMA: slope comparison ─────────────────────────────────────────────────────
cat("\n=== SMA slope comparison: Woody vs. Herbaceous ===\n")
fit_lf  <- sma(log_resp ~ log_mass * life_form, data = dat_lf, method = "SMA")
comp_lf <- summary(fit_lf)
cat("Slope heterogeneity test (SMA):\n")
print(comp_lf$coef)
```

## Slope comparison: shoot vs. root

```{r slope-comparison-part}
fit_part <- sma(log_resp ~ log_mass * plant_part,
                data = filter(dat, plant_part %in% c("Shoot", "Root")),
                method = "SMA")
comp_part <- summary(fit_part)
cat("Shoot vs. Root slope comparison:\n")
print(comp_part$coef)
```

---

# Dataset-source comparison (Mori 2010 / Kurosawa 2023 / New)

```{r sma-by-dataset, fig.cap="SMA lines stratified by dataset source within the woody component."}
woody_dat <- dat |>
  filter(life_form == "Woody", !is.na(dataset_source)) |>
  mutate(dataset_source = factor(dataset_source,
                                 levels = c("Mori2010", "Kurosawa2023", "New")))

if (length(unique(woody_dat$dataset_source)) > 1) {
  ds_fits <- woody_dat |>
    group_by(dataset_source, plant_part) |>
    filter(n() >= 10) |>
    nest() |>
    mutate(
      fit        = map(data, ~ sma(log_resp ~ log_mass, data = .x, method = "SMA")),
      slope      = map_dbl(fit, ~ coef(.x)[2]),
      intercept  = map_dbl(fit, ~ coef(.x)[1]),
      ci_lo      = map_dbl(fit, ~ .x$coef[[1]][2,2]),
      ci_hi      = map_dbl(fit, ~ .x$coef[[1]][2,3]),
      n          = map_int(data, nrow)
    )

  kable(
    ds_fits |>
      select(dataset_source, plant_part, n, slope, ci_lo, ci_hi) |>
      rename(`Dataset source` = dataset_source, `Plant part` = plant_part,
             N = n, Slope = slope, `CI low` = ci_lo, `CI high` = ci_hi) |>
      mutate(across(where(is.numeric), ~ round(.x, 3))),
    caption = "SMA slopes by dataset source (woody plants only)"
  )

  ggplot(woody_dat, aes(x = log_mass, y = log_resp, colour = dataset_source)) +
    geom_point(alpha = 0.25, size = 1.2) +
    geom_abline(data = ds_fits,
                aes(intercept = intercept, slope = slope, colour = dataset_source),
                linewidth = 1.1) +
    facet_wrap(~ plant_part, scales = "free") +
    scale_colour_brewer(palette = "Set1") +
    labs(x = expression(log[10]~"Fresh mass (g)"),
         y = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
         colour = "Dataset source",
         title = "Woody-plant respiration by dataset source") +
    theme(legend.position = "bottom")
}
```

---

# Taxonomy and family assignment

```{r taxonomy-audit}
# ── Demo-mode family assignment ────────────────────────────────────────────
# In demo mode, family is hardcoded from APG IV classification (see species_lookup above).
# In real-data mode (USE_REAL_DATA = TRUE), the family column must be populated
# via a taxonomic API (e.g. taxize::tax_name or rgbif::name_backbone) or
# a curated lookup table from the Dryad README.
#
# ⚠ All family assignments below are for demo data only and have NOT been
# verified through a live taxonomic service. Verify via GBIF Backbone or TNRS
# (https://tnrs.biendata.org) before any publication use.

# Build species-level audit table
taxonomy_audit <- dat |>
  group_by(species, family, life_form) |>
  summarise(
    n_records    = n(),
    n_parts      = n_distinct(plant_part),
    parts        = paste(sort(unique(plant_part)), collapse = " + "),
    .groups = "drop"
  ) |>
  arrange(family, life_form, species)

kable(
  taxonomy_audit |>
    rename(Species = species, Family = family, `Life form` = life_form,
           `N records` = n_records, `N parts` = n_parts, Parts = parts),
  caption = if (USE_REAL_DATA)
    "Species inventory: family, life form, and record counts"
  else
    "\u26a0 DEMO DATA — Species inventory with hardcoded APG IV family assignments"
)
```

```{r family-counts}
fam_summary <- dat |>
  group_by(family, life_form) |>
  summarise(
    n_species = n_distinct(species),
    n_records = n(),
    .groups = "drop"
  ) |>
  mutate(
    pgls_eligible = n_species >= 8,
    sma_eligible  = n_species >= 5 | n_records >= 10
  ) |>
  arrange(desc(n_species))

kable(
  fam_summary |>
    rename(Family = family, `Life form` = life_form,
           `N species` = n_species, `N records` = n_records,
           `PGLS eligible (n≥8)` = pgls_eligible,
           `SMA eligible (n≥5 sp or n≥10 rec)` = sma_eligible),
  caption = "Family summary: species counts and eligibility for family-level analyses"
)
```

---

# Phylogenetically corrected scaling (PGLS)

PGLS corrects for phylogenetic non-independence among species by modelling
residual covariance as proportional to the Brownian-motion phylogenetic
covariance matrix scaled by Pagel's $\lambda$. The formal model is:

$$\log R_i = \alpha + \beta \log M_i + \varepsilon_i, \quad
\boldsymbol{\varepsilon} \sim \mathcal{N}\!\left(0,\,
\sigma^2\bigl[\lambda\mathbf{V} + (1-\lambda)\mathbf{I}\bigr]\right)$$

where $\mathbf{V}$ is the Brownian-motion phylogenetic covariance matrix
(off-diagonal entries $\propto$ shared branch length to MRCA), and $\lambda$
is estimated by maximum likelihood. $\lambda \approx 1$ indicates strong
phylogenetic signal (Brownian motion); $\lambda \approx 0$ collapses to OLS.

**Interpretation of Pagel's $\lambda$ for metabolic scaling** — four ecologically
distinct scenarios:

1. **$\lambda \approx 0$ (phylogenetically random residuals):** Metabolic residuals
   are uncorrelated with shared ancestry; species converge on the same scaling
   behaviour regardless of lineage. This is the expected outcome if WBE describes
   a universal physical optimum: natural selection continuously drives all vascular
   networks toward the same fractal geometry, erasing the phylogenetic signal in
   residual metabolic rate.

2. **$\lambda \approx 1$ (Brownian-motion conservatism):** Residual respiration
   evolves as a random walk along phylogenetic branches; closely related species
   have more similar absolute rates than chance predicts. This indicates
   physiological conservatism — perhaps through shared enzyme concentrations,
   membrane lipid composition, or ecological niche conservatism among the taxa
   sampled.

3. **Intermediate $\lambda$:** Partial phylogenetic signal, where both universal
   geometric constraints and clade-specific biochemistry shape the data. This is
   the most commonly observed pattern in broad metabolic-rate compilations
   spanning multiple life forms (Glazier 2005) and is expected when WBE governs
   the exponent while lineage-level physiology modulates the intercept.

4. **High $\lambda$ combined with $\hat{\beta} \approx 3/4$ (strongest test):**
   Phylogeny strongly predicts the *level* of respiration (intercept) but not the
   *rate of change with mass* (exponent). This dissociation — evolutionary
   conservatism in metabolic level yet convergence in the scaling exponent — is
   precisely the WBE signature: geometry constrains the exponent universally while
   biochemistry shapes the intercept clade-specifically. Observing
   $\hat{\beta} = 3/4$ after removing phylogenetic covariance provides
   substantially stronger support for WBE than the same estimate from an OLS fit
   that ignores shared ancestry.

> **Note:** PGLS uses GLS (Model I regression), not SMA. PGLS slopes are
> expected to be slightly lower than SMA slopes due to attenuation bias when
> predictor (log mass) carries measurement error. This systematic difference
> should be taken into account when comparing methods.

> **Phylogenetic analysis caveats:**
>
> 1. **Tree source:** Smith & Brown (2018) GBOTB megaphylogeny via V.PhyloMaker2 (Jin & Qian 2022). Covers seed plants only; ferns, bryophytes, and algae not placed. Intra-generic topology resolved by random polytomy splitting (Scenario S3, set.seed(42)).
> 2. **Branch-length meaning:** Branch lengths are time-calibrated (millions of years) from Zanne et al. (2014) and Smith & Brown (2018). PGLS assumes trait evolution follows Brownian motion on these branch lengths; this assumption may not hold for metabolic rates under strong selective pressures.
> 3. **Ultrametricity correction:** `phytools::force.ultrametric(method="extend")` was applied where needed. This method elongates the shortest tips, which can slightly inflate Pagel's λ by reducing tip-level residual variance. The magnitude of this effect should be checked by comparing λ with and without the correction.
> 4. **Hybrid and uncertain taxa excluded from PGLS:** *Chrysanthemum* ×*morifolium*, *Fragaria* ×*ananassa*, and *Leucophyta brownii* could not be placed in the GBOTB backbone (hybrid genomes / uncertain placement). These three species are retained in OLS and SMA analyses but excluded from PGLS.
> 5. **Singleton species (n_records = 1):** Included in species means but carry no within-species variance information. PGLS is sensitive to influential singleton tips; a sensitivity analysis excluding singletons is recommended before publication.
> 6. **GLS predictor error (EIV bias):** PGLS assumes log mass is measured without error. Species-mean log mass inherits within-species variance and individual-level measurement error. This causes attenuation of PGLS slopes relative to the true scaling exponent by an amount proportional to the signal-to-noise ratio in log mass.
> 7. **λ bounds:** Lambda is bounded at [1×10⁻⁶, 1]; values > 1 (stronger-than-BM covariance) cannot be detected. Blomberg's K is also reported; K > 1 indicates stronger-than-BM signal.
> 8. **Family assignments:** Populated from the phylo.maker match table (APG IV via GBOTB). Verify assignments against GBIF Backbone or TNRS before publication use.

## Species means for PGLS

```{r pgls-species-means}
# Aggregate individual records to species-level geometric means (= arithmetic
# mean of log10 values). Retain family, life_form, plant_part for downstream
# filtering. Singletons (n_records = 1) are included but flagged.
sp_means <- dat |>
  group_by(species, family, life_form, plant_part) |>
  summarise(
    log_resp_mean = mean(log_resp, na.rm = TRUE),
    log_mass_mean = mean(log_mass, na.rm = TRUE),
    n_records     = n(),
    sd_log_resp   = sd(log_resp, na.rm = TRUE),
    sd_log_mass   = sd(log_mass, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    singleton     = (n_records == 1L),
    # Species name with underscore for tree tip matching
    sp_tree       = gsub(" ", "_", species),
    genus         = gsub(" .*", "", species)
  ) |>
  filter(is.finite(log_resp_mean), is.finite(log_mass_mean))

cat("Total species means:", nrow(sp_means), "\n")
cat("Singletons (n_records=1):", sum(sp_means$singleton), "\n")
cat("Species with n>=3:", sum(!sp_means$singleton & sp_means$n_records >= 3), "\n")
```

## Phylogeny construction

```{r pgls-phylogeny, fig.cap="Smith & Brown (2018) megaphylogeny pruned to Kurosawa dataset species (demo). Fan layout; tip labels colored by family."}
if (HAS_PHYLOMAKER) {
  # Build species list for V.PhyloMaker2::phylo.maker()
  # Requires columns: species (binomial), genus, family
  sp_list_pgls <- sp_means |>
    select(species, genus, family) |>
    distinct(species, .keep_all = TRUE) |>
    # family NA causes phylo.maker to attempt genus-level inference
    mutate(family = ifelse(is.na(family), NA_character_, family))

  cat("Species for phylogeny:", nrow(sp_list_pgls), "\n")

  # Build tree via Scenario S3 (random polytomy resolution within genus)
  # set.seed ensures reproducibility of polytomy resolution (ape::multi2di)
  set.seed(42)
  tree_out <- tryCatch({
    V.PhyloMaker2::phylo.maker(
      sp.list   = sp_list_pgls,
      tree      = V.PhyloMaker2::GBOTB.extended.WP,
      nodes     = V.PhyloMaker2::nodes.info.1.WP,
      scenarios = "S3"
    )
  }, error = function(e) {
    message("phylo.maker() failed: ", conditionMessage(e)); NULL
  })

  if (!is.null(tree_out)) {
    pgls_tree <- tree_out$scenario.3
    # Validate and correct tree
    # ⚠ force.ultrametric(method="extend") elongates short tips and can slightly
    # inflate Pagel's λ by reducing tip-level residual variance. Check whether
    # non-ultrametricity is from floating-point rounding (tol = 1e-3) before applying.
    if (!ape::is.ultrametric(pgls_tree, tol = 1e-5)) {
      cat("Tree not ultrametric (max deviation:",
          max(abs(diag(ape::vcv(pgls_tree)) - mean(diag(ape::vcv(pgls_tree))))), ")\n")
      pgls_tree <- phytools::force.ultrametric(pgls_tree, method = "extend")
      cat("⚠ force.ultrametric(method='extend') applied — λ estimates may be slightly inflated.\n")
    }
    set.seed(42)
    pgls_tree <- ape::multi2di(pgls_tree, random = TRUE)
    pgls_tree$node.label <- NULL  # prevent caper label duplication error

    cat("Tree tips:", ape::Ntip(pgls_tree), "\n")
    cat("Is ultrametric:", ape::is.ultrametric(pgls_tree, tol = 1e-5), "\n")
    cat("Is binary:", ape::is.binary(pgls_tree), "\n")

    # BUG 3 FIX: Back-fill family from phylo.maker match.table (APG IV assignments).
    # phylo.maker returns family_in_tree / family.in.tree for every genus, even when
    # the input sp.list had NA families. Use these to populate sp_means and dat.
    mt <- tree_out$match.table
    fam_col <- intersect(c("family.in.tree", "family_in_tree", "family"), names(mt))
    if (length(fam_col) > 0) {
      fam_lookup <- mt |>
        dplyr::select(genus, family_phylo = dplyr::all_of(fam_col[1])) |>
        dplyr::distinct(genus, .keep_all = TRUE) |>
        dplyr::filter(!is.na(family_phylo), nchar(trimws(family_phylo)) > 0)
      sp_means <- sp_means |>
        dplyr::left_join(fam_lookup, by = "genus") |>
        dplyr::mutate(family = dplyr::coalesce(family, family_phylo)) |>
        dplyr::select(-family_phylo)
      dat <- dat |>
        dplyr::left_join(fam_lookup, by = "genus") |>
        dplyr::mutate(family = dplyr::coalesce(family, family_phylo)) |>
        dplyr::select(-family_phylo)
      cat("Family back-fill (APG IV via phylo.maker match.table):",
          sum(!is.na(sp_means$family)), "/", nrow(sp_means), "species assigned.\n")
    } else {
      cat("⚠ Family column not found in phylo.maker match.table — family PGLS will be skipped.\n")
    }

    # Match summary
    sp_in_tree <- gsub(" ", "_", sp_list_pgls$species) %in% pgls_tree$tip.label
    cat("Species matched to tree:", sum(sp_in_tree), "/", nrow(sp_list_pgls), "\n")
    unmatched <- sp_list_pgls$species[!sp_in_tree]
    if (length(unmatched) > 0) {
      cat("Unmatched taxa (hybrids / uncertain placements — excluded from PGLS, retained in OLS/SMA):\n")
      cat(paste(" ", unmatched), sep = "\n")
    }

    # Plot tree
    plot(pgls_tree, type = "fan", cex = 0.55, no.margin = TRUE,
         main = "Pruned megaphylogeny (Smith & Brown 2018, S3)")
  } else {
    pgls_tree <- NULL
    cat("Tree building failed — PGLS sections will be skipped.\n")
  }
  HAS_TREE <- !is.null(pgls_tree)
  USING_STAR_TREE <- FALSE
} else {
  # Fallback: star tree for demonstration when V.PhyloMaker2 not available
  message("V.PhyloMaker2 not available. Building a star tree for PGLS demonstration.")
  sp_for_star <- unique(sp_means$sp_tree)
  pgls_tree   <- ape::stree(length(sp_for_star), type = "star",
                             tip.label = sp_for_star)
  pgls_tree$edge.length <- rep(1, nrow(pgls_tree$edge))
  pgls_tree$node.label  <- NULL
  HAS_TREE <- TRUE
  USING_STAR_TREE <- TRUE
  cat("Star tree built with", ape::Ntip(pgls_tree), "tips.\n")
  cat("Note: Star tree has lambda = 0 by construction — PGLS = OLS on a star tree.\n")
}
```

## PGLS fits by group

```{r pgls-fit}
if (isTRUE(USING_STAR_TREE)) {
  cat("\n> **WARNING: PGLS results below used a star tree (V.PhyloMaker2 unavailable). All branch lengths are equal, lambda estimates are artifactually zero, and these results are NOT phylogenetically informed. Install V.PhyloMaker2 for valid PGLS: `remotes::install_github('jinyizju/V.PhyloMaker2')`**\n\n")
}
if (HAS_TREE) {
  # Helper: fit OLS + SMA + PGLS for one group subset
  fit_group_pgls <- function(sp_sub, tree, group_label) {
    # Match data to tree
    sp_sub <- sp_sub |> filter(sp_tree %in% tree$tip.label)
    if (nrow(sp_sub) < 6) {
      message("Skipping ", group_label, ": only ", nrow(sp_sub), " species in tree.")
      return(NULL)
    }
    tr <- ape::keep.tip(tree, sp_sub$sp_tree)
    tr$node.label <- NULL

    # BUG 1 FIX: restrict df to formula columns only.
    # caper::na.omit.comparative.data checks ALL columns for NAs, so any ancillary
    # column with NAs (e.g. family, sd_log_resp) will silently drop all rows.
    df           <- as.data.frame(sp_sub)[, c("sp_tree", "log_resp_mean", "log_mass_mean")]
    rownames(df) <- df$sp_tree

    cd <- tryCatch(
      caper::comparative.data(phy = tr, data = df, names.col = "sp_tree",
                              vcv = TRUE, warn.dropped = FALSE),
      error = function(e) { message("comparative.data failed for ", group_label, ": ", e$message); NULL }
    )
    if (is.null(cd)) return(NULL)

    # OLS
    ols_fit   <- lm(log_resp_mean ~ log_mass_mean, data = sp_sub)
    ols_s     <- summary(ols_fit)
    ols_slope <- coef(ols_fit)[2]
    ols_se    <- coef(ols_s)[2, 2]
    ols_ci    <- confint(ols_fit)[2, ]
    ols_r2    <- ols_s$r.squared

    # SMA
    sma_fit   <- tryCatch(smatr::sma(log_resp_mean ~ log_mass_mean, data = sp_sub, method = "SMA"),
                          error = function(e) NULL)
    if (!is.null(sma_fit)) {
      sma_slope <- sma_fit$coef[[1]][2, 1]
      sma_ci_lo <- sma_fit$coef[[1]][2, 2]
      sma_ci_hi <- sma_fit$coef[[1]][2, 3]
      sma_r2    <- sma_fit$r2[[1]]
    } else {
      sma_slope <- sma_ci_lo <- sma_ci_hi <- sma_r2 <- NA_real_
    }

    # PGLS
    pgls_fit <- tryCatch(
      caper::pgls(log_resp_mean ~ log_mass_mean, data = cd, lambda = "ML",
                  bounds = list(lambda = c(1e-6, 1))),
      error = function(e) {
        message("PGLS ML failed for ", group_label, ", trying lambda=1: ", e$message)
        tryCatch(caper::pgls(log_resp_mean ~ log_mass_mean, data = cd, lambda = 1),
                 error = function(e2) NULL)
      }
    )
    if (!is.null(pgls_fit)) {
      pgls_s      <- summary(pgls_fit)
      pgls_slope  <- coef(pgls_fit)[2]
      pgls_se     <- coef(pgls_s)[2, 2]
      pgls_n      <- length(pgls_fit$residuals)  # length() is robust; nrow() requires matrix storage
      pgls_ci     <- c(pgls_slope - qt(0.975, pgls_n - 2) * pgls_se,
                       pgls_slope + qt(0.975, pgls_n - 2) * pgls_se)
      pgls_r2     <- pgls_s$r.squared
      pgls_lambda <- pgls_fit$param["lambda"]
      # Lambda CI via profile likelihood
      lam_ci <- tryCatch({
        lam_prof <- caper::pgls.profile(pgls_fit, which = "lambda")
        caper::pgls.confint(lam_prof, 0.95)$ci.val
      }, error = function(e) c(NA_real_, NA_real_))
      pgls_p_wbe  <- 2 * pt(abs((pgls_slope - 0.75) / pgls_se), pgls_n - 2, lower.tail = FALSE)
      pgls_p_iso  <- 2 * pt(abs((pgls_slope - 1.00) / pgls_se), pgls_n - 2, lower.tail = FALSE)
    } else {
      pgls_slope <- pgls_se <- pgls_r2 <- pgls_lambda <- pgls_p_wbe <- pgls_p_iso <- NA_real_
      pgls_ci    <- c(NA_real_, NA_real_)
      lam_ci     <- c(NA_real_, NA_real_)
    }

    # Blomberg's K for log_resp_mean (on species in tree).
    # K and lambda capture different aspects: lambda scales off-diagonal covariance
    # relative to BM; K is the ratio of observed MPD to BM expectation.
    # K > 1 = stronger-than-BM clustering; K < 1 = weaker (trait evolution faster than BM).
    blom_k <- tryCatch({
      resp_vec <- setNames(sp_sub$log_resp_mean, sp_sub$sp_tree)
      phytools::phylosig(tr, resp_vec, method = "K", test = TRUE, nsim = 999)
    }, error = function(e) NULL)
    blomberg_K   <- if (!is.null(blom_k)) blom_k$K  else NA_real_
    blomberg_p   <- if (!is.null(blom_k)) blom_k$P  else NA_real_

    tibble(
      group        = group_label,
      n_species    = nrow(sp_sub),
      method       = c("OLS", "SMA", "PGLS"),
      slope        = c(ols_slope,  sma_slope,  pgls_slope),
      ci_lower     = c(ols_ci[1],  sma_ci_lo,  pgls_ci[1]),
      ci_upper     = c(ols_ci[2],  sma_ci_hi,  pgls_ci[2]),
      r_squared    = c(ols_r2,     sma_r2,     pgls_r2),
      lambda       = c(NA_real_,   NA_real_,   pgls_lambda),
      lambda_ci_lo = c(NA_real_,   NA_real_,   lam_ci[1]),
      lambda_ci_hi = c(NA_real_,   NA_real_,   lam_ci[2]),
      blomberg_K   = c(NA_real_,   NA_real_,   blomberg_K),
      blomberg_p   = c(NA_real_,   NA_real_,   blomberg_p),
      p_vs_wbe     = c(
        2 * pt(abs((ols_slope - 0.75) / ols_se), nrow(sp_sub) - 2, lower.tail = FALSE),
        tryCatch(smatr::slope.test(sp_sub$log_resp_mean, sp_sub$log_mass_mean,
                                   test.value = 0.75, method = "SMA")$p,
                 error = function(e) NA_real_),
        pgls_p_wbe
      ),
      p_vs_iso     = c(
        2 * pt(abs((ols_slope - 1.00) / ols_se), nrow(sp_sub) - 2, lower.tail = FALSE),
        tryCatch(smatr::slope.test(sp_sub$log_resp_mean, sp_sub$log_mass_mean,
                                   test.value = 1.00, method = "SMA")$p,
                 error = function(e) NA_real_),
        pgls_p_iso
      )
    )
  }

  # BUG 2 FIX: include Whole-organ groups (84 Woody-Whole + 31 Herbaceous-Whole species in tree).
  # Run for each group
  pgls_results <- bind_rows(
    fit_group_pgls(sp_means |> filter(life_form == "Woody", plant_part == "Shoot"),
                   pgls_tree, "Woody — Shoot"),
    fit_group_pgls(sp_means |> filter(life_form == "Woody", plant_part == "Root"),
                   pgls_tree, "Woody — Root"),
    fit_group_pgls(sp_means |> filter(life_form == "Woody", plant_part == "Whole"),
                   pgls_tree, "Woody — Whole"),
    fit_group_pgls(sp_means |> filter(life_form == "Herbaceous", plant_part == "Shoot"),
                   pgls_tree, "Herbaceous — Shoot"),
    fit_group_pgls(sp_means |> filter(life_form == "Herbaceous", plant_part == "Root"),
                   pgls_tree, "Herbaceous — Root"),
    fit_group_pgls(sp_means |> filter(life_form == "Herbaceous", plant_part == "Whole"),
                   pgls_tree, "Herbaceous — Whole")
  )
} else {
  pgls_results <- tibble()
}
```

## PGLS results table

```{r pgls-lambda-table}
if (nrow(pgls_results) > 0) {
  kable(
    pgls_results |>
      filter(method == "PGLS") |>
      select(group, n_species, slope, ci_lower, ci_upper, r_squared,
             lambda, lambda_ci_lo, lambda_ci_hi, blomberg_K, blomberg_p,
             p_vs_wbe, p_vs_iso) |>
      mutate(across(where(is.numeric), ~ round(.x, 3)),
             blomberg_p = ifelse(is.na(blomberg_p), "—",
                                 ifelse(blomberg_p < 0.001, "<0.001",
                                        as.character(round(blomberg_p, 3)))),
             p_vs_wbe = ifelse(is.na(p_vs_wbe), "—",
                               ifelse(p_vs_wbe < 0.001, "<0.001",
                                      as.character(round(p_vs_wbe, 3)))),
             p_vs_iso = ifelse(is.na(p_vs_iso), "—",
                               ifelse(p_vs_iso < 0.001, "<0.001",
                                      as.character(round(p_vs_iso, 3))))) |>
      rename(Group = group, `N species` = n_species,
             `PGLS slope` = slope, `CI low` = ci_lower, `CI high` = ci_upper,
             `R²` = r_squared, `λ (ML)` = lambda,
             `λ CI low` = lambda_ci_lo, `λ CI high` = lambda_ci_hi,
             `Blomberg K` = blomberg_K, `K p-value` = blomberg_p,
             `p (β≠3/4)` = p_vs_wbe, `p (β≠1)` = p_vs_iso),
    caption = "PGLS results: phylogenetically corrected scaling exponents, Pagel's λ, and Blomberg's K for log respiration. p (β≠3/4) = two-tailed test vs WBE prediction; p (β≠1) = test vs isometric null."
  )
}
```

## OLS / SMA / PGLS comparison forest plot

> **What this plot shows:**
> The forest plot below compares three estimators of the scaling exponent $\beta$ in
> **log respiration ~ log fresh mass** for each of six life-form × plant-part groups
> (Woody Shoot, Woody Root, Woody Whole, Herbaceous Shoot, Herbaceous Root, Herbaceous Whole).
> All fits are on **species means** (geometric mean of log-transformed individual records per species),
> not on raw individuals — this removes within-species pseudoreplication.
>
> | Estimator | Data used | Interpretation |
> |-----------|----------|----------------|
> | **OLS** | Species means | Minimises vertical (response-axis) residuals; underestimates true slope when x-axis (log mass) has measurement error |
> | **SMA** | Species means | Minimises residuals in both axes; preferred when both variables are measured with error |
> | **PGLS** | Species means, phylogenetically structured | Corrects for shared evolutionary history; Pagel's λ estimated by ML; slope reflects the relationship after accounting for phylogenetic covariance |
>
> OLS slopes are systematically lower than SMA slopes; PGLS slopes may differ from both
> depending on the estimated λ (λ = 0 → OLS, λ = 1 → fully Brownian-motion-corrected).

> **Competing scaling predictions (reference lines in plot below):**
>
> | $\beta$ | Prediction | Biological meaning |
> |--------|-----------|-------------------|
> | 2/3 | Rubner (1883) surface-area rule | Mass-specific metabolism ∝ M^(-1/3); heat loss ∝ surface area |
> | **3/4** | **WBE/MST** (West et al. 1997) | Fractal vascular network optimisation; universally applies to whole-organism metabolic rate |
> | 1 | Isometric null | Mass-specific respiration constant; no metabolic scaling with size |
>
> An organ-level slope of 3/4 is *consistent with* WBE but requires the additional
> assumption that the organ's fractional contribution to total plant respiration is
> size-invariant (Mori et al. 2010). The isometric null ($\beta = 1$) is the
> appropriate benchmark for "does size matter at all?" Rejection of both 3/4 *and* 1
> would suggest an intermediate physiological scaling regime.

```{r pgls-forest-plot, fig.height=6, fig.cap="OLS / SMA / PGLS slope comparison by group. Dashed = WBE 3/4 (red), dotted = isometric 1 (blue), dot-dash = Rubner 2/3 (grey). PGLS slopes incorporate phylogenetic correction; systematic offset from SMA expected due to x-error attenuation in GLS."}
if (nrow(pgls_results) > 0) {
  ggplot(pgls_results |> filter(!is.na(slope)),
         aes(x = slope, y = group, colour = method, shape = method)) +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                   position = position_dodge(width = 0.5), height = 0.25) +
    geom_vline(xintercept = 0.75, linetype = "dashed",  colour = "#d6604d",
               linewidth = 0.8) +
    geom_vline(xintercept = 1.00, linetype = "dotted",  colour = "#2166ac",
               linewidth = 0.8) +
    geom_vline(xintercept = 2/3,  linetype = "dotdash", colour = "grey50",
               linewidth = 0.6) +
    scale_colour_manual(values = c("OLS" = "steelblue", "SMA" = "darkorange",
                                   "PGLS" = "forestgreen")) +
    labs(x = "Scaling exponent  (log R ~ log Mass)",
         y = NULL, colour = "Method", shape = "Method",
         title = "Respiration scaling: OLS vs SMA vs PGLS (species means)",
         subtitle = "Red dashed = WBE 3/4; blue dotted = isometric 1; grey dot-dash = Rubner 2/3") +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom")
}
```

---

# Bayesian PGLMM: emulating Kurosawa et al. 2025

> **Modeling objective:** Estimate the interspecific scaling exponent $\hat{\beta}$
> for $\log R \sim \log M$ within each life-form × plant-part group using a
> Bayesian Phylogenetic Generalised Linear Mixed Model (PGLMM) that operates on
> **individual-level observations** rather than species means — the approach most
> consistent with Kurosawa et al. (2025). The phylogenetic covariance matrix enters
> as a correlated random-intercept term on species, partitioning variance into
> phylogenetic, non-phylogenetic species-level, and individual-level residual
> components.
>
> **Expert assessments** (phylogenetics-comparative-agent + merow-ecology, 2026-05-12):
> *  Recommend `brms` over `MCMCglmm` for superior convergence diagnostics (Rhat, Bulk/Tail ESS) and built-in `pp_check()`.
> *  Use individual-level data with `(1 | gr(phylo, cov = A))` for phylogenetic random intercept; add `(1 | sp)` for non-phylogenetic species variance only if ESS is adequate.
> *  Prior `Normal(0.75, 0.5)` on slope — wide enough to test WBE rather than assume it. Run sensitivity with `Normal(1, 0.5)`.
> *  Include `dataset_source` as a fixed covariate to absorb Japan/Spain measurement-temperature confound.
> *  Fixed-slope assumption is defensible: per-species OLS slope IQR (0.3–0.5) reflects estimation noise at n = 2–10, not true biological heterogeneity.
> *  Flag Fagus crenata ontogenetic series as potential DGP mixing; test slope sensitivity with and without.
> *  V.PhyloMaker2 single-topology tree is adequate; report as a stated uncertainty.
>
> **Known limitations:**
> * Models use a single V.PhyloMaker2 tree (topology uncertainty not propagated).
> * No temperature correction is possible if `meas_temp_c` is absent from the Dryad CSV; `dataset_source` acts as a nominal proxy.
> * Measurement error in log-mass is not modelled (errors-in-variables); this causes mild attenuation of $\hat{\beta}$ toward zero in all estimators.
> * brms sampling takes ~20–40 min per group; results are cached to `output/brm_cache/` and reloaded on subsequent renders.

## Kurosawa et al. 2025 published slopes (reference)

The following slopes are transcribed from Kurosawa et al. (2025, Table 1 /
Fig. 2) for direct comparison. **These values must be verified against the paper
before any publication use — transcription errors are possible.**

```{r kurosawa-published-slopes}
# ── Published Bayesian PGLMM slopes from Kurosawa et al. 2025 ─────────────────
# Source: Kurosawa Y et al. (2025) PRSB 292:20241910. DOI: 10.1098/rspb.2024.1910
# ⚠ All values UNVERIFIED — transcribed from abstract/figures, not from data.
# Confirm against Table 1 or SI before manuscript use.
kurosawa_published <- tibble::tribble(
  ~group,                 ~pub_slope, ~pub_ci_lo, ~pub_ci_hi, ~verified,
  "Woody — Shoot",         0.75,       NA,         NA,        "UNVERIFIED",
  "Woody — Root",          0.75,       NA,         NA,        "UNVERIFIED",
  "Herbaceous — Shoot",    0.91,       NA,         NA,        "UNVERIFIED",
  "Herbaceous — Root",     0.91,       NA,         NA,        "UNVERIFIED"
)
# NOTE: The WBE prediction is 0.75. Kurosawa et al. 2025 report that woody plants
# conform to WBE (slope ~0.75) while herbaceous plants show a steeper slope.
# The exact posterior means and 95% credible intervals must be read from the paper.
# The values above are placeholders derived from the paper's main narrative;
# replace with exact Table values once verified.

cat("⚠ Kurosawa et al. 2025 published slopes — PLACEHOLDER VALUES (unverified transcription)\n")
cat("  Update pub_slope, pub_ci_lo, pub_ci_hi from Table 1 or SI before publication use.\n\n")
kable(
  kurosawa_published |>
    mutate(across(c(pub_slope, pub_ci_lo, pub_ci_hi), ~ ifelse(is.na(.x), "—", as.character(round(.x, 3))))),
  col.names = c("Group", "Published slope", "Published CI lo", "Published CI hi", "Verification"),
  caption = "Kurosawa et al. 2025 Bayesian PGLMM slopes (placeholder — verify against paper Table 1 / SI)"
)
```

## Bayesian PGLMM setup

```{r pglmm-setup}
# ── Check required packages ───────────────────────────────────────────────────
HAS_BRMS      <- requireNamespace("brms",      quietly = TRUE)
HAS_MCMCGLMM  <- requireNamespace("MCMCglmm",  quietly = TRUE)
HAS_CODA      <- requireNamespace("coda",      quietly = TRUE)

# ── DGP tagging: flag Fagus crenata ontogenetic records ──────────────────────
# Kurosawa 2023 (Fagus crenata ontogenetic series) represents within-individual
# developmental growth — a different DGP from cross-species interspecific scaling.
# Records are tagged so we can test slope sensitivity with and without this subset.
dat <- dat |>
  dplyr::mutate(
    dgp_type = dplyr::case_when(
      grepl("Fagus crenata", species, ignore.case = TRUE) &
        grepl("Kurosawa", dataset_source, ignore.case = TRUE) ~ "ontogenetic",
      TRUE ~ "interspecific"
    )
  )
n_ontogenetic <- sum(dat$dgp_type == "ontogenetic", na.rm = TRUE)
cat(sprintf("DGP tagging: %d records tagged as 'ontogenetic' (Fagus crenata, Kurosawa 2023)\n",
            n_ontogenetic))
cat(sprintf("             %d records tagged as 'interspecific'\n",
            sum(dat$dgp_type == "interspecific", na.rm = TRUE)))

# ── Prepare species-label column for tree matching ────────────────────────────
# brms gr() requires exact match between data column and rownames of covariance matrix.
dat <- dat |>
  dplyr::mutate(
    phylo = gsub(" ", "_", species),   # phylogenetic RE (matched to tree tips)
    sp    = gsub(" ", "_", species)    # non-phylogenetic species RE
  )

# ── Dataset source as factor (covariate to absorb temperature/geographic confound)
dat <- dat |>
  dplyr::mutate(source_fct = factor(dataset_source))

cat("\nDataset sources (factor levels for temperature/geographic covariate):\n")
print(table(dat$source_fct))
```

## PGLMM fitting: brms (primary)

```{r pglmm-brms, cache=FALSE, message=FALSE, warning=FALSE}
# ── This chunk fits Bayesian PGLMMs via brms (Stan/HMC) ───────────────────────
# Each group model takes ~20–40 min on first fit; cached to output/brm_cache/.
# Subsequent renders load cached .rds files instantly.
#
# Model: log_resp ~ log_mass + source_fct + (1 | gr(phylo, cov = A))
#   Fixed: intercept, log_mass (the scaling exponent β), source_fct (confound)
#   Random: phylogenetic correlated intercepts via A = vcv.phylo(tree, corr=TRUE)
# Prior: Normal(0.75, 0.5) on β — wide enough to test WBE, not assume it.
# Prior sensitivity: also fit with Normal(1, 0.5) on β.

pglmm_brms_results <- list()

if (HAS_BRMS && HAS_TREE && !USING_STAR_TREE) {
  suppressPackageStartupMessages(library(brms))

  cache_dir <- file.path(proj_dir, "output", "brm_cache")
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

  # ── Helper: fit one PGLMM group ─────────────────────────────────────────────
  fit_pglmm_brms <- function(dat_sub, tree_full, group_label,
                              prior_slope_mean = 0.75,
                              prior_slope_sd   = 0.5,
                              prior_tag        = "wbe_informed") {
    dat_sub <- dat_sub |>
      dplyr::filter(
        phylo %in% tree_full$tip.label,
        is.finite(log_resp), is.finite(log_mass)
      )
    n_sp  <- dplyr::n_distinct(dat_sub$phylo)
    n_obs <- nrow(dat_sub)

    if (n_obs < 15 || n_sp < 6) {
      message("Skipping ", group_label, " (", prior_tag, "): n_sp=",
              n_sp, " n_obs=", n_obs, " — insufficient.")
      return(NULL)
    }

    tree_sub <- ape::keep.tip(tree_full, unique(dat_sub$phylo))
    # Phylogenetic correlation matrix (corr=TRUE → max diagonal = 1)
    A <- ape::vcv.phylo(tree_sub, corr = TRUE)
    dat_sub <- dat_sub |> dplyr::filter(phylo %in% rownames(A))

    # Priors
    priors <- c(
      brms::prior_string(
        paste0("normal(", prior_slope_mean, ", ", prior_slope_sd, ")"),
        class = "b", coef = "log_mass"
      ),
      brms::prior(normal(0, 2),    class = b),
      brms::prior(normal(0, 5),    class = Intercept),
      brms::prior(exponential(1),  class = sd),
      brms::prior(exponential(1),  class = sigma)
    )

    safe_label <- gsub("[^A-Za-z0-9]", "_",
                       paste0(group_label, "_", prior_tag))
    cache_file <- file.path(cache_dir, safe_label)

    # Drop non-phylogenetic species RE when n_sp < 15 (poorly identified)
    re_formula <- if (n_sp >= 15) {
      "log_resp ~ log_mass + source_fct + (1 | gr(phylo, cov = A)) + (1 | sp)"
    } else {
      "log_resp ~ log_mass + source_fct + (1 | gr(phylo, cov = A))"
    }

    fit <- tryCatch(
      brms::brm(
        formula = brms::bf(as.formula(re_formula)),
        data    = as.data.frame(dat_sub),
        data2   = list(A = A),
        family  = gaussian(),
        prior   = priors,
        chains  = 4, iter = 4000, warmup = 2000, cores = 4,
        seed    = 42, file = cache_file,
        control = list(adapt_delta = 0.95, max_treedepth = 12),
        silent  = 2
      ),
      error = function(e) {
        message("brms fit failed for ", group_label, " (", prior_tag, "): ", e$message)
        NULL
      }
    )
    if (is.null(fit)) return(NULL)

    # ── Extract results ──────────────────────────────────────────────────────
    fx        <- brms::fixef(fit)
    slope_m   <- fx["log_mass", "Estimate"]
    slope_lo  <- fx["log_mass", "Q2.5"]
    slope_hi  <- fx["log_mass", "Q97.5"]

    rhat_max  <- max(brms::rhat(fit), na.rm = TRUE)
    neff_r    <- brms::neff_ratio(fit)
    ess_min   <- min(neff_r, na.rm = TRUE) * (4 * 2000)

    vc        <- brms::VarCorr(fit)
    sd_phylo  <- vc$phylo$sd[1, "Estimate"]
    sd_resid  <- vc$residual__$sd[1, "Estimate"]
    sd_sp     <- if ("sp" %in% names(vc)) vc$sp$sd[1, "Estimate"] else NA_real_
    var_total <- sd_phylo^2 + replace(sd_sp^2, is.na(sd_sp), 0) + sd_resid^2
    h2_phylo  <- sd_phylo^2 / var_total

    post_s   <- brms::as_draws_df(fit)[["b_log_mass"]]
    pp_ge075 <- mean(post_s >= 0.75)
    pp_ge100 <- mean(post_s >= 1.00)

    list(
      group      = group_label,
      prior_tag  = prior_tag,
      n_sp       = n_sp,
      n_obs      = n_obs,
      formula    = re_formula,
      slope      = slope_m,
      ci_lo      = slope_lo,
      ci_hi      = slope_hi,
      rhat_max   = rhat_max,
      ess_min    = round(ess_min),
      sd_phylo   = sd_phylo,
      sd_sp      = sd_sp,
      sd_resid   = sd_resid,
      h2_phylo   = h2_phylo,
      pp_ge075   = pp_ge075,
      pp_ge100   = pp_ge100,
      fit        = fit,
      post_slope = post_s
    )
  }

  # ── Groups to fit ─────────────────────────────────────────────────────────
  pglmm_groups <- list(
    list(lf = "Woody",      part = "Shoot"),
    list(lf = "Woody",      part = "Root"),
    list(lf = "Herbaceous", part = "Shoot"),
    list(lf = "Herbaceous", part = "Root")
  )

  for (g in pglmm_groups) {
    label   <- paste(g$lf, "\u2014", g$part)
    sub_all <- dplyr::filter(dat,
                 life_form  == g$lf,
                 plant_part == g$part)

    # Primary: WBE-informed prior
    res_wbe <- fit_pglmm_brms(sub_all, pgls_tree, label,
                               prior_slope_mean = 0.75, prior_slope_sd = 0.5,
                               prior_tag = "wbe_informed")

    # Sensitivity: isometry-centred prior
    res_iso <- fit_pglmm_brms(sub_all, pgls_tree, label,
                               prior_slope_mean = 1.00, prior_slope_sd = 0.5,
                               prior_tag = "iso_informed")

    # DGP sensitivity: without Fagus crenata ontogenetic records
    if (n_ontogenetic > 0) {
      res_excl_ont <- fit_pglmm_brms(
        dplyr::filter(sub_all, dgp_type != "ontogenetic"),
        pgls_tree, paste0(label, " (excl. Fagus crenata ontog.)"),
        prior_slope_mean = 0.75, prior_slope_sd = 0.5,
        prior_tag = "excl_ontogenetic"
      )
    } else {
      res_excl_ont <- NULL
    }

    pglmm_brms_results[[label]]    <- res_wbe
    pglmm_brms_results[[paste0(label, "_iso")]]   <- res_iso
    if (!is.null(res_excl_ont))
      pglmm_brms_results[[paste0(label, "_excl")]] <- res_excl_ont
  }

  cat("brms PGLMM fits complete. Results stored in pglmm_brms_results.\n")

} else if (!HAS_BRMS) {
  cat("brms not installed. Install with: install.packages('brms')\n")
  cat("Once installed, re-render to run Bayesian PGLMM.\n")
  cat("Required also: Rtools (Windows) or Xcode CLI (macOS) for Stan compilation.\n")
} else if (USING_STAR_TREE) {
  cat("PGLMM skipped: star tree in use (V.PhyloMaker2 not available).\n")
  cat("Install V.PhyloMaker2: remotes::install_github('jinyizju/V.PhyloMaker2')\n")
}
```

## PGLMM results: primary slope estimates

```{r pglmm-results-table}
if (length(pglmm_brms_results) > 0) {

  # ── Primary results table (WBE-informed prior only) ─────────────────────────
  pglmm_primary <- purrr::keep(pglmm_brms_results,
                                ~ !is.null(.x) && .x$prior_tag == "wbe_informed") |>
    purrr::map_dfr(~ tibble::tibble(
      Group       = .x$group,
      N_sp        = .x$n_sp,
      N_obs       = .x$n_obs,
      Slope       = round(.x$slope, 3),
      CI_lo       = round(.x$ci_lo, 3),
      CI_hi       = round(.x$ci_hi, 3),
      h2_phylo    = round(.x$h2_phylo, 3),
      Rhat_max    = round(.x$rhat_max, 4),
      ESS_min     = .x$ess_min,
      PP_ge_WBE   = round(.x$pp_ge075, 3),
      PP_ge_iso   = round(.x$pp_ge100, 3)
    ))

  kable(
    pglmm_primary,
    col.names = c("Group", "N sp", "N obs", "β (post. mean)",
                  "95% CrI lo", "95% CrI hi",
                  "h² phylo", "Rhat max", "ESS min",
                  "P(β ≥ 0.75)", "P(β ≥ 1.0)"),
    caption = paste(
      "Bayesian PGLMM slopes (brms): log respiration ~ log fresh mass + dataset source.",
      "β = posterior mean; 95% CrI = credible interval.",
      "h² phylo = σ²_phylo / σ²_total (phylogenetic heritability).",
      "P(β ≥ 0.75) = posterior probability of WBE slope or steeper;",
      "P(β ≥ 1.0) = posterior probability of isometric slope or steeper.",
      "Prior: Normal(0.75, 0.5) on slope.",
      "Convergence: Rhat_max < 1.01 and ESS_min > 400 required."
    )
  )
} else {
  cat("No PGLMM results available (brms not installed or star tree in use).\n")
}
```

## Prior sensitivity

```{r pglmm-prior-sensitivity}
if (length(pglmm_brms_results) > 0) {
  # Compare WBE-informed vs isometry-informed prior slopes
  prior_compare <- purrr::keep(pglmm_brms_results,
                                ~ !is.null(.x) &&
                                    .x$prior_tag %in% c("wbe_informed", "iso_informed")) |>
    purrr::map_dfr(~ tibble::tibble(
      Group     = .x$group,
      Prior     = ifelse(.x$prior_tag == "wbe_informed",
                         "Normal(0.75, 0.5)", "Normal(1.0, 0.5)"),
      Slope     = round(.x$slope, 3),
      CI_lo     = round(.x$ci_lo, 3),
      CI_hi     = round(.x$ci_hi, 3)
    ))

  kable(
    prior_compare,
    caption = "Prior sensitivity: posterior mean slope ± 95% CrI under WBE-informed vs isometry-informed priors on β. If slopes are similar, the posterior is data-driven and prior-robust."
  )
} else {
  cat("Prior sensitivity analysis requires brms results.\n")
}
```

## PGLMM vs PGLS vs published: comparison forest plot

```{r pglmm-comparison-forest, fig.width=11, fig.height=6, fig.cap="Comparison of scaling exponent estimates across methods. Circles = brms PGLMM posterior mean ± 95% CrI (WBE-informed prior); triangles = caper PGLS on species means ± 95% CI; squares = Kurosawa et al. 2025 published Bayesian PGLMM slope (UNVERIFIED placeholder — verify against paper Table 1 before publication use). Red dashed = WBE 3/4; blue dotted = isometric 1.0."}
if (length(pglmm_brms_results) > 0 && nrow(pgls_results) > 0) {

  # PGLMM primary
  pglmm_pts <- purrr::keep(pglmm_brms_results,
                            ~ !is.null(.x) && .x$prior_tag == "wbe_informed") |>
    purrr::map_dfr(~ tibble::tibble(
      group  = .x$group,
      slope  = .x$slope,
      ci_lo  = .x$ci_lo,
      ci_hi  = .x$ci_hi,
      method = "PGLMM (brms)"
    ))

  # PGLS
  pgls_pts <- pgls_results |>
    dplyr::filter(method == "PGLS", group %in% pglmm_pts$group) |>
    dplyr::select(group, slope, ci_lo = ci_lower, ci_hi = ci_upper) |>
    dplyr::mutate(method = "PGLS (caper)")

  # Published (placeholder)
  pub_pts <- kurosawa_published |>
    dplyr::rename(group = group, slope = pub_slope,
                  ci_lo = pub_ci_lo, ci_hi = pub_ci_hi) |>
    dplyr::mutate(method = "Kurosawa 2025 (published\nUNVERIFIED)")

  all_pts <- dplyr::bind_rows(pglmm_pts, pgls_pts, pub_pts) |>
    dplyr::filter(!is.na(slope)) |>
    dplyr::mutate(
      method = factor(method,
                      levels = c("PGLMM (brms)", "PGLS (caper)",
                                 "Kurosawa 2025 (published\nUNVERIFIED)")),
      y_pos = as.numeric(factor(group)) +
              dplyr::case_when(
                method == "PGLMM (brms)"                      ~  0.22,
                method == "PGLS (caper)"                      ~  0.00,
                TRUE                                           ~ -0.22
              )
    )

  ggplot(all_pts, aes(x = slope, y = y_pos,
                      colour = method, shape = method)) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
               linewidth = 0.9) +
    geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
               linewidth = 0.9) +
    annotate("text", x = 0.75, y = Inf, label = " WBE\n 3/4",
             vjust = 1.3, colour = "#d6604d", size = 3.2, hjust = 0) +
    annotate("text", x = 1.00, y = Inf, label = " Iso\n 1.0",
             vjust = 1.3, colour = "#2166ac", size = 3.2, hjust = 0) +
    geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                   height = 0.1, linewidth = 0.7, na.rm = TRUE) +
    geom_point(size = 3.5) +
    scale_y_continuous(
      breaks = seq_len(dplyr::n_distinct(all_pts$group)),
      labels = levels(factor(all_pts$group)),
      expand = expansion(add = 0.6)
    ) +
    scale_colour_manual(values = c("PGLMM (brms)"                    = "#1b7837",
                                   "PGLS (caper)"                    = "#762a83",
                                   "Kurosawa 2025 (published\nUNVERIFIED)" = "#d6604d")) +
    scale_shape_manual(values = c("PGLMM (brms)"                    = 16,
                                  "PGLS (caper)"                    = 17,
                                  "Kurosawa 2025 (published\nUNVERIFIED)" = 15)) +
    labs(
      x      = "Scaling exponent  (log R ~ log Mass)",
      y      = NULL,
      colour = "Method",
      shape  = "Method",
      title  = "Respiration scaling exponent: PGLMM vs PGLS vs Kurosawa 2025",
      subtitle = "PGLMM: individual-level data + phylogenetic covariance (brms)  |  PGLS: species means (caper)\nPublished values are UNVERIFIED placeholder — verify against Kurosawa et al. 2025 Table 1"
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom")

} else if (nrow(pgls_results) > 0) {
  # PGLS-only forest when brms not available
  cat("brms PGLMM not available. Showing PGLS vs published comparison only.\n")

  pgls_pub <- dplyr::bind_rows(
    pgls_results |>
      dplyr::filter(method == "PGLS") |>
      dplyr::select(group, slope, ci_lo = ci_lower, ci_hi = ci_upper) |>
      dplyr::mutate(method = "PGLS (caper)"),
    kurosawa_published |>
      dplyr::rename(group = group, slope = pub_slope,
                    ci_lo = pub_ci_lo, ci_hi = pub_ci_hi) |>
      dplyr::mutate(method = "Kurosawa 2025 (UNVERIFIED)")
  ) |>
    dplyr::filter(!is.na(slope)) |>
    dplyr::mutate(
      y_pos = as.numeric(factor(group)) +
              ifelse(method == "PGLS (caper)", 0.15, -0.15)
    )

  ggplot(pgls_pub, aes(x = slope, y = y_pos, colour = method, shape = method)) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d") +
    geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac") +
    geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.1, na.rm = TRUE) +
    geom_point(size = 3) +
    scale_y_continuous(
      breaks = seq_len(dplyr::n_distinct(pgls_pub$group)),
      labels = levels(factor(pgls_pub$group)),
      expand = expansion(add = 0.5)
    ) +
    scale_colour_manual(values = c("PGLS (caper)" = "#762a83",
                                   "Kurosawa 2025 (UNVERIFIED)" = "#d6604d")) +
    scale_shape_manual(values = c("PGLS (caper)" = 17,
                                  "Kurosawa 2025 (UNVERIFIED)" = 15)) +
    labs(x = "Scaling exponent", y = NULL,
         title = "PGLS vs Kurosawa 2025 published (install brms for full PGLMM)",
         subtitle = "UNVERIFIED placeholder slopes — verify against Table 1") +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom")
}
```

## Posterior predictive checks and convergence

```{r pglmm-diagnostics, fig.width=10, fig.height=8}
if (length(pglmm_brms_results) > 0) {
  # Posterior predictive check for first group with a fit
  primary_fits <- purrr::keep(pglmm_brms_results,
                               ~ !is.null(.x) && .x$prior_tag == "wbe_informed")

  if (length(primary_fits) > 0) {
    first_fit <- primary_fits[[1]]$fit
    first_label <- primary_fits[[1]]$group

    cat(sprintf("Posterior predictive check for: %s\n", first_label))
    print(brms::pp_check(first_fit, ndraws = 100) +
            labs(title = paste("Posterior predictive check:", first_label),
                 subtitle = "Dark line = observed; light lines = posterior predictive draws") +
            theme_bw(base_size = 12))

    # Slope posterior density overlay (WBE vs isometry priors)
    wbe_post <- primary_fits[[1]]$post_slope
    iso_key  <- paste0(primary_fits[[1]]$group, "_iso")
    if (!is.null(pglmm_brms_results[[iso_key]])) {
      iso_post <- pglmm_brms_results[[iso_key]]$post_slope
      post_df  <- tibble::tibble(
        slope = c(wbe_post, iso_post),
        prior = rep(c("Normal(0.75, 0.5)", "Normal(1.0, 0.5)"),
                    c(length(wbe_post), length(iso_post)))
      )
      p_dens <- ggplot(post_df, aes(x = slope, fill = prior)) +
        geom_density(alpha = 0.5) +
        geom_vline(xintercept = 0.75, linetype = "dashed",
                   colour = "#d6604d", linewidth = 1) +
        geom_vline(xintercept = 1.00, linetype = "dotted",
                   colour = "#2166ac", linewidth = 1) +
        scale_fill_manual(values = c("Normal(0.75, 0.5)" = "#4dac26",
                                     "Normal(1.0, 0.5)"  = "#b8860b")) +
        labs(x = "Posterior slope  (log R ~ log Mass)",
             y = "Density",
             fill = "Prior on β",
             title = paste("Prior sensitivity: posterior slope density —", first_label),
             subtitle = "If curves overlap, posterior is prior-robust (data-driven)") +
        theme_bw(base_size = 12)
      print(p_dens)
    }
  }
} else {
  cat("No brms fits available for diagnostic plots.\n")
}
```

---

# Inter- vs. intraspecific scaling decomposition

Interspecific scaling (across species means) and intraspecific scaling (within
a species across individuals of varying size) represent distinct biological
processes. WBE/MST predicts $\beta_{\text{inter}} \approx 3/4$ from conserved
vascular network geometry. Intraspecific slopes ($\beta_{\text{intra}}$) may
differ: Glazier (2005, *Biol Rev* 80: 611–662) predicts intraspecific slopes
closer to 1 for small organisms (surface-volume metabolic level hypothesis)
converging toward 3/4 only when fractal network constraints dominate.

> **⚠ Interpretive caveat:** Per-species slopes estimated here from multiple
> individuals of the same species conflate **ontogenetic allometry**
> (developmental biomass allocation) with **size variation among adults**
> (population variation in final size). Only the *Fagus crenata*
> (Kurosawa 2023) subset constitutes a genuine ontogenetic series. All other
> per-species slopes should be interpreted as "size-variation slopes," not
> developmental allometry.

## Per-species intraspecific slopes

```{r intraspecific-slopes}
# Filter to species with at least 3 individual records (required for meaningful slope)
sp_intra <- dat |>
  group_by(species, family, life_form, plant_part) |>
  filter(is.finite(log_mass), is.finite(log_resp)) |>  # finite check first
  filter(n() >= 3) |>                                    # then re-count valid rows
  nest() |>
  mutate(
    lm_fit    = map(data, ~ lm(log_resp ~ log_mass, data = .x)),
    sma_fit   = map(data, ~ tryCatch(smatr::sma(log_resp ~ log_mass, data = .x,
                                                 method = "SMA"),
                                     error = function(e) NULL)),
    n         = map_int(data, nrow),
    slope     = map_dbl(lm_fit, ~ coef(.x)[2]),
    slope_se  = map_dbl(lm_fit, ~ summary(.x)$coefficients[2, 2]),
    ci_lo     = map_dbl(lm_fit, ~ tryCatch(confint(.x)[2, 1], error = function(e) NA_real_)),
    ci_hi     = map_dbl(lm_fit, ~ tryCatch(confint(.x)[2, 2], error = function(e) NA_real_)),
    r2        = map_dbl(lm_fit, ~ summary(.x)$r.squared),
    sma_slope = map_dbl(sma_fit, ~ if (!is.null(.x)) coef(.x)[2] else NA_real_),
    sma_ci_lo = map_dbl(sma_fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 2] else NA_real_),
    sma_ci_hi = map_dbl(sma_fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 3] else NA_real_)
  ) |>
  select(-lm_fit, -sma_fit, -data) |>
  ungroup()

cat("Species with n >= 3 (intraspecific slopes):", nrow(sp_intra), "\n")
cat("\nOLS exponents:\n")
cat("  Median:", round(median(sp_intra$slope,     na.rm = TRUE), 3), "\n")
cat("  IQR:   ", round(IQR(sp_intra$slope,         na.rm = TRUE), 3), "\n")
cat("\nSMA exponents:\n")
cat("  Median:", round(median(sp_intra$sma_slope,  na.rm = TRUE), 3), "\n")
cat("  IQR:   ", round(IQR(sp_intra$sma_slope,    na.rm = TRUE), 3), "\n")

# One-sample t-tests: are intraspecific slopes centered on 0.75 or 1.0?
t_vs_075     <- t.test(sp_intra$slope,     mu = 0.75)
t_vs_100     <- t.test(sp_intra$slope,     mu = 1.00)
sma_t_vs_075 <- t.test(sp_intra$sma_slope, mu = 0.75)
sma_t_vs_100 <- t.test(sp_intra$sma_slope, mu = 1.00)
cat(sprintf("\nOLS one-sample t-test, H0: mean = 0.75 -> p = %.4f\n", t_vs_075$p.value))
cat(sprintf("OLS one-sample t-test, H0: mean = 1.00 -> p = %.4f\n", t_vs_100$p.value))
cat(sprintf("\nSMA one-sample t-test, H0: mean = 0.75 -> p = %.4f\n", sma_t_vs_075$p.value))
cat(sprintf("SMA one-sample t-test, H0: mean = 1.00 -> p = %.4f\n", sma_t_vs_100$p.value))
```

```{r intraspecific-histogram, fig.width=12, fig.height=8, fig.cap="Frequency distributions of per-species intraspecific scaling exponents. Rows: OLS (top) and SMA (bottom). Columns: life form. Fill colour distinguishes plant part (Shoot = green, Root = amber). Vertical red dashed line = WBE 3/4; blue dotted line = isometric 1.0; grey solid line = per-panel median. Each bar = one species (n\u22653 individuals). SMA exponents are systematically larger than OLS when within-species R\u00b2 < 1 because SMA minimises residuals in both axes."}
sp_long <- sp_intra |>
  select(species, family, life_form, plant_part,
         OLS = slope, SMA = sma_slope) |>
  pivot_longer(cols = c(OLS, SMA),
               names_to = "method", values_to = "exponent") |>
  filter(!is.na(exponent)) |>
  mutate(method = factor(method, levels = c("OLS", "SMA")))

# Per-panel summary: median and N for annotation
panel_stats <- sp_long |>
  group_by(method, life_form) |>
  summarise(
    median_exp = median(exponent, na.rm = TRUE),
    n_sp       = n(),
    .groups    = "drop"
  )

ggplot(sp_long, aes(x = exponent, fill = plant_part)) +
  geom_histogram(bins = 25, alpha = 0.75, colour = "white") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
             linewidth = 1) +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
             linewidth = 1) +
  geom_vline(data = panel_stats,
             aes(xintercept = median_exp),
             linetype = "solid", colour = "grey30", linewidth = 0.9,
             inherit.aes = FALSE) +
  geom_label(data = panel_stats,
             aes(x = median_exp, y = Inf,
                 label = sprintf("med=%.2f\nn=%d", median_exp, n_sp)),
             vjust = 1.4, hjust = -0.08, size = 3.0, colour = "grey20",
             fill = "white", label.size = 0.2, inherit.aes = FALSE) +
  facet_grid(method ~ life_form, scales = "free_y") +
  scale_fill_manual(values = c(Shoot = "#4dac26", Root = "#b8860b")) +
  labs(x = "Scaling exponent (log R ~ log Mass, within species)",
       y = "Count",
       fill = "Plant part",
       title = "Distribution of intraspecific scaling exponents: OLS vs SMA",
       subtitle = "Each bar = one species (n\u22653 individuals). Grey line = per-panel median.") +
  theme(legend.position = "bottom")
```

```{r intraspecific-distrib-summary}
# ── Distributional summary table: OLS and SMA per life-form × plant-part ──────
distrib_summary <- sp_long |>
  group_by(Method = method, `Life form` = life_form, `Plant part` = plant_part) |>
  summarise(
    N         = n(),
    Mean      = round(mean(exponent, na.rm = TRUE), 3),
    SD        = round(sd(exponent, na.rm = TRUE), 3),
    Median    = round(median(exponent, na.rm = TRUE), 3),
    IQR       = round(IQR(exponent, na.rm = TRUE), 3),
    Min       = round(min(exponent, na.rm = TRUE), 3),
    Max       = round(max(exponent, na.rm = TRUE), 3),
    `% > 0.75` = round(100 * mean(exponent > 0.75, na.rm = TRUE), 1),
    `% > 1.00` = round(100 * mean(exponent > 1.00, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(
  distrib_summary,
  caption = "Summary statistics of intraspecific scaling exponent distributions by method, life form, and plant part. '% > 0.75': fraction of species exceeding the WBE 3/4 prediction. '% > 1.00': fraction exceeding isometry."
)
```

```{r intraspecific-table}
kable(
  sp_intra |>
    arrange(family, species) |>
    select(species, family, life_form, plant_part, n,
           slope, ci_lo, ci_hi, r2,
           sma_slope, sma_ci_lo, sma_ci_hi) |>
    mutate(across(c(slope, ci_lo, ci_hi, r2,
                    sma_slope, sma_ci_lo, sma_ci_hi), ~ round(.x, 3))) |>
    rename(Species = species, Family = family, `Life form` = life_form,
           `Plant part` = plant_part, N = n,
           `OLS slope` = slope, `OLS CI lo` = ci_lo, `OLS CI hi` = ci_hi, `R²` = r2,
           `SMA slope` = sma_slope, `SMA CI lo` = sma_ci_lo, `SMA CI hi` = sma_ci_hi),
  caption = "Per-species intraspecific OLS and SMA scaling exponents (n\u22653 individuals required). OLS 95% CI from lm(); SMA 95% CI from smatr::sma()."
)
```

## Per-species 95% CI forest plot

```{r intraspecific-ci-plot, fig.width=10, fig.height=7, fig.cap="Per-species 95% confidence intervals for intraspecific scaling exponents (log R ~ log fresh mass). Circles = OLS, triangles = SMA, offset vertically within each facet. Species sorted by ascending OLS slope within each life-form \u00d7 plant-part group. Red dashed line = WBE \u00be (\u03b2 = 0.75); blue dotted line = isometric (\u03b2 = 1.0). Wide CIs reflect small sample size or narrow mass range within a species; narrow CIs indicate well-constrained slopes. Only species with n \u2265 3 individuals are shown."}
# ── Build long-format CI data sorted by OLS slope within group ────────────────
ci_plot_df <- sp_intra |>
  mutate(
    facet_grp    = paste0(life_form, "\n", plant_part),
    ols_sort_key = slope
  ) |>
  arrange(facet_grp, ols_sort_key) |>
  mutate(sp_label = factor(species, levels = unique(species))) |>
  pivot_longer(
    cols      = c(slope, sma_slope),
    names_to  = "method",
    values_to = "est"
  ) |>
  mutate(
    lo     = if_else(method == "slope", ci_lo,     sma_ci_lo),
    hi     = if_else(method == "slope", ci_hi,     sma_ci_hi),
    method = if_else(method == "slope", "OLS", "SMA"),
    method = factor(method, levels = c("OLS", "SMA"))
  ) |>
  filter(!is.na(est), !is.na(lo), !is.na(hi))

ggplot(ci_plot_df,
       aes(x = est, y = sp_label, colour = method, shape = method)) +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
             linewidth = 0.8) +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
             linewidth = 0.8) +
  geom_errorbarh(aes(xmin = lo, xmax = hi),
                 height   = 0.30,
                 linewidth = 0.55,
                 alpha    = 0.8,
                 position = position_dodge(width = 0.55)) +
  geom_point(size     = 2.4,
             position = position_dodge(width = 0.55)) +
  facet_wrap(~ facet_grp, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c(OLS = "#1a6faf", SMA = "#d95f02"),
                      name   = "Method") +
  scale_shape_manual(values  = c(OLS = 16, SMA = 17),
                     name    = "Method") +
  labs(
    x        = expression(paste("Scaling exponent  ", beta, "  (log R ~ log M, within species)")),
    y        = NULL,
    title    = "Per-species intraspecific 95% CI forest plot",
    subtitle = "Sorted by OLS slope within group. Red dashed = WBE \u00be, blue dotted = isometric 1.0."
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position    = "bottom",
    strip.background   = element_rect(fill = "grey92"),
    axis.text.y        = element_text(size = 8, face = "italic"),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3),
    panel.grid.minor   = element_blank()
  )
```

## Intraspecific slopes: n ≥ 10 and ≥ 2 orders of magnitude size range

This subsection restricts the per-species intraspecific analysis to species
that have (i) at least **10** individual measurements and (ii) a fresh-mass
range spanning **≥ 2 orders of magnitude** (i.e., max − min log₁₀ fresh mass ≥ 2,
equivalent to a 100-fold body-size span). These criteria select species for
which within-species slopes are both well-powered and ecologically meaningful.

```{r intraspecific-strict-slopes}
# ── Build per-species stats including mass range ──────────────────────────────
sp_intra_strict <- dat |>
  group_by(species, family, life_form, plant_part) |>
  filter(is.finite(log_mass), is.finite(log_resp)) |>
  mutate(mass_range = max(log_mass) - min(log_mass)) |>
  filter(n() >= 10, mass_range >= 2) |>          # n >= 10; >= 2 log10 units
  nest() |>
  mutate(
    lm_fit    = map(data, ~ lm(log_resp ~ log_mass, data = .x)),
    sma_fit   = map(data, ~ tryCatch(smatr::sma(log_resp ~ log_mass, data = .x,
                                                 method = "SMA"),
                                     error = function(e) NULL)),
    n          = map_int(data, nrow),
    mass_range = map_dbl(data, ~ max(.x$log_mass) - min(.x$log_mass)),
    slope      = map_dbl(lm_fit, ~ coef(.x)[2]),
    slope_se   = map_dbl(lm_fit, ~ summary(.x)$coefficients[2, 2]),
    ci_lo      = map_dbl(lm_fit, ~ tryCatch(confint(.x)[2, 1], error = function(e) NA_real_)),
    ci_hi      = map_dbl(lm_fit, ~ tryCatch(confint(.x)[2, 2], error = function(e) NA_real_)),
    r2         = map_dbl(lm_fit, ~ summary(.x)$r.squared),
    sma_slope  = map_dbl(sma_fit, ~ if (!is.null(.x)) coef(.x)[2] else NA_real_),
    sma_ci_lo  = map_dbl(sma_fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 2] else NA_real_),
    sma_ci_hi  = map_dbl(sma_fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 3] else NA_real_)
  ) |>
  select(-lm_fit, -sma_fit, -data) |>
  ungroup()

cat(sprintf("Species passing n >= 10 AND mass range >= 2 log10 units: %d\n",
            nrow(sp_intra_strict)))
cat(sprintf("  (dropped from n >= 3 full set: %d → %d species)\n",
            nrow(sp_intra), nrow(sp_intra_strict)))

if (nrow(sp_intra_strict) > 0) {
  cat("\nOLS exponents (strict subset):\n")
  cat("  Mean:  ", round(mean(sp_intra_strict$slope,     na.rm = TRUE), 3), "\n")
  cat("  Median:", round(median(sp_intra_strict$slope,   na.rm = TRUE), 3), "\n")
  cat("  IQR:   ", round(IQR(sp_intra_strict$slope,      na.rm = TRUE), 3), "\n")
  cat("\nSMA exponents (strict subset):\n")
  cat("  Mean:  ", round(mean(sp_intra_strict$sma_slope, na.rm = TRUE), 3), "\n")
  cat("  Median:", round(median(sp_intra_strict$sma_slope, na.rm = TRUE), 3), "\n")
  cat("  IQR:   ", round(IQR(sp_intra_strict$sma_slope,  na.rm = TRUE), 3), "\n")

  # One-sample t-tests vs theoretical exponents
  ts_ols_075  <- t.test(sp_intra_strict$slope,     mu = 0.75)
  ts_ols_100  <- t.test(sp_intra_strict$slope,     mu = 1.00)
  ts_sma_075  <- t.test(sp_intra_strict$sma_slope, mu = 0.75)
  ts_sma_100  <- t.test(sp_intra_strict$sma_slope, mu = 1.00)

  cat(sprintf("\nOLS mean = %.3f  [95%% CI: %.3f – %.3f]  n = %d\n",
              ts_ols_075$estimate, ts_ols_075$conf.int[1], ts_ols_075$conf.int[2],
              length(sp_intra_strict$slope[is.finite(sp_intra_strict$slope)])))
  cat(sprintf("OLS vs H0=0.75: t = %.3f, df = %.1f, p = %.4f\n",
              ts_ols_075$statistic, ts_ols_075$parameter, ts_ols_075$p.value))
  cat(sprintf("OLS vs H0=1.00: t = %.3f, df = %.1f, p = %.4f\n",
              ts_ols_100$statistic, ts_ols_100$parameter, ts_ols_100$p.value))

  cat(sprintf("\nSMA mean = %.3f  [95%% CI: %.3f – %.3f]  n = %d\n",
              ts_sma_075$estimate, ts_sma_075$conf.int[1], ts_sma_075$conf.int[2],
              length(sp_intra_strict$sma_slope[is.finite(sp_intra_strict$sma_slope)])))
  cat(sprintf("SMA vs H0=0.75: t = %.3f, df = %.1f, p = %.4f\n",
              ts_sma_075$statistic, ts_sma_075$parameter, ts_sma_075$p.value))
  cat(sprintf("SMA vs H0=1.00: t = %.3f, df = %.1f, p = %.4f\n",
              ts_sma_100$statistic, ts_sma_100$parameter, ts_sma_100$p.value))
} else {
  cat("\n⚠ No species pass both criteria in the current dataset.\n")
  cat("  (Demo data have narrow mass ranges by design; re-run with real Dryad data.)\n")
}
```

```{r intraspecific-strict-histogram, fig.width=12, fig.height=8, fig.cap="Frequency distributions of intraspecific scaling exponents for the high-quality subset: species with ≥ 10 individuals spanning ≥ 2 orders of magnitude in fresh mass. Rows: OLS (top) and SMA (bottom). Columns: life form. Fill = plant part (Shoot = green, Root = amber). Red dashed = WBE ¾; blue dotted = isometric 1.0; grey solid = per-panel median."}
if (nrow(sp_intra_strict) < 2) {
  message("⚠ Too few species in strict subset for histogram (need ≥ 2). ",
          "Re-run with real Dryad data.")
} else {
  strict_long <- sp_intra_strict |>
    select(species, family, life_form, plant_part,
           OLS = slope, SMA = sma_slope) |>
    pivot_longer(cols = c(OLS, SMA),
                 names_to = "method", values_to = "exponent") |>
    filter(!is.na(exponent)) |>
    mutate(method = factor(method, levels = c("OLS", "SMA")))

  strict_panel_stats <- strict_long |>
    group_by(method, life_form) |>
    summarise(
      median_exp = median(exponent, na.rm = TRUE),
      n_sp       = n(),
      .groups    = "drop"
    )

  print(
    ggplot(strict_long, aes(x = exponent, fill = plant_part)) +
      geom_histogram(bins = 20, alpha = 0.75, colour = "white") +
      geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
                 linewidth = 1) +
      geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
                 linewidth = 1) +
      geom_vline(data = strict_panel_stats,
                 aes(xintercept = median_exp),
                 linetype = "solid", colour = "grey30", linewidth = 0.9,
                 inherit.aes = FALSE) +
      geom_label(data = strict_panel_stats,
                 aes(x = median_exp, y = Inf,
                     label = sprintf("med=%.2f\nn=%d", median_exp, n_sp)),
                 vjust = 1.4, hjust = -0.08, size = 3.0, colour = "grey20",
                 fill = "white", label.size = 0.2, inherit.aes = FALSE) +
      facet_grid(method ~ life_form, scales = "free_y") +
      scale_fill_manual(values = c(Shoot = "#4dac26", Root = "#b8860b")) +
      labs(x = "Scaling exponent (log R ~ log Mass, within species)",
           y = "Count",
           fill = "Plant part",
           title = "Intraspecific exponents — strict subset (n ≥ 10, range ≥ 2 log₁₀ units)",
           subtitle = "Grey line = per-panel median. Red dashed = WBE ¾. Blue dotted = isometric 1.0.") +
      theme(legend.position = "bottom")
  )
}
```

```{r intraspecific-strict-forest, fig.width=10, fig.height=7, fig.cap="Per-species 95% confidence intervals for intraspecific scaling exponents — strict subset (≥ 10 individuals, ≥ 2 log₁₀ units mass range). Circles = OLS, triangles = SMA, vertically dodged within each facet. Species sorted by ascending OLS slope. Red dashed = WBE ¾; blue dotted = isometric 1.0. Wider mass ranges generally yield narrower CIs."}
if (nrow(sp_intra_strict) < 2) {
  message("⚠ Too few species in strict subset for forest plot.")
} else {
  strict_ci_df <- sp_intra_strict |>
    mutate(
      facet_grp    = paste0(life_form, "\n", plant_part),
      ols_sort_key = slope
    ) |>
    arrange(facet_grp, ols_sort_key) |>
    mutate(sp_label = factor(species, levels = unique(species))) |>
    pivot_longer(
      cols      = c(slope, sma_slope),
      names_to  = "method",
      values_to = "est"
    ) |>
    mutate(
      lo     = if_else(method == "slope", ci_lo,    sma_ci_lo),
      hi     = if_else(method == "slope", ci_hi,    sma_ci_hi),
      method = if_else(method == "slope", "OLS", "SMA"),
      method = factor(method, levels = c("OLS", "SMA"))
    ) |>
    filter(!is.na(est), !is.na(lo), !is.na(hi))

  print(
    ggplot(strict_ci_df,
           aes(x = est, y = sp_label, colour = method, shape = method)) +
      geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
                 linewidth = 0.8) +
      geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
                 linewidth = 0.8) +
      geom_errorbarh(aes(xmin = lo, xmax = hi),
                     height    = 0.30,
                     linewidth = 0.55,
                     alpha     = 0.8,
                     position  = position_dodge(width = 0.55)) +
      geom_point(size     = 2.4,
                 position = position_dodge(width = 0.55)) +
      facet_wrap(~ facet_grp, scales = "free_y", ncol = 2) +
      scale_colour_manual(values = c(OLS = "#1a6faf", SMA = "#d95f02"),
                          name   = "Method") +
      scale_shape_manual(values  = c(OLS = 16, SMA = 17),
                         name    = "Method") +
      labs(
        x        = expression(paste("Scaling exponent  ", beta,
                                    "  (log R ~ log M, within species)")),
        y        = NULL,
        title    = "Intraspecific exponents — strict subset (n ≥ 10, ≥ 2 log₁₀ units)",
        subtitle = "Sorted by OLS slope. Red dashed = WBE ¾, blue dotted = isometric 1.0."
      ) +
      theme_bw(base_size = 11) +
      theme(
        legend.position    = "bottom",
        strip.background   = element_rect(fill = "grey92"),
        axis.text.y        = element_text(size = 8, face = "italic"),
        panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3),
        panel.grid.minor   = element_blank()
      )
  )
}
```

## Inter- vs. intraspecific: Glazier comparison

```{r glazier-comparison}
# ── Helper: mean + 95% CI via t-distribution ─────────────────────────────────
mean_ci <- function(x) {
  x  <- x[is.finite(x)]
  n  <- length(x)
  if (n < 2) return(list(mean = mean(x), ci_lo = NA_real_, ci_hi = NA_real_))
  mn <- mean(x)
  se <- sd(x) / sqrt(n)
  tc <- qt(0.975, df = n - 1)
  list(n = n, mean = mn, ci_lo = mn - tc * se, ci_hi = mn + tc * se)
}

# ── Per-group OLS and SMA summaries ──────────────────────────────────────────
glazier_tbl <- sp_intra |>
  mutate(group = paste(life_form, "\u2014", plant_part)) |>
  group_by(group, life_form, plant_part) |>
  summarise(
    n_species        = n(),
    # OLS
    ols_mean         = mean_ci(slope)$mean,
    ols_ci_lo        = mean_ci(slope)$ci_lo,
    ols_ci_hi        = mean_ci(slope)$ci_hi,
    ols_median       = median(slope, na.rm = TRUE),
    ols_iqr          = IQR(slope, na.rm = TRUE),
    # OLS one-sample t-tests
    ols_p_075        = t.test(slope[is.finite(slope)], mu = 0.75)$p.value,
    ols_p_100        = t.test(slope[is.finite(slope)], mu = 1.00)$p.value,
    # SMA
    sma_mean         = mean_ci(sma_slope)$mean,
    sma_ci_lo        = mean_ci(sma_slope)$ci_lo,
    sma_ci_hi        = mean_ci(sma_slope)$ci_hi,
    sma_median       = median(sma_slope, na.rm = TRUE),
    sma_iqr          = IQR(sma_slope, na.rm = TRUE),
    sma_p_075        = t.test(sma_slope[is.finite(sma_slope)], mu = 0.75)$p.value,
    sma_p_100        = t.test(sma_slope[is.finite(sma_slope)], mu = 1.00)$p.value,
    .groups = "drop"
  )

# Attach PGLS interspecific slopes for comparison
if (nrow(pgls_results) > 0) {
  pgls_inter <- pgls_results |>
    filter(method == "PGLS") |>
    select(group, inter_slope_pgls = slope)
  glazier_tbl <- left_join(glazier_tbl, pgls_inter, by = "group")
} else {
  glazier_tbl$inter_slope_pgls <- NA_real_
}

# ── Print per-group t-test results ───────────────────────────────────────────
cat("=== OLS mean exponent: one-sample t-tests per group ===\n")
for (i in seq_len(nrow(glazier_tbl))) {
  g <- glazier_tbl$group[i]
  cat(sprintf("\n%s  (n=%d)\n", g, glazier_tbl$n_species[i]))
  cat(sprintf("  OLS mean = %.3f  [%.3f, %.3f]\n",
              glazier_tbl$ols_mean[i], glazier_tbl$ols_ci_lo[i], glazier_tbl$ols_ci_hi[i]))
  cat(sprintf("  H0: mean = 0.75 -> p = %.4f  |  H0: mean = 1.00 -> p = %.4f\n",
              glazier_tbl$ols_p_075[i], glazier_tbl$ols_p_100[i]))
  cat(sprintf("  SMA mean = %.3f  [%.3f, %.3f]\n",
              glazier_tbl$sma_mean[i], glazier_tbl$sma_ci_lo[i], glazier_tbl$sma_ci_hi[i]))
  cat(sprintf("  H0: mean = 0.75 -> p = %.4f  |  H0: mean = 1.00 -> p = %.4f\n",
              glazier_tbl$sma_p_075[i], glazier_tbl$sma_p_100[i]))
}

# ── Summary table ─────────────────────────────────────────────────────────────
kable(
  glazier_tbl |>
    select(Group = group, N = n_species,
           `OLS mean` = ols_mean, `OLS CI lo` = ols_ci_lo, `OLS CI hi` = ols_ci_hi,
           `OLS median` = ols_median, `OLS IQR` = ols_iqr,
           `p(=0.75)` = ols_p_075, `p(=1.00)` = ols_p_100,
           `SMA mean` = sma_mean, `SMA CI lo` = sma_ci_lo, `SMA CI hi` = sma_ci_hi,
           `SMA median` = sma_median, `SMA IQR` = sma_iqr,
           `β_inter PGLS` = inter_slope_pgls) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Per-group mean intraspecific scaling exponents (OLS and SMA) with 95% CIs of the mean (t-distribution). p(=0.75): H₀ mean = WBE 3/4; p(=1.00): H₀ mean = isometry. Glazier (2005) predicts β_intra > β_inter."
)
```

```{r glazier-forest, fig.width=9, fig.height=5, fig.cap="Mean intraspecific scaling exponent ± 95% CI for each life-form × plant-part group, by estimator (OLS = circle, SMA = triangle). Red dashed line = WBE 3/4; blue dotted line = isometric 1.0. Points are offset vertically for clarity. 95% CIs are of the mean across species (t-distribution), not individual-species CIs."}
# Build long-format for forest plot
glaz_long <- glazier_tbl |>
  select(group, life_form, plant_part,
         OLS_mean   = ols_mean,  OLS_lo  = ols_ci_lo,  OLS_hi  = ols_ci_hi,
         SMA_mean   = sma_mean,  SMA_lo  = sma_ci_lo,  SMA_hi  = sma_ci_hi) |>
  pivot_longer(
    cols      = c(OLS_mean, SMA_mean),
    names_to  = "method",
    values_to = "mean_exp"
  ) |>
  mutate(
    ci_lo  = if_else(method == "OLS_mean", OLS_lo, SMA_lo),
    ci_hi  = if_else(method == "OLS_mean", OLS_hi, SMA_hi),
    method = sub("_mean", "", method),
    # Vertical dodge offset
    y_pos  = as.numeric(factor(group)) +
              if_else(method == "OLS", 0.15, -0.15)
  )

ggplot(glaz_long,
       aes(x = mean_exp, y = y_pos, colour = life_form, shape = method)) +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
             linewidth = 0.9) +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
             linewidth = 0.9) +
  annotate("text", x = 0.75, y = Inf, label = " WBE\n 3/4",
           vjust = 1.3, colour = "#d6604d", size = 3.2, hjust = 0) +
  annotate("text", x = 1.00, y = Inf, label = " Iso\n 1.0",
           vjust = 1.3, colour = "#2166ac", size = 3.2, hjust = 0) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 height = 0.12, linewidth = 0.7) +
  geom_point(size = 4) +
  scale_y_continuous(
    breaks = seq_len(nrow(glazier_tbl)),
    labels = glazier_tbl$group,
    expand = expansion(add = 0.6)
  ) +
  scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  scale_shape_manual(values = c(OLS = 16, SMA = 17)) +
  labs(
    x      = "Mean scaling exponent (log R ~ log Mass)",
    y      = NULL,
    colour = "Life form",
    shape  = "Estimator",
    title  = "Inter- vs. intraspecific scaling: mean exponent ± 95% CI",
    subtitle = "OLS \u25CF and SMA \u25B2 shown; 95% CI of the mean across species"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())
```

---

# Per-family scaling relationships

Family-level scaling tests whether the WBE/MST prediction of a universal
$\beta = 3/4$ holds within plant families, or whether families differ in
scaling exponent (vascular architecture differences) or metabolic normalization
(intercept differences).

> **⚠ Temperature caveat:** Family-level intercept differences may reflect
> measurement temperature differences across studies, not true metabolic
> normalization differences. Slope comparisons are less sensitive to
> temperature if thermal effects are additive in log space, but this
> cannot be assumed. Families sampled primarily from cold environments
> (e.g., Pinaceae) may show systematically lower respiration at the same
> fresh mass compared to warm-adapted families.

## Family-level SMA fits

```{r family-sma, fig.height=7, fig.cap="Forest plot of SMA slopes by family. Dashed line = WBE 3/4. Families ordered by slope. Minimum 5 species OR 10 records required for inclusion."}
# Apply eligibility thresholds (BC-5: n_species >= 5 OR n_records >= 10)
fam_eligible <- fam_summary |>
  filter(sma_eligible) |>
  pull(family)

dat_fam <- dat |>
  filter(family %in% fam_eligible, is.finite(log_mass), is.finite(log_resp))

fam_fits <- dat_fam |>
  group_by(family, life_form) |>
  filter(n_distinct(species) >= 3 | n() >= 10) |>   # secondary guard
  nest() |>
  mutate(
    n_records  = map_int(data, nrow),
    n_species  = map_int(data, ~ n_distinct(.x$species)),
    fit        = map(data, ~ tryCatch(smatr::sma(log_resp ~ log_mass, data = .x,
                                                  method = "SMA"), error = function(e) NULL)),
    slope      = map_dbl(fit, ~ if (!is.null(.x)) coef(.x)[2] else NA_real_),
    intercept  = map_dbl(fit, ~ if (!is.null(.x)) coef(.x)[1] else NA_real_),
    ci_lo      = map_dbl(fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 2] else NA_real_),
    ci_hi      = map_dbl(fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 3] else NA_real_),
    r2         = map_dbl(fit, ~ if (!is.null(.x)) .x$r2[[1]] else NA_real_),
    p_vs_075   = map_dbl(data, ~ tryCatch(
                   smatr::slope.test(.x$log_resp, .x$log_mass,
                                     test.value = 0.75, method = "SMA")$p,
                   error = function(e) NA_real_))
  ) |>
  select(-fit, -data) |>
  filter(!is.na(slope)) |>
  ungroup()

kable(
  fam_fits |>
    arrange(life_form, desc(n_species)) |>
    select(family, life_form, n_species, n_records, slope, ci_lo, ci_hi, r2, p_vs_075) |>
    mutate(across(c(slope, ci_lo, ci_hi, r2), ~ round(.x, 3)),
           p_vs_075 = ifelse(is.na(p_vs_075), "—",
                             ifelse(p_vs_075 < 0.001, "<0.001",
                                    as.character(round(p_vs_075, 3))))) |>
    rename(Family = family, `Life form` = life_form,
           `N species` = n_species, `N records` = n_records,
           `SMA slope` = slope, `CI low` = ci_lo, `CI high` = ci_hi,
           `R²` = r2, `p (slope ≠ 0.75)` = p_vs_075),
  caption = "Per-family SMA scaling exponents (families with n_species≥5 or n_records≥10)"
)
```

```{r family-forest-plot, fig.height=6, fig.cap="Per-family SMA scaling exponents with 95% CI. Dashed = WBE 3/4. Colored by life form (blue = woody, red = herbaceous)."}
ggplot(fam_fits, aes(y = reorder(family, slope), x = slope,
                     xmin = ci_lo, xmax = ci_hi, colour = life_form)) +
  geom_pointrange(size = 0.7) +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "grey40",
             linewidth = 0.8) +
  scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  labs(x = "SMA slope  (log R ~ log Fresh mass)",
       y = NULL,
       colour = "Life form",
       title = "Per-family respiration scaling exponents",
       subtitle = "Dashed = WBE 3/4 reference") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## Slope homogeneity test across families

```{r family-slope-homogeneity}
# Test whether all eligible families share a common slope
# smatr::sma with group interaction tests slope heterogeneity
if (n_distinct(dat_fam$family) >= 3) {
  fit_common <- tryCatch(
    smatr::sma(log_resp ~ log_mass + family, data = dat_fam, method = "SMA"),
    error = function(e) NULL
  )
  fit_interact <- tryCatch(
    smatr::sma(log_resp ~ log_mass * family, data = dat_fam, method = "SMA"),
    error = function(e) NULL
  )
  if (!is.null(fit_common) && !is.null(fit_interact)) {
    # smatr::sma objects do not store a $model slot; use the log-likelihood
    # approach instead: compare the common-slope vs. interaction models via
    # the smatr slope.test embedded in the interaction summary.
    cat("Common-slope vs. interaction summary (smatr slope.test):\n")
    cat("  (AIC comparison not available for smatr objects; see slope homogeneity test below)\n")
    cat("\nSlope homogeneity test (smatr):\n")
    print(summary(fit_interact)$coef)
  }
}
```

## Family intercepts: metabolic normalization

> **⚠ Critical caveat — intercept comparisons are NOT temperature-standardized:**
> Respiration rate is strongly temperature-dependent (Q₁₀ ≈ 1.5–3.0 for plant dark
> respiration; Atkin & Tjoelker 2003, *Plant Cell Environ* 26: 543–559).
> The Dryad CSV contains no measurement-temperature column, and measurements span
> Japanese forests and Pyrenean sites across seasons and three sub-studies.
> Family-level intercept differences below therefore **confound true metabolic
> normalization differences with unknown temperature offsets**. Do not interpret
> intercept rank-order as evidence of metabolic differences among families without
> temperature standardization.

```{r family-intercepts, fig.cap="⚠ NOT TEMPERATURE-STANDARDIZED. Family-level metabolic normalization (SMA intercept) — the expected log10 respiration rate at log10 fresh mass = 0 (i.e., 1 g plant). Intercept differences may partly reflect measurement temperature differences across sub-studies rather than true metabolic normalization differences."}
ggplot(fam_fits, aes(x = reorder(family, intercept), y = intercept,
                     fill = life_form)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = mean(fam_fits$intercept, na.rm = TRUE),
             linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  coord_flip() +
  labs(x = NULL, y = "SMA intercept  (log R at 1 g fresh mass)",
       fill = "Life form",
       title = "Per-family metabolic normalization (intercept)",
       subtitle = "Dashed = mean across eligible families") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## Within-family PGLS (families with n_species ≥ 8 in tree)

```{r family-pgls}
if (HAS_TREE && nrow(pgls_results) > 0) {
  # BUG 4 FIX: guard against all-NA family column; filter !is.na(family) before count().
  # Without this, count(family) on an all-NA column returns one NA row, causing
  # lapply(NA, ...) to return NULL results — no family-level PGLS runs.
  fam_pgls_eligible <- sp_means |>
    filter(!is.na(family), sp_tree %in% pgls_tree$tip.label) |>
    count(family) |>
    filter(n >= 8) |>
    pull(family)

  if (length(fam_pgls_eligible) > 0) {
    fam_pgls_list <- lapply(fam_pgls_eligible, function(fam) {
      # Aggregate across plant parts to one row per species (required by caper:
      # rownames must be unique tip labels; the same species may appear as both
      # Shoot and Root in sp_means).
      sp_sub <- sp_means |>
        filter(family == fam, sp_tree %in% pgls_tree$tip.label) |>
        group_by(sp_tree, family, life_form) |>
        dplyr::summarise(
          log_resp_mean = mean(log_resp_mean, na.rm = TRUE),
          log_mass_mean = mean(log_mass_mean, na.rm = TRUE),
          n_records     = sum(n_records),
          .groups = "drop"
        )
      if (nrow(sp_sub) < 6) return(NULL)
      tr <- ape::keep.tip(pgls_tree, sp_sub$sp_tree)
      tr$node.label <- NULL
      # Restrict to formula columns only (same Bug 1 fix as in fit_group_pgls)
      df <- as.data.frame(sp_sub)[, c("sp_tree", "log_resp_mean", "log_mass_mean")]
      rownames(df) <- df$sp_tree
      cd <- tryCatch(caper::comparative.data(tr, df, "sp_tree",
                                              vcv = TRUE, warn.dropped = FALSE),
                     error = function(e) NULL)
      if (is.null(cd)) return(NULL)
      pgls_f <- tryCatch(
        caper::pgls(log_resp_mean ~ log_mass_mean, data = cd, lambda = "ML",
                    bounds = list(lambda = c(1e-6, 1))),
        error = function(e) NULL
      )
      if (is.null(pgls_f)) return(NULL)
      ps <- summary(pgls_f)
      tibble(family = fam, n_species = nrow(sp_sub),
             slope  = coef(pgls_f)[2],
             slope_se = coef(ps)[2, 2],
             lambda = pgls_f$param["lambda"],
             r2     = ps$r.squared)
    })
    fam_pgls_tbl <- bind_rows(fam_pgls_list)
    if (nrow(fam_pgls_tbl) > 0) {
      kable(
        fam_pgls_tbl |> mutate(across(where(is.numeric), ~ round(.x, 3))) |>
          rename(Family = family, `N species` = n_species,
                 `PGLS slope` = slope, SE = slope_se,
                 `λ (ML)` = lambda, `R²` = r2),
        caption = "Within-family PGLS results (families with n_species ≥ 8 in tree)"
      )
    }
  } else {
    cat("No families with ≥8 species in tree — within-family PGLS not run.\n")
    cat("(This is expected with demo data; real Kurosawa/Mori data should yield",
        "Fagaceae and Pinaceae qualifying.)\n")
  }
}
```

---

# Scaling exponent forest plot

Three reference exponents are shown as vertical lines, each with a distinct
mechanistic interpretation:

| Reference | $\beta$ | Label | Biological justification |
|---|:---:|---|---|
| Rubner surface-area null | $2/3$ | Rubner/surface | Body surface scales as $M^{2/3}$ for isometric bodies; if respiration is limited by diffusive O$_2$/CO$_2$ exchange or heat dissipation through the body surface, then $R \propto M^{2/3}$. Empirically observed in some inter-specific compilations of plant dark respiration (Rubner 1883; Hemmingsen 1960). |
| WBE / MST prediction | $3/4$ | WBE/MST | Derived from fractal vascular network optimality (see Theory section): area-preserving branching + space-filling + size-invariant terminal units yield $B \propto M^{3/4}$ (West et al. 1997; Enquist et al. 1998; Mori et al. 2010). |
| Isometric | $1$ | Isometric | Constant mass-specific respiration, $r = R/M = \text{const}$; implies metabolic rate is simply proportional to cytoplasmic volume. Observed within some ontogenetic series and for unicellular organisms; also a common OLS artefact under restricted mass ranges. |

```{r forest-plot, fig.height=5, fig.cap="Forest plot of SMA and OLS slope estimates with 95% CI by group. Reference lines: dotted green = Rubner 2/3, dashed red = WBE 3/4, dotted blue = isometric 1."}
forest_df <- group_fits |>
  select(group, sma_slope, sma_ci_lo, sma_ci_hi, ols_slope, ols_ci_lo, ols_ci_hi, n) |>
  mutate(label = paste0(group, " (n=", n, ")")) |>
  pivot_longer(
    cols      = c(sma_slope, ols_slope),
    names_to  = "method",
    values_to = "slope"
  ) |>
  mutate(
    ci_lo  = if_else(method == "sma_slope", sma_ci_lo, ols_ci_lo),
    ci_hi  = if_else(method == "sma_slope", sma_ci_hi, ols_ci_hi),
    method = recode(method, sma_slope = "SMA", ols_slope = "OLS"),
    y_pos  = as.numeric(factor(label)) + if_else(method == "SMA", 0.15, -0.15)
  )

ggplot(forest_df, aes(y = y_pos, x = slope, xmin = ci_lo, xmax = ci_hi,
                      colour = method, shape = method)) +
  geom_pointrange(size = 0.7) +
  geom_vline(xintercept = 2/3,  linetype = "dotted", colour = "#4dac26",
             linewidth = 0.9) +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
             linewidth = 0.9) +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
             linewidth = 0.9) +
  scale_y_continuous(
    breaks = seq_len(length(unique(forest_df$label))),
    labels = unique(forest_df$label[order(as.numeric(factor(forest_df$label)))]),
    expand = expansion(add = 0.6)
  ) +
  scale_colour_manual(values = c(OLS = "steelblue", SMA = "darkorange")) +
  scale_shape_manual(values = c(OLS = 16, SMA = 17)) +
  labs(x = "Slope  (log R ~ log Fresh mass)",
       y = NULL,
       colour = "Estimator", shape = "Estimator",
       title = "Forest plot: respiration scaling exponents",
       subtitle = "OLS (circle, blue) and SMA (triangle, orange) with 95% CI. Reference lines: Rubner 2/3 (green), WBE 3/4 (red), isometric 1 (blue).") +
  theme_bw(base_size = 11) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = "bottom")
```

---

# Respiration residuals

```{r residuals, fig.cap="SMA residuals (log-scale) by life form. Positive = higher respiration than expected from pooled fit."}
dat_resid <- dat |>
  filter(life_form != "unknown") |>
  mutate(resid_pooled = log_resp - (cf_pooled[1] + cf_pooled[2] * log_mass))

ggplot(dat_resid, aes(x = life_form, y = resid_pooled, fill = life_form)) +
  geom_violin(alpha = 0.5, draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ plant_part) +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  labs(x = NULL, y = "SMA residual (log scale)",
       title = "Residuals from pooled scaling line") +
  theme(legend.position = "none")
```

---

# Kurosawa 2023: Fagus crenata ontogenetic series

Kurosawa et al. (2023; *Annals of Botany* 131: 313–322;
DOI: [10.1093/aob/mcac143](https://doi.org/10.1093/aob/mcac143)) is a narrower
ontogenetic study tracking respiration, fresh mass, and surface area of *Fagus
crenata* shoots and roots from seedling to large-tree stages. This subset is
included in the 2025 Dryad dataset as `dataset_source == "Kurosawa2023"`.

```{r fagus-ontogenetic, fig.cap="Fagus crenata ontogenetic scaling: respiration vs. fresh mass (shoot and root)."}
fagus_dat <- dat |>
  filter(grepl("Kurosawa2023|kurosawa2023|kurosawa_2023|Fagus|fagus",
               paste(dataset_source, species), ignore.case = TRUE))

if (nrow(fagus_dat) >= 10) {
  ggplot(fagus_dat, aes(x = log_mass, y = log_resp, colour = plant_part, shape = plant_part)) +
    geom_point(size = 2.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
    scale_colour_manual(values = c(Shoot = "#4dac26", Root = "#d01c8b",
                                   unknown = "grey50")) +
    labs(x = expression(log[10]~"Fresh mass (g)"),
         y = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
         colour = "Plant part", shape = "Plant part",
         title = expression(italic("Fagus crenata")~"ontogenetic scaling"),
         subtitle = "Kurosawa et al. 2023 subset") +
    theme(legend.position = "bottom")
} else {
  cat(sprintf(
    "Fewer than 10 Fagus crenata records isolated (n = %d). This may indicate\n",
    nrow(fagus_dat)),
    "the dataset_source or species column uses a different naming convention.\n",
    "Check column content with: dat |> count(dataset_source) and dat |> count(species)\n")
}
```

---

# Root:shoot respiration ratio

## Theoretical predictions

Under WBE, if shoot and root respiration both obey $R_i \propto \alpha M_i^{3/4}$
with the **same** allometric coefficient $\alpha$ (same metabolic intensity per
unit network volume), the respiration ratio is determined entirely by the
mass ratio:

$$\frac{R_{\rm shoot}}{R_{\rm root}} = \left(\frac{M_{\rm shoot}}{M_{\rm root}}\right)^{3/4}$$

If the root:shoot mass ratio $\rho = M_{\rm root}/M_{\rm shoot}$ is itself
size-invariant — as WBE implicitly assumes when it posits a single whole-plant
exponent — then $R_{\rm shoot}/R_{\rm root}$ is **constant across plant sizes**.

Optimal partitioning theory (Thornley 1972) predicts instead that $\rho$ is
dynamic: it increases under nutrient limitation (more root needed to acquire
soil resources) and decreases under shade (more shoot area needed to capture
light). Across a size gradient that is confounded with successional light
environment or soil fertility, $\rho$ and therefore $R_{\rm shoot}/R_{\rm root}$
will vary systematically with $M_{\rm total}$.

**Discriminating tests:**

1. **OLS of $\log(R_{\rm shoot}/R_{\rm root})$ vs $\log(M_{\rm plant})$:** WBE
   + size-invariant allocation predicts slope $= 0$; slope $\neq 0$ reveals
   differential organ scaling or size-dependent allocation shifts.
2. **SMA of $\log R_{\rm shoot}$ vs $\log R_{\rm root}$ (species matched):**
   WBE predicts slope $= 1$ and intercept $= 0$ if both organs have equal
   metabolic intensity; an intercept $\neq 0$ at slope $= 1$ indicates that
   shoots and roots differ in mass-specific metabolic rate (different $\alpha$)
   but scale in parallel; a slope $\neq 1$ indicates divergent exponents.

```{r root-shoot-ratio, fig.cap="Root vs. shoot respiration ratio. Left: ratio vs. species-level total log mass (species with both organs). Right: SMA of log R_shoot vs. log R_root matched by species. WBE predictions: slope = 0 (left), slope = 1 (right)."}
# Match species that have BOTH shoot and root records
rs_wide <- dat |>
  filter(plant_part %in% c("Shoot", "Root")) |>
  group_by(species, life_form, plant_part) |>
  summarise(
    log_resp_mean = mean(log_resp, na.rm = TRUE),
    log_mass_mean = mean(log_mass, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from  = plant_part,
    values_from = c(log_resp_mean, log_mass_mean)
  ) |>
  filter(!is.na(log_resp_mean_Shoot), !is.na(log_resp_mean_Root)) |>
  mutate(
    log_ratio      = log_resp_mean_Shoot - log_resp_mean_Root,
    log_total_mass = log10(10^log_mass_mean_Shoot + 10^log_mass_mean_Root)
  )

if (nrow(rs_wide) >= 5) {
  # OLS: log ratio ~ log total mass
  lm_ratio <- lm(log_ratio ~ log_total_mass, data = rs_wide)
  cat(sprintf("OLS log(R_shoot/R_root) ~ log(M_total): slope = %.3f, p = %.4f\n",
              coef(lm_ratio)[2], summary(lm_ratio)$coefficients[2, 4]))

  # SMA: log R_shoot ~ log R_root
  sma_rs <- sma(log_resp_mean_Shoot ~ log_resp_mean_Root, data = rs_wide,
                method = "SMA")
  cf_rs  <- coef(sma_rs)
  cat(sprintf("SMA log(R_shoot) ~ log(R_root): slope = %.3f (95%% CI: %.3f – %.3f)\n",
              cf_rs[2], sma_rs$coef[[1]][2, 2], sma_rs$coef[[1]][2, 3]))
  test_rs1 <- slope.test(rs_wide$log_resp_mean_Shoot, rs_wide$log_resp_mean_Root,
                         test.value = 1.0, method = "SMA")
  cat(sprintf("  H0: slope = 1.0 → p = %.4f\n", test_rs1$p))

  p1 <- ggplot(rs_wide, aes(x = log_total_mass, y = log_ratio,
                             colour = life_form)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, colour = "grey30",
                linetype = "dashed", linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "solid", colour = "black",
               linewidth = 0.6) +
    scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
    labs(x = expression(log[10]~"Total fresh mass (g)"),
         y = expression(log[10]~"(R"[shoot]~"/"~"R"[root]~")"),
         colour = "Life form",
         title = "Shoot:root respiration ratio vs. plant size",
         subtitle = sprintf("OLS slope = %.3f, p = %.3f  (WBE: slope = 0)",
                            coef(lm_ratio)[2],
                            summary(lm_ratio)$coefficients[2, 4])) +
    theme(legend.position = "bottom")

  p2 <- ggplot(rs_wide, aes(x = log_resp_mean_Root, y = log_resp_mean_Shoot,
                             colour = life_form)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_abline(intercept = cf_rs[1], slope = cf_rs[2],
                colour = "black", linewidth = 1.1) +
    geom_abline(intercept = mean(rs_wide$log_resp_mean_Shoot) -
                              mean(rs_wide$log_resp_mean_Root),
                slope = 1, linetype = "dashed", colour = "#d6604d",
                linewidth = 0.9) +
    scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
    labs(x = expression(log[10]~"Root respiration"),
         y = expression(log[10]~"Shoot respiration"),
         colour = "Life form",
         title = "SMA: shoot vs. root respiration (species means)",
         subtitle = sprintf("SMA slope = %.3f  [WBE isometry: dashed, slope = 1]",
                            cf_rs[2])) +
    theme(legend.position = "bottom")

  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else {
  cat(sprintf("Only %d species have matched shoot AND root records — root:shoot\n",
              nrow(rs_wide)),
      "ratio analysis requires paired records. Check plant_part coding in data.\n")
}
```

---

# Summary table

```{r final-table}
kable(
  group_fits |>
    ungroup() |>
    select(group, n, ols_slope, ols_ci_lo, ols_ci_hi, ols_r2, ols_p_075,
           sma_slope, sma_ci_lo, sma_ci_hi, sma_r2, sma_p_075) |>
    rename(Group = group, N = n,
           `OLS slope` = ols_slope, `OLS CI lo` = ols_ci_lo, `OLS CI hi` = ols_ci_hi,
           `OLS R2` = ols_r2, `OLS p(b=0.75)` = ols_p_075,
           `SMA slope` = sma_slope, `SMA CI lo` = sma_ci_lo, `SMA CI hi` = sma_ci_hi,
           `SMA R2` = sma_r2, `SMA p(b=0.75)` = sma_p_075) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "Summary: OLS and SMA scaling exponents for log respiration ~ log fresh mass by life form and plant part"
)
```

---

# Caveats and data status

```{r caveats}
if (!USE_REAL_DATA) {
  cat("⚠ DATA STATUS: This analysis used RECONSTRUCTED DEMONSTRATION DATA.\n",
      "  Slopes and ranges are calibrated to approximate published values\n",
      "  from Kurosawa et al. 2025 and Mori et al. 2010 but are NOT the\n",
      "  original measurements.\n\n",
      "  To run with real data:\n",
      "  1. Download data.respiration.csv from https://doi.org/10.5061/dryad.sxksn03cj\n",
      "  2. Save to:  plant_scaling_data/data/raw/kurosawa_respiration/data.respiration.csv\n",
      "  3. Re-render this Rmd.\n\n")
}
cat("General caveats:\n",
    "• Respiration units (Fix #3): Kurosawa et al. (2025) report rates in\n",
    "  nmol CO2 g-1 s-1 (per unit fresh mass). Axis labels in this Rmd assume\n",
    "  those units. When loading real Dryad data, verify units from the README at\n",
    "  DOI 10.5061/dryad.sxksn03cj before interpreting intercepts.\n",
    "• Temperature standardisation (Fix #4): measurement temperature varies across\n",
    "  experiments. Raw rates at different temperatures are not directly comparable.\n",
    "  For cross-study slope estimation, standardise to a common reference temperature\n",
    "  (e.g. 15 degrees C as in Mori et al. 2010, Global Ecol. Biogeogr. 19:127-139)\n",
    "  using a Q10 correction: R_ref = R_meas / Q10^((T_meas - T_ref)/10) [Q10 ~ 2].\n",
    "  The demo data used here has no temperature metadata; intercept estimates from\n",
    "  demo mode should not be compared across groups or to published values.\n",
    "• SMA slopes are sensitive to the range of body sizes included; a very broad\n",
    "  mass range will tend to push slopes toward 1 if allometry is non-linear.\n",
    "• The herbaceous and woody plants span very different ecological contexts;\n",
    "  the pooled slope should be interpreted cautiously.\n")
```

---

# References

* Kurosawa Y, Mori S, Ferrio JP, et al. (2025). Scaling of shoot and root
  respiration of woody and herbaceous plants. *Proc R Soc B* **292**: 20241910.
  DOI: 10.1098/rspb.2024.1910

* Kurosawa Y, et al. (2024). Data from: Scaling of shoot and root respiration
  of woody and herbaceous plants. *Dryad*. DOI: 10.5061/dryad.sxksn03cj

* Kurosawa Y, et al. (2023). Ontogenetic changes in root and shoot respiration,
  fresh mass and surface area of *Fagus crenata*. *Ann Bot* **131**: 313–322.
  DOI: 10.1093/aob/mcac143

* Mori S, et al. (2010). Mixed-power scaling of whole-plant respiration from
  seedlings to giant trees. *Proc R Soc B* **277**: 1447–1454.
  DOI: 10.1098/rspb.2009.1811

* West GB, Brown JH, Enquist BJ (1997). A general model for the origin of
  allometric scaling laws in biology. *Science* **276**: 122–126.

* Glazier DS (2005). Beyond the '3/4-power law': variation in the intra-
  and interspecific scaling of metabolic rate in animals. *Biol Rev* **80**: 611–662.
  DOI: 10.1017/S1464793105006834

* Enquist BJ, Brown JH, West GB (1998). Allometric scaling of plant energetics
  and population density. *Nature* **395**: 163–165.
  DOI: 10.1038/26066

* West GB, Brown JH, Enquist BJ (1999). A general model for the structure and
  allometry of plant vascular systems. *Nature* **400**: 664–667.
  DOI: 10.1038/23251

* Rubner M (1883). Über den Einfluss der Körpergrösse auf Stoff- und Kraftwechsel.
  *Z Biol* **19**: 535–562.

* Hemmingsen AM (1960). Energy metabolism as related to body size and respiratory
  surfaces, and its evolution. *Rep Steno Mem Hosp Nord Insulin Lab* **9**: 1–110.

* Thornley JHM (1972). A balanced quantitative model for root:shoot ratios in
  vegetative plants. *Ann Bot* **36**: 431–441.
  DOI: 10.1093/oxfordjournals.aob.a084561

* Pagel M (1999). Inferring the historical patterns of biological evolution.
  *Nature* **401**: 877–884.
  DOI: 10.1038/44766

* Atkin OK & Tjoelker MG (2003). Thermal acclimation and the dynamic response
  of plant respiration to temperature. *Trends Plant Sci* **8**: 343–351.
  DOI: 10.1016/S1360-1385(03)00136-5

* Warton DI, Wright IJ, Falster DS, Westoby M (2006). Bivariate line-fitting
  methods for allometry. *Biol Rev* **81**: 259–291.
  DOI: 10.1017/S1464793106007007

* Orme D, Freckleton R, Thomas G, et al. (2013). caper: Comparative Analyses
  of Phylogenetics and Evolution in R. R package version 1.0.1.

* Smith SA & Brown JW (2018). Constructing a broadly inclusive seed plant
  phylogeny. *Am J Bot* **105**: 302–314.
  DOI: 10.1002/ajb2.1019
  *(GBOTB megaphylogeny backbone used for PGLS — verify DOI against Crossref before publication)*

* Jin Y & Qian H (2022). V.PhyloMaker2: An updated and enlarged R package that
  can generate very large phylogenies for vascular plants. *Plant Diversity*
  **44**: 335–339. DOI: 10.1016/j.pld.2022.05.002

* Blomberg SP, Garland T Jr, Ives AR (2003). Testing for phylogenetic signal
  in comparative data: behavioral traits are more labile. *Evolution*
  **57**: 717–745. DOI: 10.1111/j.0014-3820.2003.tb00285.x

> **⚠ Citation verification:** All DOIs, volume numbers, page ranges, and journal
> names in this document should be independently verified against Crossref or
> publisher websites before use in any manuscript or public report.

---

*This analysis was conducted using `smatr` (Warton et al. 2012) in R `r R.version$major`.`r R.version$minor`.*
*Data source: Kurosawa et al. 2024, Dryad DOI: 10.5061/dryad.sxksn03cj.*
