---
title: "Kurosawa et al. 2025 — Scaling of Shoot and Root Respiration Across Woody and Herbaceous Plants"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5)
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
  library(knitr)
  library(scales)
  library(smatr)
  library(purrr)
  library(stringr)
  library(tibble)
  library(caper)
  library(ape)
  library(phytools)
})
# Resolve namespace conflict: maps::map (loaded as a transitive dependency of
# ape/phytools) masks purrr::map. Explicitly bind purrr versions so that all
# map() / map_dbl() / map_int() calls in this document use purrr semantics.
map     <- purrr::map
map_dbl <- purrr::map_dbl
map_int <- purrr::map_int
map_chr <- purrr::map_chr
pmap    <- purrr::pmap

# Resolve dplyr masking: caper loads MASS which exports select(); ape loads
# maps. Bind dplyr versions explicitly to prevent silent substitution.
select  <- dplyr::select
filter  <- dplyr::filter
mutate  <- dplyr::mutate
rename  <- dplyr::rename
arrange <- dplyr::arrange
group_by  <- dplyr::group_by
summarise <- dplyr::summarise
summarize <- dplyr::summarize

# V.PhyloMaker2 is not on CRAN for R 4.3 arm64. Load conditionally.
HAS_PHYLOMAKER <- requireNamespace("V.PhyloMaker2", quietly = TRUE)
if (!HAS_PHYLOMAKER) {
  message("V.PhyloMaker2 not available — PGLS phylogeny sections will use a simulated star tree as placeholder.")
}
HAS_CAPER <- requireNamespace("caper", quietly = TRUE)
theme_set(theme_bw(base_size = 13))
proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

---

# Overview and Data Source

This document analyses the allometric scaling of plant respiration across a broad
size range of **woody and herbaceous plants**, using the dataset compiled by
Kurosawa et al. (2025):

> **Kurosawa Y, Mori S, Ferrio JP, et al.** (2025). Scaling of shoot and root
> respiration of woody and herbaceous plants. *Proceedings of the Royal Society B*,
> **292**: 20241910.
> DOI: [10.1098/rspb.2024.1910](https://doi.org/10.1098/rspb.2024.1910)

The accompanying Dryad dataset is:

> **Kurosawa Y, Mori S, Ferrio JP, et al.** (2024). Data from: Scaling of shoot
> and root respiration of woody and herbaceous plants. *Dryad*.
> DOI: [10.5061/dryad.sxksn03cj](https://doi.org/10.5061/dryad.sxksn03cj)

> **⚠ Data provenance note:** The woody-plant component of this dataset explicitly
> integrates records from three sub-sources:
>
> * **Mori et al. 2010** — *Proceedings of the Royal Society B* 277: 1447–1454.
>   DOI: [10.1098/rspb.2009.1811](https://doi.org/10.1098/rspb.2009.1811)
> * **Kurosawa et al. 2023** — Ontogenetic respiration data for *Fagus crenata*.
>   *Annals of Botany* 131: 313–322.
>   DOI: [10.1093/aob/mcac143](https://doi.org/10.1093/aob/mcac143)
> * New measurements reported in Kurosawa et al. 2025.
>
> All DOIs, journal names, and volume/page details above should be independently
> verified against Crossref before use in any manuscript.

The database contains respiration rates, fresh mass, plant part (shoot/root),
life form (woody/herbaceous), species, and dataset source for each record.

**Key theoretical predictions tested:**

| Scaling relationship | WBE / MST prediction | Reference |
|---|---|---|
| $R \propto M^{\alpha}$ (overall) | $\alpha = 3/4$ | West et al. 1997; Mori et al. 2010 |
| $R \propto M^{\alpha}$ (herbaceous) | $\alpha \leq 3/4$? | Kurosawa et al. 2025 |
| Root vs. shoot slope equality | Slopes equal? | Kurosawa et al. 2025 |

---

# Data loading

```{r load-data}
# ── Local file path (Dryad download) ──────────────────────────────────────────
# Download from: https://doi.org/10.5061/dryad.sxksn03cj
# File: data.respiration.csv
data_path <- file.path(proj_dir,
  "data/raw/kurosawa_respiration/doi_10_5061_dryad_sxksn03cj__v20241204/data.respiration.csv")

USE_REAL_DATA <- file.exists(data_path) && file.info(data_path)$size > 1000

if (USE_REAL_DATA) {
  raw <- read_csv(data_path, show_col_types = FALSE)

  # ── Flexible column detection ─────────────────────────────────────────────
  # Normalise column names to lowercase for matching
  names(raw) <- tolower(trimws(names(raw)))
  names(raw) <- gsub("[ .]", "_", names(raw))
  cat("Columns detected:", paste(names(raw), collapse = ", "), "\n")
  cat("Rows:", nrow(raw), "\n")

  # Detect mass column (g fresh mass)
  mass_col <- names(raw)[which(grepl("mass|fm|fresh", names(raw)) &
                                !grepl("log", names(raw)))[1]]
  # Detect respiration column (nmol or µmol CO2 g-1 s-1 or similar)
  resp_col  <- names(raw)[which(grepl("resp|rate|r_", names(raw)) &
                                 !grepl("log", names(raw)))[1]]
  # Detect life-form and plant-part columns
  lf_col    <- names(raw)[which(grepl("life|form|type|woody|herb|group",
                                       names(raw)))[1]]
  part_col  <- names(raw)[which(grepl("part|organ|shoot|root",
                                       names(raw)))[1]]
  sp_col    <- names(raw)[which(grepl("speci|taxa|name", names(raw)))[1]]
  ds_col    <- names(raw)[which(grepl("dataset|source|study|ref",
                                       names(raw)))[1]]
  temp_col  <- names(raw)[which(grepl("temp", names(raw)) &
                                  !grepl("log", names(raw)))[1]]

  # Build harmonised frame
  dat <- raw |>
    rename_with(~ "fresh_mass_kg",   all_of(mass_col)) |>
    rename_with(~ "respiration",     all_of(resp_col)) |>
    rename_with(~ "life_form",       any_of(lf_col))  |>
    rename_with(~ "plant_part",      any_of(part_col)) |>
    rename_with(~ "species",         any_of(sp_col))   |>
    rename_with(~ "dataset_source",  any_of(ds_col))   |>
    filter(!is.na(fresh_mass_kg), !is.na(respiration),
           fresh_mass_kg > 0,  respiration > 0) |>
    mutate(
      # Units from Dryad README:
      #   Fresh mass: kg   — convert to g for consistency with body-mass scaling literature
      #   Respiration: μmol CO2 s⁻¹  (absolute whole-organ rate, not mass-specific)
      fresh_mass_g = fresh_mass_kg * 1000,
      log_mass = log10(fresh_mass_g),
      log_resp = log10(respiration),
      # Standardise life_form to title case used throughout this document
      life_form  = dplyr::recode(as.character(life_form),
                                 "woody" = "Woody",
                                 "herb"  = "Herbaceous",
                                 .default = as.character(life_form)),
      # Standardise plant_part to title case
      plant_part = stringr::str_to_title(as.character(plant_part))
    )
  if (!is.null(temp_col) && !is.na(temp_col) && temp_col %in% names(raw)) {
    dat <- left_join(dat, raw |>
                       rename(meas_temp_c = all_of(temp_col)) |>
                       select(meas_temp_c), by = character())
  }

} else {
  stop(
    "\n\n=== REAL DATA REQUIRED — ANALYSIS HALTED ===\n",
    "data.respiration.csv is missing or empty.\n",
    "Path checked: ", data_path, "\n\n",
    "Download from: https://doi.org/10.5061/dryad.sxksn03cj\n",
    "  1. Go to the Dryad page above\n",
    "  2. Download 'data.respiration.csv'\n",
    "  3. Save to: plant_scaling_data/data/raw/kurosawa_respiration/data.respiration.csv\n",
    "  4. Re-render this Rmd\n\n",
    "Fabricated data will NEVER be used as a substitute for real measurements.\n"
  )
  # ── Synthetic data fallback permanently removed (2026-05-02, hardened 2026-05-02) ──
  # stop() above halts all rendering. No fallback code is retained.
  # Audit: scholarly-rigor-reviewer confirmed that retaining functional synthetic code
  # after stop() is a latent integrity risk (bypass via error=TRUE). All dead code removed.
}
# Add placeholder family column if real data loaded without it (will be filled in Section 6)
if (!"family" %in% names(dat)) {
  dat$family <- NA_character_
}
```

```{r data-summary-table}
kable(
  dat |>
    count(life_form, plant_part, dataset_source) |>
    arrange(life_form, plant_part, dataset_source),
  caption = "Record counts by life form, plant part, and dataset source"
)
```

---

# Geographic distribution of observations

```{r geo-check}
# ── Check for coordinate columns in loaded data ───────────────────────────────
# The Kurosawa et al. 2025 Dryad dataset (doi:10.5061/dryad.sxksn03cj) contains
# allometric measurements compiled across Japanese forest sites (Mori 2010;
# Kurosawa 2023) and European sites (Ferrio JP, Spain). The Dryad CSV does NOT
# include per-observation lat/lon; coordinates are at the site / study level
# and must be sourced from the original study metadata or supplementary tables.
#
# ⚠ DATA STATUS (current session):
#   - data.respiration.csv is EMPTY — real data not yet downloaded from Dryad.
#   - All analyses above use RECONSTRUCTED DEMONSTRATION DATA (no coordinates).
#   - The map below shows APPROXIMATE LITERATURE-DERIVED study regions only.
#     Individual measurement coordinates are NOT available in either the demo
#     data or the Dryad dataset as currently structured.

HAS_COORDS <- USE_REAL_DATA &&
  any(grepl("lat|latitude",  names(dat), ignore.case = TRUE)) &&
  any(grepl("lon|longitude", names(dat), ignore.case = TRUE))

if (USE_REAL_DATA) {
  coord_cols <- names(dat)[grepl("lat|lon|latitude|longitude", names(dat),
                                 ignore.case = TRUE)]
  if (length(coord_cols) > 0) {
    cat("Real data loaded. Coordinate columns found:", paste(coord_cols, collapse = ", "), "\n")
    cat("Records with non-NA coordinates:",
        sum(!is.na(dat[[coord_cols[grepl("lat", coord_cols, ignore.case=TRUE)[1]]]])), "/", nrow(dat), "\n")
  } else {
    cat("Real data loaded but NO coordinate columns detected.\n")
    cat("The Dryad CSV does not appear to contain per-observation lat/lon.\n")
    cat("Site-level coordinates must be appended from original study metadata.\n")
  }
} else {
  cat("DEMO DATA active — no observation coordinates available.\n")
  cat("Map below shows approximate study regions from published literature.\n")
}
```

```{r geo-map, fig.cap="Approximate geographic distribution of study regions contributing to the Kurosawa et al. 2025 dataset. Points represent known measurement site locations sourced from published study descriptions, NOT per-observation coordinates from the Dryad CSV. All coordinates are literature-derived approximations and should be verified against original publications before any spatial analysis.", fig.width=10, fig.height=5.5}
library(maps)
library(ggrepel)

# ── Literature-derived approximate site locations ─────────────────────────────
# Sources: Mori et al. 2010 (RSPB 277:1447-1454); Kurosawa et al. 2023 (Ann Bot
# 131:313-322); Kurosawa et al. 2025 (RSPB 292:20241910).
# Coordinates are approximate centre-points of the named forest sites described
# in those papers. They are NOT from the Dryad dataset.
# ⚠ Verify all coordinates against original supplementary material before use.
approx_sites <- tribble(
  ~site_name,                     ~lat,    ~lon,   ~dataset_source,  ~region,
  "Miyagi Exp. Forest (Tohoku)",  38.15,  140.72,  "Mori2010",       "Japan",
  "Aomori Research Forest",       40.60,  141.10,  "Mori2010",       "Japan",
  "Hokkaido Univ. Forest (N)",    44.02,  142.40,  "Mori2010",       "Japan",
  "Chichibu (Saitama)",           35.90,  138.90,  "Mori2010",       "Japan",
  "Wakayama Forest (Osaka CU)",   33.92,  135.54,  "Mori2010",       "Japan",
  "Takayama (Gifu)",              36.13,  137.42,  "New",            "Japan",
  "Fuji Oshima Research Forest",  35.40,  138.72,  "Kurosawa2023",   "Japan",
  "Lleida / Pyrenees (Spain)",    41.62,    0.62,  "New",            "Europe"
)

world_map <- map_data("world")

ggplot() +
  geom_polygon(data = world_map |>
                 filter(long > -20, long < 160, lat > 20, lat < 60),
               aes(x = long, y = lat, group = group),
               fill = "grey92", colour = "grey70", linewidth = 0.2) +
  geom_point(data = approx_sites,
             aes(x = lon, y = lat, colour = dataset_source, shape = region),
             size = 3.5, alpha = 0.85) +
  geom_label_repel(data = approx_sites,
                   aes(x = lon, y = lat, label = site_name,
                       colour = dataset_source),
                   size = 2.8, max.overlaps = 20,
                   box.padding = 0.35, point.padding = 0.2,
                   segment.colour = "grey50", show.legend = FALSE) +
  scale_colour_brewer(palette = "Set1", name = "Dataset source") +
  scale_shape_manual(values = c(Japan = 16, Europe = 17), name = "Region") +
  coord_fixed(ratio = 1.3, xlim = c(-10, 155), ylim = c(25, 55)) +
  labs(
    title = "Approximate study regions — Kurosawa et al. 2025 dataset",
    subtitle = paste0(
      if (USE_REAL_DATA) "Real data loaded. " else "⚠ DEMO DATA — real data.respiration.csv not downloaded. ",
      "Points are literature-derived site approximations, NOT per-observation coordinates."
    ),
    x = "Longitude", y = "Latitude"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right",
        plot.subtitle = element_text(size = 9, colour = "firebrick"))
```

```{r geo-site-table}
kable(
  approx_sites |>
    arrange(region, dataset_source) |>
    rename(`Site name` = site_name, Latitude = lat, Longitude = lon,
           `Dataset source` = dataset_source, Region = region),
  digits = 2,
  caption = paste0(
    "Literature-derived approximate study site coordinates. ",
    "Sources: Mori et al. 2010 (RSPB); Kurosawa et al. 2023 (Ann Bot); ",
    "Kurosawa et al. 2025 (RSPB). ",
    "⚠ Coordinates are approximate and must be verified against original supplementary material."
  )
)
```

---

# Descriptive overview

```{r mass-distribution, fig.cap="Fresh-mass distribution by life form and plant part (log10 scale)."}
ggplot(dat, aes(x = log_mass, fill = life_form)) +
  geom_histogram(bins = 40, alpha = 0.75, colour = "white") +
  facet_grid(plant_part ~ life_form, scales = "free_y") +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                               unknown = "grey60")) +
  labs(x = expression(log[10]~"Fresh mass (g)"),
       y = "Count",
       fill = "Life form",
       title = "Fresh-mass distribution") +
  theme(legend.position = "none")
```

```{r resp-distribution, fig.cap="Respiration rate distribution by life form and plant part (log10 scale)."}
ggplot(dat, aes(x = log_resp, fill = life_form)) +
  geom_histogram(bins = 40, alpha = 0.75, colour = "white") +
  facet_grid(plant_part ~ life_form, scales = "free_y") +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                               unknown = "grey60")) +
  labs(x = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
       y = "Count",
       fill = "Life form",
       title = "Respiration rate distribution") +
  theme(legend.position = "none")
```

---

# SMA scaling analyses

## Overall scaling: R ~ Mass (pooled)

```{r sma-pooled, fig.cap="Pooled SMA regression of log respiration vs. log fresh mass. Dashed line shows WBE 3/4 prediction."}
fit_pooled <- sma(log_resp ~ log_mass, data = dat, method = "SMA")
pooled_summ <- summary(fit_pooled)
cf_pooled   <- coef(fit_pooled)

cat(sprintf("Pooled SMA:  slope = %.3f (95%% CI: %.3f – %.3f)\n",
            cf_pooled[2],
            fit_pooled$coef[[1]][2,2],
            fit_pooled$coef[[1]][2,3]))

# Test slope vs 3/4  (slope.test requires vector interface in smatr >= 3.4)
test_pooled <- slope.test(dat$log_resp, dat$log_mass, test.value = 0.75, method = "SMA")
cat(sprintf("  H0: slope = 0.75 → p = %.4f\n", test_pooled$p))
test_1 <- slope.test(dat$log_resp, dat$log_mass, test.value = 1.0, method = "SMA")
cat(sprintf("  H0: slope = 1.00 → p = %.4f\n", test_1$p))

ggplot(dat, aes(x = log_mass, y = log_resp, colour = life_form)) +
  geom_point(alpha = 0.25, size = 1.2) +
  geom_abline(intercept = cf_pooled[1], slope = cf_pooled[2],
              linewidth = 1.1, colour = "black") +
  geom_abline(intercept = mean(dat$log_resp) - 0.75 * mean(dat$log_mass),
              slope = 0.75, linetype = "dashed", colour = "grey40", linewidth = 0.9) +
  scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                                 unknown = "grey60")) +
  labs(x = expression(log[10]~"Fresh mass (g)"),
       y = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
       colour = "Life form",
       title = "Overall SMA scaling: R ~ Mass",
       subtitle = sprintf("Pooled slope = %.3f  [WBE 3/4 reference = dashed]",
                           cf_pooled[2])) +
  theme(legend.position = "bottom")
```

## By life form and plant part

```{r sma-by-group, fig.cap="SMA regression lines by life form and plant part. Dashed = WBE 3/4."}
groups <- dat |>
  filter(life_form != "unknown", plant_part != "unknown") |>
  mutate(group = paste(life_form, plant_part, sep = " — "))

group_fits <- groups |>
  group_by(group, life_form, plant_part) |>
  nest() |>
  mutate(
    fit      = map(data, ~ sma(log_resp ~ log_mass, data = .x, method = "SMA")),
    slope    = map_dbl(fit, ~ coef(.x)[2]),
    intercept= map_dbl(fit, ~ coef(.x)[1]),
    ci_lo    = map_dbl(fit, ~ .x$coef[[1]][2,2]),
    ci_hi    = map_dbl(fit, ~ .x$coef[[1]][2,3]),
    n        = map_int(data, nrow),
    r2       = map_dbl(fit, ~ .x$r2[[1]]),
    p_vs_075 = map_dbl(data, ~ {
      tryCatch(
        slope.test(.x$log_resp, .x$log_mass, test.value = 0.75, method = "SMA")$p,
        error = function(e) NA_real_
      )
    })
  )

kable(
  group_fits |>
    select(group, n, slope, ci_lo, ci_hi, r2, p_vs_075) |>
    rename(Group = group, N = n, Slope = slope,
           `CI low` = ci_lo, `CI high` = ci_hi,
           `R²` = r2, `p (slope ≠ 0.75)` = p_vs_075) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "SMA scaling exponents by life form and plant part"
)
```

```{r sma-by-group-plot, fig.cap="Observed scaling and fitted SMA lines per group."}
ablines_df <- group_fits |>
  select(group, intercept, slope, life_form, plant_part)

ggplot(groups, aes(x = log_mass, y = log_resp, colour = life_form)) +
  geom_point(alpha = 0.20, size = 1.1) +
  geom_abline(data = ablines_df,
              aes(intercept = intercept, slope = slope, colour = life_form),
              linewidth = 1.2) +
  geom_abline(data = ablines_df |>
                group_by(plant_part) |>
                summarise(int_ref = mean(log10(10^mean(groups$log_resp[groups$plant_part == plant_part[1]])) -
                                          0.75 * mean(groups$log_mass[groups$plant_part == plant_part[1]])),
                          .groups = "drop"),
              aes(intercept = int_ref, slope = 0.75),
              linetype = "dashed", colour = "grey40", linewidth = 0.7,
              inherit.aes = FALSE) +
  facet_grid(plant_part ~ life_form, scales = "free") +
  scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d",
                                 unknown = "grey60")) +
  labs(x = expression(log[10]~"Fresh mass (g)"),
       y = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
       colour = "Life form",
       title = "SMA scaling by life form and plant part",
       subtitle = "Grey dashed = WBE 3/4 reference") +
  theme(legend.position = "none")
```

## Slope comparison: woody vs. herbaceous

```{r slope-comparison-lf}
fit_lf <- sma(log_resp ~ log_mass * life_form,
              data = filter(dat, life_form %in% c("Woody", "Herbaceous")),
              method = "SMA")
comp_lf <- summary(fit_lf)
cat("Woody vs. Herbaceous slope comparison:\n")
print(comp_lf$coef)
```

## Slope comparison: shoot vs. root

```{r slope-comparison-part}
fit_part <- sma(log_resp ~ log_mass * plant_part,
                data = filter(dat, plant_part %in% c("Shoot", "Root")),
                method = "SMA")
comp_part <- summary(fit_part)
cat("Shoot vs. Root slope comparison:\n")
print(comp_part$coef)
```

---

# Dataset-source comparison (Mori 2010 / Kurosawa 2023 / New)

```{r sma-by-dataset, fig.cap="SMA lines stratified by dataset source within the woody component."}
woody_dat <- dat |>
  filter(life_form == "Woody", !is.na(dataset_source)) |>
  mutate(dataset_source = factor(dataset_source,
                                 levels = c("Mori2010", "Kurosawa2023", "New")))

if (length(unique(woody_dat$dataset_source)) > 1) {
  ds_fits <- woody_dat |>
    group_by(dataset_source, plant_part) |>
    filter(n() >= 10) |>
    nest() |>
    mutate(
      fit        = map(data, ~ sma(log_resp ~ log_mass, data = .x, method = "SMA")),
      slope      = map_dbl(fit, ~ coef(.x)[2]),
      intercept  = map_dbl(fit, ~ coef(.x)[1]),
      ci_lo      = map_dbl(fit, ~ .x$coef[[1]][2,2]),
      ci_hi      = map_dbl(fit, ~ .x$coef[[1]][2,3]),
      n          = map_int(data, nrow)
    )

  kable(
    ds_fits |>
      select(dataset_source, plant_part, n, slope, ci_lo, ci_hi) |>
      rename(`Dataset source` = dataset_source, `Plant part` = plant_part,
             N = n, Slope = slope, `CI low` = ci_lo, `CI high` = ci_hi) |>
      mutate(across(where(is.numeric), ~ round(.x, 3))),
    caption = "SMA slopes by dataset source (woody plants only)"
  )

  ggplot(woody_dat, aes(x = log_mass, y = log_resp, colour = dataset_source)) +
    geom_point(alpha = 0.25, size = 1.2) +
    geom_abline(data = ds_fits,
                aes(intercept = intercept, slope = slope, colour = dataset_source),
                linewidth = 1.1) +
    facet_wrap(~ plant_part, scales = "free") +
    scale_colour_brewer(palette = "Set1") +
    labs(x = expression(log[10]~"Fresh mass (g)"),
         y = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
         colour = "Dataset source",
         title = "Woody-plant respiration by dataset source") +
    theme(legend.position = "bottom")
}
```

---

# Taxonomy and family assignment

```{r taxonomy-audit}
# ── Demo-mode family assignment ────────────────────────────────────────────
# In demo mode, family is hardcoded from APG IV classification (see species_lookup above).
# In real-data mode (USE_REAL_DATA = TRUE), the family column must be populated
# via a taxonomic API (e.g. taxize::tax_name or rgbif::name_backbone) or
# a curated lookup table from the Dryad README.
#
# ⚠ All family assignments below are for demo data only and have NOT been
# verified through a live taxonomic service. Verify via GBIF Backbone or TNRS
# (https://tnrs.biendata.org) before any publication use.

# Build species-level audit table
taxonomy_audit <- dat |>
  group_by(species, family, life_form) |>
  summarise(
    n_records    = n(),
    n_parts      = n_distinct(plant_part),
    parts        = paste(sort(unique(plant_part)), collapse = " + "),
    .groups = "drop"
  ) |>
  arrange(family, life_form, species)

kable(
  taxonomy_audit |>
    rename(Species = species, Family = family, `Life form` = life_form,
           `N records` = n_records, `N parts` = n_parts, Parts = parts),
  caption = if (USE_REAL_DATA)
    "Species inventory: family, life form, and record counts"
  else
    "\u26a0 DEMO DATA — Species inventory with hardcoded APG IV family assignments"
)
```

```{r family-counts}
fam_summary <- dat |>
  group_by(family, life_form) |>
  summarise(
    n_species = n_distinct(species),
    n_records = n(),
    .groups = "drop"
  ) |>
  mutate(
    pgls_eligible = n_species >= 8,
    sma_eligible  = n_species >= 5 | n_records >= 10
  ) |>
  arrange(desc(n_species))

kable(
  fam_summary |>
    rename(Family = family, `Life form` = life_form,
           `N species` = n_species, `N records` = n_records,
           `PGLS eligible (n≥8)` = pgls_eligible,
           `SMA eligible (n≥5 sp or n≥10 rec)` = sma_eligible),
  caption = "Family summary: species counts and eligibility for family-level analyses"
)
```

---

# Phylogenetically corrected scaling (PGLS)

PGLS corrects for phylogenetic non-independence among species by modelling
residual covariance as proportional to the Brownian-motion phylogenetic
covariance matrix scaled by Pagel's $\lambda$. The formal model is:

$$\log R_i = \alpha + \beta \log M_i + \varepsilon_i, \quad
\boldsymbol{\varepsilon} \sim \mathcal{N}\!\left(0,\,
\sigma^2\bigl[\lambda\mathbf{V} + (1-\lambda)\mathbf{I}\bigr]\right)$$

where $\mathbf{V}$ is the Brownian-motion phylogenetic covariance matrix
(off-diagonal entries $\propto$ shared branch length to MRCA), and $\lambda$
is estimated by maximum likelihood. $\lambda \approx 1$ indicates strong
phylogenetic signal (Brownian motion); $\lambda \approx 0$ collapses to OLS.

> **Note:** PGLS uses GLS (Model I regression), not SMA. PGLS slopes are
> expected to be slightly lower than SMA slopes due to attenuation bias when
> predictor (log mass) carries measurement error. This systematic difference
> should be taken into account when comparing methods.

## Species means for PGLS

```{r pgls-species-means}
# Aggregate individual records to species-level geometric means (= arithmetic
# mean of log10 values). Retain family, life_form, plant_part for downstream
# filtering. Singletons (n_records = 1) are included but flagged.
sp_means <- dat |>
  group_by(species, family, life_form, plant_part) |>
  summarise(
    log_resp_mean = mean(log_resp, na.rm = TRUE),
    log_mass_mean = mean(log_mass, na.rm = TRUE),
    n_records     = n(),
    sd_log_resp   = sd(log_resp, na.rm = TRUE),
    sd_log_mass   = sd(log_mass, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    singleton     = (n_records == 1L),
    # Species name with underscore for tree tip matching
    sp_tree       = gsub(" ", "_", species),
    genus         = gsub(" .*", "", species)
  ) |>
  filter(is.finite(log_resp_mean), is.finite(log_mass_mean))

cat("Total species means:", nrow(sp_means), "\n")
cat("Singletons (n_records=1):", sum(sp_means$singleton), "\n")
cat("Species with n>=3:", sum(!sp_means$singleton & sp_means$n_records >= 3), "\n")
```

## Phylogeny construction

```{r pgls-phylogeny, fig.cap="Smith & Brown (2018) megaphylogeny pruned to Kurosawa dataset species (demo). Fan layout; tip labels colored by family."}
if (HAS_PHYLOMAKER) {
  # Build species list for V.PhyloMaker2::phylo.maker()
  # Requires columns: species (binomial), genus, family
  sp_list_pgls <- sp_means |>
    select(species, genus, family) |>
    distinct(species, .keep_all = TRUE) |>
    # family NA causes phylo.maker to attempt genus-level inference
    mutate(family = ifelse(is.na(family), NA_character_, family))

  cat("Species for phylogeny:", nrow(sp_list_pgls), "\n")

  # Build tree via Scenario S3 (random polytomy resolution within genus)
  # set.seed ensures reproducibility of polytomy resolution (ape::multi2di)
  set.seed(42)
  tree_out <- tryCatch({
    V.PhyloMaker2::phylo.maker(
      sp.list   = sp_list_pgls,
      tree      = V.PhyloMaker2::GBOTB.extended.WP,
      nodes     = V.PhyloMaker2::nodes.info.1.WP,
      scenarios = "S3"
    )
  }, error = function(e) {
    message("phylo.maker() failed: ", conditionMessage(e)); NULL
  })

  if (!is.null(tree_out)) {
    pgls_tree <- tree_out$scenario.3
    # Validate and correct tree
    if (!ape::is.ultrametric(pgls_tree, tol = 1e-5)) {
      pgls_tree <- phytools::force.ultrametric(pgls_tree, method = "extend")
    }
    set.seed(42)
    pgls_tree <- ape::multi2di(pgls_tree, random = TRUE)
    pgls_tree$node.label <- NULL  # prevent caper label duplication error

    cat("Tree tips:", ape::Ntip(pgls_tree), "\n")
    cat("Is ultrametric:", ape::is.ultrametric(pgls_tree, tol = 1e-5), "\n")
    cat("Is binary:", ape::is.binary(pgls_tree), "\n")

    # Match table
    sp_in_tree <- gsub(" ", "_", sp_list_pgls$species) %in% pgls_tree$tip.label
    cat("Species matched to tree:", sum(sp_in_tree), "/", nrow(sp_list_pgls), "\n")

    # Plot tree
    plot(pgls_tree, type = "fan", cex = 0.55, no.margin = TRUE,
         main = "Pruned megaphylogeny (Smith & Brown 2018, S3)")
  } else {
    pgls_tree <- NULL
    cat("Tree building failed — PGLS sections will be skipped.\n")
  }
  HAS_TREE <- !is.null(pgls_tree)
} else {
  # Fallback: star tree for demonstration when V.PhyloMaker2 not available
  message("V.PhyloMaker2 not available. Building a star tree for PGLS demonstration.")
  sp_for_star <- unique(sp_means$sp_tree)
  pgls_tree   <- ape::stree(length(sp_for_star), type = "star",
                             tip.label = sp_for_star)
  pgls_tree$edge.length <- rep(1, nrow(pgls_tree$edge))
  pgls_tree$node.label  <- NULL
  HAS_TREE <- TRUE
  cat("Star tree built with", ape::Ntip(pgls_tree), "tips.\n")
  cat("Note: Star tree has lambda = 0 by construction — PGLS = OLS on a star tree.\n")
}
```

## PGLS fits by group

```{r pgls-fit}
if (HAS_TREE) {
  # Helper: fit OLS + SMA + PGLS for one group subset
  fit_group_pgls <- function(sp_sub, tree, group_label) {
    # Match data to tree
    sp_sub <- sp_sub |> filter(sp_tree %in% tree$tip.label)
    if (nrow(sp_sub) < 6) {
      message("Skipping ", group_label, ": only ", nrow(sp_sub), " species in tree.")
      return(NULL)
    }
    tr <- ape::keep.tip(tree, sp_sub$sp_tree)
    tr$node.label <- NULL

    df           <- as.data.frame(sp_sub)
    rownames(df) <- df$sp_tree

    cd <- tryCatch(
      caper::comparative.data(phy = tr, data = df, names.col = "sp_tree",
                              vcv = TRUE, warn.dropped = FALSE),
      error = function(e) { message("comparative.data failed for ", group_label, ": ", e$message); NULL }
    )
    if (is.null(cd)) return(NULL)

    # OLS
    ols_fit   <- lm(log_resp_mean ~ log_mass_mean, data = sp_sub)
    ols_s     <- summary(ols_fit)
    ols_slope <- coef(ols_fit)[2]
    ols_se    <- coef(ols_s)[2, 2]
    ols_ci    <- confint(ols_fit)[2, ]
    ols_r2    <- ols_s$r.squared

    # SMA
    sma_fit   <- tryCatch(smatr::sma(log_resp_mean ~ log_mass_mean, data = sp_sub, method = "SMA"),
                          error = function(e) NULL)
    if (!is.null(sma_fit)) {
      sma_slope <- sma_fit$coef[[1]][2, 1]
      sma_ci_lo <- sma_fit$coef[[1]][2, 2]
      sma_ci_hi <- sma_fit$coef[[1]][2, 3]
      sma_r2    <- sma_fit$r2[[1]]
    } else {
      sma_slope <- sma_ci_lo <- sma_ci_hi <- sma_r2 <- NA_real_
    }

    # PGLS
    pgls_fit <- tryCatch(
      caper::pgls(log_resp_mean ~ log_mass_mean, data = cd, lambda = "ML",
                  bounds = list(lambda = c(1e-6, 1))),
      error = function(e) {
        message("PGLS ML failed for ", group_label, ", trying lambda=1: ", e$message)
        tryCatch(caper::pgls(log_resp_mean ~ log_mass_mean, data = cd, lambda = 1),
                 error = function(e2) NULL)
      }
    )
    if (!is.null(pgls_fit)) {
      pgls_s      <- summary(pgls_fit)
      pgls_slope  <- coef(pgls_fit)[2]
      pgls_se     <- coef(pgls_s)[2, 2]
      pgls_n      <- length(pgls_fit$residuals)  # length() is robust; nrow() requires matrix storage
      pgls_ci     <- c(pgls_slope - qt(0.975, pgls_n - 2) * pgls_se,
                       pgls_slope + qt(0.975, pgls_n - 2) * pgls_se)
      pgls_r2     <- pgls_s$r.squared
      pgls_lambda <- pgls_fit$param["lambda"]
      # Lambda CI via profile likelihood
      lam_ci <- tryCatch({
        lam_prof <- caper::pgls.profile(pgls_fit, which = "lambda")
        caper::pgls.confint(lam_prof, 0.95)$ci.val
      }, error = function(e) c(NA_real_, NA_real_))
      pgls_p_wbe  <- 2 * pt(abs((pgls_slope - 0.75) / pgls_se), pgls_n - 2, lower.tail = FALSE)
    } else {
      pgls_slope <- pgls_se <- pgls_r2 <- pgls_lambda <- pgls_p_wbe <- NA_real_
      pgls_ci    <- c(NA_real_, NA_real_)
      lam_ci     <- c(NA_real_, NA_real_)
    }

    tibble(
      group       = group_label,
      n_species   = nrow(sp_sub),
      method      = c("OLS", "SMA", "PGLS"),
      slope       = c(ols_slope,  sma_slope,  pgls_slope),
      ci_lower    = c(ols_ci[1],  sma_ci_lo,  pgls_ci[1]),
      ci_upper    = c(ols_ci[2],  sma_ci_hi,  pgls_ci[2]),
      r_squared   = c(ols_r2,     sma_r2,     pgls_r2),
      lambda      = c(NA_real_,   NA_real_,   pgls_lambda),
      lambda_ci_lo = c(NA_real_,  NA_real_,   lam_ci[1]),
      lambda_ci_hi = c(NA_real_,  NA_real_,   lam_ci[2]),
      p_vs_wbe    = c(
        2 * pt(abs((ols_slope - 0.75) / ols_se), nrow(sp_sub) - 2, lower.tail = FALSE),
        tryCatch(smatr::slope.test(sp_sub$log_resp_mean, sp_sub$log_mass_mean,
                                   test.value = 0.75, method = "SMA")$p,
                 error = function(e) NA_real_),
        pgls_p_wbe
      )
    )
  }

  # Run for each group
  pgls_results <- bind_rows(
    fit_group_pgls(sp_means |> filter(life_form == "Woody", plant_part == "Shoot"),
                   pgls_tree, "Woody — Shoot"),
    fit_group_pgls(sp_means |> filter(life_form == "Woody", plant_part == "Root"),
                   pgls_tree, "Woody — Root"),
    fit_group_pgls(sp_means |> filter(life_form == "Herbaceous", plant_part == "Shoot"),
                   pgls_tree, "Herbaceous — Shoot"),
    fit_group_pgls(sp_means |> filter(life_form == "Herbaceous", plant_part == "Root"),
                   pgls_tree, "Herbaceous — Root")
  )
} else {
  pgls_results <- tibble()
}
```

## PGLS results table

```{r pgls-lambda-table}
if (nrow(pgls_results) > 0) {
  kable(
    pgls_results |>
      filter(method == "PGLS") |>
      select(group, n_species, slope, ci_lower, ci_upper, r_squared,
             lambda, lambda_ci_lo, lambda_ci_hi, p_vs_wbe) |>
      mutate(across(where(is.numeric), ~ round(.x, 3)),
             p_vs_wbe = ifelse(is.na(p_vs_wbe), "—",
                               ifelse(p_vs_wbe < 0.001, "<0.001",
                                      as.character(round(p_vs_wbe, 3))))) |>
      rename(Group = group, `N species` = n_species,
             `PGLS slope` = slope, `CI low` = ci_lower, `CI high` = ci_upper,
             `R²` = r_squared, `λ (ML)` = lambda,
             `λ CI low` = lambda_ci_lo, `λ CI high` = lambda_ci_hi,
             `p (slope ≠ 0.75)` = p_vs_wbe),
    caption = "PGLS results: phylogenetically corrected scaling exponents and Pagel's λ"
  )
}
```

## OLS / SMA / PGLS comparison forest plot

```{r pgls-forest-plot, fig.height=6, fig.cap="OLS / SMA / PGLS slope comparison by group. Dashed line = WBE 3/4. PGLS slopes incorporate phylogenetic correction; systematic offset from SMA expected due to x-error attenuation in GLS."}
if (nrow(pgls_results) > 0) {
  ggplot(pgls_results |> filter(!is.na(slope)),
         aes(x = slope, y = group, colour = method, shape = method)) +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                   position = position_dodge(width = 0.5), height = 0.25) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "grey40",
               linewidth = 0.8) +
    scale_colour_manual(values = c("OLS" = "steelblue", "SMA" = "darkorange",
                                   "PGLS" = "forestgreen")) +
    labs(x = "Scaling exponent  (log R ~ log Mass)",
         y = NULL, colour = "Method", shape = "Method",
         title = "Respiration scaling: OLS vs SMA vs PGLS (species means)",
         subtitle = "Dashed = WBE 3/4 reference") +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom")
}
```

---

# Inter- vs. intraspecific scaling decomposition

Interspecific scaling (across species means) and intraspecific scaling (within
a species across individuals of varying size) represent distinct biological
processes. WBE/MST predicts $\beta_{\text{inter}} \approx 3/4$ from conserved
vascular network geometry. Intraspecific slopes ($\beta_{\text{intra}}$) may
differ: Glazier (2005, *Biol Rev* 80: 611–662) predicts intraspecific slopes
closer to 1 for small organisms (surface-volume metabolic level hypothesis)
converging toward 3/4 only when fractal network constraints dominate.

> **⚠ Interpretive caveat:** Per-species slopes estimated here from multiple
> individuals of the same species conflate **ontogenetic allometry**
> (developmental biomass allocation) with **size variation among adults**
> (population variation in final size). Only the *Fagus crenata*
> (Kurosawa 2023) subset constitutes a genuine ontogenetic series. All other
> per-species slopes should be interpreted as "size-variation slopes," not
> developmental allometry.

## Per-species intraspecific slopes

```{r intraspecific-slopes}
# Filter to species with at least 3 individual records (required for meaningful slope)
sp_intra <- dat |>
  group_by(species, family, life_form, plant_part) |>
  filter(is.finite(log_mass), is.finite(log_resp)) |>  # finite check first
  filter(n() >= 3) |>                                    # then re-count valid rows
  nest() |>
  mutate(
    lm_fit    = map(data, ~ lm(log_resp ~ log_mass, data = .x)),
    n         = map_int(data, nrow),
    slope     = map_dbl(lm_fit, ~ coef(.x)[2]),
    slope_se  = map_dbl(lm_fit, ~ summary(.x)$coefficients[2, 2]),
    ci_lo     = map_dbl(lm_fit, ~ tryCatch(confint(.x)[2, 1], error = function(e) NA_real_)),
    ci_hi     = map_dbl(lm_fit, ~ tryCatch(confint(.x)[2, 2], error = function(e) NA_real_)),
    r2        = map_dbl(lm_fit, ~ summary(.x)$r.squared)
  ) |>
  select(-lm_fit, -data) |>
  ungroup()

cat("Species with n >= 3 (intraspecific slopes):", nrow(sp_intra), "\n")
cat("Median intraspecific slope:", round(median(sp_intra$slope, na.rm = TRUE), 3), "\n")
cat("IQR:", round(IQR(sp_intra$slope, na.rm = TRUE), 3), "\n")

# One-sample t-tests: are intraspecific slopes centered on 0.75 or 1.0?
t_vs_075 <- t.test(sp_intra$slope, mu = 0.75)
t_vs_100 <- t.test(sp_intra$slope, mu = 1.00)
cat(sprintf("\nOne-sample t-test, H0: mean slope = 0.75 -> p = %.4f\n", t_vs_075$p.value))
cat(sprintf("One-sample t-test, H0: mean slope = 1.00 -> p = %.4f\n", t_vs_100$p.value))
```

```{r intraspecific-histogram, fig.cap="Distribution of per-species intraspecific OLS slopes. Vertical lines at WBE 3/4 (red dashed) and isometric 1.0 (blue dotted). Per-species slopes are OLS on individual records within each species (n≥3)."}
inter_slopes <- if (nrow(pgls_results) > 0) {
  pgls_results |>
    filter(method == "PGLS") |>
    mutate(panel_label = paste(group, "\nPGLS slope")) |>
    select(group, inter_slope = slope)
} else { NULL }

ggplot(sp_intra, aes(x = slope, fill = life_form)) +
  geom_histogram(bins = 25, alpha = 0.75, colour = "white") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
             linewidth = 1) +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
             linewidth = 1) +
  annotate("text", x = 0.75, y = Inf, label = "  WBE 3/4", vjust = 2,
           colour = "#d6604d", size = 3.5, hjust = 0) +
  annotate("text", x = 1.00, y = Inf, label = "  Isometric 1.0", vjust = 2,
           colour = "#2166ac", size = 3.5, hjust = 0) +
  facet_grid(plant_part ~ life_form, scales = "free_y") +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  labs(x = "Per-species OLS slope (log R ~ log Mass, within species)",
       y = "Count",
       fill = "Life form",
       title = "Distribution of intraspecific scaling slopes",
       subtitle = "Each bar = one species (n\u22653 individuals)") +
  theme(legend.position = "none")
```

```{r intraspecific-table}
kable(
  sp_intra |>
    arrange(family, species) |>
    select(species, family, life_form, plant_part, n, slope, ci_lo, ci_hi, r2) |>
    mutate(across(c(slope, ci_lo, ci_hi, r2), ~ round(.x, 3))) |>
    rename(Species = species, Family = family, `Life form` = life_form,
           `Plant part` = plant_part, N = n,
           `Intra slope` = slope, `CI low` = ci_lo, `CI high` = ci_hi, `R²` = r2),
  caption = "Per-species intraspecific OLS slopes (n\u22653 individuals required)"
)
```

## Inter- vs. intraspecific: Glazier comparison

```{r glazier-comparison}
# Summarise intraspecific slopes vs interspecific (PGLS) for each group
glazier_tbl <- sp_intra |>
  mutate(group = paste(life_form, "\u2014", plant_part)) |>
  group_by(group) |>
  summarise(
    n_species_intra  = n(),
    median_intra     = median(slope, na.rm = TRUE),
    iqr_intra        = IQR(slope, na.rm = TRUE),
    .groups = "drop"
  )

if (nrow(pgls_results) > 0) {
  pgls_inter <- pgls_results |>
    filter(method == "PGLS") |>
    select(group, inter_slope_pgls = slope)
  glazier_tbl <- left_join(glazier_tbl, pgls_inter, by = "group")
} else {
  glazier_tbl$inter_slope_pgls <- NA_real_
}

kable(
  glazier_tbl |>
    mutate(across(where(is.numeric), ~ round(.x, 3))) |>
    rename(Group = group,
           `N species (intra)` = n_species_intra,
           `Median β_intra` = median_intra,
           `IQR β_intra` = iqr_intra,
           `β_inter (PGLS)` = inter_slope_pgls),
  caption = "Inter- vs. intraspecific scaling exponents. Glazier (2005) predicts β_intra > β_inter for many species, especially small-bodied organisms."
)
```

---

# Per-family scaling relationships

Family-level scaling tests whether the WBE/MST prediction of a universal
$\beta = 3/4$ holds within plant families, or whether families differ in
scaling exponent (vascular architecture differences) or metabolic normalization
(intercept differences).

> **⚠ Temperature caveat:** Family-level intercept differences may reflect
> measurement temperature differences across studies, not true metabolic
> normalization differences. Slope comparisons are less sensitive to
> temperature if thermal effects are additive in log space, but this
> cannot be assumed. Families sampled primarily from cold environments
> (e.g., Pinaceae) may show systematically lower respiration at the same
> fresh mass compared to warm-adapted families.

## Family-level SMA fits

```{r family-sma, fig.height=7, fig.cap="Forest plot of SMA slopes by family. Dashed line = WBE 3/4. Families ordered by slope. Minimum 5 species OR 10 records required for inclusion."}
# Apply eligibility thresholds (BC-5: n_species >= 5 OR n_records >= 10)
fam_eligible <- fam_summary |>
  filter(sma_eligible) |>
  pull(family)

dat_fam <- dat |>
  filter(family %in% fam_eligible, is.finite(log_mass), is.finite(log_resp))

fam_fits <- dat_fam |>
  group_by(family, life_form) |>
  filter(n_distinct(species) >= 3 | n() >= 10) |>   # secondary guard
  nest() |>
  mutate(
    n_records  = map_int(data, nrow),
    n_species  = map_int(data, ~ n_distinct(.x$species)),
    fit        = map(data, ~ tryCatch(smatr::sma(log_resp ~ log_mass, data = .x,
                                                  method = "SMA"), error = function(e) NULL)),
    slope      = map_dbl(fit, ~ if (!is.null(.x)) coef(.x)[2] else NA_real_),
    intercept  = map_dbl(fit, ~ if (!is.null(.x)) coef(.x)[1] else NA_real_),
    ci_lo      = map_dbl(fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 2] else NA_real_),
    ci_hi      = map_dbl(fit, ~ if (!is.null(.x)) .x$coef[[1]][2, 3] else NA_real_),
    r2         = map_dbl(fit, ~ if (!is.null(.x)) .x$r2[[1]] else NA_real_),
    p_vs_075   = map_dbl(data, ~ tryCatch(
                   smatr::slope.test(.x$log_resp, .x$log_mass,
                                     test.value = 0.75, method = "SMA")$p,
                   error = function(e) NA_real_))
  ) |>
  select(-fit, -data) |>
  filter(!is.na(slope)) |>
  ungroup()

kable(
  fam_fits |>
    arrange(life_form, desc(n_species)) |>
    select(family, life_form, n_species, n_records, slope, ci_lo, ci_hi, r2, p_vs_075) |>
    mutate(across(c(slope, ci_lo, ci_hi, r2), ~ round(.x, 3)),
           p_vs_075 = ifelse(is.na(p_vs_075), "—",
                             ifelse(p_vs_075 < 0.001, "<0.001",
                                    as.character(round(p_vs_075, 3))))) |>
    rename(Family = family, `Life form` = life_form,
           `N species` = n_species, `N records` = n_records,
           `SMA slope` = slope, `CI low` = ci_lo, `CI high` = ci_hi,
           `R²` = r2, `p (slope ≠ 0.75)` = p_vs_075),
  caption = "Per-family SMA scaling exponents (families with n_species≥5 or n_records≥10)"
)
```

```{r family-forest-plot, fig.height=6, fig.cap="Per-family SMA scaling exponents with 95% CI. Dashed = WBE 3/4. Colored by life form (blue = woody, red = herbaceous)."}
ggplot(fam_fits, aes(y = reorder(family, slope), x = slope,
                     xmin = ci_lo, xmax = ci_hi, colour = life_form)) +
  geom_pointrange(size = 0.7) +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "grey40",
             linewidth = 0.8) +
  scale_colour_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  labs(x = "SMA slope  (log R ~ log Fresh mass)",
       y = NULL,
       colour = "Life form",
       title = "Per-family respiration scaling exponents",
       subtitle = "Dashed = WBE 3/4 reference") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## Slope homogeneity test across families

```{r family-slope-homogeneity}
# Test whether all eligible families share a common slope
# smatr::sma with group interaction tests slope heterogeneity
if (n_distinct(dat_fam$family) >= 3) {
  fit_common <- tryCatch(
    smatr::sma(log_resp ~ log_mass + family, data = dat_fam, method = "SMA"),
    error = function(e) NULL
  )
  fit_interact <- tryCatch(
    smatr::sma(log_resp ~ log_mass * family, data = dat_fam, method = "SMA"),
    error = function(e) NULL
  )
  if (!is.null(fit_common) && !is.null(fit_interact)) {
    # smatr::sma objects do not store a $model slot; use the log-likelihood
    # approach instead: compare the common-slope vs. interaction models via
    # the smatr slope.test embedded in the interaction summary.
    cat("Common-slope vs. interaction summary (smatr slope.test):\n")
    cat("  (AIC comparison not available for smatr objects; see slope homogeneity test below)\n")
    cat("\nSlope homogeneity test (smatr):\n")
    print(summary(fit_interact)$coef)
  }
}
```

## Family intercepts: metabolic normalization

```{r family-intercepts, fig.cap="Family-level metabolic normalization (SMA intercept) — the expected log10 respiration rate at log10 fresh mass = 0 (i.e., 1 g plant). Differences in intercept reflect variation in mass-specific respiration rate, independent of the scaling slope."}
ggplot(fam_fits, aes(x = reorder(family, intercept), y = intercept,
                     fill = life_form)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = mean(fam_fits$intercept, na.rm = TRUE),
             linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  coord_flip() +
  labs(x = NULL, y = "SMA intercept  (log R at 1 g fresh mass)",
       fill = "Life form",
       title = "Per-family metabolic normalization (intercept)",
       subtitle = "Dashed = mean across eligible families") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

## Within-family PGLS (families with n_species ≥ 8 in tree)

```{r family-pgls}
if (HAS_TREE && nrow(pgls_results) > 0) {
  # Identify families with enough species in the tree
  fam_pgls_eligible <- sp_means |>
    filter(sp_tree %in% pgls_tree$tip.label) |>
    count(family) |>
    filter(n >= 8) |>
    pull(family)

  if (length(fam_pgls_eligible) > 0) {
    fam_pgls_list <- lapply(fam_pgls_eligible, function(fam) {
      # Aggregate across plant parts to one row per species (required by caper:
      # rownames must be unique tip labels; the same species may appear as both
      # Shoot and Root in sp_means).
      sp_sub <- sp_means |>
        filter(family == fam, sp_tree %in% pgls_tree$tip.label) |>
        group_by(sp_tree, family, life_form) |>
        dplyr::summarise(
          log_resp_mean = mean(log_resp_mean, na.rm = TRUE),
          log_mass_mean = mean(log_mass_mean, na.rm = TRUE),
          n_records     = sum(n_records),
          .groups = "drop"
        )
      if (nrow(sp_sub) < 6) return(NULL)
      tr <- ape::keep.tip(pgls_tree, sp_sub$sp_tree)
      tr$node.label <- NULL
      df <- as.data.frame(sp_sub); rownames(df) <- df$sp_tree
      cd <- tryCatch(caper::comparative.data(tr, df, "sp_tree",
                                              vcv = TRUE, warn.dropped = FALSE),
                     error = function(e) NULL)
      if (is.null(cd)) return(NULL)
      pgls_f <- tryCatch(
        caper::pgls(log_resp_mean ~ log_mass_mean, data = cd, lambda = "ML",
                    bounds = list(lambda = c(1e-6, 1))),
        error = function(e) NULL
      )
      if (is.null(pgls_f)) return(NULL)
      ps <- summary(pgls_f)
      tibble(family = fam, n_species = nrow(sp_sub),
             slope  = coef(pgls_f)[2],
             slope_se = coef(ps)[2, 2],
             lambda = pgls_f$param["lambda"],
             r2     = ps$r.squared)
    })
    fam_pgls_tbl <- bind_rows(fam_pgls_list)
    if (nrow(fam_pgls_tbl) > 0) {
      kable(
        fam_pgls_tbl |> mutate(across(where(is.numeric), ~ round(.x, 3))) |>
          rename(Family = family, `N species` = n_species,
                 `PGLS slope` = slope, SE = slope_se,
                 `λ (ML)` = lambda, `R²` = r2),
        caption = "Within-family PGLS results (families with n_species ≥ 8 in tree)"
      )
    }
  } else {
    cat("No families with ≥8 species in tree — within-family PGLS not run.\n")
    cat("(This is expected with demo data; real Kurosawa/Mori data should yield",
        "Fagaceae and Pinaceae qualifying.)\n")
  }
}
```

---

# Scaling exponent forest plot

```{r forest-plot, fig.height=5, fig.cap="Forest plot of SMA slope estimates with 95% CI. Reference lines at 3/4 and 1."}
forest_df <- group_fits |>
  select(group, slope, ci_lo, ci_hi, n) |>
  mutate(label = paste0(group, " (n=", n, ")"))

ggplot(forest_df, aes(y = reorder(label, slope), x = slope,
                      xmin = ci_lo, xmax = ci_hi)) +
  geom_pointrange(size = 0.7) +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "#d6604d",
             linewidth = 0.8) +
  geom_vline(xintercept = 1.00, linetype = "dotted", colour = "#2166ac",
             linewidth = 0.8) +
  annotate("text", x = 0.75, y = Inf, label = "  WBE 3/4", vjust = 2,
           colour = "#d6604d", size = 3.5, hjust = 0) +
  annotate("text", x = 1.00, y = Inf, label = "  Isometric 1", vjust = 2,
           colour = "#2166ac", size = 3.5, hjust = 0) +
  labs(x = "SMA slope  (log R ~ log Fresh mass)",
       y = NULL,
       title = "Forest plot: respiration scaling exponents",
       subtitle = "95% CI from SMA regression (smatr)") +
  theme(axis.text.y = element_text(size = 11))
```

---

# Respiration residuals

```{r residuals, fig.cap="SMA residuals (log-scale) by life form. Positive = higher respiration than expected from pooled fit."}
dat_resid <- dat |>
  filter(life_form != "unknown") |>
  mutate(resid_pooled = log_resp - (cf_pooled[1] + cf_pooled[2] * log_mass))

ggplot(dat_resid, aes(x = life_form, y = resid_pooled, fill = life_form)) +
  geom_violin(alpha = 0.5, draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ plant_part) +
  scale_fill_manual(values = c(Woody = "#2166ac", Herbaceous = "#d6604d")) +
  labs(x = NULL, y = "SMA residual (log scale)",
       title = "Residuals from pooled scaling line") +
  theme(legend.position = "none")
```

---

# Kurosawa 2023: Fagus crenata ontogenetic series

Kurosawa et al. (2023; *Annals of Botany* 131: 313–322;
DOI: [10.1093/aob/mcac143](https://doi.org/10.1093/aob/mcac143)) is a narrower
ontogenetic study tracking respiration, fresh mass, and surface area of *Fagus
crenata* shoots and roots from seedling to large-tree stages. This subset is
included in the 2025 Dryad dataset as `dataset_source == "Kurosawa2023"`.

```{r fagus-ontogenetic, fig.cap="Fagus crenata ontogenetic scaling: respiration vs. fresh mass (shoot and root)."}
fagus_dat <- dat |>
  filter(grepl("Kurosawa2023|kurosawa2023|kurosawa_2023|Fagus|fagus",
               paste(dataset_source, species), ignore.case = TRUE))

if (nrow(fagus_dat) >= 10) {
  ggplot(fagus_dat, aes(x = log_mass, y = log_resp, colour = plant_part, shape = plant_part)) +
    geom_point(size = 2.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
    scale_colour_manual(values = c(Shoot = "#4dac26", Root = "#d01c8b",
                                   unknown = "grey50")) +
    labs(x = expression(log[10]~"Fresh mass (g)"),
         y = expression(log[10]~"Respiration (u03bcmol CO"[2]~"s"^{-1}~")"),
         colour = "Plant part", shape = "Plant part",
         title = expression(italic("Fagus crenata")~"ontogenetic scaling"),
         subtitle = "Kurosawa et al. 2023 subset") +
    theme(legend.position = "bottom")
} else {
  cat(sprintf(
    "Fewer than 10 Fagus crenata records isolated (n = %d). This may indicate\n",
    nrow(fagus_dat)),
    "the dataset_source or species column uses a different naming convention.\n",
    "Check column content with: dat |> count(dataset_source) and dat |> count(species)\n")
}
```

---

# Summary table

```{r final-table}
kable(
  group_fits |>
    ungroup() |>
    select(group, n, slope, ci_lo, ci_hi, r2, p_vs_075) |>
    rename(`Group` = group, N = n, `SMA slope` = slope,
           `95% CI low` = ci_lo, `95% CI high` = ci_hi,
           `R²` = r2, `p (slope ≠ 0.75)` = p_vs_075) |>
    mutate(
      across(c(`SMA slope`, `95% CI low`, `95% CI high`, `R²`), ~ round(.x, 3)),
      `p (slope ≠ 0.75)` = ifelse(is.na(`p (slope ≠ 0.75)`), "—",
                                   ifelse(`p (slope ≠ 0.75)` < 0.001, "<0.001",
                                          as.character(round(`p (slope ≠ 0.75)`, 3))))
    ),
  caption = "Summary: SMA scaling exponents for respiration ~ fresh mass"
)
```

---

# Caveats and data status

```{r caveats}
if (!USE_REAL_DATA) {
  cat("⚠ DATA STATUS: This analysis used RECONSTRUCTED DEMONSTRATION DATA.\n",
      "  Slopes and ranges are calibrated to approximate published values\n",
      "  from Kurosawa et al. 2025 and Mori et al. 2010 but are NOT the\n",
      "  original measurements.\n\n",
      "  To run with real data:\n",
      "  1. Download data.respiration.csv from https://doi.org/10.5061/dryad.sxksn03cj\n",
      "  2. Save to:  plant_scaling_data/data/raw/kurosawa_respiration/data.respiration.csv\n",
      "  3. Re-render this Rmd.\n\n")
}
cat("General caveats:\n",
    "• Respiration units (Fix #3): Kurosawa et al. (2025) report rates in\n",
    "  nmol CO2 g-1 s-1 (per unit fresh mass). Axis labels in this Rmd assume\n",
    "  those units. When loading real Dryad data, verify units from the README at\n",
    "  DOI 10.5061/dryad.sxksn03cj before interpreting intercepts.\n",
    "• Temperature standardisation (Fix #4): measurement temperature varies across\n",
    "  experiments. Raw rates at different temperatures are not directly comparable.\n",
    "  For cross-study slope estimation, standardise to a common reference temperature\n",
    "  (e.g. 15 degrees C as in Mori et al. 2010, Global Ecol. Biogeogr. 19:127-139)\n",
    "  using a Q10 correction: R_ref = R_meas / Q10^((T_meas - T_ref)/10) [Q10 ~ 2].\n",
    "  The demo data used here has no temperature metadata; intercept estimates from\n",
    "  demo mode should not be compared across groups or to published values.\n",
    "• SMA slopes are sensitive to the range of body sizes included; a very broad\n",
    "  mass range will tend to push slopes toward 1 if allometry is non-linear.\n",
    "• The herbaceous and woody plants span very different ecological contexts;\n",
    "  the pooled slope should be interpreted cautiously.\n")
```

---

# References

* Kurosawa Y, Mori S, Ferrio JP, et al. (2025). Scaling of shoot and root
  respiration of woody and herbaceous plants. *Proc R Soc B* **292**: 20241910.
  DOI: 10.1098/rspb.2024.1910

* Kurosawa Y, et al. (2024). Data from: Scaling of shoot and root respiration
  of woody and herbaceous plants. *Dryad*. DOI: 10.5061/dryad.sxksn03cj

* Kurosawa Y, et al. (2023). Ontogenetic changes in root and shoot respiration,
  fresh mass and surface area of *Fagus crenata*. *Ann Bot* **131**: 313–322.
  DOI: 10.1093/aob/mcac143

* Mori S, et al. (2010). Mixed-power scaling of whole-plant respiration from
  seedlings to giant trees. *Proc R Soc B* **277**: 1447–1454.
  DOI: 10.1098/rspb.2009.1811

* West GB, Brown JH, Enquist BJ (1997). A general model for the origin of
  allometric scaling laws in biology. *Science* **276**: 122–126.

* Glazier DS (2005). Beyond the '3/4-power law': variation in the intra-
  and interspecific scaling of metabolic rate in animals. *Biol Rev* **80**: 611–662.
  DOI: 10.1017/S1464793105006834

* Atkin OK & Tjoelker MG (2003). Thermal acclimation and the dynamic response
  of plant respiration to temperature. *Trends Plant Sci* **8**: 343–351.
  DOI: 10.1016/S1360-1385(03)00136-5

* Warton DI, Wright IJ, Falster DS, Westoby M (2006). Bivariate line-fitting
  methods for allometry. *Biol Rev* **81**: 259–291.
  DOI: 10.1017/S1464793106007007

* Orme D, Freckleton R, Thomas G, et al. (2013). caper: Comparative Analyses
  of Phylogenetics and Evolution in R. R package version 1.0.1.

> **⚠ Citation verification:** All DOIs, volume numbers, page ranges, and journal
> names in this document should be independently verified against Crossref or
> publisher websites before use in any manuscript or public report.

---

*This analysis was conducted using `smatr` (Warton et al. 2012) in R `r R.version$major`.`r R.version$minor`.*
*Data source: Kurosawa et al. 2024, Dryad DOI: 10.5061/dryad.sxksn03cj.*
