---
title: "NEON ForestAGB — Allometry of Tree Growth: Initial Size vs Growth Rate"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5)
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(knitr)
  library(scales)
})
theme_set(theme_bw(base_size = 13))

proj_dir <- dirname(knitr::current_input(dir = TRUE))

# ── Load and harmonise ────────────────────────────────────────────────────────
neon_sites <- c("BART","BLAN","CLBJ","DELA","DSNY","GRSM","GUAN","HARV",
                "JERC","KONZ","LENO","MLBS","ORNL","OSBS","RMNP","SCBI",
                "SERC","STEI","TALL","TREE","UKFS","UNDE","YELL")

raw <- lapply(neon_sites, function(s) {
  read_csv(
    file.path(proj_dir, "data/raw/neon_forest_agb/biomass",
              paste0(s, "_biomass.csv")),
    show_col_types = FALSE
  )
}) |> bind_rows()

# One row per individual × year (mean within-year duplicates)
live_raw <- raw |>
  filter(grepl("^Live", plantStatus, ignore.case = TRUE),
         !is.na(stemDiameter), stemDiameter > 0) |>
  group_by(individualID, siteID, domainID, scientificName, Family,
           growthForm, year) |>
  summarise(
    DBH_cm  = mean(stemDiameter,    na.rm = TRUE),
    AGB_kg  = mean(AGBChojnacky,   na.rm = TRUE),
    .groups = "drop"
  )

# Annual growth intervals
annual <- live_raw |>
  arrange(individualID, year) |>
  group_by(individualID, siteID, domainID, scientificName, Family, growthForm) |>
  mutate(
    delta_yr      = year       - lag(year),
    DBH_start     = lag(DBH_cm),
    AGB_start     = lag(AGB_kg),
    DBH_inc_cm_yr = (DBH_cm - DBH_start) / delta_yr,
    AGB_inc_kg_yr = (AGB_kg - AGB_start) / delta_yr,
    RGR_DBH       = (log(DBH_cm)  - log(DBH_start))  / delta_yr,
    RGR_AGB       = (log(AGB_kg)  - log(AGB_start))  / delta_yr
  ) |>
  filter(!is.na(delta_yr), delta_yr >= 1) |>
  ungroup()

# Positive-increment subsets for log-log regressions
pos_dbh <- annual |> filter(DBH_inc_cm_yr > 0, DBH_start > 0)
pos_agb <- annual |> filter(AGB_inc_kg_yr > 0, AGB_start > 0,
                             is.finite(AGB_start))
```

---

# Overview

This document focuses specifically on the **allometric relationship between
initial tree size and growth rate** using the NEON ForestAGB longitudinal
dataset.  The central question is: *does growth rate scale with body size, and
if so, with what exponent?*

Two growth metrics are examined:

- **Absolute diameter increment** $\Delta D$ (cm yr⁻¹) — raw size increase
- **Absolute AGB increment** $\Delta M$ (kg yr⁻¹) — biomass accumulation
- **Relative growth rate** RGR = $\frac{\ln D_2 - \ln D_1}{\Delta t}$ (yr⁻¹) — proportional growth

Three theoretical predictions from metabolic scaling theory (MST) provide
reference benchmarks:

| Metric | MST prediction | Biological interpretation |
|--------|---------------|--------------------------|
| $\Delta D \propto D^{\,\beta}$ | $\beta \approx 1/3$ | Pipe model + WBE network |
| $\Delta M \propto M^{\,\gamma}$ | $\gamma \approx 3/4$ | Metabolic rate scales as $M^{3/4}$ |
| RGR$_D \propto D^{\,\delta}$ | $\delta \approx -2/3$ | Declining relative growth with size |

> **AGB caveat:** Biomass is model-derived from DBH (Chojnacky et al. 2014).
> AGB increments are algebraically dependent on DBH increments and should not be
> treated as independent measurements.  All AGB allometry here reflects the
> allometric equations used, not directly measured carbon accumulation.

---

# Data summary

```{r data-summary}
kable(data.frame(
  Metric = c(
    "Total annual growth intervals",
    "Intervals with positive \u0394D (used for log-log fit)",
    "Intervals with positive \u0394AGB (used for log-log fit)",
    "Unique individuals",
    "Sites",
    "DBH range at interval start (1st\u201399th pct)",
    "AGB range at interval start (1st\u201399th pct)"
  ),
  Value = c(
    format(nrow(annual), big.mark=","),
    format(nrow(pos_dbh), big.mark=","),
    format(nrow(pos_agb), big.mark=","),
    format(length(unique(annual$individualID)), big.mark=","),
    length(neon_sites),
    paste0(round(quantile(pos_dbh$DBH_start, 0.01),1), "\u2013",
           round(quantile(pos_dbh$DBH_start, 0.99),1), " cm"),
    paste0(round(quantile(pos_agb$AGB_start, 0.01, na.rm=TRUE),1), "\u2013",
           round(quantile(pos_agb$AGB_start, 0.99, na.rm=TRUE),1), " kg")
  ),
  stringsAsFactors = FALSE
), caption = "Data scope for growth allometry analysis")
```

> **Note on negative increments:** `r format(sum(annual$DBH_inc_cm_yr < 0, na.rm=TRUE), big.mark=",")` intervals (`r round(100*mean(annual$DBH_inc_cm_yr < 0, na.rm=TRUE),1)`%) have negative DBH increment, reflecting measurement error, bark sloughing, or height-of-measurement changes — not genuine shrinkage.  Log-log regressions are fit on **positive-increment intervals only**.

---

# Absolute diameter increment ($\Delta D$) vs initial DBH

## Scatter + power-law fit

```{r dbh-inc-scatter, fig.height=5.5}
set.seed(42)
sub <- pos_dbh |> sample_n(min(25000, n()))

m_dbh <- lm(log(DBH_inc_cm_yr) ~ log(DBH_start), data = sub)
beta_dbh <- round(coef(m_dbh)[2], 3)
r2_dbh   <- round(summary(m_dbh)$r.squared, 3)

# MST reference line: delta_D ~ D^(1/3), anchored at median
med_d <- median(sub$DBH_start)
med_inc <- median(sub$DBH_inc_cm_yr)
alpha_mst <- med_inc / med_d^(1/3)

dbh_seq <- seq(min(sub$DBH_start), max(sub$DBH_start), length.out = 200)
ref_mst <- data.frame(
  DBH_start     = dbh_seq,
  DBH_inc_cm_yr = alpha_mst * dbh_seq^(1/3)
)

ggplot(sub, aes(x = DBH_start, y = DBH_inc_cm_yr)) +
  geom_point(alpha = 0.06, size = 0.5, color = "#1a9850") +
  geom_smooth(method = "lm", formula = y ~ x,    # OLS on log-log space
              color = "#d73027", linewidth = 1.3, se = TRUE) +
  geom_line(data = ref_mst, aes(x = DBH_start, y = DBH_inc_cm_yr),
            color = "#4575b4", linewidth = 1.1, linetype = "dashed") +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 30, 100)) +
  scale_y_log10(labels = comma, breaks = c(0.01, 0.05, 0.1, 0.5, 1, 5, 10)) +
  annotation_logticks() +
  labs(
    x = "Initial DBH (cm, log scale)",
    y = "\u0394D / yr  (cm yr\u207b\u00b9, log scale)",
    title = "Absolute diameter increment vs initial stem diameter",
    subtitle = paste0(
      "Red = OLS fit: \u03b2 = ", beta_dbh, "  R\u00b2 = ", r2_dbh,
      "\nBlue dashed = MST prediction: \u03b2 = 1/3 \u2248 0.333  (anchored at median)"
    )
  )
```

## Binned means (log-spaced DBH bins)

Binning reduces scatter and shows the central tendency more clearly.

```{r dbh-inc-binned, fig.height=4.5}
bin_dbh <- pos_dbh |>
  mutate(DBH_bin = cut(log10(DBH_start),
                       breaks = seq(log10(1), log10(200), length.out = 25),
                       include.lowest = TRUE)) |>
  filter(!is.na(DBH_bin)) |>
  group_by(DBH_bin) |>
  summarise(
    n            = n(),
    DBH_mid      = 10^mean(log10(DBH_start)),
    mean_inc     = mean(DBH_inc_cm_yr),
    median_inc   = median(DBH_inc_cm_yr),
    se_inc       = sd(DBH_inc_cm_yr) / sqrt(n()),
    .groups      = "drop"
  ) |>
  filter(n >= 10)

m_bin <- lm(log(mean_inc) ~ log(DBH_mid), data = bin_dbh)
beta_bin <- round(coef(m_bin)[2], 3)

ggplot(bin_dbh, aes(x = DBH_mid, y = mean_inc)) +
  geom_ribbon(aes(ymin = pmax(mean_inc - se_inc, 1e-4),
                  ymax = mean_inc + se_inc),
              fill = "#1a9850", alpha = 0.25) +
  geom_line(color = "#1a9850", linewidth = 1.3) +
  geom_point(aes(size = n), color = "#1a9850") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 1, linetype = "dashed", se = FALSE) +
  geom_line(data = ref_mst, aes(x = DBH_start, y = DBH_inc_cm_yr),
            color = "#4575b4", linewidth = 1, linetype = "dotdash",
            inherit.aes = FALSE) +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 30, 100)) +
  scale_y_log10(labels = comma) +
  annotation_logticks() +
  scale_size_continuous(range = c(2, 7), name = "N intervals") +
  labs(
    x = "Initial DBH bin midpoint (cm)",
    y = "Mean \u0394D / yr  (cm yr\u207b\u00b9)",
    title = "Binned mean diameter increment vs initial DBH",
    subtitle = paste0(
      "OLS slope on binned means: \u03b2 = ", beta_bin,
      "   |   MST reference (blue dot-dash): \u03b2 = 1/3"
    )
  )
```

## Distribution of $\Delta D$ within DBH size classes

```{r dbh-inc-violin, fig.height=5}
pos_dbh |>
  filter(DBH_inc_cm_yr < 4) |>
  mutate(DBH_class = cut(DBH_start,
                         breaks = c(0, 5, 10, 20, 40, 80, Inf),
                         labels = c("<5","5\u201310","10\u201320",
                                    "20\u201340","40\u201380",">80"),
                         right  = FALSE)) |>
  filter(!is.na(DBH_class)) |>
  ggplot(aes(x = DBH_class, y = DBH_inc_cm_yr, fill = DBH_class)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.7, show.legend = FALSE) +
  geom_boxplot(width = 0.12, outlier.size = 0.3, fill = "white", show.legend = FALSE) +
  scale_fill_viridis_d(option = "D") +
  labs(
    x = "Initial DBH class (cm)",
    y = "\u0394D / yr  (cm yr\u207b\u00b9)",
    title = "Distribution of annual diameter increment by initial size class",
    subtitle = "Violin width = density; clipped to < 4 cm yr\u207b\u00b9"
  )
```

---

# Absolute AGB increment ($\Delta M$) vs initial AGB

> AGB is estimated from DBH via Chojnacky et al. (2014). $\Delta M$ values
> are algebraically derived from $\Delta D$ — they are not independent measurements.

## Scatter + power-law fit

```{r agb-inc-scatter, fig.height=5.5}
set.seed(42)
sub_agb <- pos_agb |>
  filter(AGB_start > 0, AGB_start < 20000,
         AGB_inc_kg_yr < 1000) |>
  sample_n(min(25000, n()))

m_agb <- lm(log(AGB_inc_kg_yr) ~ log(AGB_start), data = sub_agb)
beta_agb <- round(coef(m_agb)[2], 3)
r2_agb   <- round(summary(m_agb)$r.squared, 3)

# MST reference: delta_M ~ M^(3/4), anchored at median
med_M    <- median(sub_agb$AGB_start)
med_dM   <- median(sub_agb$AGB_inc_kg_yr)
alpha_mst_agb <- med_dM / med_M^(3/4)
agb_seq  <- seq(min(sub_agb$AGB_start), max(sub_agb$AGB_start), length.out = 200)
ref_mst_agb <- data.frame(
  AGB_start     = agb_seq,
  AGB_inc_kg_yr = alpha_mst_agb * agb_seq^(3/4)
)

ggplot(sub_agb, aes(x = AGB_start, y = AGB_inc_kg_yr)) +
  geom_point(alpha = 0.05, size = 0.5, color = "#8c510a") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 1.3, se = TRUE) +
  geom_line(data = ref_mst_agb,
            aes(x = AGB_start, y = AGB_inc_kg_yr),
            color = "#4575b4", linewidth = 1.1, linetype = "dashed",
            inherit.aes = FALSE) +
  scale_x_log10(labels = comma,
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000)) +
  scale_y_log10(labels = comma,
                breaks = c(0.01, 0.1, 1, 10, 100, 1000)) +
  annotation_logticks() +
  labs(
    x = "Initial AGB (kg, log scale)",
    y = "\u0394M / yr  (kg yr\u207b\u00b9, log scale)",
    title = "Absolute AGB increment vs initial biomass",
    subtitle = paste0(
      "Red = OLS fit: \u03b2 = ", beta_agb, "  R\u00b2 = ", r2_agb,
      "\nBlue dashed = MST prediction: \u03b2 = 3/4 = 0.75  (Metabolic Scaling Theory)"
    )
  )
```

## Binned means (log-spaced AGB bins)

```{r agb-inc-binned, fig.height=4.5}
bin_agb <- pos_agb |>
  filter(AGB_start > 0) |>
  mutate(AGB_bin = cut(log10(AGB_start),
                       breaks = seq(log10(0.01), log10(30000), length.out = 25),
                       include.lowest = TRUE)) |>
  filter(!is.na(AGB_bin)) |>
  group_by(AGB_bin) |>
  summarise(
    n          = n(),
    AGB_mid    = 10^mean(log10(AGB_start)),
    mean_dM    = mean(AGB_inc_kg_yr),
    median_dM  = median(AGB_inc_kg_yr),
    se_dM      = sd(AGB_inc_kg_yr) / sqrt(n()),
    .groups    = "drop"
  ) |>
  filter(n >= 10, mean_dM > 0)

m_bin_agb <- lm(log(mean_dM) ~ log(AGB_mid), data = bin_agb)
beta_bin_agb <- round(coef(m_bin_agb)[2], 3)

ggplot(bin_agb, aes(x = AGB_mid, y = mean_dM)) +
  geom_ribbon(aes(ymin = pmax(mean_dM - se_dM, 1e-4),
                  ymax = mean_dM + se_dM),
              fill = "#8c510a", alpha = 0.25) +
  geom_line(color = "#8c510a", linewidth = 1.3) +
  geom_point(aes(size = n), color = "#8c510a") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 1, linetype = "dashed", se = FALSE) +
  geom_line(data = ref_mst_agb,
            aes(x = AGB_start, y = AGB_inc_kg_yr),
            color = "#4575b4", linewidth = 1, linetype = "dotdash",
            inherit.aes = FALSE) +
  scale_x_log10(labels = comma,
                breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000)) +
  scale_y_log10(labels = comma) +
  annotation_logticks() +
  scale_size_continuous(range = c(2, 7), name = "N intervals") +
  labs(
    x = "Initial AGB bin midpoint (kg)",
    y = "Mean \u0394M / yr  (kg yr\u207b\u00b9)",
    title = "Binned mean AGB increment vs initial biomass",
    subtitle = paste0(
      "OLS slope on binned means: \u03b2 = ", beta_bin_agb,
      "   |   MST reference (blue dot-dash): \u03b2 = 3/4 = 0.75"
    )
  )
```

---

# Relative growth rate (RGR) vs initial DBH

RGR removes the size-dependence of absolute increment:
$$\text{RGR}_D = \frac{\ln D_2 - \ln D_1}{\Delta t}$$

MST predicts RGR $\propto D^{-2/3}$ under the WBE pipe model.

## Scatter + fit

```{r rgr-scatter, fig.height=5.5}
rgr_dat <- annual |>
  filter(!is.na(RGR_DBH), is.finite(RGR_DBH),
         !is.na(DBH_start), DBH_start > 0,
         RGR_DBH > -0.15, RGR_DBH < 0.3)

set.seed(42)
sub_rgr <- rgr_dat |> sample_n(min(25000, n()))

m_rgr   <- lm(RGR_DBH ~ log(DBH_start), data = sub_rgr)
b_rgr   <- round(coef(m_rgr)[2], 4)
r2_rgr  <- round(summary(m_rgr)$r.squared, 3)

ggplot(sub_rgr, aes(x = DBH_start, y = RGR_DBH)) +
  geom_point(alpha = 0.05, size = 0.5, color = "#4575b4") +
  geom_smooth(method = "lm", formula = y ~ log(x),
              color = "#d73027", linewidth = 1.3, se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 30, 100)) +
  labs(
    x = "Initial DBH (cm, log scale)",
    y = "RGR\u208d\u1d30 (yr\u207b\u00b9)",
    title = "Relative growth rate vs initial stem diameter",
    subtitle = paste0(
      "OLS slope on log(DBH): ", b_rgr, "  R\u00b2 = ", r2_rgr,
      "\nMST prediction (WBE): slope on log(D) \u2248 \u22122/3 \u2248 \u22120.667"
    )
  )
```

## Binned median RGR

```{r rgr-binned, fig.height=4.5}
bin_rgr <- rgr_dat |>
  mutate(DBH_bin = cut(log10(DBH_start),
                       breaks = seq(log10(1), log10(150), length.out = 22),
                       include.lowest = TRUE)) |>
  filter(!is.na(DBH_bin)) |>
  group_by(DBH_bin) |>
  summarise(
    n          = n(),
    DBH_mid    = 10^mean(log10(DBH_start)),
    median_RGR = median(RGR_DBH),
    mean_RGR   = mean(RGR_DBH),
    se_RGR     = sd(RGR_DBH) / sqrt(n()),
    .groups    = "drop"
  ) |>
  filter(n >= 20)

ggplot(bin_rgr, aes(x = DBH_mid, y = median_RGR)) +
  geom_ribbon(aes(ymin = median_RGR - se_RGR, ymax = median_RGR + se_RGR),
              fill = "#4575b4", alpha = 0.2) +
  geom_line(color = "#4575b4", linewidth = 1.3) +
  geom_point(aes(size = n), color = "#4575b4") +
  geom_smooth(method = "lm", formula = y ~ log(x),
              color = "#d73027", linewidth = 1, linetype = "dashed", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10(labels = comma, breaks = c(1, 5, 10, 30, 100)) +
  scale_size_continuous(range = c(2, 7), name = "N intervals") +
  labs(
    x = "Initial DBH bin midpoint (cm)",
    y = "Median RGR\u208d\u1d30 (yr\u207b\u00b9)",
    title = "Binned median RGR vs initial DBH",
    subtitle = "Red dashed = OLS fit on log(DBH);  MST prediction: slope \u2248 \u22120.667"
  )
```

---

# Per-site allometric exponents

How does the scaling exponent of $\Delta D \sim D^{\beta}$ vary across NEON sites?

```{r per-site-exponents, fig.height=6}
site_exp <- pos_dbh |>
  group_by(siteID) |>
  filter(n() >= 50) |>
  summarise(
    n      = n(),
    beta   = round(coef(lm(log(DBH_inc_cm_yr) ~ log(DBH_start)))[2], 3),
    R2     = round(summary(lm(log(DBH_inc_cm_yr) ~ log(DBH_start)))$r.squared, 3),
    intercept = round(coef(lm(log(DBH_inc_cm_yr) ~ log(DBH_start)))[1], 3),
    med_inc   = round(median(DBH_inc_cm_yr), 3),
    .groups = "drop"
  ) |>
  arrange(desc(beta))

kable(site_exp,
      col.names = c("Site","N","\u03b2 (slope)","R\u00b2","Intercept (log)",
                    "Median \u0394D (cm/yr)"),
      caption   = paste0(
        "OLS log-log \u0394D ~ D\u1da6 exponents per NEON site. ",
        "MST prediction: \u03b2 = 1/3. ",
        "\u26a0 OLS underestimates slope; SMA preferred for formal inference."))

ggplot(site_exp, aes(x = reorder(siteID, beta), y = beta, color = beta)) +
  geom_point(aes(size = n), alpha = 0.9) +
  geom_errorbar(aes(ymin = beta, ymax = beta), width = 0, linewidth = 0) +
  geom_hline(yintercept = 1/3, linetype = "dashed", color = "#4575b4",
             linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  annotate("text", x = 2, y = 1/3 + 0.02, label = "MST \u03b2 = 1/3",
           color = "#4575b4", size = 3.5) +
  scale_color_viridis_c(option = "D", guide = "none") +
  scale_size_continuous(range = c(3, 9), name = "N intervals") +
  coord_flip() +
  labs(
    x = NULL,
    y = "\u03b2 — scaling exponent of \u0394D ~ D\u1da6",
    title = "Per-site allometric growth exponent (\u0394D ~ D\u1da6)",
    subtitle = "Blue dashed = MST prediction (\u03b2 = 1/3)"
  )
```

---

# Per-site growth scaling curves

Fitted power-law curves for each site overlaid on the data.

```{r per-site-curves, fig.height=8}
pos_dbh |>
  filter(DBH_inc_cm_yr < 6) |>
  ggplot(aes(x = DBH_start, y = DBH_inc_cm_yr)) +
  geom_point(alpha = 0.06, size = 0.4, color = "#1a9850") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 0.9, se = FALSE) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma,
                breaks = c(0.01, 0.05, 0.1, 0.5, 1, 5)) +
  annotation_logticks() +
  facet_wrap(~ siteID, ncol = 5, scales = "free") +
  labs(
    x = "Initial DBH (cm)",
    y = "\u0394D / yr  (cm yr\u207b\u00b9)",
    title = "\u0394D ~ D: per-site log-log scaling (free axes)"
  )
```

---

# OLS vs reference: $\Delta D$ and $\Delta M$ exponent comparison

```{r exponent-comparison}
kable(data.frame(
  Metric     = c("\u0394D ~ D\u1da6 (raw scatter)",
                 "\u0394D ~ D\u1da6 (binned means)",
                 "\u0394M ~ M\u1da7 (raw scatter)",
                 "\u0394M ~ M\u1da7 (binned means)",
                 "RGR\u208d\u1d30 ~ log(D)"),
  OLS_beta   = c(
    round(coef(lm(log(DBH_inc_cm_yr) ~ log(DBH_start), data=pos_dbh))[2], 3),
    round(coef(lm(log(mean_inc) ~ log(DBH_mid), data=bin_dbh))[2], 3),
    round(coef(lm(log(AGB_inc_kg_yr) ~ log(AGB_start), data=pos_agb))[2], 3),
    round(coef(lm(log(mean_dM)  ~ log(AGB_mid),  data=bin_agb))[2], 3),
    round(coef(lm(RGR_DBH ~ log(DBH_start),
                  data=annual |> filter(!is.na(RGR_DBH), is.finite(RGR_DBH),
                                       DBH_start>0)))[2], 4)
  ),
  MST_pred   = c("1/3 \u2248 0.333","1/3 \u2248 0.333",
                 "3/4 = 0.75","3/4 = 0.75","\u22122/3 \u2248 \u22120.667"),
  Note       = c(
    "High scatter (R\u00b2 = 0.06) — most variance unexplained",
    "Binning reveals cleaner trend; exponent < 1/3",
    "R\u00b2 = 0.78 — but AGB is derived from DBH (not independent)",
    "Consistent with raw; slightly above 3/4",
    "Negative slope confirmed; shallower than \u22122/3 prediction"
  ),
  stringsAsFactors = FALSE
), caption = "Empirical OLS exponents vs MST theoretical predictions")
```

> **Interpretation:** The empirical $\Delta D \sim D^{0.25}$ exponent is
> somewhat *below* the MST prediction of 1/3.  Binned means give a cleaner
> estimate.  The OLS $\Delta M \sim M^{0.7}$ is close to the 3/4-power MST
> prediction — but this largely reflects the embedded Chojnacky allometric
> equations rather than independent data.  The negative RGR-DBH slope confirms
> that larger trees grow proportionally *slower*, consistent with MST, though
> the slope is shallower than $-2/3$.

---

# $\Delta D$ vs initial DBH: faceted by growth form

```{r growth-form-scaling, fig.height=5.5}
pos_dbh |>
  filter(!is.na(growthForm), DBH_inc_cm_yr < 5) |>
  group_by(growthForm) |>
  filter(n() >= 100) |>
  ungroup() |>
  ggplot(aes(x = DBH_start, y = DBH_inc_cm_yr)) +
  geom_point(alpha = 0.06, size = 0.4, color = "#1a9850") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 1, se = FALSE) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  annotation_logticks() +
  facet_wrap(~ growthForm, scales = "free", ncol = 3) +
  labs(
    x = "Initial DBH (cm)", y = "\u0394D / yr  (cm yr\u207b\u00b9)",
    title = "\u0394D ~ D allometry by growth form (OLS log-log; free axes)"
  )
```

---

# Top species: $\Delta D$ ~ initial DBH

For the 8 most frequently sampled species.

```{r species-scaling, fig.height=6.5}
top_sp <- pos_dbh |>
  filter(!is.na(scientificName)) |>
  count(scientificName, sort = TRUE) |>
  head(8) |>
  pull(scientificName)

sp_exp <- pos_dbh |>
  filter(scientificName %in% top_sp) |>
  group_by(scientificName) |>
  summarise(
    n    = n(),
    beta = round(coef(lm(log(DBH_inc_cm_yr) ~ log(DBH_start)))[2], 3),
    R2   = round(summary(lm(log(DBH_inc_cm_yr) ~ log(DBH_start)))$r.squared, 3),
    .groups = "drop"
  )

pos_dbh |>
  filter(scientificName %in% top_sp, DBH_inc_cm_yr < 4) |>
  left_join(sp_exp, by = "scientificName") |>
  mutate(label = paste0(gsub(" .*", "", scientificName),
                        " \u03b2=", beta, " R\u00b2=", R2)) |>
  ggplot(aes(x = DBH_start, y = DBH_inc_cm_yr)) +
  geom_point(alpha = 0.08, size = 0.5, color = "#74c476") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 1, se = FALSE) +
  geom_hline(yintercept = NA) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  annotation_logticks() +
  facet_wrap(~ label, scales = "free", ncol = 4) +
  labs(
    x = "Initial DBH (cm)", y = "\u0394D / yr  (cm yr\u207b\u00b9)",
    title = "Growth allometry for 8 most-sampled species"
  )
```

---

# Caveats

1. **OLS on log-transformed data.** OLS underestimates the allometric slope
   when measurement error is present in both variables.  SMA (standardised
   major axis) regression is preferred for formal allometric estimation.

2. **Positive-increment filter.** Log-log regressions require strictly positive
   $\Delta D$ and $\Delta M$.  Excluding the 20% negative-increment intervals
   introduces a selection bias toward faster-growing or more accurately measured
   trees.

3. **AGB is model-derived.** All AGB values come from Chojnacky et al. (2014)
   applied to DBH.  The $\Delta M \sim M^{\beta}$ relationship largely reflects
   the structure of the Chojnacky equations, not independent biomass measurements.

4. **Gap normalisation.** Growth is divided by the actual year gap between
   censuses (1–9 years), which is not uniform.  Long gaps average over
   interannual variability.

5. **Mixed species and biomes.** All sites are pooled for overall regressions.
   Allometric exponents differ among species, forest types, and disturbance
   histories.  Per-site and per-species analyses reveal this heterogeneity.

6. **Theory comparison.** MST predictions (WBE model, Enquist et al.) assume
   ideal self-similar vascular networks and no resource limitation other than
   light capture geometry.  Real forests deviate from this for many reasons
   (competition, disturbance, drought, ontogenetic stage).

---

# Session info

```{r session}
sessionInfo()
```
