---
title: "NEON ForestAGB — Tree Growth Rate Analysis"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5)
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(knitr)
  library(scales)
  library(maps)
})
theme_set(theme_bw(base_size = 13))

proj_dir <- dirname(knitr::current_input(dir = TRUE))

# ── Load all NEON site files ─────────────────────────────────────────────────
neon_sites <- c("BART","BLAN","CLBJ","DELA","DSNY","GRSM","GUAN","HARV",
                "JERC","KONZ","LENO","MLBS","ORNL","OSBS","RMNP","SCBI",
                "SERC","STEI","TALL","TREE","UKFS","UNDE","YELL")

raw <- lapply(neon_sites, function(s) {
  read_csv(
    file.path(proj_dir, "data/raw/neon_forest_agb/biomass",
              paste0(s, "_biomass.csv")),
    show_col_types = FALSE
  )
}) |> bind_rows()

# ── Harmonise: all "Live*" statuses, one row per individual × year ───────────
# Some individuals appear more than once in a year (different stems / flags).
# We take the mean DBH and AGB per individual per year for those cases.
live_raw <- raw |>
  filter(grepl("^Live", plantStatus, ignore.case = TRUE),
         !is.na(stemDiameter), stemDiameter > 0) |>
  group_by(individualID, siteID, domainID, scientificName, Family,
           growthForm, year) |>
  summarise(
    DBH_cm       = mean(stemDiameter,    na.rm = TRUE),
    Height_m     = mean(height,          na.rm = TRUE),
    AGB_Choj_kg  = mean(AGBChojnacky,    na.rm = TRUE),
    AGB_Jenk_kg  = mean(AGBJenkins,      na.rm = TRUE),
    .groups = "drop"
  )

# ── Annual growth increments (consecutive year pairs per individual) ──────────
annual_growth <- live_raw |>
  arrange(individualID, year) |>
  group_by(individualID, siteID, domainID, scientificName, Family, growthForm) |>
  mutate(
    delta_yr     = year      - lag(year),
    delta_DBH    = DBH_cm    - lag(DBH_cm),
    delta_AGB    = AGB_Choj_kg - lag(AGB_Choj_kg),
    delta_Ht     = Height_m  - lag(Height_m),
    DBH_start    = lag(DBH_cm),
    AGB_start    = lag(AGB_Choj_kg),
    # annualised rates (divide by actual year gap — not always 1)
    DBH_inc_cm_yr  = delta_DBH / delta_yr,
    AGB_inc_kg_yr  = delta_AGB / delta_yr,
    Ht_inc_m_yr    = delta_Ht  / delta_yr,
    # relative growth rate (RGR) — normalised by size at start
    RGR_DBH        = (log(DBH_cm) - log(DBH_start)) / delta_yr,
    RGR_AGB        = (log(AGB_Choj_kg) - log(AGB_start)) / delta_yr
  ) |>
  filter(!is.na(delta_yr), delta_yr >= 1) |>
  ungroup()

# ── Per-individual lifetime summaries ────────────────────────────────────────
lifetime <- live_raw |>
  group_by(individualID, siteID, domainID, scientificName, Family, growthForm) |>
  filter(n() >= 2) |>
  summarise(
    n_census           = n(),
    year_first         = min(year),
    year_last          = max(year),
    duration_yr        = max(year) - min(year),
    DBH_first_cm       = DBH_cm[which.min(year)],
    DBH_last_cm        = DBH_cm[which.max(year)],
    AGB_first_kg       = AGB_Choj_kg[which.min(year)],
    AGB_last_kg        = AGB_Choj_kg[which.max(year)],
    total_DBH_inc_cm   = DBH_last_cm  - DBH_first_cm,
    total_AGB_inc_kg   = AGB_last_kg  - AGB_first_kg,
    mean_annual_DBH_inc = total_DBH_inc_cm / duration_yr,
    mean_annual_AGB_inc = total_AGB_inc_kg  / duration_yr,
    RGR_DBH_lifetime   = (log(DBH_last_cm)  - log(DBH_first_cm))  / duration_yr,
    RGR_AGB_lifetime   = (log(AGB_last_kg)  - log(AGB_first_kg))  / duration_yr,
    .groups = "drop"
  )
```

---

# Overview

This document derives individual-tree growth rates from the **NEON ForestAGB**
longitudinal dataset.  The same tagged trees (`individualID`) are measured across
multiple years, enabling computation of:

- **Annual DBH increment** (cm yr⁻¹): diameter growth between consecutive censuses
- **Annual AGB increment** (kg yr⁻¹): biomass growth between consecutive censuses
- **Relative growth rate (RGR)**: proportional diameter or biomass change per year
- **Lifetime totals**: cumulative change from first to last live census per tree

> **AGB caveat:** Biomass is estimated from DBH via Chojnacky et al. (2014)
> allometric equations.  AGB increments are therefore *model-derived* and
> correlated with DBH increments — they do not represent independent
> measurements.

> **⚠ Growth rates can be negative** (measurement error, bark loss, or DBH
> measured at different heights across visits). Negative values are retained
> as data but flagged in relevant plots.

---

# Dataset summary

```{r summary-table}
kable(data.frame(
  Metric = c(
    "Total raw live-status rows",
    "Live individuals with >= 1 census",
    "Live individuals with >= 2 censuses (growth derivable)",
    "Annual growth intervals computed",
    "Year range",
    "Sites"
  ),
  Value = c(
    format(nrow(live_raw), big.mark = ","),
    format(length(unique(live_raw$individualID)), big.mark = ","),
    format(nrow(lifetime), big.mark = ","),
    format(nrow(annual_growth), big.mark = ","),
    paste0(min(live_raw$year, na.rm=TRUE), "\u2013", max(live_raw$year, na.rm=TRUE)),
    length(neon_sites)
  ),
  stringsAsFactors = FALSE
), caption = "NEON growth analysis — dataset scope")
```

---

# Census visit distribution

```{r visit-dist, fig.height=3.5}
lifetime |>
  count(n_census) |>
  ggplot(aes(x = factor(n_census), y = n)) +
  geom_col(fill = "#1a9850") +
  geom_text(aes(label = format(n, big.mark=",")), vjust = -0.4, size = 3.5) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Number of censuses per individual (live visits)",
       y = "N individuals",
       title = "Distribution of census visits per tree (live status only)")
```

```{r duration-dist, fig.height=4}
lifetime |>
  ggplot(aes(x = duration_yr)) +
  geom_histogram(bins = 30, fill = "#4575b4", color = "white", linewidth = 0.2) +
  scale_y_continuous(labels = comma) +
  labs(x = "Duration from first to last live census (years)",
       y = "N individuals",
       title = paste0("Observation duration per individual  (n = ",
                      format(nrow(lifetime), big.mark=","), ")"))
```

---

# Annual DBH increment

## Distribution

```{r dbh-inc-dist, fig.height=4.5}
annual_growth |>
  ggplot(aes(x = DBH_inc_cm_yr)) +
  geom_histogram(bins = 80, fill = "#1a9850", color = "white", linewidth = 0.15) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 0.8) +
  scale_y_continuous(labels = comma) +
  coord_cartesian(xlim = c(-3, 6)) +
  labs(
    x = "Annual DBH increment (cm yr\u207b\u00b9)",
    y = "N intervals",
    title = paste0("Annual DBH increment distribution  (n = ",
                   format(nrow(annual_growth), big.mark=","), " intervals)"),
    subtitle = paste0(
      "Median = ",
      round(median(annual_growth$DBH_inc_cm_yr, na.rm=TRUE), 3),
      " cm yr\u207b\u00b9   |   Mean = ",
      round(mean(annual_growth$DBH_inc_cm_yr, na.rm=TRUE), 3),
      " cm yr\u207b\u00b9   |   ",
      format(sum(annual_growth$DBH_inc_cm_yr < 0, na.rm=TRUE), big.mark=","),
      " intervals with negative increment (",
      round(100*mean(annual_growth$DBH_inc_cm_yr < 0, na.rm=TRUE), 1), "%)"
    )
  )
```

## Annual DBH increment vs initial DBH

```{r dbh-inc-vs-dbh, fig.height=5}
sub <- annual_growth |>
  filter(!is.na(DBH_start), DBH_start > 0,
         !is.na(DBH_inc_cm_yr),
         DBH_inc_cm_yr > -2, DBH_inc_cm_yr < 8) |>
  sample_n(min(20000, n()))

lm_dbhinc <- lm(DBH_inc_cm_yr ~ log(DBH_start), data = sub)
r2_dbhinc  <- round(summary(lm_dbhinc)$r.squared, 3)

ggplot(sub, aes(x = DBH_start, y = DBH_inc_cm_yr)) +
  geom_point(alpha = 0.05, size = 0.5, color = "#1a9850") +
  geom_smooth(method = "lm", formula = y ~ log(x),
              color = "#d73027", linewidth = 1.2, se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10(labels = comma) +
  labs(
    x = "Initial DBH at census start (cm, log scale)",
    y = "Annual DBH increment (cm yr\u207b\u00b9)",
    title = "Does growth rate depend on tree size?",
    subtitle = paste0("OLS fit (log DBH predictor)   R\u00b2 = ", r2_dbhinc,
                      "   n = 20,000 subsample")
  )
```

## Annual DBH increment by site

```{r dbh-inc-by-site, fig.height=6}
annual_growth |>
  filter(!is.na(DBH_inc_cm_yr)) |>
  group_by(siteID) |>
  mutate(med = median(DBH_inc_cm_yr, na.rm=TRUE)) |>
  ungroup() |>
  ggplot(aes(x = reorder(siteID, med), y = DBH_inc_cm_yr, fill = siteID)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, show.legend = FALSE,
               coef = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_flip(ylim = c(-1.5, 4)) +
  scale_fill_viridis_d(option = "D") +
  labs(
    x = NULL,
    y = "Annual DBH increment (cm yr\u207b\u00b9)",
    title = "Annual DBH increment distribution by NEON site",
    subtitle = "Boxes ordered by median increment; clipped to [\u22121.5, 4] cm yr\u207b\u00b9"
  )
```

## Annual DBH increment per-site statistics

```{r dbh-site-stats}
site_stats <- annual_growth |>
  filter(!is.na(DBH_inc_cm_yr)) |>
  group_by(siteID) |>
  summarise(
    N_intervals  = n(),
    Mean_cm_yr   = round(mean(DBH_inc_cm_yr, na.rm=TRUE), 3),
    Median_cm_yr = round(median(DBH_inc_cm_yr, na.rm=TRUE), 3),
    SD_cm_yr     = round(sd(DBH_inc_cm_yr, na.rm=TRUE), 3),
    P10_cm_yr    = round(quantile(DBH_inc_cm_yr, 0.10, na.rm=TRUE), 3),
    P90_cm_yr    = round(quantile(DBH_inc_cm_yr, 0.90, na.rm=TRUE), 3),
    Pct_negative = round(100*mean(DBH_inc_cm_yr < 0, na.rm=TRUE), 1),
    .groups = "drop"
  ) |>
  arrange(desc(Median_cm_yr))

kable(site_stats,
      col.names = c("Site","N intervals","Mean (cm/yr)","Median (cm/yr)",
                    "SD","P10","P90","% negative"),
      caption = "Annual DBH increment summary per NEON site")
```

---

# Annual AGB increment

> **Reminder:** AGB is model-derived from DBH — AGB increments are not
> independent of DBH increments.

## Distribution

```{r agb-inc-dist, fig.height=4.5}
annual_growth |>
  filter(!is.na(AGB_inc_kg_yr)) |>
  ggplot(aes(x = AGB_inc_kg_yr)) +
  geom_histogram(bins = 80, fill = "#8c510a", color = "white", linewidth = 0.15) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 0.8) +
  scale_y_continuous(labels = comma) +
  coord_cartesian(xlim = c(-50, 150)) +
  labs(
    x = "Annual AGB increment (kg yr\u207b\u00b9, Chojnacky)",
    y = "N intervals",
    title = "Annual AGB increment distribution",
    subtitle = paste0(
      "Median = ",
      round(median(annual_growth$AGB_inc_kg_yr, na.rm=TRUE), 2),
      " kg yr\u207b\u00b9   |   Mean = ",
      round(mean(annual_growth$AGB_inc_kg_yr, na.rm=TRUE), 2),
      " kg yr\u207b\u00b9"
    )
  )
```

## Annual AGB increment vs initial AGB

```{r agb-inc-vs-agb, fig.height=5}
sub_agb <- annual_growth |>
  filter(!is.na(AGB_start), AGB_start > 0,
         !is.na(AGB_inc_kg_yr), AGB_inc_kg_yr > -50, AGB_inc_kg_yr < 500) |>
  sample_n(min(20000, n()))

lm_agbinc <- lm(AGB_inc_kg_yr ~ log(AGB_start), data = sub_agb)
r2_agbinc  <- round(summary(lm_agbinc)$r.squared, 3)

ggplot(sub_agb, aes(x = AGB_start, y = AGB_inc_kg_yr)) +
  geom_point(alpha = 0.05, size = 0.5, color = "#8c510a") +
  geom_smooth(method = "lm", formula = y ~ log(x),
              color = "#4575b4", linewidth = 1.2, se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10(labels = comma) +
  labs(
    x = "Initial AGB (kg, log scale)",
    y = "Annual AGB increment (kg yr\u207b\u00b9)",
    title = "Annual AGB increment vs initial biomass",
    subtitle = paste0("OLS fit (log AGB predictor)   R\u00b2 = ", r2_agbinc)
  )
```

## Annual AGB increment by site

```{r agb-inc-by-site, fig.height=6}
annual_growth |>
  filter(!is.na(AGB_inc_kg_yr)) |>
  group_by(siteID) |>
  mutate(med = median(AGB_inc_kg_yr, na.rm=TRUE)) |>
  ungroup() |>
  ggplot(aes(x = reorder(siteID, med), y = AGB_inc_kg_yr, fill = siteID)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, show.legend = FALSE,
               coef = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_flip(ylim = c(-20, 80)) +
  scale_fill_viridis_d(option = "C") +
  labs(
    x = NULL,
    y = "Annual AGB increment (kg yr\u207b\u00b9)",
    title = "Annual AGB increment distribution by NEON site",
    subtitle = "Boxes ordered by median; clipped to [\u221220, 80] kg yr\u207b\u00b9"
  )
```

---

# Relative growth rate (RGR)

RGR normalises growth by initial size:
$$\text{RGR}_{\text{DBH}} = \frac{\ln(D_2) - \ln(D_1)}{t_2 - t_1} \quad (\text{yr}^{-1})$$

This removes the strong size-growth correlation and reveals proportional growth
rates.

## RGR-DBH distribution

```{r rgr-dist, fig.height=4.5}
annual_growth |>
  filter(!is.na(RGR_DBH), is.finite(RGR_DBH)) |>
  ggplot(aes(x = RGR_DBH)) +
  geom_histogram(bins = 80, fill = "#4575b4", color = "white", linewidth = 0.15) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  scale_y_continuous(labels = comma) +
  coord_cartesian(xlim = c(-0.15, 0.25)) +
  labs(
    x = "RGR-DBH (yr\u207b\u00b9)",
    y = "N intervals",
    title = "Relative growth rate — DBH (annual census intervals)",
    subtitle = paste0(
      "Median = ",
      round(median(annual_growth$RGR_DBH, na.rm=TRUE), 4), " yr\u207b\u00b9   |   ",
      "Mean = ",
      round(mean(annual_growth$RGR_DBH, na.rm=TRUE), 4), " yr\u207b\u00b9"
    )
  )
```

## RGR vs initial DBH

Metabolic scaling theory (Enquist et al.) predicts RGR ∝ DBH^(−1/3) for
size-symmetric light competition.

```{r rgr-vs-dbh, fig.height=5}
sub_rgr <- annual_growth |>
  filter(!is.na(RGR_DBH), is.finite(RGR_DBH),
         !is.na(DBH_start), DBH_start > 0,
         RGR_DBH > -0.15, RGR_DBH < 0.25) |>
  sample_n(min(20000, n()))

lm_rgr <- lm(RGR_DBH ~ log(DBH_start), data = sub_rgr)
beta_rgr <- round(coef(lm_rgr)[2], 4)
r2_rgr   <- round(summary(lm_rgr)$r.squared, 3)

ggplot(sub_rgr, aes(x = DBH_start, y = RGR_DBH)) +
  geom_point(alpha = 0.04, size = 0.5, color = "#4575b4") +
  geom_smooth(method = "lm", formula = y ~ log(x),
              color = "#d73027", linewidth = 1.2, se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10(labels = comma) +
  labs(
    x = "Initial DBH (cm, log scale)",
    y = "RGR-DBH (yr\u207b\u00b9)",
    title = "Relative growth rate vs tree size",
    subtitle = paste0(
      "OLS slope on log(DBH) = ", beta_rgr,
      "   R\u00b2 = ", r2_rgr,
      "   (MST prediction: slope \u2248 \u22121/3 \u2248 \u22120.33)"
    )
  )
```

## RGR by site

```{r rgr-by-site, fig.height=6}
annual_growth |>
  filter(!is.na(RGR_DBH), is.finite(RGR_DBH)) |>
  group_by(siteID) |>
  mutate(med = median(RGR_DBH, na.rm=TRUE)) |>
  ungroup() |>
  ggplot(aes(x = reorder(siteID, med), y = RGR_DBH, fill = siteID)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_flip(ylim = c(-0.1, 0.2)) +
  scale_fill_viridis_d(option = "E") +
  labs(
    x = NULL,
    y = "RGR-DBH (yr\u207b\u00b9)",
    title = "Relative DBH growth rate by NEON site"
  )
```

---

# Temporal trends in growth

## Mean annual DBH increment by year

```{r temporal-dbh, fig.height=4.5}
annual_growth |>
  group_by(siteID, year) |>
  summarise(
    mean_inc = mean(DBH_inc_cm_yr, na.rm=TRUE),
    se_inc   = sd(DBH_inc_cm_yr, na.rm=TRUE) / sqrt(n()),
    .groups  = "drop"
  ) |>
  ggplot(aes(x = year, y = mean_inc, color = siteID, group = siteID)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_viridis_d(option = "D") +
  labs(
    x = "Census year (end of interval)",
    y = "Mean annual DBH increment (cm yr\u207b\u00b9)",
    title = "Temporal trend in mean annual DBH increment per site",
    subtitle = "Each point = mean across all live trees measured in that interval at that site",
    color = "Site"
  ) +
  theme(legend.position = "right",
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"))
```

## All-site mean with trend

```{r temporal-all-sites, fig.height=4}
annual_growth |>
  group_by(year) |>
  summarise(
    mean_inc = mean(DBH_inc_cm_yr, na.rm=TRUE),
    se_inc   = sd(DBH_inc_cm_yr, na.rm=TRUE) / sqrt(n()),
    n        = n(),
    .groups  = "drop"
  ) |>
  ggplot(aes(x = year, y = mean_inc)) +
  geom_ribbon(aes(ymin = mean_inc - se_inc, ymax = mean_inc + se_inc),
              fill = "#1a9850", alpha = 0.25) +
  geom_line(color = "#1a9850", linewidth = 1.2) +
  geom_point(aes(size = n), color = "#1a9850") +
  geom_smooth(method = "lm", formula = y ~ x,
              color = "#d73027", linewidth = 1, linetype = "dashed", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_size_continuous(range = c(2, 6), labels = comma,
                        name = "N intervals") +
  labs(
    x = "Census year",
    y = "Mean annual DBH increment (cm yr\u207b\u00b9)",
    title = "All-site mean annual DBH increment over time",
    subtitle = "Ribbon = \u00b1 1 SE; dashed red line = OLS temporal trend; point size = N intervals"
  )
```

---

# Lifetime growth summaries

## Lifetime total DBH increment distribution

```{r lifetime-dbh-total, fig.height=4.5}
lifetime |>
  ggplot(aes(x = total_DBH_inc_cm)) +
  geom_histogram(bins = 70, fill = "#1a9850", color = "white", linewidth = 0.15) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  scale_y_continuous(labels = comma) +
  coord_cartesian(xlim = c(-10, 40)) +
  labs(
    x = "Total DBH increment over full observation period (cm)",
    y = "N individuals",
    title = paste0("Lifetime total DBH growth  (n = ",
                   format(nrow(lifetime), big.mark=","), " individuals)"),
    subtitle = paste0(
      "Median = ", round(median(lifetime$total_DBH_inc_cm, na.rm=TRUE), 2), " cm   |   ",
      "Mean = ", round(mean(lifetime$total_DBH_inc_cm, na.rm=TRUE), 2), " cm   |   ",
      format(sum(lifetime$total_DBH_inc_cm < 0, na.rm=TRUE), big.mark=","),
      " trees with net decline"
    )
  )
```

## Mean annual increment vs duration

```{r mean-annual-vs-duration, fig.height=5}
lifetime |>
  filter(duration_yr >= 1, !is.na(mean_annual_DBH_inc)) |>
  ggplot(aes(x = duration_yr, y = mean_annual_DBH_inc, color = siteID)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_smooth(aes(group = 1), method = "lm", formula = y ~ x,
              color = "black", linewidth = 1.2, se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_viridis_d(option = "D", guide = "none") +
  labs(
    x = "Observation duration (years)",
    y = "Mean annual DBH increment (cm yr\u207b\u00b9)",
    title = "Does estimated growth rate depend on observation window length?",
    subtitle = "Systematic trends indicate regression-to-mean or temporal autocorrelation effects"
  )
```

## Lifetime mean annual DBH increment by site

```{r lifetime-by-site, fig.height=6}
lifetime |>
  filter(!is.na(mean_annual_DBH_inc)) |>
  group_by(siteID) |>
  mutate(med = median(mean_annual_DBH_inc, na.rm=TRUE)) |>
  ungroup() |>
  ggplot(aes(x = reorder(siteID, med), y = mean_annual_DBH_inc, fill = siteID)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  coord_flip(ylim = c(-1, 3)) +
  scale_fill_viridis_d(option = "D") +
  labs(
    x = NULL,
    y = "Mean annual DBH increment (cm yr\u207b\u00b9)",
    title = "Lifetime mean annual DBH increment by NEON site"
  )
```

## Lifetime total AGB increment by site

```{r lifetime-agb-by-site, fig.height=6}
lifetime |>
  filter(!is.na(total_AGB_inc_kg)) |>
  group_by(siteID) |>
  summarise(
    N            = n(),
    median_total = round(median(total_AGB_inc_kg, na.rm=TRUE), 1),
    mean_total   = round(mean(total_AGB_inc_kg, na.rm=TRUE), 1),
    .groups      = "drop"
  ) |>
  ggplot(aes(x = reorder(siteID, median_total), y = median_total, fill = siteID)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(median_total, " kg")),
            hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.25))) +
  scale_fill_viridis_d(option = "C") +
  labs(
    x = NULL,
    y = "Median lifetime total AGB increment (kg)",
    title = "Median lifetime AGB gained per individual by NEON site",
    subtitle = "AGB Chojnacky (model-derived from DBH)"
  )
```

---

# Growth rates by species

## Top 15 fastest-growing species (median annual DBH increment)

```{r top-species-growth, fig.height=5.5}
sp_growth <- annual_growth |>
  filter(!is.na(DBH_inc_cm_yr), !is.na(scientificName)) |>
  group_by(scientificName) |>
  filter(n() >= 50) |>
  summarise(
    N_intervals  = n(),
    median_inc   = round(median(DBH_inc_cm_yr, na.rm=TRUE), 3),
    mean_inc     = round(mean(DBH_inc_cm_yr, na.rm=TRUE), 3),
    median_RGR   = round(median(RGR_DBH, na.rm=TRUE), 4),
    .groups = "drop"
  ) |>
  arrange(desc(median_inc))

sp_growth |>
  head(15) |>
  ggplot(aes(x = reorder(scientificName, median_inc), y = median_inc)) +
  geom_col(fill = "#74c476") +
  geom_text(aes(label = paste0(median_inc, " cm/yr")), hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(
    x = NULL,
    y = "Median annual DBH increment (cm yr\u207b\u00b9)",
    title = "Top 15 fastest-growing species (median annual DBH increment)",
    subtitle = "Species with \u2265 50 growth intervals across all NEON sites"
  )
```

## Top 15 fastest-growing species (median RGR-DBH)

```{r top-species-rgr, fig.height=5.5}
sp_growth |>
  arrange(desc(median_RGR)) |>
  head(15) |>
  ggplot(aes(x = reorder(scientificName, median_RGR), y = median_RGR)) +
  geom_col(fill = "#9ecae1") +
  geom_text(aes(label = sprintf("%.4f", median_RGR)), hjust = -0.05, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(
    x = NULL,
    y = "Median RGR-DBH (yr\u207b\u00b9)",
    title = "Top 15 fastest-growing species by relative growth rate",
    subtitle = "Species with \u2265 50 growth intervals"
  )
```

## Full species growth table

```{r species-growth-table}
kable(sp_growth |> head(40),
      col.names = c("Species","N intervals","Median DBH inc (cm/yr)",
                    "Mean DBH inc (cm/yr)","Median RGR (yr\u207b\u00b9)"),
      caption   = "Top 40 species by median annual DBH increment (>= 50 intervals)")
```

---

# Growth rate vs canopy position

```{r canopy-growth, fig.height=4.5}
raw_canopy <- raw |>
  filter(grepl("^Live", plantStatus, ignore.case=TRUE),
         !is.na(canopyPosition), !is.na(stemDiameter), stemDiameter > 0) |>
  group_by(individualID, year, canopyPosition) |>
  summarise(DBH_cm = mean(stemDiameter, na.rm=TRUE), .groups="drop")

# Join canopy position to annual growth
annual_growth_cp <- annual_growth |>
  left_join(
    raw_canopy |> rename(year_end = year),
    by = c("individualID", "year" = "year_end")
  )

annual_growth_cp |>
  filter(!is.na(canopyPosition), !is.na(DBH_inc_cm_yr)) |>
  group_by(canopyPosition) |>
  mutate(med = median(DBH_inc_cm_yr, na.rm=TRUE)) |>
  ungroup() |>
  ggplot(aes(x = reorder(canopyPosition, med), y = DBH_inc_cm_yr,
             fill = canopyPosition)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip(ylim = c(-1, 4)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Annual DBH increment (cm yr\u207b\u00b9)",
    title = "Annual DBH increment by canopy position",
    subtitle = "Clipped to [\u22121, 4] cm yr\u207b\u00b9"
  )
```

---

# Lifetime summary table by site

```{r lifetime-site-table}
lt_site <- lifetime |>
  group_by(siteID) |>
  summarise(
    N_trees             = n(),
    Median_dur_yr       = round(median(duration_yr), 1),
    Median_total_DBH_cm = round(median(total_DBH_inc_cm, na.rm=TRUE), 2),
    Mean_ann_DBH_cm_yr  = round(mean(mean_annual_DBH_inc, na.rm=TRUE), 3),
    Median_ann_DBH_cm_yr = round(median(mean_annual_DBH_inc, na.rm=TRUE), 3),
    Median_total_AGB_kg = round(median(total_AGB_inc_kg, na.rm=TRUE), 1),
    Pct_net_decline     = round(100*mean(total_DBH_inc_cm < 0, na.rm=TRUE), 1),
    .groups = "drop"
  ) |>
  arrange(desc(Median_ann_DBH_cm_yr))

kable(lt_site,
      col.names = c("Site","N trees","Median obs. duration (yr)",
                    "Median total DBH inc (cm)","Mean annual DBH inc (cm/yr)",
                    "Median annual DBH inc (cm/yr)",
                    "Median total AGB inc (kg)","% trees with net decline"),
      caption   = "Per-site lifetime growth summary (live trees with >= 2 censuses)")
```

---

# Caveats and interpretation notes

1. **AGB increments are model-derived.** All AGB values are estimated from DBH
   via Chojnacky et al. (2014). AGB growth rates are therefore algebraically
   related to DBH growth rates and do not represent independent measurements.

2. **Negative increments** reflect measurement error, bark sloughing, changes
   in measurement height across visits, or recorder differences — not genuine
   shrinkage. Trees with large negative increments should be examined for flag
   columns before including in formal analyses.

3. **Year gap normalisation.** NEON plots are not revisited every year uniformly.
   Growth is normalised by the actual year gap between consecutive censuses.
   Intervals of >2 years (common) average over interannual variation.

4. **Multiple live-status categories.** This analysis includes all `plantStatus`
   values beginning with "Live" (Live, Live broken bole, Live insect damaged,
   etc.) to maximise sample size. Restricting to "Live" only changes results
   quantitatively but not qualitatively.

5. **No dead-to-live transitions.** Each growth interval requires both years to
   carry a "Live\*" status. Trees that died between censuses are excluded from
   all growth calculations.

6. **RGR interpretation.** RGR-DBH measures proportional diameter growth.
   It is not equivalent to basal area increment (BAI) or volumetric growth,
   which are better proxies for carbon accumulation.

7. **OLS vs SMA.** All regressions here are OLS for rapid visualisation. SMA
   (standardised major axis) is appropriate for formal allometric inference
   when both variables carry error.

---

# Session info

```{r session}
sessionInfo()
```
