---
title: "Niklas & Enquist Biomass Database — Detailed Summary"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)
library(stringr)

theme_set(theme_bw(base_size = 12))

# ── Load and clean ────────────────────────────────────────────────────────────
raw_path <- file.path(dirname(knitr::current_input(dir = TRUE)),
                      "data/raw/niklas_enquist/niklas_enquist_biomass_20040122.csv")

dat <- read_csv(raw_path, skip = 2,
  col_names = c(
    "taxa", "citation", "footnotes", "record_num",
    "age_yr", "log_age", "height_m", "log_height",
    "root_biomass_kg", "log_root", "stem_biomass_kg", "log_stem",
    "leaf_biomass_kg", "log_leaf", "shoot_biomass_kg", "log_shoot",
    "repro_biomass_kg", "log_repro", "total_biomass_kg", "log_total",
    "stem_growth_kgyr", "log_stem_growth", "leaf_growth_kgyr", "log_leaf_growth",
    "root_growth_kgyr", "log_root_growth", "total_growth_kgyr", "log_total_growth"
  ),
  locale = locale(encoding = "latin1"),
  show_col_types = FALSE, skip_empty_rows = TRUE) |>
  mutate(across(-c(taxa, citation, footnotes), as.numeric)) |>
  mutate(across(where(is.numeric), ~ ifelse(. == -999, NA_real_, .)))
```

---

# Overview

The **Niklas & Enquist biomass dataset** is a cross-species compilation of
plant biomass, height, age, and growth-rate measurements drawn from the primary
literature, originally assembled by Enquist and Niklas as the empirical backbone
for their metabolic scaling theory papers published in *Nature* and related journals
(2001–2004). The primary published data compilation is described in:

> Niklas KJ & Enquist BJ (2001). Invariant scaling relationships for interspecific
> plant biomass production rates and body size. *PNAS*, 98(5): 2922–2927.
> DOI: [10.1073/pnas.041590298](https://doi.org/10.1073/pnas.041590298)

> Enquist BJ & Niklas KJ (2002). Global allocation rules for patterns of
> biomass partitioning in seed plants. *Science*, 295: 1517–1520.
> DOI: [10.1126/science.1066360](https://doi.org/10.1126/science.1066360)

The source CSV (`niklas_enquist_biomass_20040122.csv`) was compiled January 2004
and sourced from Cannell (1982) and other primary literature. Missing values are
encoded as −999 in the original file and have been recoded to NA here.

> **⚠ Citation note:** The citations above should be independently verified
> before use in any manuscript. DOIs have not been checked against Crossref
> in this exploratory summary.

```{r top-counts}
n_rows  <- nrow(dat)
n_taxa  <- length(unique(dat$taxa))
n_cites <- length(unique(dat$citation))

kable(data.frame(
  Metric  = c("Total records", "Unique taxa strings", "Unique citations"),
  Value   = format(c(n_rows, n_taxa, n_cites), big.mark = ",")
), caption = "Top-level database summary")
```

---

# Data structure

## Column dictionary

```{r dict-table}
col_dict <- data.frame(
  Column            = c("taxa","citation","footnotes","record_num",
                        "age_yr","height_m",
                        "root_biomass_kg","stem_biomass_kg","leaf_biomass_kg",
                        "shoot_biomass_kg","repro_biomass_kg","total_biomass_kg",
                        "stem_growth_kgyr","leaf_growth_kgyr",
                        "root_growth_kgyr","total_growth_kgyr"),
  Unit              = c("—","—","—","—",
                        "yr","m",
                        "kg dry wt","kg dry wt","kg dry wt",
                        "kg dry wt","kg dry wt","kg dry wt",
                        "kg yr⁻¹","kg yr⁻¹","kg yr⁻¹","kg yr⁻¹"),
  Description       = c("Taxon name (may include mixed stands)",
                        "Literature citation string",
                        "Footnotes from Cannell (1982) pages",
                        "Original sort order from source table",
                        "Plant age","Plant height",
                        "Root dry biomass per plant",
                        "Stem dry biomass per plant",
                        "Leaf dry biomass per plant",
                        "Shoot (stem+leaf) dry biomass per plant",
                        "Reproductive organ dry biomass",
                        "Total dry biomass (root+stem+leaf)",
                        "Annual stem dry mass increment",
                        "Annual leaf dry mass increment",
                        "Annual root dry mass increment",
                        "Annual total dry mass increment")
)
kable(col_dict, caption = "Column dictionary for Niklas-Enquist dataset")
```

## Data availability per column

```{r availability, fig.height=6}
num_cols <- c("age_yr","height_m","root_biomass_kg","stem_biomass_kg",
              "leaf_biomass_kg","shoot_biomass_kg","repro_biomass_kg",
              "total_biomass_kg","stem_growth_kgyr","leaf_growth_kgyr",
              "root_growth_kgyr","total_growth_kgyr")

avail <- dat |>
  summarise(across(all_of(num_cols), ~ sum(!is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n") |>
  arrange(desc(n))

ggplot(avail, aes(x = reorder(variable, n), y = n)) +
  geom_col(fill = "#2166ac") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Non-missing records",
       title = "Data availability per column",
       subtitle = paste0("Total rows = ", format(n_rows, big.mark = ",")))
```

```{r range-table}
rng <- dat |>
  summarise(across(all_of(num_cols),
    list(n = ~ sum(!is.na(.)),
         min = ~ round(min(., na.rm = TRUE), 4),
         max = ~ round(max(., na.rm = TRUE), 2)),
    .names = "{.col}__{.fn}")) |>
  pivot_longer(everything(),
    names_to = c("variable", "stat"), names_sep = "__") |>
  pivot_wider(names_from = stat, values_from = value)

kable(rng, caption = "Non-missing counts and value ranges for numeric columns")
```

---

# Taxonomic coverage

## Top taxa by record count

```{r top-taxa}
taxa_counts <- dat |>
  filter(!is.na(taxa)) |>
  count(taxa, sort = TRUE)

kable(head(taxa_counts, 50),
      col.names = c("Taxon string", "N records"),
      caption = "Top 50 taxa by record count (note: includes mixed stands and crop species)")
```

## Taxa string composition

The dataset spans vascular plants including wild woody species, crops, and
annual herbs. Taxon strings are free-text from the original publications
and have not been harmonized to a taxonomic backbone in this exploratory summary.

```{r taxa-type-guess}
# Rough classification by keywords
dat2 <- dat |>
  filter(!is.na(taxa)) |>
  mutate(likely_type = case_when(
    str_detect(taxa, regex("Pinus|Picea|Abies|Larix|Tsuga|Pseudotsuga|Cryptomeria|Chamaecyparis|Cedrus|Taxus", ignore_case=TRUE)) ~ "Conifer",
    str_detect(taxa, regex("wheat|barley|maize|corn|rice|soya|soybean|pea|bean|millet|oat|rye|cotton|sunflower|tobacco|potato", ignore_case=TRUE)) ~ "Crop/herb",
    str_detect(taxa, regex("Quercus|Fagus|Betula|Populus|Acer|Fraxinus|Alnus|Eucalyptus|Acacia", ignore_case=TRUE)) ~ "Broadleaf tree",
    TRUE ~ "Other / unclassified"
  ))

dat2 |>
  count(likely_type, sort = TRUE) |>
  ggplot(aes(x = reorder(likely_type, n), y = n, fill = likely_type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Records",
       title = "Rough functional group classification (keyword-based)",
       subtitle = "Formal taxonomy reconciliation required before analysis")
```

---

# Age and size distributions

```{r age-dist}
dat |> filter(!is.na(age_yr), age_yr > 0) |>
  ggplot(aes(x = age_yr)) +
  geom_histogram(bins = 60, fill = "#4dac26", color = "white") +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +
  annotation_logticks(sides = "b") +
  labs(x = "Age (yr, log scale)", y = "Count",
       title = paste0("Age distribution  (n = ", sum(!is.na(dat$age_yr) & dat$age_yr > 0), ")"))
```

```{r height-dist}
dat |> filter(!is.na(height_m), height_m > 0) |>
  ggplot(aes(x = height_m)) +
  geom_histogram(bins = 60, fill = "#1b7837", color = "white") +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +
  annotation_logticks(sides = "b") +
  labs(x = "Height (m, log scale)", y = "Count",
       title = paste0("Height distribution  (n = ", sum(!is.na(dat$height_m) & dat$height_m > 0), ")"))
```

---

# Scaling relationships

All biomass variables are in **kg dry weight per plant**.
Height is in **m**. Age is in **yr**.
All regressions below are **exploratory OLS on log10-transformed data**.
Standardized Major Axis (SMA) regression using `smatr` is required for
unbiased allometric exponent estimation and will be applied in the formal
analysis phase.

## Total biomass vs. height

```{r total-height}
d1 <- dat |> filter(total_biomass_kg > 0, height_m > 0)
lm1 <- lm(log10(total_biomass_kg) ~ log10(height_m), data = d1)
b1 <- round(coef(lm1)[2], 3)
a1 <- round(10^coef(lm1)[1], 4)
n1 <- nrow(d1)

ggplot(d1, aes(x = height_m, y = total_biomass_kg)) +
  geom_point(alpha = 0.35, size = 1, color = "#1b7837") +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(x = "Height (m)", y = "Total biomass (kg)",
       title = paste0("Total biomass ~ height  |  β = ", b1, "  |  n = ", format(n1, big.mark=",")))
```

> OLS: total_biomass = `r a1` × height^`r b1`  (n = `r format(n1, big.mark=",")`).

## Stem biomass vs. total biomass

```{r stem-total}
d2 <- dat |> filter(stem_biomass_kg > 0, total_biomass_kg > 0)
lm2 <- lm(log10(stem_biomass_kg) ~ log10(total_biomass_kg), data = d2)
b2 <- round(coef(lm2)[2], 3); n2 <- nrow(d2)

ggplot(d2, aes(x = total_biomass_kg, y = stem_biomass_kg)) +
  geom_point(alpha = 0.35, size = 1, color = "#2166ac") +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(x = "Total biomass (kg)", y = "Stem biomass (kg)",
       title = paste0("Stem ~ total biomass  |  β = ", b2, "  |  n = ", format(n2, big.mark=",")))
```

## Leaf biomass vs. total biomass

```{r leaf-total}
d3 <- dat |> filter(leaf_biomass_kg > 0, total_biomass_kg > 0)
lm3 <- lm(log10(leaf_biomass_kg) ~ log10(total_biomass_kg), data = d3)
b3 <- round(coef(lm3)[2], 3); n3 <- nrow(d3)

ggplot(d3, aes(x = total_biomass_kg, y = leaf_biomass_kg)) +
  geom_point(alpha = 0.35, size = 1, color = "#d6604d") +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(x = "Total biomass (kg)", y = "Leaf biomass (kg)",
       title = paste0("Leaf ~ total biomass  |  β = ", b3, "  |  n = ", format(n3, big.mark=",")))
```

## Root biomass vs. total biomass

```{r root-total}
d4 <- dat |> filter(root_biomass_kg > 0, total_biomass_kg > 0)
lm4 <- lm(log10(root_biomass_kg) ~ log10(total_biomass_kg), data = d4)
b4 <- round(coef(lm4)[2], 3); n4 <- nrow(d4)

ggplot(d4, aes(x = total_biomass_kg, y = root_biomass_kg)) +
  geom_point(alpha = 0.35, size = 1, color = "#762a83") +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(x = "Total biomass (kg)", y = "Root biomass (kg)",
       title = paste0("Root ~ total biomass  |  β = ", b4, "  |  n = ", format(n4, big.mark=",")))
```

## Total growth rate vs. total biomass

```{r growth-total}
d5 <- dat |> filter(total_growth_kgyr > 0, total_biomass_kg > 0)
lm5 <- lm(log10(total_growth_kgyr) ~ log10(total_biomass_kg), data = d5)
b5 <- round(coef(lm5)[2], 3); n5 <- nrow(d5)

ggplot(d5, aes(x = total_biomass_kg, y = total_growth_kgyr)) +
  geom_point(alpha = 0.35, size = 1, color = "#f4a582") +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(x = "Total biomass (kg)", y = "Total growth rate (kg yr⁻¹)",
       title = paste0("Growth rate ~ total biomass  |  β = ", b5, "  |  n = ", format(n5, big.mark=",")))
```

> The WBE metabolic scaling prediction is β ≈ 0.75 for production ∝ biomass.

## Summary of scaling exponents (exploratory OLS)

```{r exponent-table}
exponents <- data.frame(
  Relationship      = c("total_biomass ~ height",
                        "stem_biomass ~ total_biomass",
                        "leaf_biomass ~ total_biomass",
                        "root_biomass ~ total_biomass",
                        "total_growth ~ total_biomass"),
  OLS_beta          = c(b1, b2, b3, b4, b5),
  n                 = c(n1, n2, n3, n4, n5)
)
kable(exponents, caption = "Exploratory OLS scaling exponents (log10-log10). Use SMA for formal inference.")
```

---

# Source citations

```{r citations-table}
cites <- dat |>
  filter(!is.na(citation)) |>
  mutate(citation = iconv(citation, to = "UTF-8", sub = "?"),
         citation = substr(citation, 1, 200)) |>
  count(citation, sort = TRUE)

kable(cites, col.names = c("Citation string", "N records"), escape = TRUE,
      caption = paste0("All ", nrow(cites), " source citation strings in the dataset"))
```

---

# Key caveats

1. **Taxon strings are unharmonized.** Many entries are free-text names from
   Cannell (1982) and include mixed-stand descriptions, common names, and
   synonyms. Taxonomy reconciliation against a backbone (TPL, TNRS, or GBIF)
   is required before cross-dataset merging.

2. **Missing data sentinel.** The original file uses −999 for missing values.
   These have been recoded to NA here; downstream scripts must not treat −999
   as a valid observation.

3. **Crops and annuals.** The dataset includes crop species (pearl millet,
   wheat, pea, soybean) alongside forest trees. Analyses of interspecific
   woody-plant scaling should filter on functional group or growth form.

4. **Shoot vs. total biomass.** `shoot_biomass_kg` = stem + leaf (above-ground
   except reproduction). `total_biomass_kg` = root + stem + leaf. These must
   not be used interchangeably.

5. **Log columns.** Pre-computed log10 columns are present in the source file
   but are derived quantities. The raw biomass columns should be used for
   independent analysis.

6. **OLS bias.** All exponents reported here are OLS on log-transformed data.
   These underestimate true allometric exponents when both axes have error.
   Use SMA (`smatr::sma()`) for publication-quality inference.

7. **Citation verification required.** The Niklas & Enquist (2001) and
   Enquist & Niklas (2002) references listed in the Overview need independent
   DOI and metadata verification before inclusion in any methods narrative.

---

# Session info

```{r session}
sessionInfo()
```
