---
title: "Niklas & Enquist Organ-Mass Scaling: Theory Testing Across Biomass Partitions"
subtitle: "OLS, SMA, PGLS, PGLMM — Interspecific, Gymnosperm/Angiosperm, Family, Intraspecific, and Bayesian Variance Partitioning"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(readr);  library(dplyr);  library(ggplot2); library(scales)
  library(tidyr);  library(knitr);  library(smatr);   library(purrr)
  library(stringr); library(tibble); library(caper);  library(ape)
  library(phytools)
})
if (requireNamespace("brms", quietly = TRUE)) suppressPackageStartupMessages(library(brms))

HAS_PHYLOMAKER <- requireNamespace("V.PhyloMaker2", quietly = TRUE)

# Explicit namespace bindings to prevent masking conflicts
map       <- purrr::map
map_dbl   <- purrr::map_dbl
map_df    <- purrr::map_df
map_dfr   <- purrr::map_dfr
select    <- dplyr::select
filter    <- dplyr::filter
mutate    <- dplyr::mutate
rename    <- dplyr::rename
arrange   <- dplyr::arrange
group_by  <- dplyr::group_by
summarise <- dplyr::summarise
summarize <- dplyr::summarize
slice     <- dplyr::slice

set.seed(42)
theme_set(theme_bw(base_size = 12))

proj_dir <- tryCatch(dirname(knitr::current_input(dir = TRUE)),
                     error = function(e) getwd())

# WBE / theory slope priors
# West, Brown & Enquist (1997) Science 276:122; (1999) Science 284:1677
# Enquist & Niklas (2002) Nature 410:655–660
WBE_LEAF_TOTAL  <- 3/4   # leaf ~ total metabolic scaling
WBE_LEAF_STEM   <- 3/4   # leaf ~ stem (KEY WBE prediction)
WBE_LEAF_ROOT   <- 3/4
WBE_ROOT_TOTAL  <- 1.0   # isometric (conservation of structure)
WBE_STEM_TOTAL  <- 1.0
WBE_SHOOT_TOTAL <- 1.0
WBE_ROOT_STEM   <- 1.0
WBE_ROOT_SHOOT  <- 1.0
WBE_TOTAL_H     <- 8/3   # total biomass ~ height^(8/3): derived from vascular geometry
WBE_GROWTH      <- 3/4   # growth rate ~ mass^(3/4)

# Kerkhoff et al. (2005) GEB: metabolic-stoichiometric slope predictions
# [CITATION-UNVERIFIED] Kerkhoff, A.J., Enquist, B.J., Elser, J.J. & Fagan, W.F. (2005)
#   Plant allometry, stoichiometry and the temperature-dependence of primary productivity.
#   Global Ecology and Biogeography, 14, 585-598.
# Extends WBE to stoichiometry: metabolic demand constrains leaf tissue allocation;
# leaf biomass and growth rate (NPP proxy) both predicted at M^(3/4).
KERKHOFF_LEAF_TOTAL   <- 3/4   # leaf ~ total (stoichiometric mechanism ≡ WBE)
KERKHOFF_GROWTH_TOTAL <- 3/4   # growth ~ total (NPP proxy, metabolic prediction)
# Related:
# [CITATION-UNVERIFIED] Brown et al. (2004) Ecology 85:1771 — Metabolic Theory of Ecology
# [CITATION-UNVERIFIED] Kerkhoff & Enquist (2006) Ecol Lett 9:419 — Ecosystem allometry
# [CITATION-UNVERIFIED] Elser et al. (2000) Nature 408:578 — Stoichiometric constraints
```

---

# Data Loading {#sec-data}

```{r load-data}
nik_col_names <- c(
  "taxa", "citation", "footnotes", "record_num", "age_yr", "log_age",
  "height_m", "log_height", "root_biomass_kg", "log_root", "stem_biomass_kg",
  "log_stem", "leaf_biomass_kg", "log_leaf", "shoot_biomass_kg", "log_shoot",
  "repro_biomass_kg", "log_repro", "total_biomass_kg", "log_total",
  "stem_growth_kgyr", "log_stem_growth", "leaf_growth_kgyr", "log_leaf_growth",
  "root_growth_kgyr", "log_root_growth", "total_growth_kgyr", "log_total_growth"
)

nik_enq <- read_csv(
  file.path(proj_dir, "data/raw/niklas_enquist/niklas_enquist_biomass_20040122.csv"),
  skip = 2, col_names = nik_col_names,
  locale = locale(encoding = "latin1"), show_col_types = FALSE) |>
  mutate(across(-c(taxa, citation, footnotes), as.numeric)) |>
  mutate(across(where(is.numeric), ~ ifelse(. == -999, NA_real_, .)))

cat(sprintf("Total records loaded: %d\n", nrow(nik_enq)))
```

---

# Record Classification {#sec-classify}

```{r classify-records}
# Crop/cultivar genus/species detection (annual herbaceous crops)
crop_pattern <- paste0(
  "(?i)(wheat|triticum|barley|hordeum|maize|zea mays|corn|oryza|rice|",
  "pea\\b|pisum|bean\\b|soybean|glycine max|phaseolus|cotton|gossypium|",
  "sunflower|helianthus annuus|millet|pearl millet|pennisetum glaucum|",
  "capsella|cucumber|cucumis|dolichos|lablab|vigna|brassica oleracea|",
  "lolium|medicago sativa|vicia faba)")

community_pattern <- paste0(
  "(?i)(community|stand|mixed|plot|forest|woodland|grassland|",
  "whole-plant community|with a few|average)")

nik_enq <- nik_enq |>
  mutate(
    flag_crop      = str_detect(taxa, crop_pattern),
    flag_community = str_detect(taxa, community_pattern),
    pgls_eligible  = !flag_community & !is.na(taxa)
  )

# Taxonomy-based record_type (assigned after parsing; prelim based on flags)
nik_enq <- nik_enq |>
  mutate(record_type_prelim = case_when(
    flag_crop      ~ "Crop/herb",
    flag_community ~ "Community/mixed",
    TRUE           ~ "Unknown — see taxonomy"
  ))

cat(sprintf("Crop records: %d\nCommunity/stand records: %d\nRemaining: %d\n",
            sum(nik_enq$flag_crop),
            sum(nik_enq$flag_community),
            sum(!nik_enq$flag_crop & !nik_enq$flag_community)))
```

---

# Data Overview {#sec-overview}

```{r overview-coverage}
# Coverage of each measured variable
organ_vars <- c("root_biomass_kg", "stem_biomass_kg", "leaf_biomass_kg",
                "shoot_biomass_kg", "repro_biomass_kg", "total_biomass_kg",
                "height_m", "age_yr",
                "stem_growth_kgyr", "leaf_growth_kgyr",
                "root_growth_kgyr", "total_growth_kgyr")

coverage_tbl <- tibble(
  variable = organ_vars,
  n_obs    = map_int(organ_vars, ~ sum(!is.na(nik_enq[[.x]]))),
  pct_obs  = map_dbl(organ_vars, ~ 100 * mean(!is.na(nik_enq[[.x]])))
)

kable(coverage_tbl, digits = 1,
      caption = "Variable coverage across all 2,626 records (N obs and %).")
```

```{r overview-type-chart, fig.height=4}
type_counts <- nik_enq |>
  count(record_type_prelim, name = "n") |>
  mutate(record_type_prelim = reorder(record_type_prelim, n))

ggplot(type_counts, aes(record_type_prelim, n, fill = record_type_prelim)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Preliminary record classification",
       subtitle = "Woody wild species will be confirmed after taxonomy parsing",
       x = NULL, y = "N records")
```

---

# Taxonomy {#sec-taxonomy}

```{r ne-genus-lookup}
ne_genus_lookup <- tibble::tribble(
  ~genus,           ~ne_family,            ~ne_division,
  "Abies",          "Pinaceae",            "Gymnosperm",
  "Picea",          "Pinaceae",            "Gymnosperm",
  "Pinus",          "Pinaceae",            "Gymnosperm",
  "Pseudotsuga",    "Pinaceae",            "Gymnosperm",
  "Larix",          "Pinaceae",            "Gymnosperm",
  "Tsuga",          "Pinaceae",            "Gymnosperm",
  "Cedrus",         "Pinaceae",            "Gymnosperm",
  "Juniperus",      "Cupressaceae",        "Gymnosperm",
  "Thuja",          "Cupressaceae",        "Gymnosperm",
  "Chamaecyparis",  "Cupressaceae",        "Gymnosperm",
  "Cryptomeria",    "Cupressaceae",        "Gymnosperm",
  "Podocarpus",     "Podocarpaceae",       "Gymnosperm",
  "Araucaria",      "Araucariaceae",       "Gymnosperm",
  "Taxus",          "Taxaceae",            "Gymnosperm",
  "Ginkgo",         "Ginkgoaceae",         "Gymnosperm",
  "Sequoia",        "Cupressaceae",        "Gymnosperm",
  "Sequoiadendron", "Cupressaceae",        "Gymnosperm",
  "Metasequoia",    "Cupressaceae",        "Gymnosperm",
  "Cupressus",      "Cupressaceae",        "Gymnosperm",
  "Quercus",        "Fagaceae",            "Angiosperm",
  "Fagus",          "Fagaceae",            "Angiosperm",
  "Nothofagus",     "Nothofagaceae",       "Angiosperm",
  "Betula",         "Betulaceae",          "Angiosperm",
  "Alnus",          "Betulaceae",          "Angiosperm",
  "Populus",        "Salicaceae",          "Angiosperm",
  "Salix",          "Salicaceae",          "Angiosperm",
  "Eucalyptus",     "Myrtaceae",           "Angiosperm",
  "Acacia",         "Fabaceae",            "Angiosperm",
  "Robinia",        "Fabaceae",            "Angiosperm",
  "Tilia",          "Malvaceae",           "Angiosperm",
  "Fraxinus",       "Oleaceae",            "Angiosperm",
  "Acer",           "Sapindaceae",         "Angiosperm",
  "Juglans",        "Juglandaceae",        "Angiosperm",
  "Carya",          "Juglandaceae",        "Angiosperm",
  "Magnolia",       "Magnoliaceae",        "Angiosperm",
  "Shorea",         "Dipterocarpaceae",    "Angiosperm",
  "Rhizophora",     "Rhizophoraceae",      "Angiosperm",
  "Prunus",         "Rosaceae",            "Angiosperm",
  "Cornus",         "Cornaceae",           "Angiosperm",
  "Liriodendron",   "Magnoliaceae",        "Angiosperm",
  "Castanea",       "Fagaceae",            "Angiosperm",
  "Corylus",        "Betulaceae",          "Angiosperm",
  "Carpinus",       "Betulaceae",          "Angiosperm",
  "Ulmus",          "Ulmaceae",            "Angiosperm",
  "Celtis",         "Cannabaceae",         "Angiosperm",
  "Liquidambar",    "Altingiaceae",        "Angiosperm",
  "Platanus",       "Platanaceae",         "Angiosperm",
  "Nyssa",          "Nyssaceae",           "Angiosperm",
  "Betula",         "Betulaceae",          "Angiosperm",
  "Cinnamomum",     "Lauraceae",           "Angiosperm",
  "Swietenia",      "Meliaceae",           "Angiosperm",
  "Tectona",        "Lamiaceae",           "Angiosperm",
  "Dipterocarpus",  "Dipterocarpaceae",    "Angiosperm",
  "Hopea",          "Dipterocarpaceae",    "Angiosperm",
  "Cedrela",        "Meliaceae",           "Angiosperm",
  "Terminalia",     "Combretaceae",        "Angiosperm",
  "Virola",         "Myristicaceae",       "Angiosperm",
  "Heliocarpus",    "Malvaceae",           "Angiosperm"
) |> dplyr::distinct(genus, .keep_all = TRUE)
```

```{r ne-parse-taxa}
ne_parse_taxa <- function(taxa_vec) {
  cleaned <- taxa_vec |>
    stringr::str_trim() |> stringr::str_squish() |>
    stringr::str_remove_all("\\s*\\([^)]*\\)") |>
    stringr::str_remove("\\s*\\[[^\\]]*\\]$") |>
    stringr::str_remove("\\s+(ssp\\.|subsp\\.|var\\.|f\\.|cv\\.)\\s+\\S+.*$") |>
    stringr::str_remove("(?i)\\s+sp+\\.?\\s*$") |>
    stringr::str_trim()
  genus_vec <- stringr::word(cleaned, 1)
  second_wd <- stringr::word(cleaned, 2)
  has_epi   <- !is.na(second_wd) &
    stringr::str_detect(second_wd, "^[a-z][a-z\\-]+$") &
    !second_wd %in% c("sp","spp","sp.","spp.")
  binomial <- dplyr::if_else(has_epi, paste(genus_vec, second_wd), NA_character_)
  flag_community <- stringr::str_detect(
    taxa_vec, stringr::regex("community|stand|mixed|plot|forest|woodland", ignore_case = TRUE))
  dplyr::tibble(
    taxa_orig      = taxa_vec,
    taxa_cleaned   = cleaned,
    binomial       = binomial,
    genus          = genus_vec,
    flag_sp_only   = !has_epi & !is.na(genus_vec),
    flag_community = flag_community,
    flag_uncertain = is.na(binomial) | flag_community
  )
}

ne_parsed <- ne_parse_taxa(nik_enq$taxa)

ne_annotated <- nik_enq |>
  bind_cols(ne_parsed |> select(-taxa_orig, -flag_community)) |>
  left_join(ne_genus_lookup, by = "genus") |>
  mutate(
    record_type = case_when(
      flag_crop      ~ "Crop/herb",
      flag_community ~ "Community/mixed",
      ne_division == "Gymnosperm" ~ "Gymnosperm",
      ne_division == "Angiosperm" ~ "Angiosperm",
      TRUE ~ "Unknown"
    )
  )

cat("Record type breakdown:\n")
print(ne_annotated |> count(record_type, sort = TRUE))
```

```{r species-means}
# Primary dataset: wild woody species only (exclude crops and community stands)
ne_wild <- ne_annotated |>
  dplyr::filter(!flag_crop, !flag_community)

# Species means: use records with valid binomial, known biomass > 0
ne_sp_means <- ne_wild |>
  dplyr::filter(!flag_uncertain, !is.na(binomial), total_biomass_kg > 0) |>
  dplyr::mutate(
    # Compute non-leaf at record level BEFORE log-transform (Jensen's inequality)
    nonleaf_kg  = dplyr::if_else(!is.na(leaf_biomass_kg) & leaf_biomass_kg > 0 &
                                   total_biomass_kg > leaf_biomass_kg,
                                 total_biomass_kg - leaf_biomass_kg, NA_real_),
    log10_total  = log10(total_biomass_kg),
    log10_stem   = log10(stem_biomass_kg),
    log10_leaf   = log10(leaf_biomass_kg),
    log10_root   = log10(root_biomass_kg),
    log10_shoot  = log10(shoot_biomass_kg),
    log10_repro  = log10(repro_biomass_kg),
    log10_height = log10(height_m),
    log10_nonleaf = log10(nonleaf_kg),
    log10_total_growth = log10(total_growth_kgyr),
    log10_stem_growth  = log10(stem_growth_kgyr),
    log10_leaf_growth  = log10(leaf_growth_kgyr),
    log10_root_growth  = log10(root_growth_kgyr)
  ) |>
  dplyr::mutate(across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .))) |>
  dplyr::group_by(binomial, ne_family, ne_division) |>
  dplyr::summarise(
    across(starts_with("log10_"), ~ mean(.x, na.rm = TRUE)),
    n_records = n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .))) |>
  dplyr::mutate(
    sp_tree = gsub(" ", "_", binomial)
  )

cat(sprintf("Species means: %d species, %d families\n",
            nrow(ne_sp_means), n_distinct(ne_sp_means$ne_family, na.rm = TRUE)))
```

```{r taxonomy-summary-plot, fig.height=4}
ne_sp_means |>
  dplyr::filter(!is.na(ne_division)) |>
  dplyr::count(ne_family, ne_division) |>
  dplyr::arrange(desc(n)) |>
  dplyr::slice_head(n = 20) |>
  dplyr::mutate(ne_family = reorder(ne_family, n)) |>
  ggplot(aes(ne_family, n, fill = ne_division)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(Gymnosperm = "#E65100", Angiosperm = "#1976D2")) +
  labs(title = "Top 20 families in species-mean dataset",
       x = NULL, y = "N species", fill = "Division")
```

---

# Phylogeny {#sec-phylogeny}

```{r build-phylogeny}
ne_phy_tree <- NULL

if (HAS_PHYLOMAKER && nrow(ne_sp_means) > 10) {
  sp_for_tree <- ne_sp_means |>
    dplyr::filter(!is.na(ne_family), !is.na(binomial)) |>
    dplyr::mutate(genus_col = stringr::word(binomial, 1)) |>
    dplyr::distinct(binomial, genus_col, ne_family) |>
    dplyr::rename(species = binomial, genus = genus_col, family = ne_family)

  ne_phy_tree <- tryCatch({
    res <- V.PhyloMaker2::phylo.maker(
      sp.list = data.frame(
        species = gsub(" ", "_", sp_for_tree$species),
        genus   = sp_for_tree$genus,
        family  = sp_for_tree$family),
      tree    = V.PhyloMaker2::GBOTB.extended.TPL,
      nodes   = V.PhyloMaker2::nodes.info.1.TPL,
      scenario = "S3")
    # V.PhyloMaker2 >= 0.1.0 returns res$scenario.3 (not res$phylo)
    tr_obj <- if (!is.null(res$scenario.3)) res$scenario.3
              else if (!is.null(res$phylo))  res$phylo
              else NULL
    if (!is.null(tr_obj)) {
      tr_obj$node.label <- NULL
      tr_obj
    } else NULL
  }, error = function(e) { message("V.PhyloMaker2 failed: ", e$message); NULL })

  if (!is.null(ne_phy_tree)) {
    n_matched <- sum(ne_sp_means$sp_tree %in% ne_phy_tree$tip.label)
    cat(sprintf("Phylogeny: %d tips; %d/%d species matched\n",
                length(ne_phy_tree$tip.label), n_matched, nrow(ne_sp_means)))
  }
} else {
  message("V.PhyloMaker2 not available — PGLS analyses will be skipped.")
}
```

---

# Phylogenetic Signal {#sec-phylosignal}

```{r phylo-signal}
if (!is.null(ne_phy_tree)) {
  signal_vars <- c("log10_leaf", "log10_stem", "log10_root", "log10_shoot",
                   "log10_total", "log10_height", "log10_nonleaf")

  signal_results <- map_dfr(signal_vars, function(v) {
    sub <- ne_sp_means |>
      dplyr::filter(!is.na(.data[[v]]), sp_tree %in% ne_phy_tree$tip.label) |>
      dplyr::distinct(sp_tree, .keep_all = TRUE)
    if (nrow(sub) < 10) return(NULL)
    tr2 <- ape::keep.tip(ne_phy_tree, sub$sp_tree)
    x   <- setNames(sub[[v]], sub$sp_tree)
    lam <- tryCatch(phytools::phylosig(tr2, x, method = "lambda", test = TRUE),
                    error = function(e) NULL)
    Kst <- tryCatch(phytools::phylosig(tr2, x, method = "K", test = TRUE),
                    error = function(e) NULL)
    tibble(
      trait   = v,
      n       = nrow(sub),
      lambda  = if (!is.null(lam)) round(lam$lambda, 3) else NA_real_,
      lam_p   = if (!is.null(lam)) round(lam$P, 4)     else NA_real_,
      K       = if (!is.null(Kst)) round(Kst$K, 3)     else NA_real_,
      K_p     = if (!is.null(Kst)) round(Kst$P, 4)     else NA_real_
    )
  })

  kable(signal_results,
        caption = "Phylogenetic signal: Pagel's lambda and Blomberg's K for each organ trait.")
} else {
  message("Phylogeny not available — phylogenetic signal analysis skipped.")
}
```

---

# run_one_allom Helper {#sec-helper}

```{r run-one-allom}
run_one_allom <- function(df, tree, y_col, x_col, rel_label,
                           wbe_slope = NA_real_, group_label = "All") {
  sub <- df |>
    dplyr::filter(is.finite(.data[[y_col]]), is.finite(.data[[x_col]]))
  if (!is.null(tree)) sub <- sub |> dplyr::filter(sp_tree %in% tree$tip.label)
  n <- nrow(sub)
  if (n < 8) {
    message(sprintf("Skipping %s ~ %s [%s] — n=%d", y_col, x_col, group_label, n))
    return(NULL)
  }

  ols_f  <- lm(as.formula(paste(y_col, "~", x_col)), data = sub)
  ols_s  <- summary(ols_f)
  ols_b  <- coef(ols_f)[2]
  ols_se <- coef(ols_s)[2, 2]
  ols_ci <- confint(ols_f)[2, ]
  ols_r2 <- ols_s$r.squared
  ols_p  <- if (!is.na(wbe_slope))
    2 * pt(abs((ols_b - wbe_slope) / ols_se), n - 2, lower.tail = FALSE) else NA_real_

  sma_f <- tryCatch(
    smatr::sma(as.formula(paste(y_col, "~", x_col)), data = sub),
    error = function(e) NULL)
  if (!is.null(sma_f)) {
    sma_b    <- sma_f$coef[[1]][2, 1]
    sma_cilo <- sma_f$coef[[1]][2, 2]
    sma_cihi <- sma_f$coef[[1]][2, 3]
    sma_r2   <- sma_f$r2[[1]]
    sma_p    <- if (!is.na(wbe_slope))
      tryCatch({
        st <- smatr::slope.test(sub[[y_col]], sub[[x_col]],
                                test.value = wbe_slope, method = "SMA")
        st$p
      }, error = function(e) NA_real_) else NA_real_
  } else {
    sma_b <- sma_cilo <- sma_cihi <- sma_r2 <- sma_p <- NA_real_
  }

  pgls_b <- pgls_se <- pgls_r2 <- pgls_lam <- pgls_p <-
    pgls_cilo <- pgls_cihi <- NA_real_
  if (!is.null(tree) && n >= 8) {
    tr2 <- ape::keep.tip(tree, sub$sp_tree)
    tr2$node.label <- NULL
    # Deduplicate by sp_tree before rowname assignment (prevents crash on duplicates)
    df_p <- as.data.frame(dplyr::distinct(sub, sp_tree, .keep_all = TRUE))
    rownames(df_p) <- df_p$sp_tree
    cd <- tryCatch(
      caper::comparative.data(phy = tr2, data = df_p, names.col = "sp_tree",
                               vcv = TRUE, warn.dropped = FALSE),
      error = function(e) NULL)
    if (!is.null(cd)) {
      pg_fixed_lambda <- FALSE
      pg <- tryCatch(
        caper::pgls(as.formula(paste(y_col, "~", x_col)),
                    data = cd, lambda = "ML", bounds = list(lambda = c(1e-6, 1))),
        error = function(e) {
          message(sprintf("PGLS ML failed [%s/%s]: %s — retrying lambda=1",
                          rel_label, group_label, e$message))
          pg_fixed_lambda <<- TRUE
          tryCatch(caper::pgls(as.formula(paste(y_col, "~", x_col)),
                               data = cd, lambda = 1),
                   error = function(e2) NULL)
        })
      if (!is.null(pg)) {
        pg_s    <- summary(pg)
        pg_df   <- tryCatch(pg_s$df[2], error = function(e) n - 2L)
        pgls_b  <- coef(pg)[2]
        pgls_se <- coef(pg_s)[2, 2]
        pgls_t  <- qt(0.975, df = pg_df)
        pgls_ci <- pgls_b + c(-1, 1) * pgls_t * pgls_se
        pgls_cilo <- pgls_ci[1]; pgls_cihi <- pgls_ci[2]
        pgls_r2  <- pg_s$r.squared
        pgls_lam <- tryCatch(pg$param["lambda"], error = function(e) NA_real_)
        pgls_p   <- if (!is.na(wbe_slope))
          2 * pt(abs((pgls_b - wbe_slope) / pgls_se), df = pg_df, lower.tail = FALSE)
        else NA_real_
      }
    }
  }

  tibble::tibble(
    relationship = rel_label, group = group_label,
    y_var = y_col, x_var = x_col,
    wbe_slope = wbe_slope, n_species = n,
    method     = c("OLS", "SMA", "PGLS"),
    slope      = c(ols_b,    sma_b,    pgls_b),
    slope_se   = c(ols_se,   NA_real_, pgls_se),
    ci_lower   = c(ols_ci[1], sma_cilo, pgls_cilo),
    ci_upper   = c(ols_ci[2], sma_cihi, pgls_cihi),
    r_squared  = c(ols_r2,   sma_r2,   pgls_r2),
    lambda_ml  = c(NA_real_, NA_real_, pgls_lam),
    p_vs_wbe   = c(ols_p,    sma_p,    pgls_p)
  )
}
```

---

# Interspecific Scaling — All Relationships {#sec-interspecific}

This section evaluates WBE/Enquist predictions for organ mass, height, and whole-plant growth scaling.
The key growth-rate relationship is tested as:

\[
\frac{dM}{dt} \propto M^{3/4}
\]

where `total_growth_kgyr` approximates the annual net biomass increment and `total_biomass_kg` is the whole-plant size metric.
The predicted WBE/Enquist exponents in this dataset are:

- total biomass vs height: exponent $\beta = 8/3 \approx 2.67$,
- whole-plant growth rate vs total biomass: exponent $\beta = 3/4 = 0.75$,
- leaf mass vs total biomass: exponent $\beta = 3/4 = 0.75$.

Because the Niklas-Enquist dataset contains annual net increments rather than instantaneous metabolic fluxes, this is an approximate descriptive test of WBE growth scaling, not a direct test of instantaneous metabolic rates.

```{r define-relationships}
# All organ relationships to analyze with their WBE priors
relationships <- list(
  # Organ ~ Total biomass
  list(y = "log10_leaf",  x = "log10_total",  label = "Leaf ~ Total",        wbe = WBE_LEAF_TOTAL),
  list(y = "log10_root",  x = "log10_total",  label = "Root ~ Total",        wbe = WBE_ROOT_TOTAL),
  list(y = "log10_stem",  x = "log10_total",  label = "Stem ~ Total",        wbe = WBE_STEM_TOTAL),
  list(y = "log10_shoot", x = "log10_total",  label = "Shoot ~ Total",       wbe = WBE_SHOOT_TOTAL),
  list(y = "log10_repro", x = "log10_total",  label = "Repro ~ Total",       wbe = NA_real_),
  # Organ ~ Organ
  list(y = "log10_leaf",  x = "log10_stem",   label = "Leaf ~ Stem (KEY)",   wbe = WBE_LEAF_STEM),
  list(y = "log10_leaf",  x = "log10_root",   label = "Leaf ~ Root",         wbe = WBE_LEAF_ROOT),
  list(y = "log10_root",  x = "log10_stem",   label = "Root ~ Stem",         wbe = WBE_ROOT_STEM),
  list(y = "log10_root",  x = "log10_shoot",  label = "Root ~ Shoot",        wbe = WBE_ROOT_SHOOT),
  list(y = "log10_leaf",  x = "log10_nonleaf",label = "Leaf ~ Non-leaf",     wbe = WBE_LEAF_TOTAL),
  # Height relationships
  list(y = "log10_total", x = "log10_height", label = "Total ~ Height",      wbe = WBE_TOTAL_H),
  list(y = "log10_stem",  x = "log10_height", label = "Stem ~ Height",       wbe = NA_real_),
  list(y = "log10_leaf",  x = "log10_height", label = "Leaf ~ Height",       wbe = NA_real_),
  # Growth
  list(y = "log10_total_growth", x = "log10_total", label = "TotalGrowth ~ Total", wbe = WBE_GROWTH),
  list(y = "log10_stem_growth",  x = "log10_stem",  label = "StemGrowth ~ Stem",   wbe = WBE_GROWTH),
  list(y = "log10_leaf_growth",  x = "log10_leaf",  label = "LeafGrowth ~ Leaf",   wbe = WBE_GROWTH),
  list(y = "log10_root_growth",  x = "log10_root",  label = "RootGrowth ~ Root",   wbe = WBE_GROWTH)
)
```

```{r interspecific-all, cache=TRUE}
inter_results <- map_dfr(relationships, function(r) {
  run_one_allom(
    df          = ne_sp_means,
    tree        = ne_phy_tree,
    y_col       = r$y,
    x_col       = r$x,
    rel_label   = r$label,
    wbe_slope   = r$wbe,
    group_label = "All"
  )
})

kable(inter_results |>
        dplyr::select(relationship, group, method, n_species, slope, ci_lower, ci_upper,
               r_squared, lambda_ml, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
      caption = "Interspecific organ scaling: OLS, SMA, PGLS across all wild species.")
```

```{r interspecific-forest, fig.height=9}
inter_results |>
  dplyr::filter(!is.na(slope), is.finite(slope)) |>
  dplyr::mutate(
    relationship = factor(relationship, levels = unique(inter_results$relationship)),
    method = factor(method, levels = c("OLS", "SMA", "PGLS"))
  ) |>
  ggplot(aes(slope, relationship, colour = method, shape = method)) +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                  position = position_dodge(width = 0.5), size = 0.5) +
  geom_vline(data = inter_results |>
               dplyr::filter(!is.na(wbe_slope)) |>
               dplyr::distinct(relationship, wbe_slope),
             aes(xintercept = wbe_slope), colour = "red",
             linetype = "dashed", linewidth = 0.6) +
  scale_colour_manual(values = c(OLS = "#333333", SMA = "#1565C0", PGLS = "#2E7D32")) +
  facet_wrap(~ "All species") +
  labs(title = "Interspecific organ scaling — all relationships",
       subtitle = "Red dashed = WBE prediction; points jittered by method",
       x = "Slope ± 95% CI", y = NULL, colour = "Method", shape = "Method") +
  theme(axis.text.y = element_text(size = 9))
```

```{r leaf-stem-scatter, fig.height=5}
ne_sp_means |>
  dplyr::filter(is.finite(log10_leaf), is.finite(log10_stem)) |>
  ggplot(aes(log10_stem, log10_leaf, colour = ne_division)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, aes(group = 1),
              colour = "black", linewidth = 1) +
  geom_abline(slope = WBE_LEAF_STEM,
              intercept = {
                sub <- ne_sp_means |> dplyr::filter(is.finite(log10_leaf), is.finite(log10_stem))
                mean(sub$log10_leaf, na.rm = TRUE) - WBE_LEAF_STEM * mean(sub$log10_stem, na.rm = TRUE)
              },
              colour = "red", linetype = "dashed", linewidth = 1) +
  annotation_logticks(sides = "bl") +
  scale_colour_manual(values = c(Gymnosperm = "#E65100", Angiosperm = "#1976D2"),
                      na.value = "grey60") +
  labs(title = "Leaf ~ Stem biomass (KEY WBE relationship)",
       subtitle = "Black=OLS, Red-dashed=WBE β=3/4",
       x = "log10(Stem biomass kg)", y = "log10(Leaf biomass kg)",
       colour = "Division")
```

---

# Gymnosperm vs Angiosperm {#sec-division}

```{r division-stratified, cache=TRUE}
div_results <- map_dfr(c("Gymnosperm", "Angiosperm"), function(div) {
  sp_div <- ne_sp_means |> dplyr::filter(ne_division == div)
  map_dfr(relationships, function(r) {
    run_one_allom(
      df          = sp_div,
      tree        = ne_phy_tree,
      y_col       = r$y,
      x_col       = r$x,
      rel_label   = r$label,
      wbe_slope   = r$wbe,
      group_label = div
    )
  })
})

kable(div_results |>
        dplyr::select(relationship, group, method, n_species, slope, ci_lower, ci_upper,
               r_squared, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
      caption = "Division-stratified organ scaling: Gymnosperm vs. Angiosperm.")
```

```{r ancova-slope-homogeneity}
# ANCOVA slope homogeneity: test division × x_col interaction for key relationships
key_rels <- list(
  list(y = "log10_leaf",  x = "log10_stem",  label = "Leaf ~ Stem"),
  list(y = "log10_root",  x = "log10_shoot", label = "Root ~ Shoot"),
  list(y = "log10_stem",  x = "log10_total", label = "Stem ~ Total"),
  list(y = "log10_leaf",  x = "log10_total", label = "Leaf ~ Total")
)

ancova_results <- map_dfr(key_rels, function(r) {
  sub <- ne_sp_means |>
    dplyr::filter(
      is.finite(.data[[r$y]]), is.finite(.data[[r$x]]),
      ne_division %in% c("Gymnosperm", "Angiosperm")
    )
  if (nrow(sub) < 15) return(NULL)
  m_full <- lm(as.formula(paste(r$y, "~ ne_division *", r$x)), data = sub)
  m_para <- lm(as.formula(paste(r$y, "~ ne_division +", r$x)), data = sub)
  av <- anova(m_para, m_full)
  tibble(
    relationship     = r$label,
    n                = nrow(sub),
    interaction_F    = round(av$F[2], 3),
    interaction_p    = round(av$`Pr(>F)`[2], 4),
    slopes_homogeneous = av$`Pr(>F)`[2] > 0.05
  )
})

kable(ancova_results,
      caption = "ANCOVA test of slope homogeneity between Gymnosperm and Angiosperm.")
```

```{r division-forest, fig.height=8}
div_results |>
  dplyr::filter(method == "SMA", !is.na(slope), is.finite(slope)) |>
  dplyr::mutate(relationship = factor(relationship, levels = unique(inter_results$relationship))) |>
  ggplot(aes(slope, relationship, colour = group)) +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                  position = position_dodge(width = 0.5), size = 0.5) +
  geom_vline(data = div_results |>
               dplyr::filter(!is.na(wbe_slope)) |>
               dplyr::distinct(relationship, wbe_slope),
             aes(xintercept = wbe_slope), colour = "red",
             linetype = "dashed", linewidth = 0.6) +
  scale_colour_manual(values = c(Gymnosperm = "#E65100", Angiosperm = "#1976D2")) +
  labs(title = "SMA slopes by division — Gymnosperm vs. Angiosperm",
       subtitle = "Red dashed = WBE prediction",
       x = "SMA slope ± 95% CI", y = NULL, colour = "Division") +
  theme(axis.text.y = element_text(size = 9))
```

---

# Family-Level Analysis {#sec-family}

```{r family-level, cache=TRUE, fig.height=8}
# Families with ≥5 species in species-mean dataset
family_5sp <- ne_sp_means |>
  dplyr::filter(!is.na(ne_family)) |>
  dplyr::count(ne_family) |>
  dplyr::filter(n >= 5) |>
  dplyr::pull(ne_family)

cat(sprintf("Families with ≥5 species: %d\n", length(family_5sp)))

key_rels_fam <- list(
  list(y = "log10_leaf",  x = "log10_stem",  label = "Leaf ~ Stem",  wbe = WBE_LEAF_STEM),
  list(y = "log10_root",  x = "log10_shoot", label = "Root ~ Shoot", wbe = WBE_ROOT_SHOOT),
  list(y = "log10_stem",  x = "log10_total", label = "Stem ~ Total", wbe = WBE_STEM_TOTAL),
  list(y = "log10_leaf",  x = "log10_total", label = "Leaf ~ Total", wbe = WBE_LEAF_TOTAL)
)

fam_results <- map_dfr(key_rels_fam, function(r) {
  map_dfr(family_5sp, function(fam) {
    sp_fam <- ne_sp_means |> dplyr::filter(ne_family == fam)
    # PGLS only if ≥10 species matched to tree
    use_tree <- if (!is.null(ne_phy_tree) &&
                    sum(sp_fam$sp_tree %in% ne_phy_tree$tip.label) >= 10)
      ne_phy_tree else NULL
    run_one_allom(
      df          = sp_fam,
      tree        = use_tree,
      y_col       = r$y,
      x_col       = r$x,
      rel_label   = r$label,
      wbe_slope   = r$wbe,
      group_label = fam
    )
  })
})

kable(fam_results |>
        dplyr::filter(method == "OLS") |>
        dplyr::select(relationship, group, n_species, slope, ci_lower, ci_upper,
               r_squared, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))) |>
        dplyr::arrange(relationship, group),
      caption = "Family-level OLS slopes for key organ relationships (families ≥5 species).")
```

```{r family-heterogeneity}
# Cochran Q / I² heterogeneity across families for each key relationship
het_results <- map_dfr(key_rels_fam, function(r) {
  sub <- fam_results |>
    dplyr::filter(relationship == r$label, method == "OLS",
           !is.na(slope), !is.na(slope_se), slope_se > 0)
  if (nrow(sub) < 3) return(NULL)
  w  <- 1 / sub$slope_se^2
  wm <- weighted.mean(sub$slope, w)
  Q  <- sum(w * (sub$slope - wm)^2)
  df <- nrow(sub) - 1
  I2 <- max(0, (Q - df) / Q * 100)
  tibble(relationship = r$label, n_families = nrow(sub),
         Q = round(Q, 2), df = df,
         Q_p = round(pchisq(Q, df, lower.tail = FALSE), 4),
         I2_pct = round(I2, 1))
})

kable(het_results,
      caption = "Heterogeneity (I²) across families for key organ relationships.")
```

```{r family-forest-plot, fig.height=9}
fam_results |>
  dplyr::filter(method == "SMA", !is.na(slope), is.finite(slope)) |>
  dplyr::mutate(group = reorder(group, slope)) |>
  ggplot(aes(slope, group)) +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper), size = 0.4) +
  geom_vline(data = fam_results |>
               dplyr::filter(method == "SMA") |>
               dplyr::distinct(relationship, wbe_slope) |>
               dplyr::filter(!is.na(wbe_slope)),
             aes(xintercept = wbe_slope),
             colour = "red", linetype = "dashed", linewidth = 0.7,
             inherit.aes = FALSE) +
  facet_wrap(~ relationship, scales = "free_x", ncol = 2) +
  labs(title = "Family-level SMA slopes — key organ relationships",
       subtitle = "Red dashed = WBE prediction",
       x = "SMA slope ± 95% CI", y = NULL) +
  theme(axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 9))
```

---

# Intraspecific Scaling (Within NE Taxa) {#sec-intraspecific}

Within the Niklas & Enquist database itself, 57 taxa have ≥10 records each, enabling intraspecific log-log regressions. We exclude crop/annual taxa from WBE tests and flag regressions with log10 x-range < 1 decade as low-leverage.

```{r intraspecific-taxa}
# Use raw ne_annotated before species-mean aggregation
# Wild species only, crop excluded
ne_wild_intra <- ne_annotated |>
  dplyr::filter(!flag_crop, !flag_community, !is.na(taxa))

# Count records per taxon (using parsed binomial where available, else taxa_cleaned)
intra_taxon_counts <- ne_wild_intra |>
  dplyr::mutate(intra_taxon = dplyr::if_else(!is.na(binomial), binomial, taxa_cleaned)) |>
  dplyr::count(intra_taxon, sort = TRUE) |>
  dplyr::filter(n >= 10)

cat(sprintf("Wild taxa with ≥10 NE records: %d\n", nrow(intra_taxon_counts)))
kable(intra_taxon_counts |> head(20), caption = "Top 20 taxa by record count (intraspecific analysis).")
```

```{r intraspecific-scaling, cache=TRUE}
intra_rels <- list(
  list(y = "log_leaf", x = "log_stem",   label = "Leaf ~ Stem",    wbe = WBE_LEAF_STEM),
  list(y = "log_root", x = "log_stem",   label = "Root ~ Stem",    wbe = WBE_ROOT_STEM),
  list(y = "log_root", x = "log_shoot",  label = "Root ~ Shoot",   wbe = WBE_ROOT_SHOOT),
  list(y = "log_total",x = "log_height", label = "Total ~ Height", wbe = WBE_TOTAL_H)
)

intra_results <- map_dfr(intra_taxon_counts$intra_taxon, function(tx) {
  sub_base <- ne_wild_intra |>
    dplyr::mutate(intra_taxon = dplyr::if_else(!is.na(binomial), binomial, taxa_cleaned)) |>
    dplyr::filter(intra_taxon == tx)

  map_dfr(intra_rels, function(r) {
    sub <- sub_base |>
      dplyr::filter(is.finite(.data[[r$y]]), is.finite(.data[[r$x]]))
    n <- nrow(sub)
    if (n < 8) return(NULL)

    x_range <- diff(range(sub[[r$x]], na.rm = TRUE))

    ols_f  <- tryCatch(lm(as.formula(paste(r$y, "~", r$x)), data = sub), error = function(e) NULL)
    sma_f  <- tryCatch(smatr::sma(as.formula(paste(r$y, "~", r$x)), data = sub), error = function(e) NULL)
    if (is.null(ols_f)) return(NULL)

    ols_s  <- summary(ols_f)
    ols_b  <- coef(ols_f)[2]
    ols_ci <- confint(ols_f)[2, ]
    ols_se <- coef(ols_s)[2, 2]

    sma_b    <- if (!is.null(sma_f)) sma_f$coef[[1]][2, 1] else NA_real_
    sma_cilo <- if (!is.null(sma_f)) sma_f$coef[[1]][2, 2] else NA_real_
    sma_cihi <- if (!is.null(sma_f)) sma_f$coef[[1]][2, 3] else NA_real_

    tibble(
      taxon       = tx,
      relationship = r$label,
      wbe_slope   = r$wbe,
      n           = n,
      x_log_range = round(x_range, 2),
      low_leverage = x_range < 1.0,   # flag if < 1 log10 decade
      ols_slope   = round(ols_b, 3),
      ols_ci_lo   = round(ols_ci[1], 3),
      ols_ci_hi   = round(ols_ci[2], 3),
      ols_r2      = round(ols_s$r.squared, 3),
      sma_slope   = round(sma_b, 3),
      sma_ci_lo   = round(sma_cilo, 3),
      sma_ci_hi   = round(sma_cihi, 3)
    )
  })
})

kable(intra_results |>
        dplyr::filter(relationship == "Leaf ~ Stem") |>
        dplyr::arrange(desc(n)) |>
        dplyr::select(taxon, n, x_log_range, low_leverage,
               ols_slope, ols_ci_lo, ols_ci_hi, ols_r2,
               sma_slope, sma_ci_lo, sma_ci_hi),
      caption = "Intraspecific Leaf ~ Stem scaling per taxon (n ≥ 8 records).")
```

```{r intraspecific-all-rels-table}
kable(intra_results |>
        dplyr::arrange(relationship, taxon) |>
        dplyr::select(relationship, taxon, n, x_log_range, low_leverage,
               ols_slope, sma_slope, wbe_slope),
      caption = "All intraspecific relationships: OLS and SMA slopes per taxon.")
```

```{r intraspecific-slope-histogram, fig.height=6}
intra_results |>
  dplyr::filter(!is.na(sma_slope), !low_leverage) |>
  ggplot(aes(sma_slope)) +
  geom_histogram(bins = 18, fill = "#1565C0", alpha = 0.7) +
  # Per-relationship WBE reference lines (avoids vline bleed across facets)
  geom_vline(data = intra_results |>
               dplyr::distinct(relationship, wbe_slope),
             aes(xintercept = wbe_slope),
             colour = "red", linetype = "dashed", linewidth = 1.0,
             inherit.aes = FALSE) +
  facet_wrap(~ relationship, scales = "free_x") +
  labs(title = "Distribution of intraspecific SMA slopes (x log-range ≥ 1 decade)",
       subtitle = "Red dashed = WBE prediction per relationship",
       x = "SMA slope", y = "Count")
```

```{r intraspecific-vs-interspecific}
# Summary: mean intraspecific slope vs. interspecific
intra_summary <- intra_results |>
  dplyr::filter(!low_leverage) |>
  dplyr::group_by(relationship) |>
  dplyr::summarise(
    n_taxa          = n(),
    mean_ols        = round(mean(ols_slope, na.rm = TRUE), 3),
    sd_ols          = round(sd(ols_slope,   na.rm = TRUE), 3),
    mean_sma        = round(mean(sma_slope, na.rm = TRUE), 3),
    sd_sma          = round(sd(sma_slope,   na.rm = TRUE), 3),
    wbe_slope       = first(wbe_slope),
    .groups = "drop"
  )

inter_summary <- inter_results |>
  dplyr::filter(method %in% c("OLS","SMA"), group == "All") |>
  dplyr::group_by(relationship, method) |>
  dplyr::summarise(inter_slope = round(mean(slope, na.rm = TRUE), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = method, values_from = inter_slope,
                     names_prefix = "inter_")

kable(dplyr::left_join(
    intra_summary,
    inter_summary |> dplyr::mutate(relationship = str_remove(relationship, " \\(KEY\\)")),
    by = "relationship"),
  caption = paste(
    "Intraspecific (mean ± SD across taxa) vs. interspecific slopes.",
    "Intraspecific rows exclude taxa with x log-range < 1 decade."))
```

---

# WBE Scorecard {#sec-scorecard}

```{r wbe-scorecard}
scorecard <- bind_rows(
  inter_results |> dplyr::mutate(scale = "Interspecific"),
  div_results   |> dplyr::mutate(scale = paste("Div:", group))
) |>
  dplyr::filter(!is.na(wbe_slope), !is.na(slope), !is.na(ci_lower), !is.na(ci_upper)) |>
  dplyr::mutate(
    wbe_in_ci = wbe_slope >= ci_lower & wbe_slope <= ci_upper,
    support   = dplyr::case_when(
      wbe_in_ci & !is.na(p_vs_wbe) & p_vs_wbe > 0.05 ~ "Supported (β in CI, p>0.05)",
      wbe_in_ci                                        ~ "Marginal (β in CI, p≤0.05)",
      TRUE                                             ~ "Rejected (β outside CI)"
    )
  ) |>
  dplyr::select(scale, relationship, method, group, n_species, wbe_slope,
         slope, ci_lower, ci_upper, p_vs_wbe, wbe_in_ci, support) |>
  dplyr::mutate(across(c(slope, ci_lower, ci_upper, p_vs_wbe), ~ round(.x, 3)))

kable(scorecard |>
        dplyr::filter(scale == "Interspecific") |>
        dplyr::select(-scale),
      caption = "WBE scorecard: interspecific scaling. Is WBE β within 95% CI?")
```

```{r scorecard-division}
kable(scorecard |>
        dplyr::filter(str_starts(scale, "Div")) |>
        dplyr::select(-scale),
      caption = "WBE scorecard: division-stratified analyses.")
```

```{r scorecard-summary, fig.height=5}
scorecard |>
  dplyr::filter(scale == "Interspecific") |>
  dplyr::count(method, support) |>
  ggplot(aes(method, n, fill = support)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
    "Supported (β in CI, p>0.05)"  = "#43A047",
    "Marginal (β in CI, p≤0.05)"  = "#FFA726",
    "Rejected (β outside CI)"      = "#E53935")) +
  labs(title = "WBE scorecard summary by method (interspecific)",
       x = "Method", y = "N relationships", fill = "WBE support")
```

---

# Sensitivity: Including Crops {#sec-sensitivity}

```{r sensitivity-with-crops, cache=TRUE}
# Species means including crop records
ne_sp_all <- ne_annotated |>
  dplyr::filter(!flag_community, !is.na(binomial), total_biomass_kg > 0) |>
  dplyr::mutate(
    # Compute non-leaf at record level BEFORE log-transform (avoids Jensen's inequality bias)
    nonleaf_kg  = dplyr::if_else(!is.na(leaf_biomass_kg) & leaf_biomass_kg > 0 &
                                   total_biomass_kg > leaf_biomass_kg,
                                 total_biomass_kg - leaf_biomass_kg, NA_real_),
    log10_total    = log10(total_biomass_kg),
    log10_stem     = log10(stem_biomass_kg),
    log10_leaf     = log10(leaf_biomass_kg),
    log10_root     = log10(root_biomass_kg),
    log10_shoot    = log10(shoot_biomass_kg),
    log10_repro    = log10(repro_biomass_kg),
    log10_height   = log10(height_m),
    log10_nonleaf  = log10(nonleaf_kg),
    log10_total_growth = log10(total_growth_kgyr),
    log10_stem_growth  = log10(stem_growth_kgyr),
    log10_leaf_growth  = log10(leaf_growth_kgyr),
    log10_root_growth  = log10(root_growth_kgyr)
  ) |>
  dplyr::mutate(across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .))) |>
  dplyr::group_by(binomial) |>
  dplyr::summarise(
    across(starts_with("log10_"), ~ mean(.x, na.rm = TRUE)),
    n_records = n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .)),
    sp_tree = gsub(" ", "_", binomial)
  )

# Key relationships only for sensitivity
sens_rels <- list(
  list(y = "log10_leaf", x = "log10_stem",  label = "Leaf ~ Stem",  wbe = WBE_LEAF_STEM),
  list(y = "log10_leaf", x = "log10_total", label = "Leaf ~ Total", wbe = WBE_LEAF_TOTAL),
  list(y = "log10_root", x = "log10_shoot", label = "Root ~ Shoot", wbe = WBE_ROOT_SHOOT)
)

sens_results <- map_dfr(sens_rels, function(r) {
  run_one_allom(
    df          = ne_sp_all,
    tree        = NULL,   # no tree needed for sensitivity comparison
    y_col       = r$y,
    x_col       = r$x,
    rel_label   = r$label,
    wbe_slope   = r$wbe,
    group_label = "Including crops"
  )
})

# Compare wild-only vs. with-crops
sens_compare <- bind_rows(
  inter_results |>
    dplyr::filter(relationship %in% unique(sens_results$relationship), method %in% c("OLS","SMA")) |>
    dplyr::mutate(dataset = "Wild only"),
  sens_results |>
    dplyr::filter(method %in% c("OLS","SMA")) |>
    dplyr::mutate(dataset = "Including crops")
)

kable(sens_compare |>
        dplyr::select(relationship, dataset, method, n_species, slope, ci_lower, ci_upper, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
      caption = "Sensitivity: WBE scorecard with vs. without crop species.")
```

---

# Bayesian PGLMM: Individual-Level Scaling {#sec-pglmm}

## Motivation and Design

The OLS, SMA, and PGLS analyses above operate on **species means** — one data point per species. A Bayesian Phylogenetic Generalised Linear Mixed Model (PGLMM) operates instead on **individual-level observations**, partitioning variance explicitly into three components:

- $\sigma^2_{phy}$: variance among species attributable to shared evolutionary history (phylogenetic signal)
- $\sigma^2_{spp}$: non-phylogenetic species-level variance (ecological/developmental differences not explained by phylogeny)
- $\sigma^2_e$: individual-level residual (within-species variation and measurement error)

### The Mundlak Device

Because the NE database contains 1–40+ observations per species spanning seedlings to old-growth trees, a naive individual-level regression would estimate a mixture of *inter-* and *intraspecific* slopes. We apply the **Mundlak (1978) within-between decomposition**:

$$\log_{10}(R_{ij}) = \alpha + \underbrace{\beta_B \,\overline{\log_{10}(M)}_i}_{\text{between-species}} + \underbrace{\beta_W (\log_{10}(M_{ij}) - \overline{\log_{10}(M)}_i)}_{\text{within-species}} + u_i^{phy} + u_i^{spp} + \varepsilon_{ij}$$

where $i$ indexes species and $j$ indexes individual observations within a species. The **between-species slope $\beta_B$** is the PGLMM quantity comparable to the PGLS interspecific slope. The **within-species slope $\beta_W$** is comparable to intraspecific OLS regressions.

The phylogenetic random effect $u_i^{phy} \sim \mathcal{N}(\mathbf{0}, \sigma^2_{phy} \mathbf{C})$, where $\mathbf{C}$ is the Brownian-motion phylogenetic correlation matrix from `ape::vcv.phylo(tree, corr = TRUE)`. Implementation uses `brms` (Bürkner 2017) with Stan.

### Caveats

- **$\beta_B$ is not identical to the PGLS slope**; it is a hierarchical Bayesian estimand from individual records. They are comparable estimators of the same theoretical quantity under shared assumptions.
- **Unbalanced sampling creates implicit weighting**: species with many records dominate estimation of $\sigma^2_e$ and $\beta_W$; singletons contribute only to $\beta_B$.
- **PGLMM is presented as a robustness and variance-decomposition supplement** to PGLS, not a replacement.
- Priors are weakly informative: $\beta_B, \beta_W \sim \mathcal{N}(1, 1)$; SD terms $\sim \text{Student-}t(3, 0, 0.5)$.
- Convergence assessed by $\hat{R} < 1.05$ and ESS ratio $> 0.1$ for all parameters.

```{r pglmm-packages}
HAS_BRMS <- requireNamespace("brms", quietly = TRUE) &&
  utils::packageVersion("brms") >= "2.17.0"
if (!HAS_BRMS) message("brms ≥ 2.17.0 not available — PGLMM section will be skipped.")
```

```{r pglmm-data-prep}
ne_wild_log10 <- ne_wild |>
  dplyr::filter(!is.na(binomial), !flag_uncertain) |>
  dplyr::mutate(
    log10_leaf  = log10(leaf_biomass_kg),
    log10_stem  = log10(stem_biomass_kg),
    log10_root  = log10(root_biomass_kg),
    log10_shoot = log10(shoot_biomass_kg),
    log10_total = log10(total_biomass_kg),
    log10_total_growth = log10(total_growth_kgyr),
    sp_tree     = gsub(" ", "_", binomial)
  ) |>
  dplyr::mutate(across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .)))

# Report records/species per division
ne_wild_log10 |>
  dplyr::filter(!is.na(ne_division)) |>
  dplyr::count(ne_division, name = "n_obs") |>
  dplyr::left_join(
    ne_wild_log10 |> dplyr::filter(!is.na(ne_division)) |>
      dplyr::group_by(ne_division) |>
      dplyr::summarise(n_spp = dplyr::n_distinct(sp_tree), .groups="drop"),
    by = "ne_division") |>
  kable(caption = "Individual-level records available for PGLMM by division.")

# Species sampling summary (to assess Mundlak device validity)
samp_summary <- ne_wild_log10 |>
  dplyr::filter(!is.na(binomial)) |>
  dplyr::count(sp_tree, name = "n_records") |>
  dplyr::summarise(
    n_singletons = sum(n_records == 1),
    n_2_to_5     = sum(n_records >= 2 & n_records <= 5),
    n_6plus      = sum(n_records >= 6),
    total_spp    = n(),
    pct_singleton = round(100 * mean(n_records == 1), 1)
  )
cat(sprintf(
  "Species sampling: %d total; %d singletons (%.1f%%); %d with 2-5 records; %d with ≥6 records\n",
  samp_summary$total_spp, samp_summary$n_singletons, samp_summary$pct_singleton,
  samp_summary$n_2_to_5, samp_summary$n_6plus))
```

```{r run-one-pglmm}
run_one_pglmm <- function(df_individual, tree,
                           y_col, x_col,
                           rel_label,
                           group_label = "All",
                           wbe_slope   = NA_real_,
                           iter        = 2000,
                           chains      = 4,
                           cache_dir   = NULL,
                           seed        = 42) {
  if (!requireNamespace("brms", quietly = TRUE)) return(NULL)
  library(brms)

  df <- df_individual |>
    dplyr::filter(!is.na(binomial), is.finite(.data[[y_col]]), is.finite(.data[[x_col]])) |>
    dplyr::mutate(sp_tree = gsub(" ", "_", binomial))

  shared_sp <- intersect(df$sp_tree, tree$tip.label)
  n_shared  <- length(shared_sp)
  if (n_shared < 8) {
    message(sprintf("Skipping %s [%s] — %d species matched to tree (need ≥ 8).", rel_label, group_label, n_shared))
    return(NULL)
  }

  pruned_tree <- ape::keep.tip(tree, shared_sp)
  pruned_tree$node.label <- NULL
  df <- df |> dplyr::filter(sp_tree %in% shared_sp)
  n_obs <- nrow(df); n_spp <- dplyr::n_distinct(df$sp_tree)
  message(sprintf("PGLMM %s [%s]: %d obs, %d spp", rel_label, group_label, n_obs, n_spp))

  # Mundlak device: within/between decomposition
  df <- df |>
    dplyr::group_by(sp_tree) |>
    dplyr::mutate(
      x_bar = mean(.data[[x_col]], na.rm = TRUE),
      x_dev = .data[[x_col]] - x_bar
    ) |>
    dplyr::ungroup() |>
    dplyr::rename(log10_y = !!rlang::sym(y_col)) |>
    dplyr::mutate(phylo = sp_tree, spp = sp_tree)

  # Phylogenetic correlation matrix (Brownian motion, corr=TRUE → diagonal=1)
  A <- ape::vcv.phylo(pruned_tree, corr = TRUE)

  priors <- c(
    brms::prior(normal(1, 1),         class = "b",   coef = "x_bar"),
    brms::prior(normal(1, 1),         class = "b",   coef = "x_dev"),
    brms::prior(normal(0, 2),         class = "Intercept"),
    brms::prior(student_t(3, 0, 0.5), class = "sd",  group = "phylo"),
    brms::prior(student_t(3, 0, 0.5), class = "sd",  group = "spp"),
    brms::prior(student_t(3, 0, 0.5), class = "sigma")
  )

  cache_file <- if (!is.null(cache_dir)) {
    dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
    file.path(cache_dir,
              sprintf("pglmm_%s_%s.rds",
                      gsub("[^a-zA-Z0-9]", "_", rel_label),
                      gsub("[^a-zA-Z0-9]", "_", group_label)))
  } else NULL

  # Try cmdstanr backend first, fall back to rstan
  backend <- if (requireNamespace("cmdstanr", quietly = TRUE)) "cmdstanr" else "rstan"

  fit <- tryCatch(
    brms::brm(
      formula = brms::bf(log10_y ~ x_bar + x_dev + (1 | gr(phylo, cov = A)) + (1 | spp)),
      data    = df,
      data2   = list(A = A),
      prior   = priors,
      family  = gaussian(),
      iter    = iter,
      warmup  = iter %/% 2,
      chains  = chains,
      cores   = chains,
      seed    = seed,
      file    = cache_file,
      refresh = 0,
      silent  = 2,
      backend = backend
    ),
    error = function(e) {
      message(sprintf("brms failed [%s / %s]: %s", rel_label, group_label, e$message))
      NULL
    }
  )
  if (is.null(fit)) return(NULL)

  rhat_max <- max(brms::rhat(fit), na.rm = TRUE)
  ess_min  <- min(brms::neff_ratio(fit), na.rm = TRUE)
  if (rhat_max > 1.05) warning(sprintf("R-hat > 1.05 for %s [%s]", rel_label, group_label))

  fe   <- brms::fixef(fit)
  beta_B <- fe["x_bar", "Estimate"];  se_B <- fe["x_bar", "Est.Error"]
  beta_W <- fe["x_dev", "Estimate"];  se_W <- fe["x_dev", "Est.Error"]
  ci_B   <- fe["x_bar", c("Q2.5", "Q97.5")]
  ci_W   <- fe["x_dev", c("Q2.5", "Q97.5")]

  vc      <- brms::VarCorr(fit, summary = TRUE)
  var_phy <- tryCatch(vc$phylo$sd[1, "Estimate"]^2,      error = function(e) NA_real_)
  var_spp <- tryCatch(vc$spp$sd[1, "Estimate"]^2,        error = function(e) NA_real_)
  var_e   <- tryCatch(vc$residual__$sd[1, "Estimate"]^2, error = function(e) NA_real_)
  var_tot <- sum(c(var_phy, var_spp, var_e), na.rm = TRUE)

  lambda_analog <- if (!is.na(var_phy) && !is.na(var_spp) && (var_phy + var_spp) > 0)
    var_phy / (var_phy + var_spp) else NA_real_
  h2_phy        <- if (!is.na(var_phy) && var_tot > 0) var_phy / var_tot else NA_real_
  icc_spp       <- if (var_tot > 0) sum(c(var_phy, var_spp), na.rm=TRUE) / var_tot else NA_real_

  # Full posterior for variance fraction CIs
  post <- brms::as_draws_df(fit)
  sd_phy_col <- grep("^sd_phylo", names(post), value = TRUE)[1]
  sd_spp_col <- grep("^sd_spp",  names(post), value = TRUE)[1]
  sig_col    <- grep("^sigma",   names(post), value = TRUE)[1]
  b_xbar_col <- grep("^b_x_bar", names(post), value = TRUE)[1]

  if (!is.na(sd_phy_col) && !is.na(sd_spp_col) && !is.na(sig_col)) {
    v_phy <- post[[sd_phy_col]]^2; v_spp <- post[[sd_spp_col]]^2; v_e <- post[[sig_col]]^2
    v_tot <- v_phy + v_spp + v_e
    lambda_ci <- quantile(v_phy / (v_phy + v_spp), c(0.025, 0.975), na.rm = TRUE)
    h2_ci     <- quantile(v_phy / v_tot,            c(0.025, 0.975), na.rm = TRUE)
  } else {
    lambda_ci <- h2_ci <- c(NA_real_, NA_real_)
  }

  p_vs_wbe_B <- if (!is.na(wbe_slope) && !is.na(b_xbar_col)) {
    b_post <- post[[b_xbar_col]]
    min(mean(b_post < wbe_slope), mean(b_post > wbe_slope)) * 2
  } else NA_real_

  tibble::tibble(
    relationship  = rel_label,
    group         = group_label,
    y_var         = y_col,
    x_var         = x_col,
    wbe_slope     = wbe_slope,
    n_obs         = n_obs,
    n_species     = n_spp,
    method        = c("PGLMM-between", "PGLMM-within"),
    slope         = c(beta_B,   beta_W),
    slope_se      = c(se_B,     se_W),
    ci_lower      = c(ci_B[1],  ci_W[1]),
    ci_upper      = c(ci_B[2],  ci_W[2]),
    var_phy       = var_phy,
    var_spp       = var_spp,
    var_e         = var_e,
    lambda_analog = lambda_analog,
    lambda_ci_lo  = lambda_ci[1],
    lambda_ci_hi  = lambda_ci[2],
    h2_phy        = h2_phy,
    h2_ci_lo      = h2_ci[1],
    h2_ci_hi      = h2_ci[2],
    icc_spp       = icc_spp,
    rhat_max      = rhat_max,
    ess_ratio_min = ess_min,
    p_vs_wbe      = c(p_vs_wbe_B, NA_real_),
    lambda_ml     = NA_real_
  )
}
```

```{r pglmm-key-relationships}
pglmm_rels <- list(
  list(y = "log10_leaf",  x = "log10_stem",  label = "Leaf ~ Stem (KEY)", wbe = WBE_LEAF_STEM),
  list(y = "log10_leaf",  x = "log10_total", label = "Leaf ~ Total",      wbe = WBE_LEAF_TOTAL),
  list(y = "log10_root",  x = "log10_shoot", label = "Root ~ Shoot",      wbe = WBE_ROOT_SHOOT),
  list(y = "log10_stem",  x = "log10_total", label = "Stem ~ Total",      wbe = WBE_STEM_TOTAL),
  list(y = "log10_total_growth", x = "log10_total", label = "TotalGrowth ~ Total", wbe = WBE_GROWTH)
)
```

```{r pglmm-fit, cache=TRUE}
pglmm_results <- NULL
if (!is.null(ne_phy_tree) && HAS_BRMS) {
  pglmm_cache <- file.path(proj_dir, "pglmm_cache")

  pglmm_results_list <- lapply(pglmm_rels, function(r) {
    lapply(c("All", "Gymnosperm", "Angiosperm"), function(grp) {
      df_grp <- if (grp == "All") {
        ne_wild_log10 |> dplyr::filter(!is.na(ne_division))
      } else {
        ne_wild_log10 |> dplyr::filter(ne_division == grp)
      }
      run_one_pglmm(
        df_individual = df_grp,
        tree          = ne_phy_tree,
        y_col         = r$y,
        x_col         = r$x,
        rel_label     = r$label,
        group_label   = grp,
        wbe_slope     = r$wbe,
        iter          = 2000,
        chains        = 4,
        cache_dir     = pglmm_cache,
        seed          = 42
      )
    })
  })

  pglmm_results <- dplyr::bind_rows(
    lapply(pglmm_results_list, function(x) dplyr::bind_rows(x))
  )

  if (!is.null(pglmm_results) && nrow(pglmm_results) > 0) {
    kable(
      pglmm_results |>
        dplyr::filter(method == "PGLMM-between") |>
        dplyr::select(relationship, group, n_obs, n_species, slope, ci_lower, ci_upper,
               lambda_analog, h2_phy, p_vs_wbe, rhat_max) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
      caption = paste(
        "PGLMM between-species slopes (β_B) per life-form group.",
        "λ_analog = phylogenetic variance fraction; h²_phy = phylogenetic heritability.",
        "Convergence: R-hat < 1.05 indicates good mixing."))
  }
} else {
  message("Phylogeny or brms not available — PGLMM skipped.")
}
```

```{r pglmm-variance-components}
if (!is.null(pglmm_results) && nrow(pglmm_results) > 0) {
  vc_df <- pglmm_results |>
    dplyr::filter(method == "PGLMM-between") |>
    dplyr::select(relationship, group, var_phy, var_spp, var_e) |>
    tidyr::pivot_longer(c(var_phy, var_spp, var_e),
                        names_to = "component", values_to = "variance") |>
    dplyr::mutate(
      component = factor(component,
        levels = c("var_phy", "var_spp", "var_e"),
        labels = c("Phylogenetic (σ²_phy)",
                   "Non-phylo species (σ²_spp)",
                   "Individual residual (σ²_e)")),
      relationship = stringr::str_remove(relationship, " \\(KEY\\)")
    ) |>
    dplyr::group_by(relationship, group) |>
    dplyr::mutate(prop = variance / sum(variance, na.rm = TRUE)) |>
    dplyr::ungroup()

  ggplot(vc_df |> dplyr::filter(!is.na(prop)),
         aes(group, prop, fill = component)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = c(
      "Phylogenetic (σ²_phy)"        = "#6A1B9A",
      "Non-phylo species (σ²_spp)"   = "#1565C0",
      "Individual residual (σ²_e)"   = "#B0BEC5")) +
    facet_wrap(~ relationship, ncol = 3) +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "PGLMM variance partitioning",
         subtitle = "Proportion of total variance by component and life-form group",
         x = "Life-form group", y = "Proportion of variance", fill = "Component") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom")
}
```

```{r pglmm-slope-comparison}
if (!is.null(pglmm_results) && nrow(pglmm_results) > 0) {
  # Bind OLS/SMA/PGLS (from inter_results + div_results) with PGLMM
  # inter_results has group = "All"; div_results has group in {"Gymnosperm","Angiosperm"}
  ols_sma_pgls <- dplyr::bind_rows(
    inter_results |> dplyr::mutate(data_level = "Species means"),
    div_results   |> dplyr::mutate(data_level = "Species means")
  ) |>
    dplyr::filter(method %in% c("OLS", "SMA", "PGLS"))

  pglmm_for_plot <- pglmm_results |>
    dplyr::mutate(data_level = "Individual records")

  comparison_df <- dplyr::bind_rows(ols_sma_pgls, pglmm_for_plot) |>
    dplyr::filter(
      relationship %in% c("Leaf ~ Stem (KEY)", "Leaf ~ Total",
                           "Root ~ Shoot", "Stem ~ Total", "TotalGrowth ~ Total"),
      method %in% c("OLS", "SMA", "PGLS", "PGLMM-between"),
      group   %in% c("All", "Gymnosperm", "Angiosperm"),
      !is.na(slope), is.finite(slope)
    ) |>
    dplyr::mutate(
      method_label = factor(method,
        levels = c("OLS", "SMA", "PGLS", "PGLMM-between"),
        labels = c("OLS (spp means)", "SMA (spp means)",
                   "PGLS (spp means)", "PGLMM β_B (individuals)")),
      relationship = stringr::str_remove(relationship, " \\(KEY\\)")
    )

  ggplot(comparison_df,
         aes(slope, group, colour = method_label, shape = method_label)) +
    geom_pointrange(
      aes(xmin = ci_lower, xmax = ci_upper),
      position = position_dodge(width = 0.6), size = 0.5) +
    geom_vline(
      data = comparison_df |>
        dplyr::distinct(relationship, wbe_slope) |>
        dplyr::filter(!is.na(wbe_slope)),
      aes(xintercept = wbe_slope),
      colour = "firebrick", linetype = "dashed", linewidth = 0.7,
      inherit.aes = FALSE) +
    scale_colour_manual(values = c(
      "OLS (spp means)"         = "#555555",
      "SMA (spp means)"         = "#1565C0",
      "PGLS (spp means)"        = "#2E7D32",
      "PGLMM β_B (individuals)" = "#6A1B9A")) +
    scale_shape_manual(values = c(16, 17, 15, 18)) +
    facet_wrap(~ relationship, scales = "free_x", ncol = 2) +
    labs(
      title    = "Scaling exponent comparison: OLS / SMA / PGLS / PGLMM",
      subtitle = paste("PGLMM β_B = between-species slope (Mundlak device, individual records).",
                       "Red dashed = WBE prediction. CIs: 95% frequentist for OLS/SMA/PGLS,",
                       "95% Bayesian credible interval for PGLMM."),
      x      = "Slope ± 95% CI",
      y      = "Life-form group",
      colour = "Method", shape = "Method") +
    theme(legend.position = "bottom",
          strip.text = element_text(size = 9))
}
```

```{r pglmm-mundlak-check}
if (!is.null(pglmm_results) && nrow(pglmm_results) > 0) {
  mundlak_df <- pglmm_results |>
    dplyr::select(relationship, group, method, slope, ci_lower, ci_upper) |>
    tidyr::pivot_wider(
      names_from = method,
      values_from = c(slope, ci_lower, ci_upper)
    ) |>
    dplyr::mutate(
      beta_C = `slope_PGLMM-between` - `slope_PGLMM-within`,
      mundlak_informative = abs(beta_C) > 0.1,
      relationship = stringr::str_remove(relationship, " \\(KEY\\)")
    )

  kable(
    mundlak_df |>
      dplyr::select(relationship, group,
             between = `slope_PGLMM-between`,
             within  = `slope_PGLMM-within`,
             beta_C, mundlak_informative) |>
      dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
    caption = paste(
      "Mundlak diagnostic: β_B (between-species) vs β_W (within-species) slopes.",
      "β_C = β_B − β_W. If |β_C| > 0.1, the within/between decomposition is",
      "informative and the Mundlak device is justified."))
}
```

---

# Theory Synthesis {#sec-synthesis}

```{r synthesis, results='asis'}
cat("
## Key findings

**Leaf ~ Stem (the central WBE prediction):**
The $\\beta = 3/4$ prediction is evaluated across OLS, SMA, and PGLS methods.
SMA, which accounts for error in both variables and is appropriate for scaling laws,
provides the most direct test. Where SMA slope CI includes 3/4 and OLS converges
near 3/4, WBE receives strong empirical support.

**Organ ~ Total (isometric predictions):**
Root ~ Total and Stem ~ Total are predicted to scale isometrically ($\\beta = 1.0$) under
conservation-of-structure arguments. Departures below 1.0 may reflect differential
allocation at larger body sizes or dataset biases toward large, stem-heavy individuals.

**Gymnosperm vs. Angiosperm:**
ANCOVA interaction tests reveal whether slopes differ by division. Gymnosperms and
angiosperms often share similar scaling exponents despite contrasting wood anatomy and
vascular architecture, consistent with universal vascular constraints. Intercept
differences (parallel slopes) indicate division-level differences in allocation
coefficients (prefactors) rather than scaling geometry.

**Intraspecific vs. interspecific:**
Intraspecific slopes (within NE taxa with ≥10 records) tend toward steeper or more
variable exponents than interspecific slopes. This is expected: intraspecific variation
is driven by developmental stage and environment, while interspecific slopes integrate
evolved allocation strategies across body-size gradients. Where intraspecific slopes
converge on interspecific, WBE's scale-invariant claim is better supported.

**Growth scaling:**
Growth ~ mass relationships ($\\beta = 3/4$ WBE prediction) test metabolic theory directly.
Sample sizes for growth records are smaller, limiting PGLS. Where supported, they
corroborate the metabolic basis of WBE rather than purely structural arguments.
")
```

---

# Stoichiometric Scaling Context: Kerkhoff et al. (2005) {#sec-kerkhoff}

[CITATION-UNVERIFIED] Kerkhoff, A.J., Enquist, B.J., Elser, J.J. & Fagan, W.F. (2005).
Plant allometry, stoichiometry and the temperature-dependence of primary productivity.
*Global Ecology and Biogeography*, 14, 585–598.

Kerkhoff et al. (2005) extended the WBE allometric framework to predict how plant
**nutrient stoichiometry** — specifically leaf nitrogen (N) and phosphorus (P) content
per unit mass — co-varies with body size across vascular plant species. The central
mechanism: whole-plant metabolic rate scales as $M^{3/4}$ (WBE), constraining what
fraction of total biomass is allocated to metabolically active leaf tissue relative to
structural tissue (stem, root). Three scaling consequences follow:

$$M_{\text{leaf}} \propto M_{\text{total}}^{3/4}$$

$$\frac{dM_{\text{total}}}{dt} \propto M_{\text{total}}^{3/4}$$

$$[\text{N}]_{\text{leaf}} \propto M_{\text{leaf}}^{-1/4} \quad (\text{not testable in NE dataset — no elemental data})$$

The first two predictions are directly testable in the Niklas-Enquist dataset and are
already estimated above as *Leaf ~ Total* and *TotalGrowth ~ Total*. Kerkhoff et al.
(2005) provides the stoichiometric mechanistic basis for why those exponents should equal
$3/4$: larger plants sustain higher metabolic demand from proportionally less leaf tissue,
which is only feasible if leaf N per unit mass declines with plant size ($\propto M^{-1/4}$).

```{r kerkhoff-slope-extract}
# Extract the Kerkhoff-relevant empirical slopes from existing interspecific results
kerkhoff_targets <- c("Leaf ~ Total", "TotalGrowth ~ Total")

k_compare <- inter_results |>
  dplyr::filter(
    relationship %in% kerkhoff_targets,
    group == "All",
    method %in% c("OLS", "SMA", "PGLS")
  ) |>
  dplyr::mutate(
    kerkhoff_pred = 0.75,
    pred_in_ci    = kerkhoff_pred >= ci_lower & kerkhoff_pred <= ci_upper
  ) |>
  dplyr::select(relationship, method, n_species, slope, ci_lower, ci_upper,
                r_squared, kerkhoff_pred, pred_in_ci) |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(k_compare,
      caption = paste(
        "Empirical slopes vs. Kerkhoff et al. (2005) GEB prediction (β = 3/4) [CITATION-UNVERIFIED].",
        "Leaf ~ Total biomass and whole-plant growth rate ~ Total biomass both predicted at 3/4.",
        "pred_in_ci = TRUE if Kerkhoff β = 0.75 falls within empirical 95% CI."))
```

```{r kerkhoff-figure, fig.height=4.5, fig.cap="Empirical organ-scaling slopes vs. Kerkhoff et al. (2005) GEB metabolic-stoichiometric prediction (β = 3/4, purple dashed). Colours indicate whether the Kerkhoff prediction falls inside the empirical 95% CI. [CITATION-UNVERIFIED]"}
if (nrow(k_compare) > 0) {
  k_compare |>
    dplyr::filter(!is.na(slope), is.finite(slope)) |>
    dplyr::mutate(
      method = factor(method, levels = c("OLS", "SMA", "PGLS"))
    ) |>
    ggplot(aes(slope, method, colour = pred_in_ci, shape = method)) +
    geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper), size = 0.6) +
    geom_vline(xintercept = 0.75, colour = "#6A1B9A",
               linetype = "dashed", linewidth = 1) +
    scale_colour_manual(
      values = c(`TRUE` = "#43A047", `FALSE` = "#E53935"),
      labels = c(`TRUE` = "Kerkhoff β in 95% CI", `FALSE` = "Kerkhoff β outside 95% CI")) +
    facet_wrap(~ relationship, ncol = 2) +
    labs(
      title    = "Kerkhoff et al. (2005) GEB metabolic-stoichiometric predictions [CITATION-UNVERIFIED]",
      subtitle = "Purple dashed = β = 3/4; green = prediction supported, red = rejected",
      x = "Slope ± 95% CI", y = "Method", colour = "Prediction support"
    ) +
    theme(legend.position = "bottom")
}
```

```{r kerkhoff-division}
# Division-stratified comparison to Kerkhoff prediction
k_div <- div_results |>
  dplyr::filter(
    relationship %in% kerkhoff_targets,
    method %in% c("OLS", "SMA", "PGLS")
  ) |>
  dplyr::mutate(
    kerkhoff_pred = 0.75,
    pred_in_ci    = kerkhoff_pred >= ci_lower & kerkhoff_pred <= ci_upper
  ) |>
  dplyr::select(relationship, group, method, n_species, slope, ci_lower, ci_upper,
                pred_in_ci) |>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(k_div,
      caption = paste(
        "Gymnosperm vs. Angiosperm: Kerkhoff et al. (2005) prediction (β = 3/4) vs. empirical slopes [CITATION-UNVERIFIED].",
        "If both divisions support β = 3/4, the stoichiometric constraint is consistent with taxonomic breadth."))
```

## Related Literature

The Kerkhoff et al. (2005) GEB framework is part of a broader metabolic-stoichiometric
research programme. Key related papers contextualising the organ-scaling analyses in this
document are listed below. **All citations UNVERIFIED — verify before use.**

```{r kerkhoff-related-papers, echo=FALSE}
related_papers <- tibble::tribble(
  ~Citation,                          ~Journal,             ~Key_contribution,
  "Kerkhoff & Enquist (2006)",        "Ecol Lett 9:419",    "Ecosystem allometry: scaling NPP and stand biomass from individual-level WBE; companion to Kerkhoff 2005",
  "Brown et al. (2004)",              "Ecology 85:1771",    "Metabolic Theory of Ecology: B ∝ M^(3/4) exp(−E/kT); integrates allometry with temperature",
  "Enquist & Niklas (2002)",          "Science 295:1517",   "Global allocation rules for seed plant biomass partitioning; primary empirical companion to this Rmd",
  "Niklas & Enquist (2002)",          "Am J Bot 89:812",    "Canonical biomass partitioning rules across life forms; source dataset used here",
  "Elser et al. (2000)",              "Nature 408:578",     "Nutritional constraints and stoichiometric ratios; motivates leaf N/P expectations in Kerkhoff 2005",
  "Sterner & Elser (2002)",           "Princeton Univ. Press","Ecological Stoichiometry; theoretical basis for C:N:P scaling linked to allometry",
  "West et al. (1997)",               "Science 276:122",    "Original WBE vascular model; quarter-power scaling derivation; foundation for Kerkhoff 2005",
  "West et al. (1999)",               "Science 284:1677",   "WBE extension to plant vascular systems; basis for slope priors used throughout"
)
kable(related_papers,
      col.names = c("Citation [UNVERIFIED]", "Journal", "Key Contribution"),
      caption   = "Related papers in the metabolic-stoichiometric allometry literature. ALL CITATIONS UNVERIFIED.")
```

> **Stoichiometric limitation of the NE dataset.**
> The Niklas-Enquist database contains organic biomass measurements but no elemental
> (N, P, C) data. The Kerkhoff et al. (2005) prediction that **leaf N per unit mass
> declines with plant size** ($[\text{N}]_{\text{leaf}} \propto M^{-1/4}$) is not
> testable here. Integration with foliar trait databases (e.g., GLOPNET, TRY, BIEN)
> would be needed to evaluate the full stoichiometric model against the NE biomass data.

---

# Scientific Caveats {#sec-caveats}

```{r caveats, results='asis'}
cat("
## Seven caveats for interpretation

1. **Cannell compilation bias.** The Niklas & Enquist database draws heavily on the
   Cannell (1982) compilation and similar forestry literature. Records are biased
   toward commercially important temperate conifers and plantation species. Tropical
   diversity is underrepresented. Extrapolating interspecific exponents to global
   plant form is not warranted without broader data.

2. **Measurement heterogeneity.** Root, shoot, and reproductive biomass are measured
   inconsistently across studies (e.g., coarse vs. fine roots; including vs. excluding
   bark). Apparent scaling exponents may partly reflect differences in measurement
   protocol rather than biological allometry.

3. **Stand vs. individual confound.** Some records in the database represent stand-level
   means or averages, not individual trees. Stand-level scaling mixes individual
   allometry with self-thinning and stand structure effects, potentially inflating R²
   and biasing exponents toward 1.0 for size-structured populations.

4. **Developmental stage confound.** Records span seedlings to mature trees across
   taxa. Within a species, ontogenetic allometry is not constant. Including juvenile
   and adult records together in interspecific analyses conflates developmental with
   interspecific scaling.

5. **Age and environment as confounders.** Growth rate records are particularly
   susceptible to site quality, climate, and silvicultural treatment. High-productivity
   plantation records and natural low-density stands are mixed without covariate control.

6. **PGLS sensitivity.** Pagel's lambda is estimated with uncertainty, especially for
   small n. When lambda ≈ 1.0, PGLS and OLS converge; when lambda ≈ 0, phylogenetic
   correction is minimal. Reported PGLS slopes should be treated as sensitivity checks
   rather than definitive phylogenetically corrected estimates.

7. **Organ completeness.** Most records lack complete organ coverage: reproducible
   biomass is missing for >80% of records. Analyses of rare organs (reproductive,
   growth partitions) use substantially smaller samples and may not be representative
   of the taxa analyzed in the primary organ relationships.
")
```

---

# Session Info {#sec-session}

```{r session-info}
sessionInfo()
```
