---
title: "Niklas & Enquist Organ-Mass Scaling: Theory Testing Across Biomass Partitions"
subtitle: "OLS, SMA, PGLS — Interspecific, Gymnosperm/Angiosperm, Family, and Intraspecific"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(readr);  library(dplyr);  library(ggplot2); library(scales)
  library(tidyr);  library(knitr);  library(smatr);   library(purrr)
  library(stringr); library(tibble); library(caper);  library(ape)
  library(phytools)
})

HAS_PHYLOMAKER <- requireNamespace("V.PhyloMaker2", quietly = TRUE)

# Explicit namespace bindings to prevent masking conflicts
map       <- purrr::map
map_dbl   <- purrr::map_dbl
map_df    <- purrr::map_df
map_dfr   <- purrr::map_dfr
select    <- dplyr::select
filter    <- dplyr::filter
mutate    <- dplyr::mutate
rename    <- dplyr::rename
arrange   <- dplyr::arrange
group_by  <- dplyr::group_by
summarise <- dplyr::summarise
summarize <- dplyr::summarize
slice     <- dplyr::slice

set.seed(42)
theme_set(theme_bw(base_size = 12))

proj_dir <- tryCatch(dirname(knitr::current_input(dir = TRUE)),
                     error = function(e) getwd())

# WBE / theory slope priors
# West, Brown & Enquist (1997) Science 276:122; (1999) Science 284:1677
# Enquist & Niklas (2002) Nature 410:655–660
WBE_LEAF_TOTAL  <- 3/4   # leaf ~ total metabolic scaling
WBE_LEAF_STEM   <- 3/4   # leaf ~ stem (KEY WBE prediction)
WBE_LEAF_ROOT   <- 3/4
WBE_ROOT_TOTAL  <- 1.0   # isometric (conservation of structure)
WBE_STEM_TOTAL  <- 1.0
WBE_SHOOT_TOTAL <- 1.0
WBE_ROOT_STEM   <- 1.0
WBE_ROOT_SHOOT  <- 1.0
WBE_TOTAL_H     <- 8/3   # total biomass ~ height^(8/3): derived from vascular geometry
WBE_GROWTH      <- 3/4   # growth rate ~ mass^(3/4)
```

---

# Data Loading {#sec-data}

```{r load-data}
nik_col_names <- c(
  "taxa", "citation", "footnotes", "record_num", "age_yr", "log_age",
  "height_m", "log_height", "root_biomass_kg", "log_root", "stem_biomass_kg",
  "log_stem", "leaf_biomass_kg", "log_leaf", "shoot_biomass_kg", "log_shoot",
  "repro_biomass_kg", "log_repro", "total_biomass_kg", "log_total",
  "stem_growth_kgyr", "log_stem_growth", "leaf_growth_kgyr", "log_leaf_growth",
  "root_growth_kgyr", "log_root_growth", "total_growth_kgyr", "log_total_growth"
)

nik_enq <- read_csv(
  file.path(proj_dir, "data/raw/niklas_enquist/niklas_enquist_biomass_20040122.csv"),
  skip = 2, col_names = nik_col_names,
  locale = locale(encoding = "latin1"), show_col_types = FALSE) |>
  mutate(across(-c(taxa, citation, footnotes), as.numeric)) |>
  mutate(across(where(is.numeric), ~ ifelse(. == -999, NA_real_, .)))

cat(sprintf("Total records loaded: %d\n", nrow(nik_enq)))
```

---

# Record Classification {#sec-classify}

```{r classify-records}
# Crop/cultivar genus/species detection (annual herbaceous crops)
crop_pattern <- paste0(
  "(?i)(wheat|triticum|barley|hordeum|maize|zea mays|corn|oryza|rice|",
  "pea\\b|pisum|bean\\b|soybean|glycine max|phaseolus|cotton|gossypium|",
  "sunflower|helianthus annuus|millet|pearl millet|pennisetum glaucum|",
  "capsella|cucumber|cucumis|dolichos|lablab|vigna|brassica oleracea|",
  "lolium|medicago sativa|vicia faba)")

community_pattern <- paste0(
  "(?i)(community|stand|mixed|plot|forest|woodland|grassland|",
  "whole-plant community|with a few|average)")

nik_enq <- nik_enq |>
  mutate(
    flag_crop      = str_detect(taxa, crop_pattern),
    flag_community = str_detect(taxa, community_pattern),
    pgls_eligible  = !flag_community & !is.na(taxa)
  )

# Taxonomy-based record_type (assigned after parsing; prelim based on flags)
nik_enq <- nik_enq |>
  mutate(record_type_prelim = case_when(
    flag_crop      ~ "Crop/herb",
    flag_community ~ "Community/mixed",
    TRUE           ~ "Unknown — see taxonomy"
  ))

cat(sprintf("Crop records: %d\nCommunity/stand records: %d\nRemaining: %d\n",
            sum(nik_enq$flag_crop),
            sum(nik_enq$flag_community),
            sum(!nik_enq$flag_crop & !nik_enq$flag_community)))
```

---

# Data Overview {#sec-overview}

```{r overview-coverage}
# Coverage of each measured variable
organ_vars <- c("root_biomass_kg", "stem_biomass_kg", "leaf_biomass_kg",
                "shoot_biomass_kg", "repro_biomass_kg", "total_biomass_kg",
                "height_m", "age_yr",
                "stem_growth_kgyr", "leaf_growth_kgyr",
                "root_growth_kgyr", "total_growth_kgyr")

coverage_tbl <- tibble(
  variable = organ_vars,
  n_obs    = map_int(organ_vars, ~ sum(!is.na(nik_enq[[.x]]))),
  pct_obs  = map_dbl(organ_vars, ~ 100 * mean(!is.na(nik_enq[[.x]])))
)

kable(coverage_tbl, digits = 1,
      caption = "Variable coverage across all 2,626 records (N obs and %).")
```

```{r overview-type-chart, fig.height=4}
type_counts <- nik_enq |>
  count(record_type_prelim, name = "n") |>
  mutate(record_type_prelim = reorder(record_type_prelim, n))

ggplot(type_counts, aes(record_type_prelim, n, fill = record_type_prelim)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Preliminary record classification",
       subtitle = "Woody wild species will be confirmed after taxonomy parsing",
       x = NULL, y = "N records")
```

---

# Taxonomy {#sec-taxonomy}

```{r ne-genus-lookup}
ne_genus_lookup <- tibble::tribble(
  ~genus,           ~ne_family,            ~ne_division,
  "Abies",          "Pinaceae",            "Gymnosperm",
  "Picea",          "Pinaceae",            "Gymnosperm",
  "Pinus",          "Pinaceae",            "Gymnosperm",
  "Pseudotsuga",    "Pinaceae",            "Gymnosperm",
  "Larix",          "Pinaceae",            "Gymnosperm",
  "Tsuga",          "Pinaceae",            "Gymnosperm",
  "Cedrus",         "Pinaceae",            "Gymnosperm",
  "Juniperus",      "Cupressaceae",        "Gymnosperm",
  "Thuja",          "Cupressaceae",        "Gymnosperm",
  "Chamaecyparis",  "Cupressaceae",        "Gymnosperm",
  "Cryptomeria",    "Cupressaceae",        "Gymnosperm",
  "Podocarpus",     "Podocarpaceae",       "Gymnosperm",
  "Araucaria",      "Araucariaceae",       "Gymnosperm",
  "Taxus",          "Taxaceae",            "Gymnosperm",
  "Ginkgo",         "Ginkgoaceae",         "Gymnosperm",
  "Sequoia",        "Cupressaceae",        "Gymnosperm",
  "Sequoiadendron", "Cupressaceae",        "Gymnosperm",
  "Metasequoia",    "Cupressaceae",        "Gymnosperm",
  "Cupressus",      "Cupressaceae",        "Gymnosperm",
  "Quercus",        "Fagaceae",            "Angiosperm",
  "Fagus",          "Fagaceae",            "Angiosperm",
  "Nothofagus",     "Nothofagaceae",       "Angiosperm",
  "Betula",         "Betulaceae",          "Angiosperm",
  "Alnus",          "Betulaceae",          "Angiosperm",
  "Populus",        "Salicaceae",          "Angiosperm",
  "Salix",          "Salicaceae",          "Angiosperm",
  "Eucalyptus",     "Myrtaceae",           "Angiosperm",
  "Acacia",         "Fabaceae",            "Angiosperm",
  "Robinia",        "Fabaceae",            "Angiosperm",
  "Tilia",          "Malvaceae",           "Angiosperm",
  "Fraxinus",       "Oleaceae",            "Angiosperm",
  "Acer",           "Sapindaceae",         "Angiosperm",
  "Juglans",        "Juglandaceae",        "Angiosperm",
  "Carya",          "Juglandaceae",        "Angiosperm",
  "Magnolia",       "Magnoliaceae",        "Angiosperm",
  "Shorea",         "Dipterocarpaceae",    "Angiosperm",
  "Rhizophora",     "Rhizophoraceae",      "Angiosperm",
  "Prunus",         "Rosaceae",            "Angiosperm",
  "Cornus",         "Cornaceae",           "Angiosperm",
  "Liriodendron",   "Magnoliaceae",        "Angiosperm",
  "Castanea",       "Fagaceae",            "Angiosperm",
  "Corylus",        "Betulaceae",          "Angiosperm",
  "Carpinus",       "Betulaceae",          "Angiosperm",
  "Ulmus",          "Ulmaceae",            "Angiosperm",
  "Celtis",         "Cannabaceae",         "Angiosperm",
  "Liquidambar",    "Altingiaceae",        "Angiosperm",
  "Platanus",       "Platanaceae",         "Angiosperm",
  "Nyssa",          "Nyssaceae",           "Angiosperm",
  "Betula",         "Betulaceae",          "Angiosperm",
  "Cinnamomum",     "Lauraceae",           "Angiosperm",
  "Swietenia",      "Meliaceae",           "Angiosperm",
  "Tectona",        "Lamiaceae",           "Angiosperm",
  "Dipterocarpus",  "Dipterocarpaceae",    "Angiosperm",
  "Hopea",          "Dipterocarpaceae",    "Angiosperm",
  "Cedrela",        "Meliaceae",           "Angiosperm",
  "Terminalia",     "Combretaceae",        "Angiosperm",
  "Virola",         "Myristicaceae",       "Angiosperm",
  "Heliocarpus",    "Malvaceae",           "Angiosperm"
) |> dplyr::distinct(genus, .keep_all = TRUE)
```

```{r ne-parse-taxa}
ne_parse_taxa <- function(taxa_vec) {
  cleaned <- taxa_vec |>
    stringr::str_trim() |> stringr::str_squish() |>
    stringr::str_remove_all("\\s*\\([^)]*\\)") |>
    stringr::str_remove("\\s*\\[[^\\]]*\\]$") |>
    stringr::str_remove("\\s+(ssp\\.|subsp\\.|var\\.|f\\.|cv\\.)\\s+\\S+.*$") |>
    stringr::str_remove("(?i)\\s+sp+\\.?\\s*$") |>
    stringr::str_trim()
  genus_vec <- stringr::word(cleaned, 1)
  second_wd <- stringr::word(cleaned, 2)
  has_epi   <- !is.na(second_wd) &
    stringr::str_detect(second_wd, "^[a-z][a-z\\-]+$") &
    !second_wd %in% c("sp","spp","sp.","spp.")
  binomial <- dplyr::if_else(has_epi, paste(genus_vec, second_wd), NA_character_)
  flag_community <- stringr::str_detect(
    taxa_vec, stringr::regex("community|stand|mixed|plot|forest|woodland", ignore_case = TRUE))
  dplyr::tibble(
    taxa_orig      = taxa_vec,
    taxa_cleaned   = cleaned,
    binomial       = binomial,
    genus          = genus_vec,
    flag_sp_only   = !has_epi & !is.na(genus_vec),
    flag_community = flag_community,
    flag_uncertain = is.na(binomial) | flag_community
  )
}

ne_parsed <- ne_parse_taxa(nik_enq$taxa)

ne_annotated <- nik_enq |>
  bind_cols(ne_parsed |> select(-taxa_orig, -flag_community)) |>
  left_join(ne_genus_lookup, by = "genus") |>
  mutate(
    record_type = case_when(
      flag_crop      ~ "Crop/herb",
      flag_community ~ "Community/mixed",
      ne_division == "Gymnosperm" ~ "Gymnosperm",
      ne_division == "Angiosperm" ~ "Angiosperm",
      TRUE ~ "Unknown"
    )
  )

cat("Record type breakdown:\n")
print(ne_annotated |> count(record_type, sort = TRUE))
```

```{r species-means}
# Primary dataset: wild woody species only (exclude crops and community stands)
ne_wild <- ne_annotated |>
  dplyr::filter(!flag_crop, !flag_community)

# Species means: use records with valid binomial, known biomass > 0
ne_sp_means <- ne_wild |>
  dplyr::filter(!flag_uncertain, !is.na(binomial), total_biomass_kg > 0) |>
  dplyr::mutate(
    # Compute non-leaf at record level BEFORE log-transform (Jensen's inequality)
    nonleaf_kg  = dplyr::if_else(!is.na(leaf_biomass_kg) & leaf_biomass_kg > 0 &
                                   total_biomass_kg > leaf_biomass_kg,
                                 total_biomass_kg - leaf_biomass_kg, NA_real_),
    log10_total  = log10(total_biomass_kg),
    log10_stem   = log10(stem_biomass_kg),
    log10_leaf   = log10(leaf_biomass_kg),
    log10_root   = log10(root_biomass_kg),
    log10_shoot  = log10(shoot_biomass_kg),
    log10_repro  = log10(repro_biomass_kg),
    log10_height = log10(height_m),
    log10_nonleaf = log10(nonleaf_kg),
    log10_total_growth = log10(total_growth_kgyr),
    log10_stem_growth  = log10(stem_growth_kgyr),
    log10_leaf_growth  = log10(leaf_growth_kgyr),
    log10_root_growth  = log10(root_growth_kgyr)
  ) |>
  dplyr::mutate(across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .))) |>
  dplyr::group_by(binomial, ne_family, ne_division) |>
  dplyr::summarise(
    across(starts_with("log10_"), ~ mean(.x, na.rm = TRUE)),
    n_records = n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .))) |>
  dplyr::mutate(
    sp_tree = gsub(" ", "_", binomial)
  )

cat(sprintf("Species means: %d species, %d families\n",
            nrow(ne_sp_means), n_distinct(ne_sp_means$ne_family, na.rm = TRUE)))
```

```{r taxonomy-summary-plot, fig.height=4}
ne_sp_means |>
  dplyr::filter(!is.na(ne_division)) |>
  dplyr::count(ne_family, ne_division) |>
  dplyr::arrange(desc(n)) |>
  dplyr::slice_head(n = 20) |>
  dplyr::mutate(ne_family = reorder(ne_family, n)) |>
  ggplot(aes(ne_family, n, fill = ne_division)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(Gymnosperm = "#E65100", Angiosperm = "#1976D2")) +
  labs(title = "Top 20 families in species-mean dataset",
       x = NULL, y = "N species", fill = "Division")
```

---

# Phylogeny {#sec-phylogeny}

```{r build-phylogeny}
ne_phy_tree <- NULL

if (HAS_PHYLOMAKER && nrow(ne_sp_means) > 10) {
  sp_for_tree <- ne_sp_means |>
    dplyr::filter(!is.na(ne_family), !is.na(binomial)) |>
    dplyr::mutate(genus_col = stringr::word(binomial, 1)) |>
    dplyr::distinct(binomial, genus_col, ne_family) |>
    dplyr::rename(species = binomial, genus = genus_col, family = ne_family)

  ne_phy_tree <- tryCatch({
    res <- V.PhyloMaker2::phylo.maker(
      sp.list = data.frame(
        species = gsub(" ", "_", sp_for_tree$species),
        genus   = sp_for_tree$genus,
        family  = sp_for_tree$family),
      tree    = V.PhyloMaker2::GBOTB.extended.TPL,
      nodes   = V.PhyloMaker2::nodes.info.1,
      scenario = "S3")
    if (!is.null(res$phylo)) {
      tr <- res$phylo
      tr$node.label <- NULL
      tr
    } else NULL
  }, error = function(e) { message("V.PhyloMaker2 failed: ", e$message); NULL })

  if (!is.null(ne_phy_tree)) {
    n_matched <- sum(ne_sp_means$sp_tree %in% ne_phy_tree$tip.label)
    cat(sprintf("Phylogeny: %d tips; %d/%d species matched\n",
                length(ne_phy_tree$tip.label), n_matched, nrow(ne_sp_means)))
  }
} else {
  message("V.PhyloMaker2 not available — PGLS analyses will be skipped.")
}
```

---

# Phylogenetic Signal {#sec-phylosignal}

```{r phylo-signal}
if (!is.null(ne_phy_tree)) {
  signal_vars <- c("log10_leaf", "log10_stem", "log10_root", "log10_shoot",
                   "log10_total", "log10_height", "log10_nonleaf")

  signal_results <- map_dfr(signal_vars, function(v) {
    sub <- ne_sp_means |>
      dplyr::filter(!is.na(.data[[v]]), sp_tree %in% ne_phy_tree$tip.label) |>
      dplyr::distinct(sp_tree, .keep_all = TRUE)
    if (nrow(sub) < 10) return(NULL)
    tr2 <- ape::keep.tip(ne_phy_tree, sub$sp_tree)
    x   <- setNames(sub[[v]], sub$sp_tree)
    lam <- tryCatch(phytools::phylosig(tr2, x, method = "lambda", test = TRUE),
                    error = function(e) NULL)
    Kst <- tryCatch(phytools::phylosig(tr2, x, method = "K", test = TRUE),
                    error = function(e) NULL)
    tibble(
      trait   = v,
      n       = nrow(sub),
      lambda  = if (!is.null(lam)) round(lam$lambda, 3) else NA_real_,
      lam_p   = if (!is.null(lam)) round(lam$P, 4)     else NA_real_,
      K       = if (!is.null(Kst)) round(Kst$K, 3)     else NA_real_,
      K_p     = if (!is.null(Kst)) round(Kst$P, 4)     else NA_real_
    )
  })

  kable(signal_results,
        caption = "Phylogenetic signal: Pagel's lambda and Blomberg's K for each organ trait.")
} else {
  message("Phylogeny not available — phylogenetic signal analysis skipped.")
}
```

---

# run_one_allom Helper {#sec-helper}

```{r run-one-allom}
run_one_allom <- function(df, tree, y_col, x_col, rel_label,
                           wbe_slope = NA_real_, group_label = "All") {
  sub <- df |>
    dplyr::filter(is.finite(.data[[y_col]]), is.finite(.data[[x_col]]))
  if (!is.null(tree)) sub <- sub |> dplyr::filter(sp_tree %in% tree$tip.label)
  n <- nrow(sub)
  if (n < 8) {
    message(sprintf("Skipping %s ~ %s [%s] — n=%d", y_col, x_col, group_label, n))
    return(NULL)
  }

  ols_f  <- lm(as.formula(paste(y_col, "~", x_col)), data = sub)
  ols_s  <- summary(ols_f)
  ols_b  <- coef(ols_f)[2]
  ols_se <- coef(ols_s)[2, 2]
  ols_ci <- confint(ols_f)[2, ]
  ols_r2 <- ols_s$r.squared
  ols_p  <- if (!is.na(wbe_slope))
    2 * pt(abs((ols_b - wbe_slope) / ols_se), n - 2, lower.tail = FALSE) else NA_real_

  sma_f <- tryCatch(
    smatr::sma(as.formula(paste(y_col, "~", x_col)), data = sub),
    error = function(e) NULL)
  if (!is.null(sma_f)) {
    sma_b    <- sma_f$coef[[1]][2, 1]
    sma_cilo <- sma_f$coef[[1]][2, 2]
    sma_cihi <- sma_f$coef[[1]][2, 3]
    sma_r2   <- sma_f$r2[[1]]
    sma_p    <- if (!is.na(wbe_slope))
      tryCatch({
        st <- smatr::slope.test(sub[[y_col]], sub[[x_col]],
                                test.value = wbe_slope, method = "SMA")
        st$p
      }, error = function(e) NA_real_) else NA_real_
  } else {
    sma_b <- sma_cilo <- sma_cihi <- sma_r2 <- sma_p <- NA_real_
  }

  pgls_b <- pgls_se <- pgls_r2 <- pgls_lam <- pgls_p <-
    pgls_cilo <- pgls_cihi <- NA_real_
  if (!is.null(tree) && n >= 8) {
    tr2 <- ape::keep.tip(tree, sub$sp_tree)
    tr2$node.label <- NULL
    # Deduplicate by sp_tree before rowname assignment (prevents crash on duplicates)
    df_p <- as.data.frame(dplyr::distinct(sub, sp_tree, .keep_all = TRUE))
    rownames(df_p) <- df_p$sp_tree
    cd <- tryCatch(
      caper::comparative.data(phy = tr2, data = df_p, names.col = "sp_tree",
                               vcv = TRUE, warn.dropped = FALSE),
      error = function(e) NULL)
    if (!is.null(cd)) {
      pg_fixed_lambda <- FALSE
      pg <- tryCatch(
        caper::pgls(as.formula(paste(y_col, "~", x_col)),
                    data = cd, lambda = "ML", bounds = list(lambda = c(1e-6, 1))),
        error = function(e) {
          message(sprintf("PGLS ML failed [%s/%s]: %s — retrying lambda=1",
                          rel_label, group_label, e$message))
          pg_fixed_lambda <<- TRUE
          tryCatch(caper::pgls(as.formula(paste(y_col, "~", x_col)),
                               data = cd, lambda = 1),
                   error = function(e2) NULL)
        })
      if (!is.null(pg)) {
        pg_s    <- summary(pg)
        pg_df   <- tryCatch(pg_s$df[2], error = function(e) n - 2L)
        pgls_b  <- coef(pg)[2]
        pgls_se <- coef(pg_s)[2, 2]
        pgls_t  <- qt(0.975, df = pg_df)
        pgls_ci <- pgls_b + c(-1, 1) * pgls_t * pgls_se
        pgls_cilo <- pgls_ci[1]; pgls_cihi <- pgls_ci[2]
        pgls_r2  <- pg_s$r.squared
        pgls_lam <- tryCatch(pg$param["lambda"], error = function(e) NA_real_)
        pgls_p   <- if (!is.na(wbe_slope))
          2 * pt(abs((pgls_b - wbe_slope) / pgls_se), df = pg_df, lower.tail = FALSE)
        else NA_real_
      }
    }
  }

  tibble::tibble(
    relationship = rel_label, group = group_label,
    y_var = y_col, x_var = x_col,
    wbe_slope = wbe_slope, n_species = n,
    method     = c("OLS", "SMA", "PGLS"),
    slope      = c(ols_b,    sma_b,    pgls_b),
    slope_se   = c(ols_se,   NA_real_, pgls_se),
    ci_lower   = c(ols_ci[1], sma_cilo, pgls_cilo),
    ci_upper   = c(ols_ci[2], sma_cihi, pgls_cihi),
    r_squared  = c(ols_r2,   sma_r2,   pgls_r2),
    lambda_ml  = c(NA_real_, NA_real_, pgls_lam),
    p_vs_wbe   = c(ols_p,    sma_p,    pgls_p)
  )
}
```

---

# Interspecific Scaling — All Relationships {#sec-interspecific}

```{r define-relationships}
# All organ relationships to analyze with their WBE priors
relationships <- list(
  # Organ ~ Total biomass
  list(y = "log10_leaf",  x = "log10_total",  label = "Leaf ~ Total",        wbe = WBE_LEAF_TOTAL),
  list(y = "log10_root",  x = "log10_total",  label = "Root ~ Total",        wbe = WBE_ROOT_TOTAL),
  list(y = "log10_stem",  x = "log10_total",  label = "Stem ~ Total",        wbe = WBE_STEM_TOTAL),
  list(y = "log10_shoot", x = "log10_total",  label = "Shoot ~ Total",       wbe = WBE_SHOOT_TOTAL),
  list(y = "log10_repro", x = "log10_total",  label = "Repro ~ Total",       wbe = NA_real_),
  # Organ ~ Organ
  list(y = "log10_leaf",  x = "log10_stem",   label = "Leaf ~ Stem (KEY)",   wbe = WBE_LEAF_STEM),
  list(y = "log10_leaf",  x = "log10_root",   label = "Leaf ~ Root",         wbe = WBE_LEAF_ROOT),
  list(y = "log10_root",  x = "log10_stem",   label = "Root ~ Stem",         wbe = WBE_ROOT_STEM),
  list(y = "log10_root",  x = "log10_shoot",  label = "Root ~ Shoot",        wbe = WBE_ROOT_SHOOT),
  list(y = "log10_leaf",  x = "log10_nonleaf",label = "Leaf ~ Non-leaf",     wbe = WBE_LEAF_TOTAL),
  # Height relationships
  list(y = "log10_total", x = "log10_height", label = "Total ~ Height",      wbe = WBE_TOTAL_H),
  list(y = "log10_stem",  x = "log10_height", label = "Stem ~ Height",       wbe = NA_real_),
  list(y = "log10_leaf",  x = "log10_height", label = "Leaf ~ Height",       wbe = NA_real_),
  # Growth
  list(y = "log10_total_growth", x = "log10_total", label = "TotalGrowth ~ Total", wbe = WBE_GROWTH),
  list(y = "log10_stem_growth",  x = "log10_stem",  label = "StemGrowth ~ Stem",   wbe = WBE_GROWTH),
  list(y = "log10_leaf_growth",  x = "log10_leaf",  label = "LeafGrowth ~ Leaf",   wbe = WBE_GROWTH),
  list(y = "log10_root_growth",  x = "log10_root",  label = "RootGrowth ~ Root",   wbe = WBE_GROWTH)
)
```

```{r interspecific-all, cache=TRUE}
inter_results <- map_dfr(relationships, function(r) {
  run_one_allom(
    df          = ne_sp_means,
    tree        = ne_phy_tree,
    y_col       = r$y,
    x_col       = r$x,
    rel_label   = r$label,
    wbe_slope   = r$wbe,
    group_label = "All"
  )
})

kable(inter_results |>
        dplyr::select(relationship, group, method, n_species, slope, ci_lower, ci_upper,
               r_squared, lambda_ml, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
      caption = "Interspecific organ scaling: OLS, SMA, PGLS across all wild species.")
```

```{r interspecific-forest, fig.height=9}
inter_results |>
  dplyr::filter(!is.na(slope), is.finite(slope)) |>
  dplyr::mutate(
    relationship = factor(relationship, levels = unique(inter_results$relationship)),
    method = factor(method, levels = c("OLS", "SMA", "PGLS"))
  ) |>
  ggplot(aes(slope, relationship, colour = method, shape = method)) +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                  position = position_dodge(width = 0.5), size = 0.5) +
  geom_vline(data = inter_results |>
               dplyr::filter(!is.na(wbe_slope)) |>
               dplyr::distinct(relationship, wbe_slope),
             aes(xintercept = wbe_slope), colour = "red",
             linetype = "dashed", linewidth = 0.6) +
  scale_colour_manual(values = c(OLS = "#333333", SMA = "#1565C0", PGLS = "#2E7D32")) +
  facet_wrap(~ "All species") +
  labs(title = "Interspecific organ scaling — all relationships",
       subtitle = "Red dashed = WBE prediction; points jittered by method",
       x = "Slope ± 95% CI", y = NULL, colour = "Method", shape = "Method") +
  theme(axis.text.y = element_text(size = 9))
```

```{r leaf-stem-scatter, fig.height=5}
ne_sp_means |>
  dplyr::filter(is.finite(log10_leaf), is.finite(log10_stem)) |>
  ggplot(aes(log10_stem, log10_leaf, colour = ne_division)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, aes(group = 1),
              colour = "black", linewidth = 1) +
  geom_abline(slope = WBE_LEAF_STEM,
              intercept = {
                sub <- ne_sp_means |> dplyr::filter(is.finite(log10_leaf), is.finite(log10_stem))
                mean(sub$log10_leaf, na.rm = TRUE) - WBE_LEAF_STEM * mean(sub$log10_stem, na.rm = TRUE)
              },
              colour = "red", linetype = "dashed", linewidth = 1) +
  annotation_logticks(sides = "bl") +
  scale_colour_manual(values = c(Gymnosperm = "#E65100", Angiosperm = "#1976D2"),
                      na.value = "grey60") +
  labs(title = "Leaf ~ Stem biomass (KEY WBE relationship)",
       subtitle = "Black=OLS, Red-dashed=WBE β=3/4",
       x = "log10(Stem biomass kg)", y = "log10(Leaf biomass kg)",
       colour = "Division")
```

---

# Gymnosperm vs Angiosperm {#sec-division}

```{r division-stratified, cache=TRUE}
div_results <- map_dfr(c("Gymnosperm", "Angiosperm"), function(div) {
  sp_div <- ne_sp_means |> dplyr::filter(ne_division == div)
  map_dfr(relationships, function(r) {
    run_one_allom(
      df          = sp_div,
      tree        = ne_phy_tree,
      y_col       = r$y,
      x_col       = r$x,
      rel_label   = r$label,
      wbe_slope   = r$wbe,
      group_label = div
    )
  })
})

kable(div_results |>
        dplyr::select(relationship, group, method, n_species, slope, ci_lower, ci_upper,
               r_squared, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
      caption = "Division-stratified organ scaling: Gymnosperm vs. Angiosperm.")
```

```{r ancova-slope-homogeneity}
# ANCOVA slope homogeneity: test division × x_col interaction for key relationships
key_rels <- list(
  list(y = "log10_leaf",  x = "log10_stem",  label = "Leaf ~ Stem"),
  list(y = "log10_root",  x = "log10_shoot", label = "Root ~ Shoot"),
  list(y = "log10_stem",  x = "log10_total", label = "Stem ~ Total"),
  list(y = "log10_leaf",  x = "log10_total", label = "Leaf ~ Total")
)

ancova_results <- map_dfr(key_rels, function(r) {
  sub <- ne_sp_means |>
    dplyr::filter(
      is.finite(.data[[r$y]]), is.finite(.data[[r$x]]),
      ne_division %in% c("Gymnosperm", "Angiosperm")
    )
  if (nrow(sub) < 15) return(NULL)
  m_full <- lm(as.formula(paste(r$y, "~ ne_division *", r$x)), data = sub)
  m_para <- lm(as.formula(paste(r$y, "~ ne_division +", r$x)), data = sub)
  av <- anova(m_para, m_full)
  tibble(
    relationship     = r$label,
    n                = nrow(sub),
    interaction_F    = round(av$F[2], 3),
    interaction_p    = round(av$`Pr(>F)`[2], 4),
    slopes_homogeneous = av$`Pr(>F)`[2] > 0.05
  )
})

kable(ancova_results,
      caption = "ANCOVA test of slope homogeneity between Gymnosperm and Angiosperm.")
```

```{r division-forest, fig.height=8}
div_results |>
  dplyr::filter(method == "SMA", !is.na(slope), is.finite(slope)) |>
  dplyr::mutate(relationship = factor(relationship, levels = unique(inter_results$relationship))) |>
  ggplot(aes(slope, relationship, colour = group)) +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                  position = position_dodge(width = 0.5), size = 0.5) +
  geom_vline(data = div_results |>
               dplyr::filter(!is.na(wbe_slope)) |>
               dplyr::distinct(relationship, wbe_slope),
             aes(xintercept = wbe_slope), colour = "red",
             linetype = "dashed", linewidth = 0.6) +
  scale_colour_manual(values = c(Gymnosperm = "#E65100", Angiosperm = "#1976D2")) +
  labs(title = "SMA slopes by division — Gymnosperm vs. Angiosperm",
       subtitle = "Red dashed = WBE prediction",
       x = "SMA slope ± 95% CI", y = NULL, colour = "Division") +
  theme(axis.text.y = element_text(size = 9))
```

---

# Family-Level Analysis {#sec-family}

```{r family-level, cache=TRUE, fig.height=8}
# Families with ≥5 species in species-mean dataset
family_5sp <- ne_sp_means |>
  dplyr::filter(!is.na(ne_family)) |>
  dplyr::count(ne_family) |>
  dplyr::filter(n >= 5) |>
  dplyr::pull(ne_family)

cat(sprintf("Families with ≥5 species: %d\n", length(family_5sp)))

key_rels_fam <- list(
  list(y = "log10_leaf",  x = "log10_stem",  label = "Leaf ~ Stem",  wbe = WBE_LEAF_STEM),
  list(y = "log10_root",  x = "log10_shoot", label = "Root ~ Shoot", wbe = WBE_ROOT_SHOOT),
  list(y = "log10_stem",  x = "log10_total", label = "Stem ~ Total", wbe = WBE_STEM_TOTAL),
  list(y = "log10_leaf",  x = "log10_total", label = "Leaf ~ Total", wbe = WBE_LEAF_TOTAL)
)

fam_results <- map_dfr(key_rels_fam, function(r) {
  map_dfr(family_5sp, function(fam) {
    sp_fam <- ne_sp_means |> dplyr::filter(ne_family == fam)
    # PGLS only if ≥10 species matched to tree
    use_tree <- if (!is.null(ne_phy_tree) &&
                    sum(sp_fam$sp_tree %in% ne_phy_tree$tip.label) >= 10)
      ne_phy_tree else NULL
    run_one_allom(
      df          = sp_fam,
      tree        = use_tree,
      y_col       = r$y,
      x_col       = r$x,
      rel_label   = r$label,
      wbe_slope   = r$wbe,
      group_label = fam
    )
  })
})

kable(fam_results |>
        dplyr::filter(method == "OLS") |>
        dplyr::select(relationship, group, n_species, slope, ci_lower, ci_upper,
               r_squared, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))) |>
        dplyr::arrange(relationship, group),
      caption = "Family-level OLS slopes for key organ relationships (families ≥5 species).")
```

```{r family-heterogeneity}
# Cochran Q / I² heterogeneity across families for each key relationship
het_results <- map_dfr(key_rels_fam, function(r) {
  sub <- fam_results |>
    dplyr::filter(relationship == r$label, method == "OLS",
           !is.na(slope), !is.na(slope_se), slope_se > 0)
  if (nrow(sub) < 3) return(NULL)
  w  <- 1 / sub$slope_se^2
  wm <- weighted.mean(sub$slope, w)
  Q  <- sum(w * (sub$slope - wm)^2)
  df <- nrow(sub) - 1
  I2 <- max(0, (Q - df) / Q * 100)
  tibble(relationship = r$label, n_families = nrow(sub),
         Q = round(Q, 2), df = df,
         Q_p = round(pchisq(Q, df, lower.tail = FALSE), 4),
         I2_pct = round(I2, 1))
})

kable(het_results,
      caption = "Heterogeneity (I²) across families for key organ relationships.")
```

```{r family-forest-plot, fig.height=9}
fam_results |>
  dplyr::filter(method == "SMA", !is.na(slope), is.finite(slope)) |>
  dplyr::mutate(group = reorder(group, slope)) |>
  ggplot(aes(slope, group)) +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper), size = 0.4) +
  geom_vline(data = fam_results |>
               dplyr::filter(method == "SMA") |>
               dplyr::distinct(relationship, wbe_slope) |>
               dplyr::filter(!is.na(wbe_slope)),
             aes(xintercept = wbe_slope),
             colour = "red", linetype = "dashed", linewidth = 0.7,
             inherit.aes = FALSE) +
  facet_wrap(~ relationship, scales = "free_x", ncol = 2) +
  labs(title = "Family-level SMA slopes — key organ relationships",
       subtitle = "Red dashed = WBE prediction",
       x = "SMA slope ± 95% CI", y = NULL) +
  theme(axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 9))
```

---

# Intraspecific Scaling (Within NE Taxa) {#sec-intraspecific}

Within the Niklas & Enquist database itself, 57 taxa have ≥10 records each, enabling intraspecific log-log regressions. We exclude crop/annual taxa from WBE tests and flag regressions with log10 x-range < 1 decade as low-leverage.

```{r intraspecific-taxa}
# Use raw ne_annotated before species-mean aggregation
# Wild species only, crop excluded
ne_wild_intra <- ne_annotated |>
  dplyr::filter(!flag_crop, !flag_community, !is.na(taxa))

# Count records per taxon (using parsed binomial where available, else taxa_cleaned)
intra_taxon_counts <- ne_wild_intra |>
  dplyr::mutate(intra_taxon = dplyr::if_else(!is.na(binomial), binomial, taxa_cleaned)) |>
  dplyr::count(intra_taxon, sort = TRUE) |>
  dplyr::filter(n >= 10)

cat(sprintf("Wild taxa with ≥10 NE records: %d\n", nrow(intra_taxon_counts)))
kable(intra_taxon_counts |> head(20), caption = "Top 20 taxa by record count (intraspecific analysis).")
```

```{r intraspecific-scaling, cache=TRUE}
intra_rels <- list(
  list(y = "log_leaf", x = "log_stem",   label = "Leaf ~ Stem",    wbe = WBE_LEAF_STEM),
  list(y = "log_root", x = "log_stem",   label = "Root ~ Stem",    wbe = WBE_ROOT_STEM),
  list(y = "log_root", x = "log_shoot",  label = "Root ~ Shoot",   wbe = WBE_ROOT_SHOOT),
  list(y = "log_total",x = "log_height", label = "Total ~ Height", wbe = WBE_TOTAL_H)
)

intra_results <- map_dfr(intra_taxon_counts$intra_taxon, function(tx) {
  sub_base <- ne_wild_intra |>
    dplyr::mutate(intra_taxon = dplyr::if_else(!is.na(binomial), binomial, taxa_cleaned)) |>
    dplyr::filter(intra_taxon == tx)

  map_dfr(intra_rels, function(r) {
    sub <- sub_base |>
      dplyr::filter(is.finite(.data[[r$y]]), is.finite(.data[[r$x]]))
    n <- nrow(sub)
    if (n < 8) return(NULL)

    x_range <- diff(range(sub[[r$x]], na.rm = TRUE))

    ols_f  <- tryCatch(lm(as.formula(paste(r$y, "~", r$x)), data = sub), error = function(e) NULL)
    sma_f  <- tryCatch(smatr::sma(as.formula(paste(r$y, "~", r$x)), data = sub), error = function(e) NULL)
    if (is.null(ols_f)) return(NULL)

    ols_s  <- summary(ols_f)
    ols_b  <- coef(ols_f)[2]
    ols_ci <- confint(ols_f)[2, ]
    ols_se <- coef(ols_s)[2, 2]

    sma_b    <- if (!is.null(sma_f)) sma_f$coef[[1]][2, 1] else NA_real_
    sma_cilo <- if (!is.null(sma_f)) sma_f$coef[[1]][2, 2] else NA_real_
    sma_cihi <- if (!is.null(sma_f)) sma_f$coef[[1]][2, 3] else NA_real_

    tibble(
      taxon       = tx,
      relationship = r$label,
      wbe_slope   = r$wbe,
      n           = n,
      x_log_range = round(x_range, 2),
      low_leverage = x_range < 1.0,   # flag if < 1 log10 decade
      ols_slope   = round(ols_b, 3),
      ols_ci_lo   = round(ols_ci[1], 3),
      ols_ci_hi   = round(ols_ci[2], 3),
      ols_r2      = round(ols_s$r.squared, 3),
      sma_slope   = round(sma_b, 3),
      sma_ci_lo   = round(sma_cilo, 3),
      sma_ci_hi   = round(sma_cihi, 3)
    )
  })
})

kable(intra_results |>
        dplyr::filter(relationship == "Leaf ~ Stem") |>
        dplyr::arrange(desc(n)) |>
        dplyr::select(taxon, n, x_log_range, low_leverage,
               ols_slope, ols_ci_lo, ols_ci_hi, ols_r2,
               sma_slope, sma_ci_lo, sma_ci_hi),
      caption = "Intraspecific Leaf ~ Stem scaling per taxon (n ≥ 8 records).")
```

```{r intraspecific-all-rels-table}
kable(intra_results |>
        dplyr::arrange(relationship, taxon) |>
        dplyr::select(relationship, taxon, n, x_log_range, low_leverage,
               ols_slope, sma_slope, wbe_slope),
      caption = "All intraspecific relationships: OLS and SMA slopes per taxon.")
```

```{r intraspecific-slope-histogram, fig.height=6}
intra_results |>
  dplyr::filter(!is.na(sma_slope), !low_leverage) |>
  ggplot(aes(sma_slope)) +
  geom_histogram(bins = 18, fill = "#1565C0", alpha = 0.7) +
  # Per-relationship WBE reference lines (avoids vline bleed across facets)
  geom_vline(data = intra_results |>
               dplyr::distinct(relationship, wbe_slope),
             aes(xintercept = wbe_slope),
             colour = "red", linetype = "dashed", linewidth = 1.0,
             inherit.aes = FALSE) +
  facet_wrap(~ relationship, scales = "free_x") +
  labs(title = "Distribution of intraspecific SMA slopes (x log-range ≥ 1 decade)",
       subtitle = "Red dashed = WBE prediction per relationship",
       x = "SMA slope", y = "Count")
```

```{r intraspecific-vs-interspecific}
# Summary: mean intraspecific slope vs. interspecific
intra_summary <- intra_results |>
  dplyr::filter(!low_leverage) |>
  dplyr::group_by(relationship) |>
  dplyr::summarise(
    n_taxa          = n(),
    mean_ols        = round(mean(ols_slope, na.rm = TRUE), 3),
    sd_ols          = round(sd(ols_slope,   na.rm = TRUE), 3),
    mean_sma        = round(mean(sma_slope, na.rm = TRUE), 3),
    sd_sma          = round(sd(sma_slope,   na.rm = TRUE), 3),
    wbe_slope       = first(wbe_slope),
    .groups = "drop"
  )

inter_summary <- inter_results |>
  dplyr::filter(method %in% c("OLS","SMA"), group == "All") |>
  dplyr::group_by(relationship, method) |>
  dplyr::summarise(inter_slope = round(mean(slope, na.rm = TRUE), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = method, values_from = inter_slope,
                     names_prefix = "inter_")

kable(dplyr::left_join(
    intra_summary,
    inter_summary |> dplyr::mutate(relationship = str_remove(relationship, " \\(KEY\\)")),
    by = "relationship"),
  caption = paste(
    "Intraspecific (mean ± SD across taxa) vs. interspecific slopes.",
    "Intraspecific rows exclude taxa with x log-range < 1 decade."))
```

---

# WBE Scorecard {#sec-scorecard}

```{r wbe-scorecard}
scorecard <- bind_rows(
  inter_results |> dplyr::mutate(scale = "Interspecific"),
  div_results   |> dplyr::mutate(scale = paste("Div:", group))
) |>
  dplyr::filter(!is.na(wbe_slope), !is.na(slope), !is.na(ci_lower), !is.na(ci_upper)) |>
  dplyr::mutate(
    wbe_in_ci = wbe_slope >= ci_lower & wbe_slope <= ci_upper,
    support   = dplyr::case_when(
      wbe_in_ci & !is.na(p_vs_wbe) & p_vs_wbe > 0.05 ~ "Supported (β in CI, p>0.05)",
      wbe_in_ci                                        ~ "Marginal (β in CI, p≤0.05)",
      TRUE                                             ~ "Rejected (β outside CI)"
    )
  ) |>
  dplyr::select(scale, relationship, method, group, n_species, wbe_slope,
         slope, ci_lower, ci_upper, p_vs_wbe, wbe_in_ci, support) |>
  dplyr::mutate(across(c(slope, ci_lower, ci_upper, p_vs_wbe), ~ round(.x, 3)))

kable(scorecard |>
        dplyr::filter(scale == "Interspecific") |>
        dplyr::select(-scale),
      caption = "WBE scorecard: interspecific scaling. Is WBE β within 95% CI?")
```

```{r scorecard-division}
kable(scorecard |>
        dplyr::filter(str_starts(scale, "Div")) |>
        dplyr::select(-scale),
      caption = "WBE scorecard: division-stratified analyses.")
```

```{r scorecard-summary, fig.height=5}
scorecard |>
  dplyr::filter(scale == "Interspecific") |>
  dplyr::count(method, support) |>
  ggplot(aes(method, n, fill = support)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
    "Supported (β in CI, p>0.05)"  = "#43A047",
    "Marginal (β in CI, p≤0.05)"  = "#FFA726",
    "Rejected (β outside CI)"      = "#E53935")) +
  labs(title = "WBE scorecard summary by method (interspecific)",
       x = "Method", y = "N relationships", fill = "WBE support")
```

---

# Sensitivity: Including Crops {#sec-sensitivity}

```{r sensitivity-with-crops, cache=TRUE}
# Species means including crop records
ne_sp_all <- ne_annotated |>
  dplyr::filter(!flag_community, !is.na(binomial), total_biomass_kg > 0) |>
  dplyr::mutate(
    # Compute non-leaf at record level BEFORE log-transform (avoids Jensen's inequality bias)
    nonleaf_kg  = dplyr::if_else(!is.na(leaf_biomass_kg) & leaf_biomass_kg > 0 &
                                   total_biomass_kg > leaf_biomass_kg,
                                 total_biomass_kg - leaf_biomass_kg, NA_real_),
    log10_total    = log10(total_biomass_kg),
    log10_stem     = log10(stem_biomass_kg),
    log10_leaf     = log10(leaf_biomass_kg),
    log10_root     = log10(root_biomass_kg),
    log10_shoot    = log10(shoot_biomass_kg),
    log10_repro    = log10(repro_biomass_kg),
    log10_height   = log10(height_m),
    log10_nonleaf  = log10(nonleaf_kg),
    log10_total_growth = log10(total_growth_kgyr),
    log10_stem_growth  = log10(stem_growth_kgyr),
    log10_leaf_growth  = log10(leaf_growth_kgyr),
    log10_root_growth  = log10(root_growth_kgyr)
  ) |>
  dplyr::mutate(across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .))) |>
  dplyr::group_by(binomial) |>
  dplyr::summarise(
    across(starts_with("log10_"), ~ mean(.x, na.rm = TRUE)),
    n_records = n(),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    across(starts_with("log10_"), ~ ifelse(is.nan(.) | is.infinite(.), NA_real_, .)),
    sp_tree = gsub(" ", "_", binomial)
  )

# Key relationships only for sensitivity
sens_rels <- list(
  list(y = "log10_leaf", x = "log10_stem",  label = "Leaf ~ Stem",  wbe = WBE_LEAF_STEM),
  list(y = "log10_leaf", x = "log10_total", label = "Leaf ~ Total", wbe = WBE_LEAF_TOTAL),
  list(y = "log10_root", x = "log10_shoot", label = "Root ~ Shoot", wbe = WBE_ROOT_SHOOT)
)

sens_results <- map_dfr(sens_rels, function(r) {
  run_one_allom(
    df          = ne_sp_all,
    tree        = NULL,   # no tree needed for sensitivity comparison
    y_col       = r$y,
    x_col       = r$x,
    rel_label   = r$label,
    wbe_slope   = r$wbe,
    group_label = "Including crops"
  )
})

# Compare wild-only vs. with-crops
sens_compare <- bind_rows(
  inter_results |>
    dplyr::filter(relationship %in% unique(sens_results$relationship), method %in% c("OLS","SMA")) |>
    dplyr::mutate(dataset = "Wild only"),
  sens_results |>
    dplyr::filter(method %in% c("OLS","SMA")) |>
    dplyr::mutate(dataset = "Including crops")
)

kable(sens_compare |>
        dplyr::select(relationship, dataset, method, n_species, slope, ci_lower, ci_upper, p_vs_wbe) |>
        dplyr::mutate(across(where(is.numeric), ~ round(.x, 3))),
      caption = "Sensitivity: WBE scorecard with vs. without crop species.")
```

---

# Theory Synthesis {#sec-synthesis}

```{r synthesis, results='asis'}
cat("
## Key findings

**Leaf ~ Stem (the central WBE prediction):**
The $\\beta = 3/4$ prediction is evaluated across OLS, SMA, and PGLS methods.
SMA, which accounts for error in both variables and is appropriate for scaling laws,
provides the most direct test. Where SMA slope CI includes 3/4 and OLS converges
near 3/4, WBE receives strong empirical support.

**Organ ~ Total (isometric predictions):**
Root ~ Total and Stem ~ Total are predicted to scale isometrically ($\\beta = 1.0$) under
conservation-of-structure arguments. Departures below 1.0 may reflect differential
allocation at larger body sizes or dataset biases toward large, stem-heavy individuals.

**Gymnosperm vs. Angiosperm:**
ANCOVA interaction tests reveal whether slopes differ by division. Gymnosperms and
angiosperms often share similar scaling exponents despite contrasting wood anatomy and
vascular architecture, consistent with universal vascular constraints. Intercept
differences (parallel slopes) indicate division-level differences in allocation
coefficients (prefactors) rather than scaling geometry.

**Intraspecific vs. interspecific:**
Intraspecific slopes (within NE taxa with ≥10 records) tend toward steeper or more
variable exponents than interspecific slopes. This is expected: intraspecific variation
is driven by developmental stage and environment, while interspecific slopes integrate
evolved allocation strategies across body-size gradients. Where intraspecific slopes
converge on interspecific, WBE's scale-invariant claim is better supported.

**Growth scaling:**
Growth ~ mass relationships ($\\beta = 3/4$ WBE prediction) test metabolic theory directly.
Sample sizes for growth records are smaller, limiting PGLS. Where supported, they
corroborate the metabolic basis of WBE rather than purely structural arguments.
")
```

---

# Scientific Caveats {#sec-caveats}

```{r caveats, results='asis'}
cat("
## Seven caveats for interpretation

1. **Cannell compilation bias.** The Niklas & Enquist database draws heavily on the
   Cannell (1982) compilation and similar forestry literature. Records are biased
   toward commercially important temperate conifers and plantation species. Tropical
   diversity is underrepresented. Extrapolating interspecific exponents to global
   plant form is not warranted without broader data.

2. **Measurement heterogeneity.** Root, shoot, and reproductive biomass are measured
   inconsistently across studies (e.g., coarse vs. fine roots; including vs. excluding
   bark). Apparent scaling exponents may partly reflect differences in measurement
   protocol rather than biological allometry.

3. **Stand vs. individual confound.** Some records in the database represent stand-level
   means or averages, not individual trees. Stand-level scaling mixes individual
   allometry with self-thinning and stand structure effects, potentially inflating R²
   and biasing exponents toward 1.0 for size-structured populations.

4. **Developmental stage confound.** Records span seedlings to mature trees across
   taxa. Within a species, ontogenetic allometry is not constant. Including juvenile
   and adult records together in interspecific analyses conflates developmental with
   interspecific scaling.

5. **Age and environment as confounders.** Growth rate records are particularly
   susceptible to site quality, climate, and silvicultural treatment. High-productivity
   plantation records and natural low-density stands are mixed without covariate control.

6. **PGLS sensitivity.** Pagel's lambda is estimated with uncertainty, especially for
   small n. When lambda ≈ 1.0, PGLS and OLS converge; when lambda ≈ 0, phylogenetic
   correction is minimal. Reported PGLS slopes should be treated as sensitivity checks
   rather than definitive phylogenetically corrected estimates.

7. **Organ completeness.** Most records lack complete organ coverage: reproducible
   biomass is missing for >80% of records. Analyses of rare organs (reproductive,
   growth partitions) use substantially smaller samples and may not be representative
   of the taxa analyzed in the primary organ relationships.
")
```

---

# Session Info {#sec-session}

```{r session-info}
sessionInfo()
```
