---
title: "Allometric Scaling — OLS vs SMA vs PGLS Analysis"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
suppressPackageStartupMessages({
  library(caper)
  library(ape)
  library(dplyr)
  library(ggplot2)
  library(smatr)
  library(phytools)
  library(tibble)
  library(readr)
  library(tidyr)
  library(knitr)
  library(scales)
})
theme_set(theme_bw(base_size = 13))

proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

```{r load-results}
results_table    <- readRDS(file.path(proj_dir, "data/processed/pgls_results_table.rds"))
phylosig_table   <- readRDS(file.path(proj_dir, "data/processed/phylosig_table.rds"))
ne_species_means <- readRDS(file.path(proj_dir, "data/processed/ne_species_means.rds"))
baad_species_means <- readRDS(file.path(proj_dir, "data/processed/baad_species_means.rds"))
pruned_tree      <- readRDS(file.path(proj_dir, "data/processed/pgls_tree_smith2018_s2.rds"))
```

---

# Overview

This report documents the phylogenetically informed allometric scaling analysis
for two plant datasets.  Three regression methods are compared for each
scaling relationship: OLS, SMA, and PGLS.

**Datasets:**

| Dataset | N records | Aggregation unit | Primary source |
|---------|-----------|-----------------|----------------|
| Niklas-Enquist | ~2,626 cross-species records | Species means (geometric mean of log10 values) | Cannell (1982) and primary literature compiled by Niklas & Enquist (2001) |
| BAAD | ~21,084 individual-level records | Species means (geometric mean of log10 values) | Falster et al. (2015) |

For PGLS, both datasets were aggregated to species means by taking the
arithmetic mean of log10-transformed values, which is equivalent to the
geometric mean in the original scale.

**Regression methods:**

- **OLS** (ordinary least squares): treats each species as independent; ignores
  phylogenetic relatedness.
- **SMA** (standardized major axis): accounts for measurement error in both
  variables but still ignores phylogeny.  Preferred over OLS for allometric
  estimation when both axes carry error.
- **PGLS** (phylogenetic generalised least squares, `caper::pgls`): corrects
  for phylogenetic non-independence by modelling the expected covariance among
  species via the variance-covariance matrix of the pruned phylogeny.
  Pagel's lambda (λ) is estimated by maximum likelihood to scale the off-diagonal
  covariances.

**Phylogeny:** Smith & Brown (2018) GBOTB seed plant megatree, pruned to study
species via V.PhyloMaker2 (Scenario 2).

**WBE theoretical predictions (West, Brown & Enquist 1997):**

- $\text{AGB} \sim H^{8/3} \approx 2.67$
- $\frac{dM}{dt} \sim M^{3/4}$
- leaf biomass $\sim$ total biomass$^{3/4}$
- $H \sim \text{DBH}^{2/3}$

---

# Phylogeny

```{r phylogeny-summary}
cat("Tree tips:      ", ape::Ntip(pruned_tree), "\n")
cat("Ultrametric:    ", ape::is.ultrametric(pruned_tree, tol = 1e-5), "\n")
```

```{r match-table, results='asis'}
match_path <- file.path(proj_dir, "data/processed/pgls_tree_match_table.csv")
if (file.exists(match_path)) {
  match_tbl <- read_csv(match_path, show_col_types = FALSE)
  knitr::kable(match_tbl, caption = "Species-to-tree tip match summary.")
} else {
  ne_in_tree   <- nrow(ne_species_means[ne_species_means$taxa %in% pruned_tree$tip.label, ])
  baad_in_tree <- nrow(baad_species_means[baad_species_means$speciesMatched %in% pruned_tree$tip.label, ])
  knitr::kable(
    tibble(
      dataset        = c("Niklas-Enquist", "BAAD"),
      species_in_data = c(nrow(ne_species_means), nrow(baad_species_means)),
      species_in_tree = c(ne_in_tree, baad_in_tree)
    ),
    caption = "Species matched to pruned phylogeny per dataset."
  )
}
```

Tip labels follow the V.PhyloMaker2 GBOTB format.  Species from each dataset
were matched to tree tips after taxonomy cleaning.

---

# Phylogenetic Signal Before Regression

```{r phylosig-table, results='asis'}
knitr::kable(
  phylosig_table |>
    mutate(
      lambda     = round(lambda, 2),
      lambda_pval = formatC(lambda_pval, format = "g", digits = 3),
      K          = round(K, 2),
      K_pval     = formatC(K_pval, format = "g", digits = 3)
    ),
  col.names = c("Trait", "λ", "λ p-value", "K", "K p-value", "N species"),
  caption   = paste(
    "Pagel's lambda (λ) and Blomberg's K for each trait.",
    "λ = 0: no phylogenetic signal; λ = 1: signal matches Brownian motion prediction.",
    "K > 1: more signal than Brownian motion."
  )
)
```

If λ is high for a trait, OLS will underestimate standard errors by treating
closely related species as independent observations.  In that case, PGLS
confidence intervals will be wider than OLS, and the apparent statistical
support for a WBE exponent may be overstated by OLS.

---

# Scaling Results

```{r results-setup}
fmt3 <- function(x) round(x, 3)

print_relationship_table <- function(dat, y_col, x_col, ds, wbe_slope) {
  sub <- dat |>
    filter(dataset == ds, y_var == y_col, x_var == x_col) |>
    select(method, slope, ci_lower, ci_upper, r_squared, lambda, p_vs_wbe) |>
    mutate(across(where(is.numeric), fmt3))
  knitr::kable(
    sub,
    col.names = c("Method", "Slope", "CI lower", "CI upper", "R²", "λ (PGLS)", "p vs WBE"),
    caption   = paste0(ds, ": ", y_col, " ~ ", x_col,
                       if (!is.na(wbe_slope)) paste0(" (WBE predicted slope = ", wbe_slope, ")") else "")
  )
}
```

## Niklas-Enquist Dataset

```{r ne-biomass-height, results='asis'}
print_relationship_table(results_table,
  "log_total_biomass_kg", "log_height_m", "Niklas-Enquist", round(8/3, 3))
```

WBE predicts a slope of 8/3 ≈ 2.667 for total biomass ~ height.  Compare each
method's CI to this value; a CI that excludes 2.667 indicates a statistically
significant departure.

```{r ne-growth-biomass, results='asis'}
print_relationship_table(results_table,
  "log_total_growth_kgyr", "log_total_biomass_kg", "Niklas-Enquist", 0.75)
```

WBE predicts a slope of 3/4 = 0.75 for net growth rate ~ total biomass.

> **⚠️ CRITICAL CAVEAT: Growth rate measurement** — Growth rates in the
> Niklas-Enquist dataset are annual net biomass increment period averages
> compiled from Cannell (1982), not instantaneous metabolic rates.  WBE's
> $dM/dt \propto M^{3/4}$ prediction applies to gross metabolic flux, which
> includes maintenance respiration (typically 40–60% of GPP in temperate trees).
> Net increment = gross production − maintenance respiration.  Therefore this
> analysis **does not constitute a direct test of the WBE 3/4 metabolic
> scaling prediction**.  Results are descriptive: if net increment also scales
> as ~$M^{3/4}$, it is consistent with but does not confirm WBE.

```{r ne-leaf-total, results='asis'}
print_relationship_table(results_table,
  "log_leaf_biomass_kg", "log_total_biomass_kg", "Niklas-Enquist", 0.75)
```

WBE predicts leaf biomass ~ total biomass$^{3/4}$.

> **⚠️ WARNING: Compositional / mathematical coupling artifact** —
> leaf_biomass + stem_biomass + root_biomass ≈ total_biomass.  Regressing any
> component against its sum (e.g., leaf ~ total) introduces spurious allometry
> through mathematical coupling — known since Pearson (1897) and formalised
> in the ecological literature by Prairie & Bird (1989) and Kronmal (1993).
> **Citations need independent verification before manuscript use.**
> Interpret these slopes as descriptive only; they cannot confirm or
> falsify biomass partitioning theory without a mathematically independent
> analysis (e.g., regressing components against DBH or height).

```{r ne-stem-total, results='asis'}
print_relationship_table(results_table,
  "log_stem_biomass_kg", "log_total_biomass_kg", "Niklas-Enquist", 1.0)
```

WBE predicts isometric scaling (slope = 1.0) for stem biomass ~ total biomass.

> **⚠️ WARNING: Compositional / mathematical coupling artifact** — Same
> mathematical coupling concern applies here.  See note above for leaf ~ total.

---

## BAAD Dataset

```{r baad-agb-dbh, results='asis'}
print_relationship_table(results_table,
  "log_AGB_kg", "log_DBH_cm", "BAAD", NA_real_)
```

No specific WBE theoretical prediction is applied to AGB ~ DBH in log-log
space here; the slope is reported descriptively.

```{r baad-ht-dbh, results='asis'}
print_relationship_table(results_table,
  "log_height_m", "log_DBH_cm", "BAAD", round(2/3, 3))
```

WBE predicts H ~ DBH$^{2/3}$ ≈ 0.667 for height ~ diameter scaling.

---

# Visualizations

```{r forest-plot, fig.cap="Forest plot of allometric scaling exponents. Points = estimated slope; error bars = 95% CI; dashed line = WBE theoretical prediction. OLS (blue), SMA (orange), PGLS (green)."}
forest_path <- file.path(proj_dir, "output/forest_plot_slopes.png")
if (file.exists(forest_path)) {
  knitr::include_graphics(forest_path)
} else {
  message("forest_plot_slopes.png not found — run scripts/06_pgls_allometry.R first.")
}
```

```{r scatter-ne, fig.cap="Niklas-Enquist species means: log10(Total biomass [kg]) ~ log10(Height [m]). Blue line = OLS fit with 95% confidence band."}
scatter_path <- file.path(proj_dir, "output/scatter_ne_biomass_height.png")
if (file.exists(scatter_path)) {
  knitr::include_graphics(scatter_path)
} else {
  message("scatter_ne_biomass_height.png not found — run scripts/06_pgls_allometry.R first.")
}
```

---

# Interpretation

## OLS vs PGLS Slope Differences

When the PGLS slope differs from the OLS slope, this indicates that clade-level
body size stratification is influencing the estimate.  PGLS down-weights
within-clade replication through the phylogenetic variance-covariance matrix:
species that share a recent common ancestor are treated as partially redundant
observations.  A PGLS slope that is shallower than OLS is consistent with the
interpretation that the OLS slope partly reflects between-clade body size
diversification (e.g., the contrast between large-bodied trees and small-bodied
herbs) rather than a constraint operating within lineages.

## Pagel's Lambda Interpretation

The ML-estimated λ in PGLS residuals indicates how much of the unexplained
biomass variation is phylogenetically conserved.  λ near 1 means residuals
follow the Brownian motion expectation — traits have diverged proportionally to
branch length.  This is expected for architectural traits such as wood density,
crown geometry, and height–diameter ratios, which differ substantially among
major plant clades (gymnosperms vs. angiosperms, monocots vs. eudicots).  When
λ is high, OLS standard errors are anti-conservative and p-values should be
interpreted with caution.

## WBE Theory Evaluation

The recovery of exponents close to WBE predictions (8/3 for biomass ~ height,
3/4 for growth ~ biomass) in OLS regressions of cross-species data is
consistent with the theory but does not uniquely confirm it.  The same exponents
are predicted by several alternative models and may also arise from between-clade
body size diversification rather than from the fractal vascular network
constraint WBE postulated.  PGLS provides a more conservative test because it
partially removes the among-clade signal.  If the PGLS exponent still matches
the WBE prediction, that constitutes stronger within-lineage support.

## Dataset Caveats

> **⚠ WARNING: Clade bias** — The Niklas-Enquist dataset is derived
> substantially from Cannell (1982) and overrepresents temperate deciduous trees
> and conifers.  Conclusions may not generalize to tropical broadleaf species or
> herbaceous taxa.

> **⚠ WARNING: Growth rate measurement** — Growth rates in Niklas-Enquist are
> annual net biomass increment period averages, not instantaneous metabolic
> rates.  The test of $dM/dt \propto M^{3/4}$ is an approximation; net
> production ≠ gross metabolic rate.

---

# Methods

Allometric scaling was analysed for two datasets.  The **Niklas-Enquist**
database (~2,626 records) compiles cross-species biomass measurements primarily
from Cannell (1982) with additional primary literature records; see Niklas &
Enquist (2001) *PNAS* 98:2922-2927 and Enquist & Niklas (2002) *Science*
295:1517-1520.  The **BAAD** database (~21,084 individual-level records) is
described in Falster et al. (2015) *Scientific Data* 2:150034
(https://doi.org/10.1038/sdata.2015.34).

Both datasets were aggregated to species means by taking arithmetic means of
log10-transformed trait values (equivalent to geometric means).  Only species
present in both the dataset and the phylogeny were retained for PGLS.

The phylogeny is the Smith & Brown (2018) GBOTB seed plant megatree, pruned
to study species using V.PhyloMaker2 (Scenario 2).  Citation: Smith SA & Brown
JW (2018) Constructing a broadly inclusive seed plant phylogeny.
*American Journal of Botany* 105(3):302-314.
DOI: 10.1002/ajb2.1019.  **Note: this citation should be verified
independently before any manuscript use.**

OLS regressions used `stats::lm()`.  SMA regressions used the `smatr` package
(Warton et al. 2012 *Methods in Ecology and Evolution* 3:257-259).  Tests of
SMA slopes against WBE theoretical values used `smatr::slope.test()`.  PGLS
regressions used `caper::pgls()` (Orme et al. 2018), with Pagel's λ estimated
by maximum likelihood; if ML optimisation failed, λ was fixed at 1.
Phylogenetic signal was assessed with Pagel's λ and Blomberg's K via
`phytools::phylosig()` (Revell 2012 *Methods in Ecology and Evolution*
3:217-223).

---

# Pipeline Run Order

> **Required execution order before knitting this report:**
>
> 1. `Rscript scripts/04_prepare_pgls_species_means.R` — aggregate species means
> 2. `Rscript scripts/05_build_phylogeny_smith2018.R` — build pruned tree (requires V.PhyloMaker2)
> 3. `Rscript scripts/06_pgls_allometry.R` — run OLS / SMA / PGLS regressions
> 4. Knit this Rmd
>
> All scripts should be run from the `plant_scaling_data/` directory.
> Outputs are written to `data/processed/` and `output/`.

```{r session-info}
cat("Analysis date:", format(Sys.Date()), "\n")
R.Version()$version.string
tryCatch(cat("V.PhyloMaker2 version:", as.character(packageVersion("V.PhyloMaker2")), "\n"), error=function(e) NULL)
tryCatch(cat("caper version:", as.character(packageVersion("caper")), "\n"), error=function(e) NULL)
```
