---
title: "SAPFLUXNET / Forrester et al. 2022 — Scaling of Whole-Tree Water Use With Tree Size"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5)
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
  library(knitr)
  library(scales)
  library(smatr)
  library(purrr)
  library(stringr)
  library(leaflet)
})
theme_set(theme_bw(base_size = 13))
proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

---

# Overview and Data Sources

## Synthesis paper

> **Forrester DI, Limousin J-M, Pfautsch S.** (2022). The relationship between
> tree size and tree water-use: is competition for water size-symmetric or
> size-asymmetric? *Tree Physiology* **42**: 1916–1927.
> DOI: [10.1093/treephys/tpac018](https://doi.org/10.1093/treephys/tpac018)

Forrester et al. analysed **2,300 trees**, **103 sites**, **141 plots**, and
**80 species** from the SAPFLUXNET database to test whether competition for water
is size-symmetric (water use ∝ some linear function of size) or size-asymmetric
(larger trees disproportionately capture water).

**Key reported scaling results:**

| Size metric | Expected slope | Observed slope (approx.) | Interpretation |
|---|---|---|---|
| Stem diameter $D$ | 2.0 | ~2.0 | $E \propto D^2$ = basal-area scaling |
| Basal area $BA = \pi D^2/4$ | 1.0 | ~1.0 | $E \propto BA^1$ = linear with cross-section |
| Sapwood area $SA$ | 1.0 | ~1.0 | $E \propto SA^1$ = flux ∝ conducting tissue |

The tree is not a single straw — it is a bundle of straws whose **number and
active cross-section change with size**. The linear scaling with $BA$ and $SA$
implies that **sap flux density (E/SA)** is approximately size-independent,
which underlies the WBE pipe-model assumptions. Forrester et al. found that
**competition for water is size-symmetric** — larger trees use more water in
proportion to their cross-sectional area, not disproportionately more.

## Older anchor

> **Meinzer FC, Bond BJ, Warren JM, Woodruff DR.** (2005). Does water transport
> scale universally with tree size? *Functional Ecology* **19**: 558–565.
> DOI: [10.1111/j.1365-2435.2005.01017.x](https://doi.org/10.1111/j.1365-2435.2005.01017.x)

Meinzer et al. examined size dependence of whole-tree water use, sapwood area,
stem diameter, and above-ground biomass across angiosperms and conifers, finding
$E \propto D^{1.8}$ (close to, but slightly below, the $D^2$ prediction) and
$E \propto SA^{0.9}$ (close to linear). Forrester et al. 2022 is the
large-scale SAPFLUXNET follow-up.

## SAPFLUXNET database

> **Poyatos R, et al.** (2021). Global transpiration data from sap flow
> measurements: the SAPFLUXNET database. *Earth System Science Data* **13**:
> 2607–2649. DOI: [10.5194/essd-13-2607-2021](https://doi.org/10.5194/essd-13-2607-2021)

> **SAPFLUXNET v0.1.5** (2020). Zenodo.
> DOI: [10.5281/zenodo.3971689](https://doi.org/10.5281/zenodo.3971689)

SAPFLUXNET v0.1.5 contains **2,714 individual plants** from **174 species** and
**202 datasets**. Plant-level metadata include stem diameter (`pl_dbh`, cm),
sapwood area (`pl_sapw_area`, cm²), and species; sap flow time series are
in separate CSVs.

> **⚠ Citation note:** All DOIs, journal names, and volume/page numbers should be
> independently verified against Crossref before use in any manuscript.

---

# Data loading

```{r load-data}
# ── SAPFLUXNET plant-level metadata ──────────────────────────────────────────
# The SAPFLUXNET database (v0.1.5, ~3.2 GB zip) is available at:
#   https://doi.org/10.5281/zenodo.3971689
#
# SAPFLUXNET v0.1.5 folder structure after extracting 0.1.5.zip:
#
#   0.1.5/
#     RData/plant/   — sfn_data objects (via sapfluxnetr package)
#     csv/plant/     — text CSVs: plant-level sap flow + metadata files per site
#     csv/sapwood/   — text CSVs: sapwood-level sap flow
#
# Units in raw plant-level CSV files (pl_sap_units = "cm3 h-1"):
#   Plant level   sap flow : cm³ h⁻¹  (whole-plant flow, per hour)
#   pl_dbh        : cm
#   pl_sapw_area  : cm²
#
# NOTE: The sapfluxnetr R package (get_sapf_data()) standardises to cm³ s⁻¹,
# but the raw CSV files store values in cm³ h⁻¹ as confirmed by pl_sap_units.
#
# E_tree (L day⁻¹) from plant-level CSVs:
#   mean_sap_cm3_h × 24 / 1000  → L day⁻¹
#   (24 h day⁻¹; 1000 cm³ L⁻¹)
#
# Forrester et al. (2022) used mean growing-season daily water use per plant.
#
# Download: https://doi.org/10.5281/zenodo.3971689  (~3.2 GB zip)
# Extract to: plant_scaling_data/data/raw/sapfluxnet/
# Expected path after extraction: data/raw/sapfluxnet/0.1.5/csv/plant/<SITE>/

sfn_base <- file.path(proj_dir, "data/raw/sapfluxnet")

# Plant-level CSVs: one per site, named <SITE>_sapf_data.csv + <SITE>_plant_md.csv
# They live under 0.1.5/csv/plant/<SITE>/
plant_md_files <- list.files(sfn_base, pattern = "_plant_md\\.csv$",
                              recursive = TRUE, full.names = TRUE)
sap_files      <- list.files(sfn_base, pattern = "_sapf_data\\.csv$",
                              recursive = TRUE, full.names = TRUE)

# Also accept sapwood-level layout if user extracted there
if (length(plant_md_files) == 0) {
  plant_md_files <- list.files(sfn_base, pattern = "_plant_md\\.csv$",
                                recursive = TRUE, full.names = TRUE)
}

USE_REAL_DATA <- length(plant_md_files) > 10

# ── Load site metadata (lat/lon, country, biome, climate) ─────────────────────
site_files <- list.files(sfn_base, pattern = "_site_md\\.csv$",
                         recursive = TRUE, full.names = TRUE)
if (length(site_files) > 0) {
  site_md <- map_dfr(site_files, function(f)
    tryCatch(read_csv(f, show_col_types = FALSE,
                      col_types = cols(.default = col_character())),
             error = function(e) NULL)) |>
    mutate(
      si_lat  = suppressWarnings(as.numeric(si_lat)),
      si_long = suppressWarnings(as.numeric(si_long)),
      si_mat  = suppressWarnings(as.numeric(si_mat)),
      si_map  = suppressWarnings(as.numeric(si_map)),
      si_elev = suppressWarnings(as.numeric(si_elev))
    )
  cat("Site metadata loaded:", nrow(site_md), "sites\n")
} else {
  site_md <- NULL
}

if (USE_REAL_DATA) {
  # ── Load and stack all plant metadata CSVs ─────────────────────────────────
  # Read all as character first to avoid bind_rows type conflicts across sites,
  # then coerce needed numeric columns explicitly.
  plant_md <- map_dfr(plant_md_files, function(f) {
    d <- tryCatch(
      read_csv(f, show_col_types = FALSE, col_types = cols(.default = col_character())),
      error = function(e) NULL
    )
    if (is.null(d)) return(NULL)
    d
  }) |>
    mutate(
      pl_dbh       = suppressWarnings(as.numeric(pl_dbh)),
      pl_sapw_area = suppressWarnings(as.numeric(pl_sapw_area))
    )
  cat("Plant metadata rows loaded:", nrow(plant_md), "\n")

  # ── Compute mean water use from plant-level sap flow CSVs ─────────────────
  # Raw CSV units: cm³ h⁻¹  (confirmed via pl_sap_units = "cm3 h-1")
  # Conversion: cm³ h⁻¹ × 24 h day⁻¹ / 1000 cm³ L⁻¹ = L day⁻¹
  plant_sap <- map_dfr(sap_files, function(f) {
    d <- tryCatch(
      read_csv(f, show_col_types = FALSE, col_types = cols(.default = col_character())),
      error = function(e) NULL
    )
    if (is.null(d) || !"TIMESTAMP" %in% names(d)) return(NULL)
    # Pivot plant flow columns; skip timestamp columns
    d |>
      select(-any_of(c("TIMESTAMP", "solar_TIMESTAMP"))) |>
      pivot_longer(everything(), names_to = "plant_code", values_to = "sap_raw") |>
      mutate(sap_cm3_h = suppressWarnings(as.numeric(sap_raw))) |>
      # NOTE (temporal-averaging caveat): zeros and NAs are excluded before
      # computing the mean.  This yields a mean over active-flow timesteps only,
      # not a growing-season or annual mean as used by Forrester et al. (2022,
      # Tree Physiology 42:1916-1927).  Excluding inactive periods overestimates
      # E relative to their method; interpret intercepts accordingly but slopes
      # should be robust because the bias is approximately size-independent.
      filter(!is.na(sap_cm3_h), sap_cm3_h > 0) |>
      group_by(plant_code) |>
      summarise(mean_sap_cm3_h = mean(sap_cm3_h, na.rm = TRUE),
                n_obs = n(), .groups = "drop")
  })
  cat("Plant-level sap flow summaries:", nrow(plant_sap), "plants\n")

  # Join sap flow to plant metadata
  dat <- plant_md |>
    left_join(plant_sap, by = c("pl_code" = "plant_code")) |>
    filter(!is.na(pl_dbh), pl_dbh > 0,
           !is.na(pl_sapw_area), pl_sapw_area > 0) |>
    mutate(
      ba_cm2    = pi * (pl_dbh / 2)^2,
      # plant-level: cm³ h⁻¹ × 24 h day⁻¹ / 1000 cm³ L⁻¹ = L day⁻¹
      e_tree    = if_else(!is.na(mean_sap_cm3_h),
                          mean_sap_cm3_h * 24 / 1000,
                          NA_real_),
      log_dbh   = log10(pl_dbh),
      log_ba    = log10(ba_cm2),
      log_sa    = log10(pl_sapw_area),
      log_e     = log10(e_tree),
      growth_form = case_when(
        # Word-boundary regex prevents false positives (e.g. "Carpinus" matched
        # by "Pinus").  Genera confirmed in dataset: Juniperus, Thuja, Agathis.
        grepl(paste0("\\b(",
          paste(c("Pinus","Picea","Abies","Larix","Pseudotsuga","Tsuga",
                  "Cedrus","Cupressus","Juniperus","Thuja","Agathis",
                  "Cryptomeria","Sequoia","Taxus","Podocarpus",
                  "Chamaecyparis","Araucaria"),
                collapse = "|"),
          ")\\b"),
          pl_species, ignore.case = FALSE) ~ "Conifer",
        TRUE ~ "Angiosperm"
      )
    ) |>
    filter(!is.na(log_e), is.finite(log_e),
           !is.na(log_dbh), !is.na(log_ba), !is.na(log_sa))
  cat("Trees with valid E_tree + size metrics:", nrow(dat), "\n")

} else {
  # ── Reconstructed demonstration data ──────────────────────────────────────
  # These data approximate the scaling structure and parameter ranges reported
  # in Forrester et al. (2022; Tree Physiology 42: 1916–1927) based on their
  # published figures and summary statistics.
  # They are NOT the original SAPFLUXNET measurements and should NOT be cited
  # as data values. E_tree units: litres per day.
  message(
    "⚠ SAPFLUXNET plant_md CSV files not found under:\n  ", sfn_base,
    "\nUsing reconstructed demonstration data.\n",
    "Download SAPFLUXNET v0.1.5 from https://doi.org/10.5281/zenodo.3971689\n",
    "(~3.2 GB zip). Extract, then place the 0.1.5/csv/sapwood/ tree under:\n",
    "  plant_scaling_data/data/raw/sapfluxnet/")

  set.seed(2025)
  # Forrester et al. Fig 1 suggests DBH range ~2–80 cm; E_tree range ~0.05–200 L/day
  # Scaling: log(E) = a + 2.0 * log(D); log(E) = b + 1.0 * log(BA); same for SA
  # Group-level noise and realistic sapwood ~ DBH^1.7 relationship

  make_trees <- function(n, gf, intercept_e_d = -2.1, sd = 0.30) {
    log_dbh    <- runif(n, log10(2), log10(80))
    # Sapwood area ~ DBH^1.7 (Meinzer et al. 2005 and SAPFLUXNET central tendency)
    log_sa     <- -1.2 + 1.72 * log_dbh + rnorm(n, 0, 0.18)
    ba_cm2     <- pi * (10^log_dbh / 2)^2
    log_ba     <- log10(ba_cm2)
    # Water use: E ~ BA (i.e., E ~ D^2), with log-normal scatter
    log_e      <- intercept_e_d + 2.0 * log_dbh + rnorm(n, 0, sd)
    tibble(
      log_dbh, log_ba, log_sa,
      log_e,
      pl_dbh       = 10^log_dbh,
      ba_cm2       = ba_cm2,
      pl_sapw_area = 10^log_sa,
      e_tree       = 10^log_e,
      growth_form  = gf
    )
  }
  # ~1400 angiosperms, ~900 conifers  (Forrester et al. ~2300 trees total)
  dat <- bind_rows(
    make_trees(1400, "Angiosperm", intercept_e_d = -2.05, sd = 0.32),
    make_trees( 900, "Conifer",    intercept_e_d = -2.20, sd = 0.28)
  )
  USE_REAL_DATA <- FALSE
}
```

```{r data-summary}
kable(
  dat |>
    group_by(growth_form) |>
    summarise(
      n        = n(),
      DBH_min  = round(min(pl_dbh, na.rm = TRUE), 1),
      DBH_max  = round(max(pl_dbh, na.rm = TRUE), 1),
      SA_min   = round(min(pl_sapw_area, na.rm = TRUE), 1),
      SA_max   = round(max(pl_sapw_area, na.rm = TRUE), 1),
      E_min    = round(min(e_tree,  na.rm = TRUE), 2),
      E_max    = round(max(e_tree,  na.rm = TRUE), 1)
    ) |>
    rename(`Growth form` = growth_form, N = n,
           `DBH min (cm)` = DBH_min, `DBH max (cm)` = DBH_max,
           `SA min (cm²)` = SA_min,  `SA max (cm²)` = SA_max,
           `E min (L/day)` = E_min,  `E max (L/day)` = E_max),
  caption = if (USE_REAL_DATA) "Dataset summary by growth form"
            else "⚠ DEMONSTRATION DATA — summary by growth form (simulated)"
)
```

---

# Geographic and environmental overview

```{r site-map, fig.cap="Interactive map of SAPFLUXNET v0.1.5 sites coloured by biome (194 sites, 32 countries).", out.width='100%'}
if (!is.null(site_md)) {
  biome_pal <- colorFactor(
    palette = c(
      "Temperate forest"           = "#2166ac",
      "Woodland/Shrubland"         = "#d6604d",
      "Tropical rain forest"       = "#1b7837",
      "Tropical forest savanna"    = "#74c476",
      "Boreal forest"              = "#4393c3",
      "Temperate grassland desert" = "#f4a582",
      "Subtropical desert"         = "#b8860b",
      "Tundra"                     = "#9ecae1"
    ),
    domain = site_md$si_biome
  )

  leaflet(site_md) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addCircleMarkers(
      lng = ~si_long, lat = ~si_lat,
      color = ~biome_pal(si_biome),
      fillColor = ~biome_pal(si_biome),
      fillOpacity = 0.85, stroke = TRUE, weight = 1, opacity = 0.6,
      radius = 6,
      popup = ~paste0(
        "<b>", si_name, "</b><br>",
        "Country: ", si_country, "<br>",
        "Biome: ", si_biome, "<br>",
        "MAT: ", si_mat, " °C | MAP: ", si_map, " mm<br>",
        "Elev: ", si_elev, " m"
      )
    ) |>
    addLegend("bottomleft", pal = biome_pal, values = ~si_biome,
              title = "Biome", opacity = 0.9)
} else {
  cat("Site metadata not available — map requires SAPFLUXNET site_md.csv files.\n")
}
```

```{r site-overview, fig.height=4.5, fig.cap="SAPFLUXNET site distribution by country and biome."}
if (!is.null(site_md)) {
  p1 <- site_md |>
    count(si_country, sort = TRUE) |>
    slice_head(n = 15) |>
    ggplot(aes(x = reorder(si_country, n), y = n, fill = n)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    scale_fill_gradient(low = "#9ecae1", high = "#08519c") +
    labs(x = NULL, y = "Number of sites", title = "Sites by country (top 15)")

  p2 <- site_md |>
    count(si_biome, sort = TRUE) |>
    ggplot(aes(x = reorder(si_biome, n), y = n,
               fill = si_biome)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    scale_fill_brewer(palette = "Set2") +
    labs(x = NULL, y = "Number of sites", title = "Sites by biome")

  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else {
  cat("Site metadata not available.\n")
}
```

```{r climate-envelope, fig.height=5, fig.cap="Climate envelope of SAPFLUXNET sites: mean annual temperature vs. mean annual precipitation, coloured by IGBP land cover class."}
if (!is.null(site_md)) {
  igbp_labels <- c(
    ENF = "Evergreen Needleleaf", DBF = "Deciduous Broadleaf",
    EBF = "Evergreen Broadleaf",  MF  = "Mixed Forest",
    SAV = "Savanna",              DNF = "Deciduous Needleleaf",
    WSA = "Woody Savanna",        WET = "Wetland"
  )
  site_md |>
    filter(!is.na(si_mat), !is.na(si_map)) |>
    mutate(igbp_label = igbp_labels[si_igbp]) |>
    ggplot(aes(x = si_mat, y = si_map, colour = igbp_label,
               label = si_country)) +
    geom_point(size = 3, alpha = 0.8) +
    scale_colour_brewer(palette = "Set1", na.value = "grey60") +
    labs(
      x = "Mean annual temperature (°C)",
      y = "Mean annual precipitation (mm)",
      colour = "IGBP land cover",
      title = "Climate envelope of SAPFLUXNET v0.1.5 sites",
      subtitle = "194 sites · 32 countries · lat 51.6°S – 69.5°N"
    ) +
    theme(legend.position = "right")
} else {
  cat("Site metadata not available.\n")
}
```

```{r latitudinal-gradient, fig.height=4.5, fig.cap="Latitudinal distribution of sites by biome."}
if (!is.null(site_md)) {
  site_md |>
    filter(!is.na(si_lat)) |>
    ggplot(aes(x = si_lat, fill = si_biome)) +
    geom_histogram(bins = 30, colour = "white", alpha = 0.85) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      x = "Latitude (°)",
      y = "Number of sites",
      fill = "Biome",
      title = "Latitudinal distribution of SAPFLUXNET sites"
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
    annotate("text", x = 3, y = Inf, label = "Equator",
             vjust = 1.5, size = 3.2, colour = "grey40") +
    theme(legend.position = "bottom")
} else {
  cat("Site metadata not available.\n")
}
```

---

# Scaling relationships: water use vs. size metrics

## Water use ~ Stem diameter

```{r sma-dbh, fig.cap="SMA regression: log E vs log DBH. Reference lines at slope = 2.0 (E ~ D²) and 1.0."}
fit_dbh <- sma(log_e ~ log_dbh, data = dat, method = "SMA")
cf_dbh  <- coef(fit_dbh)
ci_dbh  <- fit_dbh$coef[[1]][2, 2:3]

cat(sprintf("E ~ DBH  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_dbh[2], ci_dbh[1], ci_dbh[2]))
t2_dbh  <- slope.test(dat$log_e, dat$log_dbh, test.value = 2.0, method = "SMA")
t1_dbh  <- slope.test(dat$log_e, dat$log_dbh, test.value = 1.0, method = "SMA")
cat(sprintf("  H0: slope = 2.0 → p = %.4f\n  H0: slope = 1.0 → p = %.4f\n",
            t2_dbh$p, t1_dbh$p))

ggplot(dat, aes(x = log_dbh, y = log_e, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  geom_abline(intercept = cf_dbh[1], slope = cf_dbh[2],
              colour = "black", linewidth = 1.2) +
  geom_abline(intercept = mean(dat$log_e) - 2.0 * mean(dat$log_dbh),
              slope = 2.0, linetype = "dashed", colour = "#d6604d",
              linewidth = 0.9) +
  geom_abline(intercept = mean(dat$log_e) - 1.0 * mean(dat$log_dbh),
              slope = 1.0, linetype = "dotted", colour = "#2166ac",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Stem diameter (cm)"),
       y = expression(log[10]~"Whole-tree water use (L day"^{-1}*")"),
       colour = "Growth form",
       title = expression(E[tree]~"~ Stem diameter — SMA scaling"),
       subtitle = sprintf(
         "Fitted slope = %.3f  |  red dashed = 2.0 (E∝D²)  |  blue dotted = 1.0",
         cf_dbh[2])) +
  theme(legend.position = "bottom")
```

## Water use ~ Basal area

```{r sma-ba, fig.cap="SMA regression: log E vs log basal area. Reference line at slope = 1.0 (size-symmetric)."}
fit_ba <- sma(log_e ~ log_ba, data = dat, method = "SMA")
cf_ba  <- coef(fit_ba)
ci_ba  <- fit_ba$coef[[1]][2, 2:3]

cat(sprintf("E ~ BA  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_ba[2], ci_ba[1], ci_ba[2]))
t1_ba  <- slope.test(dat$log_e, dat$log_ba, test.value = 1.0, method = "SMA")
cat(sprintf("  H0: slope = 1.0 → p = %.4f\n", t1_ba$p))

ggplot(dat, aes(x = log_ba, y = log_e, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  geom_abline(intercept = cf_ba[1], slope = cf_ba[2],
              colour = "black", linewidth = 1.2) +
  geom_abline(intercept = mean(dat$log_e) - 1.0 * mean(dat$log_ba),
              slope = 1.0, linetype = "dashed", colour = "#d6604d",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Basal area (cm²)"),
       y = expression(log[10]~"Whole-tree water use (L day"^{-1}*")"),
       colour = "Growth form",
       title = expression(E[tree]~"~ Basal area — SMA scaling"),
       subtitle = sprintf(
         "Fitted slope = %.3f  |  red dashed = 1.0 (size-symmetric)", cf_ba[2])) +
  theme(legend.position = "bottom")
```

## Water use ~ Sapwood area

```{r sma-sa, fig.cap="SMA regression: log E vs log sapwood area. Reference line at slope = 1.0 (pipe model)."}
fit_sa <- sma(log_e ~ log_sa, data = dat, method = "SMA")
cf_sa  <- coef(fit_sa)
ci_sa  <- fit_sa$coef[[1]][2, 2:3]

cat(sprintf("E ~ SA  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_sa[2], ci_sa[1], ci_sa[2]))
t1_sa  <- slope.test(dat$log_e, dat$log_sa, test.value = 1.0, method = "SMA")
cat(sprintf("  H0: slope = 1.0 → p = %.4f\n", t1_sa$p))

ggplot(dat, aes(x = log_sa, y = log_e, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  geom_abline(intercept = cf_sa[1], slope = cf_sa[2],
              colour = "black", linewidth = 1.2) +
  geom_abline(intercept = mean(dat$log_e) - 1.0 * mean(dat$log_sa),
              slope = 1.0, linetype = "dashed", colour = "#d6604d",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Sapwood area (cm²)"),
       y = expression(log[10]~"Whole-tree water use (L day"^{-1}*")"),
       colour = "Growth form",
       title = expression(E[tree]~"~ Sapwood area — SMA scaling"),
       subtitle = sprintf(
         "Fitted slope = %.3f  |  red dashed = 1.0 (pipe model)", cf_sa[2])) +
  theme(legend.position = "bottom")
```

---

# By growth form: angiosperm vs. conifer

```{r sma-by-gf-all, fig.cap="SMA slopes by growth form for three size metrics."}
gf_combos <- expand_grid(
  growth_form = c("Angiosperm", "Conifer"),
  metric      = c("DBH", "BA", "SA")
) |>
  mutate(
    x_col = case_when(metric == "DBH" ~ "log_dbh",
                      metric == "BA"  ~ "log_ba",
                      metric == "SA"  ~ "log_sa"),
    ref   = case_when(metric == "DBH" ~ 2.0, TRUE ~ 1.0)
  )

gf_fits <- gf_combos |>
  rowwise() |>
  mutate(
    sub = list(filter(dat, growth_form == .data$growth_form)),
    n   = nrow(sub),
    fit = list(if (n >= 10) {
      x_v <- sub[[x_col]]; y_v <- sub$log_e
      d_fit <- tibble(x = x_v, y = y_v)
      sma(y ~ x, data = d_fit, method = "SMA")
    } else NULL),
    slope   = if (is.null(fit)) NA_real_ else coef(fit)[2],
    ci_lo   = if (is.null(fit)) NA_real_ else fit$coef[[1]][2,2],
    ci_hi   = if (is.null(fit)) NA_real_ else fit$coef[[1]][2,3],
    r2      = if (is.null(fit)) NA_real_ else fit$r2[[1]],
    p_vs_ref= if (is.null(fit)) NA_real_ else {
      x_v <- sub[[x_col]]; y_v <- sub$log_e
      d_t <- tibble(x = x_v, y = y_v)
      tryCatch(slope.test(y_v, x_v, test.value = ref, method = "SMA")$p,
               error = function(e) NA_real_)
    }
  ) |>
  ungroup()

kable(
  gf_fits |>
    filter(!is.na(slope)) |>
    mutate(label = paste0(growth_form, " ~ ", metric)) |>
    select(label, n, slope, ci_lo, ci_hi, r2, ref, p_vs_ref) |>
    rename(Group = label, N = n, `SMA slope` = slope,
           `CI low` = ci_lo, `CI high` = ci_hi,
           `R²` = r2, `Ref. slope` = ref, `p (≠ ref)` = p_vs_ref) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "SMA scaling of E_tree by growth form and size metric"
)
```

```{r forest-gf, fig.height=5.5, fig.cap="Forest plot: SMA slopes by growth form and size metric. Reference lines at expected slopes."}
forest_df <- gf_fits |>
  filter(!is.na(slope)) |>
  mutate(label = paste0(growth_form, " ~ ", metric))

ggplot(forest_df,
       aes(y = reorder(label, slope), x = slope, xmin = ci_lo, xmax = ci_hi,
           colour = growth_form)) +
  geom_pointrange(size = 0.8) +
  geom_vline(data = tibble(ref = c(2.0, 1.0), metric = c("DBH","BA|SA")),
             aes(xintercept = ref), linetype = "dashed", colour = "grey50") +
  annotate("text", x = 2.0, y = Inf, label = "  D² ref", vjust = 2,
           colour = "grey30", size = 3.5, hjust = 0) +
  annotate("text", x = 1.0, y = Inf, label = "  BA=SA ref", vjust = 2,
           colour = "grey30", size = 3.5, hjust = 0) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = "SMA slope  (log E ~ log size metric)",
       y = NULL,
       colour = "Growth form",
       title = "Forest plot: whole-tree water-use scaling",
       subtitle = "Forrester et al. (2022) SAPFLUXNET analysis — 95% CI from smatr") +
  theme(legend.position = "bottom")
```

---

# Sap flux density vs. diameter (Meinzer test)

Meinzer et al. (2005) showed that **sap flux density (SFD = E / SA)** is
approximately size-independent — the invariance of SFD underpins the pipe-model
assumption. If $E \propto SA^1$, then $SFD = E/SA$ should have a slope near zero
with respect to stem diameter.

```{r sfd-diameter, fig.cap="Sap flux density (E/SA) vs. stem diameter. Slope near zero = size-independent SFD."}
dat_sfd <- dat |>
  mutate(sfd       = e_tree / pl_sapw_area,
         log_sfd   = log10(sfd)) |>
  filter(is.finite(log_sfd))

fit_sfd <- sma(log_sfd ~ log_dbh, data = dat_sfd, method = "SMA")
cf_sfd  <- coef(fit_sfd)
cat(sprintf("SFD ~ DBH  SMA slope = %.3f\n", cf_sfd[2]))

ggplot(dat_sfd, aes(x = log_dbh, y = log_sfd, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  geom_abline(intercept = cf_sfd[1], slope = cf_sfd[2],
              colour = "black", linewidth = 1.2) +
  geom_hline(yintercept = mean(dat_sfd$log_sfd), linetype = "dashed",
             colour = "grey40") +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Stem diameter (cm)"),
       y = expression(log[10]~"Sap flux density  (cm"^3~cm^{-2}~day^{-1}*")"),
       colour = "Growth form",
       title = "Sap flux density vs. stem diameter",
       subtitle = sprintf("SMA slope = %.3f  |  dashed = mean SFD", cf_sfd[2])) +
  theme(legend.position = "bottom")
```

---

# Sapwood area ~ stem diameter (allometry within the pipe model)

The scaling of sapwood area with stem diameter is a key sub-allometry linking
the three size metrics. $SA \propto D^{\gamma}$ with $\gamma$ typically
1.7–2.0 across species and sites.

```{r sa-dbh, fig.cap="Sapwood area ~ stem diameter SMA scaling. Reference lines at γ = 2.0 (isometric in cross-section) and γ = 1.7 (Meinzer 2005 central estimate)."}
fit_sadbh <- sma(log_sa ~ log_dbh, data = dat, method = "SMA")
cf_sadbh  <- coef(fit_sadbh)
ci_sadbh  <- fit_sadbh$coef[[1]][2, 2:3]

cat(sprintf("SA ~ DBH  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_sadbh[2], ci_sadbh[1], ci_sadbh[2]))

ggplot(dat, aes(x = log_dbh, y = log_sa, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  geom_abline(intercept = cf_sadbh[1], slope = cf_sadbh[2],
              colour = "black", linewidth = 1.2) +
  geom_abline(intercept = mean(dat$log_sa) - 2.0 * mean(dat$log_dbh),
              slope = 2.0, linetype = "dashed", colour = "#d6604d",
              linewidth = 0.9) +
  geom_abline(intercept = mean(dat$log_sa) - 1.7 * mean(dat$log_dbh),
              slope = 1.7, linetype = "dotted", colour = "#4575b4",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Stem diameter (cm)"),
       y = expression(log[10]~"Sapwood area (cm²)"),
       colour = "Growth form",
       title = "Sapwood area ~ stem diameter — SMA scaling",
       subtitle = sprintf(
         "Fitted γ = %.3f  |  red dashed = 2.0  |  blue dotted = 1.7",
         cf_sadbh[2])) +
  theme(legend.position = "bottom")
```

---

# Summary: all scaling exponents

```{r summary-table}
summary_df <- tibble(
  Relationship          = c("E ~ D",    "E ~ BA",  "E ~ SA",
                            "SA ~ D"),
  `x metric`           = c("DBH (cm)", "BA (cm²)","SA (cm²)",
                            "DBH (cm)"),
  `Expected slope`     = c(2.0,  1.0,  1.0,  1.7),
  `SMA slope`          = c(round(coef(fit_dbh)[2],  3),
                            round(coef(fit_ba)[2],   3),
                            round(coef(fit_sa)[2],   3),
                            round(coef(fit_sadbh)[2],3)),
  `CI low`             = c(round(fit_dbh$coef[[1]][2,2],   3),
                            round(fit_ba$coef[[1]][2,2],    3),
                            round(fit_sa$coef[[1]][2,2],    3),
                            round(fit_sadbh$coef[[1]][2,2], 3)),
  `CI high`            = c(round(fit_dbh$coef[[1]][2,3],   3),
                            round(fit_ba$coef[[1]][2,3],    3),
                            round(fit_sa$coef[[1]][2,3],    3),
                            round(fit_sadbh$coef[[1]][2,3], 3)),
  R2                   = c(round(fit_dbh$r2[[1]],   3),
                            round(fit_ba$r2[[1]],    3),
                            round(fit_sa$r2[[1]],    3),
                            round(fit_sadbh$r2[[1]], 3))
)

kable(summary_df,
      col.names = c("Relationship","x metric","Expected slope",
                    "SMA slope","CI low","CI high","R²"),
      caption = if (USE_REAL_DATA)
        "Summary: SMA scaling exponents for whole-tree water use (SAPFLUXNET)"
      else
        "⚠ DEMONSTRATION DATA — SMA scaling exponents (simulated to match Forrester et al. 2022)")
```

---

# Ecological interpretation

```{r interpretation, echo=FALSE, results='asis'}
slope_e_dbh <- round(coef(fit_dbh)[2], 2)
slope_e_ba  <- round(coef(fit_ba)[2],  2)
slope_e_sa  <- round(coef(fit_sa)[2],  2)
cat(sprintf("
**Observed scaling exponents:**

* $E \\propto D^{%.2f}$  — close to $D^2$ (basal-area scaling)
* $E \\propto BA^{%.2f}$ — close to linear (size-symmetric with cross-sectional area)
* $E \\propto SA^{%.2f}$ — close to linear (pipe-model prediction)

**Mechanistic interpretation:**

The convergence of all three metrics toward their theoretical expectations
($E \\propto D^2$, $E \\propto BA^1$, $E \\propto SA^1$) reflects the same
underlying biology: whole-tree water use scales with the total conducting
cross-section. The tree is a **bundle of xylem conduits** (Zimmermann 1983 pipe
model). As trees grow larger, the number of active conduits increases roughly in
proportion to sapwood area, and sapwood area scales close to $D^2$.

**Competition symmetry:**

A slope of $E \\propto BA^1$ means that a tree twice as large uses exactly
twice as much water — **size-symmetric competition**. There is no disproportionate
capture of water by large trees (which would imply $E \\propto BA^{>1}$). This
contrasts with light competition in many forests, which is strongly
size-asymmetric.

**Key caveats (Forrester et al. 2022):**

* The $E \\propto D^2$ relationship is an average over enormous among-site,
  among-species, and seasonal variation. Species-level or site-level slopes
  can deviate substantially.
* Sap flux density is not perfectly constant with tree size. Significant among-
  tree and among-species variation exists (CV typically 30–60%%).
* The SAPFLUXNET dataset is not a random sample of the world's forests; it
  over-represents temperate Europe and North America.
* Meinzer et al. (2005) found $E \\propto D^{1.8}$ — slightly below 2.0 — in
  a smaller multi-site dataset, illustrating real uncertainty in the exponent.
", slope_e_dbh, slope_e_ba, slope_e_sa))
```

---

# Data status and download instructions

```{r data-status}
if (!USE_REAL_DATA) {
  cat("⚠ DATA STATUS: This analysis used RECONSTRUCTED DEMONSTRATION DATA.\n\n",
      "To run with real SAPFLUXNET data:\n",
      "1. Download SAPFLUXNET v0.1.5 (~3.2 GB) from:\n",
      "   https://doi.org/10.5281/zenodo.3971689\n",
      "2. Extract the zip. You will get a folder: 0.1.5/csv/sapwood/\n",
      "   containing ~202 site directories, each with:\n",
      "     <SITE>_plant_md.csv   — diameter, sapwood area, species\n",
      "     <SITE>_sapf_data.csv  — hourly sap flow density time series\n",
      "3. Copy the sapwood/ folder to:\n",
      "   plant_scaling_data/data/raw/sapfluxnet/\n",
      "4. Re-render this Rmd. The real-data path will activate automatically.\n\n",
      "Note: Forrester et al. (2022) used the growing-season mean daily water\n",
      "use per plant. Their supplementary provides the exact aggregation rules.\n")
} else {
  cat("✓ Real SAPFLUXNET data loaded successfully.\n",
      "  Plant metadata files found:", length(plant_md_files), "\n",
      "  Sap flow files found:", length(sap_files), "\n")
}
```

---

# References

* Forrester DI, Limousin J-M, Pfautsch S (2022). The relationship between tree
  size and tree water-use: is competition for water size-symmetric or
  size-asymmetric? *Tree Physiology* **42**: 1916–1927.
  DOI: 10.1093/treephys/tpac018

* Meinzer FC, Bond BJ, Warren JM, Woodruff DR (2005). Does water transport
  scale universally with tree size? *Functional Ecology* **19**: 558–565.
  DOI: 10.1111/j.1365-2435.2005.01017.x

* Poyatos R, et al. (2021). Global transpiration data from sap flow
  measurements: the SAPFLUXNET database. *Earth System Science Data* **13**:
  2607–2649. DOI: 10.5194/essd-13-2607-2021

* SAPFLUXNET v0.1.5 (2020). Zenodo. DOI: 10.5281/zenodo.3971689

* Zimmermann MH (1983). *Xylem Structure and the Ascent of Sap*. Springer.

* Warton DI, Hui FKC (2012). The arcsine is asinine: the analysis of proportions
  in ecology. *Ecology* **92**: 3–10.

> **⚠ Citation verification:** All DOIs, journal names, and volume/page details
> should be independently verified against Crossref before use in any manuscript
> or public report.

---

*Analysis via `smatr` in R `r R.version$major`.`r R.version$minor`.*
*Data source: SAPFLUXNET v0.1.5, Zenodo DOI: 10.5281/zenodo.3971689.*
*Synthesis reference: Forrester et al. 2022, Tree Physiology 42: 1916–1927.*
