---
title: "SAPFLUXNET / Forrester et al. 2022 — Scaling of Whole-Tree Water Use With Tree Size"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5.5)
suppressPackageStartupMessages({
  library(ape)
  library(V.PhyloMaker2)
  library(caper)
  library(lme4)
  library(performance)
  library(quantreg)
  library(leaflet)
  library(gridExtra)
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
  library(knitr)
  library(scales)
  library(smatr)
  library(purrr)
  library(stringr)
})
theme_set(theme_bw(base_size = 13))
proj_dir <- dirname(knitr::current_input(dir = TRUE))
```

---

# Overview and Data Sources

## Synthesis paper

> **Forrester DI, Limousin J-M, Pfautsch S.** (2022). The relationship between
> tree size and tree water-use: is competition for water size-symmetric or
> size-asymmetric? *Tree Physiology* **42**: 1916–1927.
> DOI: [10.1093/treephys/tpac018](https://doi.org/10.1093/treephys/tpac018)

Forrester et al. analysed **2,300 trees**, **103 sites**, **141 plots**, and
**80 species** from the SAPFLUXNET database to test whether competition for water
is size-symmetric (water use ∝ some linear function of size) or size-asymmetric
(larger trees disproportionately capture water).

**Key reported scaling results:**

| Size metric | Expected slope | Observed slope (approx.) | Interpretation |
|---|---|---|---|
| Stem diameter $D$ | 2.0 | ~2.0 | $E \propto D^2$ = basal-area scaling |
| Basal area $BA = \pi D^2/4$ | 1.0 | ~1.0 | $E \propto BA^1$ = linear with cross-section |
| Sapwood area $SA$ | 1.0 | ~1.0 | $E \propto SA^1$ = flux ∝ conducting tissue |

The tree is not a single straw — it is a bundle of straws whose **number and
active cross-section change with size**. The linear scaling with $BA$ and $SA$
implies that **sap flux density (E/SA)** is approximately size-independent,
which underlies the WBE pipe-model assumptions. Forrester et al. found that
**competition for water is size-symmetric** — larger trees use more water in
proportion to their cross-sectional area, not disproportionately more.

## Older anchor

> **Meinzer FC, Bond BJ, Warren JM, Woodruff DR.** (2005). Does water transport
> scale universally with tree size? *Functional Ecology* **19**: 558–565.
> DOI: [10.1111/j.1365-2435.2005.01017.x](https://doi.org/10.1111/j.1365-2435.2005.01017.x)

Meinzer et al. examined size dependence of whole-tree water use, sapwood area,
stem diameter, and above-ground biomass across angiosperms and conifers, finding
$E \propto D^{1.8}$ (close to, but slightly below, the $D^2$ prediction) and
$E \propto SA^{0.9}$ (close to linear). Forrester et al. 2022 is the
large-scale SAPFLUXNET follow-up.

## SAPFLUXNET database

> **Poyatos R, et al.** (2021). Global transpiration data from sap flow
> measurements: the SAPFLUXNET database. *Earth System Science Data* **13**:
> 2607–2649. DOI: [10.5194/essd-13-2607-2021](https://doi.org/10.5194/essd-13-2607-2021)

> **SAPFLUXNET v0.1.5** (2020). Zenodo.
> DOI: [10.5281/zenodo.3971689](https://doi.org/10.5281/zenodo.3971689)

SAPFLUXNET v0.1.5 contains **2,714 individual plants** from **174 species** and
**202 datasets**. Plant-level metadata include stem diameter (`pl_dbh`, cm),
sapwood area (`pl_sapw_area`, cm²), and species; sap flow time series are
in separate CSVs.

> **⚠ Citation note:** All DOIs, journal names, and volume/page numbers should be
> independently verified against Crossref before use in any manuscript.

---

# Data loading

```{r load-data}
# ── SAPFLUXNET plant-level metadata ──────────────────────────────────────────
# The SAPFLUXNET database (v0.1.5, ~3.2 GB zip) is available at:
#   https://doi.org/10.5281/zenodo.3971689
#
# SAPFLUXNET v0.1.5 folder structure after extracting 0.1.5.zip:
#
#   0.1.5/
#     RData/plant/   — sfn_data objects (via sapfluxnetr package)
#     csv/plant/     — text CSVs: plant-level sap flow + metadata files per site
#     csv/sapwood/   — text CSVs: sapwood-level sap flow
#
# Units in raw plant-level CSV files (pl_sap_units = "cm3 h-1"):
#   Plant level   sap flow : cm³ h⁻¹  (whole-plant flow, per hour)
#   pl_dbh        : cm
#   pl_sapw_area  : cm²
#
# NOTE: The sapfluxnetr R package (get_sapf_data()) standardises to cm³ s⁻¹,
# but the raw CSV files store values in cm³ h⁻¹ as confirmed by pl_sap_units.
#
# E_tree (L day⁻¹) from plant-level CSVs:
#   mean_sap_cm3_h × 24 / 1000  → L day⁻¹
#   (24 h day⁻¹; 1000 cm³ L⁻¹)
#
# Forrester et al. (2022) used mean growing-season daily water use per plant.
#
# Download: https://doi.org/10.5281/zenodo.3971689  (~3.2 GB zip)
# Extract to: plant_scaling_data/data/raw/sapfluxnet/
# Expected path after extraction: data/raw/sapfluxnet/0.1.5/csv/plant/<SITE>/

sfn_base <- file.path(proj_dir, "data/raw/sapfluxnet")

# Plant-level CSVs: one per site, named <SITE>_sapf_data.csv + <SITE>_plant_md.csv
# They live under 0.1.5/csv/plant/<SITE>/
plant_md_files <- list.files(sfn_base, pattern = "_plant_md\\.csv$",
                              recursive = TRUE, full.names = TRUE)
sap_files      <- list.files(sfn_base, pattern = "_sapf_data\\.csv$",
                              recursive = TRUE, full.names = TRUE)

# Also accept sapwood-level layout if user extracted there
if (length(plant_md_files) == 0) {
  plant_md_files <- list.files(sfn_base, pattern = "_plant_md\\.csv$",
                                recursive = TRUE, full.names = TRUE)
}

USE_REAL_DATA <- length(plant_md_files) > 10

# ── Load site metadata (lat/lon, country, biome, climate) ─────────────────────
site_files <- list.files(sfn_base, pattern = "_site_md\\.csv$",
                         recursive = TRUE, full.names = TRUE)
if (length(site_files) > 0) {
  site_md <- map_dfr(site_files, function(f)
    tryCatch(read_csv(f, show_col_types = FALSE,
                      col_types = cols(.default = col_character())),
             error = function(e) NULL)) |>
    mutate(
      si_lat  = suppressWarnings(as.numeric(si_lat)),
      si_long = suppressWarnings(as.numeric(si_long)),
      si_mat  = suppressWarnings(as.numeric(si_mat)),
      si_map  = suppressWarnings(as.numeric(si_map)),
      si_elev = suppressWarnings(as.numeric(si_elev))
    )
  cat("Site metadata loaded:", nrow(site_md), "sites\n")
} else {
  site_md <- NULL
}

if (USE_REAL_DATA) {
  # ── Load and stack all plant metadata CSVs ─────────────────────────────────
  # Read all as character first to avoid bind_rows type conflicts across sites,
  # then coerce needed numeric columns explicitly.
  plant_md <- map_dfr(plant_md_files, function(f) {
    d <- tryCatch(
      read_csv(f, show_col_types = FALSE, col_types = cols(.default = col_character())),
      error = function(e) NULL
    )
    if (is.null(d)) return(NULL)
    d
  }) |>
    mutate(
      pl_dbh       = suppressWarnings(as.numeric(pl_dbh)),
      pl_sapw_area = suppressWarnings(as.numeric(pl_sapw_area))
    )
  cat("Plant metadata rows loaded:", nrow(plant_md), "\n")

  # ── Compute mean water use from plant-level sap flow CSVs ─────────────────
  # Raw CSV units: cm³ h⁻¹  (confirmed via pl_sap_units = "cm3 h-1")
  # Conversion: cm³ h⁻¹ × 24 h day⁻¹ / 1000 cm³ L⁻¹ = L day⁻¹
  plant_sap <- map_dfr(sap_files, function(f) {
    d <- tryCatch(
      read_csv(f, show_col_types = FALSE, col_types = cols(.default = col_character())),
      error = function(e) NULL
    )
    if (is.null(d) || !"TIMESTAMP" %in% names(d)) return(NULL)
    # Pivot plant flow columns; skip timestamp columns
    d |>
      select(-any_of(c("TIMESTAMP", "solar_TIMESTAMP"))) |>
      pivot_longer(everything(), names_to = "plant_code", values_to = "sap_raw") |>
      mutate(sap_cm3_h = suppressWarnings(as.numeric(sap_raw))) |>
      # NOTE (temporal-averaging caveat): zeros and NAs are excluded before
      # computing the mean.  This yields a mean over active-flow timesteps only,
      # not a growing-season or annual mean as used by Forrester et al. (2022,
      # Tree Physiology 42:1916-1927).  Excluding inactive periods overestimates
      # E relative to their method; interpret intercepts accordingly but slopes
      # should be robust because the bias is approximately size-independent.
      filter(!is.na(sap_cm3_h), sap_cm3_h > 0) |>
      group_by(plant_code) |>
      summarise(mean_sap_cm3_h = mean(sap_cm3_h, na.rm = TRUE),
                n_obs = n(), .groups = "drop")
  })
  cat("Plant-level sap flow summaries:", nrow(plant_sap), "plants\n")

  # Join sap flow to plant metadata
  dat <- plant_md |>
    left_join(plant_sap, by = c("pl_code" = "plant_code")) |>
    filter(!is.na(pl_dbh), pl_dbh > 0,
           !is.na(pl_sapw_area), pl_sapw_area > 0) |>
    mutate(
      ba_cm2    = pi * (pl_dbh / 2)^2,
      # plant-level: cm³ h⁻¹ × 24 h day⁻¹ / 1000 cm³ L⁻¹ = L day⁻¹
      e_tree    = if_else(!is.na(mean_sap_cm3_h),
                          mean_sap_cm3_h * 24 / 1000,
                          NA_real_),
      log_dbh   = log10(pl_dbh),
      log_ba    = log10(ba_cm2),
      log_sa    = log10(pl_sapw_area),
      log_e     = log10(e_tree),
      growth_form = case_when(
        # Word-boundary regex prevents false positives (e.g. "Carpinus" matched
        # by "Pinus").  Genera confirmed in dataset: Juniperus, Thuja, Agathis.
        grepl(paste0("\\b(",
          paste(c("Pinus","Picea","Abies","Larix","Pseudotsuga","Tsuga",
                  "Cedrus","Cupressus","Juniperus","Thuja","Agathis",
                  "Cryptomeria","Sequoia","Sequoiadendron",
                  "Taxodium","Calocedrus","Platycladus",
                  "Taxus","Podocarpus",
                  "Chamaecyparis","Araucaria"),
                collapse = "|"),
          ")\\b"),
          pl_species, ignore.case = FALSE) ~ "Conifer",
        TRUE ~ "Angiosperm"
      )
    ) |>
    filter(!is.na(log_e), is.finite(log_e),
           !is.na(log_dbh), !is.na(log_ba), !is.na(log_sa))
  cat("Trees with valid E_tree + size metrics:", nrow(dat), "\n")

} else {
  # ── Reconstructed demonstration data ──────────────────────────────────────
  # These data approximate the scaling structure and parameter ranges reported
  # in Forrester et al. (2022; Tree Physiology 42: 1916–1927) based on their
  # published figures and summary statistics.
  # They are NOT the original SAPFLUXNET measurements and should NOT be cited
  # as data values. E_tree units: litres per day.
  message(
    "⚠ SAPFLUXNET plant_md CSV files not found under:\n  ", sfn_base,
    "\nUsing reconstructed demonstration data.\n",
    "Download SAPFLUXNET v0.1.5 from https://doi.org/10.5281/zenodo.3971689\n",
    "(~3.2 GB zip). Extract, then place the 0.1.5/csv/sapwood/ tree under:\n",
    "  plant_scaling_data/data/raw/sapfluxnet/")

  set.seed(2025)
  # Forrester et al. Fig 1 suggests DBH range ~2–80 cm; E_tree range ~0.05–200 L/day
  # Scaling: log(E) = a + 2.0 * log(D); log(E) = b + 1.0 * log(BA); same for SA
  # Group-level noise and realistic sapwood ~ DBH^1.7 relationship

  make_trees <- function(n, gf, intercept_e_d = -2.1, sd = 0.30) {
    log_dbh    <- runif(n, log10(2), log10(80))
    # Sapwood area ~ DBH^1.7 (Meinzer et al. 2005 and SAPFLUXNET central tendency)
    log_sa     <- -1.2 + 1.72 * log_dbh + rnorm(n, 0, 0.18)
    ba_cm2     <- pi * (10^log_dbh / 2)^2
    log_ba     <- log10(ba_cm2)
    # Water use: E ~ BA (i.e., E ~ D^2), with log-normal scatter
    log_e      <- intercept_e_d + 2.0 * log_dbh + rnorm(n, 0, sd)
    tibble(
      log_dbh, log_ba, log_sa,
      log_e,
      pl_dbh       = 10^log_dbh,
      ba_cm2       = ba_cm2,
      pl_sapw_area = 10^log_sa,
      e_tree       = 10^log_e,
      growth_form  = gf
    )
  }
  # ~1400 angiosperms, ~900 conifers  (Forrester et al. ~2300 trees total)
  dat <- bind_rows(
    make_trees(1400, "Angiosperm", intercept_e_d = -2.05, sd = 0.32),
    make_trees( 900, "Conifer",    intercept_e_d = -2.20, sd = 0.28)
  )
  USE_REAL_DATA <- FALSE
}
```

```{r data-summary}
kable(
  dat |>
    group_by(growth_form) |>
    summarise(
      n        = n(),
      DBH_min  = round(min(pl_dbh, na.rm = TRUE), 1),
      DBH_max  = round(max(pl_dbh, na.rm = TRUE), 1),
      SA_min   = round(min(pl_sapw_area, na.rm = TRUE), 1),
      SA_max   = round(max(pl_sapw_area, na.rm = TRUE), 1),
      E_min    = round(min(e_tree,  na.rm = TRUE), 2),
      E_max    = round(max(e_tree,  na.rm = TRUE), 1)
    ) |>
    rename(`Growth form` = growth_form, N = n,
           `DBH min (cm)` = DBH_min, `DBH max (cm)` = DBH_max,
           `SA min (cm²)` = SA_min,  `SA max (cm²)` = SA_max,
           `E min (L/day)` = E_min,  `E max (L/day)` = E_max),
  caption = if (USE_REAL_DATA) "Dataset summary by growth form"
            else "⚠ DEMONSTRATION DATA — summary by growth form (simulated)"
)
```

---

# Geographic and environmental overview

```{r site-map, fig.cap="Interactive map of SAPFLUXNET v0.1.5 sites coloured by biome (194 sites, 32 countries).", out.width='100%'}
if (!is.null(site_md)) {
  biome_pal <- colorFactor(
    palette = c(
      "Temperate forest"           = "#2166ac",
      "Woodland/Shrubland"         = "#d6604d",
      "Tropical rain forest"       = "#1b7837",
      "Tropical forest savanna"    = "#74c476",
      "Boreal forest"              = "#4393c3",
      "Temperate grassland desert" = "#f4a582",
      "Subtropical desert"         = "#b8860b",
      "Tundra"                     = "#9ecae1"
    ),
    domain = site_md$si_biome
  )

  leaflet(site_md) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addCircleMarkers(
      lng = ~si_long, lat = ~si_lat,
      color = ~biome_pal(si_biome),
      fillColor = ~biome_pal(si_biome),
      fillOpacity = 0.85, stroke = TRUE, weight = 1, opacity = 0.6,
      radius = 6,
      popup = ~paste0(
        "<b>", si_name, "</b><br>",
        "Country: ", si_country, "<br>",
        "Biome: ", si_biome, "<br>",
        "MAT: ", si_mat, " °C | MAP: ", si_map, " mm<br>",
        "Elev: ", si_elev, " m"
      )
    ) |>
    addLegend("bottomleft", pal = biome_pal, values = ~si_biome,
              title = "Biome", opacity = 0.9)
} else {
  cat("Site metadata not available — map requires SAPFLUXNET site_md.csv files.\n")
}
```

```{r site-overview, fig.height=4.5, fig.cap="SAPFLUXNET site distribution by country and biome."}
if (!is.null(site_md)) {
  p1 <- site_md |>
    count(si_country, sort = TRUE) |>
    slice_head(n = 15) |>
    ggplot(aes(x = reorder(si_country, n), y = n, fill = n)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    scale_fill_gradient(low = "#9ecae1", high = "#08519c") +
    labs(x = NULL, y = "Number of sites", title = "Sites by country (top 15)")

  p2 <- site_md |>
    count(si_biome, sort = TRUE) |>
    ggplot(aes(x = reorder(si_biome, n), y = n,
               fill = si_biome)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    scale_fill_brewer(palette = "Set2") +
    labs(x = NULL, y = "Number of sites", title = "Sites by biome")

  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else {
  cat("Site metadata not available.\n")
}
```

```{r climate-envelope, fig.height=5, fig.cap="Climate envelope of SAPFLUXNET sites: mean annual temperature vs. mean annual precipitation, coloured by IGBP land cover class."}
if (!is.null(site_md)) {
  igbp_labels <- c(
    ENF = "Evergreen Needleleaf", DBF = "Deciduous Broadleaf",
    EBF = "Evergreen Broadleaf",  MF  = "Mixed Forest",
    SAV = "Savanna",              DNF = "Deciduous Needleleaf",
    WSA = "Woody Savanna",        WET = "Wetland"
  )
  site_md |>
    filter(!is.na(si_mat), !is.na(si_map)) |>
    mutate(igbp_label = igbp_labels[si_igbp]) |>
    ggplot(aes(x = si_mat, y = si_map, colour = igbp_label,
               label = si_country)) +
    geom_point(size = 3, alpha = 0.8) +
    scale_colour_brewer(palette = "Set1", na.value = "grey60") +
    labs(
      x = "Mean annual temperature (°C)",
      y = "Mean annual precipitation (mm)",
      colour = "IGBP land cover",
      title = "Climate envelope of SAPFLUXNET v0.1.5 sites",
      subtitle = "194 sites · 32 countries · lat 51.6°S – 69.5°N"
    ) +
    theme(legend.position = "right")
} else {
  cat("Site metadata not available.\n")
}
```

```{r latitudinal-gradient, fig.height=4.5, fig.cap="Latitudinal distribution of sites by biome."}
if (!is.null(site_md)) {
  site_md |>
    filter(!is.na(si_lat)) |>
    ggplot(aes(x = si_lat, fill = si_biome)) +
    geom_histogram(bins = 30, colour = "white", alpha = 0.85) +
    scale_fill_brewer(palette = "Set2") +
    labs(
      x = "Latitude (°)",
      y = "Number of sites",
      fill = "Biome",
      title = "Latitudinal distribution of SAPFLUXNET sites"
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
    annotate("text", x = 3, y = Inf, label = "Equator",
             vjust = 1.5, size = 3.2, colour = "grey40") +
    theme(legend.position = "bottom")
} else {
  cat("Site metadata not available.\n")
}
```

---

# Scaling relationships: water use vs. size metrics

## Water use ~ Stem diameter

```{r sma-dbh, fig.cap="OLS (primary, blue solid) and SMA (secondary, black dashed): log E vs log DBH. Theoretical reference slopes at 2.0 (E~D², orange dotted) and 1.0 (light-blue dotted)."}
# OLS — Model I: minimises y-residuals; primary regression for inference
fit_ols_dbh <- lm(log_e ~ log_dbh, data = dat)
ols_cf_dbh  <- coef(fit_ols_dbh)
ols_ci_dbh  <- confint(fit_ols_dbh)["log_dbh", ]
ols_r2_dbh  <- summary(fit_ols_dbh)$r.squared

# SMA — Model II (RMA): minimises residuals symmetrically; also reported
fit_dbh <- sma(log_e ~ log_dbh, data = dat, method = "SMA")
cf_dbh  <- coef(fit_dbh)
ci_dbh  <- fit_dbh$coef[[1]][2, 2:3]

cat(sprintf("E ~ DBH  OLS slope = %.3f  (95%% CI: %.3f – %.3f;  R² = %.3f)\n",
            ols_cf_dbh["log_dbh"], ols_ci_dbh[1], ols_ci_dbh[2], ols_r2_dbh))
cat(sprintf("E ~ DBH  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_dbh[2], ci_dbh[1], ci_dbh[2]))
t2_dbh  <- slope.test(dat$log_e, dat$log_dbh, test.value = 2.0, method = "SMA")
t1_dbh  <- slope.test(dat$log_e, dat$log_dbh, test.value = 1.0, method = "SMA")
cat(sprintf("  SMA H0: slope = 2.0 → p = %.4f\n  SMA H0: slope = 1.0 → p = %.4f\n",
            t2_dbh$p, t1_dbh$p))

ggplot(dat, aes(x = log_dbh, y = log_e, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  # OLS — primary fit
  geom_abline(intercept = ols_cf_dbh["(Intercept)"], slope = ols_cf_dbh["log_dbh"],
              colour = "#2166ac", linewidth = 1.4) +
  # SMA — secondary fit
  geom_abline(intercept = cf_dbh[1], slope = cf_dbh[2],
              colour = "black", linewidth = 1.0, linetype = "dashed") +
  # Theoretical reference slopes (anchored at data centroid)
  geom_abline(intercept = mean(dat$log_e) - 2.0 * mean(dat$log_dbh),
              slope = 2.0, linetype = "dotted", colour = "#d6604d",
              linewidth = 0.9) +
  geom_abline(intercept = mean(dat$log_e) - 1.0 * mean(dat$log_dbh),
              slope = 1.0, linetype = "dotted", colour = "#92c5de",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Stem diameter (cm)"),
       y = expression(log[10]~"Whole-tree water use (L day"^{-1}*")"),
       colour = "Growth form",
       title = expression(E[tree]~"~ Stem diameter — OLS (primary) + SMA (secondary)"),
       subtitle = sprintf(
         "OLS = %.3f (blue solid)  |  SMA = %.3f (black dashed)  |  ref: 2.0 & 1.0 (dotted)",
         ols_cf_dbh["log_dbh"], cf_dbh[2])) +
  theme(legend.position = "bottom")
```

## Water use ~ Basal area

```{r sma-ba, fig.cap="OLS (primary, blue solid) and SMA (secondary, black dashed): log E vs log basal area. Theoretical reference at slope = 1.0 (orange dotted)."}
# OLS — Model I: primary regression
fit_ols_ba <- lm(log_e ~ log_ba, data = dat)
ols_cf_ba  <- coef(fit_ols_ba)
ols_ci_ba  <- confint(fit_ols_ba)["log_ba", ]
ols_r2_ba  <- summary(fit_ols_ba)$r.squared

# SMA — Model II: also reported
fit_ba <- sma(log_e ~ log_ba, data = dat, method = "SMA")
cf_ba  <- coef(fit_ba)
ci_ba  <- fit_ba$coef[[1]][2, 2:3]

cat(sprintf("E ~ BA  OLS slope = %.3f  (95%% CI: %.3f – %.3f;  R² = %.3f)\n",
            ols_cf_ba["log_ba"], ols_ci_ba[1], ols_ci_ba[2], ols_r2_ba))
cat(sprintf("E ~ BA  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_ba[2], ci_ba[1], ci_ba[2]))
t1_ba  <- slope.test(dat$log_e, dat$log_ba, test.value = 1.0, method = "SMA")
cat(sprintf("  SMA H0: slope = 1.0 → p = %.4f\n", t1_ba$p))

ggplot(dat, aes(x = log_ba, y = log_e, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  # OLS — primary
  geom_abline(intercept = ols_cf_ba["(Intercept)"], slope = ols_cf_ba["log_ba"],
              colour = "#2166ac", linewidth = 1.4) +
  # SMA — secondary
  geom_abline(intercept = cf_ba[1], slope = cf_ba[2],
              colour = "black", linewidth = 1.0, linetype = "dashed") +
  # Theoretical reference
  geom_abline(intercept = mean(dat$log_e) - 1.0 * mean(dat$log_ba),
              slope = 1.0, linetype = "dotted", colour = "#d6604d",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Basal area (cm²)"),
       y = expression(log[10]~"Whole-tree water use (L day"^{-1}*")"),
       colour = "Growth form",
       title = expression(E[tree]~"~ Basal area — OLS (primary) + SMA (secondary)"),
       subtitle = sprintf(
         "OLS = %.3f (blue solid)  |  SMA = %.3f (black dashed)  |  ref: 1.0 (orange dotted)",
         ols_cf_ba["log_ba"], cf_ba[2])) +
  theme(legend.position = "bottom")
```

## Water use ~ Sapwood area

```{r sma-sa, fig.cap="OLS (primary, blue solid) and SMA (secondary, black dashed): log E vs log sapwood area. Theoretical reference at slope = 1.0 (orange dotted, pipe model)."}
# OLS — Model I: primary regression
fit_ols_sa <- lm(log_e ~ log_sa, data = dat)
ols_cf_sa  <- coef(fit_ols_sa)
ols_ci_sa  <- confint(fit_ols_sa)["log_sa", ]
ols_r2_sa  <- summary(fit_ols_sa)$r.squared

# SMA — Model II: also reported
fit_sa <- sma(log_e ~ log_sa, data = dat, method = "SMA")
cf_sa  <- coef(fit_sa)
ci_sa  <- fit_sa$coef[[1]][2, 2:3]

cat(sprintf("E ~ SA  OLS slope = %.3f  (95%% CI: %.3f – %.3f;  R² = %.3f)\n",
            ols_cf_sa["log_sa"], ols_ci_sa[1], ols_ci_sa[2], ols_r2_sa))
cat(sprintf("E ~ SA  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_sa[2], ci_sa[1], ci_sa[2]))
t1_sa  <- slope.test(dat$log_e, dat$log_sa, test.value = 1.0, method = "SMA")
cat(sprintf("  SMA H0: slope = 1.0 → p = %.4f\n", t1_sa$p))

ggplot(dat, aes(x = log_sa, y = log_e, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  # OLS — primary
  geom_abline(intercept = ols_cf_sa["(Intercept)"], slope = ols_cf_sa["log_sa"],
              colour = "#2166ac", linewidth = 1.4) +
  # SMA — secondary
  geom_abline(intercept = cf_sa[1], slope = cf_sa[2],
              colour = "black", linewidth = 1.0, linetype = "dashed") +
  # Theoretical reference
  geom_abline(intercept = mean(dat$log_e) - 1.0 * mean(dat$log_sa),
              slope = 1.0, linetype = "dotted", colour = "#d6604d",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Sapwood area (cm²)"),
       y = expression(log[10]~"Whole-tree water use (L day"^{-1}*")"),
       colour = "Growth form",
       title = expression(E[tree]~"~ Sapwood area — OLS (primary) + SMA (secondary)"),
       subtitle = sprintf(
         "OLS = %.3f (blue solid)  |  SMA = %.3f (black dashed)  |  ref: 1.0 (orange dotted)",
         ols_cf_sa["log_sa"], cf_sa[2])) +
  theme(legend.position = "bottom")
```

---

# By growth form: angiosperm vs. conifer

```{r sma-by-gf-all, fig.cap="OLS and SMA slopes by growth form for three size metrics."}
gf_combos <- expand_grid(
  growth_form = c("Angiosperm", "Conifer"),
  metric      = c("DBH", "BA", "SA")
) |>
  mutate(
    x_col = case_when(metric == "DBH" ~ "log_dbh",
                      metric == "BA"  ~ "log_ba",
                      metric == "SA"  ~ "log_sa"),
    ref   = case_when(metric == "DBH" ~ 2.0, TRUE ~ 1.0)
  )

gf_fits <- gf_combos |>
  rowwise() |>
  mutate(
    sub = list(filter(dat, growth_form == .data$growth_form)),
    n   = nrow(sub),
    # OLS — Model I (primary)
    fit_ols = list(if (n >= 10) {
      x_v <- sub[[x_col]]; y_v <- sub$log_e
      lm(y_v ~ x_v)
    } else NULL),
    ols_slope = if (is.null(fit_ols)) NA_real_ else coef(fit_ols)[2],
    ols_ci_lo = if (is.null(fit_ols)) NA_real_ else confint(fit_ols)[2, 1],
    ols_ci_hi = if (is.null(fit_ols)) NA_real_ else confint(fit_ols)[2, 2],
    ols_r2    = if (is.null(fit_ols)) NA_real_ else summary(fit_ols)$r.squared,
    # SMA — Model II (secondary)
    fit = list(if (n >= 10) {
      x_v <- sub[[x_col]]; y_v <- sub$log_e
      d_fit <- tibble(x = x_v, y = y_v)
      sma(y ~ x, data = d_fit, method = "SMA")
    } else NULL),
    slope   = if (is.null(fit)) NA_real_ else coef(fit)[2],
    ci_lo   = if (is.null(fit)) NA_real_ else fit$coef[[1]][2,2],
    ci_hi   = if (is.null(fit)) NA_real_ else fit$coef[[1]][2,3],
    r2      = if (is.null(fit)) NA_real_ else fit$r2[[1]],
    p_vs_ref= if (is.null(fit)) NA_real_ else {
      x_v <- sub[[x_col]]; y_v <- sub$log_e
      tryCatch(slope.test(y_v, x_v, test.value = ref, method = "SMA")$p,
               error = function(e) NA_real_)
    }
  ) |>
  ungroup()

kable(
  gf_fits |>
    filter(!is.na(ols_slope)) |>
    mutate(label = paste0(growth_form, " ~ ", metric)) |>
    select(label, n, ols_slope, ols_ci_lo, ols_ci_hi, ols_r2,
           slope, ci_lo, ci_hi, ref, p_vs_ref) |>
    rename(Group = label, N = n,
           `OLS slope` = ols_slope, `OLS CI lo` = ols_ci_lo, `OLS CI hi` = ols_ci_hi,
           `OLS R²` = ols_r2,
           `SMA slope` = slope, `SMA CI lo` = ci_lo, `SMA CI hi` = ci_hi,
           `Ref.` = ref, `p (SMA ≠ ref)` = p_vs_ref) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = "OLS (primary) and SMA (secondary) scaling of E_tree by growth form and size metric"
)
```

```{r forest-gf, fig.height=6.5, fig.cap="Forest plot: OLS (filled circle) and SMA (open circle) slopes by growth form and size metric. Blue = OLS, black = SMA. Reference lines at expected slopes."}
forest_df <- gf_fits |>
  filter(!is.na(ols_slope)) |>
  mutate(label = paste0(growth_form, " ~ ", metric))

# Long format for paired OLS+SMA display
forest_long <- bind_rows(
  forest_df |> mutate(method = "OLS",
                      est = ols_slope, lo = ols_ci_lo, hi = ols_ci_hi),
  forest_df |> mutate(method = "SMA",
                      est = slope,     lo = ci_lo,     hi = ci_hi)
) |>
  filter(!is.na(est))

ggplot(forest_long,
       aes(y = reorder(label, est), x = est, xmin = lo, xmax = hi,
           colour = method, shape = method)) +
  geom_pointrange(size = 0.7, position = position_dodge(width = 0.55)) +
  geom_vline(xintercept = c(1.0, 2.0), linetype = "dashed", colour = "grey50") +
  annotate("text", x = 2.0, y = Inf, label = "  D² ref", vjust = 2,
           colour = "grey30", size = 3.5, hjust = 0) +
  annotate("text", x = 1.0, y = Inf, label = "  BA=SA ref", vjust = 2,
           colour = "grey30", size = 3.5, hjust = 0) +
  scale_colour_manual(values = c(OLS = "#2166ac", SMA = "#333333"),
                      name = "Method") +
  scale_shape_manual(values = c(OLS = 16, SMA = 21), name = "Method") +
  labs(x = "Slope  (log E ~ log size metric)",
       y = NULL,
       title = "Forest plot: whole-tree water-use scaling by growth form",
       subtitle = "OLS (blue filled) and SMA (black open) — 95% CI | SAPFLUXNET") +
  theme(legend.position = "bottom")
```

---

# Sap flux density vs. diameter (Meinzer test)

Meinzer et al. (2005) showed that **sap flux density (SFD = E / SA)** is
approximately size-independent — the invariance of SFD underpins the pipe-model
assumption. If $E \propto SA^1$, then $SFD = E/SA$ should have a slope near zero
with respect to stem diameter.

```{r sfd-diameter, fig.cap="Sap flux density (E/SA) vs. stem diameter: OLS (primary, blue) and SMA (secondary, black dashed). Slope near zero = size-independent SFD."}
dat_sfd <- dat |>
  mutate(sfd       = e_tree / pl_sapw_area,
         log_sfd   = log10(sfd)) |>
  filter(is.finite(log_sfd))

# OLS — primary
fit_ols_sfd <- lm(log_sfd ~ log_dbh, data = dat_sfd)
ols_cf_sfd  <- coef(fit_ols_sfd)
cat(sprintf("SFD ~ DBH  OLS slope = %.3f  (R² = %.3f)\n",
            ols_cf_sfd["log_dbh"], summary(fit_ols_sfd)$r.squared))

# SMA — secondary
fit_sfd <- sma(log_sfd ~ log_dbh, data = dat_sfd, method = "SMA")
cf_sfd  <- coef(fit_sfd)
cat(sprintf("SFD ~ DBH  SMA slope = %.3f\n", cf_sfd[2]))

ggplot(dat_sfd, aes(x = log_dbh, y = log_sfd, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  # OLS — primary
  geom_abline(intercept = ols_cf_sfd["(Intercept)"], slope = ols_cf_sfd["log_dbh"],
              colour = "#2166ac", linewidth = 1.4) +
  # SMA — secondary
  geom_abline(intercept = cf_sfd[1], slope = cf_sfd[2],
              colour = "black", linewidth = 1.0, linetype = "dashed") +
  geom_hline(yintercept = mean(dat_sfd$log_sfd), linetype = "dotted",
             colour = "grey40") +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Stem diameter (cm)"),
       y = expression(log[10]~"Sap flux density  (cm"^3~cm^{-2}~day^{-1}*")"),
       colour = "Growth form",
       title = "Sap flux density vs. stem diameter — OLS (primary) + SMA (secondary)",
       subtitle = sprintf("OLS = %.3f (blue solid)  |  SMA = %.3f (black dashed)  |  dotted = mean SFD",
                          ols_cf_sfd["log_dbh"], cf_sfd[2])) +
  theme(legend.position = "bottom")
```
```

---

# Sapwood area ~ stem diameter (allometry within the pipe model)

The scaling of sapwood area with stem diameter is a key sub-allometry linking
the three size metrics. $SA \propto D^{\gamma}$ with $\gamma$ typically
1.7–2.0 across species and sites.

```{r sa-dbh, fig.cap="Sapwood area ~ stem diameter: OLS (primary, blue solid) and SMA (secondary, black dashed). Reference lines at γ = 2.0 (orange dotted) and γ = 1.7 (blue dotted)."}
# OLS — Model I: primary regression
fit_ols_sadbh <- lm(log_sa ~ log_dbh, data = dat)
ols_cf_sadbh  <- coef(fit_ols_sadbh)
ols_ci_sadbh  <- confint(fit_ols_sadbh)["log_dbh", ]
ols_r2_sadbh  <- summary(fit_ols_sadbh)$r.squared

# SMA — Model II: also reported
fit_sadbh <- sma(log_sa ~ log_dbh, data = dat, method = "SMA")
cf_sadbh  <- coef(fit_sadbh)
ci_sadbh  <- fit_sadbh$coef[[1]][2, 2:3]

cat(sprintf("SA ~ DBH  OLS slope = %.3f  (95%% CI: %.3f – %.3f;  R² = %.3f)\n",
            ols_cf_sadbh["log_dbh"], ols_ci_sadbh[1], ols_ci_sadbh[2], ols_r2_sadbh))
cat(sprintf("SA ~ DBH  SMA slope = %.3f  (95%% CI: %.3f – %.3f)\n",
            cf_sadbh[2], ci_sadbh[1], ci_sadbh[2]))

ggplot(dat, aes(x = log_dbh, y = log_sa, colour = growth_form)) +
  geom_point(alpha = 0.18, size = 1.1) +
  # OLS — primary
  geom_abline(intercept = ols_cf_sadbh["(Intercept)"], slope = ols_cf_sadbh["log_dbh"],
              colour = "#2166ac", linewidth = 1.4) +
  # SMA — secondary
  geom_abline(intercept = cf_sadbh[1], slope = cf_sadbh[2],
              colour = "black", linewidth = 1.0, linetype = "dashed") +
  # Theoretical references
  geom_abline(intercept = mean(dat$log_sa) - 2.0 * mean(dat$log_dbh),
              slope = 2.0, linetype = "dotted", colour = "#d6604d",
              linewidth = 0.9) +
  geom_abline(intercept = mean(dat$log_sa) - 1.7 * mean(dat$log_dbh),
              slope = 1.7, linetype = "dotted", colour = "#4575b4",
              linewidth = 0.9) +
  scale_colour_manual(values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")) +
  labs(x = expression(log[10]~"Stem diameter (cm)"),
       y = expression(log[10]~"Sapwood area (cm²)"),
       colour = "Growth form",
       title = "Sapwood area ~ stem diameter — OLS (primary) + SMA (secondary)",
       subtitle = sprintf(
         "OLS γ = %.3f (blue solid)  |  SMA γ = %.3f (black dashed)  |  ref: 2.0 & 1.7 (dotted)",
         ols_cf_sadbh["log_dbh"], cf_sadbh[2])) +
  theme(legend.position = "bottom")
```

---

# Summary: all scaling exponents

```{r summary-table}
# Helper to pull OLS CI safely
ols_ci_val <- function(fit, term, bound) tryCatch(confint(fit)[term, bound], error = function(e) NA_real_)

summary_df <- tibble(
  Relationship     = c("E ~ D",    "E ~ BA",  "E ~ SA",   "SA ~ D"),
  `x metric`       = c("DBH (cm)", "BA (cm²)","SA (cm²)", "DBH (cm)"),
  `Exp. slope`     = c(2.0,  1.0,  1.0,  1.7),
  # OLS — primary
  `OLS slope`      = c(round(coef(fit_ols_dbh)["log_dbh"],  3),
                        round(coef(fit_ols_ba)["log_ba"],    3),
                        round(coef(fit_ols_sa)["log_sa"],    3),
                        round(coef(fit_ols_sadbh)["log_dbh"],3)),
  `OLS CI lo`      = c(round(ols_ci_val(fit_ols_dbh,  "log_dbh", 1), 3),
                        round(ols_ci_val(fit_ols_ba,   "log_ba",  1), 3),
                        round(ols_ci_val(fit_ols_sa,   "log_sa",  1), 3),
                        round(ols_ci_val(fit_ols_sadbh,"log_dbh", 1), 3)),
  `OLS CI hi`      = c(round(ols_ci_val(fit_ols_dbh,  "log_dbh", 2), 3),
                        round(ols_ci_val(fit_ols_ba,   "log_ba",  2), 3),
                        round(ols_ci_val(fit_ols_sa,   "log_sa",  2), 3),
                        round(ols_ci_val(fit_ols_sadbh,"log_dbh", 2), 3)),
  `OLS R²`         = c(round(summary(fit_ols_dbh)$r.squared,  3),
                        round(summary(fit_ols_ba)$r.squared,   3),
                        round(summary(fit_ols_sa)$r.squared,   3),
                        round(summary(fit_ols_sadbh)$r.squared,3)),
  # SMA — secondary
  `SMA slope`      = c(round(coef(fit_dbh)[2],  3),
                        round(coef(fit_ba)[2],   3),
                        round(coef(fit_sa)[2],   3),
                        round(coef(fit_sadbh)[2],3)),
  `SMA CI lo`      = c(round(fit_dbh$coef[[1]][2,2],   3),
                        round(fit_ba$coef[[1]][2,2],    3),
                        round(fit_sa$coef[[1]][2,2],    3),
                        round(fit_sadbh$coef[[1]][2,2], 3)),
  `SMA CI hi`      = c(round(fit_dbh$coef[[1]][2,3],   3),
                        round(fit_ba$coef[[1]][2,3],    3),
                        round(fit_sa$coef[[1]][2,3],    3),
                        round(fit_sadbh$coef[[1]][2,3], 3))
)

kable(summary_df,
      caption = if (USE_REAL_DATA)
        "Summary: OLS (primary) and SMA (secondary) scaling exponents. OLS minimises y-residuals (Model I); SMA minimises residuals symmetrically (Model II / RMA). See Regression method comparison section for PGLS slopes."
      else
        "⚠ DEMONSTRATION DATA — OLS and SMA scaling exponents (simulated to match Forrester et al. 2022)")
```

---

# Ecological interpretation

```{r interpretation, echo=FALSE, results='asis'}
ols_slope_e_dbh <- round(coef(fit_ols_dbh)["log_dbh"], 2)
ols_slope_e_ba  <- round(coef(fit_ols_ba)["log_ba"],   2)
ols_slope_e_sa  <- round(coef(fit_ols_sa)["log_sa"],   2)
sma_slope_e_dbh <- round(coef(fit_dbh)[2], 2)
sma_slope_e_ba  <- round(coef(fit_ba)[2],  2)
sma_slope_e_sa  <- round(coef(fit_sa)[2],  2)
cat(sprintf("
**Observed scaling exponents (OLS primary; SMA in parentheses):**

* $E \\propto D^{%.2f}$ (SMA: %.2f) — close to $D^2$ (basal-area scaling)
* $E \\propto BA^{%.2f}$ (SMA: %.2f) — close to linear (size-symmetric)
* $E \\propto SA^{%.2f}$ (SMA: %.2f) — close to linear (pipe-model prediction)

OLS (Model I) slopes are attenuated toward zero relative to SMA (Model II) because OLS does not account for measurement error in the predictor. SMA slopes are the symmetric major-axis estimates. See the *Regression method comparison* section for PGLS slopes and Pagel\\'s \\u03bb.

**Mechanistic interpretation:**

The convergence of all three metrics toward their theoretical expectations
($E \\propto D^2$, $E \\propto BA^1$, $E \\propto SA^1$) reflects the same
underlying biology: whole-tree water use scales with the total conducting
cross-section. The tree is a **bundle of xylem conduits** (Zimmermann 1983 pipe
model). As trees grow larger, the number of active conduits increases roughly in
proportion to sapwood area, and sapwood area scales close to $D^2$.

**Competition symmetry:**

A slope of $E \\propto BA^1$ means that a tree twice as large uses exactly
twice as much water — **size-symmetric competition**. There is no disproportionate
capture of water by large trees (which would imply $E \\propto BA^{>1}$). This
contrasts with light competition in many forests, which is strongly
size-asymmetric.

**Key caveats (Forrester et al. 2022):**

* The $E \\propto D^2$ relationship is an average over enormous among-site,
  among-species, and seasonal variation. Species-level or site-level slopes
  can deviate substantially.
* OLS slopes are lower bounds on the true allometric exponent; SMA slopes assume
  equal proportional error in both axes, which may not hold here.
* Sap flux density is not perfectly constant with tree size. Significant among-
  tree and among-species variation exists (CV typically 30–60%%).
* The SAPFLUXNET dataset is not a random sample of the world\\'s forests; it
  over-represents temperate Europe and North America.
* Meinzer et al. (2005) found $E \\propto D^{1.8}$ — slightly below 2.0 — in
  a smaller multi-site dataset, illustrating real uncertainty in the exponent.
", ols_slope_e_dbh, sma_slope_e_dbh,
   ols_slope_e_ba,  sma_slope_e_ba,
   ols_slope_e_sa,  sma_slope_e_sa))
```

---

# Scaling analysis: theoretical framework

## WBE network model vs. pipe model

Two independent mechanistic frameworks converge on the same scaling prediction
for whole-tree water use, providing testable expectations for each regression
exponent.

**Pipe model** (Shinozaki et al. 1964; Mäkelä 1986):
Each unit of leaf area is supplied by a fixed cross-sectional area of sapwood
(one "pipe" per leaf unit). Total transpiration demand scales linearly with
conducting sapwood area $S_A$:

$$E = k_{SA} \cdot S_A$$

where $k_{SA}$ (L day$^{-1}$ cm$^{-2}$) is **sap flux density** — the
normalization constant of the relationship. Under the null (pipe-model)
hypothesis, $k_{SA}$ is **size-independent**. Because area-preserving
branching requires $S_A \propto D^{\gamma}$ with $\gamma \approx 2$:

$$E \propto S_A \propto D^{\gamma} \approx D^{2}$$

**WBE network model** (West, Brown & Enquist 1997):
Fractal branching networks with area-preserving geometry also predict
$S_A \propto D^2$. Whole-tree hydraulic conductance from Poiseuille flow
scales as $K_{tree} \propto D^2$. For a size-independent driving gradient
$\Delta\Psi$ (water-potential difference trunk-to-leaf):

$$E = K_{tree} \cdot \Delta\Psi \propto D^2$$

The WBE result is numerically identical to the pipe-model result because both
encode the same area-preserving assumption. Deviation from slope $= 2.0$ in
$E \sim D$ (or slope $= 1.0$ in $E \sim S_A$) implies one or more of:
(i) $k_{SA}$ varies systematically with tree size; (ii) conduit taper changes
with height; (iii) $\Delta\Psi$ is not size-independent (hydraulic
limitation in tall trees).

**Dimensional check:**

$$[E] = \text{L day}^{-1} = [k_{SA}] \times [S_A] = \underbrace{\text{L day}^{-1}\,\text{cm}^{-2}}_{k_{SA}} \times \underbrace{\text{cm}^{2}}_{S_A}$$

The antilog of the SMA intercept in $\log_{10} E = b_0 + b_1 \log_{10} S_A$
gives the **effective hydraulic conductance per unit sapwood area**,
$k_{SA} = 10^{b_0}$, a physiologically interpretable quantity comparable across
biomes and functional types.

**Summary of testable predictions:**

| Relationship | Pipe model / WBE | Size-linear null |
|---|:---:|:---:|
| $E \sim D$    | **2.0** | 1.0 |
| $E \sim BA$   | **1.0** | 0.5 |
| $E \sim S_A$  | **1.0** | —   |
| $S_A \sim D$  | **2.0** | 1.0 |

```{r exponent-synthesis, fig.height=5.5, fig.cap="Exponent synthesis: OLS (blue filled, primary) and SMA (black open, secondary) slopes vs. theoretical predictions (red diamond). OLS slopes are attenuated relative to SMA; the true exponent lies between OLS and SMA when both axes carry measurement error."}
synth_df <- bind_rows(
  tibble(
    relationship = factor(
      c("E ~ D", "E ~ BA", "E ~ SA", "SA ~ D"),
      levels = c("SA ~ D", "E ~ SA", "E ~ BA", "E ~ D")
    ),
    slope  = c(coef(fit_ols_dbh)["log_dbh"],
               coef(fit_ols_ba)["log_ba"],
               coef(fit_ols_sa)["log_sa"],
               coef(fit_ols_sadbh)["log_dbh"]),
    ci_lo  = c(ols_ci_val(fit_ols_dbh,  "log_dbh", 1),
               ols_ci_val(fit_ols_ba,   "log_ba",  1),
               ols_ci_val(fit_ols_sa,   "log_sa",  1),
               ols_ci_val(fit_ols_sadbh,"log_dbh", 1)),
    ci_hi  = c(ols_ci_val(fit_ols_dbh,  "log_dbh", 2),
               ols_ci_val(fit_ols_ba,   "log_ba",  2),
               ols_ci_val(fit_ols_sa,   "log_sa",  2),
               ols_ci_val(fit_ols_sadbh,"log_dbh", 2)),
    theory = c(2.0, 1.0, 1.0, 2.0),
    method = "OLS"
  ),
  tibble(
    relationship = factor(
      c("E ~ D", "E ~ BA", "E ~ SA", "SA ~ D"),
      levels = c("SA ~ D", "E ~ SA", "E ~ BA", "E ~ D")
    ),
    slope  = c(coef(fit_dbh)[2],   coef(fit_ba)[2],
               coef(fit_sa)[2],    coef(fit_sadbh)[2]),
    ci_lo  = c(fit_dbh$coef[[1]][2,2],   fit_ba$coef[[1]][2,2],
               fit_sa$coef[[1]][2,2],    fit_sadbh$coef[[1]][2,2]),
    ci_hi  = c(fit_dbh$coef[[1]][2,3],   fit_ba$coef[[1]][2,3],
               fit_sa$coef[[1]][2,3],    fit_sadbh$coef[[1]][2,3]),
    theory = c(2.0, 1.0, 1.0, 2.0),
    method = "SMA"
  )
)

ggplot(synth_df,
       aes(y = relationship, x = slope, xmin = ci_lo, xmax = ci_hi,
           colour = method, shape = method)) +
  geom_pointrange(size = 0.9,
                  position = position_dodge(width = 0.55)) +
  geom_point(aes(x = theory, y = relationship), shape = 18, size = 6,
             colour = "#d6604d", alpha = 0.9,
             inherit.aes = FALSE,
             data = distinct(synth_df, relationship, theory)) +
  geom_vline(xintercept = c(1.0, 2.0), linetype = "dashed",
             colour = "grey65", linewidth = 0.6) +
  scale_colour_manual(values = c(OLS = "#2166ac", SMA = "#333333"),
                      name = "Method") +
  scale_shape_manual(values = c(OLS = 16, SMA = 21), name = "Method") +
  labs(
    x = "Slope (log\u2013log)",
    y = NULL,
    title = "Exponent synthesis: OLS (primary) and SMA vs. theoretical predictions",
    subtitle = "Filled = OLS (blue)  |  Open = SMA (black)  |  Red diamond = pipe-model / WBE"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

---

# Multi-level structure: site effects

Trees are nested within 194 sites spanning 32 countries.  Standard SMA treats
every tree as an independent observation, which inflates degrees of freedom and
may bias inference if sites differ in their hydraulic baseline.  This section
decomposes variance into within-site (tree-to-tree) and among-site components
and fits a linear mixed model (LMM) as a complement to the SMA.

```{r advanced-data-prep}
# ── Join site metadata (si_mat, si_map, si_biome, si_country) to dat ─────────
# Required by all subsequent advanced sections.
# Real data:  left_join on si_code from site_md.
# Demo data:  create synthetic site structure (40 sites) to enable all analyses.

if (!is.null(site_md) &&
    "si_code" %in% names(dat) &&
    !"si_mat" %in% names(dat)) {

  dat <- dat |>
    left_join(
      site_md |>
        select(si_code, si_biome, si_country, si_mat, si_map) |>
        distinct(si_code, .keep_all = TRUE),
      by = "si_code"
    )
  cat("Site metadata joined: si_mat, si_map, si_biome, si_country added.\n",
      "Trees with si_mat:", sum(!is.na(dat$si_mat)), "\n")

} else if (!USE_REAL_DATA && !"si_code" %in% names(dat)) {

  # Synthetic site structure for demonstration data
  set.seed(42)
  n_sites <- 40
  site_tbl <- tibble(
    si_code  = paste0("SITE_", sprintf("%02d", seq_len(n_sites))),
    si_biome = sample(
      c("Temperate forest", "Boreal forest",
        "Tropical forest savanna", "Woodland/Shrubland",
        "Subtropical desert"),
      n_sites, replace = TRUE,
      prob = c(0.45, 0.20, 0.20, 0.10, 0.05)
    ),
    si_mat = rnorm(n_sites, mean = 10, sd = 7),
    si_map = exp(rnorm(n_sites, mean = log(900), sd = 0.45))
  )
  dat <- dat |>
    mutate(
      si_code = sample(site_tbl$si_code, n(), replace = TRUE)
    ) |>
    left_join(site_tbl, by = "si_code")
  cat("\u26a0 Demo data: synthetic site structure assigned (",
      n_sites, "sites)\n")

} else {
  cat("Site columns already present in dat.\n")
}

cat("Columns now in dat: ",
    paste(intersect(c("si_code","si_biome","si_mat","si_map"),
                    names(dat)), collapse = ", "), "\n")
```

```{r site-variance, fig.height=4}
# ── Variance decomposition and ICC ───────────────────────────────────────────
# LMM: log(E) ~ log(DBH) + (1 | site)
# Random intercept absorbs site-level hydraulic baseline differences.

if ("si_code" %in% names(dat)) {

  lmm_dbh <- lmer(log_e ~ log_dbh + (1 | si_code),
                  data = dat, REML = TRUE)

  # Variance components
  vc <- as.data.frame(VarCorr(lmm_dbh)) |>
    select(grp, var1, vcov) |>
    rename(Source = grp, Component = var1, Variance = vcov) |>
    mutate(
      `Std. Dev.`  = round(sqrt(Variance), 4),
      Variance     = round(Variance, 4),
      `% of total` = round(100 * Variance / sum(Variance), 1)
    )
  kable(vc,
        caption = "Variance components: log(E) ~ log(DBH) + (1 | site)")

  # ICC: proportion of variance attributable to among-site differences
  icc_res <- icc(lmm_dbh)
  cat("\nICC (proportion of total variance from among-site differences):\n")
  print(icc_res)

  # Fixed-effect slope from LMM vs SMA
  fe <- fixef(lmm_dbh)
  ci_fe <- confint(lmm_dbh, parm = "log_dbh",
                   method = "Wald", quiet = TRUE)
  cat(sprintf(
    "\nLMM fixed slope: %.3f  (95%% Wald CI: %.3f \u2013 %.3f)\n",
    fe["log_dbh"], ci_fe[1], ci_fe[2]
  ))
  cat(sprintf(
    "SMA slope (pooled): %.3f  (95%% CI: %.3f \u2013 %.3f)\n",
    coef(fit_dbh)[2],
    fit_dbh$coef[[1]][2,2],
    fit_dbh$coef[[1]][2,3]
  ))

} else {
  cat("si_code not in dat \u2014 skipping LMM\n")
}
```

```{r site-residuals, fig.height=7, fig.cap="Site-level mean deviations from the overall SMA E~DBH fitted line (caterpillar plot). Large positive residuals = sites with higher water use than the global scaling predicts at a given DBH; points coloured by biome."}
# Residuals are computable for all trees regardless of site structure
dat$resid_sma_dbh <- dat$log_e -
  (coef(fit_dbh)[1] + coef(fit_dbh)[2] * dat$log_dbh)

if ("si_code" %in% names(dat)) {
  site_resid_df <- dat |>
    group_by(si_code) |>
    summarise(
      mean_resid = mean(resid_sma_dbh, na.rm = TRUE),
      se_resid   = sd(resid_sma_dbh, na.rm = TRUE) / sqrt(n()),
      n_trees    = n(),
      si_biome   = first(si_biome),
      .groups = "drop"
    ) |>
    filter(!is.na(mean_resid))

  ggplot(site_resid_df,
         aes(x = mean_resid,
             y = reorder(si_code, mean_resid),
             colour = si_biome)) +
    geom_pointrange(
      aes(xmin = mean_resid - 1.96 * se_resid,
          xmax = mean_resid + 1.96 * se_resid),
      size = 0.35, linewidth = 0.6
    ) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = "grey40", linewidth = 0.7) +
    scale_colour_brewer(palette = "Set2", na.value = "grey60") +
    labs(
      x = expression("Mean site residual  (" * log[10] * " units)"),
      y = "Site (ranked by residual)",
      colour = "Biome",
      title = "Site-level departures from global E~DBH scaling",
      subtitle = sprintf("%d sites  |  points \u00b11.96 SE",
                         nrow(site_resid_df))
    ) +
    theme(
      axis.text.y  = element_text(size = 5.5),
      legend.position = "bottom"
    )
} else {
  cat("si_code not in dat \u2014 skipping site residuals plot\n")
}
```

```{r resid-vs-climate, fig.height=4.5, fig.cap="Tree-level residuals from the SMA E~DBH fit vs. mean annual temperature (left) and log10 MAP (right). Systematic trends indicate climate-driven variation in the hydraulic baseline beyond what tree size explains."}
if (all(c("si_mat", "si_map", "resid_sma_dbh") %in% names(dat))) {

  dat_env <- dat |> filter(!is.na(si_mat), !is.na(si_map))

  p_mat <- ggplot(dat_env,
                  aes(x = si_mat, y = resid_sma_dbh,
                      colour = growth_form)) +
    geom_point(alpha = 0.12, size = 0.8) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 1.0) +
    scale_colour_manual(
      values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")
    ) +
    labs(
      x = "Mean annual temperature (\u00b0C)",
      y = expression("Residual  (" * log[10] * " E ~ DBH)"),
      colour = "Growth form",
      title = "Residuals vs. MAT"
    ) +
    theme(legend.position = "bottom")

  p_map <- ggplot(dat_env,
                  aes(x = log10(si_map), y = resid_sma_dbh,
                      colour = growth_form)) +
    geom_point(alpha = 0.12, size = 0.8) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 1.0) +
    scale_colour_manual(
      values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")
    ) +
    labs(
      x = expression(log[10] ~ "MAP (mm)"),
      y = expression("Residual  (" * log[10] * " E ~ DBH)"),
      colour = "Growth form",
      title = "Residuals vs. MAP"
    ) +
    theme(legend.position = "bottom")

  gridExtra::grid.arrange(p_mat, p_map, ncol = 2)

} else {
  cat("Climate variables or residuals not available.\n")
}
```

---

# Biome-stratified scaling

Slopes that vary across biomes reveal ecological contexts where the pipe model
or WBE predictions break down.  Colder or drier biomes may show lower
normalization constants ($k_{SA}$) due to lower leaf-area-to-sapwood ratios or
colder xylem conductance.

```{r biome-sma, fig.height=5, fig.cap="Biome-stratified SMA slopes for E~DBH and E~SA (biomes with ≥20 trees). Dashed lines mark the theoretical predictions. Biomes whose CI excludes the theoretical value deviate significantly from the pipe-model / WBE prediction."}
biome_sma <- NULL   # initialise so hydro-norm can check existence

if ("si_biome" %in% names(dat)) {

  biome_sma <- dat |>
    filter(!is.na(si_biome)) |>
    tidyr::nest(data = -si_biome) |>
    mutate(
      n    = map_int(data, nrow),
      n_sp = map_int(data, ~ n_distinct(.x$pl_species, na.rm = TRUE)),
      fit_dbh = map(data, ~ {
        if (nrow(.x) < 20 || n_distinct(.x$pl_species, na.rm=TRUE) < 3) return(NULL)
        tryCatch(sma(log_e ~ log_dbh, data = .x, method = "SMA"),
                 error = function(e) NULL)
      }),
      fit_sa = map(data, ~ {
        if (nrow(.x) < 20 || n_distinct(.x$pl_species, na.rm=TRUE) < 3) return(NULL)
        tryCatch(sma(log_e ~ log_sa,  data = .x, method = "SMA"),
                 error = function(e) NULL)
      }),
      slope_dbh    = map_dbl(fit_dbh,
                             ~ if (is.null(.x)) NA_real_ else coef(.x)[2]),
      ci_lo_dbh    = map_dbl(fit_dbh,
                             ~ if (is.null(.x)) NA_real_ else .x$coef[[1]][2,2]),
      ci_hi_dbh    = map_dbl(fit_dbh,
                             ~ if (is.null(.x)) NA_real_ else .x$coef[[1]][2,3]),
      slope_sa     = map_dbl(fit_sa,
                             ~ if (is.null(.x)) NA_real_ else coef(.x)[2]),
      ci_lo_sa     = map_dbl(fit_sa,
                             ~ if (is.null(.x)) NA_real_ else .x$coef[[1]][2,2]),
      ci_hi_sa     = map_dbl(fit_sa,
                             ~ if (is.null(.x)) NA_real_ else .x$coef[[1]][2,3]),
      intercept_sa = map_dbl(fit_sa,
                             ~ if (is.null(.x)) NA_real_ else coef(.x)[1])
    )

  forest_biome <- bind_rows(
    biome_sma |>
      filter(!is.na(slope_dbh)) |>
      transmute(si_biome, n,
                slope = slope_dbh, ci_lo = ci_lo_dbh, ci_hi = ci_hi_dbh,
                metric = "E ~ DBH", theory = 2.0),
    biome_sma |>
      filter(!is.na(slope_sa)) |>
      transmute(si_biome, n,
                slope = slope_sa,  ci_lo = ci_lo_sa,  ci_hi = ci_hi_sa,
                metric = "E ~ SA",  theory = 1.0)
  )

  ggplot(forest_biome,
         aes(y = reorder(si_biome, slope),
             x = slope, xmin = ci_lo, xmax = ci_hi,
             colour = metric)) +
    geom_pointrange(position = position_dodge(width = 0.55), size = 0.7) +
    geom_vline(data = distinct(forest_biome, metric, theory),
               aes(xintercept = theory, colour = metric),
               linetype = "dashed", linewidth = 0.7, alpha = 0.75) +
    scale_colour_manual(
      values = c("E ~ DBH" = "#2166ac", "E ~ SA" = "#d6604d")
    ) +
    facet_wrap(~ metric, scales = "free_x") +
    labs(
      x = "SMA slope (\u00b1 95% CI)",
      y = NULL,
      colour = "Relationship",
      title = "Biome-stratified SMA scaling slopes",
      subtitle = "Dashed lines = theoretical predictions"
    ) +
    theme(legend.position = "bottom")

} else {
  cat("si_biome not in dat \u2014 skipping biome-stratified SMA\n")
}
```

```{r biome-size-dist, fig.height=6, fig.cap="Size distribution (log10 DBH) by biome and growth form. Biomes with truncated large-tree distributions may show artefactually lower slopes due to restricted size range."}
if ("si_biome" %in% names(dat)) {
  dat |>
    filter(!is.na(si_biome)) |>
    ggplot(aes(x = log_dbh, fill = growth_form)) +
    geom_histogram(bins = 25, colour = "white", alpha = 0.85,
                   position = "stack") +
    facet_wrap(~ si_biome, scales = "free_y", ncol = 2) +
    scale_fill_manual(
      values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")
    ) +
    labs(
      x = expression(log[10] ~ "Stem diameter (cm)"),
      y = "Count",
      fill = "Growth form",
      title = "Size distribution by biome and growth form",
      subtitle = "Restricted size range biases SMA slopes toward the OLS slope"
    ) +
    theme(strip.text = element_text(size = 8),
          legend.position = "bottom")
} else {
  cat("si_biome not in dat \u2014 skipping size distribution\n")
}
```

---

# Climate modulation of scaling intercepts

The SMA intercept for $E \sim S_A$ is the log-scale metabolic normalisation
constant.  If the intercept varies across sites with mean annual temperature
(MAT) or precipitation (MAP), it means $k_{SA}$ — the effective sap flux
density — shifts with climate independently of tree size.  This is tested by
averaging SMA residuals at the site level and regressing them against site
climate variables.

```{r climate-intercepts, fig.height=4.5, fig.cap="Site-level mean deviation from the E~SA SMA fit vs. MAT (left) and log10 MAP (right). Points sized by number of trees per site, coloured by biome. The weighted linear regression line quantifies directional climate effects on the hydraulic normalisation constant."}
# Residuals from the global E~SA SMA fit
dat$resid_sa <- dat$log_e -
  (coef(fit_sa)[1] + coef(fit_sa)[2] * dat$log_sa)

if (all(c("si_code", "si_mat", "si_map") %in% names(dat))) {

  site_intercepts <- dat |>
    filter(!is.na(si_code), !is.na(resid_sa)) |>
    group_by(si_code) |>
    summarise(
      intercept_i = mean(resid_sa, na.rm = TRUE),
      n_trees     = n(),
      si_mat      = first(si_mat),
      si_map      = first(si_map),
      si_biome    = first(si_biome),
      .groups = "drop"
    ) |>
    filter(!is.na(si_mat), !is.na(si_map), n_trees >= 3)

  cat("Sites entering climate model:", nrow(site_intercepts), "\n")

  # Weighted LM: site intercept deviation ~ MAT + log10(MAP)
  # Weights = number of trees per site (more trees = more reliable mean residual)
  lm_clim <- lm(intercept_i ~ si_mat + log10(si_map),
                data    = site_intercepts,
                weights = n_trees)
  cat("\n=== LM: site intercept ~ MAT + log10(MAP) [weighted by n trees] ===\n")
  print(summary(lm_clim))

  p_mat2 <- ggplot(site_intercepts,
                   aes(x = si_mat, y = intercept_i,
                       colour = si_biome, size = n_trees)) +
    geom_point(alpha = 0.8) +
    geom_smooth(aes(x = si_mat, y = intercept_i),
                method = "lm", se = TRUE, colour = "grey30",
                linewidth = 1.0, inherit.aes = FALSE) +
    scale_colour_brewer(palette = "Set2", na.value = "grey60") +
    scale_size_continuous(range = c(2, 8)) +
    labs(
      x = "Mean annual temperature (\u00b0C)",
      y = "Site mean residual (E~SA)",
      colour = "Biome",
      size = "N trees"
    ) +
    theme(legend.position = "none")

  p_map2 <- ggplot(site_intercepts,
                   aes(x = log10(si_map), y = intercept_i,
                       colour = si_biome, size = n_trees)) +
    geom_point(alpha = 0.8) +
    geom_smooth(aes(x = log10(si_map), y = intercept_i),
                method = "lm", se = TRUE, colour = "grey30",
                linewidth = 1.0, inherit.aes = FALSE) +
    scale_colour_brewer(palette = "Set2", na.value = "grey60") +
    scale_size_continuous(range = c(2, 8)) +
    labs(
      x = expression(log[10] ~ "MAP (mm)"),
      y = "Site mean residual (E~SA)",
      colour = "Biome",
      size = "N trees"
    ) +
    theme(legend.position = "right")

  gridExtra::grid.arrange(p_mat2, p_map2, ncol = 2,
    top = "Climate modulation of the E~SA hydraulic normalisation constant")

} else {
  cat("Climate variables not available \u2014 skipping climate intercept analysis\n")
}
```

---

# Size-range invariance test

A true power law is **scale-invariant**: the slope should not change when the
analysis is restricted to different portions of the size distribution.  A
systematically changing slope across DBH windows is evidence against a single
power-law mechanism operating across the full size range.

The quantile regression of sap flux density vs. diameter provides an additional
test: if $k_{SA}$ is truly size-independent (pipe model), the slope at every
quantile should be near zero.

```{r size-windows, fig.height=4, fig.cap="Scale-invariance test: SMA slope for E~DBH within three non-overlapping DBH windows. If the power law is scale-invariant, slopes should be statistically indistinguishable from each other and consistent with the theoretical value of 2.0 (red dashed line)."}
dat_w <- dat |>
  mutate(
    size_class = factor(
      case_when(
        pl_dbh < 5    ~ "Small (<5 cm)",
        pl_dbh <= 20  ~ "Medium (5\u201320 cm)",
        TRUE          ~ "Large (>20 cm)"
      ),
      levels = c("Small (<5 cm)", "Medium (5\u201320 cm)", "Large (>20 cm)")
    )
  )

window_sma <- dat_w |>
  tidyr::nest(data = -size_class) |>
  mutate(
    n       = map_int(data, nrow),
    dbh_min = map_dbl(data, ~ min(.x$pl_dbh, na.rm = TRUE)),
    dbh_max = map_dbl(data, ~ max(.x$pl_dbh, na.rm = TRUE)),
    fit     = map(data, ~ {
      if (nrow(.x) < 10) return(NULL)
      tryCatch(sma(log_e ~ log_dbh, data = .x, method = "SMA"),
               error = function(e) NULL)
    }),
    slope  = map_dbl(fit, ~ if (is.null(.x)) NA_real_ else coef(.x)[2]),
    ci_lo  = map_dbl(fit, ~ if (is.null(.x)) NA_real_ else .x$coef[[1]][2,2]),
    ci_hi  = map_dbl(fit, ~ if (is.null(.x)) NA_real_ else .x$coef[[1]][2,3]),
    r2     = map_dbl(fit, ~ if (is.null(.x)) NA_real_ else .x$r2[[1]])
  )

kable(
  window_sma |>
    select(size_class, n, dbh_min, dbh_max, slope, ci_lo, ci_hi, r2) |>
    mutate(across(c(dbh_min, dbh_max, slope, ci_lo, ci_hi, r2),
                  ~ round(.x, 3))) |>
    rename(
      `Size class`  = size_class,
      N             = n,
      `DBH min (cm)` = dbh_min,
      `DBH max (cm)` = dbh_max,
      `SMA slope`   = slope,
      `CI low`      = ci_lo,
      `CI high`     = ci_hi,
      `R2`          = r2
    ),
  caption = "SMA slope for E~DBH within each DBH size window (scale-invariance test)"
)

ggplot(window_sma |> filter(!is.na(slope)),
       aes(x = size_class, y = slope, ymin = ci_lo, ymax = ci_hi)) +
  geom_hline(yintercept = 2.0, linetype = "dashed",
             colour = "#d6604d", linewidth = 0.9) +
  geom_pointrange(colour = "#2166ac", size = 1.0) +
  geom_text(aes(label = paste0("n=", n)),
            nudge_x = 0.12, size = 3.5, colour = "grey30") +
  annotate("text", x = 0.55, y = 2.0, label = "Pipe model / WBE = 2.0",
           vjust = -0.5, hjust = 0, colour = "#d6604d", size = 3.5) +
  labs(
    x = "DBH size class",
    y = "SMA slope  (log E ~ log DBH)",
    title = "Scale-invariance test: SMA slope by DBH window",
    subtitle = "Consistent slopes across windows support a single power-law mechanism"
  )
```

```{r sfd-quantile, fig.height=4.5, fig.cap="Quantile regression of log sap flux density (SFD = E/SA) vs. log stem diameter at the 10th, 50th, and 90th percentiles. Slopes near zero at all quantiles support size-independence of SFD (pipe model null). Non-zero slope at the 90th percentile but not the median would indicate that only the largest trees deviate."}
# Pipe-model invariance: SFD should be size-independent
dat_sfd_qr <- dat |>
  mutate(log_sfd = log10(e_tree / pl_sapw_area)) |>
  filter(is.finite(log_sfd), !is.na(log_dbh))

taus <- c(0.10, 0.50, 0.90)
qr_fits <- lapply(taus, function(tau) {
  rq(log_sfd ~ log_dbh, data = dat_sfd_qr, tau = tau)
})
names(qr_fits) <- paste0("tau=", taus)

# Coefficient table (sandwich SE, computationally inexpensive)
qr_coef_rows <- lapply(names(qr_fits), function(nm) {
  s    <- summary(qr_fits[[nm]], se = "nid")
  cmat <- as.data.frame(s$coefficients)
  tibble(
    Quantile  = nm,
    Term      = rownames(cmat),
    Estimate  = cmat[[1]],
    `Std. Error` = cmat[[2]],
    `t value` = cmat[[3]],
    `p value` = cmat[[4]]
  )
})
qr_table <- bind_rows(qr_coef_rows) |>
  filter(Term == "log_dbh") |>
  select(-Term) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

kable(qr_table,
      caption = "Quantile regression: log(SFD) ~ log(DBH) — slope at 10th, 50th, 90th percentile")

# Prediction lines
dbh_seq  <- seq(min(dat_sfd_qr$log_dbh), max(dat_sfd_qr$log_dbh),
                length.out = 200)
qr_lines <- lapply(names(qr_fits), function(nm) {
  cf <- coef(qr_fits[[nm]])
  tibble(
    log_dbh  = dbh_seq,
    log_sfd  = cf[1] + cf[2] * dbh_seq,
    quantile = nm
  )
}) |> bind_rows()

ggplot(dat_sfd_qr, aes(x = log_dbh, y = log_sfd)) +
  geom_point(aes(colour = growth_form), alpha = 0.14, size = 0.9) +
  geom_line(data  = qr_lines,
            aes(x = log_dbh, y = log_sfd,
                linetype = quantile, group = quantile),
            colour = "black", linewidth = 1.0) +
  geom_hline(yintercept = mean(dat_sfd_qr$log_sfd, na.rm = TRUE),
             linetype = "dotted", colour = "grey50", linewidth = 0.7) +
  scale_colour_manual(
    values = c(Angiosperm = "#4dac26", Conifer = "#7b3294")
  ) +
  scale_linetype_manual(
    values = c("tau=0.1"  = "dotdash",
               "tau=0.5"  = "solid",
               "tau=0.9"  = "longdash"),
    labels = c("tau=0.1"  = "10th pctile",
               "tau=0.5"  = "Median",
               "tau=0.9"  = "90th pctile")
  ) +
  labs(
    x        = expression(log[10] ~ "Stem diameter (cm)"),
    y        = expression(log[10] ~ "Sap flux density  (L day" ^ {-1} ~ cm ^ {-2} * ")"),
    colour   = "Growth form",
    linetype = "Quantile",
    title    = "Pipe-model invariance: sap flux density vs. stem diameter",
    subtitle = "Slope near zero = size-independent SFD (pipe model holds)"
  ) +
  theme(legend.position = "bottom")
```

---

# Hydraulic normalization constants

The antilog of the SMA intercept from $\log E = b_0 + b_1 \log S_A$ gives

$$k_{SA} = 10^{b_0} \quad [\text{L day}^{-1}\,\text{cm}^{-2}]$$

This is the **effective sap flux density** — the average water throughput per
unit sapwood cross-section.  It is a physiologically interpretable quantity
that varies with stomatal conductance regime, hydraulic architecture, and
atmospheric demand.  Differences across biomes and functional types represent
climatic and anatomical modulation of the normalization constant that is
otherwise invisible in slope-only comparisons.

```{r hydro-norm, fig.height=4.5, fig.cap="Hydraulic normalization constant k_SA (= 10^intercept from E~SA SMA) by growth form (table) and biome (dot plot). Red dashed line = global mean k_SA. Biomes above the line maintain higher sap flux density per unit sapwood area."}
# ── Global normalisation constant ─────────────────────────────────────────────
hyd_norm_global <- 10^coef(fit_sa)[1]
cat(sprintf(
  "Global k_SA: 10^%.4f = %.4f L day\u207b\u00b9 cm\u207b\u00b2\n",
  coef(fit_sa)[1], hyd_norm_global
))

# ── By growth form (from gf_fits; metric == 'SA') ─────────────────────────────
hyd_norm_gf <- gf_fits |>
  filter(metric == "SA", !is.na(slope)) |>
  rowwise() |>
  mutate(
    intercept = if (!is.null(fit)) coef(fit)[1] else NA_real_,
    hyd_norm  = 10^intercept
  ) |>
  ungroup() |>
  select(growth_form, n, intercept, hyd_norm, slope, ci_lo, ci_hi)

kable(
  hyd_norm_gf |>
    mutate(across(where(is.numeric), ~ round(.x, 4))) |>
    rename(
      `Growth form`     = growth_form,
      N                 = n,
      `log10(k_SA)`     = intercept,
      `k_SA (L/d/cm2)` = hyd_norm,
      `E~SA slope`      = slope,
      `CI low`          = ci_lo,
      `CI high`         = ci_hi
    ),
  caption = "Hydraulic normalisation constant k_SA by growth form"
)

# ── By biome (from biome_sma) ─────────────────────────────────────────────────
if (!is.null(biome_sma) && "intercept_sa" %in% names(biome_sma)) {

  hyd_norm_biome <- biome_sma |>
    filter(!is.na(intercept_sa)) |>
    mutate(hyd_norm = 10^intercept_sa) |>
    select(si_biome, n, intercept_sa, hyd_norm,
           slope_sa, ci_lo_sa, ci_hi_sa) |>
    arrange(desc(hyd_norm))

  kable(
    hyd_norm_biome |>
      mutate(across(where(is.numeric), ~ round(.x, 4))) |>
      rename(
        Biome                = si_biome,
        N                    = n,
        `log10(k_SA)`        = intercept_sa,
        `k_SA (L/d/cm2)`    = hyd_norm,
        `E~SA slope`         = slope_sa,
        `CI low (slope)`     = ci_lo_sa,
        `CI high (slope)`    = ci_hi_sa
      ),
    caption = paste0(
      "Hydraulic normalisation constant k_SA by biome. ",
      "k_SA = 10^(SMA intercept for E~SA) is interpretable as effective ",
      "hydraulic conductance per unit sapwood area (L day\u207b\u00b9 cm\u207b\u00b2) ",
      "ONLY when the E~SA slope does not significantly depart from 1.0. ",
      "For biomes where CI does not include 1.0, treat k_SA as the fitted ",
      "value at SA = 1 cm\u00b2, not a size-independent hydraulic constant."
    )
  )

  ggplot(hyd_norm_biome,
         aes(x = reorder(si_biome, hyd_norm),
             y = hyd_norm, size = n)) +
    geom_point(colour = "#2166ac", alpha = 0.85) +
    geom_hline(yintercept = hyd_norm_global,
               linetype = "dashed", colour = "#d6604d",
               linewidth = 0.8) +
    annotate("text",
             x     = 0.6,
             y     = hyd_norm_global,
             label = sprintf("Global k_SA = %.4f", hyd_norm_global),
             vjust = -0.55, hjust = 0, size = 3.5, colour = "#d6604d") +
    coord_flip() +
    scale_size_continuous(range = c(3, 10)) +
    labs(
      x        = NULL,
      y        = expression(k[SA] ~ "(L day" ^ {-1} ~ cm ^ {-2} * ")"),
      size     = "N trees",
      title    = expression("Hydraulic normalisation constant " * k[SA] * " by biome"),
      subtitle = "Antilog of E~SA SMA intercept  |  Red dashed = global mean"
    ) +
    theme(legend.position = "right")

} else {
  cat("biome_sma not available \u2014 run the biome-stratified SMA section first\n")
}
```

---

# Regression method comparison: OLS (primary), SMA, and PGLS

This section consolidates OLS (Model I), standardised major axis regression (SMA, Model II / RMA), and phylogenetic generalised least squares (PGLS) slopes for all three core scaling relationships ($E \sim D$, $E \sim BA$, $E \sim SA$).

**OLS (Model I)** minimises residuals in the response ($y$) direction only; it is the appropriate estimator for prediction and hypothesis testing about the conditional mean but is attenuated toward zero when the predictor carries measurement error.
**SMA (Model II / RMA)** minimises residuals symmetrically and is often preferred for allometric log–log relationships where both axes carry error; it provides the slope of the major axis of the bivariate distribution and is always ≥ the OLS slope in magnitude.
**PGLS** operates on species means and adjusts for shared evolutionary history via a Pagel $\lambda$ covariance structure estimated by maximum likelihood. $\lambda = 1$ = Brownian-motion covariance; $\lambda = 0$ = star phylogeny (equivalent to OLS on species means).

```{r taxon-reconciliation}
# Required for this section: V.PhyloMaker2, caper, ape
# (loaded in setup chunk above)

if (USE_REAL_DATA && "pl_species" %in% names(dat)) {

  # Count trees per species
  sp_counts <- dat |>
    filter(!is.na(pl_species), nchar(trimws(pl_species)) > 0) |>
    count(pl_species, name = "n_trees")

  # Parse genus as first word of binomial
  sp_list_raw <- sp_counts |>
    mutate(
      species = pl_species,
      genus   = stringr::word(pl_species, 1)
    )

  # Family lookup from V.PhyloMaker2 species table (LCVP backbone)
  # tips.info.LCVP columns: group, species (Genus_species), genus, family
  tips_lcvp     <- V.PhyloMaker2::tips.info.LCVP
  # Convert backbone species to space format for matching against pl_species
  tips_sp_space <- gsub("_", " ", tips_lcvp$species)

  # W1 fix: species-level family lookup first, fall back to genus-level
  # (distinct(genus) can misassign polyphyletic genera — prefer species match)
  sp_family_exact <- tips_lcvp |>
    mutate(species_sp = gsub("_", " ", species)) |>
    select(species_sp, family) |>
    distinct(species_sp, .keep_all = TRUE)

  genus_family <- tips_lcvp |>
    select(genus, family) |>
    distinct(genus, .keep_all = TRUE)

  sp_list <- sp_list_raw |>
    left_join(sp_family_exact, by = c("species" = "species_sp")) |>
    left_join(genus_family,    by = "genus", suffix = c("", "_genus")) |>
    mutate(
      family       = coalesce(family, family_genus, "Unknown"),
      match_status = case_when(
        species %in% tips_sp_space     ~ "matched",
        genus   %in% tips_lcvp$genus   ~ "genus_only",
        TRUE                            ~ "unmatched"
      )
    ) |>
    select(-family_genus)

  n_total      <- nrow(sp_list)
  n_matched    <- sum(sp_list$match_status == "matched")
  n_genus_only <- sum(sp_list$match_status == "genus_only")
  n_unmatched  <- sum(sp_list$match_status == "unmatched")

  cat(sprintf(
    "Taxonomy reconciliation \u2014 %d species total\n  matched: %d (%.1f%%)\n  genus-only: %d (%.1f%%)\n  unmatched: %d (%.1f%%)\n",
    n_total,
    n_matched,    100 * n_matched    / n_total,
    n_genus_only, 100 * n_genus_only / n_total,
    n_unmatched,  100 * n_unmatched  / n_total
  ))

  kable(
    sp_list |>
      select(species, genus, family, match_status, n_trees) |>
      arrange(match_status, species) |>
      rename(
        Species        = species,
        Genus          = genus,
        Family         = family,
        `Match status` = match_status,
        `N trees`      = n_trees
      ),
    caption = "Taxonomy reconciliation: SAPFLUXNET species matched to V.PhyloMaker2 LCVP backbone (GBOTB.extended.LCVP). match_status: matched = exact species in backbone; genus_only = genus placed, species absent; unmatched = genus absent. Family assigned at species level first, with genus-level fallback."
  )

} else {
  sp_list <- NULL
  message(
    "Taxonomy reconciliation skipped: requires USE_REAL_DATA = TRUE ",
    "and pl_species column in dat."
  )
}
```

```{r phylo-build}
if (USE_REAL_DATA && !is.null(sp_list)) {

  n_unknown        <- sum(sp_list$family == "Unknown")
  sp_list_for_tree <- sp_list |>
    filter(family != "Unknown") |>
    select(species, genus, family) |>
    distinct(species, .keep_all = TRUE)

  cat(sprintf(
    "Building genus-level phylogeny (scenario S3) for %d species\n  (%d species with Unknown family excluded from tree build)\n",
    nrow(sp_list_for_tree), n_unknown
  ))

  phy <- tryCatch({
    result <- phylo.maker(
      sp.list        = as.data.frame(sp_list_for_tree),
      tree           = V.PhyloMaker2::GBOTB.extended.LCVP,
      nodes          = V.PhyloMaker2::nodes.info.1.LCVP,
      output.sp.list = FALSE,
      scenarios      = "S3",
      r              = 1
    )
    result$scenario.3
  }, error = function(e) {
    message("phylo.maker() failed: ", conditionMessage(e))
    NULL
  })

  if (!is.null(phy)) {
    cat(sprintf(
      "Phylogeny built: %d tips\nIs ultrametric: %s\n",
      length(phy$tip.label),
      is.ultrametric(phy)
    ))
  } else {
    cat("Tree build failed \u2014 PGLS sections will be skipped.\n")
  }

} else {
  phy <- NULL
  if (!USE_REAL_DATA) {
    message("Phylogeny build skipped: PGLS requires real SAPFLUXNET species data.")
  }
}
```

```{r pgls-setup}
sp_means   <- NULL
comp_data  <- NULL
phy_pruned <- NULL

if (USE_REAL_DATA && !is.null(phy) && "pl_species" %in% names(dat)) {

  sp_means <- dat |>
    group_by(pl_species, growth_form) |>
    summarise(
      log_e   = mean(log_e,   na.rm = TRUE),
      log_dbh = mean(log_dbh, na.rm = TRUE),
      log_ba  = mean(log_ba,  na.rm = TRUE),
      log_sa  = mean(log_sa,  na.rm = TRUE),
      n       = n(),
      .groups = "drop"
    ) |>
    filter(n >= 3) |>
    mutate(tip_label = gsub(" ", "_", pl_species))

  cat(sprintf("Species means: %d species with >= 3 trees each\n", nrow(sp_means)))

  # V.PhyloMaker2 tip labels use Genus_species (underscore) format
  shared_tips <- intersect(sp_means$tip_label, phy$tip.label)
  cat(sprintf(
    "Species in common with phylogeny tips: %d of %d\n",
    length(shared_tips), nrow(sp_means)
  ))

  if (length(shared_tips) >= 10) {

    phy_pruned <- drop.tip(phy, setdiff(phy$tip.label, shared_tips))

    sp_means_df <- sp_means |>
      filter(tip_label %in% shared_tips) |>
      as.data.frame()

    comp_data <- tryCatch(
      caper::comparative.data(
        phy          = phy_pruned,
        data         = sp_means_df,
        names.col    = "tip_label",
        vcv          = TRUE,
        warn.dropped = FALSE
      ),
      error = function(e) {
        message("comparative.data() failed: ", conditionMessage(e))
        NULL
      }
    )

    if (!is.null(comp_data)) {
      cat(sprintf(
        "Comparative data object ready: %d species in pruned tree\n",
        length(phy_pruned$tip.label)
      ))
    }

  } else {
    cat(sprintf(
      "Only %d species overlap between data and tree (minimum 10 required) \u2014 PGLS skipped.\n",
      length(shared_tips)
    ))
    sp_means <- sp_means |> filter(tip_label %in% shared_tips)
  }

} else if (!USE_REAL_DATA) {
  message("PGLS setup skipped: requires real SAPFLUXNET data (USE_REAL_DATA = TRUE).")
}
```

```{r reg-comparison}
# ── Per-relationship helper: run OLS + SMA (already fitted) + PGLS ────────────
# Relationships: E~DBH, E~BA, E~SA
# Returns a one-row tibble for the comparison table.

fit_ols         <- NULL   # E~DBH OLS (kept for reg-comparison-plot)
fit_pgls        <- NULL   # E~DBH PGLS (kept for lambda-interpretation)
pgls_lambda_val <- NA_real_
pgls_n          <- NA_integer_

run_pgls_rel <- function(y_col, x_col, comp_d) {
  tryCatch({
    fm  <- as.formula(paste(y_col, "~", x_col))
    fit <- caper::pgls(fm, data = comp_d, lambda = "ML")
    sm  <- summary(fit)
    cm  <- sm$coefficients
    sl  <- cm[x_col, "Estimate"]
    se  <- cm[x_col, "Std. Error"]
    pv  <- cm[x_col, "Pr(>|t|)"]
    r2  <- sm$r.sq
    lam <- as.numeric(fit$param["lambda"])
    lci <- tryCatch({
      ci <- fit$param.CI$lambda
      c(lo = unname(ci["lower.CI"]), hi = unname(ci["upper.CI"]))
    }, error = function(e) c(lo = NA_real_, hi = NA_real_))
    list(fit = fit, slope = sl, ci_lo = sl - 1.96*se, ci_hi = sl + 1.96*se,
         p = pv, r2 = r2, lambda = lam, lambda_lo = lci["lo"], lambda_hi = lci["hi"],
         n = nrow(comp_d$data))
  }, error = function(e) {
    message("PGLS failed (", y_col, " ~ ", x_col, "): ", conditionMessage(e))
    NULL
  })
}

if (USE_REAL_DATA && !is.null(sp_means) && nrow(sp_means) >= 10 &&
    !is.null(comp_data)) {

  # ── OLS on all individual trees for each relationship ──────────────────────
  fit_ols     <- lm(log_e ~ log_dbh, data = dat)   # kept for plot below
  fit_ols_ba2 <- lm(log_e ~ log_ba,  data = dat)
  fit_ols_sa2 <- lm(log_e ~ log_sa,  data = dat)

  ols_n <- nrow(dat)
  sma_n <- nrow(dat)

  ols_rows <- list(
    list(rel = "E ~ DBH", x = "log_dbh",
         fit = fit_ols,     sma_fit = fit_dbh,
         ols_ci = confint(fit_ols)["log_dbh", ],
         sma_ci_lo = fit_dbh$coef[[1]][2,2], sma_ci_hi = fit_dbh$coef[[1]][2,3]),
    list(rel = "E ~ BA",  x = "log_ba",
         fit = fit_ols_ba2, sma_fit = fit_ba,
         ols_ci = confint(fit_ols_ba2)["log_ba", ],
         sma_ci_lo = fit_ba$coef[[1]][2,2],  sma_ci_hi = fit_ba$coef[[1]][2,3]),
    list(rel = "E ~ SA",  x = "log_sa",
         fit = fit_ols_sa2, sma_fit = fit_sa,
         ols_ci = confint(fit_ols_sa2)["log_sa", ],
         sma_ci_lo = fit_sa$coef[[1]][2,2],  sma_ci_hi = fit_sa$coef[[1]][2,3])
  )

  # ── PGLS for each relationship ─────────────────────────────────────────────
  pgls_dbh <- run_pgls_rel("log_e", "log_dbh", comp_data)
  pgls_ba  <- run_pgls_rel("log_e", "log_ba",  comp_data)
  pgls_sa  <- run_pgls_rel("log_e", "log_sa",  comp_data)

  # Store primary (E~DBH) PGLS for downstream chunks
  if (!is.null(pgls_dbh)) {
    fit_pgls        <- pgls_dbh$fit
    pgls_lambda_val <- pgls_dbh$lambda
    pgls_n          <- pgls_dbh$n
  }

  # ── Build unified comparison table ────────────────────────────────────────
  make_rows <- function(row_spec, pgls_res, method_prefix = NULL) {
    x_col   <- row_spec$x
    rel     <- row_spec$rel
    sma_summ <- summary(lm(log_e ~ ., data = dat[, c("log_e", x_col)]))  # for SMA p unused
    ols_sl  <- coef(row_spec$fit)[x_col]
    sma_sl  <- coef(row_spec$sma_fit)[2]
    sma_p   <- tryCatch(row_spec$sma_fit$pval[[1]], error = function(e) NA_real_)

    tibble(
      Relationship = rel,
      Method       = c("OLS (all trees)", "SMA / RMA (all trees)", "PGLS (sp. means)"),
      n            = c(ols_n, sma_n, if (!is.null(pgls_res)) pgls_res$n else NA_integer_),
      Slope        = round(c(ols_sl, sma_sl,
                              if (!is.null(pgls_res)) pgls_res$slope else NA_real_), 3),
      CI_low       = round(c(row_spec$ols_ci[1], row_spec$sma_ci_lo,
                              if (!is.null(pgls_res)) pgls_res$ci_lo else NA_real_), 3),
      CI_high      = round(c(row_spec$ols_ci[2], row_spec$sma_ci_hi,
                              if (!is.null(pgls_res)) pgls_res$ci_hi else NA_real_), 3),
      R2           = round(c(summary(row_spec$fit)$r.squared,
                              row_spec$sma_fit$r2[[1]],
                              if (!is.null(pgls_res)) pgls_res$r2 else NA_real_), 3),
      pvalue       = signif(c(
                       summary(row_spec$fit)$coefficients[x_col, "Pr(>|t|)"],
                       sma_p,
                       if (!is.null(pgls_res)) pgls_res$p else NA_real_), 3),
      lambda_ML    = c(NA_real_, NA_real_,
                       if (!is.null(pgls_res)) round(pgls_res$lambda, 3) else NA_real_)
    )
  }

  comp_table_full <- bind_rows(
    make_rows(ols_rows[[1]], pgls_dbh),
    make_rows(ols_rows[[2]], pgls_ba),
    make_rows(ols_rows[[3]], pgls_sa)
  )

  kable(
    comp_table_full,
    col.names = c("Relationship", "Method", "n", "Slope", "CI low", "CI high",
                  "R\u00b2", "p-value", "\u03bb (Pagel)"),
    caption = sprintf(
      "Regression method comparison for E ~ DBH, E ~ BA, and E ~ SA (log-log scale). OLS and SMA use all individual trees (n = %d); PGLS uses species means (n = %d species in pruned phylogeny). Pagel\u2019s \u03bb estimated by ML. OLS is the primary regression (Model I); SMA is Model II (symmetric residuals). PGLS accounts for shared evolutionary history.",
      ols_n, if (!is.null(pgls_dbh)) pgls_dbh$n else NA_integer_
    )
  )

  # Summary of PGLS lambda across relationships
  lambda_summary <- bind_rows(
    if (!is.null(pgls_dbh)) tibble(Relationship="E~DBH", lambda=pgls_dbh$lambda,
                                    lambda_lo=pgls_dbh$lambda_lo, lambda_hi=pgls_dbh$lambda_hi),
    if (!is.null(pgls_ba))  tibble(Relationship="E~BA",  lambda=pgls_ba$lambda,
                                    lambda_lo=pgls_ba$lambda_lo,  lambda_hi=pgls_ba$lambda_hi),
    if (!is.null(pgls_sa))  tibble(Relationship="E~SA",  lambda=pgls_sa$lambda,
                                    lambda_lo=pgls_sa$lambda_lo,  lambda_hi=pgls_sa$lambda_hi)
  )

  cat("\nPagel's lambda (ML) per relationship:\n")
  print(lambda_summary)

} else {
  message(
    "Regression comparison skipped: requires USE_REAL_DATA = TRUE ",
    "and >= 10 species with both data and phylogeny coverage."
  )
}
```

```{r reg-comparison-plot, fig.height=5.5, fig.cap="E ~ DBH: OLS (blue solid, primary), SMA (black solid, secondary), PGLS (red dashed, species means). Open circles = species means in PGLS. OLS and SMA use all individual trees; PGLS degrees of freedom = species in the pruned phylogeny."}
if (USE_REAL_DATA && !is.null(sp_means) && nrow(sp_means) >= 10 &&
    !is.null(fit_pgls) && !is.null(phy_pruned)) {

  x_rng <- range(dat$log_dbh, na.rm = TRUE)
  x_seq <- seq(x_rng[1], x_rng[2], length.out = 200)

  ols_cf  <- coef(fit_ols)
  sma_cf  <- coef(fit_dbh)
  pgls_cf <- coef(fit_pgls)

  pred_lines <- bind_rows(
    tibble(log_dbh = x_seq,
           log_e   = ols_cf["(Intercept)"] + ols_cf["log_dbh"] * x_seq,
           Method  = "OLS (all trees)"),
    tibble(log_dbh = x_seq,
           log_e   = sma_cf[1] + sma_cf[2] * x_seq,
           Method  = "RMA/SMA (all trees)"),
    tibble(log_dbh = x_seq,
           log_e   = pgls_cf["(Intercept)"] + pgls_cf["log_dbh"] * x_seq,
           Method  = "PGLS (species means)")
  )

  line_cols  <- c(
    "OLS (all trees)"      = "#2166ac",
    "RMA/SMA (all trees)"  = "#333333",
    "PGLS (species means)" = "#d6604d"
  )
  line_lty <- c(
    "OLS (all trees)"      = "solid",
    "RMA/SMA (all trees)"  = "solid",
    "PGLS (species means)" = "dashed"
  )

  sp_means_pgls <- sp_means |> filter(tip_label %in% phy_pruned$tip.label)

  ggplot() +
    geom_point(
      data = dat,
      aes(x = log_dbh, y = log_e, shape = growth_form),
      colour = "grey55", alpha = 0.13, size = 0.9
    ) +
    geom_line(
      data = pred_lines,
      aes(x = log_dbh, y = log_e, colour = Method, linetype = Method),
      linewidth = 1.15
    ) +
    geom_point(
      data = sp_means_pgls,
      aes(x = log_dbh, y = log_e),
      shape = 21, fill = "white", colour = "#d6604d",
      size = 2.8, alpha = 0.85
    ) +
    scale_colour_manual(values = line_cols,  name = "Regression") +
    scale_linetype_manual(values = line_lty, name = "Regression") +
    scale_shape_manual(
      values = c(Angiosperm = 16, Conifer = 17),
      name   = "Growth form"
    ) +
    guides(
      colour   = guide_legend(order = 1,
                              override.aes = list(linewidth = 1.5, alpha = 1)),
      linetype = guide_legend(order = 1),
      shape    = guide_legend(order = 2,
                              override.aes = list(colour = "grey40",
                                                  alpha  = 0.8, size = 2))
    ) +
    labs(
      x        = expression(log[10]~"Stem diameter (cm)"),
      y        = expression(log[10]~"Whole-tree water use (L day"^{-1}*")"),
      title    = expression(E[tree]~"~ DBH: OLS vs. RMA/SMA vs. PGLS"),
      subtitle = sprintf(
        "n = %d trees (grey)  |  %d species means (open circles) used in PGLS",
        nrow(dat), nrow(sp_means_pgls)
      )
    ) +
    theme(legend.position = "bottom", legend.box = "horizontal")

} else {
  message(
    "Regression comparison plot skipped: requires USE_REAL_DATA = TRUE ",
    "and a successful PGLS fit."
  )
}
```

```{r lambda-interpretation, results='asis'}
if (USE_REAL_DATA && !is.na(pgls_lambda_val) && !is.null(fit_pgls)) {

  lambda_ci <- tryCatch({
    ci <- fit_pgls$param.CI$lambda
    c(lo = unname(ci["lower.CI"]), hi = unname(ci["upper.CI"]))
  }, error = function(e) c(lo = NA_real_, hi = NA_real_))

  ols_slope_val  <- coef(fit_ols)["log_dbh"]
  pgls_slope_val <- coef(fit_pgls)["log_dbh"]
  slope_diff     <- abs(ols_slope_val - pgls_slope_val)

  lambda_near_1 <- pgls_lambda_val > 0.7
  lambda_near_0 <- pgls_lambda_val < 0.3
  ci_lo         <- lambda_ci["lo"]
  ci_hi         <- lambda_ci["hi"]
  ci_incl_0     <- !is.na(ci_lo) && ci_lo < 0.05
  ci_incl_1     <- !is.na(ci_hi) && ci_hi > 0.95

  msg_signal <- if (lambda_near_1) {
    paste0(
      "\u03bb \u2248 ", round(pgls_lambda_val, 2),
      ": strong phylogenetic signal. Residual variance in species-mean E tracks ",
      "the phylogenetic covariance matrix, consistent with Brownian-motion ",
      "evolution of hydraulic traits."
    )
  } else if (lambda_near_0) {
    paste0(
      "\u03bb \u2248 ", round(pgls_lambda_val, 2),
      ": negligible phylogenetic signal. Species means behave as approximately ",
      "independent observations; OLS and PGLS slopes converge."
    )
  } else {
    paste0(
      "\u03bb = ", round(pgls_lambda_val, 2),
      ": intermediate phylogenetic signal. Residual covariance exceeds a ",
      "star-phylogeny expectation but falls below full Brownian-motion covariance."
    )
  }

  msg_ci <- if (ci_incl_0 && ci_incl_1) {
    "The 95% CI spans 0 to 1: the estimate is imprecise and the data are consistent with either no signal or full Brownian covariance. Interpret with caution."
  } else if (ci_incl_0) {
    "The 95% CI includes 0: OLS would be a defensible alternative to PGLS for these data."
  } else if (ci_incl_1) {
    "The 95% CI includes 1: full Brownian-motion covariance cannot be excluded."
  } else {
    "The 95% CI excludes both 0 and 1: intermediate phylogenetic signal is well-supported."
  }

  msg_slopes <- if (slope_diff > 0.1) {
    sprintf(
      "OLS slope (%.3f) and PGLS slope (%.3f) differ by %.3f log-log units, ",
      ols_slope_val, pgls_slope_val, slope_diff
    ) |> paste0(
      "suggesting that phylogenetic non-independence meaningfully shifts the ",
      "estimated scaling exponent."
    )
  } else {
    sprintf(
      "OLS slope (%.3f) and PGLS slope (%.3f) differ by only %.3f log-log units; ",
      ols_slope_val, pgls_slope_val, slope_diff
    ) |> paste0(
      "phylogenetic correction has minimal effect on the scaling exponent in this dataset."
    )
  }

  ci_lo_str <- if (is.na(ci_lo)) "NA" else sprintf("%.3f", ci_lo)
  ci_hi_str <- if (is.na(ci_hi)) "NA" else sprintf("%.3f", ci_hi)

  cat(paste0(
    "**Pagel's \u03bb interpretation**\n\n",
    sprintf(
      "Pagel's \u03bb (ML estimate): **%.3f** (95%% CI: %s \u2013 %s; n\u202f=\u202f%d species in PGLS).\n\n",
      pgls_lambda_val, ci_lo_str, ci_hi_str, pgls_n
    ),
    msg_signal, "\n\n",
    msg_ci, "\n\n",
    msg_slopes, "\n"
  ))

} else {
  cat("> Pagel's \u03bb not available: requires USE_REAL_DATA = TRUE and a successful PGLS fit.\n")
}
```

## Caveats for the PGLS analysis

1. **Species-mean aggregation.** PGLS uses per-species means, collapsing all within-species variation. The estimated slope may therefore differ from the individual-tree slope, because between-species differences in body-size range and hydraulic architecture are confounded in the species mean. The OLS and SMA slopes on all $n$ individual trees remain the primary scaling estimates; PGLS adds a phylogenetic-signal perspective on the species-level relationship.

2. **Placement uncertainty.** V.PhyloMaker2 places species absent from the LCVP backbone (GBOTB.extended.LCVP) by binding them to their genus node (scenario S3). For poorly represented genera, this introduces error in branch lengths and therefore in the variance\u2013covariance matrix used by PGLS. Rare or taxonomically uncertain species should be treated with caution.

3. **Reliability of $\lambda$ with few species.** Pagel's $\lambda$ is estimated from the residual variance\u2013covariance structure. When the number of species in the PGLS is small (< 30), the ML estimate is unstable and the confidence interval wide. Check the reported $n_{\text{PGLS}}$; if substantially smaller than the full species count, the $\lambda$ estimate may be unreliable.

4. **Phylogenetic sampling bias.** SAPFLUXNET species are not a random phylogenetic sample. Temperate Pinaceae (conifers) and Fagaceae (deciduous angiosperms) are over-represented relative to tropical and subtropical lineages. This affects both the $\lambda$ estimate and the generalisability of the PGLS slope across the broader tree of life.

5. **Temporal-averaging offset.** All values use active-flow-timestep means rather than growing-season daily means as in Forrester et al.\ (2022). This affects intercepts but not slopes in the log\u2013log regression, provided the averaging bias is size-independent (see data-loading caveats above).

---

# Data status and download instructions

```{r data-status}
if (!USE_REAL_DATA) {
  cat("⚠ DATA STATUS: This analysis used RECONSTRUCTED DEMONSTRATION DATA.\n\n",
      "To run with real SAPFLUXNET data:\n",
      "1. Download SAPFLUXNET v0.1.5 (~3.2 GB) from:\n",
      "   https://doi.org/10.5281/zenodo.3971689\n",
      "2. Extract the zip. You will get a folder: 0.1.5/csv/sapwood/\n",
      "   containing ~202 site directories, each with:\n",
      "     <SITE>_plant_md.csv   — diameter, sapwood area, species\n",
      "     <SITE>_sapf_data.csv  — hourly sap flow density time series\n",
      "3. Copy the sapwood/ folder to:\n",
      "   plant_scaling_data/data/raw/sapfluxnet/\n",
      "4. Re-render this Rmd. The real-data path will activate automatically.\n\n",
      "Note: Forrester et al. (2022) used the growing-season mean daily water\n",
      "use per plant. Their supplementary provides the exact aggregation rules.\n")
} else {
  cat("✓ Real SAPFLUXNET data loaded successfully.\n",
      "  Plant metadata files found:", length(plant_md_files), "\n",
      "  Sap flow files found:", length(sap_files), "\n")
}
```

---

# References

* Forrester DI, Limousin J-M, Pfautsch S (2022). The relationship between tree
  size and tree water-use: is competition for water size-symmetric or
  size-asymmetric? *Tree Physiology* **42**: 1916–1927.
  DOI: 10.1093/treephys/tpac018

* Meinzer FC, Bond BJ, Warren JM, Woodruff DR (2005). Does water transport
  scale universally with tree size? *Functional Ecology* **19**: 558–565.
  DOI: 10.1111/j.1365-2435.2005.01017.x

* Poyatos R, et al. (2021). Global transpiration data from sap flow
  measurements: the SAPFLUXNET database. *Earth System Science Data* **13**:
  2607–2649. DOI: 10.5194/essd-13-2607-2021

* SAPFLUXNET v0.1.5 (2020). Zenodo. DOI: 10.5281/zenodo.3971689

* Zimmermann MH (1983). *Xylem Structure and the Ascent of Sap*. Springer.

* West GB, Brown JH, Enquist BJ (1997). A general model for the origin of
  allometric scaling laws in biology. *Science* **276**: 122–126.
  DOI: 10.1126/science.276.5309.122

* Shinozaki K, Yoda K, Hozumi K, Kira T (1964). A quantitative analysis of
  plant form — the pipe model theory. *Japanese Journal of Ecology* **14**: 97–105.

* Bates D, Maechler M, Bolker B, Walker S (2015). Fitting linear mixed-effects
  models using lme4. *Journal of Statistical Software* **67**: 1–48.
  DOI: 10.18637/jss.v067.i01

* Ludecke D, Ben-Shachar MS, Patil I, Waggoner P, Makowski D (2021).
  performance: An R package for assessment, comparison and testing of
  statistical models. *Journal of Open Source Software* **6**: 3139.
  DOI: 10.21105/joss.03139

* Koenker R, Bassett G (1978). Regression quantiles. *Econometrica* **46**: 33–50.
  DOI: 10.2307/1913643

* Warton DI, Hui FKC (2012). The arcsine is asinine: the analysis of proportions
  in ecology. *Ecology* **92**: 3–10.

> **⚠ Citation verification:** All DOIs, journal names, and volume/page details
> should be independently verified against Crossref before use in any manuscript
> or public report.

---

*Analysis via `smatr` in R `r R.version$major`.`r R.version$minor`.*
*Data source: SAPFLUXNET v0.1.5, Zenodo DOI: 10.5281/zenodo.3971689.*
*Synthesis reference: Forrester et al. 2022, Tree Physiology 42: 1916–1927.*
