---
title: "Tallo — Global Tree Allometry and Crown Architecture Database"
author: "plant_scaling_data project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)
library(stringr)

theme_set(theme_bw(base_size = 12))

raw_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "data/raw/tallo")

tallo <- read_csv(file.path(raw_dir, "Tallo.csv"), show_col_types = FALSE)
refs  <- read_csv(file.path(raw_dir, "Tallo_references.csv"),
                  locale = locale(encoding = "latin1"), show_col_types = FALSE) |>
  mutate(source = iconv(source, from = "latin1", to = "UTF-8", sub = "?"))
meta  <- read_csv(file.path(raw_dir, "Tallo_metadata.csv"), show_col_types = FALSE)
```

---

# Overview

**Tallo** (v1.0.0) is a global database of individual tree measurements compiled
by Jucker et al. (2022). It contains georeferenced records of stem diameter,
height, and crown radius for nearly 500,000 trees drawn from 61,856 sites
spanning all major biomes. All records are taxonomically standardized and
every record carries exact latitude/longitude coordinates.

> Jucker T, Fischer FO, Chave J, Coomes DA, Caspersen J, et al. (2022)
> Tallo: A global tree allometry and crown architecture database.
> *Global Change Biology*, 28: 5254–5268.
> DOI: [10.1111/gcb.16302](https://doi.org/10.1111/gcb.16302)
>
> Data: Zenodo [doi:10.5281/zenodo.6637599](https://doi.org/10.5281/zenodo.6637599)

> **⚠ Citation note:** The DOI and journal reference above should be verified
> against Crossref before use in any manuscript.

---

# Top-level summary

```{r top-counts}
kable(data.frame(
  Metric = c("Total records","Records with stem diameter","Records with height",
             "Records with crown radius","Records with all three",
             "Unique species","Unique genera","Unique families",
             "Unique sites (approx.)", "Lat/lon coverage",
             "Lat range (°)","Lon range (°)","Source studies"),
  Value = c(
    format(nrow(tallo), big.mark=","),
    format(sum(!is.na(tallo$stem_diameter_cm)), big.mark=","),
    format(sum(!is.na(tallo$height_m)),         big.mark=","),
    format(sum(!is.na(tallo$crown_radius_m)),   big.mark=","),
    format(sum(!is.na(tallo$stem_diameter_cm) & !is.na(tallo$height_m) &
               !is.na(tallo$crown_radius_m)),  big.mark=","),
    format(length(unique(tallo$species[!is.na(tallo$species)])), big.mark=","),
    format(length(unique(tallo$genus[!is.na(tallo$genus)])),     big.mark=","),
    format(length(unique(tallo$family[!is.na(tallo$family)])),   big.mark=","),
    "61,856 (per Jucker et al. 2022)",
    paste0(sum(!is.na(tallo$latitude) & !is.na(tallo$longitude)),
           " (100%)"),
    paste0(round(min(tallo$latitude),1), " to ", round(max(tallo$latitude),1)),
    paste0(round(min(tallo$longitude),1), " to ", round(max(tallo$longitude),1)),
    format(length(unique(tallo$reference_id)), big.mark=",")
  )
), caption = "Tallo database top-level summary")
```

---

# Column dictionary

```{r dict}
kable(meta, caption = "Tallo column descriptions (from Tallo_metadata.csv)")
```

---

# Data availability

```{r availability, fig.height=4}
avail <- data.frame(
  variable = c("stem_diameter_cm","height_m","crown_radius_m",
               "species","genus","family","division"),
  n_present = c(
    sum(!is.na(tallo$stem_diameter_cm)),
    sum(!is.na(tallo$height_m)),
    sum(!is.na(tallo$crown_radius_m)),
    sum(!is.na(tallo$species)),
    sum(!is.na(tallo$genus)),
    sum(!is.na(tallo$family)),
    sum(!is.na(tallo$division))
  ),
  total = nrow(tallo)
) |> mutate(pct = n_present / total * 100)

ggplot(avail, aes(x = reorder(variable, pct), y = pct)) +
  geom_col(fill = "#2166ac") +
  geom_text(aes(label = paste0(round(pct), "%")), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 110), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "% non-missing",
       title = paste0("Data completeness (n = ", format(nrow(tallo), big.mark=","), " records)"))
```

---

# Global geographic coverage

```{r geo-map, fig.width=11, fig.height=5.5}
# Thin to ~50k points for fast plotting
set.seed(42)
tallo_thin <- tallo |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  sample_n(min(50000, n()))

world_map <- map_data("world")

ggplot() +
  geom_polygon(data = world_map,
               aes(x = long, y = lat, group = group),
               fill = "grey88", color = "grey70", linewidth = 0.15) +
  geom_point(data = tallo_thin,
             aes(x = longitude, y = latitude),
             alpha = 0.15, size = 0.4, color = "#1b7837") +
  coord_fixed(1.3) +
  labs(title = paste0("Tallo — sampling locations (n = ",
                      format(nrow(tallo), big.mark=","),
                      " trees; 50,000 plotted for clarity)"),
       x = NULL, y = NULL) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

```{r geo-division, fig.width=11, fig.height=5.5}
tallo_div <- tallo |>
  filter(!is.na(latitude), !is.na(longitude), !is.na(division)) |>
  sample_n(min(50000, n()))

ggplot() +
  geom_polygon(data = world_map,
               aes(x = long, y = lat, group = group),
               fill = "grey88", color = "grey70", linewidth = 0.15) +
  geom_point(data = tallo_div,
             aes(x = longitude, y = latitude, color = division),
             alpha = 0.2, size = 0.5) +
  scale_color_manual(values = c(Angiosperm = "#d6604d", Gymnosperm = "#4393c3"),
                     na.value = "grey50") +
  coord_fixed(1.3) +
  guides(color = guide_legend(override.aes = list(alpha=1, size=3))) +
  labs(title = "Tallo — Angiosperm vs. Gymnosperm distribution",
       color = "Division", x = NULL, y = NULL) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom")
```

---

# Taxonomic coverage

## Division breakdown

```{r div-bar, fig.height=3}
tallo |>
  mutate(division = ifelse(is.na(division), "Unknown", division)) |>
  count(division) |>
  ggplot(aes(x = reorder(division, n), y = n, fill = division)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = format(n, big.mark=",")), hjust = -0.1, size = 4) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c(Angiosperm="#d6604d", Gymnosperm="#4393c3", Unknown="grey60")) +
  labs(x = NULL, y = "N trees", title = "Records by division")
```

## Top 30 families

```{r top-families, fig.height=7}
tallo |>
  filter(!is.na(family)) |>
  count(family, sort = TRUE) |>
  head(30) |>
  ggplot(aes(x = reorder(family, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = format(n, big.mark=",")), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_viridis_c(direction = -1) +
  labs(x = NULL, y = "N trees", title = "Top 30 families by record count")
```

## Top 30 species

```{r top-species, fig.height=7}
tallo |>
  filter(!is.na(species)) |>
  count(species, sort = TRUE) |>
  head(30) |>
  ggplot(aes(x = reorder(species, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = format(n, big.mark=",")), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_viridis_c(direction = -1) +
  labs(x = NULL, y = "N trees",
       title = "Top 30 species by record count",
       caption = "Species names as standardized in Tallo v1.0.0")
```

---

# Size distributions

```{r dbh-dist}
tallo |> filter(!is.na(stem_diameter_cm), stem_diameter_cm > 0) |>
  ggplot(aes(x = stem_diameter_cm)) +
  geom_histogram(bins = 80, fill = "#2166ac", color = "white") +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +
  annotation_logticks(sides = "b") +
  labs(x = "Stem diameter (cm, log scale)", y = "Count",
       title = paste0("Stem diameter distribution  (n = ",
                      format(sum(!is.na(tallo$stem_diameter_cm)), big.mark=","), ")"),
       subtitle = paste0("Range: ", round(min(tallo$stem_diameter_cm, na.rm=TRUE),1),
                         " – ", round(max(tallo$stem_diameter_cm, na.rm=TRUE),1), " cm"))
```

```{r height-dist}
tallo |> filter(!is.na(height_m), height_m > 0) |>
  ggplot(aes(x = height_m)) +
  geom_histogram(bins = 80, fill = "#1b7837", color = "white") +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +
  annotation_logticks(sides = "b") +
  labs(x = "Height (m, log scale)", y = "Count",
       title = paste0("Height distribution  (n = ",
                      format(sum(!is.na(tallo$height_m)), big.mark=","), ")"),
       subtitle = paste0("Range: ", round(min(tallo$height_m, na.rm=TRUE),1),
                         " – ", round(max(tallo$height_m, na.rm=TRUE),1), " m"))
```

```{r crown-dist}
tallo |> filter(!is.na(crown_radius_m), crown_radius_m > 0) |>
  ggplot(aes(x = crown_radius_m)) +
  geom_histogram(bins = 80, fill = "#762a83", color = "white") +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +
  annotation_logticks(sides = "b") +
  labs(x = "Crown radius (m, log scale)", y = "Count",
       title = paste0("Crown radius distribution  (n = ",
                      format(sum(!is.na(tallo$crown_radius_m)), big.mark=","), ")"),
       subtitle = paste0("Range: ", round(min(tallo$crown_radius_m, na.rm=TRUE),2),
                         " – ", round(max(tallo$crown_radius_m, na.rm=TRUE),2), " m"))
```

---

# Scaling relationships

All regressions below are **exploratory OLS on log10-transformed data** using a 
10% random sample for speed. Use SMA (`smatr`) for formal allometric inference.

```{r sample-subset}
set.seed(123)
sub <- tallo |>
  filter(!is.na(stem_diameter_cm), stem_diameter_cm > 0,
         !is.na(height_m), height_m > 0) |>
  sample_frac(0.10)
```

## Height vs. stem diameter

```{r ht-dbh}
lm1 <- lm(log10(height_m) ~ log10(stem_diameter_cm), data = sub)
b1 <- round(coef(lm1)[2], 3)
r2 <- round(summary(lm1)$r.squared, 3)
n1 <- nrow(sub)

ggplot(sub, aes(x = stem_diameter_cm, y = height_m)) +
  geom_point(alpha = 0.08, size = 0.5, color = "#1b7837") +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(x = "Stem diameter (cm)", y = "Height (m)",
       title = paste0("Height ~ DBH  |  β = ", b1, "  |  R² = ", r2,
                      "  |  n = ", format(n1, big.mark=",")),
       subtitle = "10% random subsample; OLS on log10-transformed data")
```

```{r ht-dbh-div}
sub_div <- tallo |>
  filter(!is.na(stem_diameter_cm), stem_diameter_cm > 0,
         !is.na(height_m), height_m > 0, !is.na(division)) |>
  sample_frac(0.10)

ggplot(sub_div, aes(x = stem_diameter_cm, y = height_m, color = division)) +
  geom_point(alpha = 0.06, size = 0.5) +
  geom_smooth(method = "lm", linewidth = 1.2, se = FALSE) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  scale_color_manual(values = c(Angiosperm = "#d6604d", Gymnosperm = "#4393c3")) +
  guides(color = guide_legend(override.aes = list(alpha=1, size=3))) +
  labs(x = "Stem diameter (cm)", y = "Height (m)",
       title = "Height ~ DBH by division  (10% subsample)",
       color = "Division")
```

## Crown radius vs. stem diameter

```{r crown-dbh}
sub2 <- tallo |>
  filter(!is.na(stem_diameter_cm), stem_diameter_cm > 0,
         !is.na(crown_radius_m), crown_radius_m > 0) |>
  sample_frac(0.10)

lm2 <- lm(log10(crown_radius_m) ~ log10(stem_diameter_cm), data = sub2)
b2 <- round(coef(lm2)[2], 3)
r2b <- round(summary(lm2)$r.squared, 3)
n2 <- nrow(sub2)

ggplot(sub2, aes(x = stem_diameter_cm, y = crown_radius_m)) +
  geom_point(alpha = 0.08, size = 0.5, color = "#762a83") +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  scale_x_log10(labels = comma) + scale_y_log10(labels = comma) +
  annotation_logticks() +
  labs(x = "Stem diameter (cm)", y = "Crown radius (m)",
       title = paste0("Crown radius ~ DBH  |  β = ", b2, "  |  R² = ", r2b,
                      "  |  n = ", format(n2, big.mark=",")),
       subtitle = "10% random subsample; OLS on log10-transformed data")
```

## Summary of scaling exponents (exploratory OLS)

```{r exponent-table}
kable(data.frame(
  Relationship = c("Height ~ stem_diameter", "Crown_radius ~ stem_diameter"),
  OLS_beta     = c(b1, b2),
  R_squared    = c(r2, r2b),
  n            = c(n1, n2)
), caption = "Exploratory OLS exponents (log10-log10). Use SMA for formal allometric inference.")
```

---

# Outlier flags

Tallo includes two outlier flag columns: `height_outlier` and `crown_radius_outlier`.

```{r outlier-summary}
kable(data.frame(
  Variable = c("height_outlier","crown_radius_outlier"),
  n_flagged = c(
    sum(tallo$height_outlier == "yes", na.rm=TRUE),
    sum(tallo$crown_radius_outlier == "yes", na.rm=TRUE)
  ),
  pct_flagged = round(c(
    100 * mean(tallo$height_outlier == "yes", na.rm=TRUE),
    100 * mean(tallo$crown_radius_outlier == "yes", na.rm=TRUE)
  ), 2)
), caption = "Outlier flags in Tallo (flag = 'yes' per Jucker et al. 2022 criteria)")
```

> Jucker et al. (2022) recommend excluding flagged records for allometric
> model fitting. For exploratory analysis both are retained here.

---

# Source references

```{r refs-table}
kable(refs, escape = TRUE,
      caption = paste0("All ", nrow(refs), " source references in Tallo_references.csv"))
```

---

# BIEN export readiness assessment

This section documents a coordinate, taxonomy, and trait QA performed in May 2026
to assess Tallo v1.0.0 for loading into the BIEN biodiversity database as
georeferenced species observations.

## Species completeness

```{r bien-species-completeness}
sp_complete <- tallo |>
  summarise(
    n_total          = n(),
    n_with_species   = sum(!is.na(species)),
    n_missing_species = sum(is.na(species)),
    pct_missing      = round(100 * mean(is.na(species)), 1)
  )
kable(sp_complete,
      caption = "Species completeness — records lacking a species binomial cannot be imported to BIEN")

# Top references contributing missing species
missing_by_ref <- tallo |>
  filter(is.na(species)) |>
  count(reference_id, sort = TRUE) |>
  head(10)
kable(missing_by_ref,
      caption = "Top 10 references contributing NA species (see Tallo_references.csv for source details)")
```

**Conclusion:** 61,230 records (12.3%) lack a species binomial and must be excluded from
any BIEN occurrence export. This leaves **437,608 candidate records**.

## Coordinate precision and uncertainty

Coordinates in Tallo range from GPS-precise individual-tree locations (≥3 decimal
places, ~100 m precision) to coarse plot centroids (0–2 decimal places, 1–100 km
precision). The number of decimal places provides a conservative estimate of
`coordinateUncertaintyInMeters` required by Darwin Core.

```{r bien-coord-precision}
tallo_id <- tallo |> filter(!is.na(species))

tallo_id <- tallo_id |>
  mutate(
    lat_dp  = nchar(sub("^[^.]*[.]?", "", as.character(abs(latitude)))),
    lon_dp  = nchar(sub("^[^.]*[.]?", "", as.character(abs(longitude)))),
    min_dp  = pmin(as.integer(lat_dp), as.integer(lon_dp)),
    coord_precision_class = case_when(
      min_dp == 0 ~ "0 dp  (~100 km) — integer",
      min_dp == 1 ~ "1 dp  (~11 km)  — likely centroid",
      min_dp == 2 ~ "2 dp  (~1 km)   — likely centroid",
      min_dp >= 3 ~ "3+ dp (~100 m)  — acceptable"
    ),
    coordinateUncertaintyInMeters = case_when(
      min_dp == 0 ~ 111000L,
      min_dp == 1 ~  11100L,
      min_dp == 2 ~   1100L,
      TRUE        ~     NA_integer_
    ),
    is_plot_centroid = min_dp <= 2
  )

prec_summary <- tallo_id |>
  count(coord_precision_class) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  arrange(coord_precision_class)
kable(prec_summary,
      caption = "Coordinate precision classes in BIEN-candidate records (species non-NA only)")
```

```{r bien-coord-precision-bar, fig.height=3.5}
ggplot(prec_summary, aes(x = reorder(coord_precision_class, -n), y = n, fill = coord_precision_class)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(format(n, big.mark=","), "  (", pct, "%)")),
            hjust = -0.05, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.3))) +
  scale_fill_manual(values = c(
    "0 dp  (~100 km) — integer"    = "#d73027",
    "1 dp  (~11 km)  — likely centroid" = "#fc8d59",
    "2 dp  (~1 km)   — likely centroid" = "#fee090",
    "3+ dp (~100 m)  — acceptable" = "#1a9641"
  )) +
  labs(x = NULL, y = "N records",
       title = "Coordinate precision classes — BIEN candidate records",
       subtitle = "Records with ≤2 decimal places are plot centroids, not individual tree GPS")
```

```{r bien-repeated-coords}
top_coords <- tallo_id |>
  count(latitude, longitude, sort = TRUE) |>
  head(10)
kable(top_coords,
      caption = "Top 10 most-repeated coordinate pairs (confirms plot centroid structure)")
```

## Coordinate anomalies

```{r bien-coord-anomalies}
# lon = 0 cases
lon0 <- tallo |> filter(longitude == 0)
kable(lon0 |> select(tree_id, genus, species, latitude, longitude,
                     stem_diameter_cm, height_m, reference_id),
      caption = "Records with longitude = 0 (reference 35; Quercus ilex, Spain — verify against source publication)")

# Implausible height record
implausible <- tallo |> filter(is.na(species) & !is.na(height_m) & height_m > 90)
if (nrow(implausible) > 0) {
  kable(implausible |> select(tree_id, genus, species, latitude, longitude,
                               height_m, stem_diameter_cm, reference_id),
        caption = "Unidentified records with height > 90 m — ecologically implausible for most regions; quarantine before BIEN import")
}
```

## BIEN candidate pool summary

```{r bien-candidate-summary}
pool <- tallo_id |>
  summarise(
    `Total BIEN candidates (species non-NA)` = n(),
    `With 3+ dp coords (GPS quality)`        = sum(min_dp >= 3),
    `Plot centroids (≤2 dp)`                 = sum(is_plot_centroid),
    `With height_m`                          = sum(!is.na(height_m)),
    `With crown_radius_m`                    = sum(!is.na(crown_radius_m)),
    `height_outlier flagged`                 = sum(height_outlier == "Y"),
    `crown_radius_outlier flagged`           = sum(crown_radius_outlier == "Y"),
    `Height < 2 m (sapling/juvenile)`        = sum(!is.na(height_m) & height_m < 2)
  ) |>
  tidyr::pivot_longer(everything(), names_to = "Metric", values_to = "N") |>
  mutate(Pct = round(100 * N / n(), 1))
kable(pool, caption = "BIEN candidate pool summary (437,608 records with species binomial)")
```

## Trait units — BIEN canonical mapping

All three Tallo trait fields are already in appropriate SI-derived units; no
numeric conversion is required. Field renaming and protocol flagging are needed.

```{r bien-trait-units}
kable(data.frame(
  `Tallo field`        = c("stem_diameter_cm", "height_m", "crown_radius_m"),
  `BIEN canonical`     = c("stem_diameter_cm", "plant_height_m", "crown_radius_m"),
  Unit                 = c("cm", "m", "m"),
  `Conversion needed`  = c("None", "None", "None"),
  Notes = c(
    "Multi-stemmed trees use sqrt(sum(Di²)) — flag with measurement_protocol = 'multi_stem_quadratic_sum'",
    "Outliers flagged by source authors; preserve flag as measurementRemarks",
    "37.3% missing; missingness is by-reference (non-random) — do not impute"
  )
), caption = "Tallo → BIEN canonical trait unit mapping")
```

## Outstanding items before BIEN import

```{r bien-outstanding}
kable(data.frame(
  Priority  = c("CRITICAL","CRITICAL","CRITICAL","CRITICAL","HIGH","HIGH","HIGH"),
  Issue = c(
    "Exclude 61,230 NA-species records",
    "Attach coordinateUncertaintyInMeters (see precision class above)",
    "Extract eventDate/year from Tallo_references.csv per reference_id",
    "TNRS taxonomic reconciliation of 5,164 species against BIEN/WCVP backbone",
    "Reverse-geocode country/state/county via GADM (sf + geodata)",
    "Flag multi-stem compound diameter records by reference",
    "Quarantine T_481206 (102.7 m tree, no species, Georgia USA — likely unit error)"
  ),
  Status = c(
    "Pipeline implemented in scripts/tallo_bien_preexport.R",
    "Pipeline implemented in scripts/tallo_bien_preexport.R",
    "Year range extraction requires manual review of Tallo_references.csv",
    "Requires TNRS package or API call — not yet run",
    "Pipeline implemented in scripts/tallo_bien_preexport.R",
    "Requires per-reference study metadata — flag added for known refs",
    "Record quarantined in pipeline"
  )
), caption = "BIEN import outstanding action items (May 2026)")
```

---

# Key caveats

1. **Crown radius only available for 63% of records.** Stem diameter is
   complete for all 498,838 trees; height is available for 99.7%; crown radius
   drops to 62.7%. Analyses involving crown architecture should account for
   this non-random missingness.

2. **Outlier flags.** Tallo provides `height_outlier` and `crown_radius_outlier`
   columns. Records flagged `"Y"` are statistical outliers within allometric
   regressions fitted by Jucker et al. (2022) — they are not necessarily
   biologically implausible. Preserve these flags; do not silently delete.

3. **No biomass column.** Tallo records structural dimensions only (DBH, height,
   crown radius). Biomass must be *estimated* from allometric equations (e.g.,
   Chave 2014, allodb), not measured directly. Tallo is a complement to, not
   a substitute for, the Niklas-Enquist or BAAD biomass datasets.

4. **Taxonomic standardization:** Species names were harmonized against The Plant
   List (v1.1) by Jucker et al. — additional reconciliation against a current
   backbone (World Flora Online, TNRS) is recommended before merging.

5. **Georeferenced:** All 498,838 records carry decimal-degree coordinates, but
   ~38% are plot centroids (≤2 decimal places) rather than individual-tree GPS.
   See the BIEN export section above for precision classes and
   `coordinateUncertaintyInMeters` assignments.

6. **No observation date.** Individual records carry no measurement date; date
   information must be extracted per source publication via `reference_id`.

---

# Session info

```{r session}
sessionInfo()
```
