---
title: "Tree Longevity Data Curation Catalog"
subtitle: "Running summary of datasets, download status, and QA"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(here)
library(knitr)
library(kableExtra)
source(here("R", "catalog_utils.R"))

# Load download log if present
dl_log <- tryCatch(
  read.csv(here("logs", "download_log.csv"), stringsAsFactors = FALSE),
  error = function(e) data.frame()
)

# Helper: look up most recent log entry for a dataset_id
last_status <- function(id) {
  if (nrow(dl_log) == 0) return(list(status = "NOT RUN", date = NA, size = NA, hash = NA, notes = ""))
  rows <- dl_log[dl_log$dataset_id == id, , drop = FALSE]
  if (nrow(rows) == 0) return(list(status = "NOT RUN", date = NA, size = NA, hash = NA, notes = ""))
  r <- rows[nrow(rows), ]
  list(status  = as.character(r$http_status),
       date    = as.character(r$download_timestamp),
       size    = r$file_size_bytes,
       hash    = as.character(r$sha256),
       notes   = as.character(r$notes))
}
```

---

## Overview

This catalog documents 20 published datasets related to tree longevity, traits, and climate. 
For each dataset it records: source citation, access type, download status, file provenance (SHA-256, size), 
key variables, units, and any blockers that prevent automated download.

**Framework designed by:** scholarly-rigor-reviewer, biodiversity-science-guard, biodiversity-informatics-checker  
**Project root:** `tree-longevity-data-curation/`  
**Download log:** `logs/download_log.csv` (committed; tracks every attempt)  
**Data files:** `data/raw/` (gitignored; large files not committed)

---

## Download Status Summary

```{r summary-table}
datasets <- data.frame(
  ID      = sprintf("D%02d", 1:20),
  Dataset = c(
    "Main study longevity/traits/climate (Figshare 29876984)",
    "ITRDB tree ring data (NOAA)",
    "Tropical tree rings — Locoselli et al. 2020 (Figshare 13119842)",
    "OldList West (RMTRR)",
    "OldList East (LDEO Columbia)",
    "Native Tree Society dendro ages",
    "Old Growth Canada",
    "Tree height (Zenodo 6637599)",
    "Conifer max height (conifers.org)",
    "MonumentalTrees max height",
    "Wood density (Zenodo 13322441)",
    "Wood density (CIRAD DataVerse DVN1/KRVF0E)",
    "Conduit density + P50 + HSM (Dryad 10.5061/dryad.1138)",
    "P50 + HSM (Science Advances aav1332 supp.)",
    "GLOPNET leaf traits (Wright et al. 2004 Nature)",
    "Seed mass (TRY request 30569)",
    "TreeGOER climate/soil (Zenodo 10008994)",
    "WorldClim 2.1 gridded climate/elevation",
    "CHELSA growing season + NPP",
    "GBIF species occurrence (dl.77gcvq)"
  ),
  Downloadable = c(
    "YES", "PARTIAL", "YES",
    "PARTIAL", "PARTIAL", "PARTIAL", "NO",
    "YES", "NO", "NO",
    "YES", "YES",
    "YES", "PARTIAL", "PARTIAL",
    "NO", "YES", "YES", "YES", "PARTIAL"
  ),
  Method = c(
    "Figshare API v2", "FTP/dplR subset", "Figshare API v2",
    "rvest scrape", "rvest scrape", "rvest scrape", "Manual — maintainer contact",
    "Zenodo API", "Manual — ToS review req.", "Manual — ToS review req.",
    "Zenodo API", "DataVerse API",
    "Dryad API v2", "Manual — CDN URL", "Manual — Nature supp / TRY",
    "Manual — TRY registration", "Zenodo API", "geodata R pkg",
    "direct curl CHELSA CDN", "rgbif::occ_download_get"
  ),
  stringsAsFactors = FALSE
)

# Pull actual statuses from log
datasets$Last_Status <- sapply(datasets$ID, function(id) {
  s <- last_status(id)$status
  if (is.na(s) || s == "NOT RUN") "Not attempted" else as.character(s)
})

datasets$Download_Date <- sapply(datasets$ID, function(id) {
  d <- last_status(id)$date
  if (is.na(d)) "—" else substr(d, 1, 10)
})

kable(datasets,
      col.names = c("ID", "Dataset", "Downloadable", "Method",
                    "Last Status", "Date"),
      align = "llllll",
      escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12) |>
  row_spec(which(datasets$Downloadable == "YES"),
           background = "#f0fff0") |>
  row_spec(which(datasets$Downloadable == "PARTIAL"),
           background = "#fffbea") |>
  row_spec(which(datasets$Downloadable == "NO"),
           background = "#fff0f0")
```

---

## D01 — Main Study Longevity/Traits/Climate

```{r d01}
s <- last_status("D01")
```

| Field | Value |
|---|---|
| **Citation** | Study authors (in review). Tree longevity dataset. Figshare. DOI: 10.6084/m9.figshare.29876984 |
| **DOI / URL** | https://doi.org/10.6084/m9.figshare.29876984 |
| **Folder** | `data/raw/01_main_study/` |
| **License** | To be confirmed — verify at Figshare record |
| **Access type** | Open (if not embargoed) |
| **Download status** | `r status_badge("YES")` |
| **Method** | Figshare REST API v2 |
| **Script** | `R/download_01_main_study.R` |
| **Download date** | `r s$date` |
| **File size (bytes)** | `r s$size` |
| **SHA-256** | `r s$hash` |
| **Key variables** | Species max longevity, functional traits, mean climate per species |
| **Biological unit** | Species |
| **Known issues** | **VERIFY:** DOI not independently confirmed; record may be embargoed if paper is in review |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D02 — ITRDB Tree Ring Data (NOAA)

```{r d02}
s <- last_status("D02")
```

| Field | Value |
|---|---|
| **Citation** | International Tree-Ring Data Bank. NOAA National Centers for Environmental Information. https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring |
| **DOI / URL** | https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring |
| **Folder** | `data/raw/02_itrdb/` |
| **License** | Open — NOAA public domain |
| **Access type** | Open (FTP) |
| **Download status** | `r status_badge("PARTIAL")` |
| **Method** | NOAA FTP; `dplR` package for `.rwl` parsing |
| **Script** | `R/download_02_itrdb.R` |
| **Download date** | `r s$date` |
| **Key variables** | Ring-width chronologies, site metadata, species codes |
| **Biological unit** | Chronology (site × species) |
| **Known issues** | No single bulk download. Species/site selection criteria must be defined before downloading. File format: Tucson `.rwl`. |
| **Action required** | Define chronology selection criteria (species list, geographic bounding box, minimum length). Update `download_02_itrdb.R` with selection logic. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D03 — Tropical Tree Rings (Locoselli et al. 2020)

```{r d03}
s <- last_status("D03")
```

| Field | Value |
|---|---|
| **Citation** | Locoselli GM et al. (2020) Global tree-ring analysis reveals rapid decrease in tropical tree longevity with temperature. PNAS 117(52):33358–33364. https://doi.org/10.1073/pnas.2003873117 |
| **DOI / URL** | https://figshare.com/articles/dataset/Locoselli_et_al_2020.../13119842 |
| **Folder** | `data/raw/03_tropical_rings/` |
| **License** | CC BY 4.0 (verify at Figshare) |
| **Access type** | Open |
| **Download status** | `r status_badge("YES")` |
| **Method** | Figshare REST API v2 |
| **Script** | `R/download_03_tropical_rings.R` |
| **Download date** | `r s$date` |
| **File size (bytes)** | `r s$size` |
| **SHA-256** | `r s$hash` |
| **Key variables** | Tropical tree ring widths, species, site metadata, longevity estimates |
| **Biological unit** | Individual tree / chronology |
| **Known issues** | None identified at design stage |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D04 — OldList West (RMTRR)

```{r d04}
s <- last_status("D04")
```

| Field | Value |
|---|---|
| **Citation** | Rocky Mountain Tree Ring Research. OldList — a database of maximum tree ages. http://www.rmtrr.org/oldlist.htm. Accessed: `r Sys.Date()` |
| **URL** | http://www.rmtrr.org/oldlist.htm |
| **Folder** | `data/raw/04_oldlist_west/` |
| **License** | Not stated — cite as accessed with date |
| **Access type** | Open (HTML table) |
| **Download status** | `r status_badge("PARTIAL")` |
| **Method** | `rvest` HTML table scrape; raw HTML archived |
| **Script** | `R/download_04_oldlist_west.R` |
| **Download date** | `r s$date` |
| **Key variables** | Species name, maximum age (years), location, reference |
| **Biological unit** | Species maximum longevity record |
| **Known issues** | No DOI or persistent identifier. HTTP (not HTTPS). Page structure may change. Scrape is fragile — verify column names after each run. |
| **Provenance note** | Raw HTML archived alongside CSV. Record scrape date as version. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D05 — OldList East (LDEO Columbia)

```{r d05}
s <- last_status("D05")
```

| Field | Value |
|---|---|
| **Citation** | Larson D (LDEO Columbia). OldList East. https://www.ldeo.columbia.edu/~adk/oldlisteast/. Accessed: `r Sys.Date()` |
| **URL** | https://www.ldeo.columbia.edu/~adk/oldlisteast/ |
| **Folder** | `data/raw/05_oldlist_east/` |
| **License** | Not stated |
| **Access type** | Open (HTML) |
| **Download status** | `r status_badge("PARTIAL")` |
| **Method** | `rvest` HTML scrape |
| **Script** | `R/download_05_oldlist_east.R` |
| **Download date** | `r s$date` |
| **Key variables** | Species, maximum age (years), location, reference |
| **Biological unit** | Species maximum longevity record |
| **Known issues** | Personal-directory URL (`~adk`) — inherently unstable. Columbia LDEO may reorganize at any time. Screenshot + HTML archive mandatory. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D06 — Native Tree Society Dendro Ages

```{r d06}
s <- last_status("D06")
```

| Field | Value |
|---|---|
| **Citation** | Native Tree Society. Maximum tree ages — dendrochronological data. http://www.nativetreesociety.org/dendro/ents_maximum_ages.htm. Accessed: `r Sys.Date()` |
| **URL** | http://www.nativetreesociety.org/dendro/ents_maximum_ages.htm |
| **Folder** | `data/raw/06_nts_dendro/` |
| **License** | Not stated |
| **Access type** | Open (HTML) — HTTP only |
| **Download status** | `r status_badge("PARTIAL")` |
| **Method** | `rvest` HTML scrape; may require manual extraction if no table found |
| **Script** | `R/download_06_nts_dendro.R` |
| **Download date** | `r s$date` |
| **Key variables** | Species, maximum age, location, reference |
| **Biological unit** | Species maximum longevity record |
| **Known issues** | Community-maintained page; irregular table structure. Manual review required after scrape. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D07 — Old Growth Canada

```{r d07}
s <- last_status("D07")
```

| Field | Value |
|---|---|
| **Citation** | OldGrowth.ca. Old trees database. https://www.oldgrowth.ca/oldtrees/. Accessed: `r Sys.Date()` |
| **URL** | https://www.oldgrowth.ca/oldtrees/ |
| **Folder** | `data/raw/07_oldgrowth_canada/` |
| **License** | Not stated |
| **Access type** | Open (HTML only — no structured download) |
| **Download status** | `r status_badge("NO")` |
| **Method** | Manual — maintainer contact required |
| **Script** | `R/download_07_oldgrowth_canada.R` (writes README only) |
| **Known issues** | Records presented as HTML text/galleries; no structured table or API. No DOI or persistent identifier. |
| **Action required** | Contact OldGrowth.ca maintainers to request a data export. Document: contact date, response, fields received, attribution terms. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D08 — Tree Height (Zenodo 6637599)

```{r d08}
s <- last_status("D08")
```

| Field | Value |
|---|---|
| **Citation** | Verify citation from Zenodo record metadata at https://zenodo.org/record/6637599 |
| **DOI** | https://doi.org/10.5281/zenodo.6637599 |
| **Folder** | `data/raw/08_tree_height_zenodo/` |
| **License** | Verify at Zenodo (likely CC BY 4.0) |
| **Access type** | Open |
| **Download status** | `r status_badge("YES")` |
| **Method** | Zenodo REST API v1 |
| **Script** | `R/download_08_tree_height_zenodo.R` |
| **Download date** | `r s$date` |
| **File size (bytes)** | `r s$size` |
| **SHA-256** | `r s$hash` |
| **Key variables** | Species, maximum height (m), source |
| **Biological unit** | Species |
| **Unit note** | Verify unit (meters assumed); check README in Zenodo record |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D09 — Conifer Maximum Height (conifers.org)

```{r d09}
s <- last_status("D09")
```

| Field | Value |
|---|---|
| **Citation** | Earle CJ. The Gymnosperm Database. https://www.conifers.org. Accessed: `r Sys.Date()` |
| **URL** | https://www.conifers.org |
| **Folder** | `data/raw/09_conifers_height/` |
| **License** | Not stated — ToS review required |
| **Access type** | Open website; no bulk download |
| **Download status** | `r status_badge("NO")` |
| **Method** | Per-species page scraping feasible (~900 pages) but requires ToS clearance |
| **Script** | `R/download_09_conifers_height.R` (writes README only) |
| **Known issues** | No bulk download or API. Per-species scraping requires `polite` package for robots.txt compliance + 5 s rate limit. ToS must be reviewed before any scraping. |
| **Action required** | 1. Review robots.txt at https://www.conifers.org/robots.txt. 2. If permitted, implement per-species scrape with `polite`. 3. Alternatively, contact C.J. Earle for bulk export. |
| **Unit note** | Heights reported in meters. Verify per record. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D10 — MonumentalTrees Maximum Height

```{r d10}
s <- last_status("D10")
```

| Field | Value |
|---|---|
| **Citation** | MonumentalTrees.com. Community tree measurement database. https://www.monumentaltrees.com. Accessed: `r Sys.Date()` |
| **URL** | https://www.monumentaltrees.com |
| **Folder** | `data/raw/10_monumental_height/` |
| **License** | Not stated — ToS review required |
| **Access type** | Open website; no bulk download |
| **Download status** | `r status_badge("NO")` |
| **Method** | Manual — ToS review + maintainer contact recommended |
| **Script** | `R/download_10_monumental_height.R` (writes README only) |
| **Known issues** | Community-contributed; per-record provenance unclear. Do not use as primary source without independent verification. No DOI. |
| **Action required** | 1. Review ToS and robots.txt. 2. Contact administrator for bulk export. 3. Flag all records as 'unverified community observation' in analyses. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D11 — Wood Density (Zenodo 13322441)

```{r d11}
s <- last_status("D11")
```

| Field | Value |
|---|---|
| **Citation** | Verify citation from Zenodo record at https://zenodo.org/records/13322441 |
| **DOI** | https://doi.org/10.5281/zenodo.13322441 |
| **Folder** | `data/raw/11_wood_density_zenodo/` |
| **License** | Verify at Zenodo |
| **Access type** | Open |
| **Download status** | `r status_badge("YES")` |
| **Method** | Zenodo REST API v1 |
| **Script** | `R/download_11_wood_density_zenodo.R` |
| **Download date** | `r s$date` |
| **File size (bytes)** | `r s$size` |
| **SHA-256** | `r s$hash` |
| **Key variables** | Species, wood density |
| **Biological unit** | Species (mean) |
| **Unit note** | **CRITICAL:** Verify unit — g/cm³ vs kg/m³ before merging with D12. 1 g/cm³ = 1000 kg/m³. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D12 — Wood Density (CIRAD DataVerse)

```{r d12}
s <- last_status("D12")
```

| Field | Value |
|---|---|
| **Citation** | Verify citation at https://doi.org/10.18167/DVN1/KRVF0E |
| **DOI** | https://doi.org/10.18167/DVN1/KRVF0E |
| **Folder** | `data/raw/12_wood_density_dv/` |
| **License** | Verify at CIRAD DataVerse |
| **Access type** | Open (CIRAD DataVerse, France) |
| **Download status** | `r status_badge("YES")` |
| **Method** | DataVerse native API |
| **Script** | `R/download_12_wood_density_dv.R` |
| **Download date** | `r s$date` |
| **File size (bytes)** | `r s$size` |
| **SHA-256** | `r s$hash` |
| **Known issues** | DOI prefix 10.18167 is CIRAD DataVerse (France); may require API token even for open data. Script falls back with clear error message. |
| **Unit note** | **CRITICAL:** Audit wood density units before merging with D11. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D13 — Conduit Density + P50 + HSM (Dryad)

```{r d13}
s <- last_status("D13")
```

| Field | Value |
|---|---|
| **Citation** | Verify citation at https://doi.org/10.5061/dryad.1138 |
| **DOI** | https://doi.org/10.5061/dryad.1138 |
| **Folder** | `data/raw/13_conduit_p50_dryad/` |
| **License** | CC0 (Dryad default) |
| **Access type** | Open |
| **Download status** | `r status_badge("YES")` |
| **Method** | Dryad API v2 |
| **Script** | `R/download_13_conduit_p50_dryad.R` |
| **Download date** | `r s$date` |
| **File size (bytes)** | `r s$size` |
| **SHA-256** | `r s$hash` |
| **Key variables** | Conduit density (n/mm²), P50 (MPa), HSM (MPa), species |
| **Biological unit** | Species |
| **Unit note** | **CRITICAL — P50 sign convention:** Negative MPa expected. Run `check_p50_sign()` and compare with D14 before merging. |
| **Known issues** | Short-form Dryad DOI (`dryad.1138`) — verify resolution. Dryad API v2 endpoint used. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D14 — P50 + HSM (Science Advances aav1332)

```{r d14}
s <- last_status("D14")
```

| Field | Value |
|---|---|
| **Citation** | Choat B et al. (2018) Triggers of tree mortality under drought. Science Advances 4(1):eaav1332. https://doi.org/10.1126/sciadv.aav1332 |
| **DOI** | https://doi.org/10.1126/sciadv.aav1332 (paper DOI — not data DOI) |
| **Folder** | `data/raw/14_p50_sci_advances/` |
| **License** | Open access (CC BY-NC) |
| **Access type** | Supplementary file — may require institutional access |
| **Download status** | `r status_badge("PARTIAL")` |
| **Method** | Manual — locate and download supplementary file from paper landing page |
| **Script** | `R/download_14_p50_sci_advances.R` (writes README only) |
| **Action required** | Navigate to https://doi.org/10.1126/sciadv.aav1332, download supplementary data file, save to `data/raw/14_p50_sci_advances/`. Update catalog with filename, download date, and institutional access requirement. |
| **Unit note** | P50 in MPa. **Audit sign convention** vs D13 before merging. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D15 — GLOPNET Leaf Traits (Wright et al. 2004)

```{r d15}
s <- last_status("D15")
```

| Field | Value |
|---|---|
| **Citation** | Wright IJ et al. (2004) The worldwide leaf economics spectrum. Nature 428:821–827. https://doi.org/10.1038/nature02403 |
| **DOI** | https://doi.org/10.1038/nature02403 (paper); data via Nature supplementary or TRY |
| **Folder** | `data/raw/15_glopnet_leaf/` |
| **License** | Nature supplementary — verify. TRY data: see TRY terms. |
| **Access type** | Supplementary file (Nature) or TRY database request |
| **Download status** | `r status_badge("PARTIAL")` |
| **Method** | Manual (Nature supp) or TRY database request (recommended) |
| **Script** | `R/download_15_glopnet_leaf.R` (writes README only) |
| **Key variables** | LL, LMA, Nmass, Narea, Pmass, Amass, Aarea, Rd mass |
| **Biological unit** | Species mean (leaf trait per species) |
| **Action required** | Option A: Download Nature supplementary file from https://www.nature.com/articles/nature02403. Option B (recommended): Request GLOPNET traits via TRY database (see D16 for TRY process). |
| **Unit note** | Verify units from paper Table S1: mass-based (mg/g, nmol/g/s) vs area-based (g/m², μmol/m²/s). |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D16 — Seed Mass (TRY Database Request 30569)

```{r d16}
s <- last_status("D16")
```

| Field | Value |
|---|---|
| **Citation** | Kattge J et al. (2020) TRY plant trait database — enhanced coverage and open access. Global Change Biology 26:119–188. https://doi.org/10.1111/gcb.14904 |
| **DOI / URL** | https://www.try-db.org |
| **Folder** | `data/raw/16_try_seed_mass/` |
| **License** | TRY-specific — **data cannot be redistributed** |
| **Access type** | Registration + formal data request required |
| **Download status** | `r status_badge("NO")` |
| **Method** | Manual — TRY account + new data request submission |
| **Script** | `R/download_16_try_seed_mass.R` (writes README only) |
| **Key variables** | Seed/diaspore mass (mg dry mass), species, reference |
| **TRY Trait IDs** | TraitID 26 (seed mass) or TraitID 4 (diaspore mass) |
| **Original request** | Request No. 30569 (study-specific; not reproducible by others without new request) |
| **Action required** | 1. Register at https://www.try-db.org/TryWeb/Account/Register. 2. Submit new request for TraitID 26 (seed mass). 3. Document: request ID, TRY version, submission date, approval date. |
| **DATA RESTRICTION** | TRY data MUST NOT be committed to git or redistributed. |
| **Unit note** | Standard: mg dry mass. Verify `UnitName` column — some records may be in g. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D17 — TreeGOER Climate/Soil (Zenodo 10008994)

```{r d17}
s <- last_status("D17")
```

| Field | Value |
|---|---|
| **Citation** | Verify citation from Zenodo record at https://zenodo.org/records/10008994 |
| **DOI** | https://doi.org/10.5281/zenodo.10008994 |
| **Folder** | `data/raw/17_treegOER_climate/` |
| **License** | Verify at Zenodo |
| **Access type** | Open |
| **Download status** | `r status_badge("YES")` |
| **Method** | Zenodo REST API v1 |
| **Script** | `R/download_17_treegOER_climate.R` |
| **Download date** | `r s$date` |
| **File size (bytes)** | `r s$size` |
| **SHA-256** | `r s$hash` |
| **Key variables** | Species-level mean climate (MAT, MAP, etc.) and soil variables |
| **Biological unit** | Species |
| **Known issues** | Verify that TreeGOER provides species-level MEAN climate values (not raw occurrence grid values). Unit audit required for all soil variables. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D18 — WorldClim 2.1 Gridded Climate/Elevation

```{r d18}
s <- last_status("D18")
```

| Field | Value |
|---|---|
| **Citation** | Fick SE, Hijmans RJ (2017) WorldClim 2: new 1-km spatial resolution climate surfaces for global land areas. International Journal of Climatology 37:4302–4315. https://doi.org/10.1002/joc.5086 |
| **DOI / URL** | https://www.worldclim.org/data/worldclim21.html |
| **Folder** | `data/raw/18_worldclim/` |
| **License** | CC BY 4.0 |
| **Access type** | Open |
| **Download status** | `r status_badge("YES")` |
| **Method** | `geodata::worldclim_global()` R package |
| **Script** | `R/download_18_worldclim.R` |
| **Download date** | `r s$date` |
| **Resolution used** | 10 arc-min (development default; change `RES` in script for production) |
| **Variables** | tmin, tmax, prec, bio (bioclim 1–19), elev |
| **Unit note** | Temperature: °C × 10 (BIO1–BIO11); precipitation: mm; elevation: m |
| **Known issues** | Full resolution (30 arc-sec) is ~8 GB. Use 10 arc-min for development. Document final resolution in methods. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D19 — CHELSA Growing Season + NPP

```{r d19}
s <- last_status("D19")
```

| Field | Value |
|---|---|
| **Citation** | Karger DN et al. (2017) Climatologies at high resolution for the earth's land surface areas (CHELSA). Scientific Data 4:170122. https://doi.org/10.1038/sdata.2017.122 |
| **DOI / URL** | https://chelsa-climate.org/ |
| **Folder** | `data/raw/19_chelsa/` |
| **License** | CC BY 4.0 |
| **Access type** | Open |
| **Download status** | `r status_badge("YES")` |
| **Method** | Direct `curl` — CHELSA SwitchDrive CDN |
| **Script** | `R/download_19_chelsa.R` |
| **Download date** | `r s$date` |
| **Variables** | kg_gsl (growing season length, days), npp (net primary productivity, kg C/m²/yr) |
| **Known issues** | Individual files are 200–600 MB each. CHELSA URL conventions must match v2.1 exactly — verify at https://chelsa-climate.org/downloads/. |
| **Unit note** | GSL: days; NPP: kg C/m²/yr; Verify units in CHELSA v2.1 technical specification. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## D20 — GBIF Species Occurrence (dl.77gcvq)

```{r d20}
s <- last_status("D20")
```

| Field | Value |
|---|---|
| **Citation** | GBIF.org. GBIF Occurrence Download. https://doi.org/10.15468/dl.77gcvq. Accessed: see download date |
| **DOI** | https://doi.org/10.15468/dl.77gcvq |
| **Folder** | `data/raw/20_gbif_occurrence/` |
| **License** | CC BY 4.0 (GBIF standard) |
| **Access type** | Open — GBIF prepared download (expires after ~6 months) |
| **Download status** | `r status_badge("PARTIAL")` |
| **Method** | `rgbif::occ_download_get()` |
| **Script** | `R/download_20_gbif_occurrence.R` |
| **Download date** | `r s$date` |
| **File format** | Darwin Core Archive ZIP (occurrence.txt + meta.xml + eml.xml) |
| **Key DwC fields** | species, decimalLatitude, decimalLongitude, eventDate, basisOfRecord, occurrenceID, countryCode |
| **Biological unit** | Occurrence record |
| **Known issues** | **GBIF downloads expire after ~6 months.** Verify key `dl.77gcvq` is still active. If expired, re-download requires documenting original filter criteria (taxon key, date range, basisOfRecord, spatial filters). |
| **Action required if expired** | Submit new `rgbif::occ_download()` request. Document original filter set in this catalog. |
| **QA status** | `r qa_badge("NOT RUN")` |

---

## Critical Issues Tracker

```{r critical-issues}
issues <- data.frame(
  ID      = c("D01", "D13", "D14", "D16", "D20"),
  Issue   = c(
    "Figshare record 29876984 not independently verified — may be embargoed",
    "P50 sign convention must be audited before merging with D14",
    "P50 sign convention must be audited before merging with D13; supplementary file may require institutional access",
    "TRY data cannot be redistributed; original request 30569 not reproducible by others",
    "GBIF download dl.77gcvq may have expired; original filter criteria must be documented"
  ),
  Priority = c("HIGH", "CRITICAL", "CRITICAL", "HIGH", "HIGH"),
  stringsAsFactors = FALSE
)

kable(issues, col.names = c("Dataset", "Issue", "Priority"),
      align = "lll") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE) |>
  row_spec(which(issues$Priority == "CRITICAL"), background = "#fff0f0", bold = TRUE) |>
  row_spec(which(issues$Priority == "HIGH"), background = "#fffbea")
```

---

## QA Log

```{r qa-log}
qa_log <- tryCatch(
  read.csv(here("logs", "qa_log.csv"), stringsAsFactors = FALSE),
  error = function(e) data.frame()
)
if (nrow(qa_log) == 0) {
  cat("*No QA checks run yet. Run per-dataset QA functions from `R/qa_utils.R` after download.*")
} else {
  kable(qa_log) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 11)
}
```

---

## Download Log

```{r download-log}
if (nrow(dl_log) == 0) {
  cat("*No downloads recorded yet. Run `source('download_all.R')` to start.*")
} else {
  dl_show <- dl_log[, c("dataset_id", "dataset_name", "download_timestamp",
                         "http_status", "file_size_bytes", "download_method", "notes")]
  kable(dl_show,
        col.names = c("ID", "Dataset", "Timestamp", "Status",
                      "Size (bytes)", "Method", "Notes")) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 11)
}
```

---

## R Package Requirements

```{r packages, echo=TRUE, eval=FALSE}
# Install required packages for all download scripts and catalog
install.packages(c(
  "httr2",      # REST API calls (Zenodo, Figshare, Dryad, DataVerse)
  "digest",     # SHA-256 file hashing
  "cli",        # Formatted console messages
  "here",       # Portable file paths
  "rvest",      # HTML scraping (OldList sources)
  "polite",     # Rate-limited, robots.txt-compliant scraping
  "rgbif",      # GBIF occurrence downloads
  "geodata",    # WorldClim download wrapper
  "terra",      # Geospatial raster (CHELSA, WorldClim)
  "dplR",       # ITRDB .rwl tree ring file parsing
  "rmarkdown",
  "knitr",
  "kableExtra"
))
```

---

*Catalog rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")`*  
*Framework designed by scholarly-rigor-reviewer × biodiversity-science-guard × biodiversity-informatics-checker*
