---
title: "Animal Allometric Scaling: Metabolism and Growth Across Taxa"
author: "Animal_scaling_data Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    number_sections: true
    code_folding: hide
---

```{r setup, warning=FALSE, message=FALSE}
set.seed(42)

library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(RColorBrewer)

HAS_SMATR <- requireNamespace("smatr", quietly = TRUE)
if (HAS_SMATR) {
  library(smatr)
} else {
  message("Package 'smatr' not installed. SMA regressions will be skipped. ",
          "Install with: install.packages('smatr')")
}

knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# ── Color palette (Dark2, 8 colors) ─────────────────────────────────────────
pal8 <- brewer.pal(8, "Dark2")

# ── Dataset display labels ───────────────────────────────────────────────────
LABEL_AT  <- "AnimalTraits (Herberstein 2022)"
LABEL_PN  <- "Hoehler 2023 (PNAS)"
LABEL_H19 <- "Hatton 2019 (PNAS)"
LABEL_H15 <- "Hatton 2015 (Science)"
LABEL_CH  <- "Chown 2007 (FuncEcol, corrected)"
LABEL_GR  <- "Grady et al. 2014 (Science)"

HAS_GGREPEL <- requireNamespace("ggrepel", quietly = TRUE)

# ── Load helper ──────────────────────────────────────────────────────────────
load_csv <- function(path, label) {
  if (!file.exists(path)) {
    message("File not found: ", path)
    return(NULL)
  }
  dt <- fread(path, colClasses = list(character = "qa_flag"))
  dt[, dataset_label := label]
  for (col in c("body_mass_g", "metabolic_rate_value", "lifespan_max_years",
                "growth_rate_value", "predator_mass_g_paired",
                "metabolic_rate_at_T_W", "dry_mass_g", "carbon_mass_g")) {
    if (col %in% names(dt))
      dt[, (col) := suppressWarnings(as.numeric(get(col)))]
  }
  dt
}

at   <- load_csv("output/animaltraits_compiled.csv",     LABEL_AT)
pnas <- load_csv("output/pnas_hoehler2023_compiled.csv", LABEL_PN)
if (is.null(pnas))
  pnas <- load_csv("output/pnas_2303764120_compiled.csv", LABEL_PN)
h19  <- load_csv("output/hatton2019_compiled.csv",       LABEL_H19)
h15  <- load_csv("output/hatton2015_compiled.csv",       LABEL_H15)
ch07 <- load_csv("output/chown2007_compiled.csv",        LABEL_CH)

all_list   <- list(at = at, pnas = pnas, h19 = h19, h15 = h15, ch07 = ch07)
labels_vec <- c(at   = LABEL_AT,
                pnas = LABEL_PN,
                h19  = LABEL_H19,
                h15  = LABEL_H15,
                ch07 = LABEL_CH)

# ── Slope plausibility bound ─────────────────────────────────────────────────
# Slopes with |β| > SLOPE_MAX are biologically implausible (likely caused by
# near-zero body-mass variance within a group, producing degenerate regressions).
# Applied consistently to order-, family-, and species-level slope tables.
SLOPE_MAX <- 10

# ── QA filter ────────────────────────────────────────────────────────────────
qa_ok <- function(dt) {
  if (is.null(dt)) return(NULL)
  dt[qa_flag %in% c("ok", "")]
}

# ── OLS log-log regression helper ────────────────────────────────────────────
# Returns data.table with one row per group (plus "All taxa combined").
fit_log_lm <- function(dt, x_col, y_col, group_col = NULL, min_n = 10) {
  if (is.null(dt) || nrow(dt) == 0) return(data.table())
  if (!all(c(x_col, y_col) %in% names(dt))) return(data.table())

  sub <- dt[!is.na(get(x_col)) & get(x_col) > 0 &
            !is.na(get(y_col)) & get(y_col) > 0]
  if (nrow(sub) == 0) return(data.table())

  sub <- copy(sub)
  sub[, lx := log10(get(x_col))]
  sub[, ly := log10(get(y_col))]

  fit_one <- function(d, label) {
    if (nrow(d) < min_n) return(NULL)
    if (length(unique(d$lx)) < 2) return(NULL)   # no x variance; skip
    fit <- tryCatch(lm(ly ~ lx, data = d), error = function(e) NULL)
    if (is.null(fit)) return(NULL)
    s  <- coef(summary(fit))
    if (nrow(s) < 2) return(NULL)                 # singular fit; skip
    ci <- tryCatch(confint(fit), error = function(e) matrix(NA_real_, 2, 2))
    data.table(
      Group    = label,
      n        = nrow(d),
      Intercept = round(s[1, 1], 3),
      Slope    = round(s[2, 1], 3),
      Slope_SE = round(s[2, 2], 3),
      CI_95_lo = round(ci[2, 1], 3),
      CI_95_hi = round(ci[2, 2], 3),
      R2       = round(summary(fit)$r.squared, 3),
      p_value  = signif(s[2, 4], 3)
    )
  }

  res_all <- fit_one(sub, "All taxa combined")

  if (!is.null(group_col) && group_col %in% names(sub)) {
    grps    <- sort(unique(sub[[group_col]]))
    res_grp <- rbindlist(
      lapply(grps, function(g) fit_one(sub[get(group_col) == g], g)),
      fill = TRUE
    )
    res_all <- rbind(res_all, res_grp, fill = TRUE)
  }
  res_all
}

# ── SMA regression helper ────────────────────────────────────────────────────
# Uses smatr::sma(). Returns data.table of results; skipped if smatr absent.
# smatr 4.x coef[[1]] structure: row 1 = slope, row 2 = elevation (intercept);
# columns: coef, lower CI, upper CI, Pval
fit_sma <- function(dt, x_col, y_col, group_col = NULL, min_n = 10) {
  if (!HAS_SMATR) return(data.table())
  if (is.null(dt) || nrow(dt) == 0) return(data.table())
  if (!all(c(x_col, y_col) %in% names(dt))) return(data.table())

  sub <- dt[!is.na(get(x_col)) & get(x_col) > 0 &
            !is.na(get(y_col)) & get(y_col) > 0]
  if (nrow(sub) == 0) return(data.table())

  sub <- copy(sub)
  sub[, lx := log10(get(x_col))]
  sub[, ly := log10(get(y_col))]

  fit_one_sma <- function(d, label) {
    if (nrow(d) < min_n) return(NULL)
    fit <- tryCatch(
      smatr::sma(ly ~ lx, data = as.data.frame(d[, .(lx, ly)])),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NULL)
    cf  <- tryCatch(fit$coef[[1]], error = function(e) NULL)
    if (is.null(cf) || nrow(cf) < 2) return(NULL)
    r2v <- tryCatch(as.numeric(fit$r2[[1]]), error = function(e) NA_real_)
    data.table(
      Group     = label,
      n         = nrow(d),
      Method    = "SMA",
      Intercept = round(as.numeric(cf[2, 1]), 3),
      Slope     = round(as.numeric(cf[1, 1]), 3),
      CI_95_lo  = round(as.numeric(cf[1, 2]), 3),
      CI_95_hi  = round(as.numeric(cf[1, 3]), 3),
      R2        = round(r2v, 3)
    )
  }

  res_all <- fit_one_sma(sub, "All taxa combined")

  if (!is.null(group_col) && group_col %in% names(sub)) {
    grps <- sort(unique(sub[[group_col]]))
    res_grp <- rbindlist(
      lapply(grps, function(g) fit_one_sma(sub[get(group_col) == g], g)),
      fill = TRUE
    )
    res_all <- rbind(res_all, res_grp, fill = TRUE)
  }
  res_all
}

# ── Safe column-subset for rbinding heterogeneous tables ─────────────────────
bind_common <- function(dt, cols) {
  if (is.null(dt)) return(NULL)
  avail <- intersect(cols, names(dt))
  out   <- dt[, avail, with = FALSE]
  for (mc in setdiff(cols, names(out))) out[, (mc) := NA]
  out
}
```

---

# Introduction & Theoretical Background

One of the most robust quantitative patterns in biology is that metabolic rate ($B$), growth rate ($G$), and maximum lifespan ($L$) all scale with body mass ($M$) as power laws with **quarter-power exponents**:

$$B \propto M^{3/4}, \qquad G \propto M^{3/4}, \qquad L \propto M^{1/4}$$

The **West–Brown–Enquist (WBE) metabolic theory** attributes these scaling laws to the geometry of hierarchical fractal-like resource-distribution networks (West et al. 1997, 1999). The Metabolic Theory of Ecology (MTE; Brown et al. 2004) extends these predictions to population- and ecosystem-level patterns.

| Relationship | Predicted exponent ($\beta$) | Interpretation | Reference |
|---|---|---|---|
| Metabolic rate $B \propto M^\beta$ | **0.75** | Kleiber's law; ectotherms may differ | West et al. 1997 |
| Growth rate $G \propto M^\beta$ | **0.75** | Absolute anabolic production; mass-specific scales as $M^{-0.25}$ | West et al. 1997 |
| Maximum lifespan $L \propto M^\beta$ | **0.25** | Inverse of mass-specific metabolic rate | West et al. 1997 |
| Predator mass $M_\text{pred} \propto M_\text{prey}^\beta$ | **~1.0** | Predators are typically 10–1000× heavier than prey | Hatton et al. 2015 |

**Key theoretical nuances:**

- WBE predicts $\beta = 3/4$ for whole-organism metabolic rate. Empirical estimates in mammals and birds are close to 3/4; ectotherms may show slopes closer to 2/3 or higher.
- Unicellular organisms and prokaryotes often show $\beta > 1$ (superlinear scaling), possibly driven by different metabolic architectures. The Hoehler 2023 dataset spans ~40 orders of magnitude in mass (bacteria to whales), constituting the broadest empirical test assembled to date.
- Mixing endotherms, ectotherms, and unicellular organisms in a single regression produces a composite slope that reflects taxonomic composition rather than a universal exponent. All slopes in this document must be interpreted within their sampling context.
- The WBE growth rate prediction ($G \propto M^{3/4}$) refers to **absolute** anabolic production; per-unit-mass turnover (specific growth rate) scales as $M^{-1/4}$, so larger animals grow more slowly relative to body mass.
- Predator–prey mass ratios (Hatton et al. 2015) test body mass co-scaling across trophic levels, not metabolic allometry.

This document synthesizes four complementary published datasets spanning bacteria to whales to evaluate these predictions at multiple scales of taxonomic aggregation.

```{r theory-table, echo=FALSE}
theory_tab <- data.table(
  Relationship = c("Metabolic rate (B ~ M^β)",
                   "Growth rate (G ~ M^β)",
                   "Max lifespan (L ~ M^β)",
                   "Predator mass (M_pred ~ M_prey^β)"),
  Predicted_beta = c(0.75, 0.75, 0.25, 1.00),
  Dataset_tested = c("AnimalTraits, Hoehler 2023, Hatton 2019",
                     "Hatton 2019",
                     "Hatton 2019",
                     "Hatton 2015"),
  Section = c("3–8", "9", "10", "11")
)

kable(theory_tab,
      col.names = c("Relationship", "Predicted β (WBE)", "Dataset(s) tested", "Section"),
      caption = "Summary of WBE theoretical predictions and corresponding document sections.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

```{r qa-pass-rates}
qa_tab <- rbindlist(lapply(names(all_list), function(nm) {
  dt <- all_list[[nm]]
  if (is.null(dt)) return(NULL)
  n_total <- nrow(dt)
  n_ok    <- sum(dt$qa_flag == "ok",  na.rm = TRUE)
  n_blank <- sum(dt$qa_flag == "",    na.rm = TRUE)
  n_pass  <- n_ok + n_blank
  n_fail  <- n_total - n_pass
  pct_pass <- round(100 * n_pass / n_total, 1)
  flag_warn <- if (n_fail / n_total > 0.20) "⚠ >20% failed" else "ok"
  data.table(
    Dataset      = labels_vec[nm],
    N_total      = n_total,
    N_pass_qa    = n_pass,
    N_fail_qa    = n_fail,
    Pct_pass     = pct_pass,
    Flag_warning = flag_warn
  )
}))

kable(qa_tab,
      col.names = c("Dataset", "N total", "N pass QA", "N fail QA", "% pass", "Warning"),
      caption = paste("QA flag pass rates per dataset.",
                      "Records with qa_flag == 'ok' or qa_flag == '' (blank) are treated as QA-pass.",
                      "A warning is flagged when >20% of records fail.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

---

# Data Coverage & Quality

```{r coverage-summary-table}
dois_vec <- c(
  at   = "10.1038/s41597-022-01364-9",
  pnas = "10.1073/pnas.2303764120",
  h19  = "10.1073/pnas.1907703116",
  h15  = "10.1126/science.aac6284"
)

summ_tab <- rbindlist(lapply(names(all_list), function(nm) {
  dt <- all_list[[nm]]
  if (is.null(dt)) return(NULL)
  grps <- paste(sort(unique(
    dt$input_taxonomic_group[!is.na(dt$input_taxonomic_group) & dt$input_taxonomic_group != ""]
  )), collapse = ", ")

  has_mass <- sum(!is.na(dt$body_mass_g) & dt$body_mass_g > 0) > 0
  has_met  <- "metabolic_rate_value" %in% names(dt) &&
              sum(!is.na(dt$metabolic_rate_value) & dt$metabolic_rate_value > 0) > 0
  has_lisp <- "lifespan_max_years" %in% names(dt) &&
              sum(!is.na(dt$lifespan_max_years) & dt$lifespan_max_years > 0) > 0
  has_grow <- "growth_rate_value" %in% names(dt) &&
              sum(!is.na(dt$growth_rate_value) & dt$growth_rate_value > 0) > 0
  has_pred <- "predator_mass_g_paired" %in% names(dt) &&
              sum(!is.na(dt$predator_mass_g_paired) & dt$predator_mass_g_paired > 0) > 0

  traits <- paste(c(
    if (has_mass) "body mass",
    if (has_met)  "metabolic rate",
    if (has_lisp) "lifespan",
    if (has_grow) "growth rate",
    if (has_pred) "predator mass"
  ), collapse = ", ")

  data.table(
    Dataset          = labels_vec[nm],
    DOI              = dois_vec[nm],
    N_rows           = nrow(dt),
    Taxa_covered     = grps,
    Traits_available = traits
  )
}))

kable(summ_tab,
      col.names = c("Dataset", "DOI", "N rows", "Taxa covered", "Traits available"),
      caption = "Summary of the four datasets. Taxa covered lists verbatim taxonomic group labels as supplied by each source.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

```{r coverage-trait-counts}
cov_tab <- rbindlist(lapply(names(all_list), function(nm) {
  dt <- all_list[[nm]]
  if (is.null(dt)) return(NULL)
  mass_ok <- !is.na(dt$body_mass_g) & dt$body_mass_g > 0
  met_ok  <- if ("metabolic_rate_value" %in% names(dt))
               !is.na(dt$metabolic_rate_value) & dt$metabolic_rate_value > 0
             else rep(FALSE, nrow(dt))
  lisp_ok <- if ("lifespan_max_years" %in% names(dt))
               !is.na(dt$lifespan_max_years) & dt$lifespan_max_years > 0
             else rep(FALSE, nrow(dt))
  grow_ok <- if ("growth_rate_value" %in% names(dt))
               !is.na(dt$growth_rate_value) & dt$growth_rate_value > 0
             else rep(FALSE, nrow(dt))
  pred_ok <- if ("predator_mass_g_paired" %in% names(dt))
               !is.na(dt$predator_mass_g_paired) & dt$predator_mass_g_paired > 0
             else rep(FALSE, nrow(dt))
  m_vals   <- dt$body_mass_g[mass_ok]
  mass_rng <- if (length(m_vals) > 0)
                sprintf("%.2e – %.2e", min(m_vals), max(m_vals))
              else "—"
  data.table(
    Dataset  = labels_vec[nm],
    N_total  = nrow(dt),
    N_mass   = sum(mass_ok),
    N_met    = sum(met_ok),
    N_lisp   = sum(lisp_ok),
    N_grow   = sum(grow_ok),
    N_pred   = sum(pred_ok),
    Mass_rng = mass_rng
  )
}))

kable(cov_tab,
      col.names = c("Dataset", "N total", "N body mass", "N metabolic rate",
                    "N lifespan", "N growth rate", "N predator mass", "Body mass range (g)"),
      caption = "Trait availability by dataset (all rows, pre-QA filter). N values reflect rows with positive, non-NA values for each trait.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

```{r coverage-bar, fig.width=11, fig.height=6}
common_cols <- c("dataset_label", "input_taxonomic_group", "body_mass_g",
                 "metabolic_rate_value", "lifespan_max_years", "growth_rate_value", "qa_flag")

all_bound <- rbindlist(
  lapply(all_list, bind_common, cols = common_cols),
  fill = TRUE
)

bar_dat <- all_bound[!is.na(input_taxonomic_group) & input_taxonomic_group != "",
                     .N, by = .(dataset_label, input_taxonomic_group)]

ggplot(bar_dat,
       aes(x = reorder(input_taxonomic_group, -N), y = N, fill = dataset_label)) +
  geom_col() +
  scale_fill_manual(values = pal8[1:5], name = "Dataset") +
  facet_wrap(~ dataset_label, scales = "free", ncol = 2) +
  labs(
    title = "Record count by taxonomic group and dataset",
    x     = "Taxonomic group (verbatim label from source)",
    y     = "Number of records"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

```{r dataset-bar-total, fig.width=7, fig.height=4}
ds_totals <- rbindlist(lapply(names(all_list), function(nm) {
  dt <- all_list[[nm]]
  if (is.null(dt)) return(NULL)
  data.table(Dataset = labels_vec[nm], N = nrow(dt))
}))

ggplot(ds_totals, aes(x = reorder(Dataset, -N), y = N, fill = Dataset)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = scales::comma(N)), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = pal8[1:5]) +
  labs(title = "Total record count per dataset", x = NULL, y = "N records") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r met-type-distribution}
met_type_list <- lapply(names(all_list), function(nm) {
  dt <- all_list[[nm]]
  if (is.null(dt) || !"metabolic_rate_type" %in% names(dt)) return(NULL)
  if (!any(!is.na(dt$metabolic_rate_type) & dt$metabolic_rate_type != "")) return(NULL)
  tab <- dt[!is.na(metabolic_rate_value) & metabolic_rate_value > 0,
            .N, by = metabolic_rate_type][order(-N)]
  tab[, Dataset := labels_vec[nm]]
  tab
})
met_type_tab <- rbindlist(Filter(Negate(is.null), met_type_list), fill = TRUE)

if (nrow(met_type_tab) > 0) {
  kable(met_type_tab[, .(Dataset, metabolic_rate_type, N)],
        col.names = c("Dataset", "Metabolic rate type", "N records"),
        caption = paste("Distribution of metabolic rate measurement types across datasets.",
                        "Mixing basal, resting, active, and field rates inflates variance",
                        "and may bias slope estimates. See Section 3.4 for basal-only analysis.")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

> **Caveat — unreconciled taxonomy:** Taxonomic group labels (`input_taxonomic_group`) are verbatim as supplied by each source and have **not** been harmonized across datasets. The same organism may appear under different labels in different sources (e.g., "invertebrate" vs. "crustacean"). Cross-source comparisons of taxonomic groups should be interpreted with caution until formal taxonomic reconciliation is performed.

---

# Cross-Study Metabolic Scaling — All Taxa

Metabolic theory (WBE; Kleiber 1932) predicts $B \propto M^{3/4}$. We test this across all datasets that include metabolic rate measurements in Watts: AnimalTraits, Hoehler 2023, Hatton 2019, and Chown 2007 (insect-focused, with Riveros & Enquist 2011 corrections).

> **Temperature caveat:** Hoehler et al. (2023) standardized all metabolic rates to 25°C using the Arrhenius equation, making them directly comparable across taxa. AnimalTraits rates are **not** temperature-standardized; measurement temperatures vary by species and study. Hatton et al. (2019) applied Q10-based temperature corrections. The "all combined" slope reflects both biological variation and methodological heterogeneity between datasets.

> **Metabolic rate type caveat:** This section combines all metabolic rate types (basal, resting, active, field). Mixing types inflates variance and may bias slope estimates downward. Section 3.4 presents basal- and resting-only analyses.

## Combined metabolic data preparation {#met-prep}

```{r met-prep}
met_sources <- rbindlist(
  lapply(list(qa_ok(at), qa_ok(pnas), qa_ok(h19), qa_ok(ch07)), function(d) {
    if (is.null(d)) return(NULL)
    d[!is.na(body_mass_g) & body_mass_g > 0 &
      !is.na(metabolic_rate_value) & metabolic_rate_value > 0 &
      metabolic_rate_unit == "W"]
  }),
  fill = TRUE
)
met_basal <- met_sources[metabolic_rate_type %in% c("basal", "resting")]

cat(sprintf("Metabolic data (all types, unit = W):  n = %d\n", nrow(met_sources)))
cat(sprintf("Metabolic data (basal + resting only): n = %d\n", nrow(met_basal)))
cat("\nRecords per dataset:\n")
print(met_sources[, .N, by = dataset_label])
cat("\nMetabolic rate types:\n")
print(table(met_sources$metabolic_rate_type, useNA = "ifany"))
```

## Scatter plot — all taxa combined {#met-scatter}

```{r met-all-taxa-plot, fig.width=10, fig.height=7}
if (nrow(met_sources) >= 10) {
  fit_ov <- lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data = met_sources)
  int_ov <- coef(fit_ov)[1]
  slp_ov <- coef(fit_ov)[2]
  x_rng  <- range(log10(met_sources$body_mass_g), na.rm = TRUE)

  ref_df <- data.frame(lx = x_rng, ly = int_ov + 0.75 * x_rng)
  ols_df <- data.frame(lx = x_rng, ly = int_ov + slp_ov * x_rng)

  ggplot(met_sources,
         aes(x = log10(body_mass_g),
             y = log10(metabolic_rate_value),
             colour = dataset_label)) +
    geom_point(alpha = 0.18, size = 0.7, shape = 16) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.9,
                aes(group = dataset_label)) +
    geom_line(data = ref_df, aes(x = lx, y = ly),
              inherit.aes = FALSE, linetype = "dashed",
              colour = "black", linewidth = 1.0) +
    geom_line(data = ols_df, aes(x = lx, y = ly),
              inherit.aes = FALSE, colour = "black", linewidth = 1.3) +
    scale_colour_manual(values = pal8[1:4], name = "Dataset") +
    annotate("text",
             x = x_rng[1] + 0.5,
             y = int_ov + slp_ov * (x_rng[1] + 0.5) + 1.2,
             label = sprintf("Overall OLS  β = %.3f", slp_ov),
             size = 4, hjust = 0) +
    annotate("text",
             x = x_rng[1] + 0.5,
             y = int_ov + 0.75 * (x_rng[1] + 0.5) - 0.8,
             label = "WBE reference  β = 0.75",
             size = 3.5, hjust = 0, colour = "grey30") +
    labs(
      title    = expression(paste("Cross-study metabolic scaling: ", italic(B), " ∝ ", italic(M)^{beta})),
      subtitle = "Solid black = overall OLS; dashed = WBE reference β = 0.75; colored lines = per-dataset OLS",
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

## OLS regression table — all taxa {#met-ols-table}

```{r met-ols-table}
reg_met_overall <- fit_log_lm(met_sources, "body_mass_g", "metabolic_rate_value", min_n = 10)

if (!is.null(reg_met_overall) && nrow(reg_met_overall) > 0) {
  kable(reg_met_overall,
        col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regression of metabolic rate (W) on body mass (g) — all taxa and studies combined.",
          "All metabolic rate types included; unit = W.",
          "WBE predicted slope: β = 0.75.",
          "OLS assumes no measurement error in body mass; slopes may be attenuated relative to the true relationship.",
          "The combined row mixes datasets with different temperature normalizations."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    row_spec(1, bold = TRUE, background = "#dce8ff")
} else {
  cat("No regression results available for the all-taxa metabolic scaling table.\n")
}
```

## SMA regression — all taxa {#met-sma}

```{r met-sma-plot, fig.width=10, fig.height=7}
if (nrow(met_sources) >= 10 && HAS_SMATR) {
  sma_all <- fit_sma(met_sources, "body_mass_g", "metabolic_rate_value", min_n = 10)

  if (!is.null(sma_all) && nrow(sma_all) > 0) {
    sma_slp <- sma_all$Slope[1]
    sma_int <- sma_all$Intercept[1]
    x_rng2  <- range(log10(met_sources$body_mass_g), na.rm = TRUE)
    sma_line <- data.frame(lx = x_rng2, ly = sma_int + sma_slp * x_rng2)
    ref_df2  <- data.frame(lx = x_rng2, ly = sma_int + 0.75 * x_rng2)

    ggplot(met_sources,
           aes(x = log10(body_mass_g), y = log10(metabolic_rate_value),
               colour = dataset_label)) +
      geom_point(alpha = 0.15, size = 0.6, shape = 16) +
      geom_line(data = ref_df2, aes(x = lx, y = ly),
                inherit.aes = FALSE, linetype = "dashed",
                colour = "black", linewidth = 1.0) +
      geom_line(data = sma_line, aes(x = lx, y = ly),
                inherit.aes = FALSE, colour = "firebrick", linewidth = 1.3) +
      scale_colour_manual(values = pal8[1:4], name = "Dataset") +
      annotate("text",
               x = x_rng2[1] + 0.5,
               y = sma_int + sma_slp * (x_rng2[1] + 0.5) + 1.0,
               label = sprintf("Overall SMA  β = %.3f", sma_slp),
               colour = "firebrick", size = 4, hjust = 0) +
      labs(
        title    = expression(paste("Metabolic scaling — SMA: ", italic(B), " ∝ ", italic(M)^{beta})),
        subtitle = "Red = overall SMA line; dashed = WBE reference β = 0.75",
        x        = expression(log[10] ~ "Body mass (g)"),
        y        = expression(log[10] ~ "Metabolic rate (W)")
      ) +
      theme_bw(base_size = 13) +
      theme(legend.position = "bottom")
  }
}
```

```{r met-sma-table}
if (HAS_SMATR && nrow(met_sources) >= 10) {
  reg_sma_all <- fit_sma(met_sources, "body_mass_g", "metabolic_rate_value", min_n = 10)
  if (!is.null(reg_sma_all) && nrow(reg_sma_all) > 0) {
    kable(reg_sma_all,
          col.names = c("Group", "n", "Method", "log₁₀ Intercept", "Slope (β)",
                        "95% CI lo", "95% CI hi", "R²"),
          caption = paste(
            "SMA log-log regression of metabolic rate (W) on body mass (g) — all taxa combined.",
            "SMA (standardised major axis) corrects for measurement error in both axes;",
            "SMA slopes are ≥ OLS slopes when R² < 1.",
            "WBE predicted slope: β = 0.75."
          )) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
      row_spec(1, bold = TRUE, background = "#ffe8dc")
  }
}
```

## Basal and resting metabolic rates only {#met-basal}

```{r met-basal-analysis, fig.width=10, fig.height=6}
if (nrow(met_basal) >= 10) {
  cat(sprintf("Basal + resting metabolic data: n = %d\n", nrow(met_basal)))
  cat("By dataset:\n")
  print(met_basal[, .N, by = dataset_label])

  reg_basal <- fit_log_lm(met_basal, "body_mass_g", "metabolic_rate_value",
                           group_col = "input_taxonomic_group", min_n = 10)

  fit_bov <- lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data = met_basal)
  int_bov <- coef(fit_bov)[1]
  slp_bov <- coef(fit_bov)[2]
  xrb     <- range(log10(met_basal$body_mass_g), na.rm = TRUE)

  p_basal <- ggplot(met_basal,
         aes(x = log10(body_mass_g), y = log10(metabolic_rate_value),
             colour = input_taxonomic_group)) +
    geom_point(alpha = 0.4, size = 1.2, shape = 16) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8,
                aes(group = input_taxonomic_group)) +
    geom_abline(intercept = int_bov, slope = slp_bov,
                colour = "black", linewidth = 1.2) +
    geom_abline(intercept = int_bov, slope = 0.75,
                linetype = "dashed", colour = "grey30", linewidth = 1.0) +
    scale_colour_viridis_d(option = "turbo", name = "Taxonomic group") +
    labs(
      title    = "Basal and resting metabolic rate scaling",
      subtitle = sprintf("Solid black = overall OLS β = %.3f; dashed = WBE reference β = 0.75",
                         slp_bov),
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
  print(p_basal)

  kable(reg_basal,
        col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = "OLS log-log regression — basal and resting metabolic rates only. Reduced type heterogeneity yields more interpretable slopes.") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    row_spec(1, bold = TRUE, background = "#dce8ff")
} else {
  cat("Fewer than 10 records classified as basal or resting; basal-only analysis skipped.\n")
}
```

---

# Metabolic Scaling by Study

Per-dataset regressions allow direct comparison with each paper's published estimates and avoid conflating temperature-normalization differences.

```{r met-by-study-plot, fig.width=11, fig.height=8}
if (nrow(met_sources) >= 10) {
  ggplot(met_sources,
         aes(x = log10(body_mass_g), y = log10(metabolic_rate_value),
             colour = input_taxonomic_group)) +
    geom_point(alpha = 0.25, size = 0.8, shape = 16) +
    geom_smooth(method = "lm", se = TRUE, colour = "black", fill = "grey70",
                linewidth = 1.0, aes(group = dataset_label)) +
    geom_abline(
      data = met_sources[!is.na(body_mass_g) & body_mass_g > 0 &
                          !is.na(metabolic_rate_value) & metabolic_rate_value > 0,
                         .(ref_int = mean(log10(metabolic_rate_value), na.rm = TRUE) -
                                     0.75 * mean(log10(body_mass_g), na.rm = TRUE)),
                         by = dataset_label],
      aes(intercept = ref_int, slope = 0.75),
      colour = "firebrick", linetype = "dashed", linewidth = 0.9,
      inherit.aes = FALSE) +
    scale_colour_viridis_d(option = "turbo", name = "Taxonomic group") +
    facet_wrap(~ dataset_label, scales = "free", ncol = 2) +
    labs(
      title    = "Metabolic scaling by dataset",
      subtitle = "Black = per-dataset OLS (one line per dataset) with 95% CI; red dashed = WBE β = 0.75 (anchored at per-dataset mean)",
      caption  = "WBE reference line intercept is anchored at each dataset's empirical mean — vertical position reflects data, not theory.",
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r met-by-study-table}
reg_by_ds <- rbindlist(lapply(c(LABEL_AT, LABEL_PN, LABEL_H19), function(lbl) {
  sub <- met_sources[dataset_label == lbl]
  if (nrow(sub) < 10) return(NULL)
  res <- fit_log_lm(sub, "body_mass_g", "metabolic_rate_value", min_n = 10)
  if (!is.null(res) && nrow(res) > 0) res[, Group := lbl]
  res
}), fill = TRUE)

if (!is.null(reg_by_ds) && nrow(reg_by_ds) > 0) {
  kable(reg_by_ds,
        col.names = c("Dataset", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "Per-dataset OLS log-log regressions of metabolic rate (W) on body mass (g).",
          "All metabolic rate types combined within each dataset.",
          "Datasets differ in temperature normalization, taxonomic scope, and rate type distribution.",
          "WBE predicted slope: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r met-by-study-sma}
if (HAS_SMATR) {
  sma_by_ds <- rbindlist(lapply(c(LABEL_AT, LABEL_PN, LABEL_H19), function(lbl) {
    sub <- met_sources[dataset_label == lbl]
    if (nrow(sub) < 10) return(NULL)
    res <- fit_sma(sub, "body_mass_g", "metabolic_rate_value", min_n = 10)
    if (!is.null(res) && nrow(res) > 0) res[, Group := lbl]
    res
  }), fill = TRUE)

  if (!is.null(sma_by_ds) && nrow(sma_by_ds) > 0) {
    kable(sma_by_ds,
          col.names = c("Dataset", "n", "Method", "log₁₀ Intercept", "Slope (β)",
                        "95% CI lo", "95% CI hi", "R²"),
          caption = "SMA log-log regressions of metabolic rate (W) on body mass (g) per dataset.") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
  }
}
```

---

# Metabolic Scaling by Taxonomic Group

Grouping by `input_taxonomic_group` across all datasets reveals whether the slope differs among major clades and life forms.

```{r met-by-group-plot, fig.width=11, fig.height=9}
if (nrow(met_sources) >= 10) {
  grp_counts <- met_sources[!is.na(input_taxonomic_group) & input_taxonomic_group != "",
                             .N, by = input_taxonomic_group]
  eligible_grps <- grp_counts[N >= 10]$input_taxonomic_group

  met_grp <- met_sources[input_taxonomic_group %in% eligible_grps]

  ggplot(met_grp,
         aes(x = log10(body_mass_g), y = log10(metabolic_rate_value),
             colour = dataset_label)) +
    geom_point(alpha = 0.25, size = 0.8, shape = 16) +
    geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 1.0,
                aes(group = input_taxonomic_group)) +
    facet_wrap(~ input_taxonomic_group, scales = "free", ncol = 3) +
    scale_colour_manual(values = pal8[1:4], name = "Dataset") +
    labs(
      title = "Metabolic scaling by taxonomic group (across all datasets)",
      subtitle = "Black line = per-group OLS; groups with N < 10 excluded",
      x = expression(log[10] ~ "Body mass (g)"),
      y = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom",
          strip.text = element_text(size = 9))
}
```

```{r met-by-group-table}
reg_by_grp <- fit_log_lm(met_sources, "body_mass_g", "metabolic_rate_value",
                          group_col = "input_taxonomic_group", min_n = 10)
reg_by_grp_only <- reg_by_grp[Group != "All taxa combined"]

if (!is.null(reg_by_grp_only) && nrow(reg_by_grp_only) > 0) {
  kable(reg_by_grp_only[order(-n)],
        col.names = c("Taxonomic group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regressions of metabolic rate (W) on body mass (g) by taxonomic group.",
          "Groups with N < 10 are excluded.",
          "Taxonomic group labels are verbatim from source; cross-dataset harmonization was not performed.",
          "WBE predicted slope: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r met-slope-comparison-plot, fig.width=9, fig.height=5}
if (!is.null(reg_by_grp_only) && nrow(reg_by_grp_only) > 0 &&
    all(c("Slope", "CI_95_lo", "CI_95_hi") %in% names(reg_by_grp_only))) {
  ggplot(reg_by_grp_only[!is.na(Slope)],
         aes(x = reorder(Group, Slope), y = Slope,
             ymin = CI_95_lo, ymax = CI_95_hi)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick", linewidth = 0.9) +
    geom_errorbar(width = 0.25, colour = "grey50") +
    geom_point(size = 3, colour = pal8[1]) +
    coord_flip() +
    labs(
      title    = "Metabolic scaling slope by taxonomic group (OLS)",
      subtitle = "Error bars = 95% CI; dashed = WBE prediction (β = 0.75)",
      x        = NULL,
      y        = "OLS slope (β)"
    ) +
    theme_bw(base_size = 12)
}
```

---

# Metabolic Scaling Within Order

Within each order, OLS is fit across all records to estimate how metabolic rate scales with body mass among members of that order. This analysis uses records from datasets where the `order` column is populated (primarily Hatton 2019 and AnimalTraits).

```{r met-within-order-prep}
order_met <- met_sources[!is.na(order) & order != "" &
                          !is.na(body_mass_g) & body_mass_g > 0 &
                          !is.na(metabolic_rate_value) & metabolic_rate_value > 0]

cat(sprintf("Records with non-blank order and metabolic rate: n = %d\n", nrow(order_met)))
cat(sprintf("Number of orders represented: %d\n", length(unique(order_met$order))))
cat("\nOrders with N >= 10 records:\n")
print(order_met[, .N, by = order][N >= 10][order(-N)])
```

```{r met-within-order-table}
reg_order_all <- fit_log_lm(order_met, "body_mass_g", "metabolic_rate_value",
                             group_col = "order", min_n = 10)
reg_order_raw <- reg_order_all[Group != "All taxa combined"]
n_ord_excl <- sum(abs(reg_order_raw$Slope) > SLOPE_MAX, na.rm = TRUE)
if (n_ord_excl > 0)
  cat(sprintf("Excluded %d order(s) with |β| > %g (biologically implausible; near-zero mass variance):\n",
              n_ord_excl, SLOPE_MAX),
      paste(reg_order_raw[abs(Slope) > SLOPE_MAX]$Group, collapse = ", "), "\n")
reg_order <- reg_order_raw[is.na(Slope) | abs(Slope) <= SLOPE_MAX]

if (!is.null(reg_order) && nrow(reg_order) > 0) {
  kable(reg_order[order(-n)],
        col.names = c("Order", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regressions of metabolic rate (W) on body mass (g) within each order.",
          "Orders with fewer than 10 records excluded.",
          "Slopes describe how metabolic rate scales with mass across species (and repeated measures) within each order.",
          "OLS is not phylogenetically corrected; slopes are descriptive, not PCM estimates.",
          "WBE predicted slope: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("No orders met the minimum sample size threshold (N >= 10).\n")
}
```

```{r met-within-order-plot, fig.width=9, fig.height=5}
if (!is.null(reg_order) && nrow(reg_order) > 0) {
  ggplot(reg_order[!is.na(Slope)],
         aes(x = reorder(Group, Slope), y = Slope,
             ymin = CI_95_lo, ymax = CI_95_hi, colour = R2)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick", linewidth = 0.9) +
    geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.5) +
    geom_errorbar(width = 0.3) +
    geom_point(size = 2.5) +
    scale_colour_gradient(low = "grey60", high = pal8[3], name = "R²") +
    coord_flip() +
    labs(
      title    = "Within-order metabolic scaling slopes (OLS)",
      subtitle = "Error bars = 95% CI; dashed = WBE β = 0.75",
      x        = NULL,
      y        = "OLS slope (β)"
    ) +
    theme_bw(base_size = 11)
}
```

## Full slope frequency distribution — within order {#order-slope-dist}

```{r met-within-order-histogram, fig.width=10, fig.height=5}
if (!is.null(reg_order) && nrow(reg_order) >= 5 && sum(!is.na(reg_order$Slope)) >= 5) {
  med_ord <- median(reg_order$Slope, na.rm = TRUE)
  ggplot(reg_order[!is.na(Slope)], aes(x = Slope)) +
    geom_histogram(binwidth = 0.1, fill = pal8[3], colour = "white", alpha = 0.85) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "firebrick", linewidth = 1.1) +
    geom_vline(xintercept = med_ord, linetype = "solid", colour = pal8[1], linewidth = 1.0) +
    annotate("text", x = 0.75 + 0.05, y = Inf, label = "WBE β = 0.75",
             colour = "firebrick", vjust = 1.5, hjust = 0, size = 3.5) +
    annotate("text", x = med_ord - 0.05, y = Inf,
             label = sprintf("Median β = %.3f", med_ord),
             colour = pal8[1], vjust = 3.2, hjust = 1, size = 3.5) +
    labs(
      title    = "Frequency distribution of within-order metabolic scaling slopes",
      subtitle = sprintf("N = %d orders with ≥ 10 records; dashed = WBE β = 0.75; solid = observed median",
                         sum(!is.na(reg_order$Slope))),
      x        = "OLS slope (β) within order",
      y        = "Number of orders"
    ) +
    theme_bw(base_size = 13)
}
```

## Slope distribution by major clade — within order {#order-slope-by-clade}

```{r met-within-order-clade-prep}
# Attach the dominant taxonomic group to each order using the raw records
order_clade <- order_met[!is.na(order) & order != "" &
                          !is.na(input_taxonomic_group) & input_taxonomic_group != "",
                         .(clade = {
                             tb <- sort(table(input_taxonomic_group), decreasing = TRUE)
                             names(tb)[1]
                           }),
                         by = order]

reg_order_clade <- merge(reg_order, order_clade, by.x = "Group", by.y = "order", all.x = TRUE)

# Keep only clades with >= 3 orders so boxplots are meaningful
clade_order_counts <- reg_order_clade[!is.na(clade) & !is.na(Slope),
                                      .N, by = clade][N >= 3]
rod_plot <- reg_order_clade[clade %in% clade_order_counts$clade & !is.na(Slope)]
cat("Orders available for clade breakdown:", nrow(rod_plot), "\n")
print(rod_plot[, .N, by = clade][order(-N)])
```

```{r met-within-order-clade-boxplot, fig.width=11, fig.height=6}
if (nrow(rod_plot) >= 5) {
  ggplot(rod_plot, aes(x = reorder(clade, Slope, FUN = median), y = Slope,
                        fill = clade)) +
    geom_boxplot(alpha = 0.75, outlier.shape = 16, outlier.size = 1.8) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_jitter(width = 0.18, size = 1.5, alpha = 0.6, colour = "grey30") +
    scale_fill_brewer(palette = "Dark2", guide = "none") +
    coord_flip() +
    labs(
      title    = "Within-order metabolic scaling slopes by major clade",
      subtitle = "Each point = one order; box = median ± IQR; dashed = WBE β = 0.75",
      x        = "Major clade (dominant taxonomic group in order)",
      y        = "OLS slope (β) within order"
    ) +
    theme_bw(base_size = 13)
}
```

```{r met-within-order-clade-histogram, fig.width=12, fig.height=7}
if (nrow(rod_plot) >= 5) {
  ggplot(rod_plot, aes(x = Slope, fill = clade)) +
    geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.85) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 0.9) +
    scale_fill_brewer(palette = "Dark2", guide = "none") +
    facet_wrap(~ clade, scales = "free_y", ncol = 3) +
    labs(
      title    = "Frequency distribution of within-order slopes by major clade",
      subtitle = "Dashed = WBE β = 0.75; each panel = one major clade",
      x        = "OLS slope (β) within order",
      y        = "Number of orders"
    ) +
    theme_bw(base_size = 12)
}
```

---

# Metabolic Scaling Within Family

Within each family (N ≥ 10 records), OLS is fit across all records to estimate how metabolic rate scales with body mass among members of that family. This is primarily enabled by AnimalTraits (~372 families) and Hatton 2019.

> **Interpretation note:** Family-level slopes describe the relationship across species (and repeated measurements) within a family. When body mass variation within a family is limited, slope estimates become unreliable regardless of N. R² values can help diagnose this: low R² at small mass ranges suggests poor leverage.

```{r met-within-family-prep}
fam_met <- met_sources[!is.na(family) & family != "" &
                        !is.na(body_mass_g) & body_mass_g > 0 &
                        !is.na(metabolic_rate_value) & metabolic_rate_value > 0]

cat(sprintf("Records with non-blank family and metabolic rate: n = %d\n", nrow(fam_met)))
cat(sprintf("Number of families represented: %d\n", length(unique(fam_met$family))))

fam_counts <- fam_met[, .N, by = family][N >= 10][order(-N)]
cat(sprintf("Families with N >= 10 records: %d\n", nrow(fam_counts)))
print(head(fam_counts, 20))
```

```{r met-within-family-table}
reg_fam_all <- fit_log_lm(fam_met, "body_mass_g", "metabolic_rate_value",
                           group_col = "family", min_n = 10)
reg_fam_raw <- reg_fam_all[Group != "All taxa combined"]
n_fam_excl <- sum(abs(reg_fam_raw$Slope) > SLOPE_MAX, na.rm = TRUE)
if (n_fam_excl > 0)
  cat(sprintf("Excluded %d family/families with |β| > %g (biologically implausible; near-zero mass variance):\n",
              n_fam_excl, SLOPE_MAX),
      paste(reg_fam_raw[abs(Slope) > SLOPE_MAX]$Group, collapse = ", "), "\n")
reg_fam <- reg_fam_raw[is.na(Slope) | abs(Slope) <= SLOPE_MAX]

if (!is.null(reg_fam) && nrow(reg_fam) > 0) {
  kable(reg_fam[order(-n)],
        col.names = c("Family", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regressions of metabolic rate (W) on body mass (g) within each family.",
          "Families with N < 10 records excluded.",
          "WBE predicted slope: β = 0.75.",
          "OLS is not phylogenetically corrected."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("No families met the minimum sample size threshold (N >= 10).\n")
}
```

```{r met-within-family-histogram, fig.width=10, fig.height=5}
if (!is.null(reg_fam) && nrow(reg_fam) > 0 && sum(!is.na(reg_fam$Slope)) > 5) {
  med_slope <- median(reg_fam$Slope, na.rm = TRUE)
  ggplot(reg_fam[!is.na(Slope)], aes(x = Slope)) +
    geom_histogram(binwidth = 0.1, fill = pal8[2], colour = "white", alpha = 0.85) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_vline(xintercept = med_slope, linetype = "solid", colour = pal8[1],
               linewidth = 1.0) +
    annotate("text", x = 0.75 + 0.05, y = Inf, label = "WBE β = 0.75",
             colour = "firebrick", vjust = 1.5, hjust = 0, size = 3.5) +
    annotate("text", x = med_slope - 0.05, y = Inf,
             label = sprintf("Median β = %.3f", med_slope),
             colour = pal8[1], vjust = 3, hjust = 1, size = 3.5) +
    labs(
      title    = "Distribution of family-level metabolic scaling slopes (OLS)",
      subtitle = sprintf("N = %d families with ≥ 10 records; dashed = WBE 0.75; solid = observed median",
                         sum(!is.na(reg_fam$Slope))),
      x        = "OLS slope (β)",
      y        = "Number of families"
    ) +
    theme_bw(base_size = 13)
}
```

```{r met-within-family-boxplot, fig.width=6, fig.height=5}
if (!is.null(reg_fam) && nrow(reg_fam) > 0 && sum(!is.na(reg_fam$Slope)) > 5) {
  ggplot(reg_fam[!is.na(Slope)], aes(x = "Family-level OLS", y = Slope)) +
    geom_boxplot(fill = pal8[2], alpha = 0.7, outlier.shape = 1) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    labs(
      title = "Family-level slope distribution",
      x     = NULL,
      y     = "OLS slope (β)"
    ) +
    annotate("text", x = 1.45, y = 0.75, label = "WBE β = 0.75",
             colour = "firebrick", hjust = 0, size = 3.5) +
    theme_bw(base_size = 13)
}
```

## Slope distribution by major clade — within family {#family-slope-by-clade}

```{r met-within-family-clade-prep}
# Attach dominant clade to each family
fam_clade <- fam_met[!is.na(family) & family != "" &
                      !is.na(input_taxonomic_group) & input_taxonomic_group != "",
                     .(clade = {
                         tb <- sort(table(input_taxonomic_group), decreasing = TRUE)
                         names(tb)[1]
                       }),
                     by = family]

reg_fam_clade <- merge(reg_fam, fam_clade, by.x = "Group", by.y = "family", all.x = TRUE)

# Keep clades with >= 3 families
clade_fam_counts <- reg_fam_clade[!is.na(clade) & !is.na(Slope), .N, by = clade][N >= 3]
rfp_plot <- reg_fam_clade[clade %in% clade_fam_counts$clade & !is.na(Slope)]
cat("Families available for clade breakdown:", nrow(rfp_plot), "\n")
print(rfp_plot[, .N, by = clade][order(-N)])
```

```{r met-within-family-clade-boxplot, fig.width=11, fig.height=6}
if (nrow(rfp_plot) >= 5) {
  ggplot(rfp_plot, aes(x = reorder(clade, Slope, FUN = median), y = Slope,
                        fill = clade)) +
    geom_boxplot(alpha = 0.75, outlier.shape = 16, outlier.size = 1.5) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_jitter(width = 0.18, size = 1.2, alpha = 0.5, colour = "grey30") +
    scale_fill_brewer(palette = "Dark2", guide = "none") +
    coord_flip() +
    labs(
      title    = "Within-family metabolic scaling slopes by major clade",
      subtitle = "Each point = one family; box = median ± IQR; dashed = WBE β = 0.75",
      x        = "Major clade (dominant taxonomic group in family)",
      y        = "OLS slope (β) within family"
    ) +
    theme_bw(base_size = 13)
}
```

```{r met-within-family-clade-histogram, fig.width=13, fig.height=8}
if (nrow(rfp_plot) >= 5) {
  ggplot(rfp_plot, aes(x = Slope, fill = clade)) +
    geom_histogram(binwidth = 0.1, colour = "white", alpha = 0.85) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 0.9) +
    scale_fill_brewer(palette = "Dark2", guide = "none") +
    facet_wrap(~ clade, scales = "free_y", ncol = 3) +
    labs(
      title    = "Frequency distribution of within-family slopes by major clade",
      subtitle = "Dashed = WBE β = 0.75; each panel = one major clade",
      x        = "OLS slope (β) within family",
      y        = "Number of families"
    ) +
    theme_bw(base_size = 12)
}
```

---

# Insect Metabolic Scaling (Chown 2007, corrected)

The Chown et al. (2007) dataset provides a focused, flow-through respirometry compilation of 346 insect species across 16 orders, supplemented with unit-conversion corrections from Riveros & Enquist (2011) for three Coleoptera. This section analyses insect metabolic scaling as a stand-alone dataset and in relation to the broader cross-taxa compilation.

> **Key predictions:** WBE theory predicts β = 0.75 for whole-organism metabolic rate. Riveros & Enquist (2011) reported that insects broadly support β = 0.74 (95% CI: 0.71–0.77, n = 367) after excluding Termitidae. Wing status (winged vs. wingless) significantly affects the intercept but not the slope.

```{r insect-prep}
ins <- qa_ok(ch07)
ins <- ins[!is.na(body_mass_g) & body_mass_g > 0 &
           !is.na(metabolic_rate_value) & metabolic_rate_value > 0 &
           metabolic_rate_unit == "W"]

cat(sprintf("Insect records (QA-pass, unit = W): n = %d\n", nrow(ins)))
cat(sprintf("Orders:   %d\n", length(unique(ins$order))))
cat(sprintf("Families: %d\n", length(unique(ins$family[ins$family != ""]))))
cat(sprintf("Winged (1): %d   Wingless (0): %d\n",
            sum(ins$wing_status == 1, na.rm = TRUE),
            sum(ins$wing_status == 0, na.rm = TRUE)))
cat("\nRecords by order:\n")
print(ins[, .N, by = order][order(-N)])
```

## Overall insect metabolic scaling {#insect-overall}

```{r insect-overall-plot, fig.width=9, fig.height=6}
if (nrow(ins) >= 10) {
  fit_ins <- lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data = ins)
  slp_ins <- coef(fit_ins)[2]
  int_ins <- coef(fit_ins)[1]
  r2_ins  <- summary(fit_ins)$r.squared
  x_rng_ins <- range(log10(ins$body_mass_g), na.rm = TRUE)
  ref_ins   <- data.frame(lx = x_rng_ins, ly = int_ins + 0.75 * x_rng_ins)
  ols_ins   <- data.frame(lx = x_rng_ins, ly = int_ins + slp_ins * x_rng_ins)

  ggplot(ins, aes(x = log10(body_mass_g), y = log10(metabolic_rate_value),
                  colour = order)) +
    geom_point(alpha = 0.55, size = 1.5, shape = 16) +
    geom_line(data = ols_ins, aes(x = lx, y = ly),
              inherit.aes = FALSE, colour = "black", linewidth = 1.3) +
    geom_line(data = ref_ins, aes(x = lx, y = ly),
              inherit.aes = FALSE, linetype = "dashed", colour = "firebrick",
              linewidth = 0.9) +
    scale_colour_brewer(palette = "Set3", name = "Order") +
    annotate("text", x = x_rng_ins[1] + 0.3,
             y = int_ins + slp_ins * (x_rng_ins[1] + 0.3) + 1.2,
             label = sprintf("OLS  β = %.3f  R² = %.2f", slp_ins, r2_ins),
             size = 4, hjust = 0) +
    annotate("text", x = x_rng_ins[2] - 0.3,
             y = int_ins + 0.75 * (x_rng_ins[2] - 0.3) - 1.0,
             label = "WBE β = 0.75", colour = "firebrick", size = 3.5, hjust = 1) +
    labs(
      title    = expression(paste("Insect metabolic scaling: ", italic(B), " ∝ ", italic(M)^{beta})),
      subtitle = "Chown et al. 2007 (corrected per Riveros & Enquist 2011). Black = OLS; dashed red = WBE β = 0.75.",
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "right")
}
```

```{r insect-overall-table}
reg_ins_overall <- fit_log_lm(ins, "body_mass_g", "metabolic_rate_value", min_n = 10)
if (!is.null(reg_ins_overall) && nrow(reg_ins_overall) > 0) {
  kable(reg_ins_overall,
        col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regression of metabolic rate (W) on body mass (g) for all insects combined.",
          "WBE predicted slope: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
}
```

```{r insect-sma}
if (HAS_SMATR) {
  reg_ins_sma <- fit_sma(ins, "body_mass_g", "metabolic_rate_value", min_n = 10)
  if (!is.null(reg_ins_sma) && nrow(reg_ins_sma) > 0) {
    kable(reg_ins_sma,
          col.names = c("Group", "n", "Method", "log₁₀ Intercept", "SMA Slope (β)",
                        "95% CI lo", "95% CI hi", "R²"),
          caption = "SMA regression of metabolic rate (W) on body mass (g) — all insects combined.") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
  }
}
```

## Scaling by insect order {#insect-by-order}

```{r insect-order-table}
reg_ins_ord_raw <- fit_log_lm(ins, "body_mass_g", "metabolic_rate_value",
                               group_col = "order", min_n = 10)
reg_ins_ord_raw <- reg_ins_ord_raw[Group != "All taxa combined"]
n_ins_ord_excl <- sum(abs(reg_ins_ord_raw$Slope) > SLOPE_MAX, na.rm = TRUE)
if (n_ins_ord_excl > 0)
  cat(sprintf("Excluded %d order(s) with |β| > %g:\n", n_ins_ord_excl, SLOPE_MAX),
      paste(reg_ins_ord_raw[abs(Slope) > SLOPE_MAX]$Group, collapse = ", "), "\n")
reg_ins_ord <- reg_ins_ord_raw[is.na(Slope) | abs(Slope) <= SLOPE_MAX]

if (!is.null(reg_ins_ord) && nrow(reg_ins_ord) > 0) {
  kable(reg_ins_ord[order(-n)],
        col.names = c("Order", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS within-order metabolic scaling for insects (N ≥ 10 per order).",
          "WBE predicted slope: β = 0.75.",
          "Isoptera (termites) are expected to deviate toward β ≈ 1 (see Riveros & Enquist 2011)."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r insect-order-plot, fig.width=9, fig.height=6}
if (!is.null(reg_ins_ord) && nrow(reg_ins_ord) > 0) {
  ggplot(reg_ins_ord[!is.na(Slope)],
         aes(x = reorder(Group, Slope), y = Slope,
             ymin = CI_95_lo, ymax = CI_95_hi,
             colour = Group == "Isoptera")) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick", linewidth = 1.0) +
    geom_hline(yintercept = 0,    colour = "grey70",   linewidth = 0.5) +
    geom_errorbar(width = 0.3) +
    geom_point(size = 3) +
    scale_colour_manual(values = c("FALSE" = pal8[1], "TRUE" = "darkorange2"),
                        labels = c("FALSE" = "Other orders", "TRUE" = "Isoptera (termites)"),
                        name = NULL) +
    coord_flip() +
    labs(
      title    = "Within-order insect metabolic scaling exponents (OLS)",
      subtitle = "Dashed = WBE β = 0.75. Orange = Isoptera (Termitidae deviation noted by Riveros & Enquist 2011).",
      x        = NULL, y = "OLS slope (β)"
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

```{r insect-order-scatter, fig.width=12, fig.height=9}
ord_n10 <- ins[order %in% reg_ins_ord$Group]
if (nrow(ord_n10) > 0) {
  ggplot(ord_n10, aes(x = log10(body_mass_g), y = log10(metabolic_rate_value))) +
    geom_point(aes(colour = order), alpha = 0.6, size = 1.4, shape = 16) +
    geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.9,
                fill = "grey85", alpha = 0.4) +
    facet_wrap(~ order, scales = "free", ncol = 3) +
    scale_colour_brewer(palette = "Dark2", guide = "none") +
    labs(
      title    = "Insect metabolic scaling within each order",
      subtitle = "Black line = OLS fit with 95% CI ribbon; orders with N ≥ 10 shown.",
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 11) +
    theme(strip.text = element_text(face = "bold", size = 10))
}
```

## Wing status effect {#insect-wing}

Wing status (0 = wingless / workers / secondary wingless; 1 = winged) is recorded in the Chown dataset. Winged insects generally have larger flight muscle mass and higher mass-specific metabolic rates, so winged individuals are expected to show higher metabolic rates at the same body mass (intercept shift) while the scaling exponent may be conserved.

```{r insect-wing-prep}
ins_wing <- ins[!is.na(wing_status)]
ins_wing[, wing_label := ifelse(wing_status == 1, "Winged (1)", "Wingless (0)")]
cat(sprintf("Records with wing status: n = %d\n", nrow(ins_wing)))
print(ins_wing[, .N, by = wing_label])
```

```{r insect-wing-plot, fig.width=9, fig.height=6}
if (nrow(ins_wing) >= 20) {
  ggplot(ins_wing, aes(x = log10(body_mass_g), y = log10(metabolic_rate_value),
                       colour = wing_label)) +
    geom_point(alpha = 0.5, size = 1.5, shape = 16) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 1.1,
                aes(group = wing_label)) +
    scale_colour_manual(values = c("Winged (1)" = pal8[2], "Wingless (0)" = pal8[4]),
                        name = "Wing status") +
    labs(
      title    = "Insect metabolic scaling by wing status",
      subtitle = "Wing status 1 = winged; 0 = wingless/worker. Lines = per-group OLS.",
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

```{r insect-wing-table}
reg_ins_wing <- fit_log_lm(ins_wing, "body_mass_g", "metabolic_rate_value",
                            group_col = "wing_label", min_n = 10)
if (!is.null(reg_ins_wing) && nrow(reg_ins_wing) > 0) {
  kable(reg_ins_wing,
        col.names = c("Wing status", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS regressions by wing status.",
          "Intercept differences reflect mass-specific metabolic rate offsets between winged and wingless insects.",
          "WBE predicted slope: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
}
```

## Termitidae deviation {#insect-termitidae}

Termitidae (true termites) were identified by Riveros & Enquist (2011) as the sole significant outlier from WBE metabolic scaling among insects. Their metabolic rates are lower than predicted at a given mass, and the order-level (Isoptera) slope approaches isometry (β ≈ 1). Two potential explanations are: (1) CH₄-based respirometry used for many termite measurements may systematically underestimate CO₂-equivalent metabolic rates; (2) colonial living and eusociality impose metabolic suppression on workers measured in isolation.

```{r insect-termitidae}
ins_termit <- ins[family == "Termitidae"]
ins_other  <- ins[family != "Termitidae" | family == ""]

cat(sprintf("Termitidae: n = %d\n",    nrow(ins_termit)))
cat(sprintf("Non-Termitidae: n = %d\n", nrow(ins_other)))

if (nrow(ins_termit) >= 5 && nrow(ins_other) >= 10) {
  fit_t  <- lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data = ins_termit)
  fit_nt <- lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data = ins_other)

  termit_comp <- data.table(
    Group       = c("Termitidae only", "All insects excl. Termitidae", "All insects"),
    n           = c(nrow(ins_termit), nrow(ins_other), nrow(ins)),
    Intercept   = round(c(coef(fit_t)[1], coef(fit_nt)[1],
                          coef(lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data=ins))[1]), 3),
    Slope_beta  = round(c(coef(fit_t)[2], coef(fit_nt)[2],
                          coef(lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data=ins))[2]), 3),
    R2          = round(c(summary(fit_t)$r.squared,
                          summary(fit_nt)$r.squared,
                          summary(lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data=ins))$r.squared), 3)
  )

  kable(termit_comp,
        col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "R²"),
        caption = paste(
          "Comparison of metabolic scaling slopes with and without Termitidae.",
          "Termitidae show a near-isometric slope (β ≈ 1) and a depressed intercept,",
          "consistent with the Riveros & Enquist (2011) finding that termites are the principal insect outlier from WBE."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
}
```

```{r insect-termitidae-plot, fig.width=9, fig.height=6}
if (nrow(ins_termit) >= 5) {
  ins_plot <- copy(ins)
  ins_plot[, taxon_grp := ifelse(family == "Termitidae", "Termitidae",
                          ifelse(order == "Hymenoptera", "Hymenoptera (other)",
                                 "All other insects"))]
  x_rng2 <- range(log10(ins_plot$body_mass_g), na.rm = TRUE)
  ref_df2 <- data.frame(lx = x_rng2,
                        ly = coef(lm(log10(metabolic_rate_value) ~ log10(body_mass_g),
                                     data = ins))[1] + 0.75 * x_rng2)

  ggplot(ins_plot, aes(x = log10(body_mass_g), y = log10(metabolic_rate_value),
                       colour = taxon_grp)) +
    geom_point(alpha = 0.55, size = 1.8, shape = 16) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 1.1,
                aes(group = taxon_grp)) +
    geom_line(data = ref_df2, aes(x = lx, y = ly),
              inherit.aes = FALSE, linetype = "dashed", colour = "firebrick",
              linewidth = 1.0) +
    scale_colour_manual(values = c("Termitidae" = "darkorange2",
                                   "Hymenoptera (other)" = pal8[3],
                                   "All other insects" = "grey60"),
                        name = NULL) +
    labs(
      title    = "Termitidae vs. other insects: metabolic scaling",
      subtitle = "Dashed red = WBE β = 0.75. Termitidae deviate toward near-isometric scaling.",
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Metabolic rate (W)")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

## Insect metabolic scaling within family {#insect-family}

```{r insect-family-table}
reg_ins_fam_raw <- fit_log_lm(ins, "body_mass_g", "metabolic_rate_value",
                               group_col = "family", min_n = 10)
reg_ins_fam_raw <- reg_ins_fam_raw[Group != "All taxa combined"]
n_ins_fam_excl  <- sum(abs(reg_ins_fam_raw$Slope) > SLOPE_MAX, na.rm = TRUE)
if (n_ins_fam_excl > 0)
  cat(sprintf("Excluded %d family/ies with |β| > %g:\n", n_ins_fam_excl, SLOPE_MAX),
      paste(reg_ins_fam_raw[abs(Slope) > SLOPE_MAX]$Group, collapse = ", "), "\n")
reg_ins_fam <- reg_ins_fam_raw[is.na(Slope) | abs(Slope) <= SLOPE_MAX]

if (!is.null(reg_ins_fam) && nrow(reg_ins_fam) > 0) {
  kable(reg_ins_fam[order(-n)],
        col.names = c("Family", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS within-family metabolic scaling for insects (N ≥ 10 per family).",
          "WBE predicted slope: β = 0.75.",
          "Formicidae (ants) and Termitidae (termites) are highlighted in interpretation."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r insect-family-plot, fig.width=9, fig.height=5}
if (!is.null(reg_ins_fam) && nrow(reg_ins_fam) > 0) {
  reg_ins_fam_plot <- copy(reg_ins_fam)
  reg_ins_fam_plot[, highlight := fcase(
    Group == "Termitidae",  "Termitidae",
    Group == "Formicidae",  "Formicidae (ants)",
    default = "Other family"
  )]

  ggplot(reg_ins_fam_plot[!is.na(Slope)],
         aes(x = reorder(Group, Slope), y = Slope,
             ymin = CI_95_lo, ymax = CI_95_hi,
             colour = highlight)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick", linewidth = 1.0) +
    geom_hline(yintercept = 0,    colour = "grey70",   linewidth = 0.5) +
    geom_errorbar(width = 0.3) +
    geom_point(size = 3) +
    scale_colour_manual(values = c("Termitidae"         = "darkorange2",
                                   "Formicidae (ants)"  = pal8[3],
                                   "Other family"       = pal8[1]),
                        name = NULL) +
    coord_flip() +
    labs(
      title    = "Within-family insect metabolic scaling exponents (OLS)",
      subtitle = "Dashed = WBE β = 0.75. Orange = Termitidae; green = Formicidae.",
      x = NULL, y = "OLS slope (β)"
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "bottom")
}
```

---

# Intraspecific Metabolic Scaling

For species with N ≥ 5 metabolic rate records, OLS is fit within each species to estimate the intraspecific slope. This analysis addresses whether metabolic rate scales with body mass within individual species, reflecting ontogenetic or individual variation rather than cross-species allometry.

> **Key caveat:** Most allometric databases report one representative value per species. Intraspecific scaling requires multiple measurements on individuals or populations of the same species spanning a range of body masses. If all observations within a species have identical or nearly identical body masses, the regression slope is unreliable regardless of N. Species where the body mass coefficient of variation (CV) is very low are flagged below.

```{r intraspecific-prep}
# Identify species column: prefer resolved_taxon_name
has_resolved <- "resolved_taxon_name" %in% names(met_sources) &&
  sum(!is.na(met_sources$resolved_taxon_name) & met_sources$resolved_taxon_name != "") > 0
sp_col <- if (has_resolved) "resolved_taxon_name" else "verbatim_taxon_name"

cat("Using species column:", sp_col, "\n")

# Count records per species
sp_counts <- met_sources[
  !is.na(body_mass_g) & body_mass_g > 0 &
  !is.na(metabolic_rate_value) & metabolic_rate_value > 0 &
  !is.na(get(sp_col)) & get(sp_col) != "",
  .(N = .N,
    mass_cv   = sd(body_mass_g, na.rm = TRUE) / mean(body_mass_g, na.rm = TRUE),
    mass_range = diff(range(body_mass_g, na.rm = TRUE))),
  by = .(species = get(sp_col))
]

intra_eligible <- sp_counts[N >= 5]
cat(sprintf("Species with N >= 5 metabolic records: %d\n", nrow(intra_eligible)))
if (nrow(intra_eligible) > 0) {
  cat(sprintf("Body mass CV > 0.01 (meaningful variation): %d\n",
              sum(intra_eligible$mass_cv > 0.01, na.rm = TRUE)))
  print(intra_eligible[order(-N)])
}
```

```{r intraspecific-slopes}
if (nrow(intra_eligible) >= 1) {
  intra_spp <- intra_eligible$species

  intra_slopes <- rbindlist(lapply(intra_spp, function(sp) {
    d <- met_sources[get(sp_col) == sp &
                     !is.na(body_mass_g) & body_mass_g > 0 &
                     !is.na(metabolic_rate_value) & metabolic_rate_value > 0]
    if (nrow(d) < 5) return(NULL)
    fit <- tryCatch(
      lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data = d),
      error = function(e) NULL
    )
    if (is.null(fit)) return(NULL)
    s <- coef(summary(fit))
    if (nrow(s) < 2) return(NULL)
    mass_cv <- sd(d$body_mass_g, na.rm = TRUE) / mean(d$body_mass_g, na.rm = TRUE)
    data.table(
      species   = sp,
      n         = nrow(d),
      slope     = round(s[2, 1], 3),
      slope_se  = round(s[2, 2], 3),
      r2        = round(summary(fit)$r.squared, 3),
      p_value   = signif(s[2, 4], 3),
      mass_cv   = round(mass_cv, 3),
      flag_low_cv = mass_cv < 0.01
    )
  }), fill = TRUE)

  # Remove biologically implausible slopes
  n_intra_excl <- sum(abs(intra_slopes$slope) > SLOPE_MAX, na.rm = TRUE)
  if (n_intra_excl > 0)
    cat(sprintf("Excluded %d species with |β| > %g (biologically implausible; near-zero mass variance):\n",
                n_intra_excl, SLOPE_MAX),
        paste(intra_slopes[abs(slope) > SLOPE_MAX]$species, collapse = ", "), "\n")
  intra_slopes <- intra_slopes[is.na(slope) | abs(slope) <= SLOPE_MAX]

  if (!is.null(intra_slopes) && nrow(intra_slopes) > 0) {
    cat(sprintf("Species with fitted intraspecific slopes: %d\n", nrow(intra_slopes)))
    cat(sprintf("Median slope: %.3f\n", median(intra_slopes$slope, na.rm = TRUE)))
    cat(sprintf("IQR: %.3f – %.3f\n",
                quantile(intra_slopes$slope, 0.25, na.rm = TRUE),
                quantile(intra_slopes$slope, 0.75, na.rm = TRUE)))

    kable(intra_slopes[order(-n)],
          col.names = c("Species", "n", "Slope (β)", "SE", "R²", "p-value",
                        "Mass CV", "Flag: low CV"),
          caption = paste(
            "Intraspecific OLS log-log regressions of metabolic rate (W) on body mass (g).",
            "Species with N >= 5 metabolic records.",
            "Mass CV = coefficient of variation of body mass within species;",
            "Flag low CV indicates CV < 0.01, meaning body mass variation is minimal",
            "and the slope estimate is unreliable.",
            "WBE predicted slope: β = 0.75."
          )) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
  }
} else {
  cat("No species met the N >= 5 threshold for intraspecific analysis.\n",
      "This is expected for databases that store one representative value per species.\n")
  intra_slopes <- data.table()
}
```

```{r intraspecific-histogram, fig.width=10, fig.height=5}
if (exists("intra_slopes") && !is.null(intra_slopes) && nrow(intra_slopes) >= 5) {
  med_intra <- median(intra_slopes$slope, na.rm = TRUE)
  p_hist <- ggplot(intra_slopes[!is.na(slope)], aes(x = slope)) +
    geom_histogram(binwidth = 0.15, fill = pal8[4], colour = "white", alpha = 0.85) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.2) +
    geom_vline(xintercept = med_intra, linetype = "solid", colour = pal8[1],
               linewidth = 1.0) +
    annotate("text", x = 0.75 + 0.06, y = Inf, label = "WBE β = 0.75",
             colour = "firebrick", vjust = 1.5, hjust = 0, size = 3.5) +
    annotate("text", x = med_intra - 0.06, y = Inf,
             label = sprintf("Median β = %.3f", med_intra),
             colour = pal8[1], vjust = 3, hjust = 1, size = 3.5) +
    labs(
      title    = "Distribution of intraspecific metabolic scaling slopes (OLS)",
      subtitle = sprintf("N = %d species with ≥ 5 records; dashed = WBE 0.75",
                         sum(!is.na(intra_slopes$slope))),
      x        = "OLS slope (β) per species",
      y        = "Number of species"
    ) +
    theme_bw(base_size = 13)
  print(p_hist)

  p_box <- ggplot(intra_slopes[!is.na(slope)], aes(x = "Intraspecific slopes", y = slope)) +
    geom_boxplot(fill = pal8[4], alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    labs(
      title = "Intraspecific slope distribution",
      x     = NULL,
      y     = "OLS slope (β)"
    ) +
    annotate("text", x = 1.45, y = 0.75, label = "WBE β = 0.75",
             colour = "firebrick", hjust = 0, size = 3.5) +
    theme_bw(base_size = 13)
  print(p_box)

  # Flag outliers
  q1 <- quantile(intra_slopes$slope, 0.25, na.rm = TRUE)
  q3 <- quantile(intra_slopes$slope, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  outliers <- intra_slopes[!is.na(slope) & (slope < q1 - 1.5 * iqr | slope > q3 + 1.5 * iqr)]
  if (nrow(outliers) > 0) {
    cat(sprintf("\nOutlier species by IQR rule (slope < %.2f or > %.2f):\n",
                q1 - 1.5 * iqr, q3 + 1.5 * iqr))
    print(outliers[, .(species, n, slope, r2, mass_cv)])
  }
} else if (!exists("intra_slopes") || nrow(intra_slopes) < 5) {
  cat("Fewer than 5 species with intraspecific slopes; distribution plots skipped.\n")
}
```

## Intraspecific slopes by major clade {#intraspecific-by-clade}

```{r intraspecific-clade-prep}
if (exists("intra_slopes") && !is.null(intra_slopes) && nrow(intra_slopes) >= 5) {
  # Attach dominant clade to each species from met_sources
  sp_clade <- met_sources[!is.na(get(sp_col)) & get(sp_col) != "" &
                           !is.na(input_taxonomic_group) & input_taxonomic_group != "",
                          .(clade = {
                              tb <- sort(table(input_taxonomic_group), decreasing = TRUE)
                              names(tb)[1]
                            }),
                          by = .(species = get(sp_col))]

  intra_slopes_clade <- merge(intra_slopes, sp_clade, by = "species", all.x = TRUE)

  clade_sp_counts <- intra_slopes_clade[!is.na(clade) & !is.na(slope),
                                         .N, by = clade][N >= 3]
  isc_plot <- intra_slopes_clade[clade %in% clade_sp_counts$clade & !is.na(slope)]
  cat("Species available for clade breakdown:", nrow(isc_plot), "\n")
  print(isc_plot[, .N, by = clade][order(-N)])
} else {
  isc_plot <- data.table()
  cat("Insufficient intraspecific data for clade breakdown.\n")
}
```

```{r intraspecific-clade-boxplot, fig.width=11, fig.height=6}
if (nrow(isc_plot) >= 5) {
  ggplot(isc_plot, aes(x = reorder(clade, slope, FUN = median), y = slope,
                        fill = clade)) +
    geom_boxplot(alpha = 0.75, outlier.shape = 16, outlier.size = 1.8) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_jitter(width = 0.18, size = 1.5, alpha = 0.6, colour = "grey30") +
    scale_fill_brewer(palette = "Dark2", guide = "none") +
    coord_flip() +
    labs(
      title    = "Intraspecific metabolic scaling slopes by major clade",
      subtitle = "Each point = one species (N ≥ 5); box = median ± IQR; dashed = WBE β = 0.75",
      x        = "Major clade",
      y        = "Intraspecific OLS slope (β)"
    ) +
    theme_bw(base_size = 13)
}
```

```{r intraspecific-clade-histogram, fig.width=13, fig.height=8}
if (nrow(isc_plot) >= 5) {
  ggplot(isc_plot, aes(x = slope, fill = clade)) +
    geom_histogram(binwidth = 0.15, colour = "white", alpha = 0.85) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 0.9) +
    scale_fill_brewer(palette = "Dark2", guide = "none") +
    facet_wrap(~ clade, scales = "free_y", ncol = 3) +
    labs(
      title    = "Frequency distribution of intraspecific slopes by major clade",
      subtitle = "Dashed = WBE β = 0.75; each panel = one major clade; species with N ≥ 5 records",
      x        = "Intraspecific OLS slope (β) per species",
      y        = "Number of species"
    ) +
    theme_bw(base_size = 12)
}
```

```{r intraspecific-clade-summary-table}
if (nrow(isc_plot) >= 5) {
  clade_summ <- isc_plot[!is.na(slope),
    .(N_species     = .N,
      Median_slope  = round(median(slope), 3),
      IQR_lo        = round(quantile(slope, 0.25), 3),
      IQR_hi        = round(quantile(slope, 0.75), 3),
      Mean_slope    = round(mean(slope), 3),
      SD_slope      = round(sd(slope), 3),
      Pct_above_WBE = round(100 * mean(slope > 0.75), 1)),
    by = clade][order(-N_species)]

  kable(clade_summ,
        col.names = c("Clade", "N species", "Median β", "IQR lo", "IQR hi",
                      "Mean β", "SD β", "% slopes > WBE 0.75"),
        caption = paste(
          "Summary statistics for intraspecific OLS slopes by major clade.",
          "Species with N ≥ 5 metabolic rate records.",
          "WBE reference: β = 0.75.",
          "IQR = interquartile range of slope distribution."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

---

# Growth Rate Scaling

WBE predicts that absolute anabolic production (growth rate $G$) scales as $G \propto M^{3/4}$, the same exponent as whole-organism metabolic rate. This analysis uses Hatton et al. (2019), the only dataset in this compilation with growth rate data.

> **Unit note:** The growth rate column (`growth_rate_value`) from Hatton 2019 is expressed in units indicated by `growth_rate_unit`. Check the column carefully before interpreting the sign and magnitude of the slope. If rates are **mass-specific** (per gram per year), the WBE expectation is a negative slope (~−0.25). If rates are **absolute** (grams per year), the expected slope is **+0.75**.

```{r growth-prep}
grow_dat <- qa_ok(h19)[!is.na(growth_rate_value) & growth_rate_value > 0 &
                        !is.na(body_mass_g) & body_mass_g > 0]

cat(sprintf("Growth rate data (Hatton 2019, QA-pass): n = %d\n", nrow(grow_dat)))
cat("\nGrowth rate unit(s):\n")
print(table(grow_dat$growth_rate_unit, useNA = "ifany"))
cat("\nBy taxonomic group:\n")
print(grow_dat[, .N, by = input_taxonomic_group][order(-N)])
```

```{r growth-plot, fig.width=10, fig.height=6}
if (nrow(grow_dat) >= 10) {
  fit_grow <- lm(log10(growth_rate_value) ~ log10(body_mass_g), data = grow_dat)
  int_grow <- coef(fit_grow)[1]
  slp_grow <- coef(fit_grow)[2]
  xrg      <- range(log10(grow_dat$body_mass_g), na.rm = TRUE)

  # Reference slope depends on unit type; use 0.75 per WBE for absolute production
  ref_grow <- data.frame(lx = xrg, ly = int_grow + 0.75 * xrg)

  ggplot(grow_dat,
         aes(x = log10(body_mass_g), y = log10(growth_rate_value),
             colour = input_taxonomic_group)) +
    geom_point(alpha = 0.5, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8,
                aes(group = input_taxonomic_group)) +
    geom_abline(intercept = int_grow, slope = slp_grow,
                colour = "black", linewidth = 1.2) +
    geom_line(data = ref_grow, aes(x = lx, y = ly),
              inherit.aes = FALSE, linetype = "dashed",
              colour = "firebrick", linewidth = 1.0) +
    scale_colour_viridis_d(option = "turbo", name = "Taxonomic group") +
    labs(
      title    = expression(paste("Growth rate scaling: ", italic(G), " ∝ ", italic(M)^{beta})),
      subtitle = sprintf(
        "Solid black = overall OLS β = %.3f; dashed red = WBE reference β = 0.75 (absolute production)",
        slp_grow),
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = bquote(log[10] ~ paste("Growth rate (", .(
                    if (nrow(grow_dat) > 0) unique(grow_dat$growth_rate_unit)[1] else "value"
                  ), ")"))
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

```{r growth-ols-table}
reg_grow <- fit_log_lm(grow_dat, "body_mass_g", "growth_rate_value",
                        group_col = "input_taxonomic_group", min_n = 10)

if (!is.null(reg_grow) && nrow(reg_grow) > 0) {
  kable(reg_grow,
        col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regressions of growth rate on body mass (g).",
          "Data: Hatton et al. 2019 (PNAS).",
          "WBE predicted slope for absolute production: β = 0.75.",
          "Interpretation depends on whether growth_rate_unit reflects mass-specific or absolute rates."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    row_spec(1, bold = TRUE, background = "#dce8ff")
}
```

```{r growth-sma-table}
if (HAS_SMATR && nrow(grow_dat) >= 10) {
  reg_grow_sma <- fit_sma(grow_dat, "body_mass_g", "growth_rate_value",
                           group_col = "input_taxonomic_group", min_n = 10)
  if (!is.null(reg_grow_sma) && nrow(reg_grow_sma) > 0) {
    kable(reg_grow_sma,
          col.names = c("Group", "n", "Method", "log₁₀ Intercept", "Slope (β)",
                        "95% CI lo", "95% CI hi", "R²"),
          caption = "SMA log-log regressions of growth rate on body mass. WBE predicted slope: β = 0.75.") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
      row_spec(1, bold = TRUE, background = "#ffe8dc")
  }
}
```

---

# Grady et al. 2014 — Growth Rate Allometry in Vertebrates and Dinosaurs

Grady et al. (2014, *Science* 344:1268–1272) assembled maximum growth rates ($G_\text{max}$, g day⁻¹) and resting metabolic rates for 381 extant vertebrate species plus 28 extinct taxa (non-avian dinosaurs, stem birds, stem mammals), to test the metabolic strategy (ectothermy, endothermy, or **mesothermy**) of dinosaurs.

WBE predicts $G_\text{max} \propto M^{3/4}$, the same exponent as whole-organism metabolic rate. Endotherms (mammals, birds) lie above ectotherms (fish, reptiles) on the growth-rate–body-mass relationship because they sustain higher absolute production rates at the same body mass. **Mesotherms** are predicted to fall between these two scaling lines — the key test for dinosaurs.

> **Data source:** Grady et al. Table S1, compiled at `GlobalBodySize/output/grady2014_growth_compiled.csv`. Body mass = final adult mass (g). Growth rate = $G_\text{max}$ (g day⁻¹). Metabolic rate (W) available for a subset.

```{r grady-load}
grady_path <- "../GlobalBodySize/output/grady2014_growth_compiled.csv"
if (!file.exists(grady_path)) {
  message("Grady 2014 data not found at: ", grady_path,
          "\nSkipping Grady sections. Copy from GlobalBodySize/output/ if available.")
  HAS_GRADY <- FALSE
  gr <- data.table()
} else {
  gr <- fread(grady_path)
  HAS_GRADY <- TRUE
  cat(sprintf("Grady 2014: %d total rows\n", nrow(gr)))
}

if (HAS_GRADY) {
  ## Working dataset: both mass and gmax available
  gr_ok <- gr[!is.na(final_adult_mass_g) & final_adult_mass_g > 0 &
              !is.na(gmax_g_per_day) & gmax_g_per_day > 0 &
              data_quality_flag %in% c("ok", "", NA_character_)]

  ## Metabolic strategy label
  gr_ok[, strategy := fcase(
    is_extinct == TRUE,                          "Extinct (dinosaurs/stem lineages)",
    is_mesotherm == TRUE,                        "Mesotherm (extant)",
    input_taxonomic_group %in% c("mammal","bird"), "Endotherm (mammal/bird)",
    input_taxonomic_group %in% c("fish","reptile"), "Ectotherm (fish/reptile)",
    default = "Other"
  )]

  cat(sprintf("Records with mass + gmax: n = %d\n", nrow(gr_ok)))
  cat("\nBy strategy:\n")
  print(gr_ok[, list(n=.N,
                     mass_range_g = paste0(round(min(final_adult_mass_g),3)," – ",
                                           round(max(final_adult_mass_g),0)),
                     gmax_range = paste0(round(min(gmax_g_per_day),4)," – ",
                                         round(max(gmax_g_per_day),1))),
              by=strategy][order(-n)])
}
```

## Gmax vs. body mass — all taxa {#grady-scatter}

```{r grady-scatter, fig.width=10, fig.height=7}
if (HAS_GRADY && nrow(gr_ok) >= 10) {
  pal_strat <- c(
    "Endotherm (mammal/bird)"            = pal8[2],
    "Ectotherm (fish/reptile)"           = pal8[4],
    "Mesotherm (extant)"                 = pal8[3],
    "Extinct (dinosaurs/stem lineages)"  = "darkorange2",
    "Other"                              = "grey50"
  )
  shape_strat <- c(
    "Endotherm (mammal/bird)"            = 16,
    "Ectotherm (fish/reptile)"           = 17,
    "Mesotherm (extant)"                 = 15,
    "Extinct (dinosaurs/stem lineages)"  = 18,
    "Other"                              = 1
  )

  ## Overall OLS
  fit_gr_all <- lm(log10(gmax_g_per_day) ~ log10(final_adult_mass_g), data = gr_ok)
  slp_gr_all <- coef(fit_gr_all)[2]
  int_gr_all <- coef(fit_gr_all)[1]
  xrg_gr     <- range(log10(gr_ok$final_adult_mass_g), na.rm = TRUE)
  ref_gr     <- data.frame(lx = xrg_gr, ly = int_gr_all + 0.75 * xrg_gr)

  ggplot(gr_ok, aes(x = log10(final_adult_mass_g), y = log10(gmax_g_per_day),
                    colour = strategy, shape = strategy)) +
    geom_point(aes(size = ifelse(is_extinct, 3, 1.6)), alpha = 0.75) +
    scale_size_identity() +
    geom_smooth(data = gr_ok[strategy == "Endotherm (mammal/bird)"],
                mapping = aes(x = log10(final_adult_mass_g), y = log10(gmax_g_per_day), group = 1),
                method = "lm", se = TRUE,
                colour = pal8[2], fill = pal8[2], alpha = 0.15, linewidth = 1.0,
                inherit.aes = FALSE) +
    geom_smooth(data = gr_ok[strategy == "Ectotherm (fish/reptile)"],
                mapping = aes(x = log10(final_adult_mass_g), y = log10(gmax_g_per_day), group = 1),
                method = "lm", se = TRUE,
                colour = pal8[4], fill = pal8[4], alpha = 0.15, linewidth = 1.0,
                inherit.aes = FALSE) +
    geom_line(data = ref_gr, aes(x = lx, y = ly),
              inherit.aes = FALSE, linetype = "dashed", colour = "black",
              linewidth = 0.9) +
    scale_colour_manual(values = pal_strat, name = "Metabolic strategy") +
    scale_shape_manual(values  = shape_strat, name = "Metabolic strategy") +
    annotate("text", x = xrg_gr[2] - 0.5,
             y = int_gr_all + slp_gr_all * (xrg_gr[2] - 0.5) + 0.8,
             label = sprintf("Overall OLS β = %.3f", slp_gr_all),
             size = 3.5, hjust = 1) +
    annotate("text", x = xrg_gr[1] + 0.3,
             y = int_gr_all + 0.75 * (xrg_gr[1] + 0.3) - 0.8,
             label = "WBE β = 0.75", colour = "grey30", size = 3.5, hjust = 0) +
    labs(
      title    = expression(paste("Growth rate allometry: ", italic(G)[max], " ∝ ", italic(M)^{beta})),
      subtitle = "Grady et al. 2014 (Science). Shaded bands = per-strategy OLS 95% CI. Dashed = WBE β = 0.75.",
      x        = expression(log[10] ~ "Adult body mass (g)"),
      y        = expression(log[10] ~ italic(G)[max] ~ "(g day"^{-1}*")")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom",
          legend.text = element_text(size = 10))
}
```

## OLS regression by metabolic strategy {#grady-ols-table}

```{r grady-ols-table}
if (HAS_GRADY && nrow(gr_ok) >= 10) {
  ## Fit OLS per strategy + overall
  fit_grady_strat <- function(dt, label) {
    if (nrow(dt) < 5) return(NULL)
    dt2 <- dt[!is.na(final_adult_mass_g) & final_adult_mass_g > 0 &
              !is.na(gmax_g_per_day) & gmax_g_per_day > 0]
    if (nrow(dt2) < 5 || length(unique(log10(dt2$final_adult_mass_g))) < 2) return(NULL)
    fit  <- lm(log10(gmax_g_per_day) ~ log10(final_adult_mass_g), data = dt2)
    cf   <- coef(summary(fit))
    ci   <- confint(fit)
    data.table(
      Strategy  = label,
      n         = nrow(dt2),
      Intercept = round(coef(fit)[1], 3),
      Slope     = round(coef(fit)[2], 3),
      Slope_SE  = round(cf[2, 2], 3),
      CI_95_lo  = round(ci[2, 1], 3),
      CI_95_hi  = round(ci[2, 2], 3),
      R2        = round(summary(fit)$r.squared, 3),
      p_value   = signif(cf[2, 4], 3)
    )
  }

  strat_levels <- c("Endotherm (mammal/bird)", "Ectotherm (fish/reptile)",
                    "Mesotherm (extant)", "Extinct (dinosaurs/stem lineages)")
  reg_grady <- rbindlist(c(
    list(fit_grady_strat(gr_ok, "All taxa combined")),
    lapply(strat_levels, function(s) fit_grady_strat(gr_ok[strategy == s], s))
  ), fill = TRUE)

  if (!is.null(reg_grady) && nrow(reg_grady) > 0) {
    kable(reg_grady,
          col.names = c("Metabolic strategy", "n", "log₁₀ Intercept", "Slope (β)",
                        "SE", "95% CI lo", "95% CI hi", "R²", "p-value"),
          caption = paste(
            "OLS log-log regressions of maximum growth rate (g day⁻¹) on adult body mass (g)",
            "by metabolic strategy. Grady et al. 2014 (Science 344:1268).",
            "Dinosaurs (extinct) falling between endotherm and ectotherm intercepts",
            "support the mesothermy hypothesis. WBE predicted slope: β = 0.75."
          )) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
      row_spec(1, bold = TRUE, background = "#dce8ff") %>%
      row_spec(which(reg_grady$Strategy == "Extinct (dinosaurs/stem lineages)"),
               bold = TRUE, background = "#fff3cd")
  }
}
```

## Extinct (dinosaur) growth rates relative to extant vertebrates {#grady-extinct}

```{r grady-extinct-plot, fig.width=10, fig.height=7}
if (HAS_GRADY && nrow(gr_ok[is_extinct == TRUE]) >= 3) {
  ## Separate endo / ecto regression lines for background
  gr_endo <- gr_ok[strategy == "Endotherm (mammal/bird)"]
  gr_ecto <- gr_ok[strategy == "Ectotherm (fish/reptile)"]
  xrg2    <- range(log10(gr_ok$final_adult_mass_g), na.rm = TRUE)

  ## Compute bounding lines
  fit_endo <- lm(log10(gmax_g_per_day) ~ log10(final_adult_mass_g), data = gr_endo)
  fit_ecto <- lm(log10(gmax_g_per_day) ~ log10(final_adult_mass_g), data = gr_ecto)
  df_endo  <- data.frame(lx = xrg2, ly = coef(fit_endo)[1] + coef(fit_endo)[2] * xrg2)
  df_ecto  <- data.frame(lx = xrg2, ly = coef(fit_ecto)[1] + coef(fit_ecto)[2] * xrg2)

  gr_extinct <- gr_ok[is_extinct == TRUE]

  ggplot() +
    ## Grey ribbon between endo and ecto lines
    geom_ribbon(
      data = data.frame(
        lx    = xrg2,
        ymin  = coef(fit_ecto)[1] + coef(fit_ecto)[2] * xrg2,
        ymax  = coef(fit_endo)[1] + coef(fit_endo)[2] * xrg2
      ),
      aes(x = lx, ymin = ymin, ymax = ymax),
      fill = "grey85", alpha = 0.6, inherit.aes = FALSE
    ) +
    ## Extant background points
    geom_point(data = gr_ok[is_extinct == FALSE],
               aes(x = log10(final_adult_mass_g), y = log10(gmax_g_per_day),
                   colour = strategy),
               alpha = 0.35, size = 1.4, shape = 16) +
    ## Endo/ecto regression lines
    geom_line(data = df_endo, aes(x = lx, y = ly),
              colour = pal8[2], linewidth = 1.1, linetype = "solid") +
    geom_line(data = df_ecto, aes(x = lx, y = ly),
              colour = pal8[4], linewidth = 1.1, linetype = "solid") +
    ## Extinct taxa (large points, labelled)
    geom_point(data = gr_extinct,
               aes(x = log10(final_adult_mass_g), y = log10(gmax_g_per_day)),
               colour = "darkorange2", size = 3.5, shape = 18) +
    ggrepel::geom_text_repel(
      data = gr_extinct,
      aes(x = log10(final_adult_mass_g), y = log10(gmax_g_per_day),
          label = verbatim_taxon_name),
      size = 2.8, colour = "darkorange4", max.overlaps = 20, segment.alpha = 0.4
    ) +
    scale_colour_manual(
      values = c("Endotherm (mammal/bird)"   = pal8[2],
                 "Ectotherm (fish/reptile)"  = pal8[4],
                 "Mesotherm (extant)"        = pal8[3]),
      name = "Extant taxa") +
    labs(
      title    = "Dinosaur and stem-lineage growth rates vs. extant vertebrates",
      subtitle = paste(
        "Orange diamonds = extinct taxa. Shaded band = mesotherm zone between endo-",
        "and ectotherm scaling lines. Blue = endotherm OLS; green = ectotherm OLS."
      ),
      x = expression(log[10] ~ "Adult body mass (g)"),
      y = expression(log[10] ~ italic(G)[max] ~ "(g day"^{-1}*")")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

## Metabolic rate vs. body mass (Grady subset) {#grady-metabolic}

```{r grady-metabolic-plot, fig.width=9, fig.height=6}
if (HAS_GRADY) {
  gr_met <- gr_ok[!is.na(metabolic_rate_W) & metabolic_rate_W > 0]
  cat(sprintf("Grady subset with metabolic rate: n = %d\n", nrow(gr_met)))

  if (nrow(gr_met) >= 10) {
    fit_grmet <- lm(log10(metabolic_rate_W) ~ log10(final_adult_mass_g), data = gr_met)
    xrg_met   <- range(log10(gr_met$final_adult_mass_g), na.rm = TRUE)
    ref_met   <- data.frame(lx = xrg_met,
                            ly = coef(fit_grmet)[1] + 0.75 * xrg_met)

    ggplot(gr_met, aes(x = log10(final_adult_mass_g), y = log10(metabolic_rate_W),
                       colour = strategy, shape = strategy)) +
      geom_point(alpha = 0.7, size = 2) +
      geom_smooth(data = gr_met[strategy %in% c("Endotherm (mammal/bird)",
                                                 "Ectotherm (fish/reptile)")],
                  mapping = aes(x = log10(final_adult_mass_g), y = log10(metabolic_rate_W),
                                group = strategy, colour = strategy),
                  method = "lm", se = FALSE,
                  linewidth = 1.0, inherit.aes = FALSE) +
      geom_line(data = ref_met, aes(x = lx, y = ly),
                inherit.aes = FALSE, linetype = "dashed", colour = "black",
                linewidth = 0.9) +
      scale_colour_manual(values = pal_strat, name = "Metabolic strategy") +
      scale_shape_manual(values  = shape_strat, name = "Metabolic strategy") +
      annotate("text", x = xrg_met[2] - 0.5,
               y = coef(fit_grmet)[1] + 0.75 * (xrg_met[2] - 0.5) - 1.0,
               label = "WBE β = 0.75", colour = "grey30", size = 3.5, hjust = 1) +
      labs(
        title    = expression(paste("Metabolic rate scaling (Grady 2014 subset): ", italic(B), " ∝ ", italic(M)^{beta})),
        subtitle = sprintf("n = %d with metabolic rate. OLS β = %.3f. Dashed = WBE β = 0.75.",
                           nrow(gr_met), coef(fit_grmet)[2]),
        x        = expression(log[10] ~ "Adult body mass (g)"),
        y        = expression(log[10] ~ "Metabolic rate (W)")
      ) +
      theme_bw(base_size = 13) +
      theme(legend.position = "bottom")
  }
}
```

## Grady OLS summary table {#grady-met-table}

```{r grady-met-ols-table}
if (HAS_GRADY) {
  gr_met2 <- gr_ok[!is.na(metabolic_rate_W) & metabolic_rate_W > 0]
  if (nrow(gr_met2) >= 10) {
    reg_grmet <- rbindlist(c(
      list(fit_grady_strat(gr_met2, "All taxa combined (metabolic rate)")),
      lapply(strat_levels, function(s)
        fit_grady_strat(gr_met2[strategy == s], paste0(s, " (metabolic rate)")))
    ), fill = TRUE)
    if (!is.null(reg_grmet) && nrow(reg_grmet) > 0) {
      kable(reg_grmet,
            col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                          "95% CI lo", "95% CI hi", "R²", "p-value"),
            caption = "OLS metabolic rate (W) vs. adult body mass (g) from the Grady 2014 subset with both variables. WBE predicted slope: β = 0.75.") %>%
        kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE)
    }
  }
}
```

---

# Lifespan Scaling

WBE theory predicts maximum lifespan $L \propto M^{1/4}$ because lifespan scales inversely with mass-specific metabolic rate ($B/M \propto M^{-1/4}$). This pattern is well-documented in mammals and birds but more variable in ectotherms and unicellular organisms, where extrinsic mortality dominates. We use the Hatton 2019 dataset, which provides lifespan data spanning prokaryotes to large vertebrates.

```{r lifespan-prep}
lisp_dat <- qa_ok(h19)[!is.na(lifespan_max_years) & lifespan_max_years > 0 &
                        !is.na(body_mass_g) & body_mass_g > 0]

cat(sprintf("Lifespan data (Hatton 2019, QA-pass): n = %d\n", nrow(lisp_dat)))
cat("\nBy taxonomic group:\n")
print(lisp_dat[, .N, by = input_taxonomic_group][order(-N)])
```

```{r lifespan-plot, fig.width=10, fig.height=6}
if (nrow(lisp_dat) >= 10) {
  fit_lisp <- lm(log10(lifespan_max_years) ~ log10(body_mass_g), data = lisp_dat)
  int_lisp <- coef(fit_lisp)[1]
  slp_lisp <- coef(fit_lisp)[2]
  xrl      <- range(log10(lisp_dat$body_mass_g), na.rm = TRUE)
  ref_lisp <- data.frame(lx = xrl, ly = int_lisp + 0.25 * xrl)

  ggplot(lisp_dat,
         aes(x = log10(body_mass_g), y = log10(lifespan_max_years),
             colour = input_taxonomic_group)) +
    geom_point(alpha = 0.5, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8,
                aes(group = input_taxonomic_group)) +
    geom_abline(intercept = int_lisp, slope = slp_lisp,
                colour = "black", linewidth = 1.2) +
    geom_line(data = ref_lisp, aes(x = lx, y = ly),
              inherit.aes = FALSE, linetype = "dashed",
              colour = "firebrick", linewidth = 1.0) +
    scale_colour_viridis_d(option = "turbo", name = "Taxonomic group") +
    labs(
      title    = expression(paste("Lifespan scaling: ", italic(L), " ∝ ", italic(M)^{1/4})),
      subtitle = sprintf(
        "Solid black = overall OLS β = %.3f; dashed red = WBE reference β = 0.25",
        slp_lisp),
      x        = expression(log[10] ~ "Body mass (g)"),
      y        = expression(log[10] ~ "Maximum lifespan (years)")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

```{r lifespan-ols-table}
reg_lisp <- fit_log_lm(lisp_dat, "body_mass_g", "lifespan_max_years",
                        group_col = "input_taxonomic_group", min_n = 10)

if (!is.null(reg_lisp) && nrow(reg_lisp) > 0) {
  kable(reg_lisp,
        col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regressions of maximum lifespan (years) on body mass (g).",
          "Data: Hatton et al. 2019 (PNAS).",
          "WBE predicted slope: β = 0.25."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    row_spec(1, bold = TRUE, background = "#dce8ff")
}
```

```{r lifespan-sma-table}
if (HAS_SMATR && nrow(lisp_dat) >= 10) {
  reg_lisp_sma <- fit_sma(lisp_dat, "body_mass_g", "lifespan_max_years",
                           group_col = "input_taxonomic_group", min_n = 10)
  if (!is.null(reg_lisp_sma) && nrow(reg_lisp_sma) > 0) {
    kable(reg_lisp_sma,
          col.names = c("Group", "n", "Method", "log₁₀ Intercept", "Slope (β)",
                        "95% CI lo", "95% CI hi", "R²"),
          caption = "SMA log-log regressions of maximum lifespan (years) on body mass (g). WBE predicted slope: β = 0.25.") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
      row_spec(1, bold = TRUE, background = "#ffe8dc")
  }
}
```

---

# Predator–Prey Mass Scaling

Hatton et al. (2015) compiled a global database of predator and prey body masses across trophic levels. The theoretical expectation from optimal foraging and energetic constraints is $M_\text{pred} \propto M_\text{prey}^{1.0}$, i.e., predator and prey masses co-scale approximately isometrically, though predators are typically 10–1000× heavier than prey.

```{r predprey-prep}
pp_dat <- qa_ok(h15)[!is.na(predator_mass_g_paired) & predator_mass_g_paired > 0 &
                      !is.na(body_mass_g) & body_mass_g > 0]

cat(sprintf("Predator-prey mass data (Hatton 2015, QA-pass): n = %d\n", nrow(pp_dat)))
cat("\nBy taxonomic group:\n")
print(pp_dat[, .N, by = input_taxonomic_group][order(-N)])
cat("\nPrey mass range (g):\n")
cat(sprintf("  %.2e – %.2e\n", min(pp_dat$body_mass_g), max(pp_dat$body_mass_g)))
cat("Predator mass range (g):\n")
cat(sprintf("  %.2e – %.2e\n", min(pp_dat$predator_mass_g_paired),
            max(pp_dat$predator_mass_g_paired)))
```

```{r predprey-plot, fig.width=10, fig.height=7}
if (nrow(pp_dat) >= 10) {
  fit_pp  <- lm(log10(predator_mass_g_paired) ~ log10(body_mass_g), data = pp_dat)
  int_pp  <- coef(fit_pp)[1]
  slp_pp  <- coef(fit_pp)[2]
  xrp     <- range(log10(pp_dat$body_mass_g), na.rm = TRUE)
  ref_pp  <- data.frame(lx = xrp, ly = int_pp + 1.0 * xrp)

  ggplot(pp_dat,
         aes(x = log10(body_mass_g), y = log10(predator_mass_g_paired),
             colour = input_taxonomic_group)) +
    geom_point(alpha = 0.35, size = 1.5) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8,
                aes(group = input_taxonomic_group)) +
    geom_abline(intercept = int_pp, slope = slp_pp,
                colour = "black", linewidth = 1.2) +
    geom_line(data = ref_pp, aes(x = lx, y = ly),
              inherit.aes = FALSE, linetype = "dashed",
              colour = "firebrick", linewidth = 1.0) +
    scale_colour_viridis_d(option = "turbo", name = "Taxonomic group") +
    labs(
      title    = expression(paste("Predator–prey mass scaling: ", italic(M)[pred], " ∝ ", italic(M)[prey]^{beta})),
      subtitle = sprintf(
        "Solid black = overall OLS β = %.3f; dashed red = isometric reference β = 1.0",
        slp_pp),
      x        = expression(log[10] ~ "Prey body mass (g)"),
      y        = expression(log[10] ~ "Predator body mass (g)")
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

```{r predprey-ols-table}
reg_pp <- fit_log_lm(pp_dat, "body_mass_g", "predator_mass_g_paired",
                     group_col = "input_taxonomic_group", min_n = 10)

if (!is.null(reg_pp) && nrow(reg_pp) > 0) {
  kable(reg_pp,
        col.names = c("Group", "n", "log₁₀ Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value"),
        caption = paste(
          "OLS log-log regressions of predator body mass (g) on prey body mass (g).",
          "Data: Hatton et al. 2015 (Science).",
          "Theoretical reference: β ≈ 1.0 (isometric predator–prey mass co-scaling).",
          "Intercept reflects log₁₀ of the predator:prey mass ratio at unit prey mass."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    row_spec(1, bold = TRUE, background = "#dce8ff")
}
```

```{r predprey-sma-table}
if (HAS_SMATR && nrow(pp_dat) >= 10) {
  reg_pp_sma <- fit_sma(pp_dat, "body_mass_g", "predator_mass_g_paired",
                        group_col = "input_taxonomic_group", min_n = 10)
  if (!is.null(reg_pp_sma) && nrow(reg_pp_sma) > 0) {
    kable(reg_pp_sma,
          col.names = c("Group", "n", "Method", "log₁₀ Intercept", "Slope (β)",
                        "95% CI lo", "95% CI hi", "R²"),
          caption = paste(
            "SMA log-log regressions of predator body mass on prey body mass.",
            "SMA appropriate here since both predator and prey masses carry measurement uncertainty.",
            "Theoretical reference: β ≈ 1.0."
          )) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
      row_spec(1, bold = TRUE, background = "#ffe8dc")
  }
}
```

---

# Methods & Caveats

## Regression methods

**Ordinary Least Squares (OLS)** minimizes vertical (y-axis) residuals and is used throughout as the primary regression method. OLS is appropriate when the x-variable (log₁₀ body mass) is measured without error. When both variables carry substantial measurement error — e.g., predator vs. prey mass in field surveys — OLS underestimates the true slope. OLS slopes should be interpreted as lower bounds on the true allometric exponent.

**Standardised Major Axis (SMA)**, implemented via the `smatr` package (Warton et al. 2012), minimizes residuals perpendicular to the line and provides unbiased slope estimates when both axes have comparable error. SMA slopes are always ≥ OLS slopes (equality holds only when R² = 1). SMA is presented as a secondary analysis for comparison.

Both methods fit regressions in log₁₀–log₁₀ space. A fitted slope of $\beta$ corresponds to the power-law exponent in $Y \propto M^\beta$.

**Minimum sample sizes:**  
- Cross-taxa regressions: N ≥ 10  
- Within-family OLS: N ≥ 10  
- Intraspecific OLS: N ≥ 5

## Metabolic rate type heterogeneity

```{r met-type-summary}
if (nrow(met_sources) > 0 && "metabolic_rate_type" %in% names(met_sources)) {
  type_tab <- met_sources[, .N, by = .(dataset_label, metabolic_rate_type)][order(dataset_label, -N)]
  kable(type_tab,
        col.names = c("Dataset", "Metabolic rate type", "N"),
        caption = paste(
          "Metabolic rate measurement type distribution in the combined metabolic dataset (QA-pass, unit = W).",
          "Mixing basal, resting, active, and field metabolic rates inflates residual variance",
          "and may bias cross-taxa slope estimates. Basal-only analyses are presented in Section 3.4."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

**Implications:** Active or field metabolic rates are 2–10× higher than basal rates for endotherms, and the relationship between basal and active rates varies systematically across taxa (endotherms vs. ectotherms). Including mixed rate types elevates the residual variance around the allometric line and may produce a composite slope that does not reflect any single physiological state. Where sample sizes permit, basal-only analyses are preferred for testing WBE theory.

## Chown 2007 insect dataset: data corrections

The Chown et al. 2007 Functional Ecology dataset (Appendix S2; 346 species across 16 insect orders) was ingested from the original supplementary Word document. Three Coleoptera species from Strømme et al. (1986) contained unit-conversion errors in Chown et al. and were corrected using the values reported in Riveros & Enquist (2011, Table 1):

| Species | Reported BMR (µW) | Corrected BMR (µW) | Source |
|---|---|---|---|
| *Amara quenseli* | 1.4 | 346.5 | Riveros & Enquist 2011 |
| *Simplocaria metallica* | 0.5 | 32.7 | Riveros & Enquist 2011 |
| *Rhynchaenus flagellum* | 0.2 | 9.1 | Riveros & Enquist 2011 |

Riveros & Enquist (2011) also reported that when Termitidae are excluded, the overall insect metabolic scaling slope (OLS) is approximately 0.74 (95% CI: 0.71–0.77, n = 367), consistent with WBE predictions. Termitidae deviate toward near-isometric scaling (~β ≈ 1), potentially due to measurement artefacts (CH₄-based respirometry) or life-history differences. Wing status (winged vs. wingless) significantly affects metabolic rate intercepts but not slopes.

## Temperature standardization

No universal temperature correction was applied in this document. The datasets differ in their handling of temperature:

- **Hoehler 2023** standardized all metabolic rates to 25°C using the Arrhenius equation with an activation energy of 0.65 eV, making rates comparable across taxa and environments.
- **Hatton 2019** applied Q10-based temperature corrections to a reference temperature.
- **AnimalTraits** reports metabolic rates at the measurement temperature, which varies by species and study. No standardization was applied in our analysis.

Because metabolic rates scale with temperature according to the Boltzmann-Arrhenius factor ($e^{-E_a/kT}$), differences in measurement temperature introduce systematic offsets. Cross-dataset slope comparisons that ignore temperature may conflate thermal effects with mass effects. Users requiring temperature-corrected comparisons should apply the MTE temperature correction (Brown et al. 2004) before combining datasets.

## Phylogenetic non-independence

OLS regressions do not account for phylogenetic relatedness among species. Closely related species tend to share body mass and metabolic rate values due to common ancestry, violating the OLS assumption of independent residuals. This inflates effective sample sizes and can bias confidence intervals. The slopes and R² values reported here are statistical descriptors of the data and should not be interpreted as phylogenetically corrected partial regression coefficients. Readers requiring phylogenetically corrected estimates should use PGLS (phylogenetic generalized least squares) or related methods (e.g., BayesTraits, ape::pgls).

## QA flag pass rates

```{r methods-qa-summary}
qa_methods <- rbindlist(lapply(names(all_list), function(nm) {
  dt <- all_list[[nm]]
  if (is.null(dt)) return(NULL)
  n_total <- nrow(dt)
  n_pass  <- sum(dt$qa_flag %in% c("ok", ""), na.rm = TRUE)
  n_blank <- sum(dt$qa_flag == "", na.rm = TRUE)
  n_explicit_ok <- sum(dt$qa_flag == "ok", na.rm = TRUE)
  pct_pass <- round(100 * n_pass / n_total, 1)
  data.table(
    Dataset         = labels_vec[nm],
    N_total         = n_total,
    N_explicit_ok   = n_explicit_ok,
    N_blank_pass    = n_blank,
    N_total_pass    = n_pass,
    Pct_pass        = pct_pass,
    Flag            = if ((n_total - n_pass) / n_total > 0.20) "⚠ >20% fail" else "ok"
  )
}))

kable(qa_methods,
      col.names = c("Dataset", "N total", "N explicit 'ok'", "N blank (pass)",
                    "N total pass", "% pass", "Warning"),
      caption = paste(
        "QA flag pass rate analysis.",
        "'Explicit ok': qa_flag == 'ok' (actively audited).",
        "'Blank pass': qa_flag == '' (no flag set; treated as pass by convention).",
        "Datasets where >20% of records fail QA are flagged."
      )) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## Data provenance

| Dataset | Source | Primary DOI | N rows |
|---|---|---|---|
| AnimalTraits | Herberstein et al. 2022, Sci Data | 10.1038/s41597-022-01364-9 | `r if (!is.null(at)) nrow(at) else "N/A"` |
| Hoehler 2023 | Hoehler et al. 2023, PNAS | 10.1073/pnas.2303764120 | `r if (!is.null(pnas)) nrow(pnas) else "N/A"` |
| Hatton 2019 | Hatton et al. 2019, PNAS | 10.1073/pnas.1907703116 | `r if (!is.null(h19)) nrow(h19) else "N/A"` |
| Hatton 2015 | Hatton et al. 2015, Science | 10.1126/science.aac6284 | `r if (!is.null(h15)) nrow(h15) else "N/A"` |
| Chown 2007 (corrected) | Chown et al. 2007, Funct Ecol; corrections: Riveros & Enquist 2011 | 10.1111/j.1365-2435.2007.01245.x | `r if (!is.null(ch07)) nrow(ch07) else "N/A"` |

All data were loaded from local CSV files in `output/` compiled by the `Animal_scaling_data` ingestion pipeline. Column harmonization, QA flagging, and unit assignment were performed during ingestion; see individual provider scripts for details.

---

# Summary Statistics for Slope Distributions Across Taxonomic Levels

This section computes 95% confidence intervals on the mean slope (t-distribution) and on the median slope (bootstrap, 5000 resamples) at each taxonomic level of aggregation: order, family, genus, and intraspecific.

```{r slope-ci-helpers}
# Bootstrap 95% CI on median (5000 resamples, percentile method)
boot_median_ci <- function(x, R = 5000, alpha = 0.05) {
  x <- x[!is.na(x)]
  if (length(x) < 3) return(c(NA_real_, NA_real_))
  set.seed(42)
  boot_meds <- replicate(R, median(sample(x, length(x), replace = TRUE)))
  quantile(boot_meds, c(alpha / 2, 1 - alpha / 2))
}

# t-distribution 95% CI on the mean
mean_ci <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 2) return(c(NA_real_, NA_real_))
  se <- sd(x) / sqrt(n)
  hw <- qt(0.975, df = n - 1) * se
  c(mean(x) - hw, mean(x) + hw)
}

# Build one summary row given a slope vector and a level label
slope_summary_row <- function(slopes, level_label) {
  slopes <- slopes[!is.na(slopes)]
  n <- length(slopes)
  if (n < 3) return(NULL)
  m_ci  <- mean_ci(slopes)
  b_ci  <- boot_median_ci(slopes)
  data.table(
    Level          = level_label,
    N_groups       = n,
    Mean_slope     = round(mean(slopes), 4),
    CI95_lo_mean   = round(m_ci[1], 4),
    CI95_hi_mean   = round(m_ci[2], 4),
    Median_slope   = round(median(slopes), 4),
    CI95_lo_med    = round(b_ci[1], 4),
    CI95_hi_med    = round(b_ci[2], 4)
  )
}
```

## Order-level slope CI

```{r slope-ci-order}
ord_slopes_vec <- if (!is.null(reg_order) && nrow(reg_order) > 0)
  reg_order$Slope[!is.na(reg_order$Slope)] else numeric(0)

ci_order <- slope_summary_row(ord_slopes_vec, "Order")
if (!is.null(ci_order)) {
  kable(ci_order,
        col.names = c("Level", "N groups", "Mean slope", "95% CI lo (mean)",
                      "95% CI hi (mean)", "Median slope", "95% CI lo (median)",
                      "95% CI hi (median)"),
        caption = paste("Order-level metabolic scaling slope summary.",
                        "95% CI on mean: t-distribution.",
                        "95% CI on median: bootstrap percentile, 5000 resamples.",
                        "WBE reference: β = 0.75.")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("Insufficient order-level slopes for CI computation.\n")
}
```

## Family-level slope CI

```{r slope-ci-family}
fam_slopes_vec <- if (!is.null(reg_fam) && nrow(reg_fam) > 0)
  reg_fam$Slope[!is.na(reg_fam$Slope)] else numeric(0)

ci_fam <- slope_summary_row(fam_slopes_vec, "Family")
if (!is.null(ci_fam)) {
  kable(ci_fam,
        col.names = c("Level", "N groups", "Mean slope", "95% CI lo (mean)",
                      "95% CI hi (mean)", "Median slope", "95% CI lo (median)",
                      "95% CI hi (median)"),
        caption = paste("Family-level metabolic scaling slope summary.",
                        "95% CI on mean: t-distribution.",
                        "95% CI on median: bootstrap percentile, 5000 resamples.",
                        "WBE reference: β = 0.75.")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("Insufficient family-level slopes for CI computation.\n")
}
```

## Genus-level slope CI

```{r slope-ci-genus}
# Compute genus-level regressions if the column exists in met_sources
if ("genus" %in% names(met_sources)) {
  reg_genus_ci <- fit_log_lm(met_sources, "body_mass_g", "metabolic_rate_value",
                              group_col = "genus", min_n = 10)
  reg_genus_ci <- reg_genus_ci[Group != "All taxa combined"]
  reg_genus_ci <- reg_genus_ci[is.na(Slope) | abs(Slope) <= SLOPE_MAX]
  gen_slopes_vec <- reg_genus_ci$Slope[!is.na(reg_genus_ci$Slope)]
} else {
  reg_genus_ci   <- data.table()
  gen_slopes_vec <- numeric(0)
  cat("Column 'genus' not found in met_sources; genus-level CI skipped.\n")
}

ci_gen <- slope_summary_row(gen_slopes_vec, "Genus")
if (!is.null(ci_gen)) {
  kable(ci_gen,
        col.names = c("Level", "N groups", "Mean slope", "95% CI lo (mean)",
                      "95% CI hi (mean)", "Median slope", "95% CI lo (median)",
                      "95% CI hi (median)"),
        caption = paste("Genus-level metabolic scaling slope summary.",
                        "95% CI on mean: t-distribution.",
                        "95% CI on median: bootstrap percentile, 5000 resamples.",
                        "WBE reference: β = 0.75.")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("Insufficient genus-level slopes for CI computation.\n")
}
```

## Intraspecific slope CI

```{r slope-ci-intraspecific}
intra_slopes_vec <- if (exists("intra_slopes") && !is.null(intra_slopes) && nrow(intra_slopes) > 0)
  intra_slopes$slope[!is.na(intra_slopes$slope)] else numeric(0)

ci_intra <- slope_summary_row(intra_slopes_vec, "Intraspecific")
if (!is.null(ci_intra)) {
  kable(ci_intra,
        col.names = c("Level", "N groups", "Mean slope", "95% CI lo (mean)",
                      "95% CI hi (mean)", "Median slope", "95% CI lo (median)",
                      "95% CI hi (median)"),
        caption = paste("Intraspecific metabolic scaling slope summary.",
                        "Each group = one species with N ≥ 5 records.",
                        "95% CI on mean: t-distribution.",
                        "95% CI on median: bootstrap percentile, 5000 resamples.",
                        "WBE reference: β = 0.75.")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("Insufficient intraspecific slopes for CI computation.\n")
}
```

## Combined CI comparison across taxonomic levels

```{r slope-ci-combined-table}
ci_all_levels <- rbindlist(Filter(Negate(is.null),
                                  list(ci_order, ci_fam, ci_gen, ci_intra)),
                           fill = TRUE)

if (!is.null(ci_all_levels) && nrow(ci_all_levels) >= 2) {
  kable(ci_all_levels,
        col.names = c("Level", "N groups", "Mean β", "95% CI lo (mean)",
                      "95% CI hi (mean)", "Median β", "95% CI lo (median)",
                      "95% CI hi (median)"),
        caption = paste("Summary of mean and median metabolic scaling slopes with 95% CIs",
                        "across taxonomic levels of aggregation.",
                        "WBE reference: β = 0.75.",
                        "Mean CI: t-distribution; Median CI: bootstrap percentile (R = 5000).")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r slope-ci-combined-plot, fig.width=10, fig.height=6}
if (!is.null(ci_all_levels) && nrow(ci_all_levels) >= 2) {
  # Reshape to long for mean and median
  ci_mean_long <- ci_all_levels[, .(
    Level, N_groups,
    Estimate = Mean_slope,
    CI_lo    = CI95_lo_mean,
    CI_hi    = CI95_hi_mean,
    Type     = "Mean (t-CI)"
  )]
  ci_med_long <- ci_all_levels[, .(
    Level, N_groups,
    Estimate = Median_slope,
    CI_lo    = CI95_lo_med,
    CI_hi    = CI95_hi_med,
    Type     = "Median (bootstrap CI)"
  )]
  ci_long <- rbind(ci_mean_long, ci_med_long)
  ci_long[, Level := factor(Level, levels = c("Order", "Family", "Genus", "Intraspecific"))]

  ggplot(ci_long, aes(x = Level, y = Estimate, colour = Type,
                      ymin = CI_lo, ymax = CI_hi)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_errorbar(aes(ymin = CI_lo, ymax = CI_hi),
                  width = 0.15, position = position_dodge(width = 0.4),
                  linewidth = 0.9) +
    geom_point(size = 3.5, position = position_dodge(width = 0.4)) +
    scale_colour_manual(values = c("Mean (t-CI)" = pal8[1],
                                   "Median (bootstrap CI)" = pal8[2]),
                        name = "Statistic") +
    annotate("text", x = 0.55, y = 0.75, label = "WBE β = 0.75",
             colour = "firebrick", hjust = 0, vjust = -0.5, size = 3.5) +
    labs(
      title   = "Mean and median metabolic scaling slopes with 95% CIs across taxonomic levels",
      subtitle = "Error bars: t-CI (mean) and bootstrap percentile CI (median); dashed = WBE reference",
      x       = "Taxonomic level",
      y       = "OLS slope (β)"
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

---

# Filtered Analyses: Animals Only, N ≥ 10, Size Range ≥ 3 Orders of Magnitude

This section repeats the metabolic scaling analyses after applying three additional filters:

1. **Animals only** — rows where `input_taxonomic_group` matches non-animal patterns (bacteria, archaea, fungi, algae, plants, protists, etc.) are excluded.
2. **N ≥ 10** — each taxon group must have at least 10 valid observations.
3. **Body mass span ≥ 3 orders of magnitude** — within each group, max(body_mass_g)/min(body_mass_g) ≥ 1000, ensuring the regression spans ecologically meaningful size variation rather than spuriously extrapolating from a narrow range.

> **Rationale:** Including microbes, plants, and groups with narrow mass ranges can distort the slope distribution and inflate apparent agreement with WBE theory. These filters isolate the signal from well-sampled animal groups with broad mass coverage.

## Define animal-only dataset

```{r filtered-animals-prep}
non_animal_pattern <- "plant|microbe|bacteri|archae|fungi|alga|phyto|protist"

met_animals <- met_sources[
  !grepl(non_animal_pattern, input_taxonomic_group, ignore.case = TRUE) &
  !is.na(input_taxonomic_group) & input_taxonomic_group != ""
]

cat("Animal-only metabolic records:", nrow(met_animals), "\n")
cat("Groups retained:\n")
print(unique(met_animals$input_taxonomic_group))
cat("\nTotal records in met_sources:", nrow(met_sources), "\n")
cat("Records excluded (non-animal):", nrow(met_sources) - nrow(met_animals), "\n")
```

## Define filtered regression helper

```{r filtered-lm-helper}
# fit_log_lm_filtered: like fit_log_lm but additionally requires
#   max(body_mass_g)/min(body_mass_g) >= 1000 within each group (3 OOM)
fit_log_lm_filtered <- function(dt, x_col, y_col, group_col = NULL,
                                 min_n = 10, min_mass_ratio = 1000) {
  if (is.null(dt) || nrow(dt) == 0) return(data.table())
  if (!all(c(x_col, y_col) %in% names(dt))) return(data.table())

  sub <- dt[!is.na(get(x_col)) & get(x_col) > 0 &
            !is.na(get(y_col)) & get(y_col) > 0]
  if (nrow(sub) == 0) return(data.table())
  sub <- copy(sub)
  sub[, lx := log10(get(x_col))]
  sub[, ly := log10(get(y_col))]

  fit_one_f <- function(d, label) {
    if (nrow(d) < min_n) return(NULL)
    mass_vals  <- d[[x_col]]
    mass_ratio <- max(mass_vals, na.rm = TRUE) / min(mass_vals, na.rm = TRUE)
    if (!is.finite(mass_ratio) || mass_ratio < min_mass_ratio) return(NULL)
    if (length(unique(d$lx)) < 2) return(NULL)
    fit <- tryCatch(lm(ly ~ lx, data = d), error = function(e) NULL)
    if (is.null(fit)) return(NULL)
    s  <- coef(summary(fit))
    if (nrow(s) < 2) return(NULL)
    ci <- tryCatch(confint(fit), error = function(e) matrix(NA_real_, 2, 2))
    data.table(
      Group               = label,
      n                   = nrow(d),
      Intercept           = round(s[1, 1], 3),
      Slope               = round(s[2, 1], 3),
      Slope_SE            = round(s[2, 2], 3),
      CI_95_lo            = round(ci[2, 1], 3),
      CI_95_hi            = round(ci[2, 2], 3),
      R2                  = round(summary(fit)$r.squared, 3),
      p_value             = signif(s[2, 4], 3),
      mass_range_log10    = round(log10(mass_ratio), 2)
    )
  }

  res_all <- fit_one_f(sub, "All groups combined")

  if (!is.null(group_col) && group_col %in% names(sub)) {
    grps    <- sort(unique(sub[[group_col]]))
    res_grp <- rbindlist(
      lapply(grps, function(g) fit_one_f(sub[get(group_col) == g], g)),
      fill = TRUE
    )
    res_all <- rbind(res_all, res_grp, fill = TRUE)
  }
  res_all
}
```

## Filtered metabolic scaling by order

```{r filtered-order-compute}
reg_order_filt_all <- fit_log_lm_filtered(
  met_animals, "body_mass_g", "metabolic_rate_value",
  group_col = "order", min_n = 10, min_mass_ratio = 1000
)

if (!is.null(reg_order_filt_all) && nrow(reg_order_filt_all) > 0) {
  reg_order_filt <- reg_order_filt_all[
    Group != "All groups combined" & (is.na(Slope) | abs(Slope) <= SLOPE_MAX)
  ]
  cat(sprintf("Orders passing all filters: %d\n", nrow(reg_order_filt)))
} else {
  reg_order_filt <- data.table()
  cat("No orders passed the animal / N>=10 / 3-OOM filters.\n",
      "This may reflect limited body mass spread within orders in the current datasets.\n")
}
```

```{r filtered-order-table}
if (!is.null(reg_order_filt) && nrow(reg_order_filt) >= 1) {
  kable(reg_order_filt[order(-n)],
        col.names = c("Order", "n", "Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value",
                      "Mass range (log₁₀ decades)"),
        caption = paste(
          "Filtered order-level OLS metabolic scaling regressions.",
          "Filters: animals only; N ≥ 10 per order; body mass span ≥ 3 orders of magnitude.",
          "WBE reference: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r filtered-order-forest, fig.width=10, fig.height=6}
if (!is.null(reg_order_filt) && nrow(reg_order_filt) >= 3) {
  ggplot(reg_order_filt[!is.na(Slope)],
         aes(x = reorder(Group, Slope), y = Slope,
             ymin = CI_95_lo, ymax = CI_95_hi)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_errorbar(width = 0.25, colour = "grey50", linewidth = 0.8) +
    geom_point(size = 3.0, colour = pal8[1]) +
    coord_flip() +
    labs(
      title    = "Filtered metabolic scaling slopes by order (animals, N ≥ 10, ≥ 3 OOM mass span)",
      subtitle = "Points = OLS slope; error bars = 95% CI; dashed = WBE β = 0.75",
      x        = "Order",
      y        = "OLS slope (β)",
      caption  = "Filters: animals only; N ≥ 10 per order; max/min body mass ≥ 1000× within order."
    ) +
    theme_bw(base_size = 13)
} else if (!is.null(reg_order_filt) && nrow(reg_order_filt) > 0) {
  cat(sprintf("Only %d order(s) passed filters; forest plot requires ≥ 3.\n",
              nrow(reg_order_filt)))
}
```

## Filtered metabolic scaling by family

```{r filtered-family-compute}
reg_fam_filt_all <- fit_log_lm_filtered(
  met_animals, "body_mass_g", "metabolic_rate_value",
  group_col = "family", min_n = 10, min_mass_ratio = 1000
)

if (!is.null(reg_fam_filt_all) && nrow(reg_fam_filt_all) > 0) {
  reg_fam_filt <- reg_fam_filt_all[
    Group != "All groups combined" & (is.na(Slope) | abs(Slope) <= SLOPE_MAX)
  ]
  cat(sprintf("Families passing all filters: %d\n", nrow(reg_fam_filt)))
} else {
  reg_fam_filt <- data.table()
  cat("No families passed the animal / N>=10 / 3-OOM filters.\n")
}
```

```{r filtered-family-table}
if (!is.null(reg_fam_filt) && nrow(reg_fam_filt) >= 1) {
  kable(reg_fam_filt[order(-n)],
        col.names = c("Family", "n", "Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value",
                      "Mass range (log₁₀ decades)"),
        caption = paste(
          "Filtered family-level OLS metabolic scaling regressions.",
          "Filters: animals only; N ≥ 10 per family; body mass span ≥ 3 orders of magnitude.",
          "WBE reference: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r filtered-family-forest, fig.width=10, fig.height=7}
if (!is.null(reg_fam_filt) && nrow(reg_fam_filt) >= 3) {
  ggplot(reg_fam_filt[!is.na(Slope)],
         aes(x = reorder(Group, Slope), y = Slope,
             ymin = CI_95_lo, ymax = CI_95_hi)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_errorbar(width = 0.25, colour = "grey50", linewidth = 0.8) +
    geom_point(size = 3.0, colour = pal8[1]) +
    coord_flip() +
    labs(
      title    = "Filtered metabolic scaling slopes by family (animals, N ≥ 10, ≥ 3 OOM mass span)",
      subtitle = "Points = OLS slope; error bars = 95% CI; dashed = WBE β = 0.75",
      x        = "Family",
      y        = "OLS slope (β)",
      caption  = "Filters: animals only; N ≥ 10 per family; max/min body mass ≥ 1000× within family."
    ) +
    theme_bw(base_size = 13)
} else if (!is.null(reg_fam_filt) && nrow(reg_fam_filt) > 0) {
  cat(sprintf("Only %d family/families passed filters; forest plot requires ≥ 3.\n",
              nrow(reg_fam_filt)))
}
```

## Filtered metabolic scaling by genus

```{r filtered-genus-compute}
if ("genus" %in% names(met_animals)) {
  reg_genus_filt_all <- fit_log_lm_filtered(
    met_animals, "body_mass_g", "metabolic_rate_value",
    group_col = "genus", min_n = 10, min_mass_ratio = 1000
  )
  if (!is.null(reg_genus_filt_all) && nrow(reg_genus_filt_all) > 0) {
    reg_genus_filt <- reg_genus_filt_all[
      Group != "All groups combined" & (is.na(Slope) | abs(Slope) <= SLOPE_MAX)
    ]
    cat(sprintf("Genera passing all filters: %d\n", nrow(reg_genus_filt)))
  } else {
    reg_genus_filt <- data.table()
    cat("No genera passed the animal / N>=10 / 3-OOM filters.\n")
  }
} else {
  reg_genus_filt <- data.table()
  cat("Column 'genus' not found in met_animals; genus-level filtered analysis skipped.\n")
}
```

```{r filtered-genus-table}
if (!is.null(reg_genus_filt) && nrow(reg_genus_filt) >= 1) {
  kable(reg_genus_filt[order(-n)],
        col.names = c("Genus", "n", "Intercept", "Slope (β)", "SE",
                      "95% CI lo", "95% CI hi", "R²", "p-value",
                      "Mass range (log₁₀ decades)"),
        caption = paste(
          "Filtered genus-level OLS metabolic scaling regressions.",
          "Filters: animals only; N ≥ 10 per genus; body mass span ≥ 3 orders of magnitude.",
          "WBE reference: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r filtered-genus-forest, fig.width=10, fig.height=7}
if (!is.null(reg_genus_filt) && nrow(reg_genus_filt) >= 3) {
  ggplot(reg_genus_filt[!is.na(Slope)],
         aes(x = reorder(Group, Slope), y = Slope,
             ymin = CI_95_lo, ymax = CI_95_hi)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_errorbar(width = 0.25, colour = "grey50", linewidth = 0.8) +
    geom_point(size = 3.0, colour = pal8[1]) +
    coord_flip() +
    labs(
      title    = "Filtered metabolic scaling slopes by genus (animals, N ≥ 10, ≥ 3 OOM mass span)",
      subtitle = "Points = OLS slope; error bars = 95% CI; dashed = WBE β = 0.75",
      x        = "Genus",
      y        = "OLS slope (β)",
      caption  = "Filters: animals only; N ≥ 10 per genus; max/min body mass ≥ 1000× within genus."
    ) +
    theme_bw(base_size = 13)
} else if (!is.null(reg_genus_filt) && nrow(reg_genus_filt) > 0) {
  cat(sprintf("Only %d genus/genera passed filters; forest plot requires ≥ 3.\n",
              nrow(reg_genus_filt)))
}
```

## Filtered intraspecific scaling (animals, N ≥ 10, ≥ 3 OOM)

```{r filtered-intraspecific-compute}
if ("species" %in% names(met_animals) || exists("sp_col")) {
  sp_col_f <- if ("species" %in% names(met_animals)) "species" else sp_col

  if (sp_col_f %in% names(met_animals)) {
    intra_filt_spp <- met_animals[
      !is.na(get(sp_col_f)) & get(sp_col_f) != "" &
      !is.na(body_mass_g) & body_mass_g > 0 &
      !is.na(metabolic_rate_value) & metabolic_rate_value > 0,
      .(N           = .N,
        mass_ratio  = max(body_mass_g, na.rm = TRUE) / min(body_mass_g, na.rm = TRUE)),
      by = .(species = get(sp_col_f))
    ]
    intra_filt_spp <- intra_filt_spp[N >= 10 & is.finite(mass_ratio) & mass_ratio >= 1000]
    cat(sprintf("Species meeting animal + N>=10 + 3-OOM criteria: %d\n",
                nrow(intra_filt_spp)))

    if (nrow(intra_filt_spp) >= 1) {
      intra_slopes_filt <- rbindlist(lapply(intra_filt_spp$species, function(sp) {
        d <- met_animals[get(sp_col_f) == sp &
                         !is.na(body_mass_g) & body_mass_g > 0 &
                         !is.na(metabolic_rate_value) & metabolic_rate_value > 0]
        if (nrow(d) < 10) return(NULL)
        mass_ratio <- max(d$body_mass_g, na.rm = TRUE) / min(d$body_mass_g, na.rm = TRUE)
        if (!is.finite(mass_ratio) || mass_ratio < 1000) return(NULL)
        fit <- tryCatch(
          lm(log10(metabolic_rate_value) ~ log10(body_mass_g), data = d),
          error = function(e) NULL
        )
        if (is.null(fit)) return(NULL)
        s <- coef(summary(fit))
        if (nrow(s) < 2) return(NULL)
        ci <- tryCatch(confint(fit), error = function(e) matrix(NA_real_, 2, 2))
        data.table(
          species              = sp,
          n                    = nrow(d),
          slope                = round(s[2, 1], 3),
          slope_se             = round(s[2, 2], 3),
          ci_lo                = round(ci[2, 1], 3),
          ci_hi                = round(ci[2, 2], 3),
          r2                   = round(summary(fit)$r.squared, 3),
          p_value              = signif(s[2, 4], 3),
          mass_range_log10     = round(log10(mass_ratio), 2)
        )
      }), fill = TRUE)
      intra_slopes_filt <- intra_slopes_filt[is.na(slope) | abs(slope) <= SLOPE_MAX]
      cat(sprintf("Species with valid filtered intraspecific slopes: %d\n",
                  nrow(intra_slopes_filt)))
    } else {
      intra_slopes_filt <- data.table()
      cat("No species met all three intraspecific filters.\n")
    }
  } else {
    intra_slopes_filt <- data.table()
    cat("Species column not found in met_animals; filtered intraspecific analysis skipped.\n")
  }
} else {
  intra_slopes_filt <- data.table()
  cat("No species column available; filtered intraspecific analysis skipped.\n")
}
```

```{r filtered-intraspecific-table}
if (!is.null(intra_slopes_filt) && nrow(intra_slopes_filt) >= 1) {
  kable(intra_slopes_filt[order(-n)],
        col.names = c("Species", "n", "Slope (β)", "SE", "95% CI lo", "95% CI hi",
                      "R²", "p-value", "Mass range (log₁₀ decades)"),
        caption = paste(
          "Filtered intraspecific OLS metabolic scaling regressions.",
          "Filters: animals only; N ≥ 10 per species; body mass span ≥ 3 orders of magnitude.",
          "WBE reference: β = 0.75."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
}
```

```{r filtered-intraspecific-histogram, fig.width=10, fig.height=5}
if (!is.null(intra_slopes_filt) && nrow(intra_slopes_filt) >= 5) {
  med_filt_intra <- median(intra_slopes_filt$slope, na.rm = TRUE)
  ggplot(intra_slopes_filt[!is.na(slope)], aes(x = slope)) +
    geom_histogram(binwidth = 0.15, fill = pal8[3], colour = "white", alpha = 0.85) +
    geom_vline(xintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.2) +
    geom_vline(xintercept = med_filt_intra, linetype = "solid", colour = pal8[1],
               linewidth = 1.0) +
    annotate("text", x = 0.75 + 0.06, y = Inf, label = "WBE β = 0.75",
             colour = "firebrick", vjust = 1.5, hjust = 0, size = 3.5) +
    annotate("text", x = med_filt_intra - 0.06, y = Inf,
             label = sprintf("Median β = %.3f", med_filt_intra),
             colour = pal8[1], vjust = 3, hjust = 1, size = 3.5) +
    labs(
      title    = "Filtered intraspecific slopes: animals, N ≥ 10, ≥ 3 OOM mass span",
      subtitle = sprintf("N = %d species; dashed = WBE 0.75; solid = median",
                         sum(!is.na(intra_slopes_filt$slope))),
      x        = "OLS slope (β) per species",
      y        = "Number of species",
      caption  = "Filters: animals only; N ≥ 10 per species; max/min body mass ≥ 1000× within species."
    ) +
    theme_bw(base_size = 13)
}
```

```{r filtered-intraspecific-forest, fig.width=10, fig.height=6}
if (!is.null(intra_slopes_filt) && nrow(intra_slopes_filt) >= 3) {
  ggplot(intra_slopes_filt[!is.na(slope)],
         aes(x = reorder(species, slope), y = slope,
             ymin = ci_lo, ymax = ci_hi)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 1.0) +
    geom_errorbar(width = 0.25, colour = "grey50", linewidth = 0.8) +
    geom_point(size = 3.0, colour = pal8[1]) +
    coord_flip() +
    labs(
      title    = "Filtered intraspecific metabolic scaling slopes",
      subtitle = "Points = OLS slope; error bars = 95% CI; dashed = WBE β = 0.75",
      x        = "Species",
      y        = "OLS slope (β)",
      caption  = "Filters: animals only; N ≥ 10 per species; max/min body mass ≥ 1000× within species."
    ) +
    theme_bw(base_size = 13)
}
```

## Summary comparison: unfiltered vs filtered

```{r filtered-comparison-table}
# Collect unfiltered mean/median slopes per level
unf_rows <- rbindlist(Filter(Negate(is.null), list(
  if (length(ord_slopes_vec)   >= 3) slope_summary_row(ord_slopes_vec,   "Order (unfiltered)"),
  if (length(fam_slopes_vec)   >= 3) slope_summary_row(fam_slopes_vec,   "Family (unfiltered)"),
  if (length(gen_slopes_vec)   >= 3) slope_summary_row(gen_slopes_vec,   "Genus (unfiltered)"),
  if (length(intra_slopes_vec) >= 3) slope_summary_row(intra_slopes_vec, "Intraspecific (unfiltered)")
)), fill = TRUE)

# Collect filtered mean/median slopes per level
filt_ord_vec   <- if (!is.null(reg_order_filt) && nrow(reg_order_filt) >= 3)
                    reg_order_filt$Slope[!is.na(reg_order_filt$Slope)] else numeric(0)
filt_fam_vec   <- if (!is.null(reg_fam_filt) && nrow(reg_fam_filt) >= 3)
                    reg_fam_filt$Slope[!is.na(reg_fam_filt$Slope)] else numeric(0)
filt_gen_vec   <- if (!is.null(reg_genus_filt) && nrow(reg_genus_filt) >= 3)
                    reg_genus_filt$Slope[!is.na(reg_genus_filt$Slope)] else numeric(0)
filt_intra_vec <- if (!is.null(intra_slopes_filt) && nrow(intra_slopes_filt) >= 3)
                    intra_slopes_filt$slope[!is.na(intra_slopes_filt$slope)] else numeric(0)

filt_rows <- rbindlist(Filter(Negate(is.null), list(
  if (length(filt_ord_vec)   >= 3) slope_summary_row(filt_ord_vec,   "Order (filtered)"),
  if (length(filt_fam_vec)   >= 3) slope_summary_row(filt_fam_vec,   "Family (filtered)"),
  if (length(filt_gen_vec)   >= 3) slope_summary_row(filt_gen_vec,   "Genus (filtered)"),
  if (length(filt_intra_vec) >= 3) slope_summary_row(filt_intra_vec, "Intraspecific (filtered)")
)), fill = TRUE)

comp_table <- rbind(unf_rows, filt_rows, fill = TRUE)

if (!is.null(comp_table) && nrow(comp_table) >= 2) {
  kable(comp_table,
        col.names = c("Level", "N groups", "Mean β", "95% CI lo (mean)",
                      "95% CI hi (mean)", "Median β", "95% CI lo (median)",
                      "95% CI hi (median)"),
        caption = paste(
          "Comparison of unfiltered vs filtered slope distributions across taxonomic levels.",
          "Filtered: animals only; N ≥ 10 per group; body mass span ≥ 3 OOM.",
          "WBE reference: β = 0.75.",
          "Mean CI: t-distribution; Median CI: bootstrap percentile (R = 5000)."
        )) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("Insufficient data for unfiltered vs filtered comparison table.\n")
}
```

```{r filtered-comparison-plot, fig.width=10, fig.height=7}
if (!is.null(comp_table) && nrow(comp_table) >= 2) {
  # Derive base level label and filter status for faceting
  comp_table[, base_level := gsub(" \\(.*\\)", "", Level)]
  comp_table[, filter_status := ifelse(grepl("filtered", Level, ignore.case = TRUE),
                                        "Filtered", "Unfiltered")]

  comp_long <- rbind(
    comp_table[, .(base_level, filter_status, N_groups,
                   Estimate = Mean_slope, CI_lo = CI95_lo_mean, CI_hi = CI95_hi_mean,
                   Type = "Mean")],
    comp_table[, .(base_level, filter_status, N_groups,
                   Estimate = Median_slope, CI_lo = CI95_lo_med, CI_hi = CI95_hi_med,
                   Type = "Median")]
  )
  comp_long[, base_level := factor(base_level,
                                    levels = c("Order", "Family", "Genus", "Intraspecific"))]

  ggplot(comp_long,
         aes(x = filter_status, y = Estimate, colour = filter_status,
             ymin = CI_lo, ymax = CI_hi, shape = Type)) +
    geom_hline(yintercept = 0.75, linetype = "dashed", colour = "firebrick",
               linewidth = 0.9) +
    geom_errorbar(aes(ymin = CI_lo, ymax = CI_hi),
                  width = 0.2, position = position_dodge(width = 0.5),
                  linewidth = 0.9) +
    geom_point(size = 3.5, position = position_dodge(width = 0.5)) +
    facet_wrap(~ base_level, scales = "free_y", ncol = 2) +
    scale_colour_manual(values = c("Unfiltered" = pal8[2], "Filtered" = pal8[1]),
                        name = "Dataset") +
    scale_shape_manual(values = c("Mean" = 16, "Median" = 17), name = "Statistic") +
    annotate("text", x = 0.55, y = 0.75, label = "WBE 0.75",
             colour = "firebrick", size = 3.0, hjust = 0, vjust = -0.4) +
    labs(
      title    = "Unfiltered vs filtered metabolic scaling slopes by taxonomic level",
      subtitle = "Filtered: animals only; N ≥ 10 per group; body mass span ≥ 3 OOM",
      x        = NULL,
      y        = "OLS slope (β)",
      caption  = "Error bars: t-CI (mean) and bootstrap percentile CI (median); dashed = WBE β = 0.75."
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
} else {
  cat("Insufficient data for unfiltered vs filtered comparison plot.\n")
}
```

---

# References

West GB, Brown JH, Enquist BJ. 1997. A general model for the origin of allometric scaling laws in biology. *Science* 276:122–126. DOI: [10.1126/science.276.5309.122](https://doi.org/10.1126/science.276.5309.122)

West GB, Brown JH, Enquist BJ. 1999. The fourth dimension of life: Fractal geometry and allometric scaling of organisms. *Science* 284:1677–1679. DOI: [10.1126/science.284.5420.1677](https://doi.org/10.1126/science.284.5420.1677)

Brown JH, Gillooly JF, Allen AP, Savage VM, West GB. 2004. Toward a metabolic theory of ecology. *Ecology* 85:1771–1789. DOI: [10.1890/03-9000](https://doi.org/10.1890/03-9000)

Hatton IA, McCann KS, Fryxell JM, Davies TJ, Smerlak M, Sinclair ARE, Loreau M. 2015. The predator-prey power law: Biomass scaling across terrestrial and aquatic biomes. *Science* 349(6252):aac6284. DOI: [10.1126/science.aac6284](https://doi.org/10.1126/science.aac6284)

Hatton IA, Dobson AP, Storch D, Galbraith ED, Loreau M. 2019. Linking scaling laws across eukaryotes. *PNAS* 116(43):21616–21622. DOI: [10.1073/pnas.1907703116](https://doi.org/10.1073/pnas.1907703116)

Herberstein ME, Wignall AE, Nessler SH, Harmer AMT, Wesner JS, Fagan WF, Mirth CK, Seebacher F, Wright CM, Chung AYC, et al. 2022. AnimalTraits — a curated animal trait database for body mass, metabolic rate and brain size. *Scientific Data* 9:483. DOI: [10.1038/s41597-022-01364-9](https://doi.org/10.1038/s41597-022-01364-9)

Hoehler TM, Mankel DJ, Girguis PR, McCollom TM, Kiang NY, Bebout BM. 2023. The metabolic rate of the biosphere and its components. *PNAS* 120(25):e2303764120. DOI: [10.1073/pnas.2303764120](https://doi.org/10.1073/pnas.2303764120)

Warton DI, Duursma RA, Falster DS, Taskinen S. 2012. smatr 3 — an R package for estimation and inference about allometric lines. *Methods in Ecology and Evolution* 3:257–259. DOI: [10.1111/j.2041-210X.2011.00153.x](https://doi.org/10.1111/j.2041-210X.2011.00153.x)

Chown SL, Marais E, Terblanche JS, Klok CJ, Lighton JRB, Blackburn TM. 2007. Scaling of insect metabolic rate is inconsistent with the nutrient supply network model. *Functional Ecology* 21(2):282–290. DOI: [10.1111/j.1365-2435.2007.01245.x](https://doi.org/10.1111/j.1365-2435.2007.01245.x)

Riveros AJ, Enquist BJ. 2011. Metabolic scaling in insects supports the predictions of the WBE model. *Journal of Insect Physiology* 57:688–693. DOI: [10.1016/j.jinsphys.2011.01.011](https://doi.org/10.1016/j.jinsphys.2011.01.011)

Grady JM, Enquist BJ, Dettweiler-Robinson E, Wright NA, Smith FA. 2014. Evidence for mesothermy in dinosaurs. *Science* 344(6189):1268–1272. DOI: [10.1126/science.1253143](https://doi.org/10.1126/science.1253143)

